---
title: "DHW–Mortality Attribution on the Great Barrier Reef"
author: "Samuel Matthews"
date: "2024-05-29"
date-modified: today
# bibliography: references.bib
# link-citations: true
format:
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 3
    toc-location: left
    self-contained: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
execute:
  warning: false
  message: false
  echo: true
---

# Introduction

This document quantifies the relationship between accumulated thermal stress (Degree Heating Weeks; DHW) and coral mortality on the Great Barrier Reef (GBR). The analysis extends existing DHW–mortality frameworks (e.g., Hughes et al. 2018, ReefMod) by:

1. **Growth-adjusting** mortality estimates using reef-specific Gompertz coral growth parameters from @macneil2019, so that expected background growth is accounted for when estimating bleaching-attributable loss.
2. **Comparing data sources**: manta tow (broad spatial coverage, ~6–9 m depth), benthic photo-transects (precise, multiple depths), and MMP inshore monitoring.
3. **Stratifying by era and region** to examine whether the DHW–mortality relationship has changed across successive mass bleaching events (1998/2002, 2016/2017, 2020/2022, 2024).
4. **Assessing repeat exposure effects** — do reefs that have experienced prior bleaching show different mortality responses?
5. **Building a machine-learning mortality model** by integrating environmental predictors from Cheung et al. (2025) — SST trajectory clusters, thermal history, water clarity, currents, and cloud cover.

## Key Assumptions & Caveats

- Gompertz growth parameters are averaged at reef level from 1 km grid cells; future work should propagate parameter uncertainty.
- Mortality attribution assumes bleaching is the primary driver in filtered bleaching years; co-occurring disturbances (COTS, cyclones) are excluded via LTMP disturbance records but residual confounding is possible.
- The Cheung et al. predictor dataset covers 2016, 2017, 2020 only; the ML model is restricted to these years as a proof-of-concept.

```{r setup}
#| label: setup
# ── Core libraries ──
library(tidyr)
library(dplyr)
library(ggplot2)
library(DT)
library(arrow)
library(stringr)
library(janitor)
library(lubridate)
library(readr)
library(sf)
library(lwgeom)
library(scales)
library(patchwork)

# ── Modelling libraries ──
library(mgcv)
library(betareg)
library(MASS)
library(lme4)
library(sandwich)
library(lmtest)

# ── Shared colour palettes ──
dist_colours <- c(
    "b" = "darkorange", "B" = "darkorange",
    "c" = "maroon4", "C" = "maroon4",
    "d" = "darkolivegreen3", "D" = "darkolivegreen3",
    "f" = "burlywood4", "F" = "burlywood4",
    "s" = "dodgerblue2", "S" = "dodgerblue2",
    "u" = "grey", "U" = "grey"
)

cols_exposure <- c("First" = "#E69F00", "Repeat" = "#0072B2")
cols_outbreak <- c("COTS Outbreak" = "#D55E00", "No Outbreak" = "#009E73")

secorder <- c("CG", "PC", "CL", "CA", "IN", "TO", "CU", "WH", "PO", "SW", "CB")
secorderlabs <- c(
    "Cape Grenville", "Princess Charlotte Bay", "Cooktown / Lizard Island",
    "Cairns", "Innisfail", "Townsville", "Cape Upstart", "Whitsunday",
    "Pompey", "Swain", "Capricorn Bunker"
)

# Force dplyr functions over MASS/stats
select <- dplyr::select
filter <- dplyr::filter
```

```{r helper-functions}
#| label: helper-functions
# ── Gompertz coral growth function (simplified, single time-step) ──
# Returns c(current_cover, next_cover, change)
doCoralGrowth.Simp <- function(CoralCover, B0, HC.asym, WQ) {
    WQ.mn.sd <- c(-0.68, 0.03)
    b0.wq <- B0 + WQ * WQ.mn.sd[1]
    b1.wq <- b0.wq / log(HC.asym)
    CoralCover <- log(CoralCover)
    b0.wq <- as.numeric(b0.wq)
    b1.wq <- as.numeric(b1.wq)
    CoralCover <- as.numeric(CoralCover)
    CoralCover.t1 <- (b0.wq + (1 - b1.wq) * CoralCover)
    CoralCover.t1 <- exp(CoralCover.t1)
    Change.t1 <- CoralCover.t1 - exp(CoralCover)
    return(c(exp(CoralCover), CoralCover.t1, Change.t1))
}

# ── Multi-step growth function (n years from starting cover) ──
doCoralGrowth <- function(n, CoralCover, B0, HC.asym, WQ) {
    WQ.mn.sd <- c(-0.68, 0.03)
    b0.wq <- B0 + WQ * WQ.mn.sd[1]
    b1.wq <- b0.wq / log(HC.asym)
    CoralCover <- log(CoralCover)
    b0.wq <- as.numeric(b0.wq)
    b1.wq <- as.numeric(b1.wq)
    CoralCover <- as.numeric(CoralCover)
    df <- matrix(nrow = n + 1, ncol = 3)
    df[1, 1] <- 0
    df[1, 2] <- exp(CoralCover)
    for (i in 1:n) {
        CoralCover.t1 <- (b0.wq + (1 - b1.wq) * CoralCover)
        df[i + 1, 1] <- i
        df[i + 1, 2] <- exp(CoralCover.t1)
        df[i + 1, 3] <- exp(CoralCover.t1) - exp(CoralCover)
        CoralCover <- CoralCover.t1
    }
    colnames(df) <- c("Year", "CoralCover", "CoralGrowth")
    return(df)
}

# ── Inverse logit ──
inv_logit <- function(x) 1 / (1 + exp(-x))
```

# Growth Curves & Recovery Times

Reef-specific Gompertz growth parameters from @macneil2019 are used to model coral cover trajectories. The parameters (`B0` = intrinsic growth rate, `HC.asym` = asymptotic carrying capacity) are modified by local water quality (`WQ`) to produce site-specific recovery curves.

The growth function follows the discrete Gompertz form:
$$\log(C_{t+1}) = b_0^{WQ} + (1 - b_1^{WQ}) \cdot \log(C_t)$$

where $b_0^{WQ} = B_0 + WQ \times \beta_{WQ}$ and $b_1^{WQ} = b_0^{WQ} / \log(HC_{asym})$.

```{r growth-params}
#| label: growth-params
#| cache: true
dat.grid <- read.csv("Data/COTSModParams.csv")

dat.gridSum_base <- dat.grid %>%
    mutate(
        REEF_ID = ifelse(grepl("14-116", REEF_ID, fixed = TRUE), "14-116", REEF_ID),
        REEF_NAME = ifelse(grepl("14-116", REEF_NAME, fixed = TRUE),
            "Lizard Island Reef (14-116)", REEF_NAME
        )
    ) %>%
    group_by(REEF_NAME, REEF_ID, SECTOR) %>%
    summarise(
        B0 = mean(pred.b0.mean, na.rm = TRUE),
        HC.asym = mean(pred.HCmax.mean, na.rm = TRUE),
        WQ1 = mean(Primary, na.rm = TRUE),
        WQ2 = mean(Secondary, na.rm = TRUE),
        WQ3 = mean(Tertiary, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    mutate(
        WQ = WQ1 + WQ2 + WQ3,
        HC90 = 0.8 * HC.asym
    )

# Sector averages for fallback on ungridded inshore reefs
sector_params <- dat.gridSum_base %>%
    group_by(SECTOR) %>%
    summarise(
        B0 = mean(B0, na.rm = TRUE),
        HC.asym = mean(HC.asym, na.rm = TRUE),
        WQ = mean(WQ, na.rm = TRUE),
        .groups = "drop"
    )

# Sister reef expansions:
# 1. Normanby-Mabel Reef (17-012) uses Round-Russell Reef (17-013) parameters
nm_row <- dat.gridSum_base %>%
    filter(grepl("17-013", REEF_ID)) %>%
    head(1) %>%
    mutate(REEF_NAME = "Normanby-Mabel Reef (17-012)", REEF_ID = "17-012")

# 2. Lizard Island Reef (North West) (14-116a) from Lizard Island (14-116)
liz_params <- dat.gridSum_base %>%
    filter(grepl("14-116", REEF_ID)) %>%
    summarise(B0 = mean(B0), HC.asym = mean(HC.asym), WQ1 = mean(WQ1), WQ2 = mean(WQ2), WQ3 = mean(WQ3), WQ = mean(WQ), HC90 = mean(HC90)) %>%
    mutate(REEF_NAME = "Lizard Island Reef (North West) (14-116a)", REEF_ID = "14-116a", SECTOR = "CL")

# 3. Hook Island Reef (No 4) (20-028d) from Hook Island (20-028)
hook_params <- dat.gridSum_base %>%
    filter(grepl("20-028", REEF_ID)) %>%
    summarise(B0 = mean(B0), HC.asym = mean(HC.asym), WQ1 = mean(WQ1), WQ2 = mean(WQ2), WQ3 = mean(WQ3), WQ = mean(WQ), HC90 = mean(HC90)) %>%
    mutate(REEF_NAME = "Hook Island Reef (No 4) (20-028d)", REEF_ID = "20-028d", SECTOR = "WH")

# 4. Magnetic Island Reef (No 2) (19-009c) from Cockle Bay (19-009a)
mag_params <- dat.gridSum_base %>%
    filter(grepl("19-009", REEF_ID)) %>%
    summarise(B0 = mean(B0), HC.asym = mean(HC.asym), WQ1 = mean(WQ1), WQ2 = mean(WQ2), WQ3 = mean(WQ3), WQ = mean(WQ), HC90 = mean(HC90)) %>%
    mutate(REEF_NAME = "Magnetic Island Reef (No 2) (19-009c)", REEF_ID = "19-009c", SECTOR = "TO")

dat.gridSum <- bind_rows(dat.gridSum_base, nm_row, liz_params, hook_params, mag_params)
```

```{r growth-curves}
#| label: growth-curves
#| cache: true
#| fig-cap: "Modelled coral recovery trajectories from 0.1% cover, by sector. Ribbons show inter-reef variability."
# Simulate 30-year recovery from 0.1% cover for all reefs
CC.tmp <- data.frame()
for (i in seq_len(nrow(dat.gridSum))) {
    CCtmp2 <- as.data.frame(
        doCoralGrowth(30, 0.1, dat.gridSum$B0[i], dat.gridSum$HC.asym[i], dat.gridSum$WQ[i])
    )
    CCtmp2$ReefName <- dat.gridSum$REEF_NAME[i]
    CCtmp2$REEF_NAME <- dat.gridSum$REEF_NAME[i]
    CCtmp2$ReefID <- dat.gridSum$REEF_ID[i]
    CCtmp2$Sector <- dat.gridSum$SECTOR[i]
    CC.tmp <- bind_rows(CC.tmp, CCtmp2)
}

CC.tmp <- CC.tmp %>% left_join(select(dat.gridSum, REEF_NAME, HC90), by = "REEF_NAME")

library(ggdist)
ggplot(CC.tmp, aes(x = Year, y = CoralCover)) +
    stat_lineribbon(alpha = 0.3) +
    facet_wrap(~Sector) +
    theme_classic() +
    theme(legend.position = "none") +
    labs(x = "Years since disturbance", y = "Coral cover (%)")
```



# DHW–Mortality Relationship

This section examines the existing DHW–mortality relationship using the Hughes et al. (2018) dataset, comparing beta regression versus binomial/quasi-binomial GLM approaches. The key question is whether the sigmoidal binomial curve overpredicts mortality at high DHW values.

## Hughes et al. Data: Beta vs Binomial Regression

```{r dhw-mortality-comparison}
#| label: dhw-mortality-comparison
dat.bleach <- read.csv("Data/DHW_vs_Coral_Cover_Dataset.csv") %>%
    mutate(Mort.bin = ifelse(Change >= 0, 0.001, -Change / 100))

mod.beta <- betareg(Mort.bin ~ DHW, data = dat.bleach)
mod.bin <- glm(Mort.bin ~ DHW, data = dat.bleach, family = binomial(link = "logit"))

# Predictions
newdata <- with(dat.bleach, data.frame(DHW = seq(0, 16, len = 100)))
fit_bin <- predict(mod.bin, newdata = newdata, type = "link", se = TRUE)
q <- qt(0.975, df = df.residual(mod.bin))
newdata <- cbind(newdata,
    fit = binomial()$linkinv(fit_bin$fit),
    lower = binomial()$linkinv(fit_bin$fit - q * fit_bin$se.fit),
    upper = binomial()$linkinv(fit_bin$fit + q * fit_bin$se.fit)
)

newdata2 <- with(dat.bleach, data.frame(DHW = seq(0, 16, len = 100)))
fit_beta <- predict(mod.beta, newdata = newdata2, type = "link")
newdata2 <- cbind(newdata2, fit = binomial()$linkinv(fit_beta))
```

```{r dhw-mortality-plot}
#| label: dhw-mortality-plot
#| fig-cap: "Comparison of binomial (black) and beta regression (red) fits to the DHW–mortality relationship from Hughes et al. (2018) data."
ggplot(data = newdata, aes(y = fit, x = DHW)) +
    geom_point(data = dat.bleach, aes(y = Mort.bin)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "blue", alpha = 0.3) +
    geom_line() +
    geom_line(data = newdata2, aes(y = fit, x = DHW), colour = "red") +
    theme_classic() +
    labs(
        x = "DHW (°C-weeks)", y = "Mortality proportion",
        caption = "Black = binomial GLM, Red = beta regression"
    )
```


# Manta Tow Analysis (Updated AIMS API)

This section uses the locally cached AIMS LTMP manta tow data to construct a comprehensive DHW–mortality dataset. Growth correction is applied using reef-specific Gompertz parameters.

```{r manta-tow-data}
#| label: manta-tow-data
#| cache: true
load("Data/aims_ltmp/aims_ltmp.RData")

dat.AIMSRef <- read.csv("Data/AIMS-Reef_Reference.csv") %>%
    rename(Reef_Name = AIMS_REEF_NAME)

dat.Ref.Clean <- dat.AIMSRef %>%
    mutate(AIMS_REEF_NAME_Clean = gsub(" REEF(S)?| ISLAND(S)?| IS", "", toupper(Reef_Name))) %>%
    group_by(AIMS_REEF_NAME_Clean) %>%
    slice(1) %>%
    ungroup()

# Load complete unified disturbance dataset (from AIMS API)
Disturbances_raw <- read.csv("data/aims_ltmp/reef_disturbance.csv")

Disturbances_clean <- Disturbances_raw %>%
    mutate(
        Reef_Clean = gsub(" REEF(S)?| ISLAND(S)?| IS", "", toupper(aims_reef_name))
    ) %>%
    left_join(select(dat.Ref.Clean, AIMS_REEF_NAME_Clean, ReefName, ReefID),
        by = c("Reef_Clean" = "AIMS_REEF_NAME_Clean")
    ) %>%
    mutate(
        ReefName = case_when(
            toupper(aims_reef_name) == "FRANKLAND ISLANDS" & reef_zone == "East" ~ "Normanby-Mabel Reef (17-012)",
            toupper(aims_reef_name) == "FRANKLAND ISLANDS" & reef_zone == "West" ~ "Round-Russell Reef (17-013)",
            TRUE ~ ReefName
        )
    ) %>%
    rename(report_year = year, DISTURBANCE_TYPE = disturbance)

# Prioritize PPOINT for benthic photo transects, MANTA for manta tow
Disturbances_bent <- Disturbances_clean %>%
    arrange(ReefName, report_year, desc(sample_type == "PPOINT")) %>%
    distinct(ReefName, report_year, .keep_all = TRUE)

Disturbances_mant <- Disturbances_clean %>%
    arrange(ReefName, report_year, desc(sample_type == "MANTA")) %>%
    distinct(ReefName, report_year, .keep_all = TRUE)


# DHW data
dat.DHW <- read.csv("Data/DHW_1985_2024_GBRReefs.csv") %>%
    rename("ReefName" = "LOC_NAME_S") %>%
    mutate(
        ReefName = case_when(
            grepl("15-070", ReefName) ~ "Vicki Harriott Reef (15-070)",
            grepl("18-022", ReefName) ~ "Felicity Wishart Reef (18-022)",
            grepl("22-101", ReefName) ~ "Don Kinsey Reef (22-101)",
            grepl("23-004b", ReefName) ~ "North Keppel (Ko-no-mie) Island Reef (No 2) (23-004b)",
            grepl("15-027", ReefName) ~ "Marx Reef (15-027)",
            TRUE ~ ReefName
        )
    )

# Fill Agincourt reefs (missing individual DHW)
dat.agin <- dat.DHW |>
    filter(grepl("Agincourt", ReefName)) |>
    group_by(Year) |>
    summarise(MaxDHW.mean = mean(MaxDHW.mean, na.rm = TRUE), .groups = "drop")

dat.DHW <- dat.DHW |>
    left_join(dat.agin, by = "Year") |>
    mutate(MaxDHW.mean = ifelse(grepl("Agincourt", ReefName) & is.na(MaxDHW.mean.x),
        MaxDHW.mean.y, MaxDHW.mean.x
    )) |>
    select(-MaxDHW.mean.x, -MaxDHW.mean.y)

df.AIMS.Mant <- reef_manta_df |>
    select(
        domain_name, project_code, report_year, date, depth,
        lower:median, mean, reef_zone, id, shelf, reefpage_category
    ) |>
    mutate(Reef_Name = toupper(domain_name))

library(fuzzyjoin)

dat.AIMSRef_key <- dat.AIMSRef %>%
    filter(!ReefName %in% c("Round-Russell Reef (17-013)", "Snake Reef (14-087)")) %>%
    mutate(AIMS_REEF_NAME_case = str_to_title(AIMS_REEF_NAME_cap))

df.MANT <- df.AIMS.Mant |>
    left_join(select(dat.AIMSRef_key, AIMS_REEF_NAME_cap, ReefID, ReefName, SECT_NAME, SECTOR),
        by = join_by("Reef_Name" == "AIMS_REEF_NAME_cap")
    ) %>%
    mutate(
        year = floor(as.numeric(date)),
        frac = as.numeric(date) - year,
        date_approx = ymd(paste0(year, "-01-01")) +
            round(frac * as.numeric(ymd(paste0(year + 1, "-01-01")) - ymd(paste0(year, "-01-01"))))
    ) %>%
    select(-year, -frac)

# DHW year assignment
dat.yrs.mant <- df.MANT |>
    ungroup() |>
    filter(!is.na(ReefName)) |>
    mutate(
        Date = date_approx, Prev = report_year - 1,
        Year = year(Date), Month = month(Date),
        DHWYear = ifelse(Month >= 5, Year, Year - 1)
    ) |>
    left_join(dplyr::select(dat.DHW, ReefName, Year, MaxDHW.mean),
        by = c("ReefName", "DHWYear" = "Year")
    )

# Calculate change per reef
df.MANT <- df.MANT %>%
    arrange(ReefName, Reef_Name, reefpage_category, depth, report_year, date_approx) |>
    group_by(ReefName, Reef_Name, depth, reefpage_category) |>
    mutate(
        Change = mean - lag(mean, n = 1),
        Rel.Change = Change / lag(mean, n = 1),
        Lag = report_year - lag(report_year, n = 1),
        Baseline.Date = lag(date_approx, n = 1),
        Interval.Days = as.integer(date_approx - Baseline.Date)
    ) |>
    relocate(c("Change", "Lag", "Rel.Change"), .before = median) |>
    left_join(dat.yrs.mant)

# Growth correction
df.MANT <- df.MANT |>
    mutate(Change.Cor = Change, Prev.HC = mean - Change.Cor) |>
    filter(
        ReefName %in% dat.gridSum$REEF_NAME, !is.na(Lag),
        Reef_Name != "MILLN REEF"
    ) |>
    rowwise() |>
    mutate(
        Pred.Grth = doCoralGrowth.Simp(
            CoralCover = Prev.HC * 100,
            B0 = dat.gridSum[which(dat.gridSum$REEF_NAME == ReefName), "B0"],
            WQ = dat.gridSum[which(dat.gridSum$REEF_NAME == ReefName), "WQ"],
            HC.asym = dat.gridSum[which(dat.gridSum$REEF_NAME == ReefName), "HC.asym"]
        )[3] / 100
    ) |>
    mutate(
        Pred.Grth2 = ifelse(Lag == 2,
            Pred.Grth + doCoralGrowth.Simp(
                CoralCover = (Prev.HC + Pred.Grth) * 100,
                B0 = dat.gridSum[which(dat.gridSum$REEF_NAME == ReefName), "B0"],
                WQ = dat.gridSum[which(dat.gridSum$REEF_NAME == ReefName), "WQ"],
                HC.asym = dat.gridSum[which(dat.gridSum$REEF_NAME == ReefName), "HC.asym"]
            )[3] / 100,
            Pred.Grth
        ),
        Abs.Change = Change.Cor - Pred.Grth2,
        Rel.Change = Abs.Change / (Prev.HC + Pred.Grth2)
    )
```

## Disturbance Attribution (100% Stacked Bar & Composite Plots)

This section calculates total coral loss attributable to each disturbance type in the AIMS LTMP Manta Tow survey dataset. 
For survey observations where multiple disturbances (`m`) were recorded, we parse detailed `description` and `tooltip` fields from `reef_disturbance.csv` to identify constituent drivers (Bleaching, COTS, Cyclones/Storms, Disease) and evenly partition the coral loss across those drivers.

We present academic composite visualizations linking long-term temporal loss trends (Mean % cover loss per affected reef $\pm$ Standard Error) with 100% stacked disturbance attribution across both four broad decadal epochs (1986–1995, 1996–2005, 2006–2015, 2016–2025) and individual survey years.

```{r disturbance-attribution-stacked-bars}
#| label: disturbance-attribution-stacked-bars
#| fig.width: 11
#| fig.height: 8.5
#| dpi: 300
#| warning: false
#| message: false

library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)
library(patchwork)
library(scales)

# Load AIMS LTMP disturbance metadata
reef_dist <- read.csv("data/aims_ltmp/reef_disturbance.csv") %>%
  filter(sample_type == "MANTA") %>%
  distinct(aims_reef_name, year, .keep_all = TRUE)

# Filter manta dataset to all surveyed reefs (raw_loss = 0 if Change >= 0) & assign decadal/regional groups
df_surveyed <- df.MANT %>%
  filter(!is.na(Lag), Lag <= 3) %>%
  left_join(reef_dist, by = c("domain_name" = "aims_reef_name", "report_year" = "year")) %>%
  mutate(
    disturbance_cat = case_when(
      is.na(disturbance) | disturbance == "" ~ "u",
      TRUE ~ disturbance
    ),
    net_change = Change * 100,
    raw_loss = ifelse(Change < 0, -Change, 0),
    Decade = case_when(
      report_year >= 1986 & report_year <= 1995 ~ "1986–1995",
      report_year >= 1996 & report_year <= 2005 ~ "1996–2005",
      report_year >= 2006 & report_year <= 2015 ~ "2006–2015",
      report_year >= 2016 & report_year <= 2025 ~ "2016–2025",
      TRUE ~ "Other"
    ),
    Region = case_when(
      SECTOR %in% c("PB", "CG", "CL", "CA") ~ "Northern GBR",
      SECTOR %in% c("IN", "TO", "CU", "WH") ~ "Central GBR",
      SECTOR %in% c("PO", "SW", "CB")       ~ "Southern GBR",
      TRUE ~ "Other"
    )
  ) %>%
  filter(Decade != "Other")

# Filtered to positive loss for disturbance attribution stacked bar partitioning
df_loss <- df_surveyed %>% filter(raw_loss > 0)

# Function to parse text description for 'm' records
parse_m_record <- function(desc, tooltip, storm_name) {
  text <- paste(coalesce(desc, ""), coalesce(tooltip, ""), coalesce(storm_name, ""), sep = " ")
  
  has_b <- grepl("bleach|bl |dhw", text, ignore.case = TRUE)
  has_c <- grepl("cots|crown-of-thorns", text, ignore.case = TRUE)
  has_s <- grepl("cyclone|storm|tc |jasper|steve|nathan|ita|yasi|kirrily|alfred|larry|hamish|felicity|marcia|debbie|rona|grace|tessi|joy|charlie", text, ignore.case = TRUE)
  has_d <- grepl("disease|ws", text, ignore.case = TRUE)
  
  types <- c()
  if (has_b) types <- c(types, "b")
  if (has_c) types <- c(types, "c")
  if (has_s) types <- c(types, "s")
  if (has_d) types <- c(types, "d")
  
  if (length(types) == 0) types <- "u"
  return(types)
}

# --- Plot 1 Dataset: 'm' as its own category ---
df_plot1 <- df_loss %>%
  mutate(
    Disturbance = case_when(
      disturbance_cat == "b" ~ "Bleaching",
      disturbance_cat == "c" ~ "COTS",
      disturbance_cat == "s" ~ "Cyclone/Storm",
      disturbance_cat == "d" ~ "Disease",
      disturbance_cat == "m" ~ "Multiple (m)",
      TRUE ~ "Unknown/Other"
    )
  )

# --- Plot 2 Dataset: 'm' partitioned evenly ---
partitioned_rows <- list()
for(i in 1:nrow(df_loss)) {
  row_i <- df_loss[i, ]
  if (row_i$disturbance_cat == "m") {
    matched <- parse_m_record(row_i$description, row_i$tooltip, row_i$storm_name)
    n_match <- length(matched)
    split_loss <- row_i$raw_loss / n_match
    for(m_type in matched) {
      new_row <- row_i
      new_row$disturbance_cat <- m_type
      new_row$raw_loss <- split_loss
      partitioned_rows[[length(partitioned_rows) + 1]] <- new_row
    }
  } else {
    partitioned_rows[[length(partitioned_rows) + 1]] <- row_i
  }
}
df_partitioned <- bind_rows(partitioned_rows)

df_plot2 <- df_partitioned %>%
  mutate(
    Disturbance = case_when(
      disturbance_cat == "b" ~ "Bleaching",
      disturbance_cat == "c" ~ "COTS",
      disturbance_cat == "s" ~ "Cyclone/Storm",
      disturbance_cat == "d" ~ "Disease",
      TRUE ~ "Unknown/Other"
    )
  )

# Academic Color Palette (COTS = Vibrant Academic Purple #7B1FA2)
dist_colors_p1 <- c(
  "Bleaching"      = "#D95F02", # Red-Orange
  "COTS"           = "#7B1FA2", # Purple
  "Cyclone/Storm"  = "#2B5C8F", # Navy
  "Disease"        = "#8C510A", # Dark Ochre
  "Multiple (m)"   = "#1B9E77", # Teal
  "Unknown/Other"  = "#7F7F7F"  # Grey
)

dist_colors_p2 <- c(
  "Bleaching"      = "#D95F02",
  "COTS"           = "#7B1FA2",
  "Cyclone/Storm"  = "#2B5C8F",
  "Disease"        = "#8C510A",
  "Unknown/Other"  = "#7F7F7F"
)

# Custom Academic Theme
theme_academic <- function(base_size = 12) {
  theme_bw(base_size = base_size) +
    theme(
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
      panel.grid.major.y = element_line(color = "#E5E7EB", linewidth = 0.4),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.title = element_text(face = "bold", size = 12, color = "black"),
      axis.text = element_text(color = "black", size = 10),
      axis.ticks = element_line(color = "black", linewidth = 0.5),
      legend.title = element_text(face = "bold", size = 11),
      legend.text = element_text(size = 10),
      plot.title = element_text(face = "bold", size = 14, hjust = 0),
      plot.subtitle = element_text(size = 11, color = "#333333", hjust = 0),
      strip.background = element_rect(fill = "#F3F4F6", color = "black", linewidth = 0.8),
      strip.text = element_text(face = "bold", size = 11)
    )
}

# --- DECADAL COMPOSITE VISUALIZATION ---
decadal_loss_stats <- df_surveyed %>%
  group_by(Decade) %>%
  summarise(
    N = n(),
    Mean_Loss = mean(raw_loss * 100),
    SD_Loss = sd(raw_loss * 100),
    SE_Loss = SD_Loss / sqrt(N),
    .groups = "drop"
  )

p_top_decadal <- ggplot(decadal_loss_stats, aes(x = Decade, y = Mean_Loss)) +
  geom_bar(stat = "identity", fill = "#374151", width = 0.45, alpha = 0.85, color = "black") +
  geom_errorbar(aes(ymin = Mean_Loss - SE_Loss, ymax = Mean_Loss + SE_Loss),
                width = 0.15, linewidth = 0.8, color = "black") +
  geom_text(aes(label = sprintf("%.1f%% ± %.2f", Mean_Loss, SE_Loss)),
            vjust = -1.2, size = 3.8, fontface = "bold") +
  scale_y_continuous(limits = c(0, 6), breaks = seq(0, 6, 2), expand = c(0, 0)) +
  labs(
    title = "Coral Disturbance Attribution Across Four Decades (1986–2025)",
    subtitle = "A: Mean Coral Cover Loss per Surveyed Reef (% cover ± SE)",
    x = NULL,
    y = "Mean Loss (% cover)"
  ) +
  theme_academic(base_size = 11) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

# Decadal Plot 1: Unpartitioned
p1_dec_summary <- df_plot1 %>%
  group_by(Decade, Disturbance) %>%
  summarise(Total_Loss = sum(raw_loss), .groups = "drop_last") %>%
  mutate(
    Decade_Total = sum(Total_Loss),
    Prop = Total_Loss / Decade_Total,
    Label = ifelse(Prop >= 0.03, sprintf("%.1f%%", Prop * 100), "")
  )

p_bot_decadal_p1 <- ggplot(p1_dec_summary, aes(x = Decade, y = Prop, fill = factor(Disturbance, levels = names(dist_colors_p1)))) +
  geom_bar(stat = "identity", width = 0.55, color = "black", linewidth = 0.3) +
  geom_text(aes(label = Label), position = position_stack(vjust = 0.5),
            color = "white", fontface = "bold", size = 3.6) +
  scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
  scale_fill_manual(values = dist_colors_p1, name = "Disturbance Type") +
  labs(
    subtitle = "B: 100% Stacked Bar Chart of Disturbance Attribution (Plot 1: 'm' as Separate Category)",
    x = "Decade",
    y = "Proportion of Total Loss (%)"
  ) +
  theme_academic(base_size = 11) +
  theme(legend.position = "bottom")

composite_decadal_p1 <- p_top_decadal / p_bot_decadal_p1 + plot_layout(heights = c(1, 2.2))

# Decadal Plot 2: Partitioned
p2_dec_summary <- df_plot2 %>%
  group_by(Decade, Disturbance) %>%
  summarise(Total_Loss = sum(raw_loss), .groups = "drop_last") %>%
  mutate(
    Decade_Total = sum(Total_Loss),
    Prop = Total_Loss / Decade_Total,
    Label = ifelse(Prop >= 0.03, sprintf("%.1f%%", Prop * 100), "")
  )

p_bot_decadal_p2 <- ggplot(p2_dec_summary, aes(x = Decade, y = Prop, fill = factor(Disturbance, levels = names(dist_colors_p2)))) +
  geom_bar(stat = "identity", width = 0.55, color = "black", linewidth = 0.3) +
  geom_text(aes(label = Label), position = position_stack(vjust = 0.5),
            color = "white", fontface = "bold", size = 3.6) +
  scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
  scale_fill_manual(values = dist_colors_p2, name = "Disturbance Type") +
  labs(
    subtitle = "B: 100% Stacked Bar Chart of Disturbance Attribution (Plot 2: 'm' Partitioned)",
    x = "Decade",
    y = "Proportion of Total Loss (%)"
  ) +
  theme_academic(base_size = 11) +
  theme(legend.position = "bottom")

composite_decadal_p2 <- p_top_decadal / p_bot_decadal_p2 + plot_layout(heights = c(1, 2.2))

# Display Decadal Plots
print(composite_decadal_p1)
print(composite_decadal_p2)
```

### Annual Disturbance Attribution (1987–2026)

```{r annual-disturbance-attribution-composite}
#| label: annual-disturbance-attribution-composite
#| fig.width: 13
#| fig.height: 9
#| dpi: 300
#| warning: false
#| message: false

# --- ANNUAL COMPOSITE VISUALIZATION ---
annual_loss_stats <- df_surveyed %>%
  group_by(report_year) %>%
  summarise(
    N = n(),
    Mean_Loss = mean(raw_loss * 100),
    SD_Loss = sd(raw_loss * 100),
    SE_Loss = ifelse(N > 1, SD_Loss / sqrt(N), 0),
    .groups = "drop"
  )

p_top_annual <- ggplot(annual_loss_stats, aes(x = factor(report_year), y = Mean_Loss, group = 1)) +
  geom_ribbon(aes(ymin = pmax(0, Mean_Loss - SE_Loss), ymax = Mean_Loss + SE_Loss),
              fill = "#4B5563", alpha = 0.25) +
  geom_line(color = "#111827", linewidth = 0.8) +
  geom_point(color = "#111827", size = 1.8) +
  scale_y_continuous(limits = c(0, max(annual_loss_stats$Mean_Loss + annual_loss_stats$SE_Loss, na.rm = TRUE) * 1.1),
                     expand = c(0, 0)) +
  labs(
    title = "Annual Coral Cover Loss and Disturbance Attribution (1987–2026)",
    subtitle = "A: Mean Hard Coral Cover Loss per Surveyed Reef (% cover ± SE)",
    x = NULL,
    y = "Mean Loss (% cover)"
  ) +
  theme_academic(base_size = 11) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

# Annual Plot 1: Unpartitioned
p_bot_annual_p1 <- df_plot1 %>%
  group_by(report_year, Disturbance) %>%
  summarise(Total_Loss = sum(raw_loss), .groups = "drop") %>%
  ggplot(aes(x = factor(report_year), y = Total_Loss, fill = factor(Disturbance, levels = names(dist_colors_p1)))) +
  geom_bar(position = "fill", stat = "identity", width = 0.8, color = "black", linewidth = 0.15) +
  scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
  scale_fill_manual(values = dist_colors_p1, name = "Disturbance Type") +
  labs(
    subtitle = "B: Annual 100% Stacked Bar Chart (Plot 1: 'm' as Separate Category)",
    x = "Survey Year",
    y = "Proportion of Total Loss (%)"
  ) +
  theme_academic(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    legend.position = "bottom"
  )

composite_annual_p1 <- p_top_annual / p_bot_annual_p1 + plot_layout(heights = c(1, 2.2))

# Annual Plot 2: Partitioned
p_bot_annual_p2 <- df_plot2 %>%
  group_by(report_year, Disturbance) %>%
  summarise(Total_Loss = sum(raw_loss), .groups = "drop") %>%
  ggplot(aes(x = factor(report_year), y = Total_Loss, fill = factor(Disturbance, levels = names(dist_colors_p2)))) +
  geom_bar(position = "fill", stat = "identity", width = 0.8, color = "black", linewidth = 0.15) +
  scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
  scale_fill_manual(values = dist_colors_p2, name = "Disturbance Type") +
  labs(
    subtitle = "B: Annual 100% Stacked Bar Chart (Plot 2: 'm' Partitioned)",
    x = "Survey Year",
    y = "Proportion of Total Loss (%)"
  ) +
  theme_academic(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    legend.position = "bottom"
  )

composite_annual_p2 <- p_top_annual / p_bot_annual_p2 + plot_layout(heights = c(1, 2.2))

# Display Annual Plots
print(composite_annual_p1)
print(composite_annual_p2)
```

### Regional Disturbance Attribution Faceted Visualizations (Overall GBR & Regions)

This section presents the partitioned disturbance attribution (`m` partitioned) for the **Overall GBR** baseline stacked above the 3 regional facets (**Northern GBR**, **Central GBR**, **Southern GBR**) with a single shared legend at the bottom.

```{r regional-disturbance-attribution-composite}
#| label: regional-disturbance-attribution-composite
#| fig.width: 14
#| fig.height: 16
#| dpi: 300
#| warning: false
#| message: false

# ==============================================================================
# 1. DECADAL STACKED COMPOSITE (6-PANEL HIERARCHY WITH NET CHANGE GRADIENT)
# ==============================================================================

# --- Overall GBR Decadal Aggregations ---
overall_dec_net <- df_surveyed %>%
  group_by(Decade) %>%
  summarise(
    N = n(),
    Mean_Change = mean(net_change, na.rm = TRUE),
    SE_Change = ifelse(N > 1, sd(net_change, na.rm = TRUE) / sqrt(N), 0),
    .groups = "drop"
  ) %>%
  mutate(Region = "Overall GBR")

overall_dec_loss <- df_surveyed %>%
  group_by(Decade) %>%
  summarise(
    N = n(),
    Mean_Loss = mean(raw_loss * 100),
    SE_Loss = ifelse(N > 1, sd(raw_loss * 100) / sqrt(N), 0),
    .groups = "drop"
  ) %>%
  mutate(Region = "Overall GBR")

overall_p2_dec <- df_plot2 %>%
  group_by(Decade, Disturbance) %>%
  summarise(Total_Loss = sum(raw_loss), .groups = "drop_last") %>%
  mutate(
    Decade_Total = sum(Total_Loss),
    Prop = Total_Loss / Decade_Total,
    Label = ifelse(Prop >= 0.04, sprintf("%.1f%%", Prop * 100), ""),
    Region = "Overall GBR"
  )

# --- Regional Decadal Aggregations ---
reg_dec_net <- df_surveyed %>%
  filter(Region != "Other") %>%
  group_by(Region, Decade) %>%
  summarise(
    N = n(),
    Mean_Change = mean(net_change, na.rm = TRUE),
    SE_Change = ifelse(N > 1, sd(net_change, na.rm = TRUE) / sqrt(N), 0),
    .groups = "drop"
  ) %>%
  mutate(Region = factor(Region, levels = c("Northern GBR", "Central GBR", "Southern GBR")))

reg_dec_loss <- df_surveyed %>%
  filter(Region != "Other") %>%
  group_by(Region, Decade) %>%
  summarise(
    N = n(),
    Mean_Loss = mean(raw_loss * 100),
    SE_Loss = ifelse(N > 1, sd(raw_loss * 100) / sqrt(N), 0),
    .groups = "drop"
  ) %>%
  mutate(Region = factor(Region, levels = c("Northern GBR", "Central GBR", "Southern GBR")))

reg_p2_dec <- df_plot2 %>%
  filter(Region != "Other") %>%
  group_by(Region, Decade, Disturbance) %>%
  summarise(Total_Loss = sum(raw_loss), .groups = "drop_last") %>%
  mutate(
    Decade_Total = sum(Total_Loss),
    Prop = Total_Loss / Decade_Total,
    Label = ifelse(Prop >= 0.04, sprintf("%.1f%%", Prop * 100), ""),
    Region = factor(Region, levels = c("Northern GBR", "Central GBR", "Southern GBR"))
  )

# --- Decadal Subplots ---
# Panel A: Overall Net Change Bar Chart
p_top_net_dec <- ggplot(overall_dec_net, aes(x = Decade, y = Mean_Change, fill = Mean_Change)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.6) +
  geom_bar(stat = "identity", width = 0.45, color = "black", linewidth = 0.4) +
  geom_errorbar(aes(ymin = Mean_Change - SE_Change, ymax = Mean_Change + SE_Change), width = 0.15, linewidth = 0.7) +
  geom_text(aes(y = ifelse(Mean_Change >= 0, Mean_Change + SE_Change + 0.35, Mean_Change - SE_Change - 0.45),
                label = sprintf("%+.1f%%", Mean_Change)), size = 3.2, fontface = "bold") +
  facet_wrap(~Region) +
  scale_fill_gradient2(low = "#B2182B", mid = "#E0E0E0", high = "#2166AC", midpoint = 0, guide = "none") +
  scale_y_continuous(limits = c(-2, 2.5), breaks = seq(-2, 2, 1), expand = c(0, 0)) +
  labs(
    title = "Coral Trajectory & Disturbance Attribution: Overall GBR & Regional Breakdown",
    subtitle = "A: Overall GBR Mean Net Coral Cover Change per Surveyed Reef (% cover ± SE)",
    x = NULL,
    y = "Net Change (%)"
  ) +
  theme_academic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

# Panel B: Overall Gross Loss Bar Chart
p_top_loss_dec <- ggplot(overall_dec_loss, aes(x = Decade, y = Mean_Loss)) +
  geom_bar(stat = "identity", fill = "#4B5563", width = 0.45, alpha = 0.9, color = "black", linewidth = 0.4) +
  geom_errorbar(aes(ymin = Mean_Loss - SE_Loss, ymax = Mean_Loss + SE_Loss), width = 0.15, linewidth = 0.7, color = "black") +
  geom_text(aes(label = sprintf("%.1f%%", Mean_Loss)), vjust = -1.2, size = 3.2, fontface = "bold") +
  facet_wrap(~Region) +
  scale_y_continuous(limits = c(0, 5.2), breaks = seq(0, 5, 1), expand = c(0, 0)) +
  labs(
    subtitle = "B: Overall GBR Mean Hard Coral Cover Loss per Surveyed Reef (% cover ± SE)",
    x = NULL,
    y = "Mean Loss (%)"
  ) +
  theme_academic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

# Panel C: Overall Stacked Attribution
p_top_attr_dec <- ggplot(overall_p2_dec, aes(x = Decade, y = Prop, fill = factor(Disturbance, levels = names(dist_colors_p2)))) +
  geom_bar(stat = "identity", width = 0.5, color = "black", linewidth = 0.3) +
  geom_text(aes(label = Label), position = position_stack(vjust = 0.5), color = "white", fontface = "bold", size = 3.1) +
  facet_wrap(~Region) +
  scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
  scale_fill_manual(values = dist_colors_p2, name = "Disturbance Type") +
  labs(
    subtitle = "C: Overall GBR 100% Stacked Bar Chart ('m' Partitioned)",
    x = NULL,
    y = "Proportion of Loss (%)"
  ) +
  theme_academic() +
  theme(axis.text.x = element_text(face = "bold"), legend.position = "none")

# Panel D: Regional Net Change Bar Chart
p_reg_net_dec <- ggplot(reg_dec_net, aes(x = Decade, y = Mean_Change, fill = Mean_Change)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.6) +
  geom_bar(stat = "identity", width = 0.55, color = "black", linewidth = 0.4) +
  geom_errorbar(aes(ymin = Mean_Change - SE_Change, ymax = Mean_Change + SE_Change), width = 0.18, linewidth = 0.7) +
  geom_text(aes(y = ifelse(Mean_Change >= 0, Mean_Change + SE_Change + 0.45, Mean_Change - SE_Change - 0.55),
                label = sprintf("%+.1f%%", Mean_Change)), size = 3.0, fontface = "bold") +
  facet_wrap(~Region, nrow = 1) +
  scale_fill_gradient2(low = "#B2182B", mid = "#E0E0E0", high = "#2166AC", midpoint = 0, guide = "none") +
  scale_y_continuous(limits = c(-4.8, 3.8), breaks = seq(-4, 3, 1), expand = c(0, 0)) +
  labs(
    subtitle = "D: Regional Mean Net Coral Cover Change per Surveyed Reef (% cover ± SE)",
    x = NULL,
    y = "Net Change (%)"
  ) +
  theme_academic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

# Panel E: Regional Gross Loss Bar Chart
p_reg_loss_dec <- ggplot(reg_dec_loss, aes(x = Decade, y = Mean_Loss)) +
  geom_bar(stat = "identity", fill = "#4B5563", width = 0.55, alpha = 0.9, color = "black", linewidth = 0.4) +
  geom_errorbar(aes(ymin = Mean_Loss - SE_Loss, ymax = Mean_Loss + SE_Loss), width = 0.18, linewidth = 0.7, color = "black") +
  geom_text(aes(label = sprintf("%.1f%%", Mean_Loss)), vjust = -1.2, size = 3.0, fontface = "bold") +
  facet_wrap(~Region, nrow = 1) +
  scale_y_continuous(limits = c(0, 9.5), breaks = seq(0, 8, 2), expand = c(0, 0)) +
  labs(
    subtitle = "E: Regional Mean Hard Coral Cover Loss per Surveyed Reef (% cover ± SE)",
    x = NULL,
    y = "Mean Loss (%)"
  ) +
  theme_academic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

# Panel F: Regional Stacked Attribution
p_reg_attr_dec <- ggplot(reg_p2_dec, aes(x = Decade, y = Prop, fill = factor(Disturbance, levels = names(dist_colors_p2)))) +
  geom_bar(stat = "identity", width = 0.6, color = "black", linewidth = 0.3) +
  geom_text(aes(label = Label), position = position_stack(vjust = 0.5), color = "white", fontface = "bold", size = 3.0) +
  facet_wrap(~Region, nrow = 1) +
  scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
  scale_fill_manual(values = dist_colors_p2, name = "Disturbance Type") +
  labs(
    subtitle = "F: Regional 100% Stacked Bar Charts ('m' Partitioned)",
    x = "Decade",
    y = "Proportion of Loss (%)"
  ) +
  theme_academic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    legend.position = "bottom"
  )

composite_decadal_6panel <- (p_top_net_dec / p_top_loss_dec / p_top_attr_dec / p_reg_net_dec / p_reg_loss_dec / p_reg_attr_dec) +
  plot_layout(heights = c(1, 1, 1.8, 1, 1, 2.2), guides = "collect") &
  theme(legend.position = "bottom")

print(composite_decadal_6panel)
# 2. ANNUAL STACKED COMPOSITE (6-PANEL HIERARCHY WITH NET CHANGE GRADIENT)
# ==============================================================================

# --- Overall GBR Annual Aggregations ---
overall_ann_net <- df_surveyed %>%
  group_by(report_year) %>%
  summarise(
    N = n(),
    Mean_Change = mean(net_change, na.rm = TRUE),
    SE_Change = ifelse(N > 1, sd(net_change, na.rm = TRUE) / sqrt(N), 0),
    .groups = "drop"
  ) %>%
  mutate(Region = "Overall GBR")

overall_ann_loss <- df_surveyed %>%
  group_by(report_year) %>%
  summarise(
    N = n(),
    Mean_Loss = mean(raw_loss * 100),
    SE_Loss = ifelse(N > 1, sd(raw_loss * 100) / sqrt(N), 0),
    .groups = "drop"
  ) %>%
  mutate(Region = "Overall GBR")

overall_p2_ann <- df_plot2 %>%
  group_by(report_year, Disturbance) %>%
  summarise(Total_Loss = sum(raw_loss), .groups = "drop") %>%
  mutate(Region = "Overall GBR")

# --- Regional Annual Aggregations ---
reg_ann_net <- df_surveyed %>%
  filter(Region != "Other") %>%
  group_by(Region, report_year) %>%
  summarise(
    N = n(),
    Mean_Change = mean(net_change, na.rm = TRUE),
    SE_Change = ifelse(N > 1, sd(net_change, na.rm = TRUE) / sqrt(N), 0),
    .groups = "drop"
  ) %>%
  mutate(Region = factor(Region, levels = c("Northern GBR", "Central GBR", "Southern GBR")))

reg_ann_loss <- df_surveyed %>%
  filter(Region != "Other") %>%
  group_by(Region, report_year) %>%
  summarise(
    N = n(),
    Mean_Loss = mean(raw_loss * 100),
    SE_Loss = ifelse(N > 1, sd(raw_loss * 100) / sqrt(N), 0),
    .groups = "drop"
  ) %>%
  mutate(Region = factor(Region, levels = c("Northern GBR", "Central GBR", "Southern GBR")))

reg_p2_ann <- df_plot2 %>%
  filter(Region != "Other") %>%
  group_by(Region, report_year, Disturbance) %>%
  summarise(Total_Loss = sum(raw_loss), .groups = "drop") %>%
  mutate(Region = factor(Region, levels = c("Northern GBR", "Central GBR", "Southern GBR")))

# --- Annual Subplots ---
p_top_net_ann <- ggplot(overall_ann_net, aes(x = factor(report_year), y = Mean_Change, fill = Mean_Change)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.6) +
  geom_bar(stat = "identity", width = 0.8, color = "black", linewidth = 0.2) +
  geom_errorbar(aes(ymin = Mean_Change - SE_Change, ymax = Mean_Change + SE_Change), width = 0.25, linewidth = 0.5) +
  facet_wrap(~Region) +
  scale_fill_gradient2(low = "#B2182B", mid = "#E0E0E0", high = "#2166AC", midpoint = 0, guide = "none") +
  scale_y_continuous(limits = c(min(overall_ann_net$Mean_Change - overall_ann_net$SE_Change) * 1.15,
                                max(overall_ann_net$Mean_Change + overall_ann_net$SE_Change) * 1.15),
                     expand = c(0, 0)) +
  labs(
    title = "Annual Coral Trajectory and Disturbance Attribution (1987–2026)",
    subtitle = "A: Overall GBR Mean Net Coral Cover Change per Surveyed Reef (% cover ± SE)",
    x = NULL,
    y = "Net Change (%)"
  ) +
  theme_academic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

p_top_loss_ann <- ggplot(overall_ann_loss, aes(x = factor(report_year), y = Mean_Loss, group = 1)) +
  geom_ribbon(aes(ymin = pmax(0, Mean_Loss - SE_Loss), ymax = Mean_Loss + SE_Loss), fill = "#4B5563", alpha = 0.25) +
  geom_line(color = "#111827", linewidth = 0.8) +
  geom_point(color = "#111827", size = 1.6) +
  facet_wrap(~Region) +
  scale_y_continuous(limits = c(0, max(overall_ann_loss$Mean_Loss + overall_ann_loss$SE_Loss) * 1.1), expand = c(0, 0)) +
  labs(
    subtitle = "B: Overall GBR Mean Hard Coral Cover Loss per Surveyed Reef (% cover ± SE)",
    x = NULL,
    y = "Mean Loss (%)"
  ) +
  theme_academic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

p_top_attr_ann <- ggplot(overall_p2_ann, aes(x = factor(report_year), y = Total_Loss, fill = factor(Disturbance, levels = names(dist_colors_p2)))) +
  geom_bar(position = "fill", stat = "identity", width = 0.8, color = "black", linewidth = 0.15) +
  facet_wrap(~Region) +
  scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
  scale_fill_manual(values = dist_colors_p2, name = "Disturbance Type") +
  labs(
    subtitle = "C: Overall GBR Annual 100% Stacked Bar Chart ('m' Partitioned)",
    x = NULL,
    y = "Proportion of Loss (%)"
  ) +
  theme_academic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "none")

p_reg_net_ann <- ggplot(reg_ann_net, aes(x = factor(report_year), y = Mean_Change, fill = Mean_Change)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.6) +
  geom_bar(stat = "identity", width = 0.85, color = "black", linewidth = 0.15) +
  geom_errorbar(aes(ymin = Mean_Change - SE_Change, ymax = Mean_Change + SE_Change), width = 0.25, linewidth = 0.4) +
  facet_wrap(~Region, nrow = 1) +
  scale_fill_gradient2(low = "#B2182B", mid = "#E0E0E0", high = "#2166AC", midpoint = 0, guide = "none") +
  scale_y_continuous(limits = c(min(reg_ann_net$Mean_Change - reg_ann_net$SE_Change) * 1.15,
                                max(reg_ann_net$Mean_Change + reg_ann_net$SE_Change) * 1.15),
                     expand = c(0, 0)) +
  labs(
    subtitle = "D: Regional Mean Net Coral Cover Change per Surveyed Reef (% cover ± SE)",
    x = NULL,
    y = "Net Change (%)"
  ) +
  theme_academic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

p_reg_loss_ann <- ggplot(reg_ann_loss, aes(x = factor(report_year), y = Mean_Loss, group = 1)) +
  geom_ribbon(aes(ymin = pmax(0, Mean_Loss - SE_Loss), ymax = Mean_Loss + SE_Loss), fill = "#4B5563", alpha = 0.25) +
  geom_line(color = "#111827", linewidth = 0.7) +
  geom_point(color = "#111827", size = 1.3) +
  facet_wrap(~Region, nrow = 1) +
  scale_y_continuous(limits = c(0, max(reg_ann_loss$Mean_Loss + reg_ann_loss$SE_Loss) * 1.1), expand = c(0, 0)) +
  labs(
    subtitle = "E: Regional Mean Hard Coral Cover Loss per Surveyed Reef (% cover ± SE)",
    x = NULL,
    y = "Mean Loss (%)"
  ) +
  theme_academic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

p_reg_attr_ann <- ggplot(reg_p2_ann, aes(x = factor(report_year), y = Total_Loss, fill = factor(Disturbance, levels = names(dist_colors_p2)))) +
  geom_bar(position = "fill", stat = "identity", width = 0.85, color = "black", linewidth = 0.15) +
  facet_wrap(~Region, nrow = 1) +
  scale_y_continuous(labels = percent_format(), expand = c(0, 0)) +
  scale_fill_manual(values = dist_colors_p2, name = "Disturbance Type") +
  labs(
    subtitle = "F: Regional Annual 100% Stacked Bar Charts ('m' Partitioned)",
    x = "Survey Year",
    y = "Proportion of Loss (%)"
  ) +
  theme_academic() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7, face = "bold"),
    legend.position = "bottom"
  )

composite_annual_6panel <- (p_top_net_ann / p_top_loss_ann / p_top_attr_ann / p_reg_net_ann / p_reg_loss_ann / p_reg_attr_ann) +
  plot_layout(heights = c(1, 1, 1.8, 1, 1, 2.2), guides = "collect") &
  theme(legend.position = "bottom")

print(composite_annual_6panel)
```

```{r manta-model-data}
#| label: manta-model-data
#| cache: true
# Load spatial regions
gbr <- st_read("Data/GBR_AIMS") |> st_make_valid()
data_bucket <- s3_bucket("s3://gbr-dms-data-public/gbrmpa-complete-gbr-features/data.parquet")
reefs <- open_dataset(data_bucket) %>%
    filter(FEAT_NAME == "Reef", DATASET == "GBR Features") |>
    distinct(UNIQUE_ID, GBR_NAME, LOC_NAME_S, geometry, FEAT_NAME) %>%
    collect() %>%
    st_as_sf(crs = 4283) |>
    st_make_valid()

coast <- open_dataset(data_bucket) %>%
    filter(FEAT_NAME %in% c("Mainland", "Island"), DATASET == "GBR Features") |>
    distinct(UNIQUE_ID, GBR_NAME, LOC_NAME_S, geometry, FEAT_NAME) %>%
    collect() %>%
    st_as_sf(crs = 4283) |>
    st_make_valid()
reefs_centroids <- st_centroid(reefs)
reefs_join <- st_join(reefs_centroids, gbr)

dat.mod.mant <- df.MANT %>%
    filter(Lag < 3, project_code != "MMP") %>%
    left_join(select(Disturbances_mant, ReefName, report_year, DISTURBANCE_TYPE, storm_name, description),
        by = c("ReefName", "report_year")
    ) %>%
    filter(
        is.na(DISTURBANCE_TYPE) | DISTURBANCE_TYPE %in% c("b", "m", "n"),
        report_year %in% c(1999, 2003, 2017, 2018, 2021, 2023, 2024, 2025)
    ) %>%
    mutate(
        Era = case_when(
            DHWYear %in% c(1998, 2002) ~ "1998/2002",
            DHWYear %in% c(2016, 2017) ~ "2016/2017",
            DHWYear %in% c(2020, 2022) ~ "2020/2022",
            DHWYear == 2024 ~ "2024",
            TRUE ~ NA_character_
        )
    ) |>
    left_join(reefs_join %>% st_drop_geometry() %>% select(LOC_NAME_S, Region) %>% distinct(), by = c("ReefName" = "LOC_NAME_S"))
```

```{r manta-plots}
#| label: manta-plots
#| fig-cap: "Manta tow DHW vs relative change by era and region."
ggplot(
    dat.mod.mant,
    aes(x = MaxDHW.mean, y = Rel.Change, fill = Era, colour = Era)
) +
    geom_point(alpha = 0.5) +
    scale_fill_viridis_d(end = 0.8) +
    scale_color_viridis_d(end = 0.8) +
    coord_cartesian(xlim = c(0, 12), ylim = c(-1, 1)) +
    scale_y_continuous(labels = scales::percent) +
    geom_smooth(method = "gam", formula = y ~ s(x, k = 4)) +
    theme_bw() +
    labs(x = "DHW (°C-weeks)", y = "Relative change (growth-adjusted)")
```

# Model Comparison: LM vs GAM

We compare linear model and GAM fits to the DHW–mortality relationship, stratified by era and region. This helps assess whether the relationship is stable across events or whether more recent events show different sensitivity.

```{r lm-vs-gam}
#| label: lm-vs-gam
#| fig-cap: "LM vs GAM fits stratified by bleaching era."
dat.mod.mantTerry <- dat.mod.mant |>
    bind_rows(rename(dat.bleach, "Rel.Change" = "Change", "MaxDHW.mean" = "DHW") |>
        mutate(Era = "Hughes et al 2018", Rel.Change = Rel.Change / 100))

dat_plot <- dat.mod.mantTerry %>%
    filter(is.finite(MaxDHW.mean), is.finite(Rel.Change)) %>%
    mutate(Region = case_when(
        Era == "Hughes et al 2018" ~ "Hughes et al 2018",
        TRUE ~ Region
    )) %>%
    mutate(Era = factor(Era), Region = factor(Region)) %>%
    filter(!is.na(Era), !is.na(Region))

# Prediction function for LM + GAM
fit_predict <- function(dat_sub, grid_sub, k = 4) {
    m_lm <- lm(Rel.Change ~ MaxDHW.mean, data = dat_sub)
    pr_lm <- predict(m_lm, newdata = grid_sub, se.fit = TRUE)
    out_lm <- grid_sub %>%
        mutate(
            model = "LM", fit = pr_lm$fit,
            lwr = pr_lm$fit - 1.96 * pr_lm$se.fit,
            upr = pr_lm$fit + 1.96 * pr_lm$se.fit
        )

    m_gam <- mgcv::gam(Rel.Change ~ s(MaxDHW.mean, k = k), data = dat_sub, method = "REML")
    pr_gam <- predict(m_gam, newdata = grid_sub, se.fit = TRUE)
    out_gam <- grid_sub %>%
        mutate(
            model = paste0("GAM (k=", k, ")"), fit = pr_gam$fit,
            lwr = pr_gam$fit - 1.96 * pr_gam$se.fit,
            upr = pr_gam$fit + 1.96 * pr_gam$se.fit
        )

    bind_rows(out_lm, out_gam)
}

grid <- dat_plot %>%
    distinct(Era) %>%
    tidyr::expand_grid(MaxDHW.mean = seq(0, 20, by = 0.05))

pred.era <- dat_plot %>%
    group_by(Era) %>%
    group_modify(~ {
        grid_sub <- grid %>%
            filter(Era == .y$Era) %>%
            dplyr::select(MaxDHW.mean)
        fit_predict(.x, grid_sub, k = 3)
    }) %>%
    ungroup()

x_max <- max(dat_plot$MaxDHW.mean, na.rm = TRUE)

Fig3.Eras <- ggplot(dat_plot, aes(MaxDHW.mean, Rel.Change, colour = Era, fill = Era)) +
    annotate("rect",
        xmin = x_max, xmax = 20, ymin = -Inf, ymax = Inf,
        fill = "grey80", alpha = 0.5
    ) +
    geom_vline(xintercept = x_max, linetype = "dotted", linewidth = 0.8) +
    geom_point(shape = 17, alpha = 0.5) +
    geom_ribbon(
        data = pred.era,
        aes(x = MaxDHW.mean, y = fit, ymin = lwr, ymax = upr, group = Era),
        alpha = 0.12, colour = NA, show.legend = FALSE
    ) +
    geom_line(data = pred.era, aes(x = MaxDHW.mean, y = fit), linewidth = 1) +
    scale_color_viridis_d(option = "A", begin = 0.2, end = 0.9) +
    scale_fill_viridis_d(option = "A", begin = 0.2, end = 0.9) +
    scale_y_continuous(labels = scales::percent) +
    coord_cartesian(xlim = c(0, 20), ylim = c(-1, 1)) +
    theme_classic() +
    facet_grid(~model) +
    labs(
        x = "DHW (°C-weeks)", y = "Relative change (growth-adjusted)",
        colour = NULL, fill = NULL
    )

Fig3.Eras
ggsave("output/plots/Fig3_Eras.png", Fig3.Eras, width = 8, height = 5, dpi = 300)
```

# Model Choice & Extrapolation Sensitivity

A critical but often overlooked question in coral mortality modelling is: **what happens to our predictions when we extrapolate beyond the observed DHW range?** As mass bleaching events intensify, reefs are increasingly exposed to DHW values that exceed historical observations. The choice of statistical model profoundly affects these out-of-range predictions — some models predict physically impossible outcomes (mortality >100%, or coral cover below zero).

This section compares four model families applied to the DHW–mortality relationship, and contrasts their behaviour with the ReefMod process-based curve (Bozec et al.) to assess which statistical approaches produce physically plausible extrapolations.

## Models Fitted

We fit three model families to the mortality proportion (bounded 0–1):

1. **Bounded GAM** — `gam(Mort ~ s(DHW), family = quasibinomial("logit"))`. Logit-linked, structurally bounded to (0, 1). This is the recommended approach as it allows for a flexible non-linear shape while preventing impossible predictions.
2. **Beta regression** — `betareg(Mort ~ DHW)`. Designed for proportions with heteroscedastic variance.
3. **Quasi-binomial GLM** — `glm(Mort ~ DHW, family = quasibinomial("logit"))`. Linear on logit scale, bounded.

```{r model-extrapolation-fits}
#| label: model-extrapolation-fits
# ── ReefMod mortality function (Bozec et al.) ──
reefmod_M <- function(DHW, d = 7, s = 1, threshold = 3) {
    w <- 1 / (0.420 + 0.272 * d)
    m <- w * s * (exp(0.168 + 0.347 * DHW) - 1) / 100
    m <- pmin(1, pmax(0, m))
    m[DHW < threshold] <- 0
    M <- 1 - (1 - m)^6
    pmin(1, pmax(0, M))
}

# ── Helper: fit all four models and return prediction grid ──
fit_four_models <- function(dat, dhw_col, mort_col, max_dhw = 20) {
    d <- dat %>%
        transmute(DHW = .data[[dhw_col]], Mort = .data[[mort_col]]) %>%
        filter(is.finite(DHW), is.finite(Mort), Mort >= 0, Mort <= 1)

    # Nudge exact 0/1 for beta regression
    n <- nrow(d)
    d <- d %>% mutate(
        Mort_beta = pmin(pmax(Mort, 0), 1),
        Mort_beta = (Mort_beta * (n - 1) + 0.5) / n
    )

    grid <- data.frame(DHW = seq(0, max_dhw, by = 0.1))
    obs_max <- max(d$DHW, na.rm = TRUE)

    # 1. Unbounded GAM
    m1 <- gam(Mort ~ s(DHW, k = 4), data = d, method = "REML")
    pr1 <- predict(m1, newdata = grid, se.fit = TRUE)
    grid$unb_fit <- pr1$fit
    grid$unb_lwr <- pr1$fit - 1.96 * pr1$se.fit
    grid$unb_upr <- pr1$fit + 1.96 * pr1$se.fit

    # 2. Bounded GAM (quasibinomial logit)
    m2 <- gam(Mort ~ s(DHW, k = 4), data = d, family = quasibinomial(link = "logit"), method = "REML")
    pr2 <- predict(m2, newdata = grid, type = "link", se.fit = TRUE)
    grid$bnd_fit <- inv_logit(pr2$fit)
    grid$bnd_lwr <- inv_logit(pr2$fit - 1.96 * pr2$se.fit)
    grid$bnd_upr <- inv_logit(pr2$fit + 1.96 * pr2$se.fit)

    # 3. Beta regression
    m3 <- tryCatch(betareg(Mort_beta ~ DHW, data = d, link = "logit"), error = function(e) NULL)
    if (!is.null(m3)) {
        grid$beta_fit <- predict(m3, newdata = grid, type = "response")
        # Bootstrap CIs via coefficient simulation
        Xg <- model.matrix(~DHW, data = grid)
        b_mean <- coef(m3)[c("(Intercept)", "DHW")]
        V_mean <- vcov(m3)[c("(Intercept)", "DHW"), c("(Intercept)", "DHW")]
        set.seed(42)
        b_sim <- MASS::mvrnorm(2000, mu = b_mean, Sigma = V_mean)
        mu_sim <- inv_logit(b_sim %*% t(Xg))
        grid$beta_lwr <- apply(mu_sim, 2, quantile, 0.025)
        grid$beta_upr <- apply(mu_sim, 2, quantile, 0.975)
    } else {
        grid$beta_fit <- grid$beta_lwr <- grid$beta_upr <- NA
    }

    # 4. Quasi-binomial GLM
    m4 <- glm(Mort ~ DHW, data = d, family = quasibinomial(link = "logit"))
    pr4 <- predict(m4, newdata = grid, type = "link", se.fit = TRUE)
    grid$glm_fit <- inv_logit(pr4$fit)
    grid$glm_lwr <- inv_logit(pr4$fit - 1.96 * pr4$se.fit)
    grid$glm_upr <- inv_logit(pr4$fit + 1.96 * pr4$se.fit)

    # ReefMod reference curves
    sens_s <- c(1.4, 1.5, 1.6, 1.7)
    grid$reefmod_sens <- reefmod_M(grid$DHW, d = 7, s = mean(sens_s))
    grid$reefmod_sens_lwr <- reefmod_M(grid$DHW, d = 7, s = min(sens_s))
    grid$reefmod_sens_upr <- reefmod_M(grid$DHW, d = 7, s = max(sens_s))
    grid$reefmod_tol <- reefmod_M(grid$DHW, d = 7, s = 0.25)

    # Calculate R2 values
    r2_vals <- list(
        unb_gam = summary(m1)$r.sq,
        bnd_gam = summary(m2)$r.sq,
        beta    = if (!is.null(m3)) m3$pseudo.r.squared else NA,
        glm     = 1 - (m4$deviance / m4$null.deviance)
    )

    list(
        data = d, grid = grid, obs_max = obs_max, r2 = r2_vals,
        models = list(unbounded_gam = m1, bounded_gam = m2, beta = m3, glm = m4)
    )
}

# ── Fit to Hughes et al. (2018) data ──
hughes_fits <- fit_four_models(dat.bleach, "DHW", "Mort.bin", max_dhw = 20)

# ── Fit to AIMS LTMP manta tow data ──
# Transform growth-adjusted Rel.Change to mortality proportion
dat_aims_mort <- dat.mod.mant %>%
    ungroup() %>%
    filter(is.finite(MaxDHW.mean), is.finite(Rel.Change)) %>%
    mutate(
        Mort.prop = pmin(pmax(-Rel.Change, 0), 1),
        Mort.prop = ifelse(Mort.prop == 0, 0.001, Mort.prop)
    )

aims_fits <- fit_four_models(dat_aims_mort, "MaxDHW.mean", "Mort.prop", max_dhw = 25)
```

## Extrapolation Comparison

```{r model-extrapolation-plot}
#| label: model-extrapolation-plot
#| fig-cap: "Model choice & extrapolation sensitivity. Left: Hughes et al. (2018) single-event data. Right: AIMS LTMP manta tow data (growth-corrected, multi-event). Grey shading indicates extrapolation beyond the observed DHW range. Black lines show the ReefMod process-based mortality curve (solid = heat-sensitive, dashed = heat-tolerant)."
#| fig-height: 7
#| fig-width: 12
# ── Helper: build a plot from fits ──
build_extrap_plot <- function(fits, title_label, caption_text = NULL, x_limit = 20) {
    g <- fits$grid %>% filter(DHW <= x_limit)
    d <- fits$data
    x_max <- fits$obs_max
    r2 <- fits$r2

    # Reshape to long format for plotting
    pred_long <- bind_rows(
        g %>% transmute(DHW, model = "Bounded GAM (logit)", fit = bnd_fit, lwr = bnd_lwr, upr = bnd_upr),
        g %>% transmute(DHW, model = "Beta regression", fit = beta_fit, lwr = beta_lwr, upr = beta_upr),
        g %>% transmute(DHW, model = "Quasi-binomial GLM", fit = glm_fit, lwr = glm_lwr, upr = glm_upr)
    ) %>%
        mutate(model = factor(model, levels = c("Bounded GAM (logit)", "Beta regression", "Quasi-binomial GLM")))

    # Create R2 annotation text
    r2_text <- paste0(
        "R\u00B2 values:\n",
        "Bounded GAM: ", round(r2$bnd_gam, 3), "\n",
        "Beta Reg: ", round(r2$beta, 3), "\n",
        "GLM: ", round(r2$glm, 3)
    )

    p <- ggplot(d, aes(DHW, Mort)) +
        # Extrapolation shading
        annotate("rect",
            xmin = x_max, xmax = x_limit, ymin = -Inf, ymax = Inf,
            fill = "grey80", alpha = 0.5
        ) +
        geom_vline(xintercept = x_max, linetype = "dotted", linewidth = 0.8) +
        annotate("text",
            x = x_max + (x_limit - x_max) * 0.5, y = 0.05,
            label = "Extrapolation", colour = "grey30", size = 3
        ) +

        # Data points
        geom_point(shape = 17, colour = "grey30", size = 1.2, alpha = 0.3) +

        # Model ribbons and lines
        geom_ribbon(
            data = pred_long,
            aes(x = DHW, ymin = lwr, ymax = upr, fill = model),
            alpha = 0.12, inherit.aes = FALSE
        ) +
        geom_line(
            data = pred_long,
            aes(x = DHW, y = fit, colour = model),
            linewidth = 0.9, inherit.aes = FALSE
        ) +

        # ReefMod reference (heat-sensitive)
        geom_ribbon(
            data = g, aes(x = DHW, ymin = reefmod_sens_lwr, ymax = reefmod_sens_upr),
            fill = "black", alpha = 0.06, inherit.aes = FALSE
        ) +
        geom_line(
            data = g, aes(x = DHW, y = reefmod_sens),
            colour = "black", linewidth = 1, inherit.aes = FALSE
        ) +
        # ReefMod (heat-tolerant)
        geom_line(
            data = g, aes(x = DHW, y = reefmod_tol),
            colour = "black", linewidth = 0.8, linetype = "dashed", inherit.aes = FALSE
        ) +

        # R2 Annotation
        annotate("label",
            x = x_limit, y = 0, label = r2_text,
            hjust = 1.05, vjust = -0.1, size = 3, label.size = NA, fill = "white", alpha = 0.7
        ) +
        scale_color_viridis_d(option = "A", begin = 0.15, end = 0.85) +
        scale_fill_viridis_d(option = "A", begin = 0.15, end = 0.85) +
        coord_cartesian(xlim = c(0, x_limit), ylim = c(0, 1.05)) +
        theme_classic(base_size = 11) +
        labs(
            x = "DHW (°C-weeks)", y = "Mortality proportion",
            colour = NULL, fill = NULL, title = title_label,
            caption = caption_text
        )

    return(p)
}

p_hughes <- build_extrap_plot(
    hughes_fits,
    "A) Hughes et al. (2018) — Single event (2016)"
)

p_aims <- build_extrap_plot(
    aims_fits,
    "B) AIMS LTMP Manta Tow \u2014 Multi-event (1998\u20132024)",
    caption_text = "Note: AIMS mortality proportions are growth-corrected.\nReefs that gained cover are set to mortality \u2248 0.",
    x_limit = 25
)

Fig_Extrapolation <- p_hughes + p_aims +
    plot_layout(ncol = 2, guides = "collect") &
    theme(legend.position = "bottom")

Fig_Extrapolation
ggsave("output/plots/Fig_ModelChoice_Extrapolation.png", Fig_Extrapolation,
    width = 12, height = 7, dpi = 300
)
```

## Predicted Mortality at Key DHW Thresholds

```{r model-extrapolation-table}
#| label: model-extrapolation-table
# Extract predictions at key DHW values
extract_at_dhw <- function(fits, dataset_label, dhw_vals = c(8, 10, 12, 15, 20)) {
    g <- fits$grid %>% filter(round(DHW, 1) %in% dhw_vals)

    bind_rows(
        g %>% transmute(Dataset = dataset_label, DHW, Model = "Bounded GAM", Predicted_Mortality = bnd_fit),
        g %>% transmute(Dataset = dataset_label, DHW, Model = "Beta regression", Predicted_Mortality = beta_fit),
        g %>% transmute(Dataset = dataset_label, DHW, Model = "Quasi-binomial GLM", Predicted_Mortality = glm_fit),
        g %>% transmute(Dataset = dataset_label, DHW, Model = "ReefMod (sensitive)", Predicted_Mortality = reefmod_sens)
    )
}

extrap_table <- bind_rows(
    extract_at_dhw(hughes_fits, "Hughes et al. 2018"),
    extract_at_dhw(aims_fits, "AIMS LTMP Manta Tow")
) %>%
    tidyr::pivot_wider(
        names_from = DHW, values_from = Predicted_Mortality,
        names_prefix = "DHW_"
    ) %>%
    mutate(across(starts_with("DHW_"), ~ round(., 3)))

datatable(extrap_table,
    caption = "Predicted mortality proportion at key DHW thresholds.",
    options = list(pageLength = 12, dom = "t")
)
```

# Benthic Transect Analysis

In addition to manta tow data, AIMS LTMP uses benthic photo-transects to record detailed cover estimates, primarily at 5 m depth (with some sites at 2 m and 8 m). This section extracts and processes benthic hard coral cover for mortality attribution.

```{r benthic-data}
#| label: benthic-data
#| cache: true
load("Data/aims_ltmp/aims_ltmp.RData")
df.AIMS.full <- reef_photo_df

df.AIMS.Bent <- df.AIMS.full |>
    filter(
        data_type == "photo-transect", domain_category == "reef",
        project_code %in% c("LTMP", "MMP"), purpose == "GROUP_LEVEL", variable == "HARD CORAL"
    ) |>
    mutate(
        Reef_Name = toupper(domain_name),
        Reef_Name_Clean = gsub(" REEF(S)?| ISLAND| IS", "", Reef_Name)
    ) |>
    select(Reef_Name, Reef_Name_Clean, project_code, report_year, date, depth, lower:median, mean, reef_zone, id, shelf, reefpage_category)

# Create a cleaned version of the reference for matching, ensuring uniqueness
dat.Ref.Clean <- dat.AIMSRef %>%
    mutate(AIMS_REEF_NAME_Clean = gsub(" REEF(S)?| ISLAND| IS", "", toupper(Reef_Name))) %>%
    group_by(AIMS_REEF_NAME_Clean) %>%
    slice(1) %>% # Take the first match to avoid many-to-many explosion
    ungroup()

df.BENT.HC <- df.AIMS.Bent |>
    left_join(select(dat.Ref.Clean, AIMS_REEF_NAME_Clean, ReefID, ReefName, SECT_NAME, SECTOR),
        by = c("Reef_Name_Clean" = "AIMS_REEF_NAME_Clean")
    ) %>%
    mutate(
        ReefName = case_when(
            Reef_Name == "FRANKLAND ISLANDS" & reef_zone == "East" ~ "Normanby-Mabel Reef (17-012)",
            Reef_Name == "FRANKLAND ISLANDS" & reef_zone == "West" ~ "Round-Russell Reef (17-013)",
            TRUE ~ ReefName
        ),
        ReefID = case_when(
            Reef_Name == "FRANKLAND ISLANDS" & reef_zone == "East" ~ "17-012",
            Reef_Name == "FRANKLAND ISLANDS" & reef_zone == "West" ~ "17-013",
            TRUE ~ ReefID
        )
    )

# Add fallback growth parameters for any remaining ungridded inshore reefs
missing_benth_reefs <- df.BENT.HC %>%
    filter(!ReefName %in% dat.gridSum$REEF_NAME) %>%
    distinct(ReefName, ReefID, SECTOR)

if (nrow(missing_benth_reefs) > 0) {
    fallback_rows <- missing_benth_reefs %>%
        inner_join(sector_params, by = "SECTOR") %>%
        mutate(
            REEF_NAME = ReefName,
            REEF_ID = ReefID,
            WQ1 = WQ / 3, WQ2 = WQ / 3, WQ3 = WQ / 3,
            HC90 = 0.8 * HC.asym
        )
    dat.gridSum <- bind_rows(dat.gridSum, fallback_rows)
}

# Find max DHW for each reef year
dat.yrs.bent.hc <- df.BENT.HC |>
    ungroup() |>
    distinct(ReefName, date, report_year) |>
    filter(!is.na(ReefName)) |>
    mutate(
        year_int = floor(as.numeric(date)),
        frac = as.numeric(date) - year_int,
        Date = ymd(paste0(year_int, "-01-01")) +
            round(frac * as.numeric(ymd(paste0(year_int + 1, "-01-01")) - ymd(paste0(year_int, "-01-01")))),
        Prev = report_year - 1,
        Year = year(Date), Month = month(Date),
        DHWYear = ifelse(Month >= 5, Year, Year - 1)
    ) |>
    select(-year_int, -frac) |>
    left_join(dplyr::select(dat.DHW, ReefName, Year, MaxDHW.mean),
        by = c("ReefName", "DHWYear" = "Year")
    )

# Calculate % decline for each reef
df.BENT.HC <- df.BENT.HC %>%
    left_join(dat.yrs.bent.hc, by = join_by(ReefName, date, report_year)) |>
    arrange(ReefName, Reef_Name, depth, reef_zone, reefpage_category, report_year, Date) |>
    group_by(ReefName, Reef_Name, depth, reef_zone, reefpage_category) |>
    mutate(
        Change = mean - lag(mean, n = 1),
        Rel.Change = Change / lag(mean, n = 1),
        Lag = report_year - lag(report_year, n = 1),
        Baseline.Date = lag(Date, n = 1),
        Interval.Days = as.integer(Date - Baseline.Date)
    ) |>
    relocate(c("Change", "Lag", "Rel.Change", "Baseline.Date", "Interval.Days"), .before = median)

# Apply growth correction
df.BENT.HC <- df.BENT.HC |>
    mutate(Change.Cor = Change, Prev.HC = mean - Change.Cor) |>
    filter(
        ReefName %in% dat.gridSum$REEF_NAME, !is.na(Lag),
        Reef_Name != "MILLN REEF"
    ) |>
    rowwise() |>
    mutate(
        idx = which(dat.gridSum$REEF_NAME == ReefName)[1],
        Pred.Grth = doCoralGrowth.Simp(
            CoralCover = Prev.HC * 100,
            B0 = dat.gridSum$B0[idx],
            WQ = dat.gridSum$WQ[idx],
            HC.asym = dat.gridSum$HC.asym[idx]
        )[3] / 100
    ) |>
    mutate(
        Pred.Grth2 = ifelse(Lag == 2,
            Pred.Grth + doCoralGrowth.Simp(
                CoralCover = (Prev.HC + Pred.Grth) * 100,
                B0 = dat.gridSum$B0[idx],
                WQ = dat.gridSum$WQ[idx],
                HC.asym = dat.gridSum$HC.asym[idx]
            )[3] / 100,
            Pred.Grth
        ),
        Abs.Change = Change.Cor - Pred.Grth2,
        Rel.Change = Abs.Change / (Prev.HC + Pred.Grth2 + 0.001) # Small epsilon to avoid division by zero
    )

dat.mod.bent.hc <- df.BENT.HC %>%
    filter(Lag < 3) %>%
    left_join(select(Disturbances_bent, ReefName, report_year, DISTURBANCE_TYPE, storm_name, description),
        by = c("ReefName", "report_year")
    ) %>%
    filter(
        is.na(DISTURBANCE_TYPE) | DISTURBANCE_TYPE %in% c("b", "m", "n"),
        report_year %in% c(1999, 2003, 2017, 2018, 2021, 2023, 2024, 2025)
    ) %>%
    mutate(
        Era = case_when(
            DHWYear %in% c(1998, 2002) ~ "1998/2002",
            DHWYear %in% c(2016, 2017) ~ "2016/2017",
            DHWYear %in% c(2020, 2022) ~ "2020/2022",
            DHWYear == 2024 ~ "2024",
            TRUE ~ NA_character_
        )
    ) |>
    left_join(reefs_join %>% st_drop_geometry() %>% select(LOC_NAME_S, Region) %>% distinct(), by = c("ReefName" = "LOC_NAME_S"))
```

```{r benthic-plots}
#| label: benthic-plots
#| fig-cap: "Benthic transect DHW vs relative change by era and region."
ggplot(dat.mod.bent.hc, aes(x = MaxDHW.mean, y = Rel.Change, fill = Era, colour = Era)) +
    geom_point() +
    scale_fill_viridis_d(end = 0.8) +
    scale_color_viridis_d(end = 0.8) +
    coord_cartesian(xlim = c(0, 12), ylim = c(-1, 1)) +
    scale_y_continuous(labels = scales::percent) +
    geom_smooth(method = "gam", formula = y ~ s(x, k = 3)) +
    facet_wrap(~project_code) +
    theme_bw() +
    labs(
        x = "DHW (°C-weeks)", y = "Relative change (growth-adjusted)",
        title = "Benthic Transect: LTMP (Offshore) vs MMP (Inshore)"
    )
```

# Repeat Exposure & Frequency

Does the DHW–mortality relationship change for reefs that have been bleached previously? We stratify by first vs repeat exposure, and also look at the overall frequency of >8 DHW events since 2015.

## First vs Repeat Exposure

```{r repeat-exposure}
#| label: repeat-exposure
dat.mod.repeat <- dat.mod.mant %>%
    filter(report_year > 2015) %>%
    group_by(Reef_Name) %>%
    arrange(report_year, .by_group = TRUE) %>%
    mutate(
        exposed4 = MaxDHW.mean > 4,
        ExposureNo = cumsum(exposed4),
        ExposureCat = case_when(
            exposed4 & ExposureNo == 1 ~ "First",
            exposed4 & ExposureNo > 1 ~ "Repeat",
            TRUE ~ "NotExposed"
        ),
        StressCat = case_when(
            MaxDHW.mean < 4 ~ "0-4DHW",
            MaxDHW.mean < 8 ~ "4-8DHW",
            TRUE ~ "8-14DHW"
        )
    ) %>%
    ungroup()

dat_exp <- dat.mod.repeat %>%
    filter(ExposureCat %in% c("First", "Repeat")) %>%
    filter(is.finite(Abs.Change), is.finite(MaxDHW.mean))

# Mixed effects model to account for repeated measures on reefs
m1_re <- lmer(Abs.Change ~ ExposureCat + MaxDHW.mean + Region + (1 | Reef_Name), data = dat_exp)
summary(m1_re)
```

```{r repeat-exposure-plots}
#| label: repeat-exposure-plots
#| fig-cap: "Relative coral cover change by first vs repeat DHW exposure >4."
Fig4Repeat <- ggplot(
    dat_exp |> filter(!is.na(ExposureCat)),
    aes(x = MaxDHW.mean, y = Rel.Change, fill = ExposureCat, shape = ExposureCat, colour = ExposureCat)
) +
    geom_point(alpha = 0.6) +
    scale_fill_manual(values = cols_exposure) +
    scale_color_manual(values = cols_exposure) +
    coord_cartesian(xlim = c(4, 12)) +
    geom_smooth(method = "gam", formula = y ~ s(x, k = 3)) +
    scale_y_continuous(labels = scales::percent) +
    labs(
        x = "DHW (°C-weeks)", y = "Relative coral cover change",
        fill = "Exposure", shape = "Exposure", colour = "Exposure"
    ) +
    theme_bw() +
    theme(legend.position = "bottom")

Fig4Repeat
```

## Thermal Stress Frequency

```{r frequency-analysis}
#| label: frequency-analysis
# Total cover change vs frequency of >8 DHW since 2015
reef_keys <- dat.mod.repeat %>%
    distinct(ReefName) %>%
    filter(!is.na(ReefName))

dhw_2015 <- dat.DHW %>%
    filter(ReefName %in% reef_keys$ReefName, Year >= 2015) %>%
    arrange(ReefName, Year) %>%
    group_by(ReefName) %>%
    mutate(ev8 = MaxDHW.mean > 8) %>%
    ungroup()

exposure_summary <- dhw_2015 %>%
    group_by(ReefName) %>%
    summarise(
        dhw_years = n_distinct(Year),
        n_ev8 = sum(ev8, na.rm = TRUE),
        freq_ev8 = n_ev8 / dhw_years,
        .groups = "drop"
    )

reef_summary <- dat.mod.repeat %>%
    filter(report_year > 2015) %>%
    filter(!is.na(ReefName), is.finite(mean)) %>%
    group_by(ReefName) %>%
    summarise(
        Region = first(Region),
        coral_years = n_distinct(report_year),
        year_start = min(report_year, na.rm = TRUE),
        year_end = max(report_year, na.rm = TRUE),
        cover_start = mean[which.min(report_year)],
        cover_end = mean[which.max(report_year)],
        rel_total_change = (cover_end - cover_start) / cover_start,
        total_cover_change = cover_end - cover_start,
        .groups = "drop"
    ) %>%
    filter(coral_years >= 3, is.finite(rel_total_change), cover_start > 0, year_end > 2023)

reef_disturbance_df <- read_csv("Data/aims_ltmp/reef_disturbance.csv")
dist_summary <- reef_disturbance_df %>%
    filter(year > 2015, sample_type == "MANTA", disturbance == "c") %>%
    group_by(aims_reef_name, disturbance) %>%
    summarise(n_disturbances = n(), .groups = "drop") %>%
    left_join(dat.AIMSRef_key |> dplyr::select(aims_reef_name = AIMS_REEF_NAME_case, ReefName),
        by = "aims_reef_name"
    )

reef_summary2 <- reef_summary %>%
    left_join(exposure_summary, by = "ReefName") %>%
    left_join(dist_summary, by = "ReefName") %>%
    mutate(
        COTSOutbreak = case_when(
            n_disturbances > 0 ~ "COTS Outbreak",
            is.na(n_disturbances) ~ "No Outbreak"
        ),
        BleachingFreq = case_when(
            freq_ev8 == 0 ~ "0",
            freq_ev8 < 0.2 ~ "0-0.2/year",
            TRUE ~ "0.2/year +"
        )
    )

Fig4Freq <- ggplot(reef_summary2, aes(BleachingFreq, total_cover_change, fill = COTSOutbreak)) +
    geom_boxplot() +
    scale_fill_manual(values = cols_outbreak) +
    scale_y_continuous(labels = scales::percent) +
    theme_bw() +
    labs(
        x = "Frequency of >8 DHW years since 2015",
        y = "Total coral cover change since 2015", fill = NULL
    ) +
    theme(legend.position = "bottom")

Fig4Freq
```

## Reef Condition Index (RCI) Response

```{r rci-data}
#| label: rci-data
# ── Load LTMP/CSMP metrics and compute RCI ──
dat_metrics <- read_csv("data/aims_ltmp/metricsLTMPCSMP.csv", show_col_types = FALSE) %>%
    rename(
        Reef = Reef,
        Year = Year,
        Site = Site,
        Coral_Cover_raw = coral,
        Structural_Complexity = Complexity,
        Coral_Diversity_Raw = simpD,
        Fish_Biomass_raw = f.biomass,
        Fish_Diversity_raw = FsimpD,
        Algal_Cover_raw = cca # CCA cover as algal metric
    )

# ── Scale each metric to 0–1 and compute RCI ──
dat_rci <- dat_metrics %>%
    filter(
        !is.na(Coral_Cover_raw), !is.na(Structural_Complexity),
        !is.na(Coral_Diversity_Raw), !is.na(Fish_Biomass_raw),
        !is.na(Fish_Diversity_raw), !is.na(Algal_Cover_raw)
    ) %>%
    mutate(
        # Convert coral cover proportion (if % divide by 100)
        Coral_Cover_prop = Coral_Cover_raw / 100,
        # Log transform fish biomass
        Fish_Biomass_log = log(Fish_Biomass_raw + 1),
        # Scale each metric to 0–1 using min-max across the dataset
        Coral_sc = (Coral_Cover_prop - min(Coral_Cover_prop, na.rm = TRUE)) /
            (max(Coral_Cover_prop, na.rm = TRUE) - min(Coral_Cover_prop, na.rm = TRUE)),
        Complexity_sc = (Structural_Complexity - min(Structural_Complexity, na.rm = TRUE)) /
            (max(Structural_Complexity, na.rm = TRUE) - min(Structural_Complexity, na.rm = TRUE)),
        CoralDiv_sc = (Coral_Diversity_Raw - min(Coral_Diversity_Raw, na.rm = TRUE)) /
            (max(Coral_Diversity_Raw, na.rm = TRUE) - min(Coral_Diversity_Raw, na.rm = TRUE)),
        FishBiomass_sc = (Fish_Biomass_log - min(Fish_Biomass_log, na.rm = TRUE)) /
            (max(Fish_Biomass_log, na.rm = TRUE) - min(Fish_Biomass_log, na.rm = TRUE)),
        FishDiv_sc = (Fish_Diversity_raw - min(Fish_Diversity_raw, na.rm = TRUE)) /
            (max(Fish_Diversity_raw, na.rm = TRUE) - min(Fish_Diversity_raw, na.rm = TRUE)),
        AlgalCover_sc = (Algal_Cover_raw - min(Algal_Cover_raw, na.rm = TRUE)) /
            (max(Algal_Cover_raw, na.rm = TRUE) - min(Algal_Cover_raw, na.rm = TRUE)),
        # RCI = mean of 6 scaled metrics
        RCI = (Coral_sc + Complexity_sc + CoralDiv_sc + FishBiomass_sc + FishDiv_sc + AlgalCover_sc) / 6
    )

# ── Average RCI per reef-year (across sites/transects) ──
rci_reef_year <- dat_rci %>%
    group_by(Reef, Year) %>%
    summarise(RCI = mean(RCI, na.rm = TRUE), .groups = "drop")

# ── Join to DHW via AIMS reference key ──
# New CSV Reef column is uppercase, matching AIMS_REEF_NAME_cap
rci_dhw <- rci_reef_year %>%
    left_join(
        dat.AIMSRef_key %>% select(AIMS_REEF_NAME_cap, ReefName),
        by = c("Reef" = "AIMS_REEF_NAME_cap")
    ) %>%
    filter(!is.na(ReefName)) %>%
    left_join(
        dat.DHW %>% select(ReefName, Year, MaxDHW.mean),
        by = c("ReefName", "Year")
    )

# ── Calculate year-over-year RCI change per reef ──
rci_change <- rci_dhw %>%
    arrange(ReefName, Year) %>%
    group_by(ReefName) %>%
    mutate(
        RCI_lag = lag(RCI),
        RCI_change = RCI - RCI_lag,
        Year_lag = lag(Year)
    ) %>%
    filter(!is.na(RCI_change)) %>%
    ungroup()

# ── First vs Repeat exposure classification (mirroring dat.mod.repeat) ──
rci_repeat <- rci_change %>%
    filter(Year > 2015) %>%
    group_by(ReefName) %>%
    arrange(Year, .by_group = TRUE) %>%
    mutate(
        exposed4 = MaxDHW.mean > 4,
        ExposureNo = cumsum(exposed4),
        ExposureCat = case_when(
            exposed4 & ExposureNo == 1 ~ "First",
            exposed4 & ExposureNo > 1 ~ "Repeat",
            TRUE ~ "NotExposed"
        )
    ) %>%
    ungroup()

rci_exp <- rci_repeat %>%
    filter(ExposureCat %in% c("First", "Repeat")) %>%
    filter(is.finite(RCI_change), is.finite(MaxDHW.mean))
```

```{r rci-repeat-plot}
#| label: rci-repeat-plot
#| fig-cap: "RCI change by first vs repeat DHW exposure >4 (GAM smooth)."
Fig4RCI_Repeat <- ggplot(
    rci_exp,
    aes(
        x = MaxDHW.mean, y = RCI_change, fill = ExposureCat,
        shape = ExposureCat, colour = ExposureCat
    )
) +
    geom_point(alpha = 0.6) +
    scale_fill_manual(values = cols_exposure) +
    scale_color_manual(values = cols_exposure) +
    coord_cartesian(xlim = c(4, 12)) +
    geom_smooth(method = "gam", formula = y ~ s(x, k = 3)) +
    labs(
        x = "DHW (°C-weeks)", y = "RCI change",
        fill = "Exposure", shape = "Exposure", colour = "Exposure"
    ) +
    theme_bw() +
    theme(legend.position = "bottom")

Fig4RCI_Repeat
```

```{r rci-frequency}
#| label: rci-frequency
#| fig-cap: "RCI change by frequency of >8 DHW years since 2015."
# ── Total RCI change since 2015 per reef (mirroring reef_summary) ──
rci_summary <- rci_dhw %>%
    filter(Year > 2015) %>%
    filter(!is.na(ReefName), is.finite(RCI)) %>%
    group_by(ReefName) %>%
    summarise(
        rci_years = n_distinct(Year),
        year_start = min(Year, na.rm = TRUE),
        year_end = max(Year, na.rm = TRUE),
        rci_start = RCI[which.min(Year)],
        rci_end = RCI[which.max(Year)],
        total_rci_change = rci_end - rci_start,
        .groups = "drop"
    ) %>%
    filter(rci_years >= 3, is.finite(total_rci_change), year_end > 2023)

# ── Join with exposure frequency and COTS disturbance ──
rci_summary2 <- rci_summary %>%
    left_join(exposure_summary, by = "ReefName") %>%
    left_join(dist_summary, by = "ReefName") %>%
    mutate(
        COTSOutbreak = case_when(
            n_disturbances > 0 ~ "COTS Outbreak",
            is.na(n_disturbances) ~ "No Outbreak"
        ),
        BleachingFreq = case_when(
            freq_ev8 == 0 ~ "0",
            freq_ev8 < 0.2 ~ "0-0.2/year",
            TRUE ~ "0.2/year +"
        )
    )

Fig4RCI_Freq <- ggplot(rci_summary2, aes(BleachingFreq, total_rci_change, fill = COTSOutbreak)) +
    geom_boxplot() +
    scale_fill_manual(values = cols_outbreak) +
    theme_bw() +
    labs(
        x = "Frequency of >8 DHW years since 2015",
        y = "Total RCI change since 2015", fill = NULL
    ) +
    theme(legend.position = "bottom")

Fig4RCI_Freq
```

# Spatial Mapping

We can map the frequency of >8 DHW events since 2015 and the corresponding net coral cover change per sector.

```{r spatial-mapping}
#| label: spatial-mapping
#| fig-cap: "Spatial distribution of thermal stress frequency ( >8 DHW) across the GBR."
shp <- list.files("data/SectorShapefile", pattern = "\\.shp$", full.names = TRUE)[1]
sectors <- st_read(shp, quiet = TRUE) |> st_make_valid()
sect_id <- intersect(names(sectors), c("SECT_NAME", "SECTOR", "Sector", "NAME", "Name"))[1]

target_crs <- st_crs(sectors)
coast_sp <- st_transform(coast, target_crs)
reefs_sp <- st_transform(reefs, target_crs)
reef_id <- intersect(names(reefs_sp), c("ReefName", "REEF_NAME", "LOC_NAME_S", "domain_name", "name"))[1]

reef_pts <- st_point_on_surface(reefs_sp) |>
    mutate(ReefName = .data[[reef_id]]) |>
    left_join(reef_summary2, by = "ReefName") |>
    filter(is.finite(freq_ev8), is.finite(rel_total_change))

reef_pts2 <- st_join(reef_pts, sectors |> dplyr::select(!!sect_id), left = TRUE)

sector_stats <- reef_pts2 |>
    st_drop_geometry() |>
    group_by(sector = .data[[sect_id]]) |>
    summarise(
        n_reefs = n(),
        freq_mean = mean(freq_ev8, na.rm = TRUE),
        freq_max = max(freq_ev8, na.rm = TRUE),
        chg_mean = mean(total_cover_change, na.rm = TRUE),
        chg_se = sd(total_cover_change, na.rm = TRUE) / sqrt(sum(is.finite(total_cover_change))),
        .groups = "drop"
    )

sectors2 <- sectors |>
    mutate(sector = .data[[sect_id]]) |>
    left_join(sector_stats, by = "sector")

sector_labs <- sectors2 |>
    filter(is.finite(freq_mean)) |>
    mutate(
        lab_pt = st_point_on_surface(geometry),
        label = sprintf(
            "freq>8: %.2f\nΔcover: %+.0f ± %.0f%%",
            freq_mean, 100 * chg_mean, 100 * chg_se
        )
    )

pal <- c("#2A9D8F", "#E9C46A", "#E76F51")

FigMapFreq <- ggplot() +
    geom_sf(data = coast_sp, fill = "grey92", colour = "grey50", linewidth = 0.2) +
    geom_sf(data = sectors2, aes(fill = freq_mean), colour = "grey25", linewidth = 0.25, alpha = 0.95) +
    geom_sf(data = reef_pts2, aes(colour = freq_ev8, size = freq_ev8 * 10), alpha = 0.9) +
    geom_sf_text(data = sector_labs, aes(geometry = lab_pt, label = label), size = 3.0, lineheight = 0.95) +
    scale_fill_gradientn(colours = pal, limits = c(0, 0.5), name = "Freq >8 DHW") +
    scale_colour_gradientn(colours = pal, limits = c(0, 0.5), name = "Freq >8 DHW") +
    scale_size(range = c(0.8, 4.2), limits = c(0, 1), guide = "none") +
    coord_sf(expand = FALSE) +
    theme_classic(base_size = 12) +
    theme(
        axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank(), legend.position = "right"
    )

FigMapFreq
```


# Spatial-Temporal IDW % Acropora Interpolation

To resolve missing benthic composition in Manta Tow surveys and un-sampled reef-year pairs, we apply a spatial-temporal Inverse Distance Weighting (IDW) interpolation pipeline ($w = 1/d^2$, radius $\le 150\text{ km}$, temporal window $\pm 2$ years) across LTMP Benthic photo-transects and MMP Inshore observations.

```{r idw-acropora-load}
#| label: idw-acropora-load
#| cache: true
if (!file.exists('data/processed/acropora_interpolated_df.rds')) {
  source('scripts/idw_acropora_interpolation.R')
}
acrop_interpolated <- readRDS('data/processed/acropora_interpolated_df.rds')
```

```{r idw-acropora-plot}
#| label: idw-acropora-plot
#| fig-cap: 'Distribution of interpolated % Acropora cover across GBR reefs and survey methods.'
#| fig-width: 9
#| fig-height: 5
ggplot(acrop_interpolated, aes(x = prop_acropora_idw * 100, fill = imputation_method)) +
  geom_histogram(bins = 35, color = 'black', alpha = 0.8) +
  scale_fill_brewer(palette = 'Set2', name = 'Imputation Tier') +
  theme_classic() +
  labs(x = 'Interpolated % Acropora Cover', y = 'Reef-Year Observation Count',
       title = 'Spatial-Temporal IDW % Acropora Distribution (1985–2024)')
```

# Harmonised Mortality Outcomes

The canonical response is the proportional loss of total hard-coral cover relative to observed pre-event cover. It is bounded to `[0, 1]`; cover gains are represented as zero mortality, while true zero and complete-loss observations are retained without nudging. Absolute loss in percentage points is the secondary response. The former Gompertz growth-adjusted outcomes and their component cover estimates are retained as sensitivity variables rather than built into the primary estimand.

Manta tow, LTMP benthic, and MMP inshore benthic observations remain in separate tables. Each row is the finest survey observation supplied by AIMS. Eligible observations use the most recent baseline no more than 24 months earlier and retain the established May-to-April event window: May--December surveys map to that calendar year's bleaching event and January--April surveys map to the preceding event year. This allows mortality to be assessed after the required post-bleaching delay without mixing survey programmes.

```{r harmonise-mortality-outcomes}
#| label: harmonise-mortality-outcomes
source('scripts/harmonise_mortality_outcomes.R')

outcome_tables <- build_mortality_outcome_tables(
    dat.mod.mant,
    dat.mod.bent.hc,
    raw_manta = df.AIMS.Mant,
    raw_benthic = df.AIMS.Bent
)
mortality_outcomes_manta <- outcome_tables$manta
mortality_outcomes_ltmp <- outcome_tables$ltmp
mortality_outcomes_mmp <- outcome_tables$mmp
mortality_outcomes_manifest <- outcome_tables$manifest

mortality_outcomes_manifest
```

# Workspace Export for Downstream Modelling

All cleaned exploratory datasets, growth correction functions, and IDW acropora spatial objects are saved to `data/processed/01_exploratory_workspace.RData` for high-performance loading in `02_DHW_Mortality_Modelling.qmd`.

```{r export-workspace}
#| label: export-workspace
save.image('data/processed/01_exploratory_workspace.RData')
cat('Successfully exported 01_exploratory_workspace.RData\n')
```
