---
title: "DHWMortality_Attribution"
author: "Samuel Matthews"
date: "2024-05-29"
bibliography: references.bib
link-citations: true
output: 
  rmdformats::downcute:
    # downcute_theme: "chaos"
    code_folding: hide
    self_contained: true
    lightbox: true
    gallery: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyr)
library(dplyr)
library(ggplot2)
library(DT)
library(arrow)
library(stringr)
library(janitor)
library(lubridate)
```

# Import and Plot Growth Curves

## Asumptions to be fixed

-   Gompertz Growth parameters are supplied at 1km grid cell size. THey have just been averaged for the prototype version.
-   Future version needs to use the SD and total variation for a reef to estimate. COuld retain the grid cells and then combine the reef level estimate
-   Need to simulate the posterior distribution of the gorwth parameters and the starting coral value to propagate uncertainty (HELP! Murray)

```{r Model Time to Recovery, cache=TRUE}
# 1. Get Growth Rates and Cmax from Aaron's paper
# 2. Calculate years to full recovery (undisturbed) from full range of 
dat.grid = read.csv("Data/COTSModParams.csv") 
dat.gridSum = dat.grid %>% 
  mutate(REEF_ID = ifelse(grepl("14-116",REEF_ID, fixed = T),"14-116", REEF_ID),
         REEF_NAME = ifelse(grepl("14-116",REEF_NAME, fixed = T),
                            "Lizard Island Reef (14-116)", REEF_NAME)) %>%
  group_by(REEF_NAME, REEF_ID,SECTOR,
           # CROSS_SHELF
           ) %>%
  summarise(B0 = mean(pred.b0.mean),
            HC.asym = mean(pred.HCmax.mean),
            WQ1 = mean(Primary),
            WQ2 = mean(Secondary),
            WQ3 = mean(Tertiary)) %>%
  mutate(WQ = WQ1+WQ2+WQ3)
doCoralGrowth = function(n, CoralCover, B0, HC.asym,WQ) {
  WQ.mn.sd = c(-0.68, 0.03)
  # b0.wq <- B0 + WQ * rnorm(length(WQ), mean=WQ.mn.sd[1], sd=WQ.mn.sd[2])
  b0.wq <- B0 + WQ * WQ.mn.sd
  b1.wq <- b0.wq / log(HC.asym)
  b1.wq <- b0.wq / log(HC.asym)
  CoralCover <- log(CoralCover)
  b0.wq = as.numeric(b0.wq); b1.wq = as.numeric(b1.wq); CoralCover=as.numeric(CoralCover)
  df = matrix(nrow=n+1, ncol=3)
  # B1 <- B0/log(HC.asym)
  # CoralCover <- log(CoralCover)
  # B0 = as.numeric(B0); B1 = as.numeric(B1); CoralCover=as.numeric(CoralCover)
  # browser()
   df[1,1] = 0
   df[1,2] = exp(CoralCover)
  for (i in 1:n) {
 CoralCover.t1 <- (b0.wq + (1 - b1.wq)*CoralCover)
  # browser()
  
  df[i+1,1] = i
  df[i+1,2] = exp(CoralCover.t1)
  df[i+1,3] = exp(CoralCover.t1)-exp(CoralCover)
  
  CoralCover = CoralCover.t1
  }
  colnames(df) = c("Year", "CoralCover", "CoralGrowth")
  # browser()
  return(df)
}

doCoralGrowth2 = function(n,CoralCover, B0,  HC.asym) {
  # b0.wq <- B0 + WQ * rnorm(length(WQ), mean=WQ.mn.sd[1], sd=WQ.mn.sd[2])
  # b1.wq <- b0.wq / log(HC.asym)
  df = matrix(nrow=n+1, ncol=3)
  B1 <- B0/log(HC.asym)
  CoralCover <- log(CoralCover)
  B0 = as.numeric(B0); B1 = as.numeric(B1); CoralCover=as.numeric(CoralCover)
  # browser()
   df[1,1] = 0
   df[1,2] = exp(CoralCover)
  for (i in 1:n) {
  CoralCover.t1 <- (B0 + (1 - B1)*CoralCover)
  # browser()
  
  df[i+1,1] = i
  df[i+1,2] = exp(CoralCover.t1)
  df[i+1,3] = exp(CoralCover.t1)-exp(CoralCover)
  
  CoralCover = CoralCover.t1
  }
  colnames(df) = c("Year", "CoralCover", "CoralGrowth")
  # browser()
  return(df)
}

CC.tmp = as.data.frame(doCoralGrowth(30,0.1, dat.gridSum[1,4], dat.gridSum[1,5],dat.gridSum[1,9])) %>% mutate(ReefName = NA, Start=NA)

CC.tmp = CC.tmp[-c(1:31),]
CCs = c(0.1, 1:20)
for (i in 1:length(dat.gridSum$REEF_NAME)){
  # for (j in CCs){
CCtmp2 = as.data.frame(doCoralGrowth(30,0.1, dat.gridSum[i,4], dat.gridSum[i,5],dat.gridSum[i,9]))
# CCtmp2$Start = j
CC.tmp = bind_rows(CC.tmp, CCtmp2)
  # }
}
CC.tmp$ReefName = rep(dat.gridSum$REEF_NAME, each=31)
CC.tmp$REEF_NAME = rep(dat.gridSum$REEF_NAME, each=31)
CC.tmp$ReefID = rep(dat.gridSum$REEF_ID, each=31)
CC.tmp$Sector = rep(dat.gridSum$SECTOR, each=31)

dat.gridSum = dat.gridSum %>% mutate(HC90 = 0.8 * HC.asym)
CC.tmp = CC.tmp %>% left_join(select(dat.gridSum, REEF_NAME, HC90))
CC.tmp$isabove = ifelse(CC.tmp$CoralCover > CC.tmp$HC90, 1,0)
CC = CC.tmp %>% group_by(ReefName, isabove) %>% mutate(Max = max(Year)) %>%
  ungroup() %>% group_by(ReefName) %>% 
  mutate(Max = (min(Max))+1, Years2Rec = Max-Year, CC2 = round(CoralCover,0)) %>%
  filter(Years2Rec>0)

library(ggdist)
p1.2  = ggplot(CC.tmp %>% filter(ReefID %in% c("14-116", "17-034", "18-075", "23-082a")), aes(x=Year, y=CoralCover, colour = ReefName)) +
  stat_lineribbon(alpha=0.3) +
  facet_wrap(~ReefName) + theme_classic() + theme(legend.position = "none")
p1.2
p1.3 = ggplot(CC.tmp, aes(x=Year, y=CoralCover)) +
  stat_lineribbon(alpha=0.3) +
  facet_wrap(~Sector) + theme_classic() + theme(legend.position = "none")
p1.3
# FOr all starts
CC.tmp1 = as.data.frame(doCoralGrowth(30,0.1, dat.gridSum[1,4], dat.gridSum[1,5],dat.gridSum[1,9])) %>% mutate(ReefName = NA, Start=NA)
```



```{r, cache=T, eval=FALSE}
CC.tmp1 = CC.tmp1[-c(1:31),]
CCs = c(0.1, 1:20)
for (i in 1:length(dat.gridSum$REEF_NAME)){
  print(i)
for (j in CCs){
  print(j)
CCtmp2 = as.data.frame(doCoralGrowth(20,j, dat.gridSum[i,4], dat.gridSum[i,5],dat.gridSum[i,9]))
CCtmp2$Start = j

CC.tmp1 = bind_rows(CC.tmp1, CCtmp2)
  }
}

CC.tmp1$ReefName = rep(dat.gridSum$REEF_NAME, each=21*length(CCs))
CC.tmp1$REEF_NAME = rep(dat.gridSum$REEF_NAME, each=21*length(CCs))
CC.tmp1$ReefID = rep(dat.gridSum$REEF_ID, each=21*length(CCs))
CC.tmp1$Sector = rep(dat.gridSum$SECTOR, each=21*length(CCs))




CC.tmp1 = CC.tmp1 %>% left_join(select(dat.gridSum, REEF_NAME, HC90))
CC.tmp1$isabove = ifelse(CC.tmp1$CoralCover > CC.tmp1$HC90, 1,0)
CC.tmp1$above = (CC.tmp1$CoralCover - CC.tmp1$HC90)
CC1 = CC.tmp1 %>% group_by(ReefName,Start,isabove) %>% mutate(Max = max(Year)) %>%
  ungroup() %>% group_by(ReefName, Start) %>% 
  mutate(Max = (min(Max))+1, Years2Rec = Max-Year, CC2 = round(CoralCover,0)) %>%
  filter(Years2Rec>0) %>% 
  group_by(ReefName, REEF_NAME, REEF_ID, Sector, Start) %>%
  summarise(Years2Rec = max(Years2Rec), above = max(above)) #%>% filter(Years2Rec != 21) 
CC1Sum = CC1 %>% group_by(Sector, Start) %>%
  summarise(Mn = mean(Years2Rec),se = sd(Years2Rec)/sqrt(n()))
CCJB = CC1 %>% filter(REEF_ID == "18-075") #%>%
  # group_by(Sector,Years2Rec) %>% 
  # summarise(Start = min(Start))
pRec.JB = ggplot(CCJB, aes(x=Start, y=Years2Rec)) + 
  geom_smooth() +
  geom_point() + xlab("Coral Cover") + ylab("years to Recovery") +
  geom_line() + theme_light() +ggtitle("John Brewer Reef")

CCKP = CC1 %>% filter(REEF_ID == "18-079") #%>%
  # group_by(Sector,Years2Rec) %>% 
  # summarise(Start = min(Start))
pRec.KP = ggplot(CCKP, aes(x=Start, y=Years2Rec)) + 
  geom_smooth() +
  geom_point() + xlab("Coral Cover") + ylab("years to Recovery") +
  geom_line() + theme_light() +ggtitle("Keeper Reef")
```

# Import and Plot AIMS modelled times series

This section will load in the AIMS data accessible through their website which models coral cover trajectories

## Assumptions and Issues to be addressed

-   This is using the manta tow hard coral estimates, I believe the growth parameter estimates [@macneil2019] are based off the benthic transect data. This was corrected for in the[@mellin2019] paper but need to see how.
-   Looks like Cooktown / Lizard Reefs had their growth parameters underestimated ***will need to be remodelled***

```{r Import AIMS time series, cache=F}
dat.AIMSRef = read.csv("Data/AIMS-Reef_Reference.csv") %>% rename(Reef_Name = AIMS_REEF_NAME)
dat.AIMS = read.csv("Data/AIMS_ModelledCoralCover_Reef.csv") %>% 
  left_join(dat.AIMSRef) %>% #filter(!domain_name %in% c("CG", "PC", "PO")) %>%
  mutate(ReefID = ifelse(grepl("14-116",ReefID, fixed = T),"14-116", ReefID),
         ReefName = ifelse(grepl("14-116",ReefName, fixed = T),
                            "Lizard Island Reef (14-116)", ReefName)) %>%
  mutate(domain_name = factor(SECT_NAME),
         date = as.Date(date)) %>%
  rename(median_HC = median, lower_HC_CI = lower, 
         upper_HC_CI = upper, mean_HC = mean) %>%
  mutate(median_HC = median_HC*100, lower_HC_CI = lower_HC_CI*100, 
         upper_HC_CI = upper_HC_CI*100, mean_HC = mean_HC*100) %>%
  relocate(ReefName, .after = Reef_Name)
  

datatable(dat.AIMS %>%  
            mutate(across(where(is.numeric), ~ round(., 4))),
          style = "default")
```

```{r Plot Time Series}
ggplot(dat.AIMS %>% filter(ReefID %in% c("14-116", "17-034", "18-075", "23-082a")), aes(x=report_year, y=median_HC, colour = ReefName, fill=ReefName)) +
  geom_line(colour="black")+
  geom_pointrange(aes(ymin=lower_HC_CI, ymax=upper_HC_CI)) +
  geom_ribbon(aes(ymin=lower_HC_CI, ymax=upper_HC_CI), alpha=0.3, colour=NA) +
  facet_wrap(~ReefName, scales = "free_y") + theme_bw() + theme(legend.position = "none")

ggplot(dat.AIMS %>% filter(ReefID %in% c("14-116")), aes(x=report_year, y=median_HC, colour = ReefName, fill=ReefName)) +
  geom_line(colour="black")+
  geom_pointrange(aes(ymin=lower_HC_CI, ymax=upper_HC_CI)) +
  geom_ribbon(aes(ymin=lower_HC_CI, ymax=upper_HC_CI), alpha=0.3, colour=NA) +
  facet_wrap(~ReefName, scales = "free_y") + theme_bw() + theme(legend.position = "none")

```

# Import and plot LTMP recorded disturbance events

```{r}

secorder = c("CG", "PC",  "CL", "CA","IN", "TO","CU", "WH", "PO", "SW","CB")
secorderlabs = c("Cape Grenville", "Princess Charlotte Bay",  "Cooktown / Lizard Island",   "Cairns","Innisfail", "Townsville", "Cape Upstart", "Whitsunday", "Pompey","Swain","Capricorn Bunker")
#Importing data
dat.DIST <- read.csv("Data/Disturbances_manta.csv") 
###


#Plotting ALL DISTURBANCES across ALL REEFS
df.DIST <- dat.DIST %>%
  left_join(dat.AIMSRef, by = c( "FULLREEF_ID")) %>%
  mutate(DISTURBANCE_TYPE=replace(DISTURBANCE_TYPE, DISTURBANCE_TYPE=="m","u")) %>% #said these need to be changed to unknown in the email
  filter(!SECT_NAME %in% c("Outside AIMS field sectors","PO","CG","PB")) %>%
  filter(!DISTURBANCE_TYPE == "n") %>%
  mutate(DISTURBANCE_TYPE = gsub(".*\\b[b]\\b.*", "B", DISTURBANCE_TYPE)) %>%
  mutate(DISTURBANCE_TYPE = gsub(".*\\b[c]\\b.*", "C", DISTURBANCE_TYPE)) %>%
  mutate(DISTURBANCE_TYPE = gsub(".*\\b[d]\\b.*", "D", DISTURBANCE_TYPE)) %>%
  mutate(DISTURBANCE_TYPE = gsub(".*\\b[f]\\b.*", "F", DISTURBANCE_TYPE)) %>%
  mutate(DISTURBANCE_TYPE = gsub(".*\\b[n]\\b.*", "N", DISTURBANCE_TYPE)) %>%
  mutate(DISTURBANCE_TYPE = gsub(".*\\b[s]\\b.*", "S", DISTURBANCE_TYPE)) %>%
  mutate(DISTURBANCE_TYPE = gsub(".*\\b[u]\\b.*", "U", DISTURBANCE_TYPE)) %>%
  mutate(DISTURBANCE_TYPE=as.factor(DISTURBANCE_TYPE)) %>%
  #filter(REPORT_YEAR > 1990) %>%
  group_by(SECT_NAME, REPORT_YEAR, DISTURBANCE_TYPE) %>%
  distinct(AIMS_REEF_NAME,REEF_NAME, REPORT_YEAR, DISTURBANCE_TYPE) %>%
  mutate(SECTOR=factor(SECT_NAME, levels=secorderlabs, labels=secorderlabs))

ggplot(df.DIST %>% filter(AIMS_REEF_NAME == "Lizard Isles"), aes(x=REPORT_YEAR, fill = DISTURBANCE_TYPE)) +
  geom_histogram(binwidth=1) +
  scale_fill_manual(values = c("B"="darkorange", "C" = "maroon4", "D" = "darkolivegreen3",
                               "F"="burlywood4", "S" = "dodgerblue2", "U" = "grey")) +
  theme_bw() + ggtitle("Lizard Island - Recorded Disturbances LTMP Manta Tow") 

df.DIST.liz = df.DIST %>% filter(AIMS_REEF_NAME == "Lizard Isles")

ggplot(dat.AIMS %>% filter(ReefID %in% c("14-116")), 
       aes(x=report_year)) +
  geom_line(aes(y=median_HC),colour="black")+
  geom_linerange(data = df.DIST.liz, 
                 position = position_dodge2(width = 0.8),
               # arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
               aes(x = REPORT_YEAR, ymax = 25, 
                   ymin = 15,color = DISTURBANCE_TYPE), size=0.8) +
  ggrepel::geom_label_repel(data = df.DIST.liz,
                           aes(x = REPORT_YEAR, y = 27 * 0.9,
                               label = DISTURBANCE_TYPE,
                               color = DISTURBANCE_TYPE)) +
  geom_pointrange(aes(y=median_HC, ymin=lower_HC_CI, ymax=upper_HC_CI)) +
  scale_colour_manual(values = c("B"="darkorange", "C" = "maroon4", 
                                 "D" = "darkolivegreen3","F"="burlywood4", 
                                 "S" = "dodgerblue2", "U" = "grey")) +
  geom_ribbon(aes(ymin=lower_HC_CI, ymax=upper_HC_CI), fill="coral",
              alpha=0.3, colour=NA) +
  facet_wrap(~ReefName, scales = "free_y") + 
  theme_bw() + 
  theme(legend.position = "none")

```

## Infilling Lizard Island time series

The data points for 2014,16,18,20 need to be estimates using the gompertz growth model. We will a complete time series for each year, with an additional column that holds the estimate for what coral "should have been"without disturbance

```{r Infilling Lizard}
dat.Liz = dat.AIMS %>% filter(ReefID %in% c("14-116")) %>%
  select(ReefName, Reef_Name, FULLREEF_ID, ReefID, SECTOR, report_year,
             median_HC:mean_HC)
dat.grth.Liz = dat.gridSum[dat.gridSum$REEF_ID %in% c("14-116"),]


# Simplified coral growth function
doCoralGrowth.Simp = function(CoralCover, B0, HC.asym,WQ) {
  # browser()
  WQ.mn.sd = c(-0.68, 0.03)
  # b0.wq <- B0 + WQ * rnorm(length(WQ), mean=WQ.mn.sd[1], sd=WQ.mn.sd[2])
  b0.wq <- B0 + WQ * WQ.mn.sd
  b1.wq <- b0.wq / log(HC.asym)
  b1.wq <- b0.wq / log(HC.asym)
  CoralCover <- log(CoralCover)
  b0.wq = as.numeric(b0.wq); b1.wq = as.numeric(b1.wq); 
  CoralCover=as.numeric(CoralCover)
  # df = matrix(nrow=n+1, ncol=3)
  # B1 <- B0/log(HC.asym)
  # CoralCover <- log(CoralCover)
  # B0 = as.numeric(B0); B1 = as.numeric(B1); CoralCover=as.numeric(CoralCover)
  # browser()
  #  df[1,1] = 0
  #  df[1,2] = exp(CoralCover)
  # for (i in 1:n) {
  CoralCover.t1 = (b0.wq + (1 - b1.wq)*CoralCover)

  CoralCover.t1 = exp(CoralCover.t1)
  Change.t1 = CoralCover.t1-exp(CoralCover)
  return(c(exp(CoralCover), CoralCover.t1, Change.t1))
}


dat.Liz.Infill = dat.AIMS %>% filter(ReefID %in% c("14-116")) %>%
  select(ReefName, Reef_Name, FULLREEF_ID, ReefID, SECTOR) %>% distinct() %>%
  expand_grid(report_year=1986:2024) %>%
  left_join(dat.Liz) %>%
  # mutate(mean_HC.lag = lag(mean_HC, n=1)) %>%
  rowwise() %>%
  mutate(Pred.Grth = doCoralGrowth.Simp(CoralCover=mean_HC,B0=dat.grth.Liz[,"B0"],
                                     WQ =  dat.grth.Liz[,"WQ"], 
                                     HC.asym = dat.grth.Liz[,"HC.asym"])[3]) %>%
  ungroup() %>%
  mutate(mean_HC.Pred = lag(mean_HC, n=1) + lag(Pred.Grth, n=1)) %>%
  rowwise() %>%
  mutate(Pred.Grth = ifelse(is.na(mean_HC),
                            doCoralGrowth.Simp(CoralCover=mean_HC.Pred,
                                     B0=dat.grth.Liz[,"B0"],
                                     WQ =  dat.grth.Liz[,"WQ"], 
                                     HC.asym = dat.grth.Liz[,"HC.asym"])[3],
                            Pred.Grth)) %>%
  ungroup() %>%
  mutate(mean_HC.Pred = ifelse(is.na(mean_HC.Pred), 
                               lag(mean_HC.Pred, n=1) + lag(Pred.Grth, n=1),
                               mean_HC.Pred),
         mean_HC.infill = ifelse(is.na(mean_HC),mean_HC.Pred, mean_HC)) %>%
  select(-Pred.Grth, -mean_HC.Pred) %>%
  rowwise() %>%
  mutate(Pred.Grth = doCoralGrowth.Simp(CoralCover=mean_HC.infill,
                                        B0=dat.grth.Liz[,"B0"],
                                     WQ =  dat.grth.Liz[,"WQ"], 
                                     HC.asym = dat.grth.Liz[,"HC.asym"])[3]) %>%
  ungroup() %>%
  mutate(mean_HC.Pred = lag(mean_HC.infill, n=1) + lag(Pred.Grth, n=1))%>%
  mutate(meanHC.diff =  mean_HC.infill -mean_HC.Pred) %>%
  mutate(meanHC.reldiff = meanHC.diff/mean_HC.Pred)

ggplot(dat.AIMS %>% filter(ReefID %in% c("14-116")),
       aes(x=report_year)) +
  geom_line(aes(y=median_HC),colour="black")+
  geom_line(data=dat.Liz.Infill, aes(y=mean_HC.infill),colour="red")+
   geom_line(data=dat.Liz.Infill, aes(y=mean_HC.Pred),colour="darkred")+
  geom_linerange(data = df.DIST.liz %>% filter(REPORT_YEAR >2010), 
                 position = position_dodge2(width = 0.8),
               # arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
               aes(x = REPORT_YEAR, ymax = 25, 
                   ymin = 15,color = DISTURBANCE_TYPE), size=0.8) +
  ggrepel::geom_label_repel(data = df.DIST.liz %>% filter(REPORT_YEAR >2010),
                           aes(x = REPORT_YEAR, y = 27 * 0.9,
                               label = DISTURBANCE_TYPE,
                               color = DISTURBANCE_TYPE)) +
  geom_pointrange(aes(y=median_HC, ymin=lower_HC_CI, ymax=upper_HC_CI)) +
  scale_colour_manual(values = c("B"="darkorange", "C" = "maroon4", 
                                 "D" = "darkolivegreen3","F"="burlywood4", 
                                 "S" = "dodgerblue2", "U" = "grey")) +
  geom_ribbon(aes(ymin=lower_HC_CI, ymax=upper_HC_CI), fill="coral",
              alpha=0.3, colour=NA) +
  facet_wrap(~ReefName, scales = "free_y") + 
  theme_bw() + 
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(2010,2024))

```

# Existing DHW-Mortality Relationship

This section will focus on the existing dataset used to estimate the DHW - MOratlity relationship. Importantly it will compare different distributions for fitting the realtionship as the binomial model may force too steep of a slope thus predciting close to 100% mortality at high DHW.

## Compare beta vs binomial regression

```{r}
dat.bleach = read.csv("Data/DHW_vs_Coral_Cover_Dataset.csv")
# colnames(dat.bleach) = c("DHW", "Change")

dat.bleach = dat.bleach %>%
  mutate(Mort.bin = ifelse(Change>=0, 0.001, -Change/100))

ggplot(dat.bleach, aes(x=DHW, y=Change)) +
  geom_point() +
  geom_smooth() +
  # geom_smooth(method="glm",method.args=list(family="binomial")) +
  theme_bw()

ggplot(dat.bleach, aes(x=DHW, y=Mort.bin)) +
  geom_point() +
  coord_cartesian(xlim=c(0,16)) +
  # geom_smooth() +
  geom_smooth(method="glm",method.args=list(family="binomial")) +
  theme_bw()

# Compare betareg to binomail
library(betareg)
mod.beta = betareg(Mort.bin ~ DHW, data=dat.bleach)
mod.bin = glm(Mort.bin ~ DHW, data=dat.bleach, family = binomial(link="logit"))

plot(resid(mod.beta) ~ fitted(mod.beta))
lines(lowess(resid(mod.beta) ~ fitted(mod.beta)))

plot(resid(mod.bin) ~ fitted(mod.bin))
lines(lowess(resid(mod.bin) ~ fitted(mod.bin)))
DHARMa::testUniformity(mod.bin)

dat.resid <- sum(resid(mod.beta, type = "pearson")^2)
1 - pchisq(dat.resid, mod.beta$df.resid)

dat.resid <- sum(resid(mod.bin, type = "pearson")^2)
1 - pchisq(dat.resid, mod.bin$df.resid)

car::Anova(mod.beta)
car::Anova(mod.bin)

newdata = with(dat.bleach, 
               data.frame(DHW = seq(0, 16, len = 100)))
fit = predict(mod.bin, newdata = newdata, type = "link", se = TRUE)
q = qt(0.975, df = df.residual(mod.bin))
newdata = cbind(newdata, fit = binomial()$linkinv(fit$fit),
    lower = binomial()$linkinv(fit$fit - q * fit$se.fit), 
    upper = binomial()$linkinv(fit$fit +
        q * fit$se.fit))

newdata2 = with(dat.bleach, 
               data.frame(DHW = seq(0, 16, len = 100)))
fit = predict(mod.beta, newdata = newdata2, type = "link")
q = qt(0.975, df = df.residual(mod.beta))
newdata2 =  cbind(newdata2, fit = binomial()$linkinv(fit))
ggplot(data = newdata, aes(y = fit, x = DHW)) + 
  geom_point(data = dat.bleach,aes(y = Mort.bin)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper),
    fill = "blue", alpha = 0.3) + geom_line() +
  geom_line(data = newdata2, aes(y = fit, x = DHW), colour = "red") +
  theme_classic()


```

# Model existing LTMP mortality dataset

## Asssumptions

-   Here we will take the the LMTP time series and manta tow disturbance datasets.

-   We will initially assume that all mortality in multiple events are from bleaching to look at the extreme case

    -   The point here is that if the % mortality estimates are lower than Hughes et al estimates even without accounting for other drivers of mortality than its definitely worth proceeding

```{r LTMP DHW Realtionship}
Disturbances_manta <- read.csv("Data/Disturbances_manta.csv") %>% 
  left_join(select(dat.AIMSRef, Reef_Name, ReefName), 
            by=c("AIMS_REEF_NAME"="Reef_Name"))  %>%
  rename("report_year"="REPORT_YEAR")
dat.mort = read.csv("Data/DisturbanceMortality.csv") %>%
  filter(!is.na(Lag), !Lag > 2) %>% 
  left_join(select(dat.AIMSRef, Reef_Name, ReefName)) %>%
  select(-DISTURBANCE_TYPE) %>% distinct() |> 
  left_join(select(Disturbances_manta, ReefName, report_year, DISTURBANCE_TYPE),
            by=c("ReefName", "report_year"))
dat.mort.bl = dat.mort %>% 
  filter(!DISTURBANCE_TYPE %in% c("c", "d", "s","f"), 
         #Abs.Change <0,
         report_year %in% c(1999,2003,2017,2018, 2021, 2024))

ggplot(dat.mort.bl, aes(x=report_year)) +
  geom_bar(stat="count")

hist(dat.mort.bl$Abs.Change); hist(dat.mort.bl$Rel.Change)



dat.DHW = read.csv("Data/DHW_1985_2024_GBRReefs.csv") %>%
  rename("ReefName"="LOC_NAME_S") 

dat.agin = dat.DHW |> filter(grepl('Agincourt', ReefName)) |> 
  group_by(Year) |> summarise(MaxDHW.mean = mean(MaxDHW.mean, na.rm = T))

dat.DHW = dat.DHW |> 
  left_join(dat.agin, by="Year") |> 
  mutate(MaxDHW.mean = ifelse(grepl('Agincourt', ReefName)& is.na(MaxDHW.mean.x),
                         MaxDHW.mean.y,MaxDHW.mean.x)) |> 
  select(-MaxDHW.mean.x, -MaxDHW.mean.y)
#need to get the max of both years
dat.yrs = dat.mort.bl %>% select(ReefName, report_year) |> 
  mutate(Prev = report_year-1) |> 
  left_join(dplyr::select(dat.DHW, ReefName, Year,MaxDHW.mean), 
            by=c("ReefName", "report_year"="Year")) |> 
  left_join(dplyr::select(dat.DHW, ReefName, Year,MaxDHW.mean), 
            by=c("ReefName", "Prev"="Year")) |> rowwise() |> 
  mutate(MaxDHW = max(MaxDHW.mean.x, MaxDHW.mean.y)) |> 
  select(-MaxDHW.mean.x, -MaxDHW.mean.y)
  

dat.mod = dat.mort.bl |> left_join(dat.yrs, by=c("ReefName", "report_year")) #|> 
  #filter(MaxDHW > 0.1) 

ggplot(dat.mod , aes(x=MaxDHW, y=Rel.Change)) +
  geom_point() +
  coord_cartesian(xlim=c(0,12), ylim = c(-1,1)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  theme_bw() +facet_grid(~report_year)

ggplot(dat.mod , aes(x=MaxDHW, y=Abs.Change.Cor)) +
  geom_point() +
  coord_cartesian(xlim=c(0,12), ylim = c(-20,20)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  theme_bw() +facet_grid(~report_year)

# dat.mod2 = dat.mort.bl |> 
#   left_join(dat.DHW, by=c("ReefName", "report_year"="Year")) |> 
#   filter(MaxDHW.mean > 0.5) 
#   #filter(Rel.Change <0.50) #|> 
#   #filter(!Reef_Name =="Fitzroy Island", !report_year==2019)
# 
# ggplot(dat.mod2 , aes(x=MaxDHW.mean, y=Rel.Change)) +
#   geom_point() +
#   coord_cartesian(xlim=c(0,12), ylim = c(-1,1)) +
#   # geom_smooth() +
#   geom_smooth(method="gam") +
#   theme_bw() +facet_grid(~report_year)

# need to add lots of other years with minimal DHW and disturbance
```

# Mortality Attribution Plotting

```{r nAive attribution plot}
dat.mort.plot = dat.mort |> 
  filter(Abs.Change <0, !DISTURBANCE_TYPE %in% c('n')) |> 
  mutate(Decade = plyr::round_any(report_year,10, floor),
         AllTime = "AllTime")
ggplot(dat.mort.plot, aes(x=report_year, colour=DISTURBANCE_TYPE,
                     fill=DISTURBANCE_TYPE)) +
  geom_bar(stat="count") +
  theme_bw()
#Counts
ggplot(dat.mort.plot, aes(x=report_year, colour=DISTURBANCE_TYPE,
                     fill=DISTURBANCE_TYPE)) +
  geom_bar(stat="count", position = "fill") +
  theme_bw()
#years coral change
ggplot(dat.mort.plot, aes(x=report_year, y=Abs.Change.Cor, colour=DISTURBANCE_TYPE,
                     fill=DISTURBANCE_TYPE)) +
  geom_bar(stat="identity", position = "fill") +
  theme_bw()

ggplot(dat.mort.plot, aes(x=report_year, y=Abs.Change.Cor, colour=DISTURBANCE_TYPE,
                     fill=DISTURBANCE_TYPE)) +
  geom_bar(stat="identity") +
  theme_bw()
#Decades
ggplot(dat.mort.plot, aes(x=Decade, y=Abs.Change.Cor, colour=DISTURBANCE_TYPE,
                     fill=DISTURBANCE_TYPE)) +
  geom_bar(stat="identity", position = "fill") +
  theme_bw() +
  scale_fill_manual(values = c("b"="darkorange", "c" = "maroon4", 
                                 "d" = "darkolivegreen3","f"="burlywood4", 
                                 "s" = "dodgerblue2", "u" = "grey")) +
  scale_colour_manual(values = c("b"="darkorange", "c" = "maroon4", 
                                 "d" = "darkolivegreen3","f"="burlywood4", 
                                 "s" = "dodgerblue2", "u" = "grey"))


ggplot(dat.mort.plot, aes(x=Decade, y=Abs.Change.Cor, colour=DISTURBANCE_TYPE,
                     fill=DISTURBANCE_TYPE)) +
  geom_bar(stat="identity") +
  theme_bw()
# All Time
ggplot(dat.mort.plot, aes(x=AllTime, y=Abs.Change.Cor, colour=DISTURBANCE_TYPE,
                     fill=DISTURBANCE_TYPE)) +
  geom_bar(stat="identity", position = "fill") +
  theme_bw()

ggplot(dat.mort.plot, aes(x=AllTime, y=Abs.Change.Cor, colour=DISTURBANCE_TYPE,
                     fill=DISTURBANCE_TYPE)) +
  geom_bar(stat="identity") +
  theme_bw()
```

## Update Disturbance table from AIMS API

```{r}

```

```{r Updated Disturbance Table}
library(readr)
# THis is  the benthic data

dat.mort.att = read_csv("Data/240522 dist.lookup_AIMSLTMP.csv") |> 
  # filter(Abs.Change <0, !DISTURBANCE_TYPE %in% c('n')) |> 
  mutate(Decade = plyr::round_any(year,10, floor),
         AllTime = "AllTime")

ggplot(dat.mort.att |> filter(change <0), 
       aes(x=Decade, y=change, colour=DISTURBANCE_TYPE,
                     fill=DISTURBANCE_TYPE)) +
  geom_bar(stat="identity", position = "fill") +
  theme_bw() +
  scale_fill_manual(values = c("b"="darkorange", "c" = "maroon4", 
                                 "d" = "darkolivegreen3","f"="burlywood4", 
                                 "s" = "dodgerblue2", "u" = "grey")) +
  scale_colour_manual(values = c("b"="darkorange", "c" = "maroon4", 
                                 "d" = "darkolivegreen3","f"="burlywood4", 
                                 "s" = "dodgerblue2", "u" = "grey")) #+ 
  facet_wrap(~A_SECTOR)



ggplot(dat.mort.att |> filter(change <0), 
       aes(x=Decade, y=change, colour=DISTURBANCE_TYPE,
                     fill=DISTURBANCE_TYPE)) +
  geom_bar(stat="identity", position = "fill") +
  theme_bw() +
  scale_fill_manual(values = c("b"="darkorange", "c" = "maroon4", 
                                 "d" = "darkolivegreen3","f"="burlywood4", 
                                 "s" = "dodgerblue2", "u" = "grey")) +
  scale_colour_manual(values = c("b"="darkorange", "c" = "maroon4", 
                                 "d" = "darkolivegreen3","f"="burlywood4", 
                                 "s" = "dodgerblue2", "u" = "grey")) + 
  facet_wrap(~A_SECTOR)


# Ensure the dat.mort.att dataframe is filtered for negative 'change' values
dat.mort.att_filtered <- dat.mort.att |> 
  filter(change < 0) |>
  group_by(Decade, DISTURBANCE_TYPE) |>
  summarize(mean_change = mean(change, na.rm = TRUE)) |>
  ungroup()

# Plotting with ggplot2
ggplot(dat.mort.att_filtered, aes(x = Decade, y = mean_change, colour = DISTURBANCE_TYPE, fill = DISTURBANCE_TYPE)) +
  geom_col() +  # Create stacked bar plot with identity statistic
  theme_bw() +  # Use a clean theme
  scale_fill_manual(values = c(
    "b" = "darkorange", 
    "c" = "maroon4", 
    "d" = "darkolivegreen3", 
    "f" = "burlywood4", 
    "s" = "dodgerblue2", 
    "u" = "grey"
  )) +  # Manual fill colors for disturbance types
  scale_colour_manual(values = c(
    "b" = "darkorange", 
    "c" = "maroon4", 
    "d" = "darkolivegreen3", 
    "f" = "burlywood4", 
    "s" = "dodgerblue2", 
    "u" = "grey"
  )) #+  # Manual color outline for disturbance types
  facet_wrap(~A_SECTOR)  # Facet plot by sector
  ggplot(dat.mort.att_filtered, aes(x = Decade, y = mean_change, colour = DISTURBANCE_TYPE, fill = DISTURBANCE_TYPE)) +
  geom_col(position = position_dodge()) +  # Create side-by-side bar plot
  theme_bw() +  # Use a clean theme
  scale_fill_manual(values = c(
    "b" = "darkorange", 
    "c" = "maroon4", 
    "d" = "darkolivegreen3", 
    "f" = "burlywood4", 
    "s" = "dodgerblue2", 
    "u" = "grey"
  )) +  # Manual fill colors for disturbance types
  scale_colour_manual(values = c(
    "b" = "darkorange", 
    "c" = "maroon4", 
    "d" = "darkolivegreen3", 
    "f" = "burlywood4", 
    "s" = "dodgerblue2", 
    "u" = "grey"
  )) #+  # Manual color outline for disturbance types
  facet_wrap(~A_SECTOR)  # Facet plot by sector
  
```

## Apply predicted growth to mortality estimates

Next step is to apply growth curves. This will increase the loss as we will be comparing against what should be expected without coral disturbance.

```{r Expected growth and decline}

dat.grth = dat.mort.bl |> 
  mutate(Prev.HC = mean_HC - Abs.Change.Cor) |> 
  filter(ReefName %in% dat.gridSum$REEF_NAME) |> 
  rowwise() %>%
  mutate(Pred.Grth = doCoralGrowth.Simp(CoralCover=Prev.HC,
                  B0=dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),"B0"],
                  WQ = dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),"WQ"], 
                  HC.asym = dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),
                                                        "HC.asym"])[3]
         ) |> 
  mutate(Abs.Change.Cor = Abs.Change.Cor - Pred.Grth,
         Rel.Change.Cor = Abs.Change.Cor / (Prev.HC + Pred.Grth))

dat.mod = dat.grth |> left_join(dat.yrs, by=c("ReefName", "report_year")) |> 
  #filter(MaxDHW > 0.1) 
  mutate(Era = ifelse(report_year %in% c(1999,2003), "1998/2002",
                      ifelse(report_year %in% c(2017,2018), "2016/2017",
                             ifelse(report_year %in% c(2021,2023), "2020/2022",NA))))
ggplot(dat.mod , aes(x=MaxDHW, y=Rel.Change.Cor)) +
  geom_point() +
  coord_cartesian(xlim=c(0,12), ylim = c(-1,1)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  theme_bw() +facet_grid(~report_year)

ggplot(dat.mod , aes(x=MaxDHW, y=Abs.Change.Cor, 
                     fill = as.factor(report_year),
                     colour=as.factor(report_year))) +
  geom_point() +
  coord_cartesian(xlim=c(0,12), ylim = c(-20,20)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  # facet_grid(~report_year) +
  theme_bw() 
ggplot(dat.mod , aes(x=MaxDHW, y=Abs.Change.Cor, 
                     fill = Era,
                     colour=Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + scale_color_viridis_d(end=0.8) +
  coord_cartesian(xlim=c(0,12), ylim = c(-20,20)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  # facet_grid(~report_year) +
  theme_bw() 

ggplot(dat.mod , aes(x=MaxDHW, y=Rel.Change.Cor, 
                     fill = Era,
                     colour=Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + scale_color_viridis_d(end=0.8) +
  coord_cartesian(xlim=c(0,12)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  scale_y_continuous(labels = scales::percent) +
  # facet_grid(~report_year) +
  theme_bw() 

ggplot(dat.mod , aes(x=MaxDHW, y=Abs.Change.Cor)) +
  geom_point() +
  coord_cartesian(xlim=c(0,12), ylim = c(-20,20)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  # facet_grid(~report_year) +
  theme_bw() 

# ReefName = dat.grth$ReefName[1]
# B0=dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),"B0"]

# add in terry data
dat.mod.Terry = dat.mod |> 
  bind_rows(rename(dat.bleach, "Rel.Change.Cor" = "Change", "MaxDHW" = "DHW") |> 
              mutate(Era = "Hughes et al 2018",
                     Rel.Change.Cor = Rel.Change.Cor/100 )) |> 
  mutate(Region = ifelse(SECTOR_NAME %in% c("CG", "PC", "CL"), "Northern",
                         ifelse(SECTOR_NAME %in% c("CA", "IN", "TO", "WH"),
                                "Central",
                                 ifelse(SECTOR_NAME %in% c("PO", "SW", "CB"),
                                        "Southern", "NA"))))
library(ggokabeito)
ggplot(dat.mod.Terry , aes(x=MaxDHW, y=Rel.Change.Cor, 
                     fill = Era, shape=Era,
                     colour=Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.98,option = "D", begin = 0.1, direction = -1) +
  scale_color_viridis_d(end= 0.98,option = "D", begin = 0.1,direction = -1) +
  # scale_fill_okabe_ito() + scale_color_okabe_ito() +
  coord_cartesian(xlim=c(0,12)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  scale_y_continuous(labels = scales::percent) +
  # facet_wrap(~SECTOR_NAME) +
  theme_bw() 

ggplot(dat.mod.Terry , aes(x=MaxDHW, y=Rel.Change.Cor, 
                     fill = Era, shape=Era,
                     colour=Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.98,option = "D", begin = 0.1, direction = -1) +
  scale_color_viridis_d(end= 0.98,option = "D", begin = 0.1,direction = -1) +
  # scale_fill_okabe_ito() + scale_color_okabe_ito() +
  coord_cartesian(xlim=c(0,12)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~Region) +
  theme_bw() 

```

# DMS API - Taxa Specific mortality curves

Repeat the above analysis but it compares the for each of the main taxa groups. Use DMS to extract Taxa specific as well as disturbance data

-   Need to sort out SNAKE REEF and FRANKLAND ISLANDS to work out which is which

```{r LTMP MMP Benthic, cache=T}
# Connecting to RIMReP collection
#Accessing S3 bucket

#Establishing connection
data_bucket <- s3_bucket("s3://gbr-dms-data-public/aims-ltmp-mmp-coralreef-model/data.parquet")

#Accessing dataset
data_df <- open_dataset(data_bucket)

df.AIMS.full <- data_df |> 
  #We will ignore the geometry column for now
  select(!geometry) |> 
  #Get the first 10 rows
  # slice_head(n = 10) |> 
  collect() 

df.AIMS.full = df.AIMS.full |> mutate(across(where(is.character), as.factor))

df.AIMS <- df.AIMS.full |> 
  mutate(across(where(is.character), as.factor)) |> 
  filter(data_type == "photo-transect", domain_category == "reef",
         purpose == "COMPOSITION", variable == "HARD CORAL") |> 
  select(domain_name,project_code,purpose, date,depth, lower:median, reef_zone:fid)

library(fuzzyjoin)
ci_str_detect <- function(x, y){str_detect(x, regex(y, ignore_case = TRUE))}
# create % change column
# df.BENT = df.AIMS |> rename("Reef_Name" = "domain_name") |>
#   fuzzy_left_join(select(dat.AIMSRef, Reef_Name, ReefID, ReefName, SECT_NAME, SECTOR),match_fun = ci_str_detect)

df.BENT.Exp = df.AIMS |> rename("Reef_Name" = "domain_name") |>
  left_join(select(dat.AIMSRef, AIMS_REEF_NAME_cap, ReefID, ReefName, SECT_NAME, SECTOR), by=c("Reef_Name"="AIMS_REEF_NAME_cap"))

df.BENT = df.AIMS |> rename("Reef_Name" = "domain_name") |>
  left_join(select(dat.AIMSRef, AIMS_REEF_NAME_cap, ReefID, ReefName, SECT_NAME, SECTOR), by=c("Reef_Name"="AIMS_REEF_NAME_cap"))

# df.BENT |> select(Reef_Name.x, Reef_Name.y:SECTOR) |> distinct() |> 
#   write.csv("Data/moreAIMSmatches.csv")

nomatch = df.BENT |> filter(is.na(ReefName)) |> ungroup() |> distinct(Reef_Name)
#Inshore MMP reef
datatable(df.BENT %>% slice_head(n=50) |> 
            mutate(across(where(is.numeric), ~ round(., 4))),
          style = "default")

# find the max DHW for each reef
dat.yrs.bent = df.BENT |>  ungroup() |> distinct(ReefName, date, report_year) |>
  filter(!is.na(ReefName)) |> 
  mutate(Date = as.Date(as.character(date)), Prev = report_year-1,
         Year = year(date), Month = month(date),
         DHWYear = ifelse(Month >=7 , Year, Year-1)) |> 
  # left_join(dplyr::select(dat.DHW, ReefName, Year,MaxDHW.mean), 
  #           by=c("ReefName", "report_year"="Year")) |> 
  # left_join(dplyr::select(dat.DHW, ReefName, Year,MaxDHW.mean), 
  #           by=c("ReefName", "Prev"="Year")) |> rowwise() |> 
  # mutate(MaxDHW = max(MaxDHW.mean.x, MaxDHW.mean.y)) |> 
  # select(-MaxDHW.mean.x, -MaxDHW.mean.y) |> 
  # mutate(MaxDHW = ifelse(is.na(MaxDHW), 0, MaxDHW))
   left_join(dplyr::select(dat.DHW, ReefName, Year,MaxDHW.mean), 
            by=c("ReefName", "DHWYear"="Year")) #|

# calculate % decline for each reef / taxa paairn
df.BENT = df.BENT %>% arrange(ReefName, Reef_Name, reefpage_category, depth, report_year)|>
  group_by(ReefName,Reef_Name, depth, reefpage_category) |> 
  mutate(Change = mean - lag(mean, n=1),
         Rel.Change = Change / lag(mean, n=1),
         Lag = report_year - lag(report_year, n=1)) |> 
  relocate(c("Change", "Lag", "Rel.Change"), .before = median) |> 
  left_join(dat.yrs.bent)


dat.mod.bent = df.BENT |> filter(Lag <3) |> 
  filter(!project_code =="MMP") |> 
  # filter(abs(Change) > 0.005) |> 
  filter(reefpage_category %in% c("Acropora", "Porites", "Pocilloporidae",
                                  "Merulinidae", "Rare groups", "Other")) |> 
  #filter(MaxDHW > 0.1) 
  left_join(select(Disturbances_manta, ReefName, report_year, DISTURBANCE_TYPE),
            by=c("ReefName", "report_year")) |> 
  filter(
    !DISTURBANCE_TYPE %in% c("c", "d", "s","f"), 
    report_year %in% c(1999, 2003, 2017, 2018, 2021, 2023, 2024)
    ) |> 
  mutate(
    Era = case_when(
      report_year %in% c(1999, 2003) ~ "1998/2002",
      report_year %in% c(2017, 2018) ~ "2016/2017",
      report_year %in% c(2021, 2023) ~ "2020/2022",
      report_year == 2024 ~ "2024",
      TRUE ~ NA_character_
    ))



ggplot(dat.mod.bent , aes(x=MaxDHW.mean, y=Change, 
                     fill = Era,
                     colour=Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + scale_color_viridis_d(end=0.8) +
  coord_cartesian(xlim=c(0,12), c(-0.1,0.1)) +
  scale_y_continuous(labels = scales::percent) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  facet_wrap(~reefpage_category, scales = "free_y") +
  theme_bw() 

ggplot(dat.mod.bent , aes(x=MaxDHW.mean, y=Rel.Change, 
                     fill = Era,
                     colour=Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.95) + scale_color_viridis_d(end=0.95) +
  coord_cartesian(xlim=c(0,12), ylim = c(-1,1)) +
  scale_y_continuous(labels = scales::percent) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  facet_wrap(~reefpage_category, scales = "free_y") +
  theme_bw() 

ggplot(dat.mod.bent, aes(x=MaxDHW.mean, y=Rel.Change)) +
  geom_point() +
  coord_cartesian(xlim=c(0,12),ylim = c(-1,1)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  scale_y_continuous(labels = scales::percent) +
    facet_wrap(~reefpage_category, scales = "free_y") +
  theme_bw() 

ggplot(dat.mod.bent, aes(x=MaxDHW.mean, y=Rel.Change)) +
  geom_point() +
  coord_cartesian(xlim=c(0,12),ylim = c(-1,1)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  scale_y_continuous(labels = scales::percent) +
  # facet_wrap(~reefpage_category, scales = "free_y") +
  theme_bw() 
  

```

## MMP Data

```{r MMP}
#### Fix this section
# calculate % decline for each reef / for MMP

df.MMP.Exp = df.AIMS.full |> rename("Reef_Name" = "domain_name") |>
  filter(!Reef_Name %in% c("CA", "TO", "CB", "IN", "WH")) |> 
  left_join(select(dat.AIMSRef, AIMS_REEF_NAME_cap, ReefID,
                   ReefName, SECT_NAME, SECTOR),
            by=c("Reef_Name"="AIMS_REEF_NAME_cap")) |> 
  filter(project_code == "MMP",data_type == "photo-transect",
         purpose == "GROUP_LEVEL",
         variable == "HARD CORAL" ) |> 
  select(ReefName,Reef_Name,project_code,purpose, date,depth, 
         lower:median, reef_zone:fid) |> 
  arrange(ReefName, Reef_Name, reefpage_category, depth, reef_zone, report_year)


df.MMP <- df.AIMS.full |> 
  mutate(across(where(is.character), as.factor)) |> 
  filter(data_type == "photo-transect", project_code == "MMP", 
         purpose == "GROUP_LEVEL", domain_category == "reef", 
         variable == "HARD CORAL") |> 
  select(domain_name,project_code,purpose, date,depth, lower:median, reef_zone:fid) 

df.MMP = df.AIMS.full |> rename("Reef_Name" = "domain_name") |>
  filter(!Reef_Name %in% c("CA", "TO", "CB", "IN", "WH")) |> 
  left_join(select(dat.AIMSRef, AIMS_REEF_NAME_cap, ReefID,
                   ReefName, SECT_NAME, SECTOR),
            by=c("Reef_Name"="AIMS_REEF_NAME_cap")) |> 
  filter(project_code == "MMP",data_type == "photo-transect",
         purpose == "GROUP_LEVEL",
         variable == "HARD CORAL" ) |> 
  select(ReefName,Reef_Name,project_code,purpose, date,depth, 
         lower:median, reef_zone:fid) |> 
  arrange(ReefName, Reef_Name, reefpage_category, depth, reef_zone, report_year)|>
  group_by(ReefName,Reef_Name, depth, reefpage_category, reef_zone) |> 
  mutate(Change = mean - lag(mean, n=1),
         Rel.Change = Change / lag(mean, n=1),
         Lag = report_year - lag(report_year, n=1)) |> 
  relocate(c("Change", "Lag", "Rel.Change"), .before = median) 

nomatch = df.MMP |> filter(is.na(ReefName)) |> ungroup() |> distinct(Reef_Name)
# find the max DHW for each reef
dat.yrs.MMP = df.MMP |>  ungroup() |> distinct(ReefName, date, report_year) |>
  filter(!is.na(ReefName)) |> 
  mutate(Date = as.Date(as.character(date)), Prev = report_year-1,
         Year = year(date), Month = month(date),
         DHWYear = ifelse(Month >=7 , Year, Year-1)) |> 
   left_join(dplyr::select(dat.DHW, ReefName, Year,MaxDHW.mean), 
            by=c("ReefName", "DHWYear"="Year")) #|> 

df.MMP = df.MMP |> left_join(dat.yrs.MMP)

dat.mod.MMP = df.MMP |> filter(Lag <3) |> 
  filter(project_code =="MMP") |> 
  # filter(abs(Change) > 0.005) |> 
  # filter(reefpage_category %in% c("Acropora", "Porites", "Pocilloporidae",
  #                                 "Merulinidae", "Rare groups", "Other")) |> 
  #filter(MaxDHW > 0.1) 
  left_join(select(Disturbances_manta, ReefName, report_year, DISTURBANCE_TYPE),
            by=c("ReefName", "report_year")) |> 
  filter(
    !DISTURBANCE_TYPE %in% c("c", "d", "s","f"), 
    report_year %in% c(1999, 2003, 2017, 2018, 2021, 2023, 2024)
    ) |> 
  mutate(
    Era = case_when(
      DHWYear %in% c(1998, 2002) ~ "1998/2002",
      DHWYear %in% c(2016, 2017) ~ "2016/2017",
      DHWYear %in% c(2020, 2022) ~ "2020/2022",
      DHWYear == 2024 ~ "2024",
      TRUE ~ NA_character_
    ))


ggplot(dat.mod.MMP, aes(x = MaxDHW.mean, y = Change, 
                         fill = Era, colour = Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + 
  scale_color_viridis_d(end = 0.8) +
  coord_cartesian(xlim = c(0, 12), ylim = c(-0.15, 0.15)) +
  scale_y_continuous(labels = scales::percent) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 3)) +  # k = 3 means 2 interior knots (3 knots total)
  theme_bw()

ggplot(dat.mod.MMP, aes(x = MaxDHW.mean, y = Rel.Change, 
                         fill = Era, colour = Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + 
  scale_color_viridis_d(end = 0.8) +
  coord_cartesian(xlim = c(0, 12), ylim = c(-0.5, 0.5)) +
  scale_y_continuous(labels = scales::percent) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 3)) +  # k = 3 means 2 interior knots (3 knots total)
  theme_bw()
```

# DMS API - Manta Tow and repeat exposure

Repeat the above analysis but it compares the for each of the main taxa groups. Use DMS to extract Taxa specific as well as disturbance data

```{r Retrieve data from AIMS API, cache=T}

# Add AIMS region to data
library(sf)
library(lwgeom)

gbr <- st_read("Data/GBR_AIMS") |> st_make_valid()
# source("https://raw.githubusercontent.com/aodn/rimrep-training/main/CoTS_training_Jan2024/R_notebooks/useful_functions.R")

data_bucket <- s3_bucket("s3://gbr-dms-data-public/gbrmpa-complete-gbr-features/data.parquet")
  #Accessing dataset
reefs = open_dataset(data_bucket) %>% 
  filter(FEAT_NAME == "Reef", DATASET == "GBR Features") |> 
  distinct(UNIQUE_ID, GBR_NAME, LOC_NAME_S, geometry, FEAT_NAME) %>%
  collect() %>% 
  #Turning into sf object and assigning reference system: GDA94 (EPSG: 4283)
  st_as_sf(crs = 4283) |> st_make_valid()

coast = open_dataset(data_bucket) %>% 
  filter(DATASET == "GBR Features", FEAT_NAME == "Mainland") |> 
  distinct(UNIQUE_ID, GBR_NAME, LOC_NAME_S, geometry, FEAT_NAME) %>%
  collect() %>% 
  #Turning into sf object and assigning reference system: GDA94 (EPSG: 4283)
  st_as_sf(crs = 4283) |> st_make_valid()

reefs_centroids <- st_centroid(reefs)
reefs_join <- st_join(reefs_centroids, gbr)



duplicates <- reefs_join %>%
  group_by(LOC_NAME_S) %>%
  filter(n() > 1)

df.AIMS.Mant.Exp <- df.AIMS.full |> 
  mutate(across(where(is.character), as.factor)) |> 
  filter(data_type == "manta", domain_category == "reef",
         purpose == "MANTA", variable == "HC") |> 
  select(domain_name,project_code,date,depth, lower:median, reef_zone:fid) |> 
  rename("Reef_Name" = "domain_name") |>
  left_join(select(dat.AIMSRef, AIMS_REEF_NAME_cap, ReefID, ReefName, 
                   SECT_NAME, SECTOR), by=c("Reef_Name"="AIMS_REEF_NAME_cap")) 

write.csv(df.AIMS.Mant.Exp, "Manta.csv")
write.csv(df.MMP.Exp, "MMP.csv")
write.csv(df.BENT.Exp, "Benthic.csv")


df.AIMS.Mant <- df.AIMS.full |> 
  mutate(across(where(is.character), as.factor)) |> 
  filter(data_type == "manta", domain_category == "reef",
         purpose == "MANTA", variable == "HC") |> 
  select(domain_name,project_code,date,depth, lower:median, reef_zone:fid)

library(fuzzyjoiAIMSn)
ci_str_detect <- function(x, y){str_detect(x, regex(y, ignore_case = TRUE))}
# create % change column
# df.BENT = df.AIMS |> rename("Reef_Name" = "domain_name") |>
#   fuzzy_left_join(select(dat.AIMSRef, Reef_Name, ReefID, ReefName, SECT_NAME, SECTOR),match_fun = ci_str_detect)

df.MANT = df.AIMS.Mant |> rename("Reef_Name" = "domain_name") |>
  left_join(select(dat.AIMSRef, AIMS_REEF_NAME_cap, ReefID, ReefName, SECT_NAME, SECTOR), by=c("Reef_Name"="AIMS_REEF_NAME_cap"))

# df.BENT |> select(Reef_Name.x, Reef_Name.y:SECTOR) |> distinct() |> 
#   write.csv("Data/moreAIMSmatches.csv")

nomatch = df.MANT|> filter(is.na(ReefName))
#Inshore MMP reef
datatable(df.MANT %>% slice_head(n=50) |> 
            mutate(across(where(is.numeric), ~ round(., 4))),
          style = "default")

# find the max DHW for each reef
dat.yrs.mant = df.MANT |>  ungroup() |> distinct(ReefName, date, report_year) |>
  filter(!is.na(ReefName)) |> 
  mutate(Date = as.Date(as.character(date)), Prev = report_year-1,
         Year = year(date), Month = month(date),
         DHWYear = ifelse(Month >=7 , Year, Year-1)) |> 
  # left_join(dplyr::select(dat.DHW, ReefName, Year,MaxDHW.mean), 
  #           by=c("ReefName", "report_year"="Year")) |> 
  # left_join(dplyr::select(dat.DHW, ReefName, Year,MaxDHW.mean), 
  #           by=c("ReefName", "Prev"="Year")) |> 
   left_join(dplyr::select(dat.DHW, ReefName, Year,MaxDHW.mean), 
            by=c("ReefName", "DHWYear"="Year")) #|> 
  # rowwise() |> 
  # mutate(MaxDHW = max(MaxDHW.mean.x, MaxDHW.mean.y)) |> 
  # select(-MaxDHW.mean.x, -MaxDHW.mean.y) |> 
  # mutate(MaxDHW = ifelse(is.na(MaxDHW), 0, MaxDHW))

# calculate % decline for each reef / taxa paairn
df.MANT = df.MANT %>% arrange(ReefName, Reef_Name, reefpage_category, depth, report_year)|>
  group_by(ReefName,Reef_Name, depth, reefpage_category) |> 
  mutate(Change = mean - lag(mean, n=1),
         Rel.Change = Change / lag(mean, n=1),
         Lag = report_year - lag(report_year, n=1)) |> 
  relocate(c("Change", "Lag", "Rel.Change"), .before = median) |> 
  left_join(dat.yrs.mant)



df.MANT = df.MANT |> 
  mutate(Change.Cor = Change,
         Prev.HC = mean - Change.Cor) |> 
  filter(ReefName %in% dat.gridSum$REEF_NAME, !is.na(Lag),
         Reef_Name != "MILLN REEF") |> 
  rowwise() |> 
  mutate(Pred.Grth = doCoralGrowth.Simp(CoralCover=Prev.HC*100,
                  B0=dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),"B0"],
                  WQ = dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),"WQ"], 
                  HC.asym = dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),
                                                        "HC.asym"])[3]/100
         ) |> 
  mutate(Pred.Grth2 = ifelse(Lag==2, Pred.Grth+
                  doCoralGrowth.Simp(CoralCover=(Prev.HC+Pred.Grth)*100,
                  B0=dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),"B0"],
                  WQ = dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),"WQ"], 
                  HC.asym = dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),
                                                        "HC.asym"])[3]/100,
                  Pred.Grth)) |> 
  mutate(Abs.Change = Change.Cor - Pred.Grth2,
         Rel.Change = Abs.Change / (Prev.HC + Pred.Grth2))


library(dplyr)

dat.mod.mant <- df.MANT %>%
  filter(Lag < 3, project_code != "MMP", abs(Change)>0.01) %>% 
  left_join(select(Disturbances_manta, ReefName, report_year, DISTURBANCE_TYPE), 
            by = c("ReefName", "report_year")) %>%
  filter(
    !DISTURBANCE_TYPE %in% c("c", "d", "s", "f"),
    report_year %in% c(1999, 2003, 2017, 2018, 2021, 2023, 2024)
  ) %>%
  mutate(
    Era = case_when(
      DHWYear %in% c(1998, 2002) ~ "1998/2002",
      DHWYear %in% c(2016, 2017) ~ "2016/2017",
      DHWYear %in% c(2020, 2022) ~ "2020/2022",
      DHWYear == 2024 ~ "2024",
      TRUE ~ NA_character_
    )
  ) |> 
  left_join(select(reefs_join,LOC_NAME_S,Region), by =c("ReefName"="LOC_NAME_S"))

dat.mod.mantTerry = dat.mod.mant |> 
  bind_rows(rename(dat.bleach, "Rel.Change" = "Change", "MaxDHW.mean" = "DHW") |> 
              mutate(Era = "Hughes et al 2018",
                     Rel.Change = Rel.Change/100 ))

ggplot(dat.mod.mant, aes(x = MaxDHW.mean, y = Abs.Change, 
                         fill = Era, colour = Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + 
  scale_color_viridis_d(end = 0.8) +
  coord_cartesian(xlim = c(0, 12), ylim = c(-0.15, 0.15)) +
  scale_y_continuous(labels = scales::percent) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 3)) +  # k = 3 means 2 interior knots (3 knots total)
  theme_bw() + 
  facet_wrap(~Region)

ggplot(dat.mod.mant, 
       aes(x = MaxDHW.mean, y = Rel.Change, 
                         fill = Era, colour = Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + 
  scale_color_viridis_d(end = 0.8) +
  coord_cartesian(xlim = c(0, 12), 
                  ylim = c(-1, 1)) +
  scale_y_continuous(labels = scales::percent) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4)) +  # k = 3 means 2 interior knots (3 knots total)
  theme_bw()# + 
  # facet_wrap(~Region)

ggplot(dat.mod.mantTerry, 
       aes(x = MaxDHW.mean, y = Rel.Change, 
                         fill = Era, colour = Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + 
  scale_color_viridis_d(end = 0.8) +
  coord_cartesian(xlim = c(0, 12), 
                  ylim = c(-1, 1)) +
  scale_y_continuous(labels = scales::percent) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4)) +  # k = 3 means 2 interior knots (3 knots total)
  theme_bw()# + 

ggplot(dat.mod.mant , aes(x=MaxDHW.mean, y=Rel.Change, 
                     fill = Era,
                     colour=Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + scale_color_viridis_d(end=0.8) +
  coord_cartesian(xlim=c(0,12), ylim = c(-1,1)) +
  scale_y_continuous(labels = scales::percent) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  theme_bw() 

ggplot(dat.mod.mant |> filter(Region=="Southern GBR") , 
       aes(x=MaxDHW.mean, y=Rel.Change, 
                     fill = Era,
                     colour=Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + scale_color_viridis_d(end=0.8) +
  coord_cartesian(xlim=c(0,12), ylim = c(-1,1)) +
  scale_y_continuous(labels = scales::percent) +
  # geom_smooth() +
  geom_smooth(method="gam", formula = y ~ s(x, k = 3)) +
  theme_bw() 

ggplot(dat.mod.mant, aes(x=MaxDHW.mean, y=Rel.Change)) +
  geom_point() +
  coord_cartesian(xlim=c(0,12),ylim = c(-1,1)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  scale_y_continuous(labels = scales::percent) +
  theme_bw() 

  
```

### Updated - AIMS API - Manta Tow and repeat exposure

```{r}
load("Data/aims_ltmp/aims_ltmp.RData")
df.AIMS.Mant <- reef_manta_df |> 
  select(domain_name,project_code,report_year,date,depth, lower:median,mean, reef_zone, id, shelf,reefpage_category) |>
  mutate(Reef_Name = toupper(domain_name)) 

library(fuzzyjoin)
ci_str_detect <- function(x, y){str_detect(x, regex(y, ignore_case = TRUE))}
# create % change column
# df.BENT = df.AIMS |> rename("Reef_Name" = "domain_name") |>
#   fuzzy_left_join(select(dat.AIMSRef, Reef_Name, ReefID, ReefName, SECT_NAME, SECTOR),match_fun = ci_str_detect)
dat.AIMSRef_key <- dat.AIMSRef %>%
  filter(!ReefName %in% c("Round-Russell Reef (17-013)", "Snake Reef (14-087)")) %>%
  mutate(AIMS_REEF_NAME_case = str_to_title(AIMS_REEF_NAME_cap))

dup_ref <- dat.AIMSRef %>%
  count(AIMS_REEF_NAME_cap, sort = TRUE) %>%
  filter(n > 1)



df.MANT = df.AIMS.Mant |> 
  left_join(select(dat.AIMSRef_key, AIMS_REEF_NAME_cap, ReefID, ReefName, SECT_NAME, SECTOR), 
            by=join_by("Reef_Name"=="AIMS_REEF_NAME_cap")) %>%
  mutate(
    year = floor(as.numeric(date)),
    frac = as.numeric(date) - year,
    date_approx = ymd(paste0(year, "-01-01")) +
      round(frac * as.numeric(ymd(paste0(year + 1, "-01-01")) - ymd(paste0(year, "-01-01"))))
  ) %>%
  select(-year, -frac)

dup_years <- df.MANT %>%
  group_by(ReefName) %>%
  count(report_year, sort = TRUE) %>%
  filter(n > 1)

nomatch = df.MANT|> filter(is.na(ReefName))
#Inshore MMP reef
datatable(df.MANT %>% slice_head(n=50) |> 
            mutate(across(where(is.numeric), ~ round(., 4))),
          style = "default")

df.MANT <- df.MANT %>%
  mutate(
    year = floor(as.numeric(date)),
    frac = as.numeric(date) - year,
    date_approx = ymd(paste0(year, "-01-01")) +
      round(frac * as.numeric(ymd(paste0(year + 1, "-01-01")) - ymd(paste0(year, "-01-01"))))
  ) %>%
  select(-year, -frac)

# find the max DHW for each reef
dat.yrs.mant = df.MANT |>  ungroup() |> 
  # distinct(ReefName, date, report_year) |>
  filter(!is.na(ReefName)) |> 
  mutate(Date = date_approx, Prev = report_year-1,
         Year = year(Date), Month = month(Date),
         DHWYear = ifelse(Month >=7 , Year, Year-1)) |> 
  # left_join(dplyr::select(dat.DHW, ReefName, Year,MaxDHW.mean), 
  #           by=c("ReefName", "report_year"="Year")) |> 
  # left_join(dplyr::select(dat.DHW, ReefName, Year,MaxDHW.mean), 
  #           by=c("ReefName", "Prev"="Year")) |> 
   left_join(dplyr::select(dat.DHW, ReefName, Year,MaxDHW.mean), 
            by=c("ReefName", "DHWYear"="Year")) #|> 
  # rowwise() |> 
  # mutate(MaxDHW = max(MaxDHW.mean.x, MaxDHW.mean.y)) |> 
  # select(-MaxDHW.mean.x, -MaxDHW.mean.y) |> 
  # mutate(MaxDHW = ifelse(is.na(MaxDHW), 0, MaxDHW))

# calculate % decline for each reef / taxa paairn
df.MANT = df.MANT %>% arrange(ReefName, Reef_Name, reefpage_category, depth, report_year)|>
  group_by(ReefName,Reef_Name, depth, reefpage_category) |> 
  mutate(Change = mean - lag(mean, n=1),
         Rel.Change = Change / lag(mean, n=1),
         Lag = report_year - lag(report_year, n=1)) |> 
  relocate(c("Change", "Lag", "Rel.Change"), .before = median) |> 
  left_join(dat.yrs.mant)



df.MANT = df.MANT |> 
  mutate(Change.Cor = Change,
         Prev.HC = mean - Change.Cor) |> 
  filter(ReefName %in% dat.gridSum$REEF_NAME, !is.na(Lag),
         Reef_Name != "MILLN REEF") |> 
  rowwise() |> 
  mutate(Pred.Grth = doCoralGrowth.Simp(CoralCover=Prev.HC*100,
                  B0=dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),"B0"],
                  WQ = dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),"WQ"], 
                  HC.asym = dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),
                                                        "HC.asym"])[3]/100
         ) |> 
  mutate(Pred.Grth2 = ifelse(Lag==2, Pred.Grth+
                  doCoralGrowth.Simp(CoralCover=(Prev.HC+Pred.Grth)*100,
                  B0=dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),"B0"],
                  WQ = dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),"WQ"], 
                  HC.asym = dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),
                                                        "HC.asym"])[3]/100,
                  Pred.Grth)) |> 
  mutate(Abs.Change = Change.Cor - Pred.Grth2,
         Rel.Change = Abs.Change / (Prev.HC + Pred.Grth2))


library(dplyr)

dat.mod.mant <- df.MANT %>%
  filter(Lag < 3, project_code != "MMP", abs(Change)>0.01) %>% 
  left_join(select(Disturbances_manta, ReefName, report_year, DISTURBANCE_TYPE), 
            by = c("ReefName", "report_year")) %>%
  filter(
    !DISTURBANCE_TYPE %in% c("c", "d", "s", "f"),
    report_year %in% c(1999, 2003, 2017, 2018, 2021, 2023, 2024,2025)
  ) %>%
  mutate(
    Era = case_when(
      DHWYear %in% c(1998, 2002) ~ "1998/2002",
      DHWYear %in% c(2016, 2017) ~ "2016/2017",
      DHWYear %in% c(2020, 2022) ~ "2020/2022",
      DHWYear == 2024 ~ "2024",
      TRUE ~ NA_character_
    )
  ) |> 
  left_join(select(reefs_join,LOC_NAME_S,Region), by =c("ReefName"="LOC_NAME_S"))

dat.mod.mantTerry = dat.mod.mant |> 
  bind_rows(rename(dat.bleach, "Rel.Change" = "Change", "MaxDHW.mean" = "DHW") |> 
              mutate(Era = "Hughes et al 2018",
                     Rel.Change = Rel.Change/100 ))

ggplot(dat.mod.mant, aes(x = MaxDHW.mean, y = Abs.Change, 
                         fill = Era, colour = Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + 
  scale_color_viridis_d(end = 0.8) +
  coord_cartesian(xlim = c(0, 12), ylim = c(-0.15, 0.15)) +
  scale_y_continuous(labels = scales::percent) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 2)) +  # k = 3 means 2 interior knots (3 knots total)
  # geom_smooth(method = "glm") +
  theme_bw() + 
  facet_wrap(~Region)

ggplot(dat.mod.mant, 
       aes(x = MaxDHW.mean, y = Rel.Change, 
                         fill = Era, colour = Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + 
  scale_color_viridis_d(end = 0.8) +
  coord_cartesian(xlim = c(0, 12), 
                  ylim = c(-1, 1)) +
  scale_y_continuous(labels = scales::percent) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4)) +  # k = 3 means 2 interior knots (3 knots total)
  theme_bw()# + 
  # facet_wrap(~Region)

ggplot(dat.mod.mantTerry, 
       aes(x = MaxDHW.mean, y = Rel.Change, 
                         fill = Era, colour = Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + 
  scale_color_viridis_d(end = 0.8) +
  coord_cartesian(xlim = c(0, 12), 
                  ylim = c(-1, 1)) +
  scale_y_continuous(labels = scales::percent) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4)) +  # k = 3 means 2 interior knots (3 knots total)
  theme_bw()# + 

ggplot(dat.mod.mant , aes(x=MaxDHW.mean, y=Rel.Change, 
                     fill = Era,
                     colour=Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + scale_color_viridis_d(end=0.8) +
  coord_cartesian(xlim=c(0,12), ylim = c(-1,1)) +
  scale_y_continuous(labels = scales::percent) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  theme_bw() 

ggplot(dat.mod.mant , aes(x=MaxDHW.mean, y=Change.Cor, 
                     fill = Era,
                     colour=Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + scale_color_viridis_d(end=0.8) +
  coord_cartesian(xlim=c(0,12), ylim = c(-0.5,0.5)) +
  scale_y_continuous(labels = scales::percent) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  theme_bw() 

ggplot(dat.mod.mant |> filter(Region=="Southern GBR") , 
       aes(x=MaxDHW.mean, y=Rel.Change, 
                     fill = Era,
                     colour=Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + scale_color_viridis_d(end=0.8) +
  coord_cartesian(xlim=c(0,12), ylim = c(-1,1)) +
  scale_y_continuous(labels = scales::percent) +
  # geom_smooth() +
  geom_smooth(method="gam", formula = y ~ s(x, k = 3)) +
  theme_bw() 

ggplot(dat.mod.mant, aes(x=MaxDHW.mean, y=Rel.Change)) +
  geom_point() +
  coord_cartesian(xlim=c(0,12),ylim = c(-1,1)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  scale_y_continuous(labels = scales::percent) +
  theme_bw() 

```

# Compare LM vs GAUSSIAN GAM

```{r}
library(dplyr)
library(tidyr)
library(mgcv)

dat_plot <- dat.mod.mantTerry %>%
  filter(is.finite(MaxDHW.mean), is.finite(Rel.Change)) %>%
  mutate(Region = case_when(
    Era == "Hughes et al 2018" ~ "Hughes et al 2018",
    TRUE ~ Region
  )) %>%
  mutate(Era = factor(Era),
         Region = factor(Region)) %>%
  filter(!is.na(Era), !is.na(Region))

# Prediction grid: DHW 0–15, for each Era x Region
grid <- dat_plot %>%
  distinct(Era, Region) %>%
  tidyr::expand_grid(MaxDHW.mean = seq(0, 20, by = 0.05))

fit_predict <- function(dat_sub, grid_sub, k = 4) {
  # 1) Linear model
  m_lm <- lm(Rel.Change ~ MaxDHW.mean, data = dat_sub)

  pr_lm <- predict(m_lm, newdata = grid_sub, se.fit = TRUE)
  out_lm <- grid_sub %>%
    mutate(model = "LM",
           fit = pr_lm$fit,
           lwr = pr_lm$fit - 1.96 * pr_lm$se.fit,
           upr = pr_lm$fit + 1.96 * pr_lm$se.fit)

  # 2) GAM (Gaussian) on signed change
  m_gam <- mgcv::gam(Rel.Change ~ s(MaxDHW.mean, k = k), data = dat_sub, method = "REML")
  pr_gam <- predict(m_gam, newdata = grid_sub, se.fit = TRUE)
  out_gam <- grid_sub %>%
    mutate(model = paste0("GAM (k=", k, ")"),
           fit = pr_gam$fit,
           lwr = pr_gam$fit - 1.96 * pr_gam$se.fit,
           upr = pr_gam$fit + 1.96 * pr_gam$se.fit)

  bind_rows(out_lm, out_gam)
}

# Fit separately for each Era x Region (so each era has its own curve within region)
pred <- dat_plot %>%
  group_by(Era, Region) %>%
  group_modify(~{
    grid_sub <- grid %>%
      filter(Era == .y$Era, Region == .y$Region) %>%
      dplyr::select(MaxDHW.mean)   # <- drop grouping vars here
    fit_predict(.x, grid_sub, k = 3)
  }) %>%
  ungroup()

pred.era <- dat_plot %>%
  group_by(Era) %>%
  group_modify(~{
    grid_sub <- grid %>%
      filter(Era == .y$Era) %>%
      dplyr::select(MaxDHW.mean)   # <- drop grouping vars here
    fit_predict(.x, grid_sub, k = 3)
  }) %>%
  ungroup()

pred.reg <- dat_plot %>%
  group_by(Region) %>%
  group_modify(~{
    grid_sub <- grid %>%
      filter(Region == .y$Region) %>%
      dplyr::select(MaxDHW.mean)   # <- drop grouping vars here
    fit_predict(.x, grid_sub, k = 3)
  }) %>%
  ungroup()


x_max <- max(dat_plot$MaxDHW.mean, na.rm = TRUE)

Fig3.ErasRegions <- ggplot(dat_plot, aes(MaxDHW.mean, Rel.Change, colour = Era, fill = Era)) +

  # extrapolation shading
  annotate("rect", xmin = x_max, xmax = Inf, ymin = -Inf, ymax = Inf,
           fill = "grey80", alpha = 0.5) +
  geom_vline(xintercept = x_max, linetype = "dotted", linewidth = 0.8) +
  annotate("text", x = x_max + (20 - x_max) * 0.5, y = -0.9,
           label = "Extrapolation", colour = "grey20", size = 4) +

  geom_point(shape = 17, alpha = 0.5) +

  geom_ribbon(
    data = pred,
    aes(x = MaxDHW.mean, y=fit, ymin = lwr, ymax = upr, group = Era),
    alpha = 0.12,
    colour = NA,
    show.legend = FALSE     # keep legend clean (lines/points carry it)
  ) +

  geom_line(
    data = pred,
    aes(x = MaxDHW.mean, y = fit),
    linewidth = 1,
    inherit.aes = TRUE
  ) +
  scale_color_viridis_d(option = "A", begin = 0.2, end = 0.9) +
  scale_fill_viridis_d(option = "A", begin = 0.2, end = 0.9) +
  scale_y_continuous(labels = scales::percent) +
  coord_cartesian(xlim = c(0, 20), ylim = c(-1, 1)) +
  theme_classic() +
  theme(legend.position="bottom") +
  facet_grid(Region~model) +
  labs(x = "DHW (°C-weeks)",
       y = "Relative change (observed - expected growth)",
       colour = NULL, fill = NULL)

Fig3.Eras <- ggplot(dat_plot, aes(MaxDHW.mean, Rel.Change, colour = Era, fill = Era)) +

  # extrapolation shading
  annotate("rect", xmin = x_max, xmax = 20, ymin = -Inf, ymax = Inf,
           fill = "grey80", alpha = 0.5) +
  geom_vline(xintercept = x_max, linetype = "dotted", linewidth = 0.8) +
  annotate("text", x = x_max + (20 - x_max) * 0.5, y = 0.9,
           label = "Extrapolation", colour = "grey20", size = 4) +

  geom_point(shape = 17, alpha = 0.5) +

  geom_ribbon(
    data = pred.era,
    aes(x = MaxDHW.mean, y=fit, ymin = lwr, ymax = upr, group = Era),
    alpha = 0.12,
    colour = NA,
    show.legend = FALSE     # keep legend clean (lines/points carry it)
  ) +

  geom_line(
    data = pred.era,
    aes(x = MaxDHW.mean, y = fit),
    linewidth = 1,
    inherit.aes = TRUE
  ) +
  scale_color_viridis_d(option = "A", begin = 0.2, end = 0.9) +
  scale_fill_viridis_d(option = "A", begin = 0.2, end = 0.9) +
  # scale_color_okabe_ito() +
  # scale_fill_okabe_ito() +
  scale_y_continuous(labels = scales::percent) +
  coord_cartesian(xlim = c(0, 20), ylim = c(-1, 1)) +
  theme_classic() +
  facet_grid(~model) +
  labs(x = "DHW (°C-weeks)",
       y = "Relative change (observed - expected growth)",
       colour = NULL, fill = NULL)
Fig3.Regions <- ggplot(dat_plot, aes(MaxDHW.mean, Rel.Change, colour = Region, fill = Region)) +

  # extrapolation shading
  annotate("rect", xmin = x_max, xmax = 20, ymin = -Inf, ymax = Inf,
           fill = "grey80", alpha = 0.5) +
  geom_vline(xintercept = x_max, linetype = "dotted", linewidth = 0.8) +
  annotate("text", x = x_max + (20 - x_max) * 0.5, y = 0.9,
           label = "Extrapolation", colour = "grey20", size = 4) +

  geom_point(shape = 17, alpha = 0.5) +

  geom_ribbon(
    data = pred.reg,
    aes(x = MaxDHW.mean, y=fit, ymin = lwr, ymax = upr, group = Region),
    alpha = 0.12,
    colour = NA,
    show.legend = FALSE     # keep legend clean (lines/points carry it)
  ) +

  geom_line(
    data = pred.reg,
    aes(x = MaxDHW.mean, y = fit),
    linewidth = 1,
    inherit.aes = TRUE
  ) +
  scale_color_viridis_d(option = "A", begin = 0.2, end = 0.9) +
  scale_fill_viridis_d(option = "A", begin = 0.2, end = 0.9) +
  # scale_color_okabe_ito() +
  # scale_fill_okabe_ito() +
  scale_y_continuous(labels = scales::percent) +
  coord_cartesian(xlim = c(0, 20), ylim = c(-1, 1)) +
  theme_classic() +
  facet_grid(~model) +
  labs(x = "DHW (°C-weeks)",
       y = "Relative change (observed - expected growth)",
       colour = NULL, fill = NULL)

Fig3.Eras
Fig3.Regions
ggsave("output/plots/Fig3_Eras.png", Fig3.Eras,  width = 8, height = 5, dpi = 300)
ggsave("output/plots/Fig3_ErasRegions.png", Fig3.ErasRegions,  width = 5, height = 10, dpi = 300)
ggsave("output/plots/Fig3_Regions.png", Fig3.Regions,  width = 8, height = 5, dpi = 300)
```

## Force 0-1 Models

```{r}
dat_plot2 <- dat.mod.mantTerry %>%
  mutate(
    net = -Rel.Change,                 # decline positive, growth negative
    net_clip = pmax(pmin(net, 1), -1), # keep within [-1, 1] for stability
    decline_prop = pmax(0, net),        # growth clipped to 0
    Mort01 = (net_clip + 1)/2          # growth <0.5, no-change=0.5, decline>0.5
  )

library(dplyr)
library(betareg)
library(MASS)

inv_logit <- function(x) 1 / (1 + exp(-x))

dat_plot2 <- dat_plot2 %>%
  filter(is.finite(MaxDHW.mean), is.finite(Mort01)) %>%
  filter(Mort01 >= 0, Mort01 <= 1) %>%
  filter(net>0)

# Fractional logistic (quasi-binomial)
mod_logit <- glm(decline_prop ~ MaxDHW.mean,
                 data = dat_plot2,
                 family = quasibinomial(link = "logit"))

# Beta regression requires (0,1): nudge Mort01 into (0,1)
n <- nrow(dat_plot2)
dat_plot2 <- dat_plot2 %>%
  mutate(Mort01_beta = (pmin(pmax(Mort01, 0), 1) * (n - 1) + 0.5) / n)

mod_beta <- betareg(decline_prop ~ MaxDHW.mean,
                    data = dat_plot2,
                    link = "logit")

grid <- data.frame(MaxDHW.mean = seq(0, 20, by = 0.05))

# Logistic (quasi-binomial) CI (works natively)
prL <- predict(mod_logit, newdata = grid, type = "link", se.fit = TRUE)
grid$logit_fit <- inv_logit(prL$fit)
grid$logit_lwr <- inv_logit(prL$fit - 1.96 * prL$se.fit)
grid$logit_upr <- inv_logit(prL$fit + 1.96 * prL$se.fit)

# Beta regression mean
grid$beta_fit <- predict(mod_beta, newdata = grid, type = "response")

# Beta regression CI via coefficient simulation
Xg <- model.matrix(~ MaxDHW.mean, data = grid)

b_all <- coef(mod_beta)
V_all <- vcov(mod_beta)

# If phi is included, drop it for mean-curve CI
keep <- c("(Intercept)", "MaxDHW.mean")
b <- b_all[keep]
V <- V_all[keep, keep]

set.seed(1)
B <- 2000
b_sim <- MASS::mvrnorm(B, mu = b, Sigma = V)

eta_sim <- b_sim %*% t(Xg)   # B x ngrid
mu_sim  <- inv_logit(eta_sim)

grid$beta_lwr <- apply(mu_sim, 2, quantile, 0.025)
grid$beta_upr <- apply(mu_sim, 2, quantile, 0.975)


library(ggplot2)

pred <- bind_rows(
  grid %>% transmute(MaxDHW.mean, model = "Logistic (quasi-binomial)",
                     fit = logit_fit, lwr = logit_lwr, upr = logit_upr),
  grid %>% transmute(MaxDHW.mean, model = "Beta regression",
                     fit = beta_fit, lwr = beta_lwr, upr = beta_upr)
)

x_max <- max(dat_plot2$MaxDHW.mean, na.rm = TRUE)

ggplot(dat_plot2, aes(MaxDHW.mean, decline_prop)) +

  annotate("rect", xmin = x_max, xmax = Inf, ymin = -Inf, ymax = Inf,
           fill = "grey80", alpha = 0.5) +
  geom_vline(xintercept = x_max, linetype = "dotted", linewidth = 0.8) +
  annotate("text",
           x = x_max + (15 - x_max) * 0.5,
           y = 0.07,
           label = "Extrapolation",
           colour = "grey20",
           size = 4) +

  geom_point(shape = 17, colour = "red", alpha = 0.35) +

  geom_ribbon(data = pred,
              aes(x = MaxDHW.mean, ymin = lwr, ymax = upr, fill = model),
              alpha = 0.15, inherit.aes = FALSE, show.legend = FALSE) +
  geom_line(data = pred,
            aes(x = MaxDHW.mean, y = fit, colour = model, linetype = model),
            linewidth = 1, inherit.aes = FALSE) +

  coord_cartesian(xlim = c(0, 15), ylim = c(0, 1)) +
  theme_classic() +
  labs(x = "DHW (°C-weeks)", y = "Mort01 (growth<0.5, decline>0.5)",
       colour = NULL, linetype = NULL)

```



# DMS API - Benthic Data

```{r Retrieve benthic data from AIMS API, cache=T}
# Add AIMS region to data
library(sf)
library(lwgeom)




df.AIMS.Bent <- df.AIMS.full |> 
  mutate(across(where(is.character), as.factor)) |> 
  filter(data_type == "photo-transect", domain_category == "reef",
         project_code == "LTMP", purpose == "GROUP_LEVEL", variable == "HARD CORAL") |> 
  select(domain_name,project_code,date,depth, lower:median, reef_zone:fid)

library(fuzzyjoin)
ci_str_detect <- function(x, y){str_detect(x, regex(y, ignore_case = TRUE))}
# create % change column
# df.BENT = df.AIMS |> rename("Reef_Name" = "domain_name") |>
#   fuzzy_left_join(select(dat.AIMSRef, Reef_Name, ReefID, ReefName, SECT_NAME, SECTOR),match_fun = ci_str_detect)

df.BENT.HC = df.AIMS.Bent |> rename("Reef_Name" = "domain_name") |>
  left_join(select(dat.AIMSRef, AIMS_REEF_NAME_cap, ReefID, ReefName, SECT_NAME, SECTOR), by=c("Reef_Name"="AIMS_REEF_NAME_cap"))

# df.BENT |> select(Reef_Name.x, Reef_Name.y:SECTOR) |> distinct() |> 
#   write.csv("Data/moreAIMSmatches.csv")

nomatch = df.BENT.HC|> filter(is.na(ReefName))
#Inshore MMP reef
datatable(df.BENT.HC %>% slice_head(n=50) |> 
            mutate(across(where(is.numeric), ~ round(., 4))),
          style = "default")

# find the max DHW for each reef
dat.yrs.bent.hc = df.BENT.HC|>  ungroup() |> 
  distinct(ReefName, date, report_year) |>
  filter(!is.na(ReefName)) |> 
  mutate(Date = as.Date(as.character(date)), Prev = report_year-1,
         Year = year(date), Month = month(date),
         DHWYear = ifelse(Month >=7 , Year, Year-1)) |> 
  # left_join(dplyr::select(dat.DHW, ReefName, Year,MaxDHW.mean), 
  #           by=c("ReefName", "report_year"="Year")) |> 
  # left_join(dplyr::select(dat.DHW, ReefName, Year,MaxDHW.mean), 
  #           by=c("ReefName", "Prev"="Year")) |> 
   left_join(dplyr::select(dat.DHW, ReefName, Year,MaxDHW.mean), 
            by=c("ReefName", "DHWYear"="Year")) #|> 
  # rowwise() |> 
  # mutate(MaxDHW = max(MaxDHW.mean.x, MaxDHW.mean.y)) |> 
  # select(-MaxDHW.mean.x, -MaxDHW.mean.y) |> 
  # mutate(MaxDHW = ifelse(is.na(MaxDHW), 0, MaxDHW))

# calculate % decline for each reef / taxa paairn
df.BENT.HC = df.BENT.HC %>% arrange(ReefName, Reef_Name, reefpage_category, depth, report_year)|>
  group_by(ReefName,Reef_Name, depth, reefpage_category) |> 
  mutate(Change = mean - lag(mean, n=1),
         Rel.Change = Change / lag(mean, n=1),
         Lag = report_year - lag(report_year, n=1)) |> 
  relocate(c("Change", "Lag", "Rel.Change"), .before = median) |> 
  left_join(dat.yrs.bent.hc)

df.BENT.HC = df.BENT.HC |> 
  mutate(Change.Cor = Change,
         Prev.HC = mean - Change.Cor) |> 
  filter(ReefName %in% dat.gridSum$REEF_NAME, !is.na(Lag),
         Reef_Name != "MILLN REEF") |> 
  rowwise() |> 
  mutate(Pred.Grth = doCoralGrowth.Simp(CoralCover=Prev.HC*100,
                  B0=dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),"B0"],
                  WQ = dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),"WQ"], 
                  HC.asym = dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),
                                                        "HC.asym"])[3]/100
         ) |> 
  mutate(Pred.Grth2 = ifelse(Lag==2, Pred.Grth+
                  doCoralGrowth.Simp(CoralCover=(Prev.HC+Pred.Grth)*100,
                  B0=dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),"B0"],
                  WQ = dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),"WQ"], 
                  HC.asym = dat.gridSum[which(dat.gridSum$REEF_NAME==ReefName),
                                                        "HC.asym"])[3]/100,
                  Pred.Grth)) |> 
  mutate(Abs.Change = Change.Cor - Pred.Grth2,
         Rel.Change = Abs.Change / (Prev.HC + Pred.Grth2))


library(dplyr)

dat.mod.bent.hc <- df.BENT.HC %>%
  filter(Lag < 3, project_code != "MMP") %>% 
  left_join(select(Disturbances_manta, ReefName, report_year, DISTURBANCE_TYPE), 
            by = c("ReefName", "report_year")) %>%
  filter(
    !DISTURBANCE_TYPE %in% c("c", "d", "s", "f"),
    report_year %in% c(1999, 2003, 2017, 2018, 2021, 2023, 2024)
  ) %>%
  mutate(
    Era = case_when(
      DHWYear %in% c(1998, 2002) ~ "1998/2002",
      DHWYear %in% c(2016, 2017) ~ "2016/2017",
      DHWYear %in% c(2020, 2022) ~ "2020/2022",
      DHWYear == 2024 ~ "2024",
      TRUE ~ NA_character_
    )
  ) |> 
  left_join(select(reefs_join,LOC_NAME_S,Region), by =c("ReefName"="LOC_NAME_S"))
dat.mod.benthcTerry = dat.mod.bent.hc |> 
  bind_rows(rename(dat.bleach, "Rel.Change" = "Change", "MaxDHW.mean" = "DHW") |> 
              mutate(Era = "Hughes et al 2018",
                     Rel.Change = Rel.Change/100 ))

ggplot(dat.mod.bent.hc, aes(x = MaxDHW.mean, y = Abs.Change, 
                         fill = Era, colour = Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + 
  scale_color_viridis_d(end = 0.8) +
  coord_cartesian(xlim = c(0, 12), ylim = c(-0.15, 0.15)) +
  scale_y_continuous(labels = scales::percent) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 3)) +  # k = 3 means 2 interior knots (3 knots total)
  theme_bw() + 
  facet_wrap(~Region)

ggplot(dat.mod.bent.hc, 
       aes(x = MaxDHW.mean, y = Rel.Change, 
                         fill = Era, colour = Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + 
  scale_color_viridis_d(end = 0.8) +
  coord_cartesian(xlim = c(0, 12), 
                  ylim = c(-1, 1)) +
  scale_y_continuous(labels = scales::percent) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 3)) +  # k = 3 means 2 interior knots (3 knots total)
  theme_bw() #+ 

ggplot(dat.mod.benthcTerry, 
       aes(x = MaxDHW.mean, y = Rel.Change, 
                         fill = Era, colour = Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + 
  scale_color_viridis_d(end = 0.8) +
  coord_cartesian(xlim = c(0, 12), 
                  ylim = c(-1, 1)) +
  scale_y_continuous(labels = scales::percent) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 3)) +  # k = 3 means 2 interior knots (3 knots total)
  theme_bw() #+ 
  
```

```{r}
ggplot(dat.mod.benthcTerry, 
       aes(x = MaxDHW.mean, y = Rel.Change, 
                         fill = Era, colour = Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + 
  scale_color_viridis_d(end = 0.8) +
  coord_cartesian(xlim = c(0, 12), 
                  ylim = c(-1, 1)) +
  scale_y_continuous(labels = scales::percent) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 3)) +  # k = 3 means 2 interior knots (3 knots total)
  theme_bw()


ggplot(dat.mod.mant , aes(x=MaxDHW.mean, y=Rel.Change, 
                     fill = Era,
                     colour=Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + scale_color_viridis_d(end=0.8) +
  coord_cartesian(xlim=c(0,12), ylim = c(-1,1)) +
  scale_y_continuous(labels = scales::percent) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  theme_bw() 

ggplot(dat.mod.mant, aes(x=MaxDHW.mean, y=Rel.Change)) +
  geom_point() +
  coord_cartesian(xlim=c(0,12),ylim = c(-1,1)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  scale_y_continuous(labels = scales::percent) +
  theme_bw() 

```

# Compare First vs Second Exposure

This section compares the mortality associated with the first exposure to \>4HW compared to subsequent exposures. This will be done in two ways first is to look at both the DHW-Mortality relationship and also a categorical response for each of bins of Heat exposure 4-8, 8-12.

```{r Compare repeat exposure}

dat.mod.repeat <- dat.mod.mant %>%
  group_by(Reef_Name) %>%
  filter(report_year > 2015) |> 
  arrange(Reef_Name, report_year) %>%
  mutate(
    ExposureNo = ifelse(MaxDHW.mean > 4, cumsum(MaxDHW.mean > 4), NA),
    ExposureCat = case_when(
      ExposureNo == 1 ~ "First",
      ExposureNo > 1 ~ "Repeat",
      TRUE ~ NA_character_
    ),
    StressCat = case_when(
      MaxDHW.mean < 4 ~ "0-4DHW",
      MaxDHW.mean < 8 ~ "4-8DHW",
      TRUE ~ "8-14DHW"
    )
  ) 

dat.mod.repeat <- dat.mod.mant %>%
  filter(report_year > 2015) %>%
  group_by(Reef_Name) %>%
  arrange(report_year, .by_group = TRUE) %>%
  mutate(
    exposed4 = MaxDHW.mean > 4,
    ExposureNo = cumsum(exposed4),                 # 0,1,2,... across years
    ExposureCat = case_when(
      exposed4 & ExposureNo == 1 ~ "First",
      exposed4 & ExposureNo  > 1 ~ "Repeat",
      TRUE ~ "NotExposed"
    ),
    StressCat = case_when(
      MaxDHW.mean < 4 ~ "0-4DHW",
      MaxDHW.mean < 8 ~ "4-8DHW",
      TRUE ~ "8-14DHW"
    )
  ) %>%
  ungroup()

library(sandwich)
library(lmtest)

dat_exp <- dat.mod.repeat %>%
  filter(ExposureCat %in% c("First", "Repeat")) %>%
  filter(is.finite(Abs.Change), is.finite(MaxDHW.mean))

m1 <- lm(Abs.Change ~ ExposureCat + MaxDHW.mean + Region, data = dat_exp)

# Cluster-robust SEs by reef
coeftest(m1, vcov = vcovCL(m1, cluster = ~ Reef_Name))

library(lme4)

m1_re <- lmer(Abs.Change ~ ExposureCat + MaxDHW.mean + Region + (1|Reef_Name), data = dat_exp)
summary(m1_re)

library(ggplot2)

ggplot(dat_exp |> filter(!is.na(ExposureCat)) , 
                                aes(x=MaxDHW.mean, y=Rel.Change, 
                     fill = ExposureCat, shape=ExposureCat,
                     colour=ExposureCat)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.98,option = "D", begin = 0.1, direction = -1) +
  scale_color_viridis_d(end= 0.98,option = "D", begin = 0.1,direction = -1) +
  # scale_fill_okabe_ito() + scale_color_okabe_ito() +
  coord_cartesian(xlim=c(4,12)) +
  # geom_smooth() +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 3)) + 
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~Region) +
  theme_bw()

Fig4Repeat <- ggplot(dat_exp |> filter(!is.na(ExposureCat)) , 
                                aes(x=MaxDHW.mean, y=Rel.Change, 
                     fill = ExposureCat, shape=ExposureCat,
                     colour=ExposureCat)) +
  geom_point(alpha=0.6) +
  scale_fill_viridis_d(end = 0.98,option = "D", begin = 0.1, direction = -1) +
  scale_color_viridis_d(end= 0.98,option = "D", begin = 0.1,direction = -1) +
  # scale_fill_okabe_ito() + scale_color_okabe_ito() +
  coord_cartesian(xlim=c(4,12)) +
  # geom_smooth() +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 3)) + 
  scale_y_continuous(labels = scales::percent) +
   labs(x = "DHW (°C-weeks)", y = "Relative coral cover change",
       fill = "Exposure", shape = "Exposure", colour = "Exposure") +
   theme(legend.position="bottom") +
   # facet_grid(~Region)
  # facet_wrap(~Region) +
  theme_bw()
Fig4Repeat
ggsave("output/plots/Fig4_RepeatExposure.png", Fig4Repeat, width= 6, height = 4, dpi = 300)
```




## COmpare change to frquency of events

```{r}
library(dplyr)

# 1A) Reef list (from the coral dataset, but we won't filter DHW to coral years)
reef_keys <- dat.mod.repeat %>%
  distinct(ReefName) %>%            # or distinct(Reef_Name) if that's what you have
  filter(!is.na(ReefName))

# 1B) Pull DHW series for those reefs since 2015 (all years available)
dhw_2015 <- dat.DHW %>%
  filter(ReefName %in% reef_keys$ReefName, Year >= 2015) %>%
  arrange(ReefName, Year) %>%
  group_by(ReefName) %>%
  mutate(
    ev4 = MaxDHW.mean > 4,
    ev8 = MaxDHW.mean > 8,

    # event numbering for >4 DHW (first / repeat)
    exp4_no = cumsum(ev4),
    exp4_cat = case_when(
      ev4 & exp4_no == 1 ~ "First",
      ev4 & exp4_no  > 1 ~ "Repeat",
      TRUE ~ "NotExposed"
    ),

    # (optional) same idea for >8 DHW if useful
    exp8_no = cumsum(ev8),
    exp8_cat = case_when(
      ev8 & exp8_no == 1 ~ "First",
      ev8 & exp8_no  > 1 ~ "Repeat",
      TRUE ~ "NotExposed"
    )
  ) %>%
  ungroup()

# 1C) Reef-level exposure summary from DHW (this is what you join to coral summary)
exposure_summary <- dhw_2015 %>%
  group_by(ReefName) %>%
  summarise(
    dhw_years = n_distinct(Year),
    first_dhw_year = min(Year, na.rm = TRUE),
    last_dhw_year  = max(Year, na.rm = TRUE),

    n_ev4 = sum(ev4, na.rm = TRUE),
    n_ev8 = sum(ev8, na.rm = TRUE),

    freq_ev4 = n_ev4 / dhw_years,
    freq_ev8 = n_ev8 / dhw_years,

    first_ev4_year = ifelse(any(ev4), min(Year[ev4], na.rm = TRUE), NA_integer_),
    first_ev8_year = ifelse(any(ev8), min(Year[ev8], na.rm = TRUE), NA_integer_),

    ever_repeat4 = any(exp4_no >= 2, na.rm = TRUE),
    ever_repeat8 = any(exp8_no >= 2, na.rm = TRUE),

    .groups = "drop"
  )

reef_summary <- dat.mod.repeat %>%
  filter(report_year > 2015) %>%
  filter(!is.na(ReefName), is.finite(mean)) %>%
  group_by(ReefName) %>%
  summarise(
    Region = first(Region),
    coral_years = n_distinct(report_year),
    year_start = min(report_year, na.rm = TRUE),
    year_end   = max(report_year, na.rm = TRUE),

    cover_start = mean[which.min(report_year)],
    cover_end   = mean[which.max(report_year)],

    total_cover_change = cover_end - cover_start,
    rel_total_change   = (cover_end - cover_start) / cover_start,

    .groups = "drop"
  ) %>%
  filter(coral_years >= 3, is.finite(rel_total_change), cover_start > 0, year_end>2023)

dist_summary <- reef_disturbance_df %>%
  filter(year > 2015, sample_type=="MANTA", disturbance=="c") %>%
  group_by(aims_reef_name, disturbance) %>%
  summarise(
    n_disturbances = n(),
    .groups = "drop"
  ) %>%
  left_join(dat.AIMSRef_key |> dplyr::select(aims_reef_name = AIMS_REEF_NAME_case, ReefName), by = "aims_reef_name") 

reef_summary2 <- reef_summary %>%
  left_join(exposure_summary, by = "ReefName") %>%
  left_join(dist_summary, by = "ReefName") %>%
  mutate(COTSOutbreak = case_when(
      n_disturbances > 0 ~ "COTS Outbreak", 
      is.na(n_disturbances) ~ "No Outbreak"),
      BleachingFreq = case_when(
        freq_ev8 == 0 ~ "0",
        freq_ev8 < 0.2 ~ "0-0.2/year",
        TRUE ~ "0.2/year +"
      ))

m_freq <- lm(rel_total_change ~ freq_ev8 + Region, data = reef_summary2)
summary(m_freq)


library(ggplot2)
ggplot(reef_summary2, aes(freq_ev8, total_cover_change, fill=COTSOutbreak, color = COTSOutbreak)) +
  geom_point(alpha = 0.7) +
  # geom_smooth(method = "gam", formula = y ~ s(x, k = 3), se = TRUE) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_y_continuous(labels = scales::percent) +
  theme_bw() +
  labs(x = "Frequency of >8 DHW years since 2015",
       y = "Total coral cover change since 2015")

Fig4Freq <- ggplot(reef_summary2, aes(BleachingFreq, total_cover_change, fill=COTSOutbreak)) +
  # geom_point(alpha = 0.7) +
  # geom_smooth(method = "gam", formula = y ~ s(x, k = 3), se = TRUE) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::percent) +
  theme_bw() +
  labs(x = "Frequency of >8 DHW years since 2015",
       y = "Total coral cover change since 2015", fill = NULL)
ggsave("output/plots/Fig4_Frequency.png", Fig4Freq, width = 6, height = 4, dpi = 300)

```


```{r}
library(ggplot2)
library(patchwork)

# 1) Make them look consistent
common_theme <- theme_bw(base_size = 11) +
  theme(
    plot.title = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(6, 6, 6, 6),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 9)
  )

pA <- Fig4Repeat + common_theme
pB <- Fig4Freq   + common_theme

# 2) Combine + panel tags
Fig4_AB <- (pA / pB) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(face = "bold", size = 14),
        plot.tag.position = c(0.95, 0.95))

Fig4_AB

ggsave("output/plots/Fig4_AB.png", Fig4_AB, width = 6, height = 8, dpi = 300, bg = "white")

pA2 <- pA + theme(legend.position = "bottom")
pB2 <- pB + theme(legend.position = "bottom")


# Okabe-Ito palette
cols_exposure <- c(
  "First"  = "#E69F00",  # orange
  "Repeat" = "#0072B2"   # blue
)

cols_outbreak <- c(
  "COTS Outbreak" = "#D55E00",  # vermillion
  "No Outbreak"   = "#009E73"   # bluish green/teal
)

# cols_exposure <- c(
#   "First"  = "#E3B23C",  # warm gold
#   "Repeat" = "#3F2E8C"   # indigo
# )
# 
# cols_outbreak <- c(
#   "COTS Outbreak" = "#E76F51",  # coral
#   "No Outbreak"   = "#2A9D8F"   # blue-green
# )

Fig4Repeat_h <- Fig4Repeat +
  scale_colour_manual(values = cols_exposure) +
  scale_fill_manual(values = cols_exposure) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50")

Fig4Freq_h <- Fig4Freq +
  scale_colour_manual(values = cols_outbreak) +
  scale_fill_manual(values = cols_outbreak) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") 
Fig4Repeat_h <- Fig4Repeat_h +
  guides(fill = guide_legend(override.aes = list(alpha = 0.25)))  # if you use ribbons


pA <- Fig4Repeat_h + common_theme
pB <- Fig4Freq_h  + common_theme

# 2) Combine + panel tags
Fig4_AB <- (pA / pB) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(face = "bold", size = 14),
        plot.tag.position = c(0.95, 0.95))

Fig4_AB

ggsave("output/plots/Fig4_AB.png", Fig4_AB, width = 6, height = 8, dpi = 300, bg = "white")
```


### MAp of Thermal Stress

```{r}
library(sf)
library(dplyr)
library(ggplot2)
library(stringr)

# ---- inputs you said you already have ----
# reefs : sf polygons (reef outlines)
# coast : sf polygon/lines (coastline)
# reef_summary2 : data.frame with ReefName, freq_ev8, rel_total_change (since 2015)

# Read sector shapefile (pick the .shp inside that folder)
shp <- list.files("data/SectorShapefile", pattern = "\\.shp$", full.names = TRUE)[1]
sectors <- st_read(shp, quiet = TRUE) |> st_make_valid()

# Choose the sector name/id column (edit if yours differs)
sect_id <- intersect(names(sectors), c("SECT_NAME","SECTOR","Sector","NAME","Name"))[1]
stopifnot(!is.na(sect_id))

# Make sure geometries share a CRS
target_crs <- st_crs(sectors)
coast <- st_transform(coast, target_crs)
reefs <- st_transform(reefs, target_crs)

# Choose reef-name column in reefs (edit if needed)
reef_id <- intersect(names(reefs), c("ReefName","REEF_NAME","LOC_NAME_S","domain_name","name"))[1]
stopifnot(!is.na(reef_id))

# Reef points = centroids/point-on-surface (point-on-surface is safer for polygons)
reef_pts <- st_point_on_surface(reefs) |>
  mutate(ReefName = .data[[reef_id]]) |>
  left_join(reef_summary2, by = "ReefName") |>
  filter(is.finite(freq_ev8), is.finite(rel_total_change))

# Spatial join reefs -> sectors to get sector id per reef
reef_pts2 <- st_join(
  reef_pts,
  sectors |> dplyr::select(!!sect_id),
  left = TRUE
)

# Sector-level summaries
sector_stats <- reef_pts2 |>
  st_drop_geometry() |>
  group_by(sector = .data[[sect_id]]) |>
  summarise(
    n_reefs = n(),
    freq_mean = mean(freq_ev8, na.rm = TRUE),
    freq_max  = max(freq_ev8, na.rm = TRUE),
    chg_mean  = mean(total_cover_change, na.rm = TRUE),
    chg_se    = sd(total_cover_change, na.rm = TRUE) / sqrt(sum(is.finite(total_cover_change))),
    .groups = "drop"
  )

# Join stats back onto sector polygons
sectors2 <- sectors |>
  mutate(sector = .data[[sect_id]]) |>
  left_join(sector_stats, by = "sector")

# Label positions + label text
sector_labs <- sectors2 |>
  filter(is.finite(freq_mean)) |>
  mutate(
    lab_pt = st_point_on_surface(geometry),
    label = sprintf(
      "freq>8: %.2f\na max: %.2f\nΔcover: %+.0f ± %.0f%%",
      freq_mean, freq_max, 100*chg_mean, 100*chg_se
    )
  )

```
## Plot by Sectors

```{r}
# Harmonious continuous palette (low -> high)
pal <- c("#2A9D8F", "#E9C46A", "#E76F51")  # teal -> sand -> coral

FigMapFreq <- ggplot() +

  # coast behind everything
  geom_sf(data = coast, fill = "grey92", colour = "grey50", linewidth = 0.2) +

  # sector polygons coloured by mean frequency
  geom_sf(
    data = sectors2,
    aes(fill = freq_mean),
    colour = "grey25",
    linewidth = 0.25,
    alpha = 0.95
  ) +

  # reef outlines (very light)
  # geom_sf(data = reefs, fill = NA, colour = "grey60", linewidth = 0.08, alpha = 0.4) +

  # reef points (size + colour by reef-level frequency)
  geom_sf(
    data = reef_pts2,
    aes(colour = freq_ev8, size = freq_ev8*10),
    alpha = 0.9,
  ) +

  # sector labels (adjust size as needed)
  geom_sf_text(
    data = sector_labs,
    aes(geometry = lab_pt, label = label),
    size = 3.0,
    colour = "black",
    lineheight = 0.95
  ) +

  # One palette for both fill and colour
  scale_fill_gradientn(colours = pal, limits = c(0, 0.5), name = "Freq >8 DHW") +
  scale_colour_gradientn(colours = pal, limits = c(0, 0.5), name = "Freq >8 DHW") +

  # Size scale (hide legend to avoid clutter; colourbar is enough)
  scale_size(range = c(0.8, 4.2), limits = c(0, 1), guide = "none") +

  coord_sf(expand = FALSE) +
  theme_classic(base_size = 12) +
  theme(
    axis.title = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "right"
  ) +
  labs(title = NULL)

FigMapFreq

```

```{r}
library(sf)
library(dplyr)

# degrees -> 2x2 rotation matrix (sf vignette style)
rot_mat <- function(deg){
  a <- deg * pi / 180
  matrix(c(cos(a), sin(a), -sin(a), cos(a)), nrow = 2, ncol = 2)
}

# rotate an sf/sfc around a single (x,y) centre in projected units
rotate_sf <- function(x, deg, center_xy){
  stopifnot(length(center_xy) == 2)

  g <- st_geometry(x)
  ctr <- st_sfc(st_point(center_xy), crs = st_crs(x))

  # (g - ctr) * R + ctr  (sf affine ops)
  g2 <- (g - ctr) * rot_mat(deg) + ctr

  st_geometry(x) <- g2
  x
}

crs_proj <- 3577

sectors_p <- st_transform(sectors2, crs_proj) |> st_make_valid()
reefs_p   <- st_transform(reefs,    crs_proj) |> st_make_valid()
coast_p   <- st_transform(coast,    crs_proj) |> st_make_valid()

reef_pts_p    <- st_transform(reef_pts2,    crs_proj)
sector_labs_p <- st_transform(sector_labs,  crs_proj)



# choose a single rotation centre (union centroid of sectors is a good anchor)
ctr_xy <- st_coordinates(st_centroid(st_union(sectors_p)))[1,]

angle_deg <- 25 # clockwise; use +45 if you want the other direction

sectors_r <- rotate_sf(sectors_p, angle_deg, ctr_xy)
reefs_r   <- rotate_sf(reefs_p,   angle_deg, ctr_xy)
coast_r   <- rotate_sf(coast_p,   angle_deg, ctr_xy)
plot(coast_r["geometry"], main = "Rotated Coast", key.pos = NULL)
reef_pts_r    <- rotate_sf(reef_pts_p,    angle_deg, ctr_xy)
sector_labs_r <- rotate_sf(sector_labs_p, angle_deg, ctr_xy)
plot(sectors_r["freq_mean"], main = "Rotated Sectors", key.pos = NULL)

bb <- st_bbox(sectors_r)

pad <- 100  # ~120 km; tweak until it matches your reference look

xlim <- c(bb["xmin"] - pad, bb["xmax"] + pad)
ylim <- c(bb["ymin"] - pad, bb["ymax"] + pad)

bb2 <- bb
bb2[c("xmin","ymin")] <- bb2[c("xmin","ymin")] - pad
bb2[c("xmax","ymax")] <- bb2[c("xmax","ymax")] + pad

coast_r   <- st_crop(coast_r,   bb2)

pal <- c("#2A9D8F", "#E9C46A", "#E76F51")

ggplot() +
  geom_sf(data = coast_r, fill=NA,colour = "grey50", linewidth = 0.2) +
  geom_sf(data = reefs_r, fill = NA, colour = "grey60", linewidth = 0.08, alpha = 0.5) +
  geom_sf(data = sectors_r, aes(fill = freq_mean), colour = "grey25", linewidth = 0.25, alpha=0.8) +
  
  geom_sf(data = reef_pts_r, aes(colour = freq_ev8, size = freq_ev8), alpha = 0.9) +
  geom_sf_text(
    data = sector_labs_r,
    aes(geometry = lab_pt, label = label),
    size = 3.0, lineheight = 0.95
  ) +
  scale_fill_gradientn(colours = pal, limits = c(0, 0.3), name = "Freq >8 DHW") +
  scale_colour_gradientn(colours = pal, limits = c(0, 0.3), name = "Freq >8 DHW") +
  scale_size(range = c(0.8, 4.2), limits = c(0, 1), guide = "none") +
  coord_sf(xlim = xlim+9000, ylim = ylim, expand = FALSE, datum = NA) +
  theme_classic(base_size = 12) +
  theme(
    axis.title = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )


```



```{r}
ggplot(dat.mod.repeat |> filter(!is.na(ExposureCat)) , 
                                aes(x=MaxDHW.mean, y=Abs.Change, 
                     fill = ExposureCat, shape=ExposureCat,
                     colour=ExposureCat)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.98,option = "D", begin = 0.1, direction = -1) +
  scale_color_viridis_d(end= 0.98,option = "D", begin = 0.1,direction = -1) +
  # scale_fill_okabe_ito() + scale_color_okabe_ito() +
  coord_cartesian(xlim=c(4,12)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~Region) +
  theme_bw()


ggplot(dat.mod.repeat |> filter(ExposureCat=="First") , 
                                aes(x=MaxDHW.mean, y=Rel.Change, 
                     fill = Region, shape=Region,
                     colour=Region)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.98,option = "D", begin = 0.1, direction = -1) +
  scale_color_viridis_d(end= 0.98,option = "D", begin = 0.1,direction = -1) +
  # scale_fill_okabe_ito() + scale_color_okabe_ito() +
  coord_cartesian(xlim=c(4,12)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  scale_y_continuous(labels = scales::percent) +
  #facet_wrap(~Region) +
  theme_bw()

ggplot(dat.mod.repeat |> filter(!is.na(ExposureCat)) , 
       aes(x=MaxDHW.mean, y=Rel.Change, 
                     fill = ExposureCat, shape=ExposureCat,
                     colour=ExposureCat)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.98,option = "D", begin = 0.1, direction = -1) +
  scale_color_viridis_d(end= 0.98,option = "D", begin = 0.1,direction = -1) +
  # scale_fill_okabe_ito() + scale_color_okabe_ito() +
  coord_cartesian(xlim=c(4,12)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  scale_y_continuous(labels = scales::percent) +
  # facet_wrap(~SECTOR_NAME) +
  facet_wrap(~Region) +
  theme_bw()   

ggplot(dat.mod.repeat |> filter(!is.na(ExposureCat)) , 
       aes(x=MaxDHW.mean, y=Rel.Change, 
                     fill = ExposureCat, shape=ExposureCat,
                     colour=ExposureCat)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.98,option = "D", begin = 0.1, direction = -1) +
  scale_color_viridis_d(end= 0.98,option = "D", begin = 0.1,direction = -1) +
  # scale_fill_okabe_ito() + scale_color_okabe_ito() +
  coord_cartesian(xlim=c(4,12)) +
  # geom_smooth() +
  geom_smooth(method="gam") +
  scale_y_continuous(labels = scales::percent) +
  # facet_wrap(~SECTOR_NAME) +
  # facet_wrap(~Region) +
  theme_bw() 



ggplot(dat.mod.repeat |> filter(!is.na(ExposureCat)) , 
       aes(x=StressCat, y=Rel.Change, 
                     fill = ExposureCat, shape=ExposureCat)) +
  geom_boxplot() +
  scale_fill_viridis_d(end = 0.98,option = "D", begin = 0.1, direction = -1) +
  scale_color_viridis_d(end= 0.98,option = "D", begin = 0.1,direction = -1) +
  # scale_fill_okabe_ito() + scale_color_okabe_ito() +
  # geom_smooth() +
  # geom_smooth(method="gam") +
  scale_y_continuous(labels = scales::percent) +
  # facet_wrap(~SECTOR_NAME) +
  theme_bw()  

ggplot(dat.mod.repeat |> filter(!is.na(ExposureCat)) , 
       aes(x=StressCat, y=Change.Cor/100, 
                     fill = ExposureCat, shape=ExposureCat)) +
  geom_boxplot() +
  scale_fill_viridis_d(end = 0.98,option = "D", begin = 0.1, direction = -1) +
  scale_color_viridis_d(end= 0.98,option = "D", begin = 0.1,direction = -1) +
  # scale_fill_okabe_ito() + scale_color_okabe_ito() +
  # geom_smooth() +
  # geom_smooth(method="gam") +
  scale_y_continuous(labels = scales::percent) +
  # facet_wrap(~SECTOR_NAME) +
  theme_bw() 


ggplot(dat.mod.repeat |> filter(!is.na(ExposureCat)) , 
       aes(x=StressCat, y=Change.Cor/100, 
                     fill = ExposureCat, shape=ExposureCat)) +
  geom_boxplot() +
  scale_fill_viridis_d(end = 0.98,option = "D", begin = 0.1, direction = -1) +
  scale_color_viridis_d(end= 0.98,option = "D", begin = 0.1,direction = -1) +
  # scale_fill_okabe_ito() + scale_color_okabe_ito() +
  # geom_smooth() +
  # geom_smooth(method="gam") +
  scale_y_continuous(labels = scales::percent) +
  # facet_wrap(~SECTOR_NAME) +
  theme_bw()  

```

# DHW - Aerial Score Relationship

This will tell us if corals are adapting to bleaching itself or are more resilient towards the associated mortality

# Disturbance Layers - Cyclones, Bleaching, COTS

## Assumptions and Issues to be addressed

-   4km/5km resolution coarse scale for reef level only.

-   Need to check if cyclone exposure accounts for reef structure for sheltered vs exposed sides.

-   Need to decide if reefs are split into flanks or we use reef wide averages.

# References
