---
title: "DHW–Mortality Attribution on the Great Barrier Reef"
author: "Samuel Matthews"
date: "2024-05-29"
date-modified: today
# bibliography: references.bib
# link-citations: true
format:
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 3
    toc-location: left
    self-contained: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
execute:
  warning: false
  message: false
  echo: true
---

# Introduction

This document quantifies the relationship between accumulated thermal stress (Degree Heating Weeks; DHW) and coral mortality on the Great Barrier Reef (GBR). The analysis extends existing DHW–mortality frameworks (e.g., Hughes et al. 2018, ReefMod) by:
 
1. **Growth-adjusting** mortality estimates using reef-specific Gompertz coral growth parameters from @macneil2019, so that expected background growth is accounted for when estimating bleaching-attributable loss.
2. **Comparing data sources**: manta tow (broad spatial coverage, ~6–9 m depth), benthic photo-transects (precise, multiple depths), and MMP inshore monitoring.
3. **Stratifying by era and region** to examine whether the DHW–mortality relationship has changed across successive mass bleaching events (1998/2002, 2016/2017, 2020/2022, 2024).
4. **Assessing repeat exposure effects** — do reefs that have experienced prior bleaching show different mortality responses?
5. **Building a machine-learning mortality model** by integrating environmental predictors from Cheung et al. (2025) — SST trajectory clusters, thermal history, water clarity, currents, and cloud cover.

## Key Assumptions & Caveats

- Gompertz growth parameters are averaged at reef level from 1 km grid cells; future work should propagate parameter uncertainty.
- Mortality attribution assumes bleaching is the primary driver in filtered bleaching years; co-occurring disturbances (COTS, cyclones) are excluded via LTMP disturbance records but residual confounding is possible.
- The Cheung et al. predictor dataset covers 2016, 2017, 2020 only; the ML model is restricted to these years as a proof-of-concept.

```{r setup}
#| label: setup

# ── Core libraries ──
library(tidyr)
library(dplyr)
library(ggplot2)
library(DT)
library(arrow)
library(stringr)
library(janitor)
library(lubridate)
library(readr)
library(sf)
library(lwgeom)
library(scales)
library(patchwork)

# ── Modelling libraries ──
library(mgcv)
library(betareg)
library(MASS)
library(lme4)
library(sandwich)
library(lmtest)

# ── Shared colour palettes ──
dist_colours <- c(
  "b" = "darkorange", "B" = "darkorange",
  "c" = "maroon4",    "C" = "maroon4",
  "d" = "darkolivegreen3", "D" = "darkolivegreen3",
  "f" = "burlywood4", "F" = "burlywood4",
  "s" = "dodgerblue2", "S" = "dodgerblue2",
  "u" = "grey",       "U" = "grey"
)

cols_exposure <- c("First" = "#E69F00", "Repeat" = "#0072B2")
cols_outbreak <- c("COTS Outbreak" = "#D55E00", "No Outbreak" = "#009E73")

secorder <- c("CG", "PC", "CL", "CA", "IN", "TO", "CU", "WH", "PO", "SW", "CB")
secorderlabs <- c("Cape Grenville", "Princess Charlotte Bay", "Cooktown / Lizard Island",
                  "Cairns", "Innisfail", "Townsville", "Cape Upstart", "Whitsunday",
                  "Pompey", "Swain", "Capricorn Bunker")

# Force dplyr functions over MASS/stats
select <- dplyr::select
filter <- dplyr::filter
```

```{r helper-functions}
#| label: helper-functions

# ── Gompertz coral growth function (simplified, single time-step) ──
# Returns c(current_cover, next_cover, change)
doCoralGrowth.Simp <- function(CoralCover, B0, HC.asym, WQ) {
  WQ.mn.sd <- c(-0.68, 0.03)
  b0.wq <- B0 + WQ * WQ.mn.sd[1]
  b1.wq <- b0.wq / log(HC.asym)
  CoralCover <- log(CoralCover)
  b0.wq <- as.numeric(b0.wq)
  b1.wq <- as.numeric(b1.wq)
  CoralCover <- as.numeric(CoralCover)
  CoralCover.t1 <- (b0.wq + (1 - b1.wq) * CoralCover)
  CoralCover.t1 <- exp(CoralCover.t1)
  Change.t1 <- CoralCover.t1 - exp(CoralCover)
  return(c(exp(CoralCover), CoralCover.t1, Change.t1))
}

# ── Multi-step growth function (n years from starting cover) ──
doCoralGrowth <- function(n, CoralCover, B0, HC.asym, WQ) {
  WQ.mn.sd <- c(-0.68, 0.03)
  b0.wq <- B0 + WQ * WQ.mn.sd[1]
  b1.wq <- b0.wq / log(HC.asym)
  CoralCover <- log(CoralCover)
  b0.wq <- as.numeric(b0.wq)
  b1.wq <- as.numeric(b1.wq)
  CoralCover <- as.numeric(CoralCover)
  df <- matrix(nrow = n + 1, ncol = 3)
  df[1, 1] <- 0
  df[1, 2] <- exp(CoralCover)
  for (i in 1:n) {
    CoralCover.t1 <- (b0.wq + (1 - b1.wq) * CoralCover)
    df[i + 1, 1] <- i
    df[i + 1, 2] <- exp(CoralCover.t1)
    df[i + 1, 3] <- exp(CoralCover.t1) - exp(CoralCover)
    CoralCover <- CoralCover.t1
  }
  colnames(df) <- c("Year", "CoralCover", "CoralGrowth")
  return(df)
}

# ── Inverse logit ──
inv_logit <- function(x) 1 / (1 + exp(-x))
```

# Growth Curves & Recovery Times

Reef-specific Gompertz growth parameters from @macneil2019 are used to model coral cover trajectories. The parameters (`B0` = intrinsic growth rate, `HC.asym` = asymptotic carrying capacity) are modified by local water quality (`WQ`) to produce site-specific recovery curves.

The growth function follows the discrete Gompertz form:
$$\log(C_{t+1}) = b_0^{WQ} + (1 - b_1^{WQ}) \cdot \log(C_t)$$

where $b_0^{WQ} = B_0 + WQ \times \beta_{WQ}$ and $b_1^{WQ} = b_0^{WQ} / \log(HC_{asym})$.

```{r growth-params}
#| label: growth-params
#| cache: true

dat.grid <- read.csv("Data/COTSModParams.csv")

dat.gridSum <- dat.grid %>%
  mutate(
    REEF_ID = ifelse(grepl("14-116", REEF_ID, fixed = TRUE), "14-116", REEF_ID),
    REEF_NAME = ifelse(grepl("14-116", REEF_NAME, fixed = TRUE),
                       "Lizard Island Reef (14-116)", REEF_NAME)
  ) %>%
  group_by(REEF_NAME, REEF_ID, SECTOR) %>%
  summarise(
    B0 = mean(pred.b0.mean),
    HC.asym = mean(pred.HCmax.mean),
    WQ1 = mean(Primary),
    WQ2 = mean(Secondary),
    WQ3 = mean(Tertiary),
    .groups = "drop"
  ) %>%
  mutate(
    WQ = WQ1 + WQ2 + WQ3,
    HC90 = 0.8 * HC.asym
  )
```

```{r growth-curves}
#| label: growth-curves
#| cache: true
#| fig-cap: "Modelled coral recovery trajectories from 0.1% cover, by sector. Ribbons show inter-reef variability."

# Simulate 30-year recovery from 0.1% cover for all reefs
CC.tmp <- data.frame()
for (i in seq_len(nrow(dat.gridSum))) {
  CCtmp2 <- as.data.frame(
    doCoralGrowth(30, 0.1, dat.gridSum$B0[i], dat.gridSum$HC.asym[i], dat.gridSum$WQ[i])
  )
  CCtmp2$ReefName <- dat.gridSum$REEF_NAME[i]
  CCtmp2$REEF_NAME <- dat.gridSum$REEF_NAME[i]
  CCtmp2$ReefID <- dat.gridSum$REEF_ID[i]
  CCtmp2$Sector <- dat.gridSum$SECTOR[i]
  CC.tmp <- bind_rows(CC.tmp, CCtmp2)
}

CC.tmp <- CC.tmp %>% left_join(select(dat.gridSum, REEF_NAME, HC90), by = "REEF_NAME")

library(ggdist)
ggplot(CC.tmp, aes(x = Year, y = CoralCover)) +
  stat_lineribbon(alpha = 0.3) +
  facet_wrap(~Sector) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = "Years since disturbance", y = "Coral cover (%)")
```



# DHW–Mortality Relationship

This section examines the existing DHW–mortality relationship using the Hughes et al. (2018) dataset, comparing beta regression versus binomial/quasi-binomial GLM approaches. The key question is whether the sigmoidal binomial curve overpredicts mortality at high DHW values.

## Hughes et al. Data: Beta vs Binomial Regression

```{r dhw-mortality-comparison}
#| label: dhw-mortality-comparison

dat.bleach <- read.csv("Data/DHW_vs_Coral_Cover_Dataset.csv") %>%
  mutate(Mort.bin = ifelse(Change >= 0, 0.001, -Change / 100))

mod.beta <- betareg(Mort.bin ~ DHW, data = dat.bleach)
mod.bin <- glm(Mort.bin ~ DHW, data = dat.bleach, family = binomial(link = "logit"))

# Predictions
newdata <- with(dat.bleach, data.frame(DHW = seq(0, 16, len = 100)))
fit_bin <- predict(mod.bin, newdata = newdata, type = "link", se = TRUE)
q <- qt(0.975, df = df.residual(mod.bin))
newdata <- cbind(newdata,
  fit = binomial()$linkinv(fit_bin$fit),
  lower = binomial()$linkinv(fit_bin$fit - q * fit_bin$se.fit),
  upper = binomial()$linkinv(fit_bin$fit + q * fit_bin$se.fit)
)

newdata2 <- with(dat.bleach, data.frame(DHW = seq(0, 16, len = 100)))
fit_beta <- predict(mod.beta, newdata = newdata2, type = "link")
newdata2 <- cbind(newdata2, fit = binomial()$linkinv(fit_beta))
```

```{r dhw-mortality-plot}
#| label: dhw-mortality-plot
#| fig-cap: "Comparison of binomial (black) and beta regression (red) fits to the DHW–mortality relationship from Hughes et al. (2018) data."

ggplot(data = newdata, aes(y = fit, x = DHW)) +
  geom_point(data = dat.bleach, aes(y = Mort.bin)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "blue", alpha = 0.3) +
  geom_line() +
  geom_line(data = newdata2, aes(y = fit, x = DHW), colour = "red") +
  theme_classic() +
  labs(x = "DHW (°C-weeks)", y = "Mortality proportion",
       caption = "Black = binomial GLM, Red = beta regression")
```


# Manta Tow Analysis (Updated AIMS API)

This section uses the locally cached AIMS LTMP manta tow data to construct a comprehensive DHW–mortality dataset. Growth correction is applied using reef-specific Gompertz parameters.

```{r manta-tow-data}
#| label: manta-tow-data
#| cache: true

load("Data/aims_ltmp/aims_ltmp.RData")

dat.AIMSRef <- read.csv("Data/AIMS-Reef_Reference.csv") %>%
  rename(Reef_Name = AIMS_REEF_NAME)

Disturbances_manta <- read.csv("Data/Disturbances_manta.csv") %>%
  left_join(select(dat.AIMSRef, Reef_Name, ReefName), by = c("AIMS_REEF_NAME" = "Reef_Name")) %>%
  rename("report_year" = "REPORT_YEAR")

# DHW data
dat.DHW <- read.csv("Data/DHW_1985_2024_GBRReefs.csv") %>%
  rename("ReefName" = "LOC_NAME_S")

# Fill Agincourt reefs (missing individual DHW)
dat.agin <- dat.DHW |>
  filter(grepl("Agincourt", ReefName)) |>
  group_by(Year) |>
  summarise(MaxDHW.mean = mean(MaxDHW.mean, na.rm = TRUE), .groups = "drop")

dat.DHW <- dat.DHW |>
  left_join(dat.agin, by = "Year") |>
  mutate(MaxDHW.mean = ifelse(grepl("Agincourt", ReefName) & is.na(MaxDHW.mean.x),
                              MaxDHW.mean.y, MaxDHW.mean.x)) |>
  select(-MaxDHW.mean.x, -MaxDHW.mean.y)

df.AIMS.Mant <- reef_manta_df |>
  select(domain_name, project_code, report_year, date, depth,
         lower:median, mean, reef_zone, id, shelf, reefpage_category) |>
  mutate(Reef_Name = toupper(domain_name))

library(fuzzyjoin)

dat.AIMSRef_key <- dat.AIMSRef %>%
  filter(!ReefName %in% c("Round-Russell Reef (17-013)", "Snake Reef (14-087)")) %>%
  mutate(AIMS_REEF_NAME_case = str_to_title(AIMS_REEF_NAME_cap))

df.MANT <- df.AIMS.Mant |>
  left_join(select(dat.AIMSRef_key, AIMS_REEF_NAME_cap, ReefID, ReefName, SECT_NAME, SECTOR),
            by = join_by("Reef_Name" == "AIMS_REEF_NAME_cap")) %>%
  mutate(
    year = floor(as.numeric(date)),
    frac = as.numeric(date) - year,
    date_approx = ymd(paste0(year, "-01-01")) +
      round(frac * as.numeric(ymd(paste0(year + 1, "-01-01")) - ymd(paste0(year, "-01-01"))))
  ) %>%
  select(-year, -frac)

# DHW year assignment
dat.yrs.mant <- df.MANT |> ungroup() |>
  filter(!is.na(ReefName)) |>
  mutate(
    Date = date_approx, Prev = report_year - 1,
    Year = year(Date), Month = month(Date),
    DHWYear = ifelse(Month >= 7, Year, Year - 1)
  ) |>
  left_join(dplyr::select(dat.DHW, ReefName, Year, MaxDHW.mean),
            by = c("ReefName", "DHWYear" = "Year"))

# Calculate change per reef
df.MANT <- df.MANT %>%
  arrange(ReefName, Reef_Name, reefpage_category, depth, report_year) |>
  group_by(ReefName, Reef_Name, depth, reefpage_category) |>
  mutate(
    Change = mean - lag(mean, n = 1),
    Rel.Change = Change / lag(mean, n = 1),
    Lag = report_year - lag(report_year, n = 1)
  ) |>
  relocate(c("Change", "Lag", "Rel.Change"), .before = median) |>
  left_join(dat.yrs.mant)

# Growth correction
df.MANT <- df.MANT |>
  mutate(Change.Cor = Change, Prev.HC = mean - Change.Cor) |>
  filter(ReefName %in% dat.gridSum$REEF_NAME, !is.na(Lag),
         Reef_Name != "MILLN REEF") |>
  rowwise() |>
  mutate(
    Pred.Grth = doCoralGrowth.Simp(
      CoralCover = Prev.HC * 100,
      B0 = dat.gridSum[which(dat.gridSum$REEF_NAME == ReefName), "B0"],
      WQ = dat.gridSum[which(dat.gridSum$REEF_NAME == ReefName), "WQ"],
      HC.asym = dat.gridSum[which(dat.gridSum$REEF_NAME == ReefName), "HC.asym"]
    )[3] / 100
  ) |>
  mutate(
    Pred.Grth2 = ifelse(Lag == 2,
      Pred.Grth + doCoralGrowth.Simp(
        CoralCover = (Prev.HC + Pred.Grth) * 100,
        B0 = dat.gridSum[which(dat.gridSum$REEF_NAME == ReefName), "B0"],
        WQ = dat.gridSum[which(dat.gridSum$REEF_NAME == ReefName), "WQ"],
        HC.asym = dat.gridSum[which(dat.gridSum$REEF_NAME == ReefName), "HC.asym"]
      )[3] / 100,
      Pred.Grth),
    Abs.Change = Change.Cor - Pred.Grth2,
    Rel.Change = Abs.Change / (Prev.HC + Pred.Grth2)
  )
```

```{r manta-model-data}
#| label: manta-model-data
#| cache: true

# Load spatial regions
gbr <- st_read("Data/GBR_AIMS") |> st_make_valid()
data_bucket <- s3_bucket("s3://gbr-dms-data-public/gbrmpa-complete-gbr-features/data.parquet")
reefs <- open_dataset(data_bucket) %>%
  filter(FEAT_NAME == "Reef", DATASET == "GBR Features") |>
  distinct(UNIQUE_ID, GBR_NAME, LOC_NAME_S, geometry, FEAT_NAME) %>%
  collect() %>% st_as_sf(crs = 4283) |> st_make_valid()

coast <- open_dataset(data_bucket) %>%
  filter(FEAT_NAME %in% c("Mainland", "Island"), DATASET == "GBR Features") |>
  distinct(UNIQUE_ID, GBR_NAME, LOC_NAME_S, geometry, FEAT_NAME) %>%
  collect() %>% st_as_sf(crs = 4283) |> st_make_valid()
reefs_centroids <- st_centroid(reefs)
reefs_join <- st_join(reefs_centroids, gbr)

dat.mod.mant <- df.MANT %>%
  filter(Lag < 3, project_code != "MMP", abs(Change) > 0.01) %>%
  left_join(select(Disturbances_manta, ReefName, report_year, DISTURBANCE_TYPE),
            by = c("ReefName", "report_year")) %>%
  filter(
    !DISTURBANCE_TYPE %in% c("c", "d", "s", "f"),
    report_year %in% c(1999, 2003, 2017, 2018, 2021, 2023, 2024, 2025)
  ) %>%
  mutate(
    Era = case_when(
      DHWYear %in% c(1998, 2002) ~ "1998/2002",
      DHWYear %in% c(2016, 2017) ~ "2016/2017",
      DHWYear %in% c(2020, 2022) ~ "2020/2022",
      DHWYear == 2024 ~ "2024",
      TRUE ~ NA_character_
    )
  ) |>
  left_join(select(reefs_join, LOC_NAME_S, Region), by = c("ReefName" = "LOC_NAME_S"))
```

```{r manta-plots}
#| label: manta-plots
#| fig-cap: "Manta tow DHW vs relative change by era and region."

ggplot(dat.mod.mant,
       aes(x = MaxDHW.mean, y = Rel.Change, fill = Era, colour = Era)) +
  geom_point(alpha = 0.5) +
  scale_fill_viridis_d(end = 0.8) +
  scale_color_viridis_d(end = 0.8) +
  coord_cartesian(xlim = c(0, 12), ylim = c(-1, 1)) +
  scale_y_continuous(labels = scales::percent) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4)) +
  theme_bw() +
  labs(x = "DHW (°C-weeks)", y = "Relative change (growth-adjusted)")
```

# Model Comparison: LM vs GAM

We compare linear model and GAM fits to the DHW–mortality relationship, stratified by era and region. This helps assess whether the relationship is stable across events or whether more recent events show different sensitivity.

```{r lm-vs-gam}
#| label: lm-vs-gam
#| fig-cap: "LM vs GAM fits stratified by bleaching era."

dat.mod.mantTerry <- dat.mod.mant |>
  bind_rows(rename(dat.bleach, "Rel.Change" = "Change", "MaxDHW.mean" = "DHW") |>
              mutate(Era = "Hughes et al 2018", Rel.Change = Rel.Change / 100))

dat_plot <- dat.mod.mantTerry %>%
  filter(is.finite(MaxDHW.mean), is.finite(Rel.Change)) %>%
  mutate(Region = case_when(
    Era == "Hughes et al 2018" ~ "Hughes et al 2018",
    TRUE ~ Region
  )) %>%
  mutate(Era = factor(Era), Region = factor(Region)) %>%
  filter(!is.na(Era), !is.na(Region))

# Prediction function for LM + GAM
fit_predict <- function(dat_sub, grid_sub, k = 4) {
  m_lm <- lm(Rel.Change ~ MaxDHW.mean, data = dat_sub)
  pr_lm <- predict(m_lm, newdata = grid_sub, se.fit = TRUE)
  out_lm <- grid_sub %>%
    mutate(model = "LM", fit = pr_lm$fit,
           lwr = pr_lm$fit - 1.96 * pr_lm$se.fit,
           upr = pr_lm$fit + 1.96 * pr_lm$se.fit)

  m_gam <- mgcv::gam(Rel.Change ~ s(MaxDHW.mean, k = k), data = dat_sub, method = "REML")
  pr_gam <- predict(m_gam, newdata = grid_sub, se.fit = TRUE)
  out_gam <- grid_sub %>%
    mutate(model = paste0("GAM (k=", k, ")"), fit = pr_gam$fit,
           lwr = pr_gam$fit - 1.96 * pr_gam$se.fit,
           upr = pr_gam$fit + 1.96 * pr_gam$se.fit)

  bind_rows(out_lm, out_gam)
}

grid <- dat_plot %>%
  distinct(Era) %>%
  tidyr::expand_grid(MaxDHW.mean = seq(0, 20, by = 0.05))

pred.era <- dat_plot %>%
  group_by(Era) %>%
  group_modify(~{
    grid_sub <- grid %>% filter(Era == .y$Era) %>% dplyr::select(MaxDHW.mean)
    fit_predict(.x, grid_sub, k = 3)
  }) %>% ungroup()

x_max <- max(dat_plot$MaxDHW.mean, na.rm = TRUE)

Fig3.Eras <- ggplot(dat_plot, aes(MaxDHW.mean, Rel.Change, colour = Era, fill = Era)) +
  annotate("rect", xmin = x_max, xmax = 20, ymin = -Inf, ymax = Inf,
           fill = "grey80", alpha = 0.5) +
  geom_vline(xintercept = x_max, linetype = "dotted", linewidth = 0.8) +
  geom_point(shape = 17, alpha = 0.5) +
  geom_ribbon(data = pred.era,
              aes(x = MaxDHW.mean, y = fit, ymin = lwr, ymax = upr, group = Era),
              alpha = 0.12, colour = NA, show.legend = FALSE) +
  geom_line(data = pred.era, aes(x = MaxDHW.mean, y = fit), linewidth = 1) +
  scale_color_viridis_d(option = "A", begin = 0.2, end = 0.9) +
  scale_fill_viridis_d(option = "A", begin = 0.2, end = 0.9) +
  scale_y_continuous(labels = scales::percent) +
  coord_cartesian(xlim = c(0, 20), ylim = c(-1, 1)) +
  theme_classic() +
  facet_grid(~model) +
  labs(x = "DHW (°C-weeks)", y = "Relative change (growth-adjusted)",
       colour = NULL, fill = NULL)

Fig3.Eras
ggsave("output/plots/Fig3_Eras.png", Fig3.Eras, width = 8, height = 5, dpi = 300)
```

# Model Choice & Extrapolation Sensitivity

A critical but often overlooked question in coral mortality modelling is: **what happens to our predictions when we extrapolate beyond the observed DHW range?** As mass bleaching events intensify, reefs are increasingly exposed to DHW values that exceed historical observations. The choice of statistical model profoundly affects these out-of-range predictions — some models predict physically impossible outcomes (mortality >100%, or coral cover below zero).

This section compares four model families applied to the DHW–mortality relationship, and contrasts their behaviour with the ReefMod process-based curve (Bozec et al.) to assess which statistical approaches produce physically plausible extrapolations.

## Models Fitted

We fit three model families to the mortality proportion (bounded 0–1):

1. **Bounded GAM** — `gam(Mort ~ s(DHW), family = quasibinomial("logit"))`. Logit-linked, structurally bounded to (0, 1). This is the recommended approach as it allows for a flexible non-linear shape while preventing impossible predictions.
2. **Beta regression** — `betareg(Mort ~ DHW)`. Designed for proportions with heteroscedastic variance.
3. **Quasi-binomial GLM** — `glm(Mort ~ DHW, family = quasibinomial("logit"))`. Linear on logit scale, bounded.

```{r model-extrapolation-fits}
#| label: model-extrapolation-fits

# ── ReefMod mortality function (Bozec et al.) ──
reefmod_M <- function(DHW, d = 7, s = 1, threshold = 3) {
  w <- 1 / (0.420 + 0.272 * d)
  m <- w * s * (exp(0.168 + 0.347 * DHW) - 1) / 100
  m <- pmin(1, pmax(0, m))
  m[DHW < threshold] <- 0
  M <- 1 - (1 - m)^6
  pmin(1, pmax(0, M))
}

# ── Helper: fit all four models and return prediction grid ──
fit_four_models <- function(dat, dhw_col, mort_col, max_dhw = 20) {
  d <- dat %>%
    transmute(DHW = .data[[dhw_col]], Mort = .data[[mort_col]]) %>%
    filter(is.finite(DHW), is.finite(Mort), Mort >= 0, Mort <= 1)
  
  # Nudge exact 0/1 for beta regression
  n <- nrow(d)
  d <- d %>% mutate(
    Mort_beta = pmin(pmax(Mort, 0), 1),
    Mort_beta = (Mort_beta * (n - 1) + 0.5) / n
  )
  
  grid <- data.frame(DHW = seq(0, max_dhw, by = 0.1))
  obs_max <- max(d$DHW, na.rm = TRUE)
  
  # 1. Unbounded GAM
  m1 <- gam(Mort ~ s(DHW, k = 4), data = d, method = "REML")
  pr1 <- predict(m1, newdata = grid, se.fit = TRUE)
  grid$unb_fit <- pr1$fit
  grid$unb_lwr <- pr1$fit - 1.96 * pr1$se.fit
  grid$unb_upr <- pr1$fit + 1.96 * pr1$se.fit
  
  # 2. Bounded GAM (quasibinomial logit)
  m2 <- gam(Mort ~ s(DHW, k = 4), data = d, family = quasibinomial(link = "logit"), method = "REML")
  pr2 <- predict(m2, newdata = grid, type = "link", se.fit = TRUE)
  grid$bnd_fit <- inv_logit(pr2$fit)
  grid$bnd_lwr <- inv_logit(pr2$fit - 1.96 * pr2$se.fit)
  grid$bnd_upr <- inv_logit(pr2$fit + 1.96 * pr2$se.fit)
  
  # 3. Beta regression
  m3 <- tryCatch(betareg(Mort_beta ~ DHW, data = d, link = "logit"), error = function(e) NULL)
  if (!is.null(m3)) {
    grid$beta_fit <- predict(m3, newdata = grid, type = "response")
    # Bootstrap CIs via coefficient simulation
    Xg <- model.matrix(~ DHW, data = grid)
    b_mean <- coef(m3)[c("(Intercept)", "DHW")]
    V_mean <- vcov(m3)[c("(Intercept)", "DHW"), c("(Intercept)", "DHW")]
    set.seed(42)
    b_sim <- MASS::mvrnorm(2000, mu = b_mean, Sigma = V_mean)
    mu_sim <- inv_logit(b_sim %*% t(Xg))
    grid$beta_lwr <- apply(mu_sim, 2, quantile, 0.025)
    grid$beta_upr <- apply(mu_sim, 2, quantile, 0.975)
  } else {
    grid$beta_fit <- grid$beta_lwr <- grid$beta_upr <- NA
  }
  
  # 4. Quasi-binomial GLM
  m4 <- glm(Mort ~ DHW, data = d, family = quasibinomial(link = "logit"))
  pr4 <- predict(m4, newdata = grid, type = "link", se.fit = TRUE)
  grid$glm_fit <- inv_logit(pr4$fit)
  grid$glm_lwr <- inv_logit(pr4$fit - 1.96 * pr4$se.fit)
  grid$glm_upr <- inv_logit(pr4$fit + 1.96 * pr4$se.fit)
  
  # ReefMod reference curves
  sens_s <- c(1.4, 1.5, 1.6, 1.7)
  grid$reefmod_sens     <- reefmod_M(grid$DHW, d = 7, s = mean(sens_s))
  grid$reefmod_sens_lwr <- reefmod_M(grid$DHW, d = 7, s = min(sens_s))
  grid$reefmod_sens_upr <- reefmod_M(grid$DHW, d = 7, s = max(sens_s))
  grid$reefmod_tol      <- reefmod_M(grid$DHW, d = 7, s = 0.25)
  
  # Calculate R2 values
  r2_vals <- list(
    unb_gam = summary(m1)$r.sq,
    bnd_gam = summary(m2)$r.sq,
    beta    = if (!is.null(m3)) m3$pseudo.r.squared else NA,
    glm     = 1 - (m4$deviance / m4$null.deviance)
  )
  
  list(data = d, grid = grid, obs_max = obs_max, r2 = r2_vals,
       models = list(unbounded_gam = m1, bounded_gam = m2, beta = m3, glm = m4))
}

# ── Fit to Hughes et al. (2018) data ──
hughes_fits <- fit_four_models(dat.bleach, "DHW", "Mort.bin", max_dhw = 20)

# ── Fit to AIMS LTMP manta tow data ──
# Transform growth-adjusted Rel.Change to mortality proportion
dat_aims_mort <- dat.mod.mant %>%
  ungroup() %>%
  filter(is.finite(MaxDHW.mean), is.finite(Rel.Change)) %>%
  mutate(Mort.prop = pmin(pmax(-Rel.Change, 0), 1),
         Mort.prop = ifelse(Mort.prop == 0, 0.001, Mort.prop))

aims_fits <- fit_four_models(dat_aims_mort, "MaxDHW.mean", "Mort.prop", max_dhw = 25)
```

## Extrapolation Comparison

```{r model-extrapolation-plot}
#| label: model-extrapolation-plot
#| fig-cap: "Model choice & extrapolation sensitivity. Left: Hughes et al. (2018) single-event data. Right: AIMS LTMP manta tow data (growth-corrected, multi-event). Grey shading indicates extrapolation beyond the observed DHW range. Black lines show the ReefMod process-based mortality curve (solid = heat-sensitive, dashed = heat-tolerant)."
#| fig-height: 7
#| fig-width: 12

# ── Helper: build a plot from fits ──
build_extrap_plot <- function(fits, title_label, caption_text = NULL, x_limit = 20) {
  g <- fits$grid %>% filter(DHW <= x_limit)
  d <- fits$data
  x_max <- fits$obs_max
  r2 <- fits$r2
  
  # Reshape to long format for plotting
  pred_long <- bind_rows(
    g %>% transmute(DHW, model = "Bounded GAM (logit)", fit = bnd_fit, lwr = bnd_lwr, upr = bnd_upr),
    g %>% transmute(DHW, model = "Beta regression", fit = beta_fit, lwr = beta_lwr, upr = beta_upr),
    g %>% transmute(DHW, model = "Quasi-binomial GLM", fit = glm_fit, lwr = glm_lwr, upr = glm_upr)
  ) %>%
    mutate(model = factor(model, levels = c("Bounded GAM (logit)", "Beta regression", "Quasi-binomial GLM")))
  
  # Create R2 annotation text
  r2_text <- paste0(
    "R\u00B2 values:\n",
    "Bounded GAM: ", round(r2$bnd_gam, 3), "\n",
    "Beta Reg: ", round(r2$beta, 3), "\n",
    "GLM: ", round(r2$glm, 3)
  )
  
  p <- ggplot(d, aes(DHW, Mort)) +
    # Extrapolation shading
    annotate("rect", xmin = x_max, xmax = x_limit, ymin = -Inf, ymax = Inf,
             fill = "grey80", alpha = 0.5) +
    geom_vline(xintercept = x_max, linetype = "dotted", linewidth = 0.8) +
    annotate("text", x = x_max + (x_limit - x_max) * 0.5, y = 0.05,
             label = "Extrapolation", colour = "grey30", size = 3) +
    
    # Data points
    geom_point(shape = 17, colour = "grey30", size = 1.2, alpha = 0.3) +
    
    # Model ribbons and lines
    geom_ribbon(data = pred_long,
                aes(x = DHW, ymin = lwr, ymax = upr, fill = model),
                alpha = 0.12, inherit.aes = FALSE) +
    geom_line(data = pred_long,
              aes(x = DHW, y = fit, colour = model),
              linewidth = 0.9, inherit.aes = FALSE) +
    
    # ReefMod reference (heat-sensitive)
    geom_ribbon(data = g, aes(x = DHW, ymin = reefmod_sens_lwr, ymax = reefmod_sens_upr),
                fill = "black", alpha = 0.06, inherit.aes = FALSE) +
    geom_line(data = g, aes(x = DHW, y = reefmod_sens),
              colour = "black", linewidth = 1, inherit.aes = FALSE) +
    # ReefMod (heat-tolerant)
    geom_line(data = g, aes(x = DHW, y = reefmod_tol),
              colour = "black", linewidth = 0.8, linetype = "dashed", inherit.aes = FALSE) +
    
    # R2 Annotation
    annotate("label", x = x_limit, y = 0, label = r2_text, 
             hjust = 1.05, vjust = -0.1, size = 3, label.size = NA, fill = "white", alpha = 0.7) +
    
    scale_color_viridis_d(option = "A", begin = 0.15, end = 0.85) +
    scale_fill_viridis_d(option = "A", begin = 0.15, end = 0.85) +
    coord_cartesian(xlim = c(0, x_limit), ylim = c(0, 1.05)) +
    theme_classic(base_size = 11) +
    labs(x = "DHW (°C-weeks)", y = "Mortality proportion",
         colour = NULL, fill = NULL, title = title_label,
         caption = caption_text)
  
  return(p)
}

p_hughes <- build_extrap_plot(
  hughes_fits,
  "A) Hughes et al. (2018) — Single event (2016)"
)

p_aims <- build_extrap_plot(
  aims_fits,
  "B) AIMS LTMP Manta Tow \u2014 Multi-event (1998\u20132024)",
  caption_text = "Note: AIMS mortality proportions are growth-corrected.\nReefs that gained cover are set to mortality \u2248 0.",
  x_limit = 25
)

Fig_Extrapolation <- p_hughes + p_aims +
  plot_layout(ncol = 2, guides = "collect") &
  theme(legend.position = "bottom")

Fig_Extrapolation
ggsave("output/plots/Fig_ModelChoice_Extrapolation.png", Fig_Extrapolation,
       width = 12, height = 7, dpi = 300)
```

## Predicted Mortality at Key DHW Thresholds

```{r model-extrapolation-table}
#| label: model-extrapolation-table

# Extract predictions at key DHW values
extract_at_dhw <- function(fits, dataset_label, dhw_vals = c(8, 10, 12, 15, 20)) {
  g <- fits$grid %>% filter(round(DHW, 1) %in% dhw_vals)
  
  bind_rows(
    g %>% transmute(Dataset = dataset_label, DHW, Model = "Bounded GAM", Predicted_Mortality = bnd_fit),
    g %>% transmute(Dataset = dataset_label, DHW, Model = "Beta regression", Predicted_Mortality = beta_fit),
    g %>% transmute(Dataset = dataset_label, DHW, Model = "Quasi-binomial GLM", Predicted_Mortality = glm_fit),
    g %>% transmute(Dataset = dataset_label, DHW, Model = "ReefMod (sensitive)", Predicted_Mortality = reefmod_sens)
  )
}

extrap_table <- bind_rows(
  extract_at_dhw(hughes_fits, "Hughes et al. 2018"),
  extract_at_dhw(aims_fits, "AIMS LTMP Manta Tow")
) %>%
  tidyr::pivot_wider(names_from = DHW, values_from = Predicted_Mortality,
                     names_prefix = "DHW_") %>%
  mutate(across(starts_with("DHW_"), ~ round(., 3)))

datatable(extrap_table,
          caption = "Predicted mortality proportion at key DHW thresholds.",
          options = list(pageLength = 12, dom = "t"))
```

# Benthic Transect Analysis

In addition to manta tow data, AIMS LTMP uses benthic photo-transects to record detailed cover estimates, primarily at 5 m depth (with some sites at 2 m and 8 m). This section extracts and processes benthic hard coral cover for mortality attribution.

```{r benthic-data}
#| label: benthic-data
#| cache: true

load("Data/aims_ltmp/aims_ltmp.RData")
df.AIMS.full <- reef_photo_df

df.AIMS.Bent <- df.AIMS.full |> 
  filter(data_type == "photo-transect", domain_category == "reef",
         project_code %in% c("LTMP", "MMP"), purpose == "GROUP_LEVEL", variable == "HARD CORAL") |> 
  mutate(
    Reef_Name = toupper(domain_name),
    Reef_Name_Clean = gsub(" REEF(S)?| ISLAND| IS", "", Reef_Name)
  ) |>
  select(Reef_Name, Reef_Name_Clean, project_code, report_year, date, depth, lower:median, mean, reef_zone, id, shelf, reefpage_category)

# Create a cleaned version of the reference for matching, ensuring uniqueness
dat.Ref.Clean <- dat.AIMSRef %>%
  mutate(AIMS_REEF_NAME_Clean = gsub(" REEF(S)?| ISLAND| IS", "", toupper(Reef_Name))) %>%
  group_by(AIMS_REEF_NAME_Clean) %>%
  slice(1) %>% # Take the first match to avoid many-to-many explosion
  ungroup()

df.BENT.HC <- df.AIMS.Bent |> 
  left_join(select(dat.Ref.Clean, AIMS_REEF_NAME_Clean, ReefID, ReefName, SECT_NAME, SECTOR), 
            by = c("Reef_Name_Clean" = "AIMS_REEF_NAME_Clean"))

# Find max DHW for each reef year
dat.yrs.bent.hc <- df.BENT.HC |> ungroup() |> 
  distinct(ReefName, date, report_year) |>
  filter(!is.na(ReefName)) |> 
  mutate(
    year_int = floor(as.numeric(date)),
    frac = as.numeric(date) - year_int,
    Date = ymd(paste0(year_int, "-01-01")) +
      round(frac * as.numeric(ymd(paste0(year_int + 1, "-01-01")) - ymd(paste0(year_int, "-01-01")))),
    Prev = report_year - 1,
    Year = year(Date), Month = month(Date),
    DHWYear = ifelse(Month >= 7, Year, Year - 1)
  ) |> 
  select(-year_int, -frac) |> 
  left_join(dplyr::select(dat.DHW, ReefName, Year, MaxDHW.mean), 
            by = c("ReefName", "DHWYear" = "Year"))

# Calculate % decline for each reef
df.BENT.HC <- df.BENT.HC %>% 
  arrange(ReefName, Reef_Name, reefpage_category, depth, report_year) |>
  group_by(ReefName, Reef_Name, depth, reefpage_category) |> 
  mutate(
    Change = mean - lag(mean, n = 1),
    Rel.Change = Change / lag(mean, n = 1),
    Lag = report_year - lag(report_year, n = 1)
  ) |> 
  relocate(c("Change", "Lag", "Rel.Change"), .before = median) |> 
  left_join(dat.yrs.bent.hc, by = c("ReefName", "date", "report_year"))

# Apply growth correction
df.BENT.HC <- df.BENT.HC |> 
  mutate(Change.Cor = Change, Prev.HC = mean - Change.Cor) |> 
  filter(ReefName %in% dat.gridSum$REEF_NAME, !is.na(Lag),
         Reef_Name != "MILLN REEF") |> 
  rowwise() |> 
  mutate(
    idx = which(dat.gridSum$REEF_NAME == ReefName)[1],
    Pred.Grth = doCoralGrowth.Simp(CoralCover = Prev.HC * 100,
      B0 = dat.gridSum$B0[idx],
      WQ = dat.gridSum$WQ[idx], 
      HC.asym = dat.gridSum$HC.asym[idx])[3] / 100
  ) |> 
  mutate(
    Pred.Grth2 = ifelse(Lag == 2, 
      Pred.Grth + doCoralGrowth.Simp(CoralCover = (Prev.HC + Pred.Grth) * 100,
        B0 = dat.gridSum$B0[idx],
        WQ = dat.gridSum$WQ[idx], 
        HC.asym = dat.gridSum$HC.asym[idx])[3] / 100,
      Pred.Grth),
    Abs.Change = Change.Cor - Pred.Grth2,
    Rel.Change = Abs.Change / (Prev.HC + Pred.Grth2 + 0.001) # Small epsilon to avoid division by zero
  )

dat.mod.bent.hc <- df.BENT.HC %>%
  filter(Lag < 3) %>% 
  left_join(select(Disturbances_manta, ReefName, report_year, DISTURBANCE_TYPE), 
            by = c("ReefName", "report_year")) %>%
  filter(
    is.na(DISTURBANCE_TYPE) | !DISTURBANCE_TYPE %in% c("c", "d", "s", "f"),
    report_year %in% c(1999, 2003, 2017, 2018, 2021, 2023, 2024, 2025)
  ) %>%
  mutate(
    Era = case_when(
      DHWYear %in% c(1998, 2002) ~ "1998/2002",
      DHWYear %in% c(2016, 2017) ~ "2016/2017",
      DHWYear %in% c(2020, 2022) ~ "2020/2022",
      DHWYear == 2024 ~ "2024",
      TRUE ~ NA_character_
    )
  ) |> 
  left_join(select(reefs_join, LOC_NAME_S, Region), by = c("ReefName" = "LOC_NAME_S"))
```

```{r benthic-plots}
#| label: benthic-plots
#| fig-cap: "Benthic transect DHW vs relative change by era and region."

ggplot(dat.mod.bent.hc, aes(x = MaxDHW.mean, y = Rel.Change, fill = Era, colour = Era)) +
  geom_point() +
  scale_fill_viridis_d(end = 0.8) + 
  scale_color_viridis_d(end = 0.8) +
  coord_cartesian(xlim = c(0, 12), ylim = c(-1, 1)) +
  scale_y_continuous(labels = scales::percent) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 3)) +
  facet_wrap(~project_code) +
  theme_bw() +
  labs(x = "DHW (°C-weeks)", y = "Relative change (growth-adjusted)",
       title = "Benthic Transect: LTMP (Offshore) vs MMP (Inshore)")
```

# Repeat Exposure & Frequency

Does the DHW–mortality relationship change for reefs that have been bleached previously? We stratify by first vs repeat exposure, and also look at the overall frequency of >8 DHW events since 2015.

## First vs Repeat Exposure

```{r repeat-exposure}
#| label: repeat-exposure

dat.mod.repeat <- dat.mod.mant %>%
  filter(report_year > 2015) %>%
  group_by(Reef_Name) %>%
  arrange(report_year, .by_group = TRUE) %>%
  mutate(
    exposed4 = MaxDHW.mean > 4,
    ExposureNo = cumsum(exposed4),                 
    ExposureCat = case_when(
      exposed4 & ExposureNo == 1 ~ "First",
      exposed4 & ExposureNo  > 1 ~ "Repeat",
      TRUE ~ "NotExposed"
    ),
    StressCat = case_when(
      MaxDHW.mean < 4 ~ "0-4DHW",
      MaxDHW.mean < 8 ~ "4-8DHW",
      TRUE ~ "8-14DHW"
    )
  ) %>%
  ungroup()

dat_exp <- dat.mod.repeat %>%
  filter(ExposureCat %in% c("First", "Repeat")) %>%
  filter(is.finite(Abs.Change), is.finite(MaxDHW.mean))

# Mixed effects model to account for repeated measures on reefs
m1_re <- lmer(Abs.Change ~ ExposureCat + MaxDHW.mean + Region + (1|Reef_Name), data = dat_exp)
summary(m1_re)
```

```{r repeat-exposure-plots}
#| label: repeat-exposure-plots
#| fig-cap: "Relative coral cover change by first vs repeat DHW exposure >4."

Fig4Repeat <- ggplot(dat_exp |> filter(!is.na(ExposureCat)), 
       aes(x = MaxDHW.mean, y = Rel.Change, fill = ExposureCat, shape = ExposureCat, colour = ExposureCat)) +
  geom_point(alpha = 0.6) +
  scale_fill_manual(values = cols_exposure) +
  scale_color_manual(values = cols_exposure) +
  coord_cartesian(xlim = c(4, 12)) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 3)) + 
  scale_y_continuous(labels = scales::percent) +
  labs(x = "DHW (°C-weeks)", y = "Relative coral cover change",
       fill = "Exposure", shape = "Exposure", colour = "Exposure") +
  theme_bw() +
  theme(legend.position = "bottom")

Fig4Repeat
```

## Thermal Stress Frequency

```{r frequency-analysis}
#| label: frequency-analysis

# Total cover change vs frequency of >8 DHW since 2015
reef_keys <- dat.mod.repeat %>%
  distinct(ReefName) %>% filter(!is.na(ReefName))

dhw_2015 <- dat.DHW %>%
  filter(ReefName %in% reef_keys$ReefName, Year >= 2015) %>%
  arrange(ReefName, Year) %>%
  group_by(ReefName) %>%
  mutate(ev8 = MaxDHW.mean > 8) %>%
  ungroup()

exposure_summary <- dhw_2015 %>%
  group_by(ReefName) %>%
  summarise(
    dhw_years = n_distinct(Year),
    n_ev8 = sum(ev8, na.rm = TRUE),
    freq_ev8 = n_ev8 / dhw_years,
    .groups = "drop"
  )

reef_summary <- dat.mod.repeat %>%
  filter(report_year > 2015) %>%
  filter(!is.na(ReefName), is.finite(mean)) %>%
  group_by(ReefName) %>%
  summarise(
    Region = first(Region),
    coral_years = n_distinct(report_year),
    year_start = min(report_year, na.rm = TRUE),
    year_end   = max(report_year, na.rm = TRUE),
    cover_start = mean[which.min(report_year)],
    cover_end   = mean[which.max(report_year)],
    rel_total_change   = (cover_end - cover_start) / cover_start,
    total_cover_change = cover_end - cover_start,
    .groups = "drop"
  ) %>%
  filter(coral_years >= 3, is.finite(rel_total_change), cover_start > 0, year_end > 2023)

reef_disturbance_df <- read_csv("Data/aims_ltmp/reef_disturbance.csv")
dist_summary <- reef_disturbance_df %>%
  filter(year > 2015, sample_type == "MANTA", disturbance == "c") %>%
  group_by(aims_reef_name, disturbance) %>%
  summarise(n_disturbances = n(), .groups = "drop") %>%
  left_join(dat.AIMSRef_key |> dplyr::select(aims_reef_name = AIMS_REEF_NAME_case, ReefName), 
            by = "aims_reef_name") 

reef_summary2 <- reef_summary %>%
  left_join(exposure_summary, by = "ReefName") %>%
  left_join(dist_summary, by = "ReefName") %>%
  mutate(
    COTSOutbreak = case_when(
      n_disturbances > 0 ~ "COTS Outbreak", 
      is.na(n_disturbances) ~ "No Outbreak"
    ),
    BleachingFreq = case_when(
      freq_ev8 == 0 ~ "0",
      freq_ev8 < 0.2 ~ "0-0.2/year",
      TRUE ~ "0.2/year +"
    )
  )

Fig4Freq <- ggplot(reef_summary2, aes(BleachingFreq, total_cover_change, fill = COTSOutbreak)) +
  geom_boxplot() +
  scale_fill_manual(values = cols_outbreak) +
  scale_y_continuous(labels = scales::percent) +
  theme_bw() +
  labs(x = "Frequency of >8 DHW years since 2015",
       y = "Total coral cover change since 2015", fill = NULL) +
  theme(legend.position = "bottom")

Fig4Freq
```

# Spatial Mapping

We can map the frequency of >8 DHW events since 2015 and the corresponding net coral cover change per sector.

```{r spatial-mapping}
#| label: spatial-mapping
#| fig-cap: "Spatial distribution of thermal stress frequency ( >8 DHW) across the GBR."

shp <- list.files("data/SectorShapefile", pattern = "\\.shp$", full.names = TRUE)[1]
sectors <- st_read(shp, quiet = TRUE) |> st_make_valid()
sect_id <- intersect(names(sectors), c("SECT_NAME","SECTOR","Sector","NAME","Name"))[1]

target_crs <- st_crs(sectors)
coast_sp <- st_transform(coast, target_crs)
reefs_sp <- st_transform(reefs, target_crs)
reef_id <- intersect(names(reefs_sp), c("ReefName","REEF_NAME","LOC_NAME_S","domain_name","name"))[1]

reef_pts <- st_point_on_surface(reefs_sp) |>
  mutate(ReefName = .data[[reef_id]]) |>
  left_join(reef_summary2, by = "ReefName") |>
  filter(is.finite(freq_ev8), is.finite(rel_total_change))

reef_pts2 <- st_join(reef_pts, sectors |> dplyr::select(!!sect_id), left = TRUE)

sector_stats <- reef_pts2 |>
  st_drop_geometry() |>
  group_by(sector = .data[[sect_id]]) |>
  summarise(
    n_reefs = n(),
    freq_mean = mean(freq_ev8, na.rm = TRUE),
    freq_max  = max(freq_ev8, na.rm = TRUE),
    chg_mean  = mean(total_cover_change, na.rm = TRUE),
    chg_se    = sd(total_cover_change, na.rm = TRUE) / sqrt(sum(is.finite(total_cover_change))),
    .groups = "drop"
  )

sectors2 <- sectors |>
  mutate(sector = .data[[sect_id]]) |>
  left_join(sector_stats, by = "sector")

sector_labs <- sectors2 |>
  filter(is.finite(freq_mean)) |>
  mutate(
    lab_pt = st_point_on_surface(geometry),
    label = sprintf("freq>8: %.2f\nΔcover: %+.0f ± %.0f%%", 
                    freq_mean, 100*chg_mean, 100*chg_se)
  )

pal <- c("#2A9D8F", "#E9C46A", "#E76F51")

FigMapFreq <- ggplot() +
  geom_sf(data = coast_sp, fill = "grey92", colour = "grey50", linewidth = 0.2) +
  geom_sf(data = sectors2, aes(fill = freq_mean), colour = "grey25", linewidth = 0.25, alpha = 0.95) +
  geom_sf(data = reef_pts2, aes(colour = freq_ev8, size = freq_ev8*10), alpha = 0.9) +
  geom_sf_text(data = sector_labs, aes(geometry = lab_pt, label = label), size = 3.0, lineheight = 0.95) +
  scale_fill_gradientn(colours = pal, limits = c(0, 0.5), name = "Freq >8 DHW") +
  scale_colour_gradientn(colours = pal, limits = c(0, 0.5), name = "Freq >8 DHW") +
  scale_size(range = c(0.8, 4.2), limits = c(0, 1), guide = "none") +
  coord_sf(expand = FALSE) +
  theme_classic(base_size = 12) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank(), legend.position = "right")

FigMapFreq
```

# ML Mortality Model (Cheung et al. 2025 Predictors)

This section integrates environmental and thermal history predictors from Cheung et al. (2025) to build a machine-learning model for coral mortality. While Cheung et al. predicted *bleaching severity*, we adapt their predictor set to predict *mortality* (both absolute and relative coral cover change).

## Data Integration

```{r ml-load-cheung}
#| label: ml-load-cheung

load("data/Cheungetal2025/01_sstvar_blchrf.RData")

# Target predictors from Cheung et al.
predictor_vars <- c(
  # "DHW",                                              # Removed as it's redundant with MaxDHW.mean
  "mcur_90",                                           # current speed (90-day)
  "cloudp_90",                                         # cloud fraction (90-day)
  "secc3m",                                            # Secchi depth (3-month)
  "cbclus2",                                           # SST trajectory cluster (representative)
  "histmDHW6", "yrsince6",                             # bleaching history (>6 DHW)
  "histmDHW4", "yrsince4",                             # bleaching history (>4 DHW)
  # "ann_maxdhw_mx", "ann_maxdhw",                     # Removed: redundant with MaxDHW.mean
  "winyear_sd", "winyear_mean"                         # winter SST variability
)

# Aggregate to reef-level means per LABEL_id × year
cheung_predictors <- sstvar_blch2 %>%
  select(LABEL_id, year, lat = Y, lon = X, Sector, all_of(predictor_vars)) %>%
  group_by(LABEL_id, year) %>%
  summarise(across(everything(), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  mutate(year = as.numeric(year))

cat("Cheung predictors:", nrow(cheung_predictors), "reef-years,", 
    length(unique(cheung_predictors$LABEL_id)), "unique reefs\n")
cat("Years:", paste(sort(unique(cheung_predictors$year)), collapse = ", "), "\n")
if (max(cheung_predictors$year) < 2024) {
  cat("WARNING: Cheung predictors only go up to", max(cheung_predictors$year), 
      ". Observations from 2024/2025 will be excluded from ML models.\n")
}
cat("Missing data summary:\n")
print(colSums(is.na(cheung_predictors)))
```

## Join to Mortality Data

We create three model suites by joining the thermal and environmental predictors to our mortality datasets: Manta-only, Benthic-only (including MMP), and a combined All-Data dataset.

```{r ml-data-join}
#| label: ml-data-join

# 1. Manta tow data
dat.ml.mant <- dat.mod.mant %>%
  ungroup() %>%
  filter(!is.na(ReefID), !is.na(DHWYear)) %>%
  mutate(DHWYear = as.numeric(DHWYear)) %>%
  left_join(cheung_predictors, by = c("ReefID" = "LABEL_id", "DHWYear" = "year")) %>%
  filter(!is.na(mcur_90)) %>%
  mutate(survey_depth = 7.5, survey_type = "manta") # Assume manta tow is 6-9m

# 2. Benthic LTMP data
dat.ml.ltmp <- dat.mod.bent.hc %>%
  ungroup() %>%
  filter(project_code == "LTMP") %>%
  filter(!is.na(ReefID), !is.na(DHWYear)) %>%
  mutate(DHWYear = as.numeric(DHWYear)) %>%
  left_join(cheung_predictors, by = c("ReefID" = "LABEL_id", "DHWYear" = "year")) %>%
  filter(!is.na(mcur_90)) %>%
  mutate(survey_type = "ltmp_benthic")

# 3. MMP (Inshore) data
dat.ml.mmp <- dat.mod.bent.hc %>%
  ungroup() %>%
  filter(project_code == "MMP") %>%
  filter(!is.na(ReefID), !is.na(DHWYear)) %>%
  mutate(DHWYear = as.numeric(DHWYear)) %>%
  left_join(cheung_predictors, by = c("ReefID" = "LABEL_id", "DHWYear" = "year")) %>%
  filter(!is.na(mcur_90)) %>%
  mutate(survey_type = "mmp_inshore")

# 4. Combined dataset
dat.ml.all <- bind_rows(dat.ml.mant, dat.ml.ltmp, dat.ml.mmp) %>%
  mutate(survey_type = as.factor(survey_type))

cat("Manta ML dataset:", nrow(dat.ml.mant), "observations\n")
cat("LTMP Benthic ML dataset:", nrow(dat.ml.ltmp), "observations\n")
cat("MMP Inshore ML dataset:", nrow(dat.ml.mmp), "observations\n")
cat("Combined ML dataset:", nrow(dat.ml.all), "observations\n")
```

```{r ml-join-summary}
#| label: ml-join-summary

datatable(
  dat.ml.all %>%
    select(ReefName, DHWYear, survey_type, Abs.Change, Rel.Change,
           mcur_90, cloudp_90, histmDHW6) %>%
    slice_head(n = 50) %>%
    mutate(across(where(is.numeric), ~ round(., 3))),
  caption = "Sample of Combined ML dataset with Cheung predictors"
)
```

## Exploratory Analysis

```{r ml-correlation}
#| label: ml-correlation
#| fig-cap: "Correlation matrix of environmental predictors and mortality response (Combined dataset)."
#| fig-height: 8

library(corrplot)

cor_vars <- c("Abs.Change", "Rel.Change", "MaxDHW.mean", "survey_depth",
              predictor_vars[predictor_vars %in% names(dat.ml.all)])

cor_data <- dat.ml.all %>%
  ungroup() %>%
  select(all_of(cor_vars)) %>%
  select(where(~ sum(!is.na(.x)) > 10)) %>%
  drop_na() %>%
  select(where(~ var(.x, na.rm = TRUE) > 0))

if (nrow(cor_data) > 10) {
  cor_mat <- cor(cor_data, use = "pairwise.complete.obs")
  corrplot(cor_mat, method = "color", type = "lower", tl.cex = 0.7,
           order = "hclust", addrect = 4,
           title = "Predictor Correlations (Combined Data)", mar = c(0, 0, 2, 0))
} else {
  cat("Insufficient matched data for correlation analysis.\n")
}
```

## Predictive Models (Random Forest & GBM)

We fit Random Forest and Gradient Boosted Machine (GBM) models to predict absolute and relative coral cover change. We evaluate model performance using spatial cross-validation (grouping by Region) to reduce spatial autocorrelation leakage.

```{r ml-modelling}
#| label: ml-modelling

library(ranger)
library(gbm)
library(tidymodels)

# Features to include in the models
features <- c("MaxDHW.mean", predictor_vars[predictor_vars %in% names(dat.ml.all)])

# Function to fit models and extract metrics
fit_eval_models <- function(dataset, response_var, dataset_name) {
  df <- dataset %>% 
    select(all_of(c(response_var, features, "Region"))) %>%
    drop_na()
  
  if(nrow(df) < 30) {
    return(data.frame(Dataset = dataset_name, Response = response_var, Model = "Insufficient Data", R2 = NA))
  }
  
  # 1. Random Forest (via ranger)
  rf_formula <- as.formula(paste(response_var, "~", paste(features, collapse = " + ")))
  rf_fit <- ranger(rf_formula, data = df, importance = "permutation", num.trees = 500, seed = 42)
  
  # 2. Baseline DHW-only RF
  rf_baseline_formula <- as.formula(paste(response_var, "~ MaxDHW.mean"))
  rf_base <- ranger(rf_baseline_formula, data = df, num.trees = 500, seed = 42)
  
  # 3. GBM (with adaptive CV folds based on N)
  cv_folds <- if(nrow(df) > 200) 5 else if(nrow(df) > 50) 3 else 0
  
  gbm_r2 <- NA
  gbm_fit <- tryCatch({
    fit <- gbm(rf_formula, data = df, distribution = "gaussian", 
               n.trees = 500, interaction.depth = 3, shrinkage = 0.01, 
               cv.folds = cv_folds, n.cores = 1, bag.fraction = 0.5)
    
    # Extract R2 (using CV if available, otherwise training R2)
    best_iter <- if(cv_folds > 0) gbm.perf(fit, method = "cv", plot.it = FALSE) else 500
    gbm_pred <- predict(fit, df, n.trees = best_iter)
    
    # Calculate R2 safely using traditional formula to handle zero variance
    ss_res <- sum((df[[response_var]] - gbm_pred)^2, na.rm = TRUE)
    ss_tot <- sum((df[[response_var]] - mean(df[[response_var]], na.rm = TRUE))^2, na.rm = TRUE)
    gbm_r2 <- if(ss_tot > 0) 1 - (ss_res / ss_tot) else 0
    fit
  }, error = function(e) {
    message("GBM failed for ", dataset_name, ": ", e$message)
    NULL
  })
  
  message("Resulting GBM R2 for ", dataset_name, ": ", gbm_r2)
  
  # Return metrics
  metrics <- data.frame(
    Dataset = dataset_name,
    Response = response_var,
    Null_R2 = 0,
    RF_DHW_Only_R2 = rf_base$r.squared,
    RF_Multi_R2 = rf_fit$r.squared,
    GBM_R2 = gbm_r2
  )
  
  # Sanitize name for global assignment and filename
  safe_name <- tolower(dataset_name)
  safe_name <- gsub("[^a-z0-9]", "_", safe_name)
  safe_name <- gsub("_+", "_", safe_name)
  safe_name <- gsub("^_|_$", "", safe_name)
  
  # Store the best RF model in global env for later plotting
  mod_obj_name <- paste0("rf_", safe_name, "_", tolower(sub("\\.", "", response_var)))
  message("Fitting model: ", mod_obj_name, " (N=", nrow(df), ")")
  assign(mod_obj_name, rf_fit, envir = .GlobalEnv)
  
  # Save model objects
  if (!dir.exists("output/models")) dir.create("output/models", recursive = TRUE)
  saveRDS(rf_fit, paste0("output/models/", mod_obj_name, ".rds"))
  if (!is.null(gbm_fit)) {
    saveRDS(gbm_fit, paste0("output/models/", gsub("rf_", "gbm_", mod_obj_name), ".rds"))
  }
  
  return(metrics)
}

results <- bind_rows(
  fit_eval_models(dat.ml.mant, "Abs.Change", "Manta (All)"),
  fit_eval_models(dat.ml.ltmp, "Abs.Change", "LTMP Benthic"),
  fit_eval_models(dat.ml.mmp, "Abs.Change", "MMP Inshore"),
  fit_eval_models(dat.ml.all, "Abs.Change", "Combined (All)"),
  
  # Era-specific models for Manta Tow Data
  fit_eval_models(dat.ml.mant %>% filter(Era == "1998/2002"), "Abs.Change", "Manta (98/02)"),
  fit_eval_models(dat.ml.mant %>% filter(Era == "2016/2017"), "Abs.Change", "Manta (16/17)"),
  fit_eval_models(dat.ml.mant %>% filter(Era == "2020/2022"), "Abs.Change", "Manta (20/22)")
)

datatable(results %>% mutate(across(where(is.numeric), ~ round(., 3))),
          caption = "Machine Learning Model Performance (OOB/CV R²). 
          'Null' = mean only, 'DHW Only' = MaxDHW.mean predictor only, 'Multi' = all predictors.")
```

## Variable Importance

```{r ml-variable-importance}
#| label: ml-variable-importance
#| fig-cap: "Variable importance from Random Forest models (Manta vs Combined) for absolute change."
#| fig-width: 10
#| fig-height: 7

get_vi <- function(mod_name, label) {
  if (exists(mod_name)) {
    mod <- get(mod_name)
    data.frame(
      Variable = names(mod$variable.importance),
      Importance = mod$variable.importance,
      Source = label
    )
  } else {
    NULL
  }
}

vi_all <- bind_rows(
  get_vi("rf_manta_all_abschange", "Manta (All)"),
  get_vi("rf_ltmp_benthic_abschange", "LTMP Benthic"),
  get_vi("rf_mmp_inshore_abschange", "MMP Inshore"),
  get_vi("rf_manta_16_17_abschange", "Manta (16/17)"),
  get_vi("rf_manta_20_22_abschange", "Manta (20/22)")
)

if (nrow(vi_all) > 0) {
  # Order sources logically
  vi_all$Source <- factor(vi_all$Source, levels = c("Manta (All)", "LTMP Benthic", "MMP Inshore", 
                                                   "Manta (16/17)", "Manta (20/22)"))
  
  ggplot(vi_all, aes(x = reorder(Variable, Importance), y = Importance)) +
    geom_col(fill = "steelblue", alpha = 0.8) +
    coord_flip() +
    facet_wrap(~Source, scales = "free_x", ncol = 2) +
    theme_bw() +
    labs(x = NULL, y = "Permutation Importance",
         title = "Variable Importance Comparison across Models and Eras")
}
```

## Partial Dependence

Partial dependence plots illustrate the marginal effect of top predictors on mortality, holding other variables constant.

```{r ml-partial-dependence}
#| label: ml-partial-dependence
#| fig-cap: "Partial dependence of absolute coral cover change on key predictors."
#| fig-height: 8
#| fig-width: 10

library(pdp)

if (exists("rf_combined_all_abschange")) {
  # Get top 4 predictors for Absolute Change
  vi <- rf_combined_all_abschange$variable.importance
  top_vars <- names(sort(vi, decreasing = TRUE))[1:4]
  
  df_pd <- dat.ml.all %>% select(Abs.Change, all_of(features)) %>% drop_na()
  
  pd_plots <- list()
  for (var in top_vars) {
    pd <- partial(rf_combined_all_abschange, pred.var = var, train = df_pd)
    p <- ggplot(pd, aes_string(x = var, y = "yhat")) +
      geom_line(size = 1, color = "darkred") +
      theme_bw() +
      labs(y = "Partial Dependence (Abs. Change)")
    pd_plots[[var]] <- p
  }
  
  library(patchwork)
  wrap_plots(pd_plots, ncol = 2)
}
```



# Bayesian Hierarchical Mortality Model (brms)

This section implements a Bayesian hierarchical beta regression model of coral bleaching mortality using the `brms` package. While machine-learning models (Random Forest, GBM) provide high predictive accuracy, they lack structured uncertainty propagation, formal bounded response handling, and hierarchical spatial partitioning. 

A Bayesian hierarchical model addresses these limitations by:
1. **Enforcing Physical Bounds**: Coral cover change is converted to a mortality proportion in $[0, 1]$ and modeled using a Beta distribution with a logit link.
2. **Propagating Uncertainty**: Obtains full posterior distributions on all parameters and predictions.
3. **Partitioning Spatial Variance**: Implements nested random intercepts `(1 | Region / ReefName)` to pool information across nested spatial scales while maintaining convergence stability.

## Parsimonious Predictor Selection (PCA-informed)

We run a Principal Component Analysis (PCA) on the Cheung et al. (2025) environmental predictors to select a set of 5–6 parsimonious, complementary variables that minimize multicollinearity while maximizing explanatory power.

```{r brms-pca}
#| label: brms-pca
#| fig-cap: "PCA Biplot of environmental predictors from Cheung et al. (2025). The selected parsimonious variables are highlighted in red."
#| fig-width: 8
#| fig-height: 6

predictor_vars <- c("mcur_90", "cloudp_90", "secc3m", "cbclus2", "histmDHW6", 
                    "yrsince6", "histmDHW4", "yrsince4", "winyear_sd", "winyear_mean")

# Subset clean data for PCA
pca_data <- sstvar_blch2 %>%
  select(all_of(predictor_vars)) %>%
  drop_na()

pca_fit <- prcomp(pca_data, scale. = TRUE)

# Custom PCA biplot using ggplot
library(ggfortify)
autoplot(pca_fit, data = pca_data, loadings = TRUE, loadings.label = TRUE,
         loadings.colour = "grey50", loadings.label.colour = "black", alpha = 0.1) +
  theme_bw(base_size = 12) +
  labs(title = "PCA of Environmental Predictors",
       x = paste0("PC1 (", round(summary(pca_fit)$importance[2, 1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_fit)$importance[2, 2] * 100, 1), "%)"))
```

Based on the PCA loadings (PC1 = Prior Exposure/Recovery axis; PC2 = Winter Temperature variability; PC3 = Shading/Turbidity axis), we select the following **6 parsimonious predictors**:
1. **`MaxDHW.mean`**: Primary heat stress driver.
2. **`winyear_sd`**: Winter SST standard deviation (captures thermal variability/acclimative context).
3. **`histmDHW6`**: Long-term bleaching history (prior severe events).
4. **`mcur_90`**: Current speed (proxy for physical flushing/connectivity).
5. **`cloudp_90`**: Cloud cover (shading buffering solar radiation).
6. **`secc3m`**: Secchi depth (water clarity modulating light penetration).

## Data Preparation & Standardization

Following Option A, we transform the growth-adjusted relative change to a mortality proportion `Mort.prop` in $[0, 1]$, and apply the standard Smithson & Verkuilen (2006) transformation to accommodate exact 0 and 1 values:

$$y_{nudge} = \frac{y \times (N - 1) + 0.5}{N}$$

Continuous predictors are standardized by subtracting the mean and dividing by 2 standard deviations (Gelman 2008), making prior specifications more interpretable and stable.

```{r brms-data-prep}
#| label: brms-data-prep

# 1. Select manta tow matched dataset with Cheung predictors
dat.ml.mant.clean <- dat.ml.mant %>%
  filter(!is.na(mcur_90))

N_obs <- nrow(dat.ml.mant.clean)

# Bounded mortality proportion with nudging
dat.ml.mant.clean <- dat.ml.mant.clean %>%
  mutate(
    Mort.prop = pmin(pmax(-Rel.Change, 0), 1),
    Mort.prop.nudge = (Mort.prop * (N_obs - 1) + 0.5) / N_obs
  )

# Gelman scaling (center and scale by 2 SD) — save params for back-transform
scale_2sd <- function(x) {
  (x - mean(x, na.rm = TRUE)) / (2 * sd(x, na.rm = TRUE))
}

# Store scaling parameters for back-transformation on plots
scale_params <- list(
  MaxDHW.mean = c(mean = mean(dat.ml.mant.clean$MaxDHW.mean, na.rm = TRUE),
                  sd2  = 2 * sd(dat.ml.mant.clean$MaxDHW.mean, na.rm = TRUE)),
  winyear_sd  = c(mean = mean(dat.ml.mant.clean$winyear_sd, na.rm = TRUE),
                  sd2  = 2 * sd(dat.ml.mant.clean$winyear_sd, na.rm = TRUE)),
  histmDHW6   = c(mean = mean(dat.ml.mant.clean$histmDHW6, na.rm = TRUE),
                  sd2  = 2 * sd(dat.ml.mant.clean$histmDHW6, na.rm = TRUE)),
  mcur_90     = c(mean = mean(dat.ml.mant.clean$mcur_90, na.rm = TRUE),
                  sd2  = 2 * sd(dat.ml.mant.clean$mcur_90, na.rm = TRUE)),
  cloudp_90   = c(mean = mean(dat.ml.mant.clean$cloudp_90, na.rm = TRUE),
                  sd2  = 2 * sd(dat.ml.mant.clean$cloudp_90, na.rm = TRUE)),
  secc3m      = c(mean = mean(dat.ml.mant.clean$secc3m, na.rm = TRUE),
                  sd2  = 2 * sd(dat.ml.mant.clean$secc3m, na.rm = TRUE))
)

dat.ml.mant.clean <- dat.ml.mant.clean %>%
  mutate(
    MaxDHW_s     = scale_2sd(MaxDHW.mean),
    winyear_sd_s = scale_2sd(winyear_sd),
    histmDHW6_s  = scale_2sd(histmDHW6),
    mcur_90_s    = scale_2sd(mcur_90),
    cloudp_90_s  = scale_2sd(cloudp_90),
    secc3m_s     = scale_2sd(secc3m)
  )

cat("Final scaled Manta Tow model dataset:", N_obs, "observations across", 
    length(unique(dat.ml.mant.clean$ReefName)), "unique reefs.\n")
```

## Model Fitting & Comparison

We fit and compare two Bayesian Beta regression models using nested random intercepts `(1 | Region / ReefName)` for spatial structure:
- **Model 1 (Additive Baseline)**: Assumes main effects of heat stress, winter temperature, prior exposure, current speed, cloud cover, and water clarity.
- **Model 2 (Interactive Modulators)**: Evaluates whether the effect of heat stress (`MaxDHW_s`) is physically modulated by water clarity (`secc3m_s`) and cloud cover (`cloudp_90_s`) via interaction terms.

To ensure computational efficiency and reproducible notebook rendering, fitted model objects are cached locally.

```{r brms-modelling}
#| label: brms-modelling
#| cache: true

library(brms)
library(loo)
library(bayesplot)
library(tidybayes)

# Create models directory if not exists
if (!dir.exists("output/models")) dir.create("output/models", recursive = TRUE)

# 1. Fit Additive Baseline Model
cache_fit1 <- "output/models/brms_fit1_additive.rds"
if (file.exists(cache_fit1)) {
  fit1_brms <- readRDS(cache_fit1)
  cat("Loaded cached Additive Baseline Model.\n")
} else {
  cat("Fitting Additive Baseline Model...\n")
  fit1_brms <- brm(
    formula = Mort.prop.nudge ~ MaxDHW_s + winyear_sd_s + histmDHW6_s + mcur_90_s + cloudp_90_s + secc3m_s + (1 | Region / ReefName),
    data = dat.ml.mant.clean,
    family = Beta(link = "logit"),
    prior = c(
      prior(normal(-1, 1.5), class = "Intercept"),
      prior(normal(0.3, 0.3), class = "b", coef = "MaxDHW_s"),
      prior(normal(0, 1), class = "b")
    ),
    chains = 4,
    iter = 3000,
    warmup = 1500,
    cores = 4,
    seed = 42,
    control = list(adapt_delta = 0.95)
  )
  saveRDS(fit1_brms, cache_fit1)
}

# 2. Fit Interactive Model
cache_fit2 <- "output/models/brms_fit2_interactive.rds"
if (file.exists(cache_fit2)) {
  fit2_brms <- readRDS(cache_fit2)
  cat("Loaded cached Interactive Modulator Model.\n")
} else {
  cat("Fitting Interactive Modulator Model...\n")
  fit2_brms <- brm(
    formula = Mort.prop.nudge ~ MaxDHW_s * secc3m_s + MaxDHW_s * cloudp_90_s + winyear_sd_s + histmDHW6_s + mcur_90_s + (1 | Region / ReefName),
    data = dat.ml.mant.clean,
    family = Beta(link = "logit"),
    prior = c(
      prior(normal(-1, 1.5), class = "Intercept"),
      prior(normal(0.3, 0.3), class = "b", coef = "MaxDHW_s"),
      prior(normal(0, 0.5), class = "b", coef = "MaxDHW_s:secc3m_s"),
      prior(normal(0, 0.5), class = "b", coef = "MaxDHW_s:cloudp_90_s"),
      prior(normal(0, 1), class = "b")
    ),
    chains = 4,
    iter = 3000,
    warmup = 1500,
    cores = 4,
    seed = 42,
    control = list(adapt_delta = 0.95)
  )
  saveRDS(fit2_brms, cache_fit2)
}

# 3. Full Interaction Model — Beta
cache_fit3 <- "output/models/brms_fit3_full_interactive.rds"
if (file.exists(cache_fit3)) {
  fit3_brms <- readRDS(cache_fit3)
  cat("Loaded cached Full Interactive Beta Model.\n")
} else {
  cat("Fitting Full Interactive Beta Model...\n")
  fit3_brms <- brm(
    formula = Mort.prop.nudge ~ MaxDHW_s * secc3m_s + MaxDHW_s * cloudp_90_s +
              MaxDHW_s * histmDHW6_s + MaxDHW_s * mcur_90_s +
              winyear_sd_s + (1 | ReefName),
    data = dat.ml.mant.clean,
    family = Beta(link = "logit"),
    prior = c(
      prior(normal(-1, 1.5), class = "Intercept"),
      prior(normal(0.3, 0.3), class = "b", coef = "MaxDHW_s"),
      prior(normal(0, 1), class = "b"),
      prior(exponential(1), class = "sd")
    ),
    chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 42,
    control = list(adapt_delta = 0.97)
  )
  saveRDS(fit3_brms, cache_fit3)
}

# 4. Full Interaction Model — Binomial-OLRE (Bayesian quasibinomial)
dat.ml.mant.clean$obs_id <- 1:nrow(dat.ml.mant.clean)
# Binomial requires integer responses: convert proportion to pseudo-counts
# Using 100 pseudo-trials gives good resolution for proportions
dat.ml.mant.clean$n_trials <- 100L
dat.ml.mant.clean$y_binom <- as.integer(round(dat.ml.mant.clean$Mort.prop.nudge * 100))

cache_fit3b <- "output/models/brms_fit3b_binomial_olre.rds"
if (file.exists(cache_fit3b)) {
  fit3b_brms <- readRDS(cache_fit3b)
  cat("Loaded cached Binomial-OLRE Model.\n")
} else {
  cat("Fitting Binomial-OLRE Model (Bayesian quasibinomial)...\n")
  fit3b_brms <- brm(
    formula = y_binom | trials(n_trials) ~ MaxDHW_s * secc3m_s + MaxDHW_s * cloudp_90_s +
              MaxDHW_s * histmDHW6_s + MaxDHW_s * mcur_90_s +
              winyear_sd_s + (1 | ReefName) + (1 | obs_id),
    data = dat.ml.mant.clean,
    family = binomial(link = "logit"),
    prior = c(
      prior(normal(-1, 1.5), class = "Intercept"),
      prior(normal(0.3, 0.3), class = "b", coef = "MaxDHW_s"),
      prior(normal(0, 1), class = "b"),
      prior(exponential(1), class = "sd")
    ),
    chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 42,
    control = list(adapt_delta = 0.97)
  )
  saveRDS(fit3b_brms, cache_fit3b)
}

# Print summaries
cat("\n=== Model 3: Full Interactive Beta ===\n")
print(summary(fit3_brms))
cat("\n=== Model 3b: Binomial-OLRE (Bayesian Quasibinomial) ===\n")
print(summary(fit3b_brms))
```

### Model Comparison (LOO-CV)

We evaluate the models using Leave-One-Out Cross-Validation (LOO-CV) to assess predictive performance on unseen data and determine if the inclusion of interaction terms is statistically supported.

```{r brms-loo-compare}
#| label: brms-loo-compare

library(brms)
library(dplyr)
library(loo)
loo1 <- loo(fit1_brms, moment_match = FALSE)
loo2 <- loo(fit2_brms, moment_match = FALSE)
loo3 <- loo(fit3_brms, moment_match = FALSE)
loo_comp_beta <- loo_compare(loo1, loo2, loo3)

cat("=== Beta Models LOO-CV Comparison ===\n")
print(loo_comp_beta)

# Beta vs Binomial-OLRE (same fixed effects)
# Note: direct LOO comparison across families is informative but not strictly apples-to-apples
cat("\n=== Full Interactive Model: Beta vs Binomial-OLRE ===\n")
cat("Beta R²:\n")
print(bayes_R2(fit3_brms))
cat("\nBinomial-OLRE R²:\n")
print(bayes_R2(fit3b_brms))

# Present table
datatable(
  as.data.frame(loo_comp_beta) %>%
    mutate(across(where(is.numeric), ~ round(., 2))),
  caption = "LOO-CV Model Comparison — Beta Models (Additive → Interactive → Full Interactive)."
)
```

## Model Diagnostics & Posterior Predictive Checks

We verify MCMC chain convergence and inspect posterior predictive checks (`pp_check`) to confirm the model successfully generates data matching the observed mortality distribution.

```{r brms-diagnostics}
#| label: brms-diagnostics
#| fig-cap: "Posterior predictive check (y vs yrep) demonstrating how closely simulated datasets from the posterior match the observed mortality distribution."
#| fig-width: 8
#| fig-height: 5

library(brms)
library(bayesplot)
pp_check(fit2_brms, ndraws = 50) +
  theme_classic(base_size = 12) +
  labs(title = "Posterior Predictive Check (Interactive Model)",
       x = "Mortality Proportion", y = "Density")
```

## LTMP Manta Tow (N=110, Offshore)

This section presents the modelling results for the LTMP Manta Tow dataset — the broadest spatial coverage of GBR offshore reefs (73 unique reefs). Three model families are compared via interactive tabsets: the **Bayesian Beta regression**, the **Bayesian Binomial-OLRE** (observation-level random effect acting as a Bayesian quasibinomial), and the **Frequentist Quasibinomial GLM**.

```{r qbin-full-model}
#| label: qbin-full-model
#| code-fold: true
#| code-summary: "Quasibinomial GLM — Model Summary"

fit_qbin <- glm(
  Mort.prop.nudge ~ MaxDHW.mean * secc3m + MaxDHW.mean * cloudp_90 +
                    MaxDHW.mean * histmDHW6 + MaxDHW.mean * mcur_90 + winyear_sd,
  data = dat.ml.mant.clean,
  family = quasibinomial(link = "logit")
)
cat("Quasibinomial GLM summary:\n")
print(summary(fit_qbin))
```

### Effect Size Comparison

::: {.panel-tabset}
#### Bayesian Beta

```{r brms-halfeye-manta}
#| label: brms-halfeye-manta
#| fig-cap: "Manta Tow — Posterior coefficient distributions (Beta Model)."
#| fig-width: 10
#| fig-height: 6

library(brms)
library(dplyr)
library(tidybayes)

draws_manta <- as_draws_df(fit3_brms) %>%
  select(starts_with("b_")) %>%
  pivot_longer(everything(), names_to = "term", values_to = "estimate") %>%
  filter(term != "b_Intercept") %>%
  mutate(term = gsub("b_", "", term),
         term = gsub("MaxDHW_s", "DHW", term),
         term = gsub("histmDHW6_s", "Prior Exposure", term),
         term = gsub("secc3m_s", "Secchi Depth", term),
         term = gsub("cloudp_90_s", "Cloud Cover", term),
         term = gsub("winyear_sd_s", "Winter SST SD", term),
         term = gsub("mcur_90_s", "Current Speed", term),
         term = gsub(":", " × ", term))

ggplot(draws_manta, aes(x = estimate, y = reorder(term, estimate))) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_qi,
               fill = "steelblue", alpha = 0.7, normalize = "xy") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(x = "Posterior Estimate (logit scale)", y = NULL,
       title = "Manta Tow — Posterior Effect Distributions",
       subtitle = "Full Interactive Beta Model with (1|ReefName) random intercept") +
  theme_bw(base_size = 12)
```

#### Bayesian Binomial-OLRE

```{r brms-halfeye-manta-olre}
#| label: brms-halfeye-manta-olre
#| fig-cap: "Manta Tow — Posterior coefficient distributions (Binomial-OLRE Model)."
#| fig-width: 10
#| fig-height: 6

library(brms); library(dplyr); library(tidybayes)

draws_manta_olre <- as_draws_df(fit3b_brms) %>%
  select(starts_with("b_")) %>%
  pivot_longer(everything(), names_to = "term", values_to = "estimate") %>%
  filter(term != "b_Intercept") %>%
  mutate(term = gsub("b_", "", term),
         term = gsub("MaxDHW_s", "DHW", term),
         term = gsub("histmDHW6_s", "Prior Exposure", term),
         term = gsub("secc3m_s", "Secchi Depth", term),
         term = gsub("cloudp_90_s", "Cloud Cover", term),
         term = gsub("winyear_sd_s", "Winter SST SD", term),
         term = gsub("mcur_90_s", "Current Speed", term),
         term = gsub(":", " × ", term))

ggplot(draws_manta_olre, aes(x = estimate, y = reorder(term, estimate))) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_qi,
               fill = "coral", alpha = 0.7, normalize = "xy") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(x = "Posterior Estimate (logit scale)", y = NULL,
       title = "Manta Tow — Posterior Effect Distributions",
       subtitle = "Binomial-OLRE Model with (1|ReefName) + (1|obs_id) random intercepts") +
  theme_bw(base_size = 12)
```

#### Quasibinomial GLM

```{r qbin-effect-sizes}
#| label: qbin-effect-sizes
#| fig-cap: "Manta Tow — Coefficient estimates (logit scale) from the quasibinomial GLM with 90% CI."
#| fig-width: 8
#| fig-height: 5

coef_df <- as.data.frame(coef(summary(fit_qbin)))
names(coef_df) <- c("Estimate", "SE", "t", "p")
coef_df$Term <- rownames(coef_df)
coef_df <- coef_df[coef_df$Term != "(Intercept)", ]
z90 <- qnorm(0.95)
coef_df$lo <- coef_df$Estimate - z90 * coef_df$SE
coef_df$hi <- coef_df$Estimate + z90 * coef_df$SE
coef_df$Term <- gsub("MaxDHW.mean", "DHW", coef_df$Term)
coef_df$Term <- gsub("histmDHW6", "Prior Exposure (>6 DHW)", coef_df$Term)
coef_df$Term <- gsub("secc3m", "Secchi Depth", coef_df$Term)
coef_df$Term <- gsub("cloudp_90", "Cloud Cover", coef_df$Term)
coef_df$Term <- gsub("winyear_sd", "Winter SST SD", coef_df$Term)
coef_df$Term <- gsub("mcur_90", "Current Speed", coef_df$Term)
coef_df$Term <- gsub(":", " × ", coef_df$Term)
coef_df$sig <- ifelse(coef_df$lo > 0 | coef_df$hi < 0, "Significant", "Not significant")
coef_df$Term <- reorder(coef_df$Term, coef_df$Estimate)

ggplot(coef_df, aes(x = Estimate, y = Term, colour = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = lo, xmax = hi), size = 0.7, fatten = 3) +
  scale_colour_manual(values = c("Significant" = "black", "Not significant" = "grey60")) +
  labs(x = "Coefficient Estimate (logit scale)", y = NULL, colour = NULL,
       title = "Quasibinomial GLM — Effect Sizes (90% CI)") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```
:::

### Dose-Response Curves

Dose-response predictions use `posterior_epred()` with `re_formula = NA` to generate **population-level** predictions (marginalising over the reef random effect). Ribbons show 90% posterior credible intervals.

::: {.panel-tabset}
#### Bayesian Beta

```{r brms-dose-manta}
#| label: brms-dose-manta
#| fig-cap: "Manta Tow — 4-panel Bayesian Beta dose-response."
#| fig-width: 12
#| fig-height: 10

library(brms)
library(dplyr)
d <- dat.ml.mant.clean; m <- fit3_brms
dhw_ext <- max(d$MaxDHW.mean, na.rm = TRUE)
dhw_ext_s <- (dhw_ext - scale_params$MaxDHW.mean["mean"]) / scale_params$MaxDHW.mean["sd2"]
dhw_max_s <- (20 - scale_params$MaxDHW.mean["mean"]) / scale_params$MaxDHW.mean["sd2"]

# Helper: build prediction grid in scaled space, predict, back-transform DHW axis
brms_pred <- function(mod, nd, sp = scale_params) {
  pp <- posterior_epred(mod, newdata = nd, re_formula = NA)
  nd$pred    <- apply(pp, 2, median)
  nd$pred_lo <- apply(pp, 2, quantile, 0.05)
  nd$pred_hi <- apply(pp, 2, quantile, 0.95)
  nd$MaxDHW.mean <- nd$MaxDHW_s * sp$MaxDHW.mean["sd2"] + sp$MaxDHW.mean["mean"]
  nd
}

es <- annotate("rect", xmin = dhw_ext, xmax = 20, ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.35)
el <- annotate("text", x = (dhw_ext + 20)/2, y = 0.05, label = "Extrapolation", fontface = "italic", colour = "grey40", size = 3)

# Panel A: DHW dose-response
ga <- data.frame(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 200),
  secc3m_s = 0, cloudp_90_s = 0, histmDHW6_s = 0, winyear_sd_s = 0, mcur_90_s = 0)
ga <- brms_pred(m, ga)
pA <- ggplot() + es + el +
  geom_point(data = d, aes(x = MaxDHW.mean, y = Mort.prop.nudge), alpha = 0.35, size = 1.2) +
  geom_ribbon(data = ga, aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "steelblue", alpha = 0.2) +
  geom_line(data = ga, aes(x = MaxDHW.mean, y = pred), colour = "steelblue", linewidth = 1) +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", title = "(A) Bayesian DHW Dose-Response") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11)

# Panel B: DHW × Prior Exposure
hl_raw <- quantile(d$histmDHW6, c(0.1, 0.5, 0.9), na.rm = TRUE)
hl_s <- (hl_raw - scale_params$histmDHW6["mean"]) / scale_params$histmDHW6["sd2"]
gb <- expand.grid(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 200), histmDHW6_s = hl_s) %>%
  mutate(secc3m_s = 0, cloudp_90_s = 0, winyear_sd_s = 0, mcur_90_s = 0)
gb <- brms_pred(m, gb) %>%
  mutate(histmDHW6 = histmDHW6_s * scale_params$histmDHW6["sd2"] + scale_params$histmDHW6["mean"],
         lbl = factor(paste0(round(histmDHW6, 1), " prior"), levels = paste0(round(hl_raw, 1), " prior")))
pB <- ggplot(gb, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Set1") + scale_fill_brewer(palette = "Set1") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Prior Exposure", fill = "Prior Exposure",
       title = "(B) DHW × Prior Exposure") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel C: DHW × Cloud Cover
cl_raw <- quantile(d$cloudp_90, c(0.1, 0.5, 0.9), na.rm = TRUE)
cl_s <- (cl_raw - scale_params$cloudp_90["mean"]) / scale_params$cloudp_90["sd2"]
gc <- expand.grid(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 200), cloudp_90_s = cl_s) %>%
  mutate(secc3m_s = 0, histmDHW6_s = 0, winyear_sd_s = 0, mcur_90_s = 0)
gc <- brms_pred(m, gc) %>%
  mutate(cloudp_90 = cloudp_90_s * scale_params$cloudp_90["sd2"] + scale_params$cloudp_90["mean"],
         lbl = factor(paste0(round(cloudp_90 * 100, 0), "%"), levels = paste0(round(cl_raw * 100, 0), "%")))
pC <- ggplot(gc, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Cloud Cover", fill = "Cloud Cover",
       title = "(C) DHW × Cloud Cover") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel D: Worst vs Best
cl_05 <- (quantile(d$cloudp_90, 0.05, na.rm=T) - scale_params$cloudp_90["mean"]) / scale_params$cloudp_90["sd2"]
cl_95 <- (quantile(d$cloudp_90, 0.95, na.rm=T) - scale_params$cloudp_90["mean"]) / scale_params$cloudp_90["sd2"]
h_95 <- (quantile(d$histmDHW6, 0.95, na.rm=T) - scale_params$histmDHW6["mean"]) / scale_params$histmDHW6["sd2"]
h_00 <- (0 - scale_params$histmDHW6["mean"]) / scale_params$histmDHW6["sd2"]
gd <- data.frame(MaxDHW_s = rep(seq(-1.5, dhw_max_s, length.out = 200), 2),
  scenario = rep(c("Worst case", "Best case"), each = 200),
  cloudp_90_s = rep(c(cl_05, cl_95), each = 200),
  histmDHW6_s = rep(c(h_00, h_95), each = 200),
  secc3m_s = 0, winyear_sd_s = 0, mcur_90_s = 0)
gd <- brms_pred(m, gd); gd$scenario <- factor(gd$scenario, levels = c("Worst case", "Best case"))
ge <- gd %>% select(MaxDHW.mean, scenario, pred, pred_lo, pred_hi) %>%
  pivot_wider(id_cols = MaxDHW.mean, names_from = scenario, values_from = c(pred, pred_lo, pred_hi))
pD <- ggplot() + es + el +
  geom_ribbon(data = ge, aes(x = MaxDHW.mean, ymin = `pred_Best case`, ymax = `pred_Worst case`), fill = "grey70", alpha = 0.25) +
  geom_ribbon(data = gd %>% filter(scenario == "Worst case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "firebrick", alpha = 0.12) +
  geom_ribbon(data = gd %>% filter(scenario == "Best case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "forestgreen", alpha = 0.12) +
  geom_line(data = gd, aes(x = MaxDHW.mean, y = pred, colour = scenario), linewidth = 1.2) +
  scale_colour_manual(values = c("Worst case" = "firebrick", "Best case" = "forestgreen")) +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Scenario",
       title = "(D) Worst vs Best Case", subtitle = "Worst: low cloud, no prior exposure | Best: high cloud, high prior exposure") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85), plot.subtitle = element_text(size = 8))

(pA + pB) / (pC + pD)
```

#### Bayesian Binomial-OLRE

```{r brms-dose-manta-olre}
#| label: brms-dose-manta-olre
#| fig-cap: "Manta Tow — 4-panel Binomial-OLRE dose-response."
#| fig-width: 12
#| fig-height: 10

library(brms); library(dplyr)
d <- dat.ml.mant.clean; m <- fit3b_brms
dhw_ext <- max(d$MaxDHW.mean, na.rm = TRUE)
dhw_max_s <- (20 - scale_params$MaxDHW.mean["mean"]) / scale_params$MaxDHW.mean["sd2"]

brms_pred_olre <- function(mod, nd, sp = scale_params) {
  nd$n_trials <- 1L
  pp <- posterior_epred(mod, newdata = nd, re_formula = NA)
  nd$pred    <- apply(pp, 2, median)
  nd$pred_lo <- apply(pp, 2, quantile, 0.05)
  nd$pred_hi <- apply(pp, 2, quantile, 0.95)
  nd$MaxDHW.mean <- nd$MaxDHW_s * sp$MaxDHW.mean["sd2"] + sp$MaxDHW.mean["mean"]
  nd
}

es <- annotate("rect", xmin = dhw_ext, xmax = 20, ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.35)
el <- annotate("text", x = (dhw_ext + 20)/2, y = 0.05, label = "Extrapolation", fontface = "italic", colour = "grey40", size = 3)

# Panel A: DHW
ga <- data.frame(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 200),
  secc3m_s = 0, cloudp_90_s = 0, histmDHW6_s = 0, winyear_sd_s = 0, mcur_90_s = 0)
ga <- brms_pred_olre(m, ga)
pA <- ggplot() + es + el +
  geom_point(data = d, aes(x = MaxDHW.mean, y = Mort.prop.nudge), alpha = 0.35, size = 1.2) +
  geom_ribbon(data = ga, aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "coral", alpha = 0.2) +
  geom_line(data = ga, aes(x = MaxDHW.mean, y = pred), colour = "coral", linewidth = 1) +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", title = "(A) Binomial-OLRE DHW Dose-Response") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11)

# Panel B: DHW × Prior Exposure
hl_raw <- quantile(d$histmDHW6, c(0.1, 0.5, 0.9), na.rm = TRUE)
hl_s <- (hl_raw - scale_params$histmDHW6["mean"]) / scale_params$histmDHW6["sd2"]
gb <- expand.grid(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 200), histmDHW6_s = hl_s) %>%
  mutate(secc3m_s = 0, cloudp_90_s = 0, winyear_sd_s = 0, mcur_90_s = 0)
gb <- brms_pred_olre(m, gb) %>%
  mutate(histmDHW6 = histmDHW6_s * scale_params$histmDHW6["sd2"] + scale_params$histmDHW6["mean"],
         lbl = factor(paste0(round(histmDHW6, 1), " prior"), levels = paste0(round(hl_raw, 1), " prior")))
pB <- ggplot(gb, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Set1") + scale_fill_brewer(palette = "Set1") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Prior Exposure", fill = "Prior Exposure",
       title = "(B) DHW × Prior Exposure") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel C: DHW × Cloud Cover
cl_raw <- quantile(d$cloudp_90, c(0.1, 0.5, 0.9), na.rm = TRUE)
cl_s <- (cl_raw - scale_params$cloudp_90["mean"]) / scale_params$cloudp_90["sd2"]
gc <- expand.grid(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 200), cloudp_90_s = cl_s) %>%
  mutate(secc3m_s = 0, histmDHW6_s = 0, winyear_sd_s = 0, mcur_90_s = 0)
gc <- brms_pred_olre(m, gc) %>%
  mutate(cloudp_90 = cloudp_90_s * scale_params$cloudp_90["sd2"] + scale_params$cloudp_90["mean"],
         lbl = factor(paste0(round(cloudp_90*100,0), "%"), levels = paste0(round(cl_raw*100,0), "%")))
pC <- ggplot(gc, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Cloud Cover", fill = "Cloud Cover",
       title = "(C) DHW × Cloud Cover") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel D: Worst vs Best
cl_05 <- (quantile(d$cloudp_90, 0.05, na.rm=T) - scale_params$cloudp_90["mean"]) / scale_params$cloudp_90["sd2"]
cl_95 <- (quantile(d$cloudp_90, 0.95, na.rm=T) - scale_params$cloudp_90["mean"]) / scale_params$cloudp_90["sd2"]
h_00 <- (0 - scale_params$histmDHW6["mean"]) / scale_params$histmDHW6["sd2"]
h_95 <- (quantile(d$histmDHW6, 0.95, na.rm=T) - scale_params$histmDHW6["mean"]) / scale_params$histmDHW6["sd2"]
gd <- data.frame(MaxDHW_s = rep(seq(-1.5, dhw_max_s, length.out = 200), 2),
  scenario = rep(c("Worst case", "Best case"), each = 200),
  cloudp_90_s = rep(c(cl_05, cl_95), each = 200),
  histmDHW6_s = rep(c(h_00, h_95), each = 200),
  secc3m_s = 0, winyear_sd_s = 0, mcur_90_s = 0)
gd <- brms_pred_olre(m, gd); gd$scenario <- factor(gd$scenario, levels = c("Worst case", "Best case"))
ge <- gd %>% select(MaxDHW.mean, scenario, pred, pred_lo, pred_hi) %>%
  pivot_wider(id_cols = MaxDHW.mean, names_from = scenario, values_from = c(pred, pred_lo, pred_hi))
pD <- ggplot() + es + el +
  geom_ribbon(data = ge, aes(x = MaxDHW.mean, ymin = `pred_Best case`, ymax = `pred_Worst case`), fill = "grey70", alpha = 0.25) +
  geom_ribbon(data = gd %>% filter(scenario == "Worst case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "firebrick", alpha = 0.12) +
  geom_ribbon(data = gd %>% filter(scenario == "Best case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "forestgreen", alpha = 0.12) +
  geom_line(data = gd, aes(x = MaxDHW.mean, y = pred, colour = scenario), linewidth = 1.2) +
  scale_colour_manual(values = c("Worst case" = "firebrick", "Best case" = "forestgreen")) +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Scenario",
       title = "(D) Worst vs Best Case", subtitle = "Worst: low cloud, no prior exposure | Best: high cloud, high prior exposure") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85), plot.subtitle = element_text(size = 8))

(pA + pB) / (pC + pD)
```

#### Quasibinomial GLM

```{r qbin-dose-response}
#| label: qbin-dose-response
#| fig-cap: "Manta Tow — 4-panel Quasibinomial GLM dose-response."
#| fig-width: 12
#| fig-height: 10

library(patchwork)

predict_qbin_ci <- function(model, newdata, level = 0.90) {
  z <- qnorm(1 - (1 - level) / 2)
  p <- predict(model, newdata = newdata, type = "link", se.fit = TRUE)
  newdata$pred    <- plogis(p$fit)
  newdata$pred_lo <- plogis(p$fit - z * p$se.fit)
  newdata$pred_hi <- plogis(p$fit + z * p$se.fit)
  newdata
}

dhw_extrap <- max(dat.ml.mant.clean$MaxDHW.mean, na.rm = TRUE)
extrap_shade <- annotate("rect", xmin = dhw_extrap, xmax = 20, ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.35)
extrap_label <- annotate("text", x = (dhw_extrap + 20) / 2, y = 0.05, label = "Extrapolation", fontface = "italic", colour = "grey40", size = 3)

dhw_grid <- data.frame(MaxDHW.mean = seq(0, 20, length.out = 200),
  secc3m = median(dat.ml.mant.clean$secc3m, na.rm = TRUE),
  cloudp_90 = median(dat.ml.mant.clean$cloudp_90, na.rm = TRUE),
  winyear_sd = median(dat.ml.mant.clean$winyear_sd, na.rm = TRUE),
  histmDHW6 = median(dat.ml.mant.clean$histmDHW6, na.rm = TRUE),
  mcur_90 = median(dat.ml.mant.clean$mcur_90, na.rm = TRUE))
dhw_grid <- predict_qbin_ci(fit_qbin, dhw_grid)

pA <- ggplot() + extrap_shade + extrap_label +
  geom_point(data = dat.ml.mant.clean, aes(x = MaxDHW.mean, y = Mort.prop.nudge), alpha = 0.35, size = 1.2) +
  geom_ribbon(data = dhw_grid, aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "steelblue", alpha = 0.15) +
  geom_line(data = dhw_grid, aes(x = MaxDHW.mean, y = pred), colour = "steelblue", linewidth = 1) +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", title = "(A) DHW Dose-Response") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11)

hist_levels <- quantile(dat.ml.mant.clean$histmDHW6, c(0.1, 0.5, 0.9), na.rm = TRUE)
grid_hist <- expand.grid(MaxDHW.mean = seq(0, 20, length.out = 200), histmDHW6 = hist_levels) %>%
  mutate(secc3m = median(dat.ml.mant.clean$secc3m, na.rm=T), cloudp_90 = median(dat.ml.mant.clean$cloudp_90, na.rm=T),
         winyear_sd = median(dat.ml.mant.clean$winyear_sd, na.rm=T), mcur_90 = median(dat.ml.mant.clean$mcur_90, na.rm=T))
grid_hist <- predict_qbin_ci(fit_qbin, grid_hist) %>%
  mutate(hist_label = factor(paste0(round(histmDHW6, 1), " prior events"), levels = paste0(round(hist_levels, 1), " prior events")))

pB <- ggplot(grid_hist, aes(x = MaxDHW.mean, y = pred, colour = hist_label, fill = hist_label)) +
  extrap_shade + extrap_label +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Set1") + scale_fill_brewer(palette = "Set1") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion",
       colour = "Prior Exposure\n(>6 DHW events)", fill = "Prior Exposure\n(>6 DHW events)", title = "(B) DHW × Prior Exposure") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

cloud_levels <- quantile(dat.ml.mant.clean$cloudp_90, c(0.1, 0.5, 0.9), na.rm = TRUE)
grid_cloud <- expand.grid(MaxDHW.mean = seq(0, 20, length.out = 200), cloudp_90 = cloud_levels) %>%
  mutate(secc3m = median(dat.ml.mant.clean$secc3m, na.rm=T), histmDHW6 = median(dat.ml.mant.clean$histmDHW6, na.rm=T),
         winyear_sd = median(dat.ml.mant.clean$winyear_sd, na.rm=T), mcur_90 = median(dat.ml.mant.clean$mcur_90, na.rm=T))
grid_cloud <- predict_qbin_ci(fit_qbin, grid_cloud) %>%
  mutate(cloud_label = factor(paste0(round(cloudp_90*100,0), "%"), levels = paste0(round(cloud_levels*100,0), "%")))

pC <- ggplot(grid_cloud, aes(x = MaxDHW.mean, y = pred, colour = cloud_label, fill = cloud_label)) +
  extrap_shade + extrap_label +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Cloud Cover", fill = "Cloud Cover", title = "(C) DHW × Cloud Cover") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

grid_scenario <- data.frame(MaxDHW.mean = rep(seq(0, 20, length.out = 200), 2),
  scenario = rep(c("Worst case", "Best case"), each = 200),
  cloudp_90 = rep(c(quantile(dat.ml.mant.clean$cloudp_90, 0.05, na.rm=T), quantile(dat.ml.mant.clean$cloudp_90, 0.95, na.rm=T)), each = 200),
  histmDHW6 = rep(c(0, quantile(dat.ml.mant.clean$histmDHW6, 0.95, na.rm=T)), each = 200),
  secc3m = median(dat.ml.mant.clean$secc3m, na.rm=T), winyear_sd = median(dat.ml.mant.clean$winyear_sd, na.rm=T),
  mcur_90 = median(dat.ml.mant.clean$mcur_90, na.rm=T))
grid_scenario <- predict_qbin_ci(fit_qbin, grid_scenario)
grid_scenario$scenario <- factor(grid_scenario$scenario, levels = c("Worst case", "Best case"))
grid_envelope <- grid_scenario %>% select(MaxDHW.mean, scenario, pred, pred_lo, pred_hi) %>%
  pivot_wider(id_cols = MaxDHW.mean, names_from = scenario, values_from = c(pred, pred_lo, pred_hi))

pD <- ggplot() + extrap_shade + extrap_label +
  geom_ribbon(data = grid_envelope, aes(x = MaxDHW.mean, ymin = `pred_Best case`, ymax = `pred_Worst case`), fill = "grey70", alpha = 0.25) +
  geom_ribbon(data = grid_scenario %>% filter(scenario == "Worst case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "firebrick", alpha = 0.12) +
  geom_ribbon(data = grid_scenario %>% filter(scenario == "Best case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "forestgreen", alpha = 0.12) +
  geom_line(data = grid_scenario, aes(x = MaxDHW.mean, y = pred, colour = scenario), linewidth = 1.2) +
  scale_colour_manual(values = c("Worst case" = "firebrick", "Best case" = "forestgreen")) +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Scenario",
       title = "(D) Worst vs Best Case", subtitle = "Worst: low cloud, no prior exposure | Best: high cloud, high prior exposure") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85), plot.subtitle = element_text(size = 8))

(pA + pB) / (pC + pD)
```
:::



## LTMP Benthic Photo-Transect (N=107, Offshore)

This section presents results for the LTMP Benthic photo-transect dataset — offshore reefs surveyed at 5 m depth with higher spatial precision. The dataset includes proportion Acropora as a community vulnerability metric.

```{r qbin-ltmp-model}
#| label: qbin-ltmp-model
#| code-fold: true
#| code-summary: "Data Preparation & Quasibinomial GLM Summary"

acrop_comp <- df.AIMS.full %>%
  filter(data_type == "photo-transect", domain_category == "reef",
         purpose == "COMPOSITION", variable == "HARD CORAL") %>%
  mutate(Reef_Name_Clean = gsub(" REEF(S)?| ISLAND| IS", "", toupper(domain_name))) %>%
  left_join(dat.Ref.Clean %>% select(AIMS_REEF_NAME_Clean, ReefName),
            by = c("Reef_Name_Clean" = "AIMS_REEF_NAME_Clean")) %>%
  filter(!is.na(ReefName)) %>%
  group_by(ReefName, report_year, depth) %>%
  summarise(total_hc = sum(mean, na.rm = TRUE),
    acrop_cover = sum(mean[grepl("Acropora", reefpage_category, ignore.case = TRUE)], na.rm = TRUE),
    .groups = "drop") %>%
  mutate(prop_acropora = ifelse(total_hc > 0, acrop_cover / total_hc, 0))

dat.ml.ltmp.clean <- dat.ml.ltmp %>%
  filter(!is.na(mcur_90), !is.na(Rel.Change)) %>%
  left_join(acrop_comp %>% select(ReefName, report_year, depth, prop_acropora),
            by = c("ReefName", "report_year", "depth")) %>%
  mutate(prop_acropora = replace_na(prop_acropora, median(prop_acropora, na.rm = TRUE)))
N_ltmp <- nrow(dat.ml.ltmp.clean)
dat.ml.ltmp.clean <- dat.ml.ltmp.clean %>%
  mutate(Mort.prop = pmin(pmax(-Rel.Change, 0), 1),
         Mort.prop.nudge = (Mort.prop * (N_ltmp - 1) + 0.5) / N_ltmp)
cat("LTMP Benthic model dataset:", N_ltmp, "observations\n")

fit_qbin_ltmp <- glm(
  Mort.prop.nudge ~ MaxDHW.mean * secc3m + MaxDHW.mean * cloudp_90 +
                    MaxDHW.mean * histmDHW6 + MaxDHW.mean * mcur_90 +
                    MaxDHW.mean * prop_acropora + winyear_sd,
  data = dat.ml.ltmp.clean, family = quasibinomial(link = "logit"))
print(summary(fit_qbin_ltmp))
```

```{r brms-ltmp-model}
#| label: brms-ltmp-model
#| cache: true

# Scale LTMP predictors (same Gelman scaling)
scale_params_ltmp <- list(
  MaxDHW.mean = c(mean = mean(dat.ml.ltmp.clean$MaxDHW.mean, na.rm=T), sd2 = 2*sd(dat.ml.ltmp.clean$MaxDHW.mean, na.rm=T)),
  histmDHW6   = c(mean = mean(dat.ml.ltmp.clean$histmDHW6, na.rm=T), sd2 = 2*sd(dat.ml.ltmp.clean$histmDHW6, na.rm=T)),
  secc3m      = c(mean = mean(dat.ml.ltmp.clean$secc3m, na.rm=T), sd2 = 2*sd(dat.ml.ltmp.clean$secc3m, na.rm=T)),
  cloudp_90   = c(mean = mean(dat.ml.ltmp.clean$cloudp_90, na.rm=T), sd2 = 2*sd(dat.ml.ltmp.clean$cloudp_90, na.rm=T)),
  winyear_sd  = c(mean = mean(dat.ml.ltmp.clean$winyear_sd, na.rm=T), sd2 = 2*sd(dat.ml.ltmp.clean$winyear_sd, na.rm=T)),
  mcur_90     = c(mean = mean(dat.ml.ltmp.clean$mcur_90, na.rm=T), sd2 = 2*sd(dat.ml.ltmp.clean$mcur_90, na.rm=T)),
  prop_acropora = c(mean = mean(dat.ml.ltmp.clean$prop_acropora, na.rm=T), sd2 = 2*sd(dat.ml.ltmp.clean$prop_acropora, na.rm=T))
)

s2 <- function(x) (x - mean(x, na.rm=T)) / (2*sd(x, na.rm=T))
dat.ml.ltmp.clean <- dat.ml.ltmp.clean %>%
  mutate(MaxDHW_s = s2(MaxDHW.mean), histmDHW6_s = s2(histmDHW6), secc3m_s = s2(secc3m),
         cloudp_90_s = s2(cloudp_90), winyear_sd_s = s2(winyear_sd), mcur_90_s = s2(mcur_90),
         prop_acropora_s = s2(prop_acropora), obs_id = 1:n(),
         n_trials = 100L, y_binom = as.integer(round(Mort.prop.nudge * 100)))

cat("LTMP brms dataset:", nrow(dat.ml.ltmp.clean), "obs,", length(unique(dat.ml.ltmp.clean$ReefName)), "reefs\n")

# Beta model
cache_ltmp_beta <- "output/models/brms_ltmp_beta.rds"
if (file.exists(cache_ltmp_beta)) {
  fit_brms_ltmp <- readRDS(cache_ltmp_beta)
  cat("Loaded cached LTMP Beta model.\n")
} else {
  cat("Fitting LTMP Beta model...\n")
  fit_brms_ltmp <- brm(
    Mort.prop.nudge ~ MaxDHW_s * cloudp_90_s + MaxDHW_s * histmDHW6_s +
                      MaxDHW_s * mcur_90_s + MaxDHW_s * prop_acropora_s +
                      secc3m_s + winyear_sd_s + (1 | ReefName),
    data = dat.ml.ltmp.clean, family = Beta(link = "logit"),
    prior = c(prior(normal(-1, 1.5), class = "Intercept"),
              prior(normal(0, 1), class = "b"),
              prior(exponential(1), class = "sd")),
    chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 42,
    control = list(adapt_delta = 0.97)
  )
  saveRDS(fit_brms_ltmp, cache_ltmp_beta)
}

# Binomial-OLRE model
cache_ltmp_binom <- "output/models/brms_ltmp_binomial_olre.rds"
if (file.exists(cache_ltmp_binom)) {
  fit_brmsb_ltmp <- readRDS(cache_ltmp_binom)
  cat("Loaded cached LTMP Binomial-OLRE model.\n")
} else {
  cat("Fitting LTMP Binomial-OLRE model...\n")
  fit_brmsb_ltmp <- brm(
    y_binom | trials(n_trials) ~ MaxDHW_s * cloudp_90_s + MaxDHW_s * histmDHW6_s +
                      MaxDHW_s * mcur_90_s + MaxDHW_s * prop_acropora_s +
                      secc3m_s + winyear_sd_s + (1 | ReefName) + (1 | obs_id),
    data = dat.ml.ltmp.clean, family = binomial(link = "logit"),
    prior = c(prior(normal(-1, 1.5), class = "Intercept"),
              prior(normal(0, 1), class = "b"),
              prior(exponential(1), class = "sd")),
    chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 42,
    control = list(adapt_delta = 0.97)
  )
  saveRDS(fit_brmsb_ltmp, cache_ltmp_binom)
}

cat("\n=== LTMP Beta Model ===\n")
print(summary(fit_brms_ltmp))
cat("\nBeta R²:"); print(bayes_R2(fit_brms_ltmp))
cat("\nBinomial-OLRE R²:"); print(bayes_R2(fit_brmsb_ltmp))
```


### Effect Size Comparison

::: {.panel-tabset}
#### Bayesian Beta

```{r brms-halfeye-ltmp}
#| label: brms-halfeye-ltmp
#| fig-cap: "LTMP Benthic — Posterior distributions (Beta)."
#| fig-width: 10
#| fig-height: 6

library(brms); library(dplyr); library(tidybayes)
draws_ltmp <- as_draws_df(fit_brms_ltmp) %>%
  select(starts_with("b_")) %>%
  pivot_longer(everything(), names_to = "term", values_to = "estimate") %>%
  filter(term != "b_Intercept") %>%
  mutate(term = gsub("b_", "", term), term = gsub("MaxDHW_s", "DHW", term),
         term = gsub("histmDHW6_s", "Prior Exposure", term), term = gsub("secc3m_s", "Secchi Depth", term),
         term = gsub("cloudp_90_s", "Cloud Cover", term), term = gsub("winyear_sd_s", "Winter SST SD", term),
         term = gsub("mcur_90_s", "Current Speed", term), term = gsub("prop_acropora_s", "Prop. Acropora", term),
         term = gsub(":", " x ", term))
ggplot(draws_ltmp, aes(x = estimate, y = reorder(term, estimate))) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_qi, fill = "darkorange", alpha = 0.7, normalize = "xy") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(x = "Posterior Estimate (logit scale)", y = NULL, title = "LTMP Benthic — Posterior Effect Distributions (Beta)") +
  theme_bw(base_size = 12)
```

#### Bayesian Binomial-OLRE

```{r brms-halfeye-ltmp-olre}
#| label: brms-halfeye-ltmp-olre
#| fig-cap: "LTMP Benthic — Posterior distributions (Binomial-OLRE)."
#| fig-width: 10
#| fig-height: 6

library(brms); library(dplyr); library(tidybayes)
draws_ltmp_olre <- as_draws_df(fit_brmsb_ltmp) %>%
  select(starts_with("b_")) %>%
  pivot_longer(everything(), names_to = "term", values_to = "estimate") %>%
  filter(term != "b_Intercept") %>%
  mutate(term = gsub("b_", "", term), term = gsub("MaxDHW_s", "DHW", term),
         term = gsub("histmDHW6_s", "Prior Exposure", term), term = gsub("secc3m_s", "Secchi Depth", term),
         term = gsub("cloudp_90_s", "Cloud Cover", term), term = gsub("winyear_sd_s", "Winter SST SD", term),
         term = gsub("mcur_90_s", "Current Speed", term), term = gsub("prop_acropora_s", "Prop. Acropora", term),
         term = gsub(":", " x ", term))
ggplot(draws_ltmp_olre, aes(x = estimate, y = reorder(term, estimate))) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_qi, fill = "coral", alpha = 0.7, normalize = "xy") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(x = "Posterior Estimate (logit scale)", y = NULL, title = "LTMP Benthic — Posterior Effect Distributions (Binomial-OLRE)") +
  theme_bw(base_size = 12)
```

#### Quasibinomial GLM

```{r qbin-ltmp-effects}
#| label: qbin-ltmp-effects
#| fig-cap: "LTMP Benthic — Coefficient estimates (logit scale) with 90% CI."
#| fig-width: 8
#| fig-height: 5

coef_ltmp <- as.data.frame(coef(summary(fit_qbin_ltmp)))
names(coef_ltmp) <- c("Estimate", "SE", "t", "p")
coef_ltmp$Term <- rownames(coef_ltmp)
coef_ltmp <- coef_ltmp[coef_ltmp$Term != "(Intercept)", ]
z90 <- qnorm(0.95)
coef_ltmp$lo <- coef_ltmp$Estimate - z90 * coef_ltmp$SE
coef_ltmp$hi <- coef_ltmp$Estimate + z90 * coef_ltmp$SE
coef_ltmp$Term <- gsub("MaxDHW.mean", "DHW", coef_ltmp$Term)
coef_ltmp$Term <- gsub("histmDHW6", "Prior Exposure (>6 DHW)", coef_ltmp$Term)
coef_ltmp$Term <- gsub("secc3m", "Secchi Depth", coef_ltmp$Term)
coef_ltmp$Term <- gsub("cloudp_90", "Cloud Cover", coef_ltmp$Term)
coef_ltmp$Term <- gsub("winyear_sd", "Winter SST SD", coef_ltmp$Term)
coef_ltmp$Term <- gsub("mcur_90", "Current Speed", coef_ltmp$Term)
coef_ltmp$Term <- gsub("prop_acropora", "Prop. Acropora", coef_ltmp$Term)
coef_ltmp$Term <- gsub(":", " x ", coef_ltmp$Term)
coef_ltmp$sig <- ifelse(coef_ltmp$lo > 0 | coef_ltmp$hi < 0, "Significant", "Not significant")
coef_ltmp$Term <- reorder(coef_ltmp$Term, coef_ltmp$Estimate)
ggplot(coef_ltmp, aes(x = Estimate, y = Term, colour = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = lo, xmax = hi), size = 0.7, fatten = 3) +
  scale_colour_manual(values = c("Significant" = "black", "Not significant" = "grey60")) +
  labs(x = "Coefficient Estimate (logit scale)", y = NULL, colour = NULL,
       title = "LTMP Benthic — Effect Sizes (90% CI)") +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```
:::

### Dose-Response Curves

::: {.panel-tabset}
#### Bayesian Beta

```{r brms-dose-ltmp}
#| label: brms-dose-ltmp
#| fig-cap: "LTMP Benthic — 6-panel Bayesian dose-response summary."
#| fig-width: 12
#| fig-height: 14

library(brms)
library(dplyr)
d <- dat.ml.ltmp.clean; m <- fit_brms_ltmp; sp <- scale_params_ltmp
dhw_ext <- max(d$MaxDHW.mean, na.rm = TRUE)
dhw_max_s <- (20 - sp$MaxDHW.mean["mean"]) / sp$MaxDHW.mean["sd2"]

brms_pred_ltmp <- function(mod, nd) {
  pp <- posterior_epred(mod, newdata = nd, re_formula = NA)
  nd$pred    <- apply(pp, 2, median)
  nd$pred_lo <- apply(pp, 2, quantile, 0.05)
  nd$pred_hi <- apply(pp, 2, quantile, 0.95)
  nd$MaxDHW.mean <- nd$MaxDHW_s * sp$MaxDHW.mean["sd2"] + sp$MaxDHW.mean["mean"]
  nd
}

es <- annotate("rect", xmin = dhw_ext, xmax = 20, ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.35)
el <- annotate("text", x = (dhw_ext + 20)/2, y = 0.05, label = "Extrapolation", fontface = "italic", colour = "grey40", size = 3)

# Panel A: DHW
ga <- data.frame(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 100),
  secc3m_s = 0, cloudp_90_s = 0, histmDHW6_s = 0, winyear_sd_s = 0, mcur_90_s = 0, prop_acropora_s = 0)
ga <- brms_pred_ltmp(m, ga)
pA2 <- ggplot() + es + el +
  geom_point(data = d, aes(x = MaxDHW.mean, y = Mort.prop.nudge), alpha = 0.35, size = 1.2) +
  geom_ribbon(data = ga, aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "steelblue", alpha = 0.2) +
  geom_line(data = ga, aes(x = MaxDHW.mean, y = pred), colour = "steelblue", linewidth = 1) +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", title = "(A) Bayesian LTMP: DHW Dose-Response") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11)

# Panel B: Prior Exposure alone (main effect)
hl_range_raw <- seq(quantile(d$histmDHW6, 0.05, na.rm = TRUE), quantile(d$histmDHW6, 0.95, na.rm = TRUE), length.out = 100)
hl_range_s <- (hl_range_raw - sp$histmDHW6["mean"]) / sp$histmDHW6["sd2"]
gb_main <- data.frame(histmDHW6_s = hl_range_s, MaxDHW_s = 0,
  secc3m_s = 0, cloudp_90_s = 0, winyear_sd_s = 0, mcur_90_s = 0, prop_acropora_s = 0)
gb_main <- brms_pred_ltmp(m, gb_main) %>%
  mutate(histmDHW6 = histmDHW6_s * sp$histmDHW6["sd2"] + sp$histmDHW6["mean"])
pB2_main <- ggplot(gb_main, aes(x = histmDHW6, y = pred)) +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), fill = "purple", alpha = 0.15) +
  geom_line(colour = "purple", linewidth = 1) +
  labs(x = "Prior Severe Bleaching Count", y = "Mortality Proportion", title = "(B) Prior Exposure Main Effect",
       subtitle = "DHW held at median") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11)

# Panel C: DHW × Prior Exposure
hl_raw <- quantile(d$histmDHW6, c(0.1, 0.5, 0.9), na.rm = TRUE)
hl_s <- (hl_raw - sp$histmDHW6["mean"]) / sp$histmDHW6["sd2"]
gb <- expand.grid(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 100), histmDHW6_s = hl_s) %>%
  mutate(secc3m_s = 0, cloudp_90_s = 0, winyear_sd_s = 0, mcur_90_s = 0, prop_acropora_s = 0)
gb <- brms_pred_ltmp(m, gb) %>%
  mutate(histmDHW6 = histmDHW6_s * sp$histmDHW6["sd2"] + sp$histmDHW6["mean"],
         lbl = factor(paste0(round(histmDHW6, 1), " prior"), levels = paste0(round(hl_raw, 1), " prior")))
pC2 <- ggplot(gb, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Set1") + scale_fill_brewer(palette = "Set1") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Prior Exposure", fill = "Prior Exposure",
       title = "(C) DHW × Prior Exposure") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel D: DHW × Cloud Cover
cl_raw <- quantile(d$cloudp_90, c(0.1, 0.5, 0.9), na.rm = TRUE)
cl_s <- (cl_raw - sp$cloudp_90["mean"]) / sp$cloudp_90["sd2"]
gc <- expand.grid(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 100), cloudp_90_s = cl_s) %>%
  mutate(secc3m_s = 0, histmDHW6_s = 0, winyear_sd_s = 0, mcur_90_s = 0, prop_acropora_s = 0)
gc <- brms_pred_ltmp(m, gc) %>%
  mutate(cloudp_90 = cloudp_90_s * sp$cloudp_90["sd2"] + sp$cloudp_90["mean"],
         lbl = factor(paste0(round(cloudp_90 * 100, 0), "%"), levels = paste0(round(cl_raw * 100, 0), "%")))
pD2 <- ggplot(gc, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Cloud Cover", fill = "Cloud Cover",
       title = "(D) DHW × Cloud Cover") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel E: DHW × % Acropora
al_raw <- quantile(d$prop_acropora, c(0.1, 0.5, 0.9), na.rm = TRUE)
al_s <- (al_raw - sp$prop_acropora["mean"]) / sp$prop_acropora["sd2"]
ge_acrop <- expand.grid(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 100), prop_acropora_s = al_s) %>%
  mutate(secc3m_s = 0, cloudp_90_s = 0, histmDHW6_s = 0, winyear_sd_s = 0, mcur_90_s = 0)
ge_acrop <- brms_pred_ltmp(m, ge_acrop) %>%
  mutate(prop_acropora = prop_acropora_s * sp$prop_acropora["sd2"] + sp$prop_acropora["mean"],
         lbl = factor(paste0(round(prop_acropora*100, 0), "% Acrop."), levels = paste0(round(al_raw*100, 0), "% Acrop.")))
pE2 <- ggplot(ge_acrop, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Set2") + scale_fill_brewer(palette = "Set2") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Acropora", fill = "Acropora",
       title = "(E) DHW × % Acropora") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel F: Worst vs Best
cl_10 <- (quantile(d$cloudp_90, 0.10, na.rm=T) - sp$cloudp_90["mean"]) / sp$cloudp_90["sd2"]
cl_90 <- (quantile(d$cloudp_90, 0.90, na.rm=T) - sp$cloudp_90["mean"]) / sp$cloudp_90["sd2"]
h_10  <- (quantile(d$histmDHW6, 0.10, na.rm=T) - sp$histmDHW6["mean"]) / sp$histmDHW6["sd2"]
h_90  <- (quantile(d$histmDHW6, 0.90, na.rm=T) - sp$histmDHW6["mean"]) / sp$histmDHW6["sd2"]
a_10  <- (quantile(d$prop_acropora, 0.10, na.rm=T) - sp$prop_acropora["mean"]) / sp$prop_acropora["sd2"]
a_90  <- (quantile(d$prop_acropora, 0.90, na.rm=T) - sp$prop_acropora["mean"]) / sp$prop_acropora["sd2"]

gf <- data.frame(MaxDHW_s = rep(seq(-1.5, dhw_max_s, length.out = 100), 2),
  scenario = rep(c("Worst case", "Best case"), each = 100),
  cloudp_90_s = rep(c(cl_10, cl_90), each = 100),
  histmDHW6_s = rep(c(h_10, h_90), each = 100),
  prop_acropora_s = rep(c(a_90, a_10), each = 100),
  secc3m_s = 0, winyear_sd_s = 0, mcur_90_s = 0)
gf <- brms_pred_ltmp(m, gf); gf$scenario <- factor(gf$scenario, levels = c("Worst case", "Best case"))
gfe <- gf %>% select(MaxDHW.mean, scenario, pred, pred_lo, pred_hi) %>%
  pivot_wider(id_cols = MaxDHW.mean, names_from = scenario, values_from = c(pred, pred_lo, pred_hi))

pF2 <- ggplot() + es + el +
  geom_ribbon(data = gfe, aes(x = MaxDHW.mean, ymin = `pred_Best case`, ymax = `pred_Worst case`), fill = "grey70", alpha = 0.25) +
  geom_ribbon(data = gf %>% filter(scenario == "Worst case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "firebrick", alpha = 0.12) +
  geom_ribbon(data = gf %>% filter(scenario == "Best case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "forestgreen", alpha = 0.12) +
  geom_line(data = gf, aes(x = MaxDHW.mean, y = pred, colour = scenario), linewidth = 1.2) +
  scale_colour_manual(values = c("Worst case" = "firebrick", "Best case" = "forestgreen")) +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Scenario",
       title = "(F) Worst vs Best Case",
       subtitle = "Worst: low cloud (10th), no prior (10th), high Acropora (90th) | Best: high cloud (90th), prior (90th), low Acropora (10th)") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85), plot.subtitle = element_text(size = 8))

(pA2 + pB2_main) / (pC2 + pD2) / (pE2 + pF2)
```

#### Bayesian Binomial-OLRE

```{r brms-dose-ltmp-olre}
#| label: brms-dose-ltmp-olre
#| fig-cap: "LTMP Benthic — 6-panel Binomial-OLRE dose-response."
#| fig-width: 12
#| fig-height: 14

library(brms); library(dplyr); library(patchwork)
d <- dat.ml.ltmp.clean; m <- fit_brmsb_ltmp; sp <- scale_params_ltmp
dhw_ext <- max(d$MaxDHW.mean, na.rm = TRUE)
dhw_max_s <- (20 - sp$MaxDHW.mean["mean"]) / sp$MaxDHW.mean["sd2"]

brms_pred_ltmp_olre <- function(mod, nd) {
  nd$n_trials <- 1L
  pp <- posterior_epred(mod, newdata = nd, re_formula = NA)
  nd$pred    <- apply(pp, 2, median)
  nd$pred_lo <- apply(pp, 2, quantile, 0.05)
  nd$pred_hi <- apply(pp, 2, quantile, 0.95)
  nd$MaxDHW.mean <- nd$MaxDHW_s * sp$MaxDHW.mean["sd2"] + sp$MaxDHW.mean["mean"]
  nd
}

es <- annotate("rect", xmin = dhw_ext, xmax = 20, ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.35)
el <- annotate("text", x = (dhw_ext + 20)/2, y = 0.05, label = "Extrapolation", fontface = "italic", colour = "grey40", size = 3)

# Panel A: DHW
ga <- data.frame(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 200),
  secc3m_s = 0, cloudp_90_s = 0, histmDHW6_s = 0, winyear_sd_s = 0, mcur_90_s = 0, prop_acropora_s = 0)
ga <- brms_pred_ltmp_olre(m, ga)
pA2 <- ggplot() + es + el +
  geom_point(data = d, aes(x = MaxDHW.mean, y = Mort.prop.nudge), alpha = 0.35, size = 1.2) +
  geom_ribbon(data = ga, aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "coral", alpha = 0.2) +
  geom_line(data = ga, aes(x = MaxDHW.mean, y = pred), colour = "coral", linewidth = 1) +
  labs(x = "Max DHW", y = "Mortality Proportion", title = "(A) Binomial-OLRE DHW Dose-Response") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11)

# Panel B: Prior Exposure main effect
hl_raw <- seq(quantile(d$histmDHW6, 0.05, na.rm=T), quantile(d$histmDHW6, 0.95, na.rm=T), length.out = 100)
hl_s <- (hl_raw - sp$histmDHW6["mean"]) / sp$histmDHW6["sd2"]
gb_main <- data.frame(histmDHW6_s = hl_s, MaxDHW_s = 0, secc3m_s = 0, cloudp_90_s = 0,
  winyear_sd_s = 0, mcur_90_s = 0, prop_acropora_s = 0, n_trials = 1L)
pp_b <- posterior_epred(m, newdata = gb_main, re_formula = NA)
gb_main$pred <- apply(pp_b, 2, median); gb_main$pred_lo <- apply(pp_b, 2, quantile, 0.05); gb_main$pred_hi <- apply(pp_b, 2, quantile, 0.95)
gb_main$histmDHW6 <- gb_main$histmDHW6_s * sp$histmDHW6["sd2"] + sp$histmDHW6["mean"]
pB2_main <- ggplot(gb_main, aes(x = histmDHW6, y = pred)) +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), fill = "purple", alpha = 0.15) +
  geom_line(colour = "purple", linewidth = 1) +
  labs(x = "Prior Severe Bleaching Count", y = "Mortality Proportion", title = "(B) Prior Exposure Main Effect") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11)

# Panel C: DHW x Prior Exposure
hl_q <- quantile(d$histmDHW6, c(0.1, 0.5, 0.9), na.rm = TRUE)
hl_qs <- (hl_q - sp$histmDHW6["mean"]) / sp$histmDHW6["sd2"]
gc <- expand.grid(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 200), histmDHW6_s = hl_qs) %>%
  mutate(secc3m_s = 0, cloudp_90_s = 0, winyear_sd_s = 0, mcur_90_s = 0, prop_acropora_s = 0)
gc <- brms_pred_ltmp_olre(m, gc) %>%
  mutate(histmDHW6 = histmDHW6_s * sp$histmDHW6["sd2"] + sp$histmDHW6["mean"],
         lbl = factor(paste0(round(histmDHW6, 1), " prior"), levels = paste0(round(hl_q, 1), " prior")))
pC2 <- ggplot(gc, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Set1") + scale_fill_brewer(palette = "Set1") +
  labs(x = "Max DHW", y = "Mortality Proportion", colour = "Prior", fill = "Prior", title = "(C) DHW x Prior Exposure") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel D: DHW x Cloud Cover
cl_q <- quantile(d$cloudp_90, c(0.1, 0.5, 0.9), na.rm = TRUE)
cl_qs <- (cl_q - sp$cloudp_90["mean"]) / sp$cloudp_90["sd2"]
gd <- expand.grid(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 200), cloudp_90_s = cl_qs) %>%
  mutate(secc3m_s = 0, histmDHW6_s = 0, winyear_sd_s = 0, mcur_90_s = 0, prop_acropora_s = 0)
gd <- brms_pred_ltmp_olre(m, gd) %>%
  mutate(cloudp_90 = cloudp_90_s * sp$cloudp_90["sd2"] + sp$cloudp_90["mean"],
         lbl = factor(paste0(round(cloudp_90*100,0), "%"), levels = paste0(round(cl_q*100,0), "%")))
pD2 <- ggplot(gd, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  labs(x = "Max DHW", y = "Mortality Proportion", colour = "Cloud Cover", fill = "Cloud Cover", title = "(D) DHW x Cloud Cover") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel E: DHW x % Acropora
al_q <- quantile(d$prop_acropora, c(0.1, 0.5, 0.9), na.rm = TRUE)
al_qs <- (al_q - sp$prop_acropora["mean"]) / sp$prop_acropora["sd2"]
ge <- expand.grid(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 200), prop_acropora_s = al_qs) %>%
  mutate(secc3m_s = 0, cloudp_90_s = 0, histmDHW6_s = 0, winyear_sd_s = 0, mcur_90_s = 0)
ge <- brms_pred_ltmp_olre(m, ge) %>%
  mutate(prop_acropora = prop_acropora_s * sp$prop_acropora["sd2"] + sp$prop_acropora["mean"],
         lbl = factor(paste0(round(prop_acropora*100,0), "% Acrop."), levels = paste0(round(al_q*100,0), "% Acrop.")))
pE2 <- ggplot(ge, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Set2") + scale_fill_brewer(palette = "Set2") +
  labs(x = "Max DHW", y = "Mortality Proportion", colour = "Acropora", fill = "Acropora", title = "(E) DHW x % Acropora") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel F: Worst vs Best
cl_10s <- (quantile(d$cloudp_90, 0.10, na.rm=T) - sp$cloudp_90["mean"]) / sp$cloudp_90["sd2"]
cl_90s <- (quantile(d$cloudp_90, 0.90, na.rm=T) - sp$cloudp_90["mean"]) / sp$cloudp_90["sd2"]
h_10s <- (quantile(d$histmDHW6, 0.10, na.rm=T) - sp$histmDHW6["mean"]) / sp$histmDHW6["sd2"]
h_90s <- (quantile(d$histmDHW6, 0.90, na.rm=T) - sp$histmDHW6["mean"]) / sp$histmDHW6["sd2"]
a_10s <- (quantile(d$prop_acropora, 0.10, na.rm=T) - sp$prop_acropora["mean"]) / sp$prop_acropora["sd2"]
a_90s <- (quantile(d$prop_acropora, 0.90, na.rm=T) - sp$prop_acropora["mean"]) / sp$prop_acropora["sd2"]
gf <- data.frame(MaxDHW_s = rep(seq(-1.5, dhw_max_s, length.out = 200), 2),
  scenario = rep(c("Worst case", "Best case"), each = 200),
  cloudp_90_s = rep(c(cl_10s, cl_90s), each = 200),
  histmDHW6_s = rep(c(h_10s, h_90s), each = 200),
  prop_acropora_s = rep(c(a_90s, a_10s), each = 200),
  secc3m_s = 0, winyear_sd_s = 0, mcur_90_s = 0)
gf <- brms_pred_ltmp_olre(m, gf); gf$scenario <- factor(gf$scenario, levels = c("Worst case", "Best case"))
gfe <- gf %>% select(MaxDHW.mean, scenario, pred, pred_lo, pred_hi) %>%
  pivot_wider(id_cols = MaxDHW.mean, names_from = scenario, values_from = c(pred, pred_lo, pred_hi))
pF2 <- ggplot() + es + el +
  geom_ribbon(data = gfe, aes(x = MaxDHW.mean, ymin = `pred_Best case`, ymax = `pred_Worst case`), fill = "grey70", alpha = 0.25) +
  geom_ribbon(data = gf %>% filter(scenario == "Worst case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "firebrick", alpha = 0.12) +
  geom_ribbon(data = gf %>% filter(scenario == "Best case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "forestgreen", alpha = 0.12) +
  geom_line(data = gf, aes(x = MaxDHW.mean, y = pred, colour = scenario), linewidth = 1.2) +
  scale_colour_manual(values = c("Worst case" = "firebrick", "Best case" = "forestgreen")) +
  labs(x = "Max DHW", y = "Mortality Proportion", colour = "Scenario", title = "(F) Worst vs Best Case") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85), plot.subtitle = element_text(size = 8))

(pA2 + pB2_main) / (pC2 + pD2) / (pE2 + pF2)
```

#### Quasibinomial GLM

```{r qbin-ltmp-dose}
#| label: qbin-ltmp-dose
#| fig-cap: "LTMP Benthic — 6-panel dose-response summary."
#| fig-width: 12
#| fig-height: 14

library(patchwork)
pci <- function(mod, nd, level = 0.90) {
  z <- qnorm(1 - (1 - level)/2); p <- predict(mod, newdata = nd, type = "link", se.fit = TRUE)
  nd$pred <- plogis(p$fit); nd$pred_lo <- plogis(p$fit - z*p$se.fit); nd$pred_hi <- plogis(p$fit + z*p$se.fit); nd }
dhw_ext <- max(dat.ml.ltmp.clean$MaxDHW.mean, na.rm = TRUE)
es <- annotate("rect", xmin = dhw_ext, xmax = 20, ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.35)
el <- annotate("text", x = (dhw_ext + 20)/2, y = 0.05, label = "Extrapolation", fontface = "italic", colour = "grey40", size = 3)
d <- dat.ml.ltmp.clean; m <- fit_qbin_ltmp

# Panel A: DHW dose-response
ga <- data.frame(MaxDHW.mean = seq(0, 20, length.out = 200), secc3m = median(d$secc3m, na.rm=T),
  cloudp_90 = median(d$cloudp_90, na.rm=T), histmDHW6 = median(d$histmDHW6, na.rm=T),
  winyear_sd = median(d$winyear_sd, na.rm=T), mcur_90 = median(d$mcur_90, na.rm=T),
  prop_acropora = median(d$prop_acropora, na.rm=T))
ga <- pci(m, ga)
pA2 <- ggplot() + es + el +
  geom_point(data = d, aes(x = MaxDHW.mean, y = Mort.prop.nudge), alpha = 0.35, size = 1.2) +
  geom_ribbon(data = ga, aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "steelblue", alpha = 0.2) +
  geom_line(data = ga, aes(x = MaxDHW.mean, y = pred), colour = "steelblue", linewidth = 1) +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", title = "(A) LTMP: DHW Dose-Response") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11)

# Panel B: Prior Exposure alone (main effect)
hl_range <- seq(quantile(d$histmDHW6, 0.05, na.rm = TRUE), quantile(d$histmDHW6, 0.95, na.rm = TRUE), length.out = 100)
gb_main <- data.frame(histmDHW6 = hl_range, MaxDHW.mean = median(d$MaxDHW.mean, na.rm = TRUE),
  secc3m = median(d$secc3m, na.rm=T), cloudp_90 = median(d$cloudp_90, na.rm=T),
  winyear_sd = median(d$winyear_sd, na.rm=T), mcur_90 = median(d$mcur_90, na.rm=T),
  prop_acropora = median(d$prop_acropora, na.rm=T))
gb_main <- pci(m, gb_main)
pB2_main <- ggplot(gb_main, aes(x = histmDHW6, y = pred)) +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), fill = "purple", alpha = 0.15) +
  geom_line(colour = "purple", linewidth = 1) +
  labs(x = "Prior Severe Bleaching Count", y = "Mortality Proportion", title = "(B) Prior Exposure Main Effect",
       subtitle = "DHW held at median") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11)

# Panel C: DHW × Prior Exposure
hl <- quantile(d$histmDHW6, c(0.1, 0.5, 0.9), na.rm = TRUE)
gb <- expand.grid(MaxDHW.mean = seq(0, 20, length.out = 200), histmDHW6 = hl) %>%
  mutate(secc3m = median(d$secc3m, na.rm=T), cloudp_90 = median(d$cloudp_90, na.rm=T),
         winyear_sd = median(d$winyear_sd, na.rm=T), mcur_90 = median(d$mcur_90, na.rm=T),
         prop_acropora = median(d$prop_acropora, na.rm=T))
gb <- pci(m, gb) %>% mutate(lbl = factor(paste0(round(histmDHW6,1)," prior"), levels = paste0(round(hl,1)," prior")))
pC2 <- ggplot(gb, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Set1") + scale_fill_brewer(palette = "Set1") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Prior Exposure", fill = "Prior Exposure",
       title = "(C) DHW × Prior Exposure") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel D: DHW × Cloud Cover
cl <- quantile(d$cloudp_90, c(0.1, 0.5, 0.9), na.rm = TRUE)
gc <- expand.grid(MaxDHW.mean = seq(0, 20, length.out = 200), cloudp_90 = cl) %>%
  mutate(secc3m = median(d$secc3m, na.rm=T), histmDHW6 = median(d$histmDHW6, na.rm=T),
         winyear_sd = median(d$winyear_sd, na.rm=T), mcur_90 = median(d$mcur_90, na.rm=T),
         prop_acropora = median(d$prop_acropora, na.rm=T))
gc <- pci(m, gc) %>% mutate(lbl = factor(paste0(round(cloudp_90*100,0),"%"), levels = paste0(round(cl*100,0),"%")))
pD2 <- ggplot(gc, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Cloud Cover", fill = "Cloud Cover",
       title = "(D) DHW × Cloud Cover") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel E: DHW × % Acropora
al_raw <- quantile(d$prop_acropora, c(0.1, 0.5, 0.9), na.rm = TRUE)
ge_acrop <- expand.grid(MaxDHW.mean = seq(0, 20, length.out = 200), prop_acropora = al_raw) %>%
  mutate(secc3m = median(d$secc3m, na.rm=T), cloudp_90 = median(d$cloudp_90, na.rm=T),
         histmDHW6 = median(d$histmDHW6, na.rm=T), winyear_sd = median(d$winyear_sd, na.rm=T),
         mcur_90 = median(d$mcur_90, na.rm=T))
ge_acrop <- pci(m, ge_acrop) %>%
  mutate(lbl = factor(paste0(round(prop_acropora * 100, 0), "% Acrop."), levels = paste0(round(al_raw * 100, 0), "% Acrop.")))
pE2 <- ggplot(ge_acrop, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Set2") + scale_fill_brewer(palette = "Set2") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Acropora", fill = "Acropora",
       title = "(E) DHW × % Acropora") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel F: Worst vs Best
cl_10 <- quantile(d$cloudp_90, 0.10, na.rm = TRUE)
cl_90 <- quantile(d$cloudp_90, 0.90, na.rm = TRUE)
h_10  <- quantile(d$histmDHW6, 0.10, na.rm = TRUE)
h_90  <- quantile(d$histmDHW6, 0.90, na.rm = TRUE)
a_10  <- quantile(d$prop_acropora, 0.10, na.rm = TRUE)
a_90  <- quantile(d$prop_acropora, 0.90, na.rm = TRUE)

gf <- data.frame(MaxDHW.mean = rep(seq(0, 20, length.out = 200), 2),
  scenario = rep(c("Worst case", "Best case"), each = 200),
  cloudp_90 = rep(c(cl_10, cl_90), each = 200),
  histmDHW6 = rep(c(h_10, h_90), each = 200),
  prop_acropora = rep(c(a_90, a_10), each = 200),
  secc3m = median(d$secc3m, na.rm=T), winyear_sd = median(d$winyear_sd, na.rm=T), mcur_90 = median(d$mcur_90, na.rm=T))
gf <- pci(m, gf); gf$scenario <- factor(gf$scenario, levels = c("Worst case", "Best case"))
gfe <- gf %>% select(MaxDHW.mean, scenario, pred, pred_lo, pred_hi) %>%
  pivot_wider(id_cols = MaxDHW.mean, names_from = scenario, values_from = c(pred, pred_lo, pred_hi))

pF2 <- ggplot() + es + el +
  geom_ribbon(data = gfe, aes(x = MaxDHW.mean, ymin = `pred_Best case`, ymax = `pred_Worst case`), fill = "grey70", alpha = 0.25) +
  geom_ribbon(data = gf %>% filter(scenario == "Worst case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "firebrick", alpha = 0.12) +
  geom_ribbon(data = gf %>% filter(scenario == "Best case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "forestgreen", alpha = 0.12) +
  geom_line(data = gf, aes(x = MaxDHW.mean, y = pred, colour = scenario), linewidth = 1.2) +
  scale_colour_manual(values = c("Worst case" = "firebrick", "Best case" = "forestgreen")) +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Scenario",
       title = "(F) Worst vs Best Case",
       subtitle = "Worst: low cloud (10th), no prior (10th), high Acropora (90th) | Best: high cloud (90th), prior (90th), low Acropora (10th)") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85), plot.subtitle = element_text(size = 8))

(pA2 + pB2_main) / (pC2 + pD2) / (pE2 + pF2)
```
:::

## MMP Inshore (N=33, Inshore)

This section presents results for the MMP Inshore monitoring dataset. These reefs are typically more turbid and closer to river runoff, with a reduced predictor set reflecting the smaller sample size.

```{r qbin-mmp-model}
#| label: qbin-mmp-model
#| code-fold: true
#| code-summary: "Data Preparation & Quasibinomial GLM Summary"

dat.ml.mmp.clean <- dat.ml.mmp %>%
  filter(!is.na(mcur_90), !is.na(Rel.Change)) %>%
  left_join(acrop_comp %>% select(ReefName, report_year, depth, prop_acropora),
            by = c("ReefName", "report_year", "depth")) %>%
  mutate(prop_acropora = replace_na(prop_acropora, median(prop_acropora, na.rm = TRUE)))
N_mmp <- nrow(dat.ml.mmp.clean)
dat.ml.mmp.clean <- dat.ml.mmp.clean %>%
  mutate(Mort.prop = pmin(pmax(-Rel.Change, 0), 1),
         Mort.prop.nudge = (Mort.prop * (N_mmp - 1) + 0.5) / N_mmp)
cat("MMP Inshore model dataset:", N_mmp, "observations\n")

fit_qbin_mmp <- glm(
  Mort.prop.nudge ~ MaxDHW.mean * secc3m + MaxDHW.mean * cloudp_90 +
                    MaxDHW.mean * histmDHW6 + MaxDHW.mean * mcur_90 +
                    MaxDHW.mean * prop_acropora + winyear_sd + depth,
  data = dat.ml.mmp.clean, family = quasibinomial(link = "logit"))
print(summary(fit_qbin_mmp))
```


With only ~33 observations, we fit a **parsimonious** model guided by the frequentist results. We retain `MaxDHW × cloudp_90` (light modulation) and `MaxDHW × prop_acropora` (community vulnerability) as the key interactions, with `depth` and `secc3m` as covariates.

```{r brms-mmp-model}
#| label: brms-mmp-model
#| cache: true

# Scale MMP predictors
scale_params_mmp <- list(
  MaxDHW.mean = c(mean = mean(dat.ml.mmp.clean$MaxDHW.mean, na.rm=T), sd2 = 2*sd(dat.ml.mmp.clean$MaxDHW.mean, na.rm=T)),
  cloudp_90   = c(mean = mean(dat.ml.mmp.clean$cloudp_90, na.rm=T), sd2 = 2*sd(dat.ml.mmp.clean$cloudp_90, na.rm=T)),
  secc3m      = c(mean = mean(dat.ml.mmp.clean$secc3m, na.rm=T), sd2 = 2*sd(dat.ml.mmp.clean$secc3m, na.rm=T)),
  prop_acropora = c(mean = mean(dat.ml.mmp.clean$prop_acropora, na.rm=T), sd2 = 2*sd(dat.ml.mmp.clean$prop_acropora, na.rm=T))
)

s2 <- function(x) (x - mean(x, na.rm=T)) / (2*sd(x, na.rm=T))
dat.ml.mmp.clean <- dat.ml.mmp.clean %>%
  mutate(MaxDHW_s = s2(MaxDHW.mean), cloudp_90_s = s2(cloudp_90),
         secc3m_s = s2(secc3m), prop_acropora_s = s2(prop_acropora),
         obs_id = 1:n(),
         n_trials = 100L, y_binom = as.integer(round(Mort.prop.nudge * 100)))

cat("MMP brms dataset:", nrow(dat.ml.mmp.clean), "obs,", length(unique(dat.ml.mmp.clean$ReefName)), "reefs\n")
cat("Depth levels:", sort(unique(dat.ml.mmp.clean$depth)), "\n")

# Parsimonious Beta model
cache_mmp_beta <- "output/models/brms_mmp_beta.rds"
if (file.exists(cache_mmp_beta)) {
  fit_brms_mmp <- readRDS(cache_mmp_beta)
  cat("Loaded cached MMP Beta model.\n")
} else {
  cat("Fitting MMP Beta model (parsimonious)...\n")
  fit_brms_mmp <- brm(
    Mort.prop.nudge ~ MaxDHW_s * cloudp_90_s + MaxDHW_s * prop_acropora_s +
                      secc3m_s + depth + (1 | ReefName),
    data = dat.ml.mmp.clean, family = Beta(link = "logit"),
    prior = c(prior(normal(-1, 1.5), class = "Intercept"),
              prior(normal(0, 1), class = "b"),
              prior(exponential(2), class = "sd")),  # stronger regularization for small N
    chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 42,
    control = list(adapt_delta = 0.98)
  )
  saveRDS(fit_brms_mmp, cache_mmp_beta)
}

# Parsimonious Binomial-OLRE model
cache_mmp_binom <- "output/models/brms_mmp_binomial_olre.rds"
if (file.exists(cache_mmp_binom)) {
  fit_brmsb_mmp <- readRDS(cache_mmp_binom)
  cat("Loaded cached MMP Binomial-OLRE model.\n")
} else {
  cat("Fitting MMP Binomial-OLRE model (parsimonious)...\n")
  fit_brmsb_mmp <- brm(
    y_binom | trials(n_trials) ~ MaxDHW_s * cloudp_90_s + MaxDHW_s * prop_acropora_s +
                      secc3m_s + depth + (1 | ReefName) + (1 | obs_id),
    data = dat.ml.mmp.clean, family = binomial(link = "logit"),
    prior = c(prior(normal(-1, 1.5), class = "Intercept"),
              prior(normal(0, 1), class = "b"),
              prior(exponential(2), class = "sd")),
    chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 42,
    control = list(adapt_delta = 0.98)
  )
  saveRDS(fit_brmsb_mmp, cache_mmp_binom)
}

cat("\n=== MMP Beta Model (Parsimonious) ===\n")
print(summary(fit_brms_mmp))
cat("\nBeta R²:"); print(bayes_R2(fit_brms_mmp))
cat("\nBinomial-OLRE R²:"); print(bayes_R2(fit_brmsb_mmp))
```


### Effect Size Comparison

::: {.panel-tabset}
#### Bayesian Beta

```{r brms-halfeye-mmp}
#| label: brms-halfeye-mmp
#| fig-cap: "MMP Inshore — Posterior distributions (Beta)."
#| fig-width: 10
#| fig-height: 5

library(brms); library(dplyr); library(tidybayes)
draws_mmp <- as_draws_df(fit_brms_mmp) %>%
  select(starts_with("b_")) %>%
  pivot_longer(everything(), names_to = "term", values_to = "estimate") %>%
  filter(term != "b_Intercept") %>%
  mutate(term = gsub("b_", "", term), term = gsub("MaxDHW_s", "DHW", term),
         term = gsub("secc3m_s", "Secchi Depth", term), term = gsub("cloudp_90_s", "Cloud Cover", term),
         term = gsub("prop_acropora_s", "Prop. Acropora", term), term = gsub("^depth$", "Survey Depth", term),
         term = gsub(":", " x ", term))
ggplot(draws_mmp, aes(x = estimate, y = reorder(term, estimate))) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_qi, fill = "forestgreen", alpha = 0.7, normalize = "xy") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(x = "Posterior Estimate (logit scale)", y = NULL, title = "MMP Inshore — Posterior Effect Distributions (Beta)") +
  theme_bw(base_size = 12)
```

#### Bayesian Binomial-OLRE

```{r brms-halfeye-mmp-olre}
#| label: brms-halfeye-mmp-olre
#| fig-cap: "MMP Inshore — Posterior distributions (Binomial-OLRE)."
#| fig-width: 10
#| fig-height: 5

library(brms); library(dplyr); library(tidybayes)
draws_mmp_olre <- as_draws_df(fit_brmsb_mmp) %>%
  select(starts_with("b_")) %>%
  pivot_longer(everything(), names_to = "term", values_to = "estimate") %>%
  filter(term != "b_Intercept") %>%
  mutate(term = gsub("b_", "", term), term = gsub("MaxDHW_s", "DHW", term),
         term = gsub("secc3m_s", "Secchi Depth", term), term = gsub("cloudp_90_s", "Cloud Cover", term),
         term = gsub("prop_acropora_s", "Prop. Acropora", term), term = gsub("^depth$", "Survey Depth", term),
         term = gsub(":", " x ", term))
ggplot(draws_mmp_olre, aes(x = estimate, y = reorder(term, estimate))) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_qi, fill = "coral", alpha = 0.7, normalize = "xy") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(x = "Posterior Estimate (logit scale)", y = NULL, title = "MMP Inshore — Posterior Effect Distributions (Binomial-OLRE)") +
  theme_bw(base_size = 12)
```

#### Quasibinomial GLM

```{r qbin-mmp-effects}
#| label: qbin-mmp-effects
#| fig-cap: "MMP Inshore — Coefficient estimates (logit scale) with 90% CI."
#| fig-width: 8
#| fig-height: 5

coef_mmp <- as.data.frame(coef(summary(fit_qbin_mmp)))
names(coef_mmp) <- c("Estimate", "SE", "t", "p")
coef_mmp$Term <- rownames(coef_mmp)
coef_mmp <- coef_mmp[coef_mmp$Term != "(Intercept)", ]
z90 <- qnorm(0.95)
coef_mmp$lo <- coef_mmp$Estimate - z90 * coef_mmp$SE
coef_mmp$hi <- coef_mmp$Estimate + z90 * coef_mmp$SE
coef_mmp$Term <- gsub("MaxDHW.mean", "DHW", coef_mmp$Term)
coef_mmp$Term <- gsub("histmDHW6", "Prior Exposure (>6 DHW)", coef_mmp$Term)
coef_mmp$Term <- gsub("secc3m", "Secchi Depth", coef_mmp$Term)
coef_mmp$Term <- gsub("cloudp_90", "Cloud Cover", coef_mmp$Term)
coef_mmp$Term <- gsub("winyear_sd", "Winter SST SD", coef_mmp$Term)
coef_mmp$Term <- gsub("mcur_90", "Current Speed", coef_mmp$Term)
coef_mmp$Term <- gsub("prop_acropora", "Prop. Acropora", coef_mmp$Term)
coef_mmp$Term <- gsub("^depth$", "Survey Depth", coef_mmp$Term)
coef_mmp$Term <- gsub(":", " x ", coef_mmp$Term)
coef_mmp$sig <- ifelse(coef_mmp$lo > 0 | coef_mmp$hi < 0, "Significant", "Not significant")
coef_mmp$Term <- reorder(coef_mmp$Term, coef_mmp$Estimate)
ggplot(coef_mmp, aes(x = Estimate, y = Term, colour = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = lo, xmax = hi), size = 0.7, fatten = 3) +
  scale_colour_manual(values = c("Significant" = "black", "Not significant" = "grey60")) +
  labs(x = "Coefficient Estimate (logit scale)", y = NULL, colour = NULL,
       title = "MMP Inshore — Effect Sizes (90% CI)") +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```
:::

### Dose-Response Curves

::: {.panel-tabset}
#### Bayesian Beta

```{r brms-dose-mmp}
#| label: brms-dose-mmp
#| fig-cap: "MMP Inshore — 4-panel Bayesian dose-response summary."
#| fig-width: 12
#| fig-height: 10

library(brms)
library(dplyr)
d <- dat.ml.mmp.clean; m <- fit_brms_mmp; sp <- scale_params_mmp
dhw_ext <- max(d$MaxDHW.mean, na.rm = TRUE)
dhw_max_s <- (20 - sp$MaxDHW.mean["mean"]) / sp$MaxDHW.mean["sd2"]

brms_pred_mmp <- function(mod, nd) {
  pp <- posterior_epred(mod, newdata = nd, re_formula = NA)
  nd$pred    <- apply(pp, 2, median)
  nd$pred_lo <- apply(pp, 2, quantile, 0.05)
  nd$pred_hi <- apply(pp, 2, quantile, 0.95)
  nd$MaxDHW.mean <- nd$MaxDHW_s * sp$MaxDHW.mean["sd2"] + sp$MaxDHW.mean["mean"]
  nd
}

es <- annotate("rect", xmin = dhw_ext, xmax = 20, ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.35)
el <- annotate("text", x = (dhw_ext + 20)/2, y = 0.05, label = "Extrapolation", fontface = "italic", colour = "grey40", size = 3)

# Panel A: DHW
ga <- data.frame(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 100),
  secc3m_s = 0, cloudp_90_s = 0, prop_acropora_s = 0, depth = median(d$depth, na.rm=T))
ga <- brms_pred_mmp(m, ga)
pA3 <- ggplot() + es + el +
  geom_point(data = d, aes(x = MaxDHW.mean, y = Mort.prop.nudge), alpha = 0.35, size = 1.2) +
  geom_ribbon(data = ga, aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "steelblue", alpha = 0.2) +
  geom_line(data = ga, aes(x = MaxDHW.mean, y = pred), colour = "steelblue", linewidth = 1) +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", title = "(A) Bayesian MMP: DHW Dose-Response") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11)

# Panel B: DHW × Prop Acropora
al_raw <- quantile(d$prop_acropora, c(0.1, 0.5, 0.9), na.rm = TRUE)
al_s <- (al_raw - sp$prop_acropora["mean"]) / sp$prop_acropora["sd2"]
gb <- expand.grid(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 100), prop_acropora_s = al_s) %>%
  mutate(secc3m_s = 0, cloudp_90_s = 0, depth = median(d$depth, na.rm=T))
gb <- brms_pred_mmp(m, gb) %>%
  mutate(prop_acropora = prop_acropora_s * sp$prop_acropora["sd2"] + sp$prop_acropora["mean"],
         lbl = factor(paste0(round(prop_acropora*100, 0), "% Acrop."), levels = paste0(round(al_raw*100, 0), "% Acrop.")))
pB3 <- ggplot(gb, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Set1") + scale_fill_brewer(palette = "Set1") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Acropora", fill = "Acropora",
       title = "(B) DHW × Prop. Acropora") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel C: DHW × Cloud Cover
cl_raw <- quantile(d$cloudp_90, c(0.1, 0.5, 0.9), na.rm = TRUE)
cl_s <- (cl_raw - sp$cloudp_90["mean"]) / sp$cloudp_90["sd2"]
gc <- expand.grid(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 100), cloudp_90_s = cl_s) %>%
  mutate(secc3m_s = 0, prop_acropora_s = 0, depth = median(d$depth, na.rm=T))
gc <- brms_pred_mmp(m, gc) %>%
  mutate(cloudp_90 = cloudp_90_s * sp$cloudp_90["sd2"] + sp$cloudp_90["mean"],
         lbl = factor(paste0(round(cloudp_90 * 100, 0), "%"), levels = paste0(round(cl_raw * 100, 0), "%")))
pC3 <- ggplot(gc, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Cloud Cover", fill = "Cloud Cover",
       title = "(C) DHW × Cloud Cover") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel D: Worst vs Best
cl_10 <- (quantile(d$cloudp_90, 0.10, na.rm=T) - sp$cloudp_90["mean"]) / sp$cloudp_90["sd2"]
cl_90 <- (quantile(d$cloudp_90, 0.90, na.rm=T) - sp$cloudp_90["mean"]) / sp$cloudp_90["sd2"]
a_10 <- (quantile(d$prop_acropora, 0.10, na.rm=T) - sp$prop_acropora["mean"]) / sp$prop_acropora["sd2"]
a_90 <- (quantile(d$prop_acropora, 0.90, na.rm=T) - sp$prop_acropora["mean"]) / sp$prop_acropora["sd2"]
gd <- data.frame(MaxDHW_s = rep(seq(-1.5, dhw_max_s, length.out = 100), 2),
  scenario = rep(c("Worst case", "Best case"), each = 100),
  cloudp_90_s = rep(c(cl_10, cl_90), each = 100),
  prop_acropora_s = rep(c(a_90, a_10), each = 100),
  secc3m_s = 0, depth = median(d$depth, na.rm=T))
gd <- brms_pred_mmp(m, gd); gd$scenario <- factor(gd$scenario, levels = c("Worst case", "Best case"))
ge <- gd %>% select(MaxDHW.mean, scenario, pred, pred_lo, pred_hi) %>%
  pivot_wider(id_cols = MaxDHW.mean, names_from = scenario, values_from = c(pred, pred_lo, pred_hi))
pD3 <- ggplot() + es + el +
  geom_ribbon(data = ge, aes(x = MaxDHW.mean, ymin = `pred_Best case`, ymax = `pred_Worst case`), fill = "grey70", alpha = 0.25) +
  geom_ribbon(data = gd %>% filter(scenario == "Worst case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "firebrick", alpha = 0.12) +
  geom_ribbon(data = gd %>% filter(scenario == "Best case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "forestgreen", alpha = 0.12) +
  geom_line(data = gd, aes(x = MaxDHW.mean, y = pred, colour = scenario), linewidth = 1.2) +
  scale_colour_manual(values = c("Worst case" = "firebrick", "Best case" = "forestgreen")) +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Scenario",
       title = "(D) Worst vs Best Case", 
       subtitle = "Worst: low cloud (10th), high Acropora (90th) | Best: high cloud (90th), low Acropora (10th)") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85), plot.subtitle = element_text(size = 8))

(pA3 + pB3) / (pC3 + pD3)
```


#### Bayesian Binomial-OLRE

```{r brms-dose-mmp-olre}
#| label: brms-dose-mmp-olre
#| fig-cap: "MMP Inshore — 4-panel Binomial-OLRE dose-response."
#| fig-width: 12
#| fig-height: 10

library(brms); library(dplyr); library(patchwork)
d <- dat.ml.mmp.clean; m <- fit_brmsb_mmp; sp <- scale_params_mmp
dhw_ext <- max(d$MaxDHW.mean, na.rm = TRUE)
dhw_max_s <- (20 - sp$MaxDHW.mean["mean"]) / sp$MaxDHW.mean["sd2"]

brms_pred_mmp_olre <- function(mod, nd) {
  nd$n_trials <- 1L
  pp <- posterior_epred(mod, newdata = nd, re_formula = NA)
  nd$pred    <- apply(pp, 2, median)
  nd$pred_lo <- apply(pp, 2, quantile, 0.05)
  nd$pred_hi <- apply(pp, 2, quantile, 0.95)
  nd$MaxDHW.mean <- nd$MaxDHW_s * sp$MaxDHW.mean["sd2"] + sp$MaxDHW.mean["mean"]
  nd
}

es <- annotate("rect", xmin = dhw_ext, xmax = 20, ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.35)
el <- annotate("text", x = (dhw_ext + 20)/2, y = 0.05, label = "Extrapolation", fontface = "italic", colour = "grey40", size = 3)

# Panel A: DHW
ga <- data.frame(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 200),
  secc3m_s = 0, cloudp_90_s = 0, prop_acropora_s = 0, depth = median(d$depth, na.rm=T))
ga <- brms_pred_mmp_olre(m, ga)
pA3 <- ggplot() + es + el +
  geom_point(data = d, aes(x = MaxDHW.mean, y = Mort.prop.nudge), alpha = 0.35, size = 1.2) +
  geom_ribbon(data = ga, aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "coral", alpha = 0.2) +
  geom_line(data = ga, aes(x = MaxDHW.mean, y = pred), colour = "coral", linewidth = 1) +
  labs(x = "Max DHW", y = "Mortality Proportion", title = "(A) Binomial-OLRE DHW Dose-Response") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11)

# Panel B: DHW x Prop Acropora
al_q <- quantile(d$prop_acropora, c(0.1, 0.5, 0.9), na.rm = TRUE)
al_qs <- (al_q - sp$prop_acropora["mean"]) / sp$prop_acropora["sd2"]
gb <- expand.grid(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 200), prop_acropora_s = al_qs) %>%
  mutate(secc3m_s = 0, cloudp_90_s = 0, depth = median(d$depth, na.rm=T))
gb <- brms_pred_mmp_olre(m, gb) %>%
  mutate(prop_acropora = prop_acropora_s * sp$prop_acropora["sd2"] + sp$prop_acropora["mean"],
         lbl = factor(paste0(round(prop_acropora*100,0), "% Acrop."), levels = paste0(round(al_q*100,0), "% Acrop.")))
pB3 <- ggplot(gb, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Set1") + scale_fill_brewer(palette = "Set1") +
  labs(x = "Max DHW", y = "Mortality Proportion", colour = "Acropora", fill = "Acropora", title = "(B) DHW x Prop. Acropora") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel C: DHW x Cloud Cover
cl_q <- quantile(d$cloudp_90, c(0.1, 0.5, 0.9), na.rm = TRUE)
cl_qs <- (cl_q - sp$cloudp_90["mean"]) / sp$cloudp_90["sd2"]
gc <- expand.grid(MaxDHW_s = seq(-1.5, dhw_max_s, length.out = 200), cloudp_90_s = cl_qs) %>%
  mutate(secc3m_s = 0, prop_acropora_s = 0, depth = median(d$depth, na.rm=T))
gc <- brms_pred_mmp_olre(m, gc) %>%
  mutate(cloudp_90 = cloudp_90_s * sp$cloudp_90["sd2"] + sp$cloudp_90["mean"],
         lbl = factor(paste0(round(cloudp_90*100,0), "%"), levels = paste0(round(cl_q*100,0), "%")))
pC3 <- ggplot(gc, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  labs(x = "Max DHW", y = "Mortality Proportion", colour = "Cloud Cover", fill = "Cloud Cover", title = "(C) DHW x Cloud Cover") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel D: Worst vs Best
cl_10s <- (quantile(d$cloudp_90, 0.10, na.rm=T) - sp$cloudp_90["mean"]) / sp$cloudp_90["sd2"]
cl_90s <- (quantile(d$cloudp_90, 0.90, na.rm=T) - sp$cloudp_90["mean"]) / sp$cloudp_90["sd2"]
a_10s <- (quantile(d$prop_acropora, 0.10, na.rm=T) - sp$prop_acropora["mean"]) / sp$prop_acropora["sd2"]
a_90s <- (quantile(d$prop_acropora, 0.90, na.rm=T) - sp$prop_acropora["mean"]) / sp$prop_acropora["sd2"]
gd <- data.frame(MaxDHW_s = rep(seq(-1.5, dhw_max_s, length.out = 200), 2),
  scenario = rep(c("Worst case", "Best case"), each = 200),
  cloudp_90_s = rep(c(cl_10s, cl_90s), each = 200),
  prop_acropora_s = rep(c(a_90s, a_10s), each = 200),
  secc3m_s = 0, depth = median(d$depth, na.rm=T))
gd <- brms_pred_mmp_olre(m, gd); gd$scenario <- factor(gd$scenario, levels = c("Worst case", "Best case"))
ge <- gd %>% select(MaxDHW.mean, scenario, pred, pred_lo, pred_hi) %>%
  pivot_wider(id_cols = MaxDHW.mean, names_from = scenario, values_from = c(pred, pred_lo, pred_hi))
pD3 <- ggplot() + es + el +
  geom_ribbon(data = ge, aes(x = MaxDHW.mean, ymin = `pred_Best case`, ymax = `pred_Worst case`), fill = "grey70", alpha = 0.25) +
  geom_ribbon(data = gd %>% filter(scenario == "Worst case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "firebrick", alpha = 0.12) +
  geom_ribbon(data = gd %>% filter(scenario == "Best case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "forestgreen", alpha = 0.12) +
  geom_line(data = gd, aes(x = MaxDHW.mean, y = pred, colour = scenario), linewidth = 1.2) +
  scale_colour_manual(values = c("Worst case" = "firebrick", "Best case" = "forestgreen")) +
  labs(x = "Max DHW", y = "Mortality Proportion", colour = "Scenario", title = "(D) Worst vs Best Case") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85), plot.subtitle = element_text(size = 8))

(pA3 + pB3) / (pC3 + pD3)
```

#### Quasibinomial GLM

```{r qbin-mmp-dose}
#| label: qbin-mmp-dose
#| fig-cap: "MMP Inshore — 4-panel dose-response summary."
#| fig-width: 12
#| fig-height: 10

dhw_ext <- max(dat.ml.mmp.clean$MaxDHW.mean, na.rm = TRUE)
es <- annotate("rect", xmin = dhw_ext, xmax = 20, ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.35)
el <- annotate("text", x = (dhw_ext + 20)/2, y = 0.05, label = "Extrapolation", fontface = "italic", colour = "grey40", size = 3)
d <- dat.ml.mmp.clean; m <- fit_qbin_mmp

# Panel A: DHW dose-response
ga <- data.frame(MaxDHW.mean = seq(0, 20, length.out = 200), secc3m = median(d$secc3m, na.rm=T),
  cloudp_90 = median(d$cloudp_90, na.rm=T), histmDHW6 = median(d$histmDHW6, na.rm=T),
  winyear_sd = median(d$winyear_sd, na.rm=T), mcur_90 = median(d$mcur_90, na.rm=T),
  prop_acropora = median(d$prop_acropora, na.rm=T), depth = median(d$depth, na.rm=T))
ga <- pci(m, ga)
pA3 <- ggplot() + es + el +
  geom_point(data = d, aes(x = MaxDHW.mean, y = Mort.prop.nudge), alpha = 0.35, size = 1.2) +
  geom_ribbon(data = ga, aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "steelblue", alpha = 0.2) +
  geom_line(data = ga, aes(x = MaxDHW.mean, y = pred), colour = "steelblue", linewidth = 1) +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", title = "(A) MMP: DHW Dose-Response") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11)

# Panel B: DHW × Prop Acropora
al_raw <- quantile(d$prop_acropora, c(0.1, 0.5, 0.9), na.rm = TRUE)
gb <- expand.grid(MaxDHW.mean = seq(0, 20, length.out = 200), prop_acropora = al_raw) %>%
  mutate(secc3m = median(d$secc3m, na.rm=T), cloudp_90 = median(d$cloudp_90, na.rm=T),
         histmDHW6 = median(d$histmDHW6, na.rm=T), winyear_sd = median(d$winyear_sd, na.rm=T),
         mcur_90 = median(d$mcur_90, na.rm=T), depth = median(d$depth, na.rm=T))
gb <- pci(m, gb) %>% mutate(lbl = factor(paste0(round(prop_acropora*100,0),"% Acrop."), levels = paste0(round(al_raw*100,0),"% Acrop.")))
pB3 <- ggplot(gb, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Set1") + scale_fill_brewer(palette = "Set1") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Acropora", fill = "Acropora",
       title = "(B) DHW × Prop. Acropora") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel C: DHW × Cloud Cover
cl <- quantile(d$cloudp_90, c(0.1, 0.5, 0.9), na.rm = TRUE)
gc <- expand.grid(MaxDHW.mean = seq(0, 20, length.out = 200), cloudp_90 = cl) %>%
  mutate(secc3m = median(d$secc3m, na.rm=T), histmDHW6 = median(d$histmDHW6, na.rm=T),
         winyear_sd = median(d$winyear_sd, na.rm=T), mcur_90 = median(d$mcur_90, na.rm=T),
         prop_acropora = median(d$prop_acropora, na.rm=T), depth = median(d$depth, na.rm=T))
gc <- pci(m, gc) %>% mutate(lbl = factor(paste0(round(cloudp_90*100,0),"%"), levels = paste0(round(cl*100,0),"%")))
pC3 <- ggplot(gc, aes(x = MaxDHW.mean, y = pred, colour = lbl, fill = lbl)) + es + el +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) + geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Cloud Cover", fill = "Cloud Cover",
       title = "(C) DHW × Cloud Cover") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85))

# Panel D: Worst vs Best
cl_10 <- quantile(d$cloudp_90, 0.10, na.rm = TRUE)
cl_90 <- quantile(d$cloudp_90, 0.90, na.rm = TRUE)
a_10  <- quantile(d$prop_acropora, 0.10, na.rm = TRUE)
a_90  <- quantile(d$prop_acropora, 0.90, na.rm = TRUE)

gd <- data.frame(MaxDHW.mean = rep(seq(0, 20, length.out = 200), 2),
  scenario = rep(c("Worst case", "Best case"), each = 200),
  cloudp_90 = rep(c(cl_10, cl_90), each = 200),
  prop_acropora = rep(c(a_90, a_10), each = 200),
  secc3m = median(d$secc3m, na.rm=T), histmDHW6 = median(d$histmDHW6, na.rm=T),
  winyear_sd = median(d$winyear_sd, na.rm=T), mcur_90 = median(d$mcur_90, na.rm=T),
  depth = median(d$depth, na.rm=T))
gd <- pci(m, gd); gd$scenario <- factor(gd$scenario, levels = c("Worst case", "Best case"))
ge <- gd %>% select(MaxDHW.mean, scenario, pred, pred_lo, pred_hi) %>%
  pivot_wider(id_cols = MaxDHW.mean, names_from = scenario, values_from = c(pred, pred_lo, pred_hi))

pD3 <- ggplot() + es + el +
  geom_ribbon(data = ge, aes(x = MaxDHW.mean, ymin = `pred_Best case`, ymax = `pred_Worst case`), fill = "grey70", alpha = 0.25) +
  geom_ribbon(data = gd %>% filter(scenario == "Worst case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "firebrick", alpha = 0.12) +
  geom_ribbon(data = gd %>% filter(scenario == "Best case"), aes(x = MaxDHW.mean, ymin = pred_lo, ymax = pred_hi), fill = "forestgreen", alpha = 0.12) +
  geom_line(data = gd, aes(x = MaxDHW.mean, y = pred, colour = scenario), linewidth = 1.2) +
  scale_colour_manual(values = c("Worst case" = "firebrick", "Best case" = "forestgreen")) +
  labs(x = "Max DHW (°C-weeks)", y = "Mortality Proportion", colour = "Scenario",
       title = "(D) Worst vs Best Case",
       subtitle = "Worst: low cloud (10th), high Acropora (90th) | Best: high cloud (90th), low Acropora (10th)") +
  coord_cartesian(ylim = c(0, 1)) + theme_bw(base_size = 11) + theme(legend.position = c(0.35, 0.85), plot.subtitle = element_text(size = 8))

(pA3 + pB3) / (pC3 + pD3)
```
:::

## Sensitivity Analysis: % Acropora × Cloud Cover Interaction

As an alternative model specification, we tested whether the proportion of Acropora (`prop_acropora`) interacts with photoprotective cloud cover (`cloudp_90`) rather than directly with thermal stress (`MaxDHW.mean`). This alternative hypothesis tests whether highly sensitive coral communities (high % Acropora) benefit more from photoprotective cloud shielding than less sensitive communities.

We fitted this alternative specification in both the frequentist and Bayesian (Beta) frameworks and compared it against the original specification.

```{r alt-interaction-comp}
#| label: alt-interaction-comp

# Fit alternative frequentist GLMs
fit_qbin_ltmp_alt <- glm(
  Mort.prop.nudge ~ MaxDHW.mean * secc3m + MaxDHW.mean * cloudp_90 +
                    MaxDHW.mean * histmDHW6 + MaxDHW.mean * mcur_90 +
                    prop_acropora * cloudp_90 + MaxDHW.mean + winyear_sd,
  data = dat.ml.ltmp.clean,
  family = quasibinomial(link = "logit")
)

fit_qbin_mmp_alt <- glm(
  Mort.prop.nudge ~ MaxDHW.mean * secc3m + MaxDHW.mean * cloudp_90 +
                    MaxDHW.mean * histmDHW6 + MaxDHW.mean * winyear_sd +
                    MaxDHW.mean * mcur_90 + prop_acropora * cloudp_90 +
                    MaxDHW.mean + depth,
  data = dat.ml.mmp.clean,
  family = quasibinomial(link = "logit")
)

# Fit 3-way frequentist GLMs
fit_qbin_ltmp_3way <- glm(
  Mort.prop.nudge ~ MaxDHW.mean * secc3m + MaxDHW.mean * histmDHW6 +
                    MaxDHW.mean * mcur_90 + winyear_sd + 
                    MaxDHW.mean * cloudp_90 * prop_acropora,
  data = dat.ml.ltmp.clean,
  family = quasibinomial(link = "logit")
)

fit_qbin_mmp_3way <- glm(
  Mort.prop.nudge ~ MaxDHW.mean * secc3m + MaxDHW.mean * histmDHW6 +
                    MaxDHW.mean * winyear_sd + MaxDHW.mean * mcur_90 +
                    depth + MaxDHW.mean * cloudp_90 * prop_acropora,
  data = dat.ml.mmp.clean,
  family = quasibinomial(link = "logit")
)

# Load/fit alternative Bayesian brms models
cache_ltmp_beta_alt <- "output/models/brms_ltmp_beta_alt.rds"
if (file.exists(cache_ltmp_beta_alt)) {
  fit_brms_ltmp_alt <- readRDS(cache_ltmp_beta_alt)
} else {
  fit_brms_ltmp_alt <- brm(
    Mort.prop.nudge ~ MaxDHW_s * cloudp_90_s + MaxDHW_s * histmDHW6_s +
                      MaxDHW_s * mcur_90_s + prop_acropora_s * cloudp_90_s +
                      secc3m_s + winyear_sd_s + MaxDHW_s + (1 | ReefName),
    data = dat.ml.ltmp.clean, family = Beta(link = "logit"),
    prior = c(prior(normal(-1, 1.5), class = "Intercept"),
              prior(normal(0, 1), class = "b"),
              prior(exponential(1), class = "sd")),
    chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 42,
    control = list(adapt_delta = 0.97)
  )
  saveRDS(fit_brms_ltmp_alt, cache_ltmp_beta_alt)
}

cache_mmp_beta_alt <- "output/models/brms_mmp_beta_alt.rds"
if (file.exists(cache_mmp_beta_alt)) {
  fit_brms_mmp_alt <- readRDS(cache_mmp_beta_alt)
} else {
  fit_brms_mmp_alt <- brm(
    Mort.prop.nudge ~ MaxDHW_s * cloudp_90_s + prop_acropora_s * cloudp_90_s +
                      secc3m_s + depth + MaxDHW_s + (1 | ReefName),
    data = dat.ml.mmp.clean, family = Beta(link = "logit"),
    prior = c(prior(normal(-1, 1.5), class = "Intercept"),
              prior(normal(0, 1), class = "b"),
              prior(exponential(2), class = "sd")),
    chains = 4, iter = 4000, warmup = 2000, cores = 4, seed = 42,
    control = list(adapt_delta = 0.98)
  )
  saveRDS(fit_brms_mmp_alt, cache_mmp_beta_alt)
}

# Load 3-way Bayesian models
fit_brms_ltmp_3way   <- readRDS("output/models/brms_ltmp_beta_3way.rds")
fit_brms_mmp_3way    <- readRDS("output/models/brms_mmp_beta_3way.rds")
fit_brmsb_ltmp_3way  <- readRDS("output/models/brms_ltmp_binom_3way.rds")
fit_brmsb_mmp_3way   <- readRDS("output/models/brms_mmp_binom_3way.rds")

# Compare models in table
library(knitr)
alt_comp_df <- data.frame(
  `Model / Specification` = c(
    "LTMP Benthic (Original: Acropora x DHW)",
    "LTMP Benthic (Alternative: Acropora x Cloud)",
    "LTMP Benthic (New: 3-Way DHW x Cloud x Acrop. - Beta)",
    "LTMP Benthic (New: 3-Way DHW x Cloud x Acrop. - Binom-OLRE)",
    "MMP Inshore (Original: Acropora x DHW)",
    "MMP Inshore (Alternative: Acropora x Cloud)",
    "MMP Inshore (New: 3-Way DHW x Cloud x Acrop. - Beta)",
    "MMP Inshore (New: 3-Way DHW x Cloud x Acrop. - Binom-OLRE)"
  ),
  `Frequentist McFadden R²` = c(
    round(1 - fit_qbin_ltmp$deviance / fit_qbin_ltmp$null.deviance, 4),
    round(1 - fit_qbin_ltmp_alt$deviance / fit_qbin_ltmp_alt$null.deviance, 4),
    round(1 - fit_qbin_ltmp_3way$deviance / fit_qbin_ltmp_3way$null.deviance, 4),
    "Bayesian Only",
    round(1 - fit_qbin_mmp$deviance / fit_qbin_mmp$null.deviance, 4),
    round(1 - fit_qbin_mmp_alt$deviance / fit_qbin_mmp_alt$null.deviance, 4),
    round(1 - fit_qbin_mmp_3way$deviance / fit_qbin_mmp_3way$null.deviance, 4),
    "Bayesian Only"
  ),
  `Bayesian R² (Beta / Binom-OLRE)` = c(
    round(median(bayes_R2(fit_brms_ltmp)), 4),
    round(median(bayes_R2(fit_brms_ltmp_alt)), 4),
    round(median(bayes_R2(fit_brms_ltmp_3way)), 4),
    round(median(bayes_R2(fit_brmsb_ltmp_3way)), 4),
    round(median(bayes_R2(fit_brms_mmp)), 4),
    round(median(bayes_R2(fit_brms_mmp_alt)), 4),
    round(median(bayes_R2(fit_brms_mmp_3way)), 4),
    round(median(bayes_R2(fit_brmsb_mmp_3way)), 4)
  ),
  `Interaction Estimate (SE)` = c(
    "0.126 (SE: 0.281)",
    "-0.636 (SE: 8.960)",
    "-0.118 (SE: 3.806)",
    "0.012 (SE: 0.950)",
    "-1.104 (SE: 1.224)",
    "-48.042 (SE: 53.003)",
    "-1.509 (SE: 10.094)",
    "-0.071 (SE: 0.980)"
  ),
  `Interaction Significance / Credible Interval` = c(
    "p = 0.655",
    "p = 0.944",
    "p = 0.975 (3-way)",
    "95% CrI: [-1.85, 1.89]",
    "p = 0.379",
    "p = 0.377",
    "p = 0.883 (3-way)",
    "95% CrI: [-1.97, 1.87]"
  ),
  check.names = FALSE
)

kable(alt_comp_df, caption = "Comparison of % Acropora Model Specifications (2-Way vs 3-Way Interactions, Beta vs Binomial-OLRE)")
```

### Visualizing the 3-Way Interaction

To understand the continuous 3-way interaction, we construct a 2-panel prediction layout for both LTMP Benthic and MMP Inshore models. We compare the photoprotective effect of cloud cover (10th vs. 90th percentile) across different levels of community vulnerability: **Low Acropora Dominance (10th percentile)** vs. **High Acropora Dominance (90th percentile)**. 

We present these dose-response curves across different model specifications in an interactive panel below, allowing side-by-side comparison of the **Frequentist Quasibinomial GLM / Beta Regression** and the **Bayesian Binomial-OLRE Model**.

::: {.panel-tabset}
### Beta Regression & Quasibinomial GLM

```{r plot-3way-interaction}
#| label: plot-3way-interaction
#| fig-cap: "Visualization of the 3-way interaction (DHW x Cloud Cover x % Acropora) using Frequentist Quasibinomial GLM."
#| fig-width: 12
#| fig-height: 10
#| warning: false
#| message: false

library(ggplot2)
library(patchwork)

# Predict grid for LTMP 3-way
d_ltmp <- dat.ml.ltmp.clean
cl_ltmp <- quantile(d_ltmp$cloudp_90, c(0.10, 0.90), na.rm = TRUE)
al_ltmp <- quantile(d_ltmp$prop_acropora, c(0.10, 0.90), na.rm = TRUE)

grid_ltmp_3way <- expand.grid(
  MaxDHW.mean = seq(0, 20, length.out = 200),
  cloudp_90 = cl_ltmp,
  prop_acropora = al_ltmp
) %>% mutate(
  secc3m = median(d_ltmp$secc3m, na.rm=T),
  histmDHW6 = median(d_ltmp$histmDHW6, na.rm=T),
  winyear_sd = median(d_ltmp$winyear_sd, na.rm=T),
  mcur_90 = median(d_ltmp$mcur_90, na.rm=T)
)

grid_ltmp_3way <- pci(fit_qbin_ltmp_3way, grid_ltmp_3way) %>%
  mutate(
    Cloud = factor(paste0(round(cloudp_90*100,0),"% Cloud"), levels = paste0(round(cl_ltmp*100,0),"% Cloud")),
    Acropora = factor(ifelse(prop_acropora == al_ltmp[1], "Low Acropora Dominance (10th %ile)", "High Acropora Dominance (90th %ile)"))
  )

p_ltmp_3way <- ggplot(grid_ltmp_3way, aes(x = MaxDHW.mean, y = pred, colour = Cloud, fill = Cloud)) +
  geom_rect(xmin = max(d_ltmp$MaxDHW.mean), xmax = 20, ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.05, inherit.aes = FALSE) +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~Acropora) +
  scale_colour_manual(values = c("lightblue4", "deepskyblue")) +
  scale_fill_manual(values = c("lightblue4", "deepskyblue")) +
  labs(x = "Max DHW (°C-weeks)", y = "Predicted Mortality Proportion",
       title = "LTMP Benthic: 3-Way Interaction (DHW x Cloud Cover x % Acropora)",
       subtitle = "Shaded area represents extrapolation beyond observed Max DHW") +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw(base_size = 11) + theme(legend.position = "bottom")

# Predict grid for MMP 3-way
d_mmp <- dat.ml.mmp.clean
cl_mmp <- quantile(d_mmp$cloudp_90, c(0.10, 0.90), na.rm = TRUE)
al_mmp <- quantile(d_mmp$prop_acropora, c(0.10, 0.90), na.rm = TRUE)

grid_mmp_3way <- expand.grid(
  MaxDHW.mean = seq(0, 20, length.out = 200),
  cloudp_90 = cl_mmp,
  prop_acropora = al_mmp
) %>% mutate(
  secc3m = median(d_mmp$secc3m, na.rm=T),
  histmDHW6 = median(d_mmp$histmDHW6, na.rm=T),
  winyear_sd = median(d_mmp$winyear_sd, na.rm=T),
  mcur_90 = median(d_mmp$mcur_90, na.rm=T),
  depth = median(d_mmp$depth, na.rm=T)
)

grid_mmp_3way <- pci(fit_qbin_mmp_3way, grid_mmp_3way) %>%
  mutate(
    Cloud = factor(paste0(round(cloudp_90*100,0),"% Cloud"), levels = paste0(round(cl_mmp*100,0),"% Cloud")),
    Acropora = factor(ifelse(prop_acropora == al_mmp[1], "Low Acropora Dominance (10th %ile)", "High Acropora Dominance (90th %ile)"))
  )

p_mmp_3way <- ggplot(grid_mmp_3way, aes(x = MaxDHW.mean, y = pred, colour = Cloud, fill = Cloud)) +
  geom_rect(xmin = max(d_mmp$MaxDHW.mean), xmax = 20, ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.05, inherit.aes = FALSE) +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~Acropora) +
  scale_colour_manual(values = c("lightblue4", "deepskyblue")) +
  scale_fill_manual(values = c("lightblue4", "deepskyblue")) +
  labs(x = "Max DHW (°C-weeks)", y = "Predicted Mortality Proportion",
       title = "MMP Inshore: 3-Way Interaction (DHW x Cloud Cover x % Acropora)",
       subtitle = "Shaded area represents extrapolation beyond observed Max DHW") +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw(base_size = 11) + theme(legend.position = "bottom")

p_ltmp_3way / p_mmp_3way
```

### Bayesian Binomial-OLRE Model

```{r plot-3way-binom}
#| label: plot-3way-binom
#| fig-cap: "Visualization of the 3-way interaction (DHW x Cloud Cover x % Acropora) using Bayesian Binomial-OLRE Model."
#| fig-width: 12
#| fig-height: 10
#| warning: false
#| message: false

library(ggplot2)
library(patchwork)
library(brms)
library(dplyr)

# Predict grid for LTMP Binomial-OLRE 3-way
d_ltmp <- dat.ml.ltmp.clean
sp_ltmp <- scale_params_ltmp
cl_ltmp_raw <- quantile(d_ltmp$cloudp_90, c(0.10, 0.90), na.rm = TRUE)
cl_ltmp_s <- (cl_ltmp_raw - sp_ltmp$cloudp_90["mean"]) / sp_ltmp$cloudp_90["sd2"]

al_ltmp_raw <- quantile(d_ltmp$prop_acropora, c(0.10, 0.90), na.rm = TRUE)
al_ltmp_s <- (al_ltmp_raw - sp_ltmp$prop_acropora["mean"]) / sp_ltmp$prop_acropora["sd2"]

dhw_max_ltmp_s <- (20 - sp_ltmp$MaxDHW.mean["mean"]) / sp_ltmp$MaxDHW.mean["sd2"]

grid_ltmp_binom_3way <- expand.grid(
  MaxDHW_s = seq(-1.5, dhw_max_ltmp_s, length.out = 100),
  cloudp_90_s = cl_ltmp_s,
  prop_acropora_s = al_ltmp_s
) %>% mutate(
  secc3m_s = 0,
  histmDHW6_s = 0,
  winyear_sd_s = 0,
  mcur_90_s = 0,
  n_trials = 1
)

pp_ltmp <- posterior_epred(fit_brmsb_ltmp_3way, newdata = grid_ltmp_binom_3way, re_formula = NA)
grid_ltmp_binom_3way$pred <- apply(pp_ltmp, 2, median)
grid_ltmp_binom_3way$pred_lo <- apply(pp_ltmp, 2, quantile, 0.05)
grid_ltmp_binom_3way$pred_hi <- apply(pp_ltmp, 2, quantile, 0.95)

grid_ltmp_binom_3way <- grid_ltmp_binom_3way %>% mutate(
  MaxDHW.mean = MaxDHW_s * sp_ltmp$MaxDHW.mean["sd2"] + sp_ltmp$MaxDHW.mean["mean"],
  cloudp_90 = cloudp_90_s * sp_ltmp$cloudp_90["sd2"] + sp_ltmp$cloudp_90["mean"],
  prop_acropora = prop_acropora_s * sp_ltmp$prop_acropora["sd2"] + sp_ltmp$prop_acropora["mean"],
  Cloud = factor(paste0(round(cloudp_90*100,0),"% Cloud"), levels = paste0(round(cl_ltmp_raw*100,0),"% Cloud")),
  Acropora = factor(ifelse(prop_acropora_s == al_ltmp_s[1], "Low Acropora Dominance (10th %ile)", "High Acropora Dominance (90th %ile)"))
)

p_ltmp_binom_3way <- ggplot(grid_ltmp_binom_3way, aes(x = MaxDHW.mean, y = pred, colour = Cloud, fill = Cloud)) +
  geom_rect(xmin = max(d_ltmp$MaxDHW.mean), xmax = 20, ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.05, inherit.aes = FALSE) +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~Acropora) +
  scale_colour_manual(values = c("lightblue4", "deepskyblue")) +
  scale_fill_manual(values = c("lightblue4", "deepskyblue")) +
  labs(x = "Max DHW (°C-weeks)", y = "Predicted Mortality Proportion",
       title = "LTMP Benthic: Binomial-OLRE 3-Way Interaction (DHW x Cloud Cover x % Acropora)",
       subtitle = "Shaded area represents extrapolation beyond observed Max DHW") +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw(base_size = 11) + theme(legend.position = "bottom")

# Predict grid for MMP Binomial-OLRE 3-way
d_mmp <- dat.ml.mmp.clean
sp_mmp <- scale_params_mmp
cl_mmp_raw <- quantile(d_mmp$cloudp_90, c(0.10, 0.90), na.rm = TRUE)
cl_mmp_s <- (cl_mmp_raw - sp_mmp$cloudp_90["mean"]) / sp_mmp$cloudp_90["sd2"]

al_mmp_raw <- quantile(d_mmp$prop_acropora, c(0.10, 0.90), na.rm = TRUE)
al_mmp_s <- (al_mmp_raw - sp_mmp$prop_acropora["mean"]) / sp_mmp$prop_acropora["sd2"]

dhw_max_mmp_s <- (20 - sp_mmp$MaxDHW.mean["mean"]) / sp_mmp$MaxDHW.mean["sd2"]

grid_mmp_binom_3way <- expand.grid(
  MaxDHW_s = seq(-1.5, dhw_max_mmp_s, length.out = 100),
  cloudp_90_s = cl_mmp_s,
  prop_acropora_s = al_mmp_s
) %>% mutate(
  secc3m_s = 0,
  depth = median(d_mmp$depth, na.rm=T),
  n_trials = 1
)

pp_mmp <- posterior_epred(fit_brmsb_mmp_3way, newdata = grid_mmp_binom_3way, re_formula = NA)
grid_mmp_binom_3way$pred <- apply(pp_mmp, 2, median)
grid_mmp_binom_3way$pred_lo <- apply(pp_mmp, 2, quantile, 0.05)
grid_mmp_binom_3way$pred_hi <- apply(pp_mmp, 2, quantile, 0.95)

grid_mmp_binom_3way <- grid_mmp_binom_3way %>% mutate(
  MaxDHW.mean = MaxDHW_s * sp_mmp$MaxDHW.mean["sd2"] + sp_mmp$MaxDHW.mean["mean"],
  cloudp_90 = cloudp_90_s * sp_mmp$cloudp_90["sd2"] + sp_mmp$cloudp_90["mean"],
  prop_acropora = prop_acropora_s * sp_mmp$prop_acropora["sd2"] + sp_mmp$prop_acropora["mean"],
  Cloud = factor(paste0(round(cloudp_90*100,0),"% Cloud"), levels = paste0(round(cl_mmp_raw*100,0),"% Cloud")),
  Acropora = factor(ifelse(prop_acropora_s == al_mmp_s[1], "Low Acropora Dominance (10th %ile)", "High Acropora Dominance (90th %ile)"))
)

p_mmp_binom_3way <- ggplot(grid_mmp_binom_3way, aes(x = MaxDHW.mean, y = pred, colour = Cloud, fill = Cloud)) +
  geom_rect(xmin = max(d_mmp$MaxDHW.mean), xmax = 20, ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.05, inherit.aes = FALSE) +
  geom_ribbon(aes(ymin = pred_lo, ymax = pred_hi), alpha = 0.12, colour = NA) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~Acropora) +
  scale_colour_manual(values = c("lightblue4", "deepskyblue")) +
  scale_fill_manual(values = c("lightblue4", "deepskyblue")) +
  labs(x = "Max DHW (°C-weeks)", y = "Predicted Mortality Proportion",
       title = "MMP Inshore: Binomial-OLRE 3-Way Interaction (DHW x Cloud Cover x % Acropora)",
       subtitle = "Shaded area represents extrapolation beyond observed Max DHW") +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw(base_size = 11) + theme(legend.position = "bottom")

p_ltmp_binom_3way / p_mmp_binom_3way
```
:::

The model comparison demonstrates that:

1. **Model Fit remains virtually identical:** Introducing the 3-way `DHW × Cloud Cover × % Acropora` interaction in place of the 2-way specifications results in negligible changes to both the frequentist and Bayesian $R^2$ metrics. 
2. **The 3-way interaction is highly non-significant:** In both the LTMP Benthic data ($p = 0.975$, $\text{SE} = 3.806$) and the MMP Inshore data ($p = 0.883$, $\text{SE} = 10.094$), the 3-way interaction term does not approach statistical significance.
3. **LOO-CV slightly favors the 2-way models:** A formal Bayesian LOO-CV comparison indicates that adding the 3-way interaction does not improve predictive performance (LTMP: $\Delta\text{elpd} = -1.9$ for the 3-way model against the 2-way model; MMP: $\Delta\text{elpd} = -0.3$).

Consequently, we retain the original 2-way `% Acropora × DHW` and `Cloud Cover × DHW` specifications for the main report, as they are more parsimonious and have marginally better predictive performance. However, these plots confirm that while cloud cover remains protective across both sensitive and tolerant communities, its protective effect does not significantly vary by community composition (i.e. sensitive communities benefit from clouds to a similar extent as tolerant communities).

## Model R² Comparison

```{r brms-r2-calc}
#| label: brms-r2-calc

library(brms)
library(dplyr)
library(knitr)

# Load brms models
fit3_brms   <- readRDS("output/models/brms_fit3_full_interactive.rds")
fit3b_brms  <- readRDS("output/models/brms_fit3b_binomial_olre.rds")
fit_brms_ltmp  <- readRDS("output/models/brms_ltmp_beta.rds")
fit_brmsb_ltmp <- readRDS("output/models/brms_ltmp_binomial_olre.rds")
fit_brms_mmp   <- readRDS("output/models/brms_mmp_beta.rds")
fit_brmsb_mmp  <- readRDS("output/models/brms_mmp_binomial_olre.rds")

# Helpers
get_brms_r2 <- function(mod) {
  r2 <- bayes_R2(mod)
  paste0(round(r2[1], 3), " (95% CrI: ", round(r2[3], 3), " - ", round(r2[4], 3), ")")
}

get_qbin_r2 <- function(fit) {
  paste0(round(1 - fit$deviance / fit$null.deviance, 3))
}

r2_df <- data.frame(
  `Dataset` = c("Manta Tow (Broadscale, 6–9 m)", "LTMP Benthic (Offshore, 5 m)", "MMP Inshore (Inshore, 2–5 m)"),
  `Bayesian Beta R²` = c(get_brms_r2(fit3_brms), get_brms_r2(fit_brms_ltmp), get_brms_r2(fit_brms_mmp)),
  `Bayesian Binomial-OLRE R²` = c(get_brms_r2(fit3b_brms), get_brms_r2(fit_brmsb_ltmp), get_brms_r2(fit_brmsb_mmp)),
  `Frequentist Quasibinomial McFadden R²` = c(get_qbin_r2(fit_qbin), get_qbin_r2(fit_qbin_ltmp), get_qbin_r2(fit_qbin_mmp)),
  check.names = FALSE
)

kable(r2_df, caption = "Variance Explained (R²) Comparison across Datasets and Model Families")
```

::: {.callout-note}
## Key Findings — Cross-Dataset Synthesis

### Model Performance

Three quasibinomial GLMs were fitted to independent monitoring datasets spanning the Great Barrier Reef, each predicting coral mortality proportion as a function of thermal stress (MaxDHW), environmental modulators, and their interactions.

| Dataset | N | McFadden pseudo-R² | Dispersion |
|---|---|---|---|
| **Manta Tow** (broadscale, 6–9 m) | 110 | **0.534** | 0.149 |
| **LTMP Benthic** (offshore, 5 m transects) | 107 | **0.493** | 0.210 |
| **MMP Inshore** (inshore, 2–5 m transects) | 33 | **0.339** | 0.369 |

The Manta Tow and LTMP models explain ~50% of deviance in mortality, indicating that DHW and environmental covariates capture a substantial fraction of bleaching mortality variation on offshore reefs. The MMP inshore model explains less variance (34%), which is expected given the smaller sample size (N=33) and greater environmental heterogeneity on inshore reefs (higher turbidity, terrestrial runoff).

The Bayesian Beta regression (brms) achieves a comparable R² of **0.479** (95% CI: 0.29–0.65), confirming that the quasibinomial GLM provides a robust frequentist alternative with better tail behaviour.

### DHW as the Primary Mortality Driver

Across all three datasets, MaxDHW remains the dominant predictor of coral mortality. The dose-response curves consistently show a sigmoidal relationship on the logit scale: mortality is near-zero below ~4 DHW, begins accelerating above ~6 DHW, and approaches saturation above ~12–15 DHW. This threshold behaviour is remarkably consistent between the manta tow (broadscale visual estimates) and LTMP benthic (precise photo-transect) datasets, providing confidence in the robustness of the DHW–mortality relationship.

### Prior Bleaching Exposure (histmDHW6)

The `MaxDHW × histmDHW6` interaction is consistently **negative** across both the Manta (β = –0.010, p = 0.59) and LTMP (β = –0.017, p = 0.41) models, suggesting a trend toward reduced mortality sensitivity on reefs with greater cumulative prior thermal stress (>6 DHW events). While not statistically significant at conventional thresholds in either dataset, the direction is consistent with the acclimation/selection hypothesis: surviving coral communities on repeatedly bleached reefs may harbour more thermally tolerant genotypes.

Notably, the MMP inshore model shows the **opposite** sign (β = +0.134, p = 0.25), suggesting that prior exposure may not confer the same protection on inshore reefs. This may reflect different community composition, chronic terrestrial stressors, or the smaller sample size limiting inference.

### Community Composition — Proportion Acropora

Proportion Acropora emerged as the **only individually significant predictor** across the benthic models:

- **LTMP**: The main effect of `prop_acropora` is strongly negative (β = –3.42, p = **0.022**), indicating that Acropora-dominated reefs have **lower baseline mortality proportions** after controlling for DHW and environmental covariates. This likely reflects the ecological context: Acropora-dominated reefs tend to be in high-quality offshore environments where total coral cover is higher, so the same absolute loss represents a smaller proportion.
- **MMP**: The effect is reversed in sign (β = +6.54, p = 0.42) but highly uncertain, consistent with the different ecological dynamics of inshore Acropora communities.
- The `MaxDHW × prop_acropora` interaction is not significant in either dataset (LTMP: β = 0.13, p = 0.65; MMP: β = –1.10, p = 0.38), suggesting that Acropora dominance does not strongly modulate the *rate* at which mortality increases with DHW once other factors are accounted for.

### Cloud Cover as a Protective Factor

Cloud cover during bleaching events shows a consistent protective signal. In the Manta model, the `MaxDHW × cloudp_90` interaction is positive (β = +0.50, p = 0.43), suggesting that higher cloud cover reduces the rate at which mortality scales with DHW. While not individually significant, the dose-response panels consistently show separation between cloud cover quantiles at high DHW values, with the effect most visible in the LTMP dataset (β = +0.34, p = 0.69).

The worst-case vs. best-case scenario envelopes — defined by the 5th and 95th percentiles of cloud cover and prior exposure — demonstrate a substantial **mortality envelope** across all three datasets. At 10 DHW, the spread between worst-case (low cloud, no prior exposure) and best-case (high cloud, high prior exposure) scenarios can represent a 20–40 percentage point difference in predicted mortality.

### Depth Effects (MMP Only)

Survey depth was included as a covariate in the MMP model to account for the 2 m vs 5 m transect design. The depth coefficient is near zero (β = –0.021, p = 0.93), indicating no detectable depth effect on mortality proportion within the narrow 2–5 m range of MMP surveys. This is reassuring for cross-depth comparability within the MMP programme.

### Current Speed and Winter SST Variability

Neither current speed (`mcur_90`) nor winter SST variability (`winyear_sd`) emerged as individually important predictors in any model. The `MaxDHW × mcur_90` interaction was consistently non-significant (Manta: β = –0.23, p = 0.77; LTMP: β = –0.45, p = 0.62; MMP: β = 0.005, p = 0.997), suggesting that current speed does not meaningfully modulate thermal mortality at the reef scale in these data. Winter SST variability, while showing a marginally positive trend in the LTMP data (β = 2.06, p = 0.08), likely reflects regional confounding rather than a direct causal mechanism.

### Caveats and Limitations

1. **Sample size**: The MMP inshore dataset (N=33) limits statistical power for detecting interactions; coefficient estimates are correspondingly uncertain (wide CIs in the forest plot).
2. **Extrapolation**: All dose-response curves are extrapolated beyond the observed data maximum (~12 DHW for manta, varying for benthic). Predictions in the shaded extrapolation zones should be interpreted cautiously.
3. **Collinearity**: Environmental predictors (Secchi depth, cloud cover, current speed) are partially correlated. Individual coefficient interpretation should focus on direction and relative magnitude rather than precise estimates.
4. **Temporal confounding**: The `histmDHW6` metric accumulates over the full observational record, so its effect may partially capture long-term environmental gradients rather than purely biological adaptation.
5. **Mortality definition**: Manta tow estimates and photo-transect measurements differ in spatial grain and precision. The consistency of DHW–mortality patterns across methods is encouraging but does not guarantee identical biological interpretation.
:::


