---
title: "Spatially independent early-event bleaching update"
subtitle: "Aerial and RHIS evidence available before post-event mortality assessment"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    embed-resources: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(knitr)
})
root <- normalizePath('../..', winslash = '/', mustWork = TRUE)
out <- file.path(root, 'output', 'spatial_early_bleaching_update')
fig_dir <- file.path(root, 'output', 'fig')
comparison <- read_csv(file.path(out, 'model_comparison.csv'), show_col_types = FALSE)
events <- read_csv(file.path(out, 'event_metrics.csv'), show_col_types = FALSE)
sources <- read_csv(file.path(out, 'early_source_summary.csv'), show_col_types = FALSE)
coverage <- read_csv(file.path(out, 'spatial_source_coverage.csv'), show_col_types = FALSE)
correlation <- read_csv(file.path(out, 'signal_correlation.csv'), show_col_types = FALSE)
include_figure <- function(id) {
  path <- list.files(fig_dir, pattern = paste0('^', id, '.*\\.png$'), full.names = TRUE)
  if (length(path)) cat(paste0('![](../../output/fig/', basename(path[[1]]), ')\n'))
}
round_numeric <- function(x, digits = 3) mutate(x, across(where(is.numeric), ~round(.x, digits)))
```

## Decision

The spatial early-event update is now implemented and reproducible, but **no candidate is promoted into the current best forecast**. The initial operational forecast remains unchanged. Early aerial/RHIS information is retained as a gated within-event update layer that must be revalidated when dated aerial observations for 2022 and 2024 are available.

The strongest current candidate combines aerial severe-bleaching scores with RHIS community-weighted bleaching burden and bleaching-associated recently dead coral. With the target sector removed from the early-data surface, it improves RMSE from 0.1380 to **0.1338** and predictive R-squared from 0.228 to **0.275** across 2016, 2017 and 2020. However, severe RMSE worsens from 0.381 to **0.409**, principally because the update suppresses severe 2017 losses. It therefore fails the severe-tail safeguard.

RHIS-only updates cover all five events but worsen overall RMSE, predictive R-squared and occurrence Brier score. Allowing early observations to modify conditional magnitude as well as occurrence performs worse again. The event-time evidence should therefore not yet be used as an unconstrained mortality multiplier.

## What was tested

The update is separate from the initial forecast. For reef-event (i), the selected model supplies an initial occurrence probability (p_i^{(0)}) and conditional positive-loss magnitude (m_i^{(0)}). The primary update changes occurrence only:

$$
\operatorname{logit}(p_i^{\mathrm{update}})
= \operatorname{logit}(p_i^{(0)})
+ \alpha + \boldsymbol\gamma^\top\mathbf E_i,
\qquad
\widehat M_i^{\mathrm{update}}=p_i^{\mathrm{update}}m_i^{(0)},
$$

where (mathbf E_i) contains early bleaching signals. Coefficients are estimated using other bleaching events. The target sector or latitude block is removed before its spatial signal is calculated, so the update cannot validate itself with observations from the target geography.

A compact two-part sensitivity also updated (m_i^{(0)}). It was rejected because it worsened both overall and severe-event transfer.

## Leakage controls

- RHIS records are restricted to January–April of the event year. April 30 is the primary cutoff; March 31 is a stricter timing sensitivity.
- Only records with `Bleaching Present = TRUE` contribute recently dead coral. Later post-event mortality surveys and disturbance labels are never used.
- Aerial records are used only for 2016, 2017 and 2020. The supplied file has no survey-date field, so its within-event timing cannot be independently checked and is recorded as unavailable provenance.
- Each validation fold holds out an entire event for coefficient fitting and an entire target sector or latitude block for construction of the early-data field.
- The model's conditional mortality magnitude remains fixed except in the explicitly labelled two-part sensitivity.

## RHIS conversion

The eight morphology-specific severity fields are converted exactly as requested: 0 = none, 1 = upper-surface bleaching, 2 = pale/fluorescent, 3 = totally white, and 4 = recently dead with light algal cover. Severity is divided by four and weighted by live-plus-recently-dead cover of each morphology.

The primary burden also includes the percentage bleached within each morphology:

$$
B_j =
\frac{\sum_g C_{jg}(S_{jg}/4)(P_{jg}/100)}
     {\sum_g C_{jg}},
$$

where (C) is morphology cover, (S) is the 0–4 severity and (P) is percent bleached. Bleaching-associated recently dead benthos is retained as a separate 0–1 signal. Reef-event aggregation uses the maximum early burden so a brief severe observation is not averaged away by repeated low-severity RHIS visits.

```{r}
source_table <- sources |>
  filter(source == 'RHIS', cutoff == 'April 30') |>
  transmute(
    event = event_year, reefs, records,
    bleaching_prevalence,
    mean_community_severity = mean_severity,
    mean_community_burden = mean_burden,
    mean_bleaching_recent_dead = mean_recent_dead
  ) |>
  round_numeric(3)
kable(source_table, caption = 'Early RHIS source data available by April 30.')
```

```{r, results='asis'}
include_figure('Fig-NOWCAST-01_early_source_coverage')
```

Bleaching is already widespread in RHIS by April: prevalence ranges from 0.621 in 2020 to 0.978 in 2017. This explains why presence alone is a weak discriminator. Spatial burden and severity are more informative candidates, but their transfer is not yet stable.

## Spatial support

The source field uses a 150-km Gaussian kernel truncated at 450 km and requires an effective sample size of at least three. Sector exclusion retains reasonable RHIS support for most events (83–98% for April burden), whereas the latitude-block test is deliberately harsher and has poor 2016 RHIS coverage. Aerial scores have dense independent support in all three available events.

```{r}
coverage_table <- coverage |>
  filter(field %in% c('rhis_burden', 'aerial_severe_bleaching'),
         cutoff %in% c('April 30', 'not supplied')) |>
  transmute(
    source, design, event = event_year,
    signal = recode(field, rhis_burden = 'RHIS burden',
                    aerial_severe_bleaching = 'Aerial severe score'),
    coverage, median_effective_n, median_nearest_km
  ) |>
  round_numeric(2)
kable(coverage_table, caption = 'Coverage after excluding each target sector or spatial block.')
```

## Validation results

```{r}
comparison_table <- comparison |>
  transmute(
    candidate, design,
    updated_RMSE = updated_rmse,
    initial_RMSE = initial_rmse,
    RMSE_change = delta_rmse,
    updated_predictive_R2 = updated_predictive_r2,
    predictive_R2_change = delta_predictive_r2,
    updated_severe_RMSE = updated_severe_rmse,
    severe_RMSE_change = delta_severe_rmse,
    Brier_change = delta_occurrence_brier
  ) |>
  arrange(RMSE_change) |>
  round_numeric(4)
kable(comparison_table, caption = 'Event-held-out and spatially independent update validation. Negative RMSE/Brier changes are improvements.')
```

```{r, results='asis'}
include_figure('Fig-NOWCAST-02_spatial_update_validation')
```

The combined aerial–RHIS occurrence update is the only candidate with a lower aggregate RMSE, and only under sector exclusion. It also lowers occurrence Brier score, showing that the early observations contain real information. Its severe-tail penalty prevents promotion. Neither March 31 restriction nor the two-part magnitude update solves that problem.

```{r, results='asis'}
include_figure('Fig-NOWCAST-03_event_transfer')
```

The combined sector-independent candidate improves overall RMSE in each of 2016, 2017 and 2020, but its severe RMSE worsens sharply in 2017. This is the key failure mode to resolve before operational use.

## Complementarity

```{r}
kable(round_numeric(correlation, 3), caption = 'Spearman correlations among spatial RHIS signals.')
```

Community-weighted severity and burden correlate at 0.912 under sector exclusion. Burden and recently dead coral correlate at 0.815. They should not be expanded into a larger unconstrained predictor block. The parsimonious next comparison is aerial severity plus one RHIS burden signal; recently dead coral should remain a separately reported warning unless it demonstrates incremental held-out value after more events are added.

## Operational contract

1. The selected initial forecast is generated without aerial or RHIS event observations.
2. Early observations may be ingested through `src/evaluation/test_spatial_early_bleaching_update.R` and mapped to spatially cross-fitted signals.
3. The current update is **off by default**. It may be shown as a sensitivity/uncertainty layer, not blended into the official prediction.
4. Promotion requires lower overall RMSE and occurrence Brier score, no material severe-RMSE penalty, and gains that persist under both event-held-out and spatially independent validation.
5. Aerial survey dates and 2022/2024 scores must be supplied before the next promotion decision.

## Most important next steps

{{< include ../../docs/model-next-steps.md >}}

## Reproducibility

- Source-processing and validation script: `src/evaluation/test_spatial_early_bleaching_update.R`
- Processed RHIS source: `data/processed/rhis_early_bleaching_reef_event.csv`
- Processed aerial source: `data/processed/aerial_early_bleaching_reef_event.csv`
- Predictions and metrics: `output/spatial_early_bleaching_update/`
- Registered figures: `output/fig/Fig-NOWCAST-01` to `Fig-NOWCAST-03`
