---
title: "Independent BRT and environmental-envelope ensemble assessment"
subtitle: "Operational mortality prediction and within-event interpolation"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    embed-resources: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(knitr)
})
source('../../src/lib/model_registry.R')
root <- project_root()
out_dir <- file.path(root, 'output', 'standalone_brt_envelope_ensemble')
comparison <- read_csv(file.path(out_dir, 'model_comparison.csv'), show_col_types = FALSE)
event_metrics <- read_csv(file.path(out_dir, 'event_metrics.csv'), show_col_types = FALSE)
envelope_metrics <- read_csv(file.path(out_dir, 'envelope_metrics.csv'), show_col_types = FALSE)
promotion <- read_csv(file.path(out_dir, 'promotion_assessment.csv'), show_col_types = FALSE)
screen <- read_csv(file.path(out_dir, 'collinearity_screen.csv'), show_col_types = FALSE)
tuning <- read_csv(file.path(out_dir, 'nested_tuning.csv'), show_col_types = FALSE)
fig_dir <- file.path(root, 'output', 'fig')
include_figure <- function(id) {
  match <- list.files(fig_dir, pattern = paste0('^', id, '.*[.]png$'), full.names = TRUE)
  if (!length(match)) cat('*Figure has not yet been generated.*\n')
  else cat(sprintf('![](../../output/fig/%s)\n', basename(match[[1]])))
}
round_table <- function(x) mutate(x, across(where(is.numeric), ~round(.x, 4)))
```

## Purpose

This assessment implements the intended machine-learning comparison: a BRT
fitted independently to the full operational predictor set. It is distinct
from the earlier **residual BRT**, which attempted only to correct errors left
after fitting INLA. The independent BRT can compete directly with INLA and can
be blended with it without treating the INLA prediction as a required input.

The primary question is prospective: can the BRT predict a bleaching event not
represented during training? Reef-blocked validation is retained as a
secondary test of spatial interpolation within event states already represented.

## Models

### Selected INLA reference

The reference is the selected cause-aware INLA composite. It estimates a
two-part thermal/freshwater response with partially pooled programme layers,
then combines separate cyclone and COTS hazards on the mortality scale. It uses
local-first DHW and the current prospective COTS state.

### Direct standalone BRT

For mortality proportion $M_i$ and operational predictors $\mathbf X_i$:

$$
\widehat M_i^{\mathrm{BRT}}=
\mathrm{clip}_{[0,1]}\{f_{\mathrm{GBM}}(\mathbf X_i)\}.
$$

The Gaussian-loss tree fit reproduces the direct-mortality BRT approach used in
the earlier work. Clipping is applied only when converting output to a valid
mortality proportion.

### Two-part standalone BRT

The occurrence/magnitude sensitivity separates zero loss from positive loss:

$$
\widehat M_i^{\mathrm{2BRT}}=\widehat p_i\widehat\mu_i,
\qquad
\widehat p_i=f_{\mathrm{Bernoulli}}(\mathbf X_i),
\qquad
\operatorname{logit}(\widehat\mu_i)=
f_{\mathrm{Gaussian}}(\mathbf X_i\mid M_i>0).
$$

This can lower the mortality floor, but can compound underprediction when both
occurrence and positive magnitude are conservative.

## Predictor contract and complementarity

Both BRTs receive the same forecastable information: local-first DHW,
pre-event Acropora and starting cover, heat history and novelty, Secchi, cloud,
current speed, SST skewness, rainfall, coloured water, calm winds, cyclone
exposures, COTS pressure/history, NOAA disease risk, depth, programme and
region. Event identity is excluded.

The Spearman screen is fixed before fitting and uses no mortality values.
Disease applicability is removed because it is effectively a duplicate of
disease risk. SST kurtosis and median chlorophyll are not restored because the
preceding complementarity assessment retained SST skewness and Secchi as the
less redundant representatives.

```{r}
screen |>
  transmute(
    Predictor = predictor,
    Retained = retained,
    `Closest earlier predictor` = competing_predictor,
    `Absolute Spearman correlation` = round(absolute_correlation, 3)
  ) |>
  kable(caption = 'Outcome-free predictor collinearity screen used by both BRTs.')
```

## Validation and tuning

Interaction depth is tuned inside every outer fold. The primary folds hold out
2016, 2017, 2020, 2022 or 2024 in turn. The secondary design holds out reefs in
five spatially balanced groups. Reef-event and programme weights prevent repeat
observations or a single programme from dominating the fit.

```{r}
tuning |>
  filter(selected_two_part | selected_direct) |>
  select(scheme, fold, setting_id, depth, selected_two_part, selected_direct,
         inner_rmse, direct_inner_rmse) |>
  round_table() |>
  kable(caption = 'Nested tuning choices in each outer validation fold.')
```

```{r}
#| results: asis
include_figure('Fig-FULLBRT-01_validation')
```

```{r}
comparison |>
  filter(candidate %in% c(
    'INLA explicit hazards', 'Standalone direct BRT',
    'Standalone two-part BRT', 'Direct smooth 50% ensemble',
    'Direct applicability champion'
  )) |>
  select(candidate, scheme, rmse, predictive_r2, severe_rmse,
         occurrence_brier, false_extreme_rate, mean_bias) |>
  round_table() |>
  kable(caption = 'Matched validation of the reference, independent BRTs and leading envelope blends.')
```

The direct BRT improves reef-blocked RMSE from 0.1396 to 0.1354, and the direct
applicability blend reaches 0.1282. However, direct-BRT leave-one-event-out RMSE
is 0.1799 versus 0.1485 for INLA, while severe RMSE is 0.4650 versus 0.3267.
The direct BRT therefore learns useful interpolation within represented events
but does not transfer safely to a new event.

## Environmental-envelope ensemble

The ensemble gate is calculated separately within every training fold and is
independent of mortality. Numeric predictors are imputed with training
medians, standardised, projected onto whitened principal components retaining
90% of predictor variance (maximum eight), and scored by mean distance to the
ten nearest training neighbours. The nominal envelope is the 95th percentile
of training-neighbour distance with a penalty for predictors outside training
ranges.

Three pre-specified weights are evaluated:

$$
\widehat M_i^{\mathrm{ens}}=
w_i\widehat M_i^{\mathrm{BRT}}+(1-w_i)\widehat M_i^{\mathrm{INLA}},
$$

where $w_i$ is a hard inside-envelope indicator, half the smooth applicability
score, or the full applicability score. The gate never sees the outcome or
which model happened to have the smaller error.

```{r}
#| results: asis
include_figure('Fig-FULLBRT-03_envelope_performance')
```

The proposed inside-envelope preference is not supported for a new-event
forecast. Among event-held-out rows inside the envelope, INLA RMSE is 0.1354
compared with 0.1692 for the direct BRT. Represented covariate space does not
guarantee representation of the event-level mortality process.

```{r}
envelope_metrics |>
  filter(candidate %in% c(
    'INLA explicit hazards', 'Standalone direct BRT',
    'Standalone two-part BRT'
  )) |>
  select(candidate, scheme, domain, n, rmse, predictive_r2, severe_rmse) |>
  round_table() |>
  kable(caption = 'Performance inside and outside each training-fold environmental envelope.')
```

```{r}
#| results: asis
include_figure('Fig-FULLBRT-05_local_competence')
```

## Event transfer and calibration

```{r}
#| results: asis
include_figure('Fig-FULLBRT-02_event_transfer')
```

The direct BRT is the stronger independent BRT, but it loses substantial skill
when 2017 or 2024 is absent from training. The two-part model gives lower RMSE
than INLA in 2020 and 2022 by suppressing the low-mortality floor, but then
underpredicts the severe 2024 magnitude. This is the trade-off that aggregate
reef-blocked validation hides.

```{r}
event_metrics |>
  filter(candidate %in% c(
    'INLA explicit hazards', 'Standalone direct BRT',
    'Standalone two-part BRT', 'Direct smooth 50% ensemble'
  )) |>
  select(candidate, event_year, rmse, predictive_r2, severe_rmse,
         false_extreme_rate, mean_bias) |>
  round_table() |>
  kable(caption = 'Leave-one-event-out performance by held-out bleaching event.')
```

```{r}
#| results: asis
include_figure('Fig-FULLBRT-04_calibration')
```

```{r}
#| results: asis
include_figure('Fig-FULLBRT-08_severe_controls')
```

## Model inspection

```{r}
#| results: asis
include_figure('Fig-FULLBRT-06_variable_importance')
```

The full-data direct BRT is led by local-first DHW, Acropora, DHW novelty and
starting cover. Cyclone wind-distance, cloud, rainfall and damaging-wave hours
also receive material influence. Importance is descriptive: it can be shared
among related variables and does not override held-out validation.

```{r}
#| results: asis
include_figure('Fig-FULLBRT-07_bootstrap_pdp')
```

The ribbons are 95% reef-event cluster-bootstrap intervals from 25 complete
refits. They show stability of marginal curves, not prediction intervals or
causal effects.

## Promotion decision

```{r}
promotion |>
  select(candidate, rmse, severe_rmse, reef_rmse,
         max_lanina_rmse_increase, false_extremes_safe,
         reef_transfer_safe, lanina_safe, passes) |>
  round_table() |>
  kable(caption = 'Pre-specified promotion safeguards for standalone and ensemble candidates.')
```

No BRT or ensemble passes. The initial-forecast model remains
`operational_rrn_raw_plus_manta_state`. This does not imply that INLA is
universally better. The BRT interpolation gains are useful after the event
state is represented, but they do not transfer to a genuinely unseen event.

Operational roles are therefore separated:

- **Initial forecast:** selected cause-aware INLA composite.
- **Within-event spatial interpolation or sensitivity:** independent direct
  BRT and applicability blend, after a spatially independent update is
  validated.
- **Occurrence-floor research:** two-part BRT, retained as a sensitivity for
  low-mortality 2020/2022 states.
- **Mechanism inspection:** full-data importance and bootstrap partial
  dependence from both BRT formulations.

## Most important next steps

{{< include ../../docs/model-next-steps.md >}}

## Reproducibility

The pipeline is in `src/models/fit_standalone_brt_envelope_ensemble.R`.
Machine-readable predictions, tuning, collinearity decisions, metrics,
promotion tests and full fits are stored under
`output/standalone_brt_envelope_ensemble/`. Publication PNG, PDF and plotted
data are stored under `output/fig/`.
