---
title: "Decomposed annual coral-cover change"
subtitle: "Transferring bleaching modifiers into recovery and disturbance-loss prediction"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(scales)
library(knitr)
theme_set(theme_minimal(base_size = 12))

out <- '../../output/decomposed_annual_change'
metrics <- read_csv(file.path(out, 'matched_model_comparison_metrics.csv'), show_col_types = FALSE)
predictions <- read_csv(file.path(out, 'decomposed_change_predictions.csv'), show_col_types = FALSE)
events <- read_csv(file.path(out, 'decomposed_event_summary.csv'), show_col_types = FALSE)
importance <- read_csv(file.path(out, 'decomposed_component_importance.csv'), show_col_types = FALSE)
contrasts <- read_csv(file.path(out, 'modifier_contrast_summary.csv'), show_col_types = FALSE)
severe <- read_csv(file.path(out, 'severe_2024_reef_diagnostics.csv'), show_col_types = FALSE)
residual_correlations <- read_csv(file.path(out, 'severe_2024_residual_correlations.csv'), show_col_types = FALSE)
blend <- read_csv(file.path(out, 'equal_blend_diagnostics.csv'), show_col_types = FALSE)

candidates <- c(
  'Core decomposed BRT', 'Modifier-rich decomposed BRT',
  'Mortality-augmented decomposed BRT'
)
```

## Decision

The decomposed model improves on treating annual change as one unconstrained
response. It predicts recovery relative to available reef space and loss
relative to current coral cover, then combines them into a bounded next-cover
estimate.

There is not one winner for every operational use:

- Before a genuinely novel event is observed, the **core decomposition** is
  safest (forward 2024--2025 RMSE 10.05 percentage points; recent-year RMSE
  8.40).
- For spatial prediction after early observations from an event are available,
  the **modifier-rich model** is strongest across unseen reefs (RMSE 6.76;
  predictive R2 0.209).
- For reef-blocked 2024, transferring the restricted relative-mortality
  observations into the loss learner gives the best RMSE (8.59) and predicts
  mean change as -4.52 points, compared with -5.89 observed.

The severe tail remains underpredicted. For 53 reef-blocked 2024 transitions
with losses of at least 10 points, mean observed change is -21.58 points and
the mortality-augmented prediction is -11.14. This is useful progress, but not
yet adequate for the final GBR-wide product.

## Model structure

The predictive transition is

$$C_{t+1} = C_t + (1-C_t)G_t - C_tL_t,$$

where $G_t$ is expected net gain relative to available space and $L_t$ is
expected net loss relative to existing cover. Each has a Bernoulli occurrence
model and a bounded positive-magnitude model.

These components are inferred from **net annual change**. Annual snapshots do
not independently observe gross growth and mortality when both happened in the
same interval.

Three loss formulations were tested:

1. Structural state, nonlinear DHW and Acropora.
2. The same model plus thermal history/novelty, Secchi/cloud, water colour,
   coastal rainfall, wind/calm, cyclone waves, COTS and explicit DHW
   interactions.
3. The modifier-rich model additionally trained with the restricted,
   bleaching-attributed relative-mortality responses. Auxiliary rows obey the
   same year or reef exclusions as each annual fold.

Rainfall and water colour remain exposure proxies, not measured salinity.
Cyclone waves and COTS remain separate mechanisms.

## Matched validation

```{r}
models <- c('Persistence', 'Growth/environment LM', 'Change BRT', candidates)
metric_table <- metrics |>
  filter(programme_key == 'all', learner %in% models) |>
  transmute(
    Model = learner,
    Validation = recode(scheme,
      recent_year_holdout = 'Recent year held out',
      reef_blocked_5fold = 'Unseen reefs',
      forward_2024_2025 = 'Forward 2024-2025'),
    `RMSE (pp)` = round(post_cover_rmse_pp, 2),
    `Predictive R2` = round(change_predictive_r2, 3),
    `Direction accuracy` = percent(direction_accuracy, accuracy = 0.1),
    `Severe-loss MAE (pp)` = round(severe_loss_mae_pp, 2),
    `Severe underprediction (pp)` = round(severe_loss_bias_pp, 2)
  )
kable(metric_table)
```

```{r}
metrics |>
  filter(programme_key == 'all', learner %in% models) |>
  mutate(
    Validation = recode(scheme,
      recent_year_holdout = 'Recent year held out',
      reef_blocked_5fold = 'Unseen reefs',
      forward_2024_2025 = 'Forward 2024-2025'),
    learner = factor(learner, levels = rev(models))) |>
  ggplot(aes(post_cover_rmse_pp, learner, colour = Validation)) +
  geom_point(size = 2.8) +
  facet_wrap(~Validation, scales = 'free_x') +
  labs(x = 'Next-cover RMSE (percentage points)', y = NULL,
       title = 'Different validation tasks favour different routes') +
  theme(legend.position = 'none')
```

An equal modifier-rich/mortality-augmented blend has the lowest overall
unseen-reef RMSE (6.74), but its weight has not yet been cross-fitted. It is a
sensitivity, not a promoted ensemble, and it is weaker than the core model when
2024 is entirely absent.

```{r}
kable(blend |> transmute(
  Validation = scheme, `RMSE (pp)` = round(rmse_pp, 2),
  `Predictive R2` = round(predictive_r2, 3),
  `Severe-loss MAE (pp)` = round(severe_mae_pp, 2)))
```

## The 2024 event

```{r}
kable(events |>
  filter(scheme == 'reef_blocked_5fold', event_year == 2024,
         learner %in% candidates) |>
  transmute(
    Model = learner,
    `Observed mean (pp)` = round(observed_change_pp, 2),
    `Predicted mean (pp)` = round(predicted_change_pp, 2),
    `Predicted gain (pp)` = round(predicted_gain_pp, 2),
    `Predicted loss (pp)` = round(predicted_loss_pp, 2),
    `RMSE (pp)` = round(rmse_pp, 2),
    `Severe observed (pp)` = round(severe_loss_observed_pp, 2),
    `Severe predicted (pp)` = round(severe_loss_predicted_pp, 2)))
```

```{r}
predictions |>
  filter(scheme == 'reef_blocked_5fold', event_year == 2024,
         learner %in% candidates) |>
  ggplot(aes(cover_change_pp, predicted_change_pp, colour = programme_key)) +
  geom_abline(slope = 1, intercept = 0, colour = 'grey45') +
  geom_point(alpha = 0.65, size = 1.7) +
  facet_wrap(~learner) +
  coord_equal(xlim = c(-65, 35), ylim = c(-65, 35)) +
  labs(x = 'Observed annual change (percentage points)',
       y = 'Predicted annual change (percentage points)',
       colour = 'Programme',
       title = 'Mortality transfer deepens 2024 losses',
       subtitle = 'The largest losses remain underestimated')
```

For severe 2024 cases, the rich covariates alone add only 0.21 points of loss
on average and improve 43% of reefs. The restricted mortality response adds a
further 1.81 points of loss and improves 64%. Response transfer helps; simply
adding more covariates is insufficient.

```{r}
kable(contrasts |>
  filter(scheme == 'reef_blocked_5fold', event_year == 2024) |>
  transmute(
    Class = if_else(severe_loss, 'Loss >=10 pp', 'Other transitions'),
    N = transitions,
    `Rich minus core (pp)` = round(mean_rich_minus_core_change_pp, 2),
    `Rich improves` = percent(rich_improvement_fraction, accuracy = 0.1),
    `Auxiliary minus rich (pp)` = round(mean_augmented_minus_rich_change_pp, 2),
    `Auxiliary improves` = percent(auxiliary_improvement_fraction, accuracy = 0.1)))
```

## Remaining severe misses

```{r}
severe |>
  slice_head(n = 15) |>
  transmute(
    Reef = ReefName, Programme = programme_key,
    Observed = round(cover_change_pp, 1),
    Core = round(predicted_change_pp_Core_decomposed_BRT, 1),
    Rich = round(predicted_change_pp_Modifier_rich_decomposed_BRT, 1),
    Augmented = round(predicted_change_pp_Mortality_augmented_decomposed_BRT, 1),
    DHW = round(sst_maxdhw, 1),
    Acropora = percent(prop_acropora_pre, accuracy = 1),
    WQC = round(wqc_freqcc12, 2), Annotation = disturbance_text) |>
  kable()
```

Snapper remains the dominant miss: observed losses of 50--58 points are
predicted as only 5--6 despite very high water-colour exposure and explicit
Jasper/flood annotations. North Direction, Mackay, Linnet, One Tree and
Broomfield improve to varying degrees, but none is fully captured.

Residual is observed minus predicted change, so a negative value denotes
underprediction of loss. Among severe 2024 rows, greater calm exposure,
rainfall and water colour are associated with more negative residuals. The
positive DHW association means the worst misses concentrate among severe losses
at moderate rather than extreme DHW. This is consistent with an incompletely
represented freshwater/calm mechanism; it is diagnostic, not causal evidence.

```{r}
residual_correlations |>
  mutate(
    predictor = recode(predictor,
      era5_wind_calm_fraction = 'Calm fraction',
      log_coastal_rain30 = 'Coastal rainfall proxy',
      wqc_freqcc12 = 'Water colour',
      wqc_prior10_percentile = 'Reef-relative water colour',
      cyc_maxHrs4mw = 'Cyclone 4 m wave hours',
      cot_idwmeanpertow = 'COTS pressure',
      prop_acropora_pre = 'Acropora proportion', sst_maxdhw = 'DHW'),
    predictor = reorder(predictor, spearman_with_residual)) |>
  ggplot(aes(spearman_with_residual, predictor)) +
  geom_vline(xintercept = 0, colour = 'grey70') +
  geom_col(fill = '#287D8E') +
  labs(x = 'Spearman correlation with residual (observed - predicted)',
       y = NULL,
       title = 'Freshwater and calm signals remain in severe residuals')
```

The RRN cyclone-wave value is zero at several Jasper-annotated northern reefs.
It should remain in the model, but cannot alone represent Jasper or floodwater
exposure.

## Loss-variable influence

```{r}
importance |>
  filter(grepl('loss', component)) |>
  group_by(component) |>
  slice_max(influence, n = 12, with_ties = FALSE) |>
  ungroup() |>
  mutate(variable = reorder(variable, influence)) |>
  ggplot(aes(influence, variable, fill = component)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~component, scales = 'free_y') +
  labs(x = 'Relative influence (%)', y = NULL,
       title = 'Loss models retain state and disturbance modifiers')
```

The annual model has not reverted to a linear DHW response. Its loss process
uses prior trajectory, cover, cyclone exposure, thermal novelty, DHW--Acropora,
COTS, rainfall, water colour and cloud.

## Next fit

The next Bayesian/state-space fit should:

1. Use beta magnitude and binomial occurrence responses for gain and loss with
   programme-specific observation layers.
2. Retain the modifier-rich and mortality-augmented BRTs as ensemble members.
3. Evaluate stronger tail response to DHW interactions with rainfall, water
   colour and calm against reef-blocked 2024 severe loss, not overall RMSE only.
4. Keep field annotations diagnostic rather than operational predictors.
5. Add a broader community-state variable when available; Acropora alone does
   not identify reefs dominated by fast-growing weedy taxa.

Operationally, use the core decomposition before a novel event is observed,
then switch to a cross-fitted rich/augmented ensemble after early event surveys
are available. This prevents a complete 2024 holdout from teaching the wrong
freshwater response.
