---
title: 'Event-specific NOAA DHW correction'
subtitle: 'Authenticated AIMS logger extraction, depth diagnostics and mortality-model validation'
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)

logger <- read_csv('../data/processed/aims_logger_noaa_dhw_validation.csv',
  show_col_types = FALSE)
logger_ok <- logger |> filter(adequate_dhw_coverage)
range_metrics <- read_csv('../../output/dhw_correction/range_screen_metrics.csv',
  show_col_types = FALSE)
selected <- read_csv('../../output/dhw_correction/selected_configuration.csv',
  show_col_types = FALSE)
depth_pairs <- read_csv('../../output/dhw_correction/paired_depth_diagnostics.csv',
  show_col_types = FALSE)

base_name <- 'persistent_rw1_cyclone_cots_partial_pool'
rolling_name <- 'persistent_rw1_cyclone_cots_rolling_dhw_partial_pool'
event_name <- 'persistent_rw1_cyclone_cots_event_dhw_partial_pool'
candidate_labels <- c(
  'NOAA unchanged',
  'Historical logger surface',
  'Within-event logger surface')
names(candidate_labels) <- c(base_name, rolling_name, event_name)

inla_predictions <- read_csv('../../output/inla_spatiotemporal/cv_predictions.csv',
  show_col_types = FALSE)
inla_metrics <- read_csv('../../output/inla_spatiotemporal/model_comparison.csv',
  show_col_types = FALSE) |>
  filter(candidate %in% names(candidate_labels)) |>
  mutate(
    DHW_input = recode(candidate, !!!candidate_labels),
    validation = recode(scheme,
      reef_blocked_5fold = 'Reef-blocked 5-fold',
      leave_one_event_out = 'Leave-one-event-out'))

cluster_ids <- c('14-114', '14-116a', '14-118', '14-123', '14-126', '14-143')
cluster_predictions <- inla_predictions |>
  filter(event_year == 2024, ReefID %in% cluster_ids,
    scheme == 'leave_one_event_out', candidate %in% names(candidate_labels)) |>
  mutate(
    DHW_input = recode(candidate, !!!candidate_labels),
    NOAA_DHW = coalesce(ann_maxdhw_original, ann_maxdhw),
    correction_DHW = coalesce(applied_dhw_uplift, 0)) |>
  group_by(DHW_input, ReefID, ReefName) |>
  summarise(
    observed = mean(observed_mortality),
    predicted = mean(predicted_mortality), NOAA_DHW = mean(NOAA_DHW),
    correction_DHW = mean(correction_DHW), model_DHW = mean(ann_maxdhw),
    .groups = 'drop')
```

## Decision summary

The authenticated AIMS data support an event-specific NOAA correction layer,
but not a universal Lizard-region uplift. Discrepancies have moderate
cross-event persistence, while their magnitude and even their sign vary among
sites and years. The production-safe form is therefore a partially pooled
spatial surface trained only on previous events. A second, within-event surface
is retained for retrospective analysis or operational updating once enough
current-event logger data exist.

The historical correction modestly improves leave-one-event-out mortality RMSE
from `r number(inla_metrics$rmse[inla_metrics$candidate == base_name & inla_metrics$scheme == 'leave_one_event_out'], accuracy = 0.0001)`
to `r number(inla_metrics$rmse[inla_metrics$candidate == rolling_name & inla_metrics$scheme == 'leave_one_event_out'], accuracy = 0.0001)`.
The within-event layer raises exposure at the principal 2024 Lizard-region
misses, but does not resolve their severe mortality and slightly worsens
GBR-wide event-transfer RMSE to
`r number(inla_metrics$rmse[inla_metrics$candidate == event_name & inla_metrics$scheme == 'leave_one_event_out'], accuracy = 0.0001)`.
Local heat underestimation is therefore one contributor, not the whole
explanation; freshwater, community composition and other compound pressures
remain necessary.

## Reproducible authenticated extraction

`src/data/pull_aims_temperature_data.R` downloads the complete AIMS temperature
deployment catalogue, retains deployment depth, and downloads every daily
logger record in the prespecified bleaching-event windows. It can also add
automated weather/ocean-station water temperatures. Raw responses are cached
under `data/raw/aims_temperature/api/`; the key is read only from
`AIMS_DATAPLATFORM_API_KEY` and is never written into the repository or report.

The default extraction covers 2016, 2017, 2020, 2022 and 2024. Years, refresh
behaviour and automated-station scope are configuration values, so expansion
does not require code changes.

```{r}
logger |>
  count(event_year, source, depth_class, adequate_dhw_coverage,
    name = 'series_events') |>
  arrange(event_year, source, depth_class, desc(adequate_dhw_coverage)) |>
  kable(caption = 'AIMS--NOAA comparisons by event, source, depth and coverage')
```

## GBR-wide logger discrepancy

Logger DHW is reconstructed from quality-controlled daily AIMS temperatures,
the collocated NOAA maximum monthly mean, and the NOAA 84-day HotSpot threshold
definition. A comparison is accepted only when at least 70 days are available
in the 84-day window at the eligible logger-DHW peak.

```{r}
logger_ok |>
  group_by(event_year) |>
  summarise(
    adequate_series = n(), sites = n_distinct(site),
    median_logger_minus_NOAA_DHW = median(dhw_discrepancy_logger_minus_noaa),
    q10 = quantile(dhw_discrepancy_logger_minus_noaa, 0.10),
    q90 = quantile(dhw_discrepancy_logger_minus_noaa, 0.90),
    NOAA_underestimated_fraction = mean(dhw_discrepancy_logger_minus_noaa > 0),
    .groups = 'drop') |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  kable(caption = 'Coverage-qualified AIMS comparisons by bleaching event')
```

```{r, fig.width=9, fig.height=8}
ggplot(logger_ok, aes(lon, lat, colour = dhw_discrepancy_logger_minus_noaa)) +
  geom_point(alpha = 0.75, size = 1.6) +
  facet_wrap(~ event_year, ncol = 2) +
  scale_colour_gradient2(
    low = '#2c7bb6', mid = 'white', high = '#d7191c', midpoint = 0,
    limits = c(-8, 8), oob = squish, name = 'Logger minus\nNOAA DHW') +
  coord_equal() +
  labs(
    x = 'Longitude', y = 'Latitude',
    title = 'NOAA discrepancy is spatially and temporally heterogeneous',
    subtitle = 'Positive values indicate NOAA underestimation; colour capped at +/-8 DHW') +
  theme_bw(base_size = 11) +
  theme(legend.position = 'bottom')
```

Across series observed in multiple events, discrepancy ranks are moderately
persistent (event-pair Spearman correlations approximately 0.59--0.73), but
extreme individual logger anomalies mean interpolation uncertainty must be
retained.

## Depth: use it for record selection, not an unstable slope

NOAA is a surface-temperature product. The correction builder therefore
prefers shallow/reef-flat records (0--5 m) within each site-event and uses deep
records only where a shallow record is unavailable. It retains all depths in
the audit output.

Only `r nrow(depth_pairs)` coverage-qualified site-events currently contain
paired shallow and deep series. Their median shallow-minus-deep discrepancy is
`r number(median(depth_pairs$shallow_minus_deep_dhw), accuracy = 0.01)` DHW.
That is not enough replication to estimate a reliable continuous depth effect
or a separate deep correction surface.

```{r}
depth_pairs |>
  transmute(
    Event = event_year, Site = site,
    `Shallow depth (m)` = shallow_0_5m_depth,
    `Deep depth (m)` = deep_gt5m_depth,
    `Shallow minus deep discrepancy (DHW)` = shallow_minus_deep_dhw) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  kable(caption = 'Coverage-qualified paired shallow--deep diagnostics')
```

At Lizard Island in 2016, the authenticated automated station adds a 0.6 m
series and a 10.1 m series. Their reconstructed logger-minus-NOAA discrepancies
are -0.68 and -0.35 DHW respectively. Thus 2016 NOAA underestimation was not
uniform across the Lizard region: it is strong at Linnet and Martin, but not at
the Lizard Island station itself.

## Event-specific spatial correction

The correction aggregates to site-event, partially pools each site's historical
discrepancy toward zero according to the number of previous events, and
interpolates from at most eight nearby logger sites. Spatial ranges were
screened with both held-out sites and held-out whole events.

```{r, fig.width=9, fig.height=5.5}
range_metrics |>
  mutate(validation = recode(
    scheme,
    leave_event_out = 'Historical surface: event held out',
    leave_site_out = 'Historical surface: site held out',
    within_event_leave_site_out = 'Within-event surface: site held out')) |>
  ggplot(aes(range_km, rmse, colour = validation)) +
  geom_hline(
    yintercept = unique(range_metrics$zero_correction_rmse),
    linetype = 2, colour = 'grey45') +
  geom_line() +
  geom_point(size = 2.4) +
  scale_x_continuous(breaks = sort(unique(range_metrics$range_km))) +
  labs(
    x = 'Maximum logger interpolation range (km)',
    y = 'Held-out logger discrepancy RMSE (DHW)', colour = NULL,
    title = 'Blocked validation of the NOAA correction surface',
    subtitle = 'Dashed line: no correction; individual logger extremes dominate RMSE') +
  theme_bw(base_size = 11) +
  theme(legend.position = 'bottom')
```

The selected historical surface uses a
`r selected$selected_persistent_range_km` km range; the within-event update
uses `r selected$selected_within_event_range_km` km to estimate the broader
event offset. The historical surface improves event-held-out logger-discrepancy
RMSE from 4.93 to 4.73 DHW at the selected 100 km range, and its best screened
event-held-out value is 4.67 DHW at 25 km. Selection uses the site-held-out
criterion to avoid tuning only to years.

## Mortality-model validation

```{r}
inla_metrics |>
  select(
    DHW_input, validation, RMSE = rmse, MAE = mae,
    predictive_R2 = predictive_r2, bias,
    occurrence_Brier = occurrence_brier,
    severe_RMSE = severe_rmse, false_extreme_rate,
    WAIC = waic, DIC = dic) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  arrange(validation, RMSE) |>
  kable(caption = 'INLA validation with unchanged and logger-corrected DHW')
```

Reef-blocked performance is essentially unchanged across the three inputs.
For transfer to an unseen event, the historical surface is strongest by RMSE
and severe-case RMSE. The within-event surface cannot be interpreted as
prospective leave-event-out information; it is a retrospective/current-event
update.

```{r, fig.width=9, fig.height=6}
ggplot(cluster_predictions, aes(observed, predicted, colour = DHW_input)) +
  geom_abline(slope = 1, intercept = 0, colour = 'grey45') +
  geom_point(size = 2.8) +
  geom_text(
    aes(label = ReefID), nudge_y = 0.018, size = 3,
    check_overlap = TRUE, show.legend = FALSE) +
  coord_equal(xlim = c(0, 0.65), ylim = c(0, 0.65)) +
  labs(
    x = 'Observed relative mortality', y = 'Predicted relative mortality',
    colour = 'DHW input',
    title = 'Logger correction alone does not resolve the 2024 Lizard-region misses',
    subtitle = 'Predictions use the fold in which the complete 2024 event was withheld') +
  theme_bw(base_size = 11) +
  theme(legend.position = 'bottom')
```

```{r}
cluster_predictions |>
  mutate(across(
    c(observed, predicted, NOAA_DHW, correction_DHW, model_DHW),
    ~ round(.x, 3))) |>
  arrange(ReefID, DHW_input) |>
  select(
    ReefID, ReefName, DHW_input, NOAA_DHW,
    correction_DHW, model_DHW, observed, predicted) |>
  kable(caption = '2024 Lizard-region leave-event-out predictions')
```

The within-event surface adds approximately 0.96 DHW at Linnet, 0.94 at Martin
and 0.84 at Eyrie. Predictions rise, but remain far below the observed severe
losses. Temperature correction must not substitute for freshwater/WQC,
Acropora composition, COTS or cyclone effects.

## Operational recommendation

1. Retain NOAA DHW as the base exposure and store the correction as a separate,
   versioned field with distance, effective logger count and correction mode.
2. Use the historical-only correction for genuinely prospective predictions;
   it uses no logger observations from the event being predicted.
3. Update to the within-event correction only after prespecified coverage and
   spatial-support thresholds are met, and label those predictions as updated.
4. Prefer 0--5 m logger records for surface calibration, fall back to deeper
   records where necessary, and keep shallow/deep diagnostics. Do not estimate
   a continuous depth coefficient until substantially more paired events exist.
5. Propagate correction uncertainty into the INLA predictor rather than treating
   corrected DHW as known exactly. The current screen tests the mean only.
6. Keep the older fixed +4.82 DHW Lizard uplift solely as a stress-test. It is
   informative mechanistically but is not a calibrated GBR-wide product.
