---
title: "Thermal percentile, duration, intensity, rarity and cooling assessment"
date: 2026-09-15
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(readr)
  library(tidyr)
})
root <- normalizePath('../..', winslash = '/', mustWork = TRUE)
out <- file.path(root, 'output', 'imos_thermal_screen')
coverage <- read_csv(file.path(out, 'coverage_audit.csv'), show_col_types = FALSE)
metrics <- read_csv(file.path(out, 'blocked_metrics.csv'), show_col_types = FALSE)
source <- read_csv(file.path(out, 'source_comparison.csv'), show_col_types = FALSE)
complete_metrics <- read_csv(
  file.path(out, 'complete_case_metrics.csv'), show_col_types = FALSE
)
correlations <- read_csv(
  file.path(out, 'correlations.csv'), show_col_types = FALSE
)
pbd_metrics <- metrics |>
  filter(candidate %in% c(
    'noaa_dhw', 'imos_pbd12', 'noaa_plus_pbd12'
  ))
pbd_complete <- complete_metrics |>
  filter(comparison_set == 'pbd')
pbd_noaa_rho <- correlations |>
  filter(feature_1 == 'ann_maxdhw', feature_2 == 'imos_pbd12_c_weeks') |>
  pull(spearman_rho)
logger_validation <- read_csv(
  file.path(out, 'logger_cooling_validation.csv'), show_col_types = FALSE
)
```

## Decision

Do **not** replace local-first NOAA DHW or promote the percentile heat-dose,
MHW, cooling, or IDF blocks from this screen.

Coverage-normalised IMOS/SSTAARS DHW produced small leave-one-event-out RMSE
improvements for LTMP and manta tow, but worsened reef-blocked performance and
both MMP comparisons. The observed-days-only sensitivity did not change that
cross-programme conclusion. Duration, three-day peak and persistent-hot-night
terms did not improve LTMP or manta transfer.

The Li Shing Hiung et al. percentile-based dose is scientifically distinct
from the standard MHW block, but it does not improve mortality transfer. As a
replacement it worsens every pooled programme/design comparison. Adding it to
NOAA gives small full-row event-held-out gains for LTMP and MMP, but worsens
manta and every reef-blocked result; the LTMP gain reverses on matched observed
rows and matched MMP event folds are underpowered.

The standard MHW duration/intensity block is the most promising new construct.
It modestly improves LTMP in both full-row schemes and improves manta and MMP
leave-one-event-out RMSE, but the latter gains do not reproduce under reef
blocking. Matched complete-case results leave only the manta event-held-out
gain clearly favourable. MMP complete-case event folds remain underpowered.

The paired cooling proxy fails more directly. It worsens LTMP and manta, its
apparent MMP event-held-out gain reverses on observed complete cases, and it
does not track peak-window cooling in the available hourly logger subset.

The NOAA empirical intensity-duration-frequency block improves reef-blocked
LTMP and MMP performance but worsens unseen-event transfer for LTMP and manta.
This is relevant to the retained within-event BRT, not the initial forecast.

## Exposure contract

The source-control metric reconstructs NOAA-style DHW from IMOS 0.02-degree
nighttime SST and the SSTAARS 1992--2016 nighttime climatology:

1. derive maximum monthly mean (MMM) from the SSTAARS harmonic fit;
2. calculate HotSpot as bias-corrected nighttime SST minus SSTAARS MMM;
3. retain the full HotSpot only when it is at least 1 degree C;
4. sum over 84 calendar days, divide by seven, and take the calendar-year
   maximum to match `ann_maxdhw`.

Short internal gaps of at most two days are interpolated. The primary metric is
coverage-normalised only when at least 70% of an 84-day window is available;
the unscaled observed-days estimate is retained as a sensitivity. The centre
pixel is primary, with a quality-controlled fallback within 3 km and an
absolute 5 km reef-distance cap.

The coral-threshold block uses the fixed 1 November--30 April window: longest run at
or above MMM + 1 degree C, maximum three-day HotSpot, and number of hot nights.
Extreme dose above MMM + 2 degrees C appears only in the descriptive full-data
importance pool.

Standard MHW events require at least five days above a daily SSTAARS 90th
percentile threshold; qualifying events separated by no more than two days are
joined. The daily threshold is periodic interpolation of the published twelve
SSTAARS mid-month percentiles, and categories use the official median/90th-
percentile formula. The primary block contains maximum duration and maximum
intensity.

The percentile-based dose has no five-day event rule. For every night-SST day
above the interpolated SSTAARS 90th percentile, it accumulates the full positive
anomaly above the SSTAARS daily climatological mean. The primary 12-week window
is centred on the local November--April climatological peak and is
coverage-normalised only above 70% completeness. Ten- and 11-week fields are
retained for prespecified sensitivity, not selected from mortality outcomes.

Cooling pairs observed daytime skin SST on local date D with observed
nighttime skin SST on D+1. A site-event needs at least five hot-day pairs and
50% pair coverage. No interpolation or 14-day mosaic enters this contrast.

## Coverage

```{r}
coverage |>
  mutate(
    dhw_fraction = finite_imos_dhw / reef_events,
    structure_fraction = finite_structure / reef_events,
    pbd_fraction = finite_pbd / reef_events
  ) |>
  select(
    Event = year, `Reef-events` = reef_events, `Finite DHW` = finite_imos_dhw,
    `DHW fraction` = dhw_fraction, `Finite structure` = finite_structure,
    `Structure fraction` = structure_fraction,
    `Finite MHW` = finite_mhw, `Finite PBD12` = finite_pbd,
    `PBD12 fraction` = pbd_fraction, `Finite cooling` = finite_cooling,
    `Finite NOAA IDF` = finite_idf_frequency,
    `Median hot-pair coverage` = median_hot_pair_coverage,
    `Median event coverage` = median_event_coverage
  ) |>
  knitr::kable(digits = 3)
```

IMOS DHW is available for 283 of 304 reef-events. SSTAARS has no finite water
pixel within 5 km for the other cases, mainly repeated far-northern reefs. Only
211 reef-events meet the MHW/structure coverage rule, including 55 of 103 in
2024. The stricter peak-centred PBD12 window is finite for 162 reef-events,
including 50 in 2024. The observed-pair cooling contrast is available for 260
reef-events. The
DHW source test is usable, but duration results remain conservative pending a
higher-coverage GeoPolar/AusTemp assessment.

## NOAA versus IMOS-equivalent DHW

Negative RMSE and Brier changes favour IMOS; positive R-squared changes favour
IMOS.

```{r}
source |>
  transmute(
    Programme = toupper(programme_key), Validation = scheme,
    `NOAA RMSE` = rmse_noaa_dhw, `IMOS RMSE` = rmse_imos_dhw,
    `IMOS - NOAA RMSE` = rmse_change_imos_minus_noaa,
    `Observed IMOS - NOAA RMSE` = rmse_change_observed_imos_minus_noaa,
    `IMOS - NOAA R2` = r2_change_imos_minus_noaa,
    `IMOS - NOAA Brier` = brier_change_imos_minus_noaa,
    `IMOS - NOAA severe RMSE` = severe_rmse_change_imos_minus_noaa
  ) |>
  knitr::kable(digits = 4)
```

The strongest favourable result is occurrence discrimination: scaled IMOS DHW
reduces LTMP leave-one-event-out Brier score by about 0.05. That does not become
a stable mortality-scale gain. Manta event-held-out RMSE changes by less than
0.001, while MMP and reef-blocked comparisons favour NOAA.

## Percentile-gated heat dose

[Li Shing Hiung, Holbrook and Kajtar (2026)](https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2025GL119516)
replace the fixed MMM + 1 degree C gate with a local percentile threshold. At
their standard alpha of one, the gate is the daily 90th percentile; every day
above it contributes the *full* anomaly above the daily climatological mean.
They remove the Hobday five-day duration requirement and find 10--12-week
summer windows generally optimal.

This is distinct from both thermal constructs already screened here. NOAA DHW
uses a fixed MMM + 1 gate and an 84-day rolling maximum; the Hobday block
requires a five-day event and then summarises its duration and intensity. PBD12
instead integrates locally unusual hot days whether consecutive or isolated,
so it can distinguish reefs with different background temperature variance.
It is still a cumulative dose, not an intensity-frequency-duration return
period.

The published result is promising for bleaching detection, but its reported
skill gain is not directly portable. The response is binary bleaching presence
or severe bleaching, the SST is 0.25-degree OISST, and evaluation uses random
80/20 splits that mix events and locations. This project predicts post-event
mortality magnitude at 0.02 degrees and withholds whole events. The paper also
reports convergence between definitions for extreme bleaching and long
impact windows, precisely where much of this mortality dataset lies.

Our implementation is therefore an **SSTAARS/IMOS analogue**, not a numerical
replication: SSTAARS supplies both the daily night-SST climatology and
periodically interpolated 90th percentile; no NOAA climatology is mixed in.
The 12-week peak-centred window was fixed before looking at mortality. Ten- and
11-week outputs are retained only as sensitivities; their rank correlations
with PBD12 are 0.98--0.99. PBD12 itself correlates
`r round(pbd_noaa_rho, 2)` with local-first NOAA DHW, leaving some distinct
information but substantial redundancy.

The full-row screen median-imputes missing PBD within each analysis fold.
Negative RMSE changes favour PBD.

```{r pbd-full-screen}
pbd_metrics |>
  select(programme_key, scheme, candidate, rmse) |>
  pivot_wider(names_from = candidate, values_from = rmse) |>
  mutate(
    replacement_change = imos_pbd12 - noaa_dhw,
    incremental_change = noaa_plus_pbd12 - noaa_dhw
  ) |>
  transmute(
    Programme = toupper(programme_key), Validation = scheme,
    `NOAA RMSE` = noaa_dhw, `PBD12 RMSE` = imos_pbd12,
    `PBD12 - NOAA` = replacement_change,
    `NOAA + PBD12 RMSE` = noaa_plus_pbd12,
    `Addition - NOAA` = incremental_change
  ) |>
  knitr::kable(digits = 4, caption = 'Full-row percentile heat-dose screen')
```

PBD12 replacement is worse in all six programme/design comparisons. Adding it
improves event-held-out RMSE slightly for LTMP (-0.0017) and MMP (-0.0042), but
worsens manta (+0.0062) and all three reef-blocked comparisons. The apparent
MMP 2024 gain is accompanied by a manta 2024 loss, so it does not explain the
2024 cross-programme residual pattern consistently.

Only 162 of 304 reef-events pass the peak-window coverage rule. The matched
complete-case screen is consequently decisive for separating observed signal
from fold-wise imputation.

```{r pbd-complete-screen}
pbd_complete |>
  filter(candidate %in% c('noaa_dhw', 'imos_pbd12', 'noaa_plus_pbd12')) |>
  select(programme_key, scheme, candidate, rows, rmse) |>
  pivot_wider(names_from = candidate, values_from = rmse) |>
  mutate(
    replacement_change = imos_pbd12 - noaa_dhw,
    incremental_change = noaa_plus_pbd12 - noaa_dhw
  ) |>
  transmute(
    Programme = toupper(programme_key), Validation = scheme, Rows = rows,
    `PBD12 - NOAA RMSE` = replacement_change,
    `Addition - NOAA RMSE` = incremental_change
  ) |>
  knitr::kable(digits = 4, caption = 'Matched observed-row percentile sensitivity')
```

On matched rows, replacement remains worse for LTMP and MMP reef-blocking; the
manta event fold improves by 0.0024 but its reef-blocked result is essentially
neutral. The LTMP event-held-out addition reverses from a small gain to a 0.0057
RMSE loss. Manta's event-held-out addition reverses its full-row loss to a
0.0030 gain but is neutral under reef blocking. MMP event-held-out fitting is
unsupported because one training fold has fewer than eight positive
observations or only one occurrence class, and its reef-blocked addition
worsens by 0.0023. These results fail the project's
cross-programme and cross-design promotion rule. PBD12 should remain a
well-defined research field, not enter the selected model or formal INLA stage.

## Candidate blocks

```{r}
metrics |>
  mutate(
    programme_key = toupper(programme_key),
    candidate = recode(candidate,
      noaa_dhw = 'NOAA', imos_dhw = 'IMOS scaled',
      imos_dhw_observed = 'IMOS observed',
      imos_pbd12 = 'IMOS PBD12',
      noaa_plus_pbd12 = 'NOAA + PBD12',
      noaa_plus_structure = 'NOAA + structure',
      imos_plus_structure = 'IMOS + structure',
      noaa_plus_mhw = 'NOAA + MHW', imos_plus_mhw = 'IMOS + MHW',
      noaa_plus_cooling = 'NOAA + cooling',
      noaa_plus_compact = 'NOAA + MHW + cooling',
      noaa_plus_id_profile = 'NOAA + IMOS ID profile',
      noaa_plus_idf_frequency = 'NOAA + NOAA IDF')
  ) |>
  ggplot(aes(candidate, rmse, colour = programme_key, group = programme_key)) +
  geom_point(size = 2.2) + geom_line() + facet_wrap(~scheme, scales = 'free_y') +
  labs(x = NULL, y = 'Held-out mortality RMSE', colour = 'Programme') +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Adding structure worsens both schemes for LTMP and manta. NOAA plus structure
improves MMP event-held-out RMSE from about 0.235 to 0.221, but worsens its
reef-blocked RMSE and is not reproduced with IMOS DHW.

The matched MHW complete-case sensitivity retains a sizeable manta event-held-
out improvement (RMSE change -0.0073) but is approximately neutral under reef
blocking (+0.0004). LTMP's complete-case changes are small and inconsistent.
MMP event-held-out fits are unsupported because at least one training fold has
fewer than eight positive rows or only one occurrence class.

## Logger validation of the cooling proxy

```{r}
logger_validation |>
  transmute(Comparison = comparison, N = n, `Spearman rho` = spearman_rho) |>
  knitr::kable(digits = 3)
```

The comparison uses 10 AIMS water-quality logger sites across nine event years.
Hourly clock values are interpreted as local GBR time, as encoded by the
source CSV. Logger metrics are mean daily range and median afternoon-to-
following-predawn decline in a temperature-defined 30-day peak window. Neither
the raw nor event-centred relationship supports the satellite contrast as a
proxy for nocturnal reef cooling.

## Intensity-duration-frequency sensitivity

The NOAA-native IDF prototype uses maximum mean CoralTemp nighttime SST over
1, 3, 7, 14, 28, 56 and 84 days. Return periods are smoothed empirical
exceedance frequencies against a fixed 1986--2012 reference, which predates all
validation events. The candidate contains mean log return period across
durations and a long-minus-short log-return-period contrast. It does not fit a
copula or extrapolate beyond its 28-year empirical ceiling.

```{r}
idf <- metrics |>
  filter(candidate %in% c('noaa_dhw', 'noaa_plus_idf_frequency')) |>
  select(programme_key, scheme, candidate, rmse) |>
  tidyr::pivot_wider(names_from = candidate, values_from = rmse) |>
  mutate(change = noaa_plus_idf_frequency - noaa_dhw)
idf |>
  transmute(
    Programme = toupper(programme_key), Validation = scheme,
    `NOAA RMSE` = noaa_dhw, `NOAA + IDF RMSE` = noaa_plus_idf_frequency,
    `IDF - NOAA RMSE` = change
  ) |>
  knitr::kable(digits = 4)
```

The clearest gain is MMP reef-blocked RMSE (-0.0154); LTMP reef-blocked RMSE
also improves slightly (-0.0008). Leave-one-event-out RMSE worsens for LTMP
(+0.0045) and manta (+0.0070) and is essentially unchanged for MMP (-0.0002).
This split is consistent with a spatial rarity surface that may aid within-
event interpolation but does not transfer reliably to a new thermal regime.

## Full-data importance and recommendation

In the combined full-data mechanism pool, NOAA DHW ranks above IMOS DHW for
direct mortality and positive magnitude in every programme. IMOS DHW ranks
above NOAA for LTMP occurrence, consistent with its Brier-score improvement.
The PBD12 field ranks second to fifth in several apparent full-data BRT
components, illustrating why apparent importance is not promotion evidence:
its held-out gains are not stable. Some MHW, acute-tail and hot-night terms
also receive nontrivial apparent importance. MHW duration/intensity merit a
small formal assessment only after
higher-coverage extraction. The present cooling contrast should not advance;
NOAA IDF should remain a within-event BRT sensitivity.

Retain local-first NOAA DHW. Do not advance PBD12 to formal INLA assessment on
these results. Before another formal thermal candidate, evaluate official
GeoPolar/AusTemp daily SST against this reconstruction. Broader
subdaily logger coverage or a different remotely sensed cooling construct is
required before revisiting night cooling as a reef-wide predictor.

Primary evidence is in `output/imos_thermal_screen/`: `source_comparison.csv`,
`blocked_metrics.csv`, `complete_case_metrics.csv`, `coverage_audit.csv` and
`full_data_importance.csv`; logger evidence is in
`logger_cooling_validation.csv`, and the IDF construction manifest is
`data/processed/noaa_idf_metrics_validation.manifest.json`. The literature rationale is in
`docs/thermal-exposure-metric-research.md`.
