---
title: "Thermal duration, intensity, rarity and cooling assessment"
date: 2026-09-15
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(readr)
})
root <- normalizePath('../..', winslash = '/', mustWork = TRUE)
out <- file.path(root, 'output', 'imos_thermal_screen')
coverage <- read_csv(file.path(out, 'coverage_audit.csv'), show_col_types = FALSE)
metrics <- read_csv(file.path(out, 'blocked_metrics.csv'), show_col_types = FALSE)
source <- read_csv(file.path(out, 'source_comparison.csv'), show_col_types = FALSE)
logger_validation <- read_csv(
  file.path(out, 'logger_cooling_validation.csv'), show_col_types = FALSE
)
```

## Decision

Do **not** replace local-first NOAA DHW or promote the MHW, cooling, or IDF
blocks from this screen.

Coverage-normalised IMOS/SSTAARS DHW produced small leave-one-event-out RMSE
improvements for LTMP and manta tow, but worsened reef-blocked performance and
both MMP comparisons. The observed-days-only sensitivity did not change that
cross-programme conclusion. Duration, three-day peak and persistent-hot-night
terms did not improve LTMP or manta transfer.

The standard MHW duration/intensity block is the most promising new construct.
It modestly improves LTMP in both full-row schemes and improves manta and MMP
leave-one-event-out RMSE, but the latter gains do not reproduce under reef
blocking. Matched complete-case results leave only the manta event-held-out
gain clearly favourable. MMP complete-case event folds remain underpowered.

The paired cooling proxy fails more directly. It worsens LTMP and manta, its
apparent MMP event-held-out gain reverses on observed complete cases, and it
does not track peak-window cooling in the available hourly logger subset.

The NOAA empirical intensity-duration-frequency block improves reef-blocked
LTMP and MMP performance but worsens unseen-event transfer for LTMP and manta.
This is relevant to the retained within-event BRT, not the initial forecast.

## Exposure contract

The source-control metric reconstructs NOAA-style DHW from IMOS 0.02-degree
nighttime SST and the SSTAARS 1992--2016 nighttime climatology:

1. derive maximum monthly mean (MMM) from the SSTAARS harmonic fit;
2. calculate HotSpot as bias-corrected nighttime SST minus SSTAARS MMM;
3. retain the full HotSpot only when it is at least 1 degree C;
4. sum over 84 calendar days, divide by seven, and take the calendar-year
   maximum to match `ann_maxdhw`.

Short internal gaps of at most two days are interpolated. The primary metric is
coverage-normalised only when at least 70% of an 84-day window is available;
the unscaled observed-days estimate is retained as a sensitivity. The centre
pixel is primary, with a quality-controlled fallback within 3 km and an
absolute 5 km reef-distance cap.

The coral-threshold block uses the fixed 1 November--30 April window: longest run at
or above MMM + 1 degree C, maximum three-day HotSpot, and number of hot nights.
Extreme dose above MMM + 2 degrees C appears only in the descriptive full-data
importance pool.

Standard MHW events require at least five days above a daily SSTAARS 90th
percentile threshold; qualifying events separated by no more than two days are
joined. The daily threshold is periodic interpolation of the published twelve
SSTAARS mid-month percentiles, and categories use the official median/90th-
percentile formula. The primary block contains maximum duration and maximum
intensity.

Cooling pairs observed daytime skin SST on local date D with observed
nighttime skin SST on D+1. A site-event needs at least five hot-day pairs and
50% pair coverage. No interpolation or 14-day mosaic enters this contrast.

## Coverage

```{r}
coverage |>
  mutate(
    dhw_fraction = finite_imos_dhw / reef_events,
    structure_fraction = finite_structure / reef_events
  ) |>
  select(
    Event = year, `Reef-events` = reef_events, `Finite DHW` = finite_imos_dhw,
    `DHW fraction` = dhw_fraction, `Finite structure` = finite_structure,
    `Structure fraction` = structure_fraction,
    `Finite MHW` = finite_mhw, `Finite cooling` = finite_cooling,
    `Finite NOAA IDF` = finite_idf_frequency,
    `Median hot-pair coverage` = median_hot_pair_coverage,
    `Median event coverage` = median_event_coverage
  ) |>
  knitr::kable(digits = 3)
```

IMOS DHW is available for 283 of 304 reef-events. SSTAARS has no finite water
pixel within 5 km for the other cases, mainly repeated far-northern reefs. Only
211 reef-events meet the MHW/structure coverage rule, including 55 of 103 in
2024. The observed-pair cooling contrast is available for 260 reef-events. The
DHW source test is usable, but duration results remain conservative pending a
higher-coverage GeoPolar/AusTemp assessment.

## NOAA versus IMOS-equivalent DHW

Negative RMSE and Brier changes favour IMOS; positive R-squared changes favour
IMOS.

```{r}
source |>
  transmute(
    Programme = toupper(programme_key), Validation = scheme,
    `NOAA RMSE` = rmse_noaa_dhw, `IMOS RMSE` = rmse_imos_dhw,
    `IMOS - NOAA RMSE` = rmse_change_imos_minus_noaa,
    `Observed IMOS - NOAA RMSE` = rmse_change_observed_imos_minus_noaa,
    `IMOS - NOAA R2` = r2_change_imos_minus_noaa,
    `IMOS - NOAA Brier` = brier_change_imos_minus_noaa,
    `IMOS - NOAA severe RMSE` = severe_rmse_change_imos_minus_noaa
  ) |>
  knitr::kable(digits = 4)
```

The strongest favourable result is occurrence discrimination: scaled IMOS DHW
reduces LTMP leave-one-event-out Brier score by about 0.05. That does not become
a stable mortality-scale gain. Manta event-held-out RMSE changes by less than
0.001, while MMP and reef-blocked comparisons favour NOAA.

## Candidate blocks

```{r}
metrics |>
  mutate(
    programme_key = toupper(programme_key),
    candidate = recode(candidate,
      noaa_dhw = 'NOAA', imos_dhw = 'IMOS scaled',
      imos_dhw_observed = 'IMOS observed',
      noaa_plus_structure = 'NOAA + structure',
      imos_plus_structure = 'IMOS + structure',
      noaa_plus_mhw = 'NOAA + MHW', imos_plus_mhw = 'IMOS + MHW',
      noaa_plus_cooling = 'NOAA + cooling',
      noaa_plus_compact = 'NOAA + MHW + cooling',
      noaa_plus_id_profile = 'NOAA + IMOS ID profile',
      noaa_plus_idf_frequency = 'NOAA + NOAA IDF')
  ) |>
  ggplot(aes(candidate, rmse, colour = programme_key, group = programme_key)) +
  geom_point(size = 2.2) + geom_line() + facet_wrap(~scheme, scales = 'free_y') +
  labs(x = NULL, y = 'Held-out mortality RMSE', colour = 'Programme') +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Adding structure worsens both schemes for LTMP and manta. NOAA plus structure
improves MMP event-held-out RMSE from about 0.235 to 0.221, but worsens its
reef-blocked RMSE and is not reproduced with IMOS DHW.

The matched MHW complete-case sensitivity retains a sizeable manta event-held-
out improvement (RMSE change -0.0073) but is approximately neutral under reef
blocking (+0.0004). LTMP's complete-case changes are small and inconsistent.
MMP event-held-out fits are unsupported because at least one training fold has
fewer than eight positive rows or only one occurrence class.

## Logger validation of the cooling proxy

```{r}
logger_validation |>
  transmute(Comparison = comparison, N = n, `Spearman rho` = spearman_rho) |>
  knitr::kable(digits = 3)
```

The comparison uses 10 AIMS water-quality logger sites across nine event years.
Hourly clock values are interpreted as local GBR time, as encoded by the
source CSV. Logger metrics are mean daily range and median afternoon-to-
following-predawn decline in a temperature-defined 30-day peak window. Neither
the raw nor event-centred relationship supports the satellite contrast as a
proxy for nocturnal reef cooling.

## Intensity-duration-frequency sensitivity

The NOAA-native IDF prototype uses maximum mean CoralTemp nighttime SST over
1, 3, 7, 14, 28, 56 and 84 days. Return periods are smoothed empirical
exceedance frequencies against a fixed 1986--2012 reference, which predates all
validation events. The candidate contains mean log return period across
durations and a long-minus-short log-return-period contrast. It does not fit a
copula or extrapolate beyond its 28-year empirical ceiling.

```{r}
idf <- metrics |>
  filter(candidate %in% c('noaa_dhw', 'noaa_plus_idf_frequency')) |>
  select(programme_key, scheme, candidate, rmse) |>
  tidyr::pivot_wider(names_from = candidate, values_from = rmse) |>
  mutate(change = noaa_plus_idf_frequency - noaa_dhw)
idf |>
  transmute(
    Programme = toupper(programme_key), Validation = scheme,
    `NOAA RMSE` = noaa_dhw, `NOAA + IDF RMSE` = noaa_plus_idf_frequency,
    `IDF - NOAA RMSE` = change
  ) |>
  knitr::kable(digits = 4)
```

The clearest gain is MMP reef-blocked RMSE (-0.0154); LTMP reef-blocked RMSE
also improves slightly (-0.0008). Leave-one-event-out RMSE worsens for LTMP
(+0.0045) and manta (+0.0070) and is essentially unchanged for MMP (-0.0002).
This split is consistent with a spatial rarity surface that may aid within-
event interpolation but does not transfer reliably to a new thermal regime.

## Full-data importance and recommendation

In the combined full-data mechanism pool, NOAA DHW ranks above IMOS DHW for
direct mortality and positive magnitude in every programme. IMOS DHW ranks
above NOAA for LTMP occurrence, consistent with its Brier-score improvement.
Some MHW, acute-tail and hot-night terms receive nontrivial apparent
importance. MHW duration/intensity merit a small formal assessment only after
higher-coverage extraction. The present cooling contrast should not advance;
NOAA IDF should remain a within-event BRT sensitivity.

Retain local-first NOAA DHW. Before another formal model candidate, evaluate
official GeoPolar/AusTemp daily SST against this reconstruction. Broader
subdaily logger coverage or a different remotely sensed cooling construct is
required before revisiting night cooling as a reef-wide predictor.

Primary evidence is in `output/imos_thermal_screen/`: `source_comparison.csv`,
`blocked_metrics.csv`, `complete_case_metrics.csv`, `coverage_audit.csv` and
`full_data_importance.csv`; logger evidence is in
`logger_cooling_validation.csv`, and the IDF construction manifest is
`data/processed/noaa_idf_metrics_validation.manifest.json`. The literature rationale is in
`docs/thermal-exposure-metric-research.md`.
