---
title: INLA severe-tail, WQC and Lizard logger assessment
subtitle: Blocked predictive validation for GBR bleaching mortality
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)
library(tidyr)

theme_set(theme_minimal(base_size = 12))
model_colours <- c(
  'INLA only' = '#1f4e79',
  'Conservative tail (0.5 gate)' = '#d97706',
  'Soft-gated tail' = '#b91c1c'
)

pred <- read_csv('../../output/inla_severe_tail/cv_predictions.csv', show_col_types = FALSE)
metrics <- read_csv('../../output/inla_severe_tail/metrics.csv', show_col_types = FALSE)
events <- read_csv('../../output/inla_severe_tail/event_metrics.csv', show_col_types = FALSE)
influence <- read_csv('../../output/inla_severe_tail/production_influence.csv', show_col_types = FALSE)
inla_fixed <- read_csv('../../output/inla_spatiotemporal/full_fixed_effects.csv', show_col_types = FALSE)
logger <- read_csv('../data/processed/lizard_cluster_logger_noaa_daily_2024.csv', show_col_types = FALSE)
logger_summary <- read_csv('../data/processed/lizard_cluster_logger_noaa_summary_2024.csv', show_col_types = FALSE)
logger_monthly <- read_csv('../data/processed/lizard_cluster_logger_noaa_monthly_2024.csv', show_col_types = FALSE)
logger_dhw <- read_csv('../data/processed/lizard_cluster_logger_dhw_uplift_2024.csv', show_col_types = FALSE)
cyclone <- read_csv('../data/processed/bom_cyclone_reef_year.csv', show_col_types = FALSE)
reef_event <- read_csv('../../output/inla_severe_tail/reef_event_predictions.csv', show_col_types = FALSE)
miss_registry <- read_csv('../data/curated/extreme_miss_explanations.csv', show_col_types = FALSE)
disease <- read_csv('../data/processed/noaa_disease_risk_validation.csv', show_col_types = FALSE)
inla_compare <- read_csv('../../output/inla_spatiotemporal/model_comparison.csv', show_col_types = FALSE)

display <- bind_rows(
  metrics |> filter(programme_key == 'pooled', learner == 'INLA only'),
  metrics |>
    filter(
      programme_key == 'pooled',
      tail_features == 'wqc_threshold_track_tail',
      learner %in% c('Conservative tail (0.5 gate)', 'Soft-gated tail')
    )
)
```

## Main result

The severe-tail learner helps the problem it was designed for, but it is not a
universal replacement for the INLA expectation. Under reef-blocked validation,
the conservative correction raises mean prediction for observed mortality of
at least 50% from 35.9% to 39.2%, while pooled RMSE remains 0.136. The soft
gate raises the severe mean to 42.5% and lowers severe-case RMSE from 0.325 to
0.277, but increases ordinary false extremes and pooled RMSE.

When an entire event is held out, every tail correction worsens pooled RMSE.
This is a harsh test for a compound driver first strongly expressed in 2024,
but remains a real warning. The tail expert should be an event-risk ensemble
member rather than a silent replacement for the calibrated INLA mean.

```{r}
display |>
  transmute(
    Validation = recode(
      scheme,
      reef_blocked_5fold = 'New reefs within observed events',
      leave_one_event_out = 'Entire event held out'
    ),
    Model = learner, RMSE = round(rmse, 3), MAE = round(mae, 3),
    `Predictive R2` = round(predictive_r2, 3),
    `Severe observed` = percent(severe_observed, accuracy = 0.1),
    `Severe predicted` = percent(severe_predicted, accuracy = 0.1),
    `Severe RMSE` = round(severe_rmse, 3),
    `False extreme rate` = percent(false_extreme_rate, accuracy = 0.1)
  ) |>
  kable()
```

## Lizard-cluster in-water temperatures

The public AIMS daily endpoint provides 2024-event records for Lizard Island
and North Direction reef-flat and reef-slope loggers. Linnet has logger
deployments only through 2018, so it cannot directly test the 2024 event.

Season-wide means conceal a material January mismatch. Mean logger-minus-NOAA
differences over the full record are only -0.04 to +0.07 degrees C, but January
means are +0.08 to +0.44 degrees C. The signed January offset integrates to
0.37--1.94 degree-weeks and the positive-only offset to 0.87--2.05
degree-weeks. These are temperature-offset integrals rather than formal extra
DHW because the daily differences have not yet been evaluated against each
site's climatological MMM and HotSpot threshold. Nevertheless they show that
a local heat mismatch of ecologically meaningful magnitude is plausible and
should be tested explicitly for the Lizard cluster.

Using NOAA's v3.1 MMM and standard 84-day rule changes the interpretation
substantially. The reconstructed NOAA series agrees with the official daily
DHW peak within 0.18 DHW. Logger-derived peaks imply an uplift of 1.05--5.38
DHW relative to official NOAA across the full November--April bleaching
period. The median regional uplift is 3.85 DHW; the mean is 4.82 DHW for flats
and 2.25 DHW for slopes. This depth dependence should be retained in any local
sensitivity rather than applying the flat value to 9--10 m observations.

```{r}
logger_summary |>
  transmute(
    Site = site, Series = series, Context = logger_depth_context,
    Days = matched_days,
    `Mean logger minus NOAA (C)` = round(mean_logger_minus_noaa, 2),
    `Warm-quartile difference (C)` = round(mean_delta_on_warmest_noaa_quartile, 2),
    `Maximum 7-day difference (C)` = round(max_7day_logger_minus_noaa, 2)
  ) |>
  kable()
```

```{r}
logger_monthly |>
  filter(month == '2024-01') |>
  transmute(
    Site = site, Series = series, Days = matched_days,
    `January mean logger minus NOAA (C)` = round(mean_logger_minus_noaa, 2),
    `Signed offset (degree-weeks)` = round(signed_offset_degree_weeks, 2),
    `Positive-only offset (degree-weeks)` = round(positive_offset_degree_weeks, 2),
    `Maximum 7-day difference (C)` = round(max_7day_logger_minus_noaa, 2)
  ) |>
  kable()
```

```{r}
logger_dhw |>
  mutate(Context = case_when(
    grepl('FL', series) ~ 'Reef flat',
    TRUE ~ 'Reef slope'
  )) |>
  transmute(
    Site = site, Series = series, Context,
    MMM = round(noaa_mmm, 2),
    `Logger peak DHW` = round(logger_max_dhw, 2),
    `Official NOAA peak DHW` = round(noaa_official_max_dhw, 2),
    `DHW uplift` = round(dhw_uplift_vs_official_noaa, 2),
    `Logger peak date` = logger_max_dhw_date
  ) |>
  kable()
```

```{r}
logger |>
  select(date, site, series, logger_temperature, noaa_sst) |>
  pivot_longer(c(logger_temperature, noaa_sst), names_to = 'source', values_to = 'temperature') |>
  ggplot(aes(as.Date(date), temperature, colour = source)) +
  geom_line(linewidth = 0.45) +
  facet_wrap(~site + series, ncol = 2, scales = 'free_y') +
  scale_colour_manual(
    values = c(logger_temperature = '#b91c1c', noaa_sst = '#1f4e79'),
    labels = c(logger_temperature = 'AIMS logger', noaa_sst = 'NOAA CoralTemp')
  ) +
  labs(x = NULL, y = 'Daily temperature (degrees C)', colour = NULL) +
  theme(legend.position = 'top')
```

```{r}
ggplot(display, aes(learner, severe_predicted, colour = learner)) +
  geom_hline(aes(yintercept = severe_observed), linetype = 2, colour = 'grey35') +
  geom_point(size = 3) +
  facet_wrap(~scheme, labeller = as_labeller(c(
    reef_blocked_5fold = 'Reef blocked',
    leave_one_event_out = 'Event held out'
  ))) +
  scale_colour_manual(values = model_colours) +
  scale_y_continuous(labels = percent, limits = c(0, 0.7)) +
  labs(
    x = NULL, y = 'Mean prediction for observed mortality at least 50%',
    caption = 'Dashed line is the observed severe-case mean.'
  ) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1), legend.position = 'none')
```

## WQC threshold and cumulative exposure

RRN coloured-water frequency is on a zero-to-one scale, so the apparent
breakpoint near 50 in the earlier report is encoded as a hinge above 0.50. The
model carries raw WQC, current excess above 0.50, the rolling ten-year sum of
annual excess above 0.50, and their DHW interactions.

High WQC observations are enriched for severe mortality: 14 of 76 observations
at or above 0.50 have at least 50% mortality, compared with 18 of 461 below
0.50. Raw WQC is influential in the severe gate. However, the threshold and
cumulative terms do not improve pooled blocked RMSE over the otherwise
identical tail BRT. They remain useful candidates, but the current screen does
not establish that the 0.50 hinge alone solves the severe misses.

```{r}
wqc_bins <- pred |>
  filter(scheme == 'reef_blocked_5fold', learner == 'INLA only') |>
  mutate(wqc_group = cut(
    wqc_freqcc12, breaks = c(-Inf, 0.10, 0.25, 0.50, 0.75, Inf),
    labels = c('<0.10', '0.10 to 0.25', '0.25 to 0.50', '0.50 to 0.75', '>0.75')
  )) |>
  group_by(wqc_group) |>
  summarise(
    n = n(), observed = mean(observed_mortality),
    predicted = mean(predicted_mortality), .groups = 'drop'
  ) |>
  pivot_longer(c(observed, predicted), names_to = 'series', values_to = 'mortality')

ggplot(wqc_bins, aes(wqc_group, mortality, colour = series, group = series)) +
  geom_line() + geom_point(aes(size = n)) +
  scale_colour_manual(values = c(observed = 'black', predicted = '#1f4e79')) +
  scale_y_continuous(labels = percent) +
  labs(
    x = 'RRN wet-season coloured-water frequency', y = 'Mean mortality',
    colour = NULL, size = 'Observations'
  )
```

```{r}
influence |>
  filter(tail_features == 'wqc_threshold_track_tail') |>
  group_by(component) |>
  slice_max(relative_influence, n = 12, with_ties = FALSE) |>
  ungroup() |>
  mutate(predictor = reorder(gsub('_z$', '', predictor), relative_influence)) |>
  ggplot(aes(relative_influence, predictor, fill = component)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~component, scales = 'free_y') +
  labs(x = 'Relative BRT influence', y = NULL)
```

## Cyclone-track and COTS in the central INLA model

Adding provisional cyclone-track proximity, the wind-distance index and the
GBR COTS outbreak-probability hindcast improves both principal validation
schemes. Reef-blocked RMSE falls from 0.1369 to 0.1357 and predictive R2 rises
from 40.5% to 41.5%. Leave-event-out RMSE falls from 0.2008 to 0.1973. The
severe-case mean increases from 35.4% to 35.9%. This is a small but consistent
gain, so this pressure model is now the central INLA baseline used by the tail
BRT. The RRN wave and COTS fields remain in the model; these new variables add
track geometry and the independent outbreak hindcast.

```{r}
inla_compare |>
  filter(candidate %in% c(
    'persistent_rw1_full_partial_pool',
    'persistent_rw1_cyclone_cots_partial_pool',
    'persistent_rw1_cyclone_cots_disease_partial_pool'
  )) |>
  mutate(Model = recode(
    candidate,
    persistent_rw1_full_partial_pool = 'Previous INLA baseline',
    persistent_rw1_cyclone_cots_partial_pool = 'Add cyclone track + COTS',
    persistent_rw1_cyclone_cots_disease_partial_pool = 'Also add disease risk'
  )) |>
  transmute(
    Model, Validation = recode(
      scheme, reef_blocked_5fold = 'Reef blocked',
      leave_one_event_out = 'Event held out'
    ),
    RMSE = round(rmse, 3), `Predictive R2` = round(predictive_r2, 3),
    `Severe predicted` = percent(severe_predicted, accuracy = 0.1),
    `Severe RMSE` = round(severe_rmse, 3),
    `False extreme rate` = percent(false_extreme_rate, accuracy = 0.1),
    WAIC = round(waic, 1)
  ) |>
  arrange(Validation, Model) |>
  kable()
```

```{r}
inla_fixed |>
  filter(
    candidate == 'persistent_rw1_cyclone_cots_partial_pool',
    term %in% c(
      'tc_proximity100_z', 'tc_wind_distance_index_z',
      'cots_outbreak_probability_z'
    )
  ) |>
  transmute(
    Term = recode(
      term,
      tc_proximity100_z = 'Cyclone-track proximity',
      tc_wind_distance_index_z = 'Cyclone wind-distance index',
      cots_outbreak_probability_z = 'COTS outbreak probability'
    ),
    Mean = round(mean, 3), SD = round(sd, 3),
    `2.5%` = round(`0.025quant`, 3), `97.5%` = round(`0.975quant`, 3)
  ) |>
  kable()
```

## NOAA disease-risk screen

The disease product is structurally selective, not a continuous GBR-wide
covariate. It is applicable at 23 of 73 validation reef-years in 2020 and only
three of 61 in 2022; all validation locations are masked by the product's
winter prerequisite in 2016, 2017 and 2024. Negative mask codes were therefore
kept distinct from valid zero risk during extraction.

```{r}
disease |>
  group_by(year) |>
  summarise(
    `Reef-years` = n(),
    Applicable = sum(disease_risk_applicable),
    `Maximum risk` = max(disease_risk_max),
    `Days at risk >= 1 (maximum)` = max(disease_risk_days_ge1),
    .groups = 'drop'
  ) |>
  mutate(`Maximum risk` = round(`Maximum risk`, 2)) |>
  kable()
```

Adding disease intensity and duration to INLA worsens leave-event-out RMSE
from 0.197 to 0.211 and raises the false-extreme rate from 11.3% to 18.3%.
Adding the same features and a DHW interaction to the tail BRT also weakens
reef-blocked RMSE. The raw disease features receive zero production BRT
influence; only their interaction receives small influence. The metric is
ecologically relevant for interpreting Gannett Cay and other 2020 residuals,
but it is not promoted into the operational ensemble on this evidence.

```{r}
metrics |>
  filter(
    programme_key == 'pooled',
    tail_features %in% c(
      'wqc_threshold_track_tail',
      'wqc_threshold_track_disease_tail'
    ),
    learner == 'Conservative tail (0.5 gate)'
  ) |>
  transmute(
    Validation = recode(
      scheme, reef_blocked_5fold = 'Reef blocked',
      leave_one_event_out = 'Event held out'
    ),
    Model = recode(
      tail_features,
      wqc_threshold_track_tail = 'Tail BRT without disease',
      wqc_threshold_track_disease_tail = 'Tail BRT with disease'
    ),
    RMSE = round(rmse, 3), `Predictive R2` = round(predictive_r2, 3),
    `Severe predicted` = percent(severe_predicted, accuracy = 0.1),
    `Severe RMSE` = round(severe_rmse, 3)
  ) |>
  kable()
```

## 2024 manta-tow severe cases

For 2024 manta tow, the reef-blocked soft WQC-plus-track tail raises the severe
mean from 33.1% to 39.3% and lowers RMSE from 0.163 to 0.158. It still falls well
short of the observed severe mean of 58.0%.

```{r}
events |>
  filter(
    event_year == 2024, programme_key == 'manta',
    learner == 'INLA only' |
      tail_features == 'wqc_threshold_track_tail'
  ) |>
  transmute(
    Validation = recode(
      scheme, reef_blocked_5fold = 'Reef blocked',
      leave_one_event_out = 'Event held out'
    ),
    Model = learner, RMSE = round(rmse, 3), Bias = round(bias, 3),
    `Severe predicted` = percent(severe_predicted, accuracy = 0.1),
    `Severe observed` = percent(severe_observed, accuracy = 0.1),
    `False extreme rate` = percent(false_extreme_rate, accuracy = 0.1)
  ) |>
  kable()
```

## Cyclone exposure is provisional

The current best-track block uses nearest-track distance, reported maximum
wind and a wind-by-distance index alongside RRN hours above 4 m waves. It gives
a clear Jasper proximity signal at Mackay and Snapper even where the RRN wave
field is weak. These fields are isolated in the fitting script because a better
cyclone data set resolving Jasper and Kirrily is expected and should replace
this block before the operational fit.

```{r}
cyclone |>
  filter(event_year == 2024, grepl('Mackay|Snapper', ReefName, ignore.case = TRUE)) |>
  arrange(tc_min_distance_km) |>
  transmute(
    Reef = ReefName, `Nearest cyclone` = tc_nearest_name,
    `Track distance (km)` = round(tc_min_distance_km, 1),
    `Nearest max wind (m/s)` = round(tc_nearest_max_wind_ms, 1),
    `Storms within 300 km` = tc_storms_within300km
  ) |>
  head(12) |>
  kable()
```

## Remaining severe misses

Repeated observations are useful to the likelihood but should not be confused
with independent unexplained reef events. Snapper contributes three of the top
five observation-level misses. The table below therefore averages predictions
within reef, event and programme before ranking; the fit and its inclusion
logic are unchanged.

```{r}
reef_event |>
  filter(
    scheme == 'reef_blocked_5fold',
    tail_features == 'wqc_threshold_track_tail',
    learner == 'Soft-gated tail', observed_mortality >= 0.50
  ) |>
  arrange(desc(underprediction)) |>
  transmute(
    Programme = toupper(programme_key), Reef = ReefName, Year = event_year,
    Observations = observations,
    DHW = round(ann_maxdhw, 1), Observed = percent(observed_mortality, accuracy = 0.1),
    Predicted = percent(predicted_mortality, accuracy = 0.1),
    Underprediction = percent(underprediction, accuracy = 0.1),
    WQC = round(wqc_freqcc12, 2),
    `Cyclone distance (km)` = round(tc_min_distance_km, 0)
  ) |>
  head(15) |>
  kable()
```

## Ecological explanation register

The high-priority misses now have a versioned explanation register. The labels
are diagnostic evidence and do not silently remove observations or change the
disturbance logic. Gannett Cay is especially informative: its 2020 COTS
outbreak probability is 0.934 and the independent RRN IDW COTS estimate is
0.839, so the severe-tail screen now includes the hindcast probability as well
as the RRN pressure.

```{r}
miss_registry |>
  transmute(
    Reef = reef_name, Year = event_year,
    `Likely explanation` = primary_explanation,
    `Evidence status` = evidence_status,
    `Current signal` = current_numeric_signal,
    `Next model action` = model_action
  ) |>
  kable()
```

## Operational decision

1. Use the partially pooled cyclone-track plus COTS INLA as the calibrated
   central estimate; retain the previous baseline as an ablation.
2. Carry the conservative tail BRT as an ensemble member for within-event GBR
   mapping; expose the soft gate as a high-risk scenario.
3. Retain raw WQC, the 0.50 hinge and cumulative exposure, without yet claiming
   a validated threshold mechanism.
4. Replace the provisional cyclone block with the forthcoming
   Jasper/Kirrily-resolving data, then repeat the unchanged blocked comparison.
5. Do not apply a blanket DHW uplift near Lizard. Instead calculate a
   logger-consistent HotSpot/DHW diagnostic using the local MMM; the January
   offset is large enough to test as a candidate explanation.
6. Maintain the reef-event explanation register and expose both observation-
   level and reef-event-level miss rankings in formal validation.
7. Retain NOAA disease risk for ecological attribution and sensitivity plots,
   but do not include it in the operational ensemble unless new validation
   shows stable gains.
