---
title: 'INLA spatial and spatio-temporal mortality screen'
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)
library(tidyr)

out <- '../../output/inla_spatiotemporal'
comparison <- read_csv(file.path(out, 'model_comparison.csv'), show_col_types = FALSE)
predictions <- read_csv(file.path(out, 'cv_predictions.csv'), show_col_types = FALSE)
fixed <- read_csv(file.path(out, 'full_fixed_effects.csv'), show_col_types = FALSE)
hyper <- read_csv(file.path(out, 'full_hyperparameters.csv'), show_col_types = FALSE)
slopes <- read_csv(file.path(out, 'temporal_dhw_slopes.csv'), show_col_types = FALSE)
fields <- read_csv(file.path(out, 'spatial_field_nodes.csv'), show_col_types = FALSE)
deviations <- read_csv(
  file.path(out, 'partial_pool_deviations.csv'), show_col_types = FALSE
)
mesh_summary <- read_csv(file.path(out, 'mesh_summary.csv'), show_col_types = FALSE)

candidate_labels <- c(
  nonspatial = 'No spatial field',
  persistent_spatial = 'Persistent spatial field',
  event_spatial_iid = 'Event fields: independent',
  event_spatial_ar1 = 'Event fields: AR1',
  persistent_plus_dhw_rw1 = 'Persistent field + varying DHW slope',
  persistent_rw1_current = 'Add current speed',
  persistent_rw1_current_sst_chla = 'Add current + SST shape + Chl-a',
  persistent_rw1_full_partial_pool = 'Full features + partial pooling'
)
scheme_labels <- c(
  reef_blocked_5fold = 'New reefs within observed events',
  leave_one_event_out = 'Entire event held out'
)
programme_labels <- c(ltmp = 'LTMP', manta = 'Manta tow', mmp = 'MMP')
theme_set(theme_minimal(base_size = 12))

all_metrics <- comparison |>
  mutate(
    Model = factor(recode(candidate, !!!candidate_labels),
                   levels = unname(candidate_labels)),
    Validation = recode(scheme, !!!scheme_labels)
  )

event_summary <- predictions |>
  group_by(candidate, scheme, event_year) |>
  summarise(
    n = n(),
    observed_mean = mean(observed_mortality),
    predicted_mean = mean(predicted_mortality),
    rmse = sqrt(mean((observed_mortality - predicted_mortality)^2)),
    predictive_r2 = 1 -
      sum((observed_mortality - predicted_mortality)^2) /
      sum((observed_mortality - mean(observed_mortality))^2),
    severe_n = sum(observed_mortality >= 0.50),
    severe_observed = if_else(
      severe_n > 0,
      mean(observed_mortality[observed_mortality >= 0.50]), NA_real_
    ),
    severe_predicted = if_else(
      severe_n > 0,
      mean(predicted_mortality[observed_mortality >= 0.50]), NA_real_
    ),
    .groups = 'drop'
  )
```

## Decision

Use **INLA as the main bounded-response driver**, with a Bernoulli model for
whether mortality occurred and a conditional beta model for positive
mortality. Carry the persistent spatial field into the operational GBR model.
It improves prediction at unsurveyed reefs and can represent persistent
unmeasured susceptibility or buffering.

Keep independent event-specific spatial fields as a diagnostic and nowcasting
extension. They are marginally better when other reefs from the same event are
available, but their latent event surface is not known before mortality data
arrive. The AR1 event field should not be the default: the estimated correlation
between adjacent event surfaces is negative, and the five events are unevenly
spaced in time.

Restore current speed to the operational ecological core and carry summer SST
skewness, summer excess kurtosis and wet-season median chlorophyll as a grouped,
regularised candidate. Together they improve pooled reef-blocked RMSE to 13.9%
and predictive R2 to 38.4%. Partially pooling the five prespecified DHW
interactions improves these further to 13.7% and 40.5%. Individual coefficients
remain uncertain and held-event gains vary among programmes, so predictions
take precedence over single-coefficient claims.

The time-varying DHW response is worth retaining as a sensitivity, but it does
not currently demonstrate adaptation. Its event deviations and the direct
DHW-by-repeat-exposure interaction remain uncertain. Repeat exposure could
also represent acclimatisation, community filtering, selective mortality or
changes in observation coverage.

## What was fitted

The screen uses 537 observations, 139 reefs, 304 reef-events and five mass
bleaching events (2016, 2017, 2020, 2022 and 2024). Six likelihood columns are
fitted jointly: occurrence and positive magnitude for LTMP, Manta tow and MMP.
Programme/component intercepts and beta precisions differ, while the ecological
coefficients and latent reef-event, event and spatial effects are shared.

This is a deliberately strong shared-response screen. The formal model should
estimate programme/component calibration loadings rather than assume every
layer has exactly the same logit-scale response. Predictions in this screen
multiply the posterior marginal means for occurrence and magnitude. Posterior
sampling should be used for the final expected mortality and uncertainty maps.

| Candidate | Scientific role | Operational use |
|---|---|---|
| No spatial field | Baseline joint bounded model | Yes |
| Persistent spatial field | Stable reef-scale susceptibility/refugia | Yes |
| Independent event fields | Event-specific residual geography | With partial event surveys |
| AR1 event fields | Temporal borrowing between event surfaces | Sensitivity only |
| Persistent field + varying DHW slope | Changing thermal response across events | Yes, if validated |
| Add current speed | Hydrodynamic buffering and DHW modification | Yes; 2024 partly climatological |
| Add current, SST shape and Chl-a | Thermal trajectory and productivity candidate | Carry into formal partial pooling |
| Full features with partial pooling | Layer deviations for five DHW interactions | Preferred within-event mapping candidate |

The mesh has `r mesh_summary$nodes` nodes. Its PC prior assigns 50% probability
that the spatial range is below 200 km and 5% probability that latent-field
standard deviation exceeds 0.7 on the shared linear predictor.

## Predictive comparison

```{r}
all_metrics |>
  ggplot(aes(rmse, Model, colour = Validation)) +
  geom_point(size = 3) +
  scale_x_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    x = 'Cross-validated RMSE', y = NULL, colour = NULL,
    title = 'Spatial fields improve transfer to new reefs',
    subtitle = 'Holding out an entire event remains a substantially harder task'
  )
```

```{r}
comparison |>
  select(candidate, scheme, rmse, predictive_r2, bias,
         severe_observed, severe_predicted, severe_rmse,
         false_extreme_rate) |>
  mutate(
    candidate = recode(candidate, !!!candidate_labels),
    scheme = recode(scheme, !!!scheme_labels),
    across(c(rmse, bias, severe_observed, severe_predicted,
             severe_rmse, false_extreme_rate), percent, accuracy = 0.1),
    predictive_r2 = number(predictive_r2, accuracy = 0.01)
  ) |>
  rename(
    Model = candidate, Validation = scheme, RMSE = rmse,
    `Predictive R2` = predictive_r2, Bias = bias,
    `Severe observed` = severe_observed,
    `Severe predicted` = severe_predicted,
    `Severe RMSE` = severe_rmse,
    `False extreme rate` = false_extreme_rate
  ) |>
  arrange(Validation, RMSE) |>
  kable()
```

For new reefs within known events, the non-spatial model explains 24.8% of
held-out variation. Spatial candidates explain 31.1% to 33.2%, with RMSE
falling from 15.4% to about 14.5%. The independent event field is the narrow
winner, but the persistent field is operationally preferable because it can be
projected before event mortality observations exist.

All candidates have negative predictive R2 when a whole event is omitted. This
is a warning about extrapolating across only five heterogeneous events, not a
reason to abandon the bounded model. The persistent field plus varying DHW
slope has the lowest held-event RMSE (21.2%, versus 22.3% without a spatial
field), but the difference is small.

## Restored current and SST-shape features

The feature comparison holds the selected latent structure fixed: a persistent
spatial field and event-varying DHW slope. It therefore tests added information,
not a change in spatial complexity.

```{r}
feature_candidates <- c(
  'persistent_plus_dhw_rw1',
  'persistent_rw1_current',
  'persistent_rw1_current_sst_chla',
  'persistent_rw1_full_partial_pool'
)

comparison |>
  filter(candidate %in% feature_candidates) |>
  mutate(
    Model = factor(recode(candidate, !!!candidate_labels),
                   levels = unname(candidate_labels[feature_candidates])),
    Validation = recode(scheme, !!!scheme_labels)
  ) |>
  ggplot(aes(Model, rmse, colour = Validation, group = Validation)) +
  geom_line() +
  geom_point(size = 3) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    x = NULL, y = 'Cross-validated RMSE', colour = NULL,
    title = 'Current speed and SST-shape features improve pooled prediction'
  ) +
  theme(axis.text.x = element_text(angle = 22, hjust = 1))
```

Current speed reduces pooled reef-blocked RMSE from 14.6% to 14.2% and
leave-event-out RMSE from 21.2% to 20.2%. Adding SST skewness, excess kurtosis
and wet-season Chl-a reduces them further to 13.9% and 19.8%. Pooled
reef-blocked predictive R2 rises from 32.7% to 38.4%. Partial pooling reaches
13.7% RMSE and 40.5% predictive R2, while leave-event-out RMSE is 20.1%--a
small loss relative to the completely shared full-feature model.

```{r}
programme_feature <- read_csv(
  file.path(out, 'cv_metrics.csv'), show_col_types = FALSE
) |>
  filter(
    candidate %in% feature_candidates,
    programme_key != 'all'
  ) |>
  mutate(
    Model = factor(recode(candidate, !!!candidate_labels),
                   levels = unname(candidate_labels[feature_candidates])),
    Programme = recode(programme_key, !!!programme_labels),
    Validation = recode(scheme, !!!scheme_labels)
  )

programme_feature |>
  ggplot(aes(Model, rmse, colour = Validation, group = Validation)) +
  geom_line() +
  geom_point(size = 2.4) +
  facet_wrap(~Programme) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    x = NULL, y = 'Cross-validated RMSE', colour = NULL,
    title = 'The feature gain is programme dependent',
    subtitle = 'MMP improves at new reefs but worsens when an entire event is held out'
  ) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

The shared full-feature candidate improves reef-blocked R2 to 40.9% for LTMP
and 25.7% for MMP, while Manta is essentially unchanged near 40%. Partial
pooling raises these to 45.7%, 26.9% and 40.3%, respectively. Under
leave-event-out validation partial pooling improves MMP relative to the shared
full model but is slightly worse for LTMP and Manta. This is a genuine mapping
versus future-event trade-off, not a universal winner.

```{r}
feature_term_labels <- c(
  mcur_90_z = 'Current speed',
  dhw_x_current = 'DHW x current speed',
  sst_summer_skewness_z = 'Summer SST skewness',
  sst_summer_excess_kurtosis_z = 'Summer SST excess kurtosis',
  chla_wetseason_median_z = 'Wet-season median Chl-a'
)

fixed |>
  filter(
    candidate == 'persistent_rw1_current_sst_chla',
    term %in% names(feature_term_labels)
  ) |>
  mutate(
    Term = factor(recode(term, !!!feature_term_labels),
                  levels = rev(unname(feature_term_labels))),
    lower = .data[['0.025quant']],
    upper = .data[['0.975quant']]
  ) |>
  ggplot(aes(mean, Term)) +
  geom_vline(xintercept = 0, colour = 'grey55') +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.18) +
  geom_point(size = 2.5, colour = '#2C7FB8') +
  labs(
    x = 'Shared standardised logit-scale coefficient', y = NULL,
    title = 'Feature group improves prediction without a decisive single coefficient'
  )
```

The current-speed main effect and DHW interaction are negative, consistent
with hydrodynamic buffering, but their 95% intervals narrowly include zero.
Excess kurtosis has the strongest new thermal-shape coefficient; skewness and
Chl-a are more uncertain. Because summer skewness and kurtosis are strongly
correlated, these coefficients should not be interpreted independently.

All 537 mortality rows contain current speed. The post-January 2024 eReefs
values were completed with a reef climatology, so 2024 current speed primarily
represents persistent exposure rather than the full event anomaly. SST-shape
and Chl-a cover 304 of 306 distinct joint reef-event keys; Mackay Reef in 2016
and 2024 is imputed from each analysis fold.

## What partial pooling changed

The agreed interactions for Acropora, novelty, freshwater/coloured water,
cloud and current retain a common ecological coefficient. Each also receives
six zero-centred deviations: occurrence and positive magnitude for LTMP, Manta
and MMP. A PC prior shrinks unnecessary deviations back toward the common
response.

```{r}
deviation_labels <- c(
  acropora = 'DHW x Acropora',
  novelty = 'DHW x novelty',
  freshwater = 'DHW x freshwater',
  cloud = 'DHW x cloud',
  current = 'DHW x current speed'
)

deviations |>
  mutate(
    Interaction = factor(
      recode(interaction, !!!deviation_labels),
      levels = rev(unname(deviation_labels))
    ),
    Programme = recode(programme_key, !!!programme_labels),
    Component = recode(
      component, occurrence = 'Occurrence', magnitude = 'Positive magnitude'
    ),
    lower = .data[['0.025quant']],
    upper = .data[['0.975quant']]
  ) |>
  ggplot(aes(mean, Interaction, colour = Programme, shape = Component)) +
  geom_vline(xintercept = 0, colour = 'grey60') +
  geom_errorbarh(
    aes(xmin = lower, xmax = upper), height = 0,
    position = position_dodge(width = 0.6), alpha = 0.55
  ) +
  geom_point(position = position_dodge(width = 0.6), size = 2) +
  labs(
    x = 'Deviation from shared interaction', y = NULL,
    colour = NULL, shape = NULL,
    title = 'Most observation-layer interaction deviations shrink toward zero'
  )
```

Most layer deviations are small and uncertain, supporting substantial sharing.
The clearest exception is a more negative novelty effect in MMP positive
mortality, while LTMP occurrence tends in the opposite direction. This is
consistent with allowing limited programme/component flexibility rather than
fitting unrelated ecological models.

Partial pooling raises the reef-blocked severe-case mean prediction from 33.1%
to 35.4% and reduces severe RMSE from 34.4% to 33.1%. The observed severe mean
is still 61.7%, so the severe-tail learner remains necessary.

## Severe mortality remains the key miss

```{r}
all_metrics |>
  filter(scheme == 'reef_blocked_5fold') |>
  ggplot(aes(Model, severe_predicted)) +
  geom_hline(aes(yintercept = severe_observed), linetype = 2,
             colour = 'grey35') +
  geom_col(fill = '#2C7FB8') +
  scale_y_continuous(labels = percent, limits = c(0, 0.70)) +
  labs(
    x = NULL, y = 'Mean mortality among observations >=50%',
    title = 'Spatial structure helps, but severe mortality is still underpredicted',
    subtitle = 'Dashed line is the observed severe-case mean (61.7%)'
  ) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

The persistent field raises the severe-case mean prediction from 26.3% to
30.5%; the varying-slope version reaches 31.0%. This remains far below the
observed 61.7%. Spatial effects therefore recover structured residual risk but
do not replace the planned severe-tail component or the nonlinear BRT expert.

For 2024 reef-blocked predictions, spatial models improve predictive R2 from
13.7% to 26.5--28.2%. Nevertheless, the 23 observations above 50% mortality
average 62.3% observed and only about 33% predicted. When 2024 is completely
held out, the persistent varying-slope model predicts the overall event mean
slightly high (21.3% versus 17.6%) while still underpredicting its severe cases
(40.0% versus 62.3%). Mean calibration can therefore conceal failure in the
upper tail.

```{r}
event_summary |>
  filter(event_year == 2024) |>
  mutate(
    Model = recode(candidate, !!!candidate_labels),
    Validation = recode(scheme, !!!scheme_labels)
  ) |>
  select(Model, Validation, n, observed_mean, predicted_mean,
         rmse, predictive_r2, severe_n, severe_observed, severe_predicted) |>
  mutate(
    across(c(observed_mean, predicted_mean, rmse,
             severe_observed, severe_predicted), percent, accuracy = 0.1),
    predictive_r2 = number(predictive_r2, accuracy = 0.01)
  ) |>
  rename(
    N = n, `Observed mean` = observed_mean,
    `Predicted mean` = predicted_mean, RMSE = rmse,
    `Predictive R2` = predictive_r2, `Severe N` = severe_n,
    `Severe observed` = severe_observed,
    `Severe predicted` = severe_predicted
  ) |>
  arrange(Validation, RMSE) |>
  kable()
```

## Spatial signal

The persistent field has a posterior median range of about 223 km, but its 95%
interval is wide (39--1106 km). Its latent standard deviation is about 0.65
(0.28--1.42). This is meaningful residual spatial structure, but the range is
not yet sharply identified by 139 irregularly sampled reefs.

```{r}
fields |>
  filter(candidate == 'persistent_spatial') |>
  ggplot(aes(x_km, y_km, colour = mean)) +
  geom_point(size = 2.1) +
  coord_equal() +
  scale_colour_gradient2(
    low = '#2166AC', mid = 'white', high = '#B2182B', midpoint = 0
  ) +
  labs(
    x = 'Easting (km)', y = 'Northing (km)', colour = 'Latent effect',
    title = 'Persistent residual mortality field',
    subtitle = 'Red: higher mortality than measured covariates explain; blue: buffering'
  )
```

The field is a hypothesis generator, not a causal map. Persistent negative
areas may point to cooling, upwelling, exposure or community features not yet
represented. Positive areas may identify persistent freshwater, cyclone,
COTS, composition or survey-process gaps. Field uncertainty and distance from
observed reefs must accompany any GBR-wide projection.

## Event-specific geography

```{r}
fields |>
  filter(candidate == 'event_spatial_ar1') |>
  ggplot(aes(x_km, y_km, colour = mean)) +
  geom_point(size = 1.25) +
  coord_equal() +
  facet_wrap(~event_year, ncol = 3) +
  scale_colour_gradient2(
    low = '#2166AC', mid = 'white', high = '#B2182B', midpoint = 0
  ) +
  labs(
    x = 'Easting (km)', y = 'Northing (km)', colour = 'Latent effect',
    title = 'Event-specific residual fields',
    subtitle = 'Useful for diagnosing event geography after some observations are available'
  )
```

The AR1 event field has median spatial range near 138 km and latent standard
deviation near 0.74. Its adjacent-event correlation is -0.49, with a 95%
interval from -0.80 to approximately zero. That is evidence against treating
the five observed event surfaces as a smoothly persistent temporal process.
Independent event fields are simpler and slightly better under reef-blocked
validation. They are suitable for retrospective inference or updating a map
as event observations arrive, but not as an unobserved pre-survey correction.

## Repeat exposure and a changing DHW relationship

```{r}
slopes |>
  filter(candidate == 'persistent_rw1_full_partial_pool') |>
  mutate(
    lower = .data[['0.025quant']],
    upper = .data[['0.975quant']]
  ) |>
  ggplot(aes(event_year, mean)) +
  geom_hline(yintercept = 0, colour = 'grey50') +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.18,
              fill = '#2C7FB8') +
  geom_line(colour = '#2C7FB8') +
  geom_point(size = 2.5, colour = '#2C7FB8') +
  scale_x_continuous(breaks = unique(slopes$event_year)) +
  labs(
    x = 'Bleaching event', y = 'Event deviation in DHW slope',
    title = 'No event-specific DHW slope is clearly different from zero',
    subtitle = 'Full feature model; deviations are shared across occurrence, magnitude and programmes'
  )
```

The posterior means decline across the sequence, but every 95% interval crosses
zero. The direct DHW-by-number-of-events interaction is also centred close to
zero in all candidates. This screen therefore does not establish adaptation.
With only four reefs observed in all five events and 46 reefs observed in only
one, the distinction between within-reef change and changing spatial/community
composition is weak.

The adaptation question is better tested in the next model by separating:

1. within-reef repeat exposure and years since damaging exposure;
2. between-reef historical exposure;
3. event-specific environmental severity and composition;
4. persistent reef susceptibility; and
5. observation-programme calibration.

A continuous-time temporal model is preferable if enough events become
available. In this screen, AR1 groups are adjacent event positions, not equal
calendar intervals.

## Recommended formal INLA model

1. Keep the Bernoulli occurrence and conditional beta magnitude layers, with
   exact boundary handling and programme-specific intercepts and precision.
2. Restore current speed and its DHW interaction to the ecological core. Carry
   summer SST skewness, excess kurtosis and wet-season Chl-a as a jointly
   regularised feature group.
3. Use the persistent SPDE field in the operational model. Compare it with an
   independent event-field update when partial event data are available.
4. Retain repeat exposure, recovery interval and their DHW interactions as
   explicit hypotheses. Keep the RW1 event-varying DHW slope as a sensitivity,
   not an adaptation result.
5. Add programme/component calibration loadings and partially pooled feature
   slopes around the shared ecological response instead of forcing unit
   loadings.
6. Add the severe-tail expert and nonlinear BRT to the same nested validation
   workflow. Spatial INLA improves broad prediction but leaves the main extreme
   deficit intact.
7. Rank models primarily by reef-blocked, forward-event and leave-event-out
   prediction. Report WAIC, DIC, marginal likelihood and CPO as secondary
   within-likelihood diagnostics.
8. Generate GBR-wide posterior samples so occurrence, positive magnitude,
   spatial uncertainty and ensemble uncertainty are propagated jointly.

The operational deliverable should expose two maps: a pre-survey forecast using
measured covariates plus the persistent field, and an optional event-updated map
that learns an event-specific field from the first available mortality surveys.
