---
title: 'Shared latent mortality and annual coral-cover dynamics'
subtitle: 'Operational validation of programme calibration and net annual change'
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)

latent_dir <- '../../output/latent_mortality'
change_dir <- '../../output/annual_coral_change'

latent_metrics <- read_csv(
  file.path(latent_dir, 'model_comparison_metrics.csv'),
  show_col_types = FALSE
)
latent_predictions <- read_csv(
  file.path(latent_dir, 'model_comparison_predictions.csv'),
  show_col_types = FALSE
)
latent_diagnostics <- read_csv(
  file.path(latent_dir, 'latent_diagnostics.csv'),
  show_col_types = FALSE
)
latent_fit <- read_csv(
  file.path(latent_dir, 'latent_fit_summary.csv'),
  show_col_types = FALSE
)
latent_effects <- read_csv(
  file.path(latent_dir, 'latent_production_effects.csv'),
  show_col_types = FALSE
)
overlap <- read_csv(
  file.path(latent_dir, 'programme_overlap.csv'),
  show_col_types = FALSE
)
agreement <- read_csv(
  file.path(latent_dir, 'ltmp_manta_agreement.csv'),
  show_col_types = FALSE
)

transitions <- readRDS('../data/processed/annual_coral_transitions.rds')
transition_manifest <- read_csv(
  '../data/processed/annual_coral_transitions_manifest.csv',
  show_col_types = FALSE
)
change_metrics <- read_csv(
  file.path(change_dir, 'annual_change_metrics.csv'),
  show_col_types = FALSE
)
change_predictions <- read_csv(
  file.path(change_dir, 'annual_change_predictions.csv'),
  show_col_types = FALSE
)
importance <- read_csv(
  file.path(change_dir, 'annual_change_importance.csv'),
  show_col_types = FALSE
)
change_outliers <- read_csv(
  file.path(change_dir, 'annual_change_outliers.csv'),
  show_col_types = FALSE
)
cover_bins <- read_csv(
  file.path(change_dir, 'change_by_initial_cover.csv'),
  show_col_types = FALSE
)

scheme_labels <- c(
  leave_one_event_out = 'Unseen bleaching event',
  reef_blocked_2024 = 'Unseen reefs in 2024',
  recent_year_holdout = 'Unseen recent year',
  reef_blocked_5fold = 'Unseen reefs, all years',
  forward_2024_2025 = 'Forward 2024-25'
)
```

# Conclusions

The shared latent ecological model is now implemented, but it should **not
replace the current prediction ensemble**. It improves MMP transfer, where
programme calibration is most needed, but worsens LTMP and manta prediction
and trails the weather--RRN BRT overall. It is therefore retained as an MMP
calibration and ensemble candidate.

The annual coral-cover target is worth pursuing. A reproducible dataset now
contains 3,089 approximately annual transitions, including both gains and
losses. Predicting change avoids the denominator instability of relative
mortality at low cover and provides post-season coral cover directly. The
initial models are useful benchmarks, not yet a production state model:

- a growth/environment linear model slightly improves temporal and forward
  cover RMSE over persistence;
- a change BRT improves transfer to completely unseen reefs;
- the change BRT captures much of the GBR-wide mean 2024 decline but not its
  reef-level extremes.

The next dynamic model should separate expected growth from disturbance loss
instead of fitting one unrestricted change surface.

# Shared latent mortality model

## What was implemented

One latent ecological logit-mortality state is estimated per reef-event. DHW,
thermal history, cover, Acropora, optical environment, rainfall, winds, WQC and
cyclone waves enter that state once. Manta tow anchors the ecological scale.
LTMP and MMP have separate observation bias, beta precision, and zero/one
probabilities.

Programme-specific ecological response loadings were tested and rejected:
MMP has only one reef-event paired with another benthic programme, so its
loading was not identifiable. The final candidate uses a common ecological
scale and regularised programme biases.

```{r overlap-table}
overlap |>
  mutate(
    Pattern = paste0(
      if_else(ltmp == 1, 'LTMP ', ''),
      if_else(manta == 1, 'manta ', ''),
      if_else(mmp == 1, 'MMP', '')
    )
  ) |>
  select(Pattern, `Reef-events` = reef_events) |>
  kable(caption = 'Programme overlap available to identify observation layers')
```

There are 153 paired LTMP--manta reef-events. Their mortality correlation is
0.619, with a median absolute difference of 1.5 percentage points. MMP is
mostly unpaired, which limits direct identification of its observation bias.

## Predictive comparison

```{r latent-overall-plot, fig.width=10, fig.height=5.5}
latent_metrics |>
  filter(programme_key == 'all') |>
  mutate(Scheme = scheme_labels[scheme]) |>
  ggplot(aes(learner, rmse, fill = learner)) +
  geom_col(width = 0.72) +
  facet_wrap(~ Scheme, scales = 'free_x') +
  scale_fill_brewer(palette = 'Set2', guide = 'none') +
  labs(
    x = NULL, y = 'Mortality RMSE',
    title = 'The latent model does not beat the BRT overall'
  ) +
  theme(axis.text.x = element_text(angle = 22, hjust = 1))
```

```{r latent-programme-table}
latent_metrics |>
  filter(programme_key != 'all') |>
  mutate(
    Scheme = scheme_labels[scheme],
    Programme = toupper(programme_key),
    RMSE = round(rmse, 3),
    MAE = round(mae, 3),
    `Severe MAE` = round(severe_mae, 3)
  ) |>
  select(Scheme, Programme, Model = learner, RMSE, MAE, `Severe MAE`) |>
  kable(caption = 'Blocked mortality prediction by programme')
```

For unseen events, latent MMP RMSE falls from 0.324 in the previous pooled
BRMS to 0.271. LTMP worsens from 0.150 to 0.165 and manta from 0.158 to
0.163. For unseen 2024 reefs, MMP improves only slightly (0.251 to 0.247),
while LTMP and manta worsen. The weather--RRN BRT remains best overall at
0.180 unseen-event RMSE and 0.174 unseen-reef RMSE.

The four-chain production fit passes its diagnostic gate (maximum R-hat 1.01,
no divergences), but the two-chain screening folds reach R-hat 1.027. The
production Bayesian R2 is 0.123. Nineteen PSIS-LOO observations have Pareto
`k > 0.7`, so blocked validation is the primary decision evidence.

The production MMP observation bias is negative on the manta-anchored logit
scale (-0.677, 95% interval -1.26 to -0.11). LTMP bias is small and uncertain
(-0.171, interval -0.38 to 0.06). This explains why calibration helps MMP but
adds little for the already well-paired LTMP--manta data.

## Operational use

The production object includes a programme-free prediction path. For any GBR
reef-event with the required covariates it returns:

- the population ecological mortality mean, excluding fitted reef effects;
- a process-marginalised mean for a new reef-event;
- 5%, 50% and 95% process intervals generated from a new reef-event effect.

This prevents fitted historical random effects from leaking into GBR-wide
maps. Because overall blocked performance is weaker, these predictions should
enter ensemble testing rather than being issued alone.

# Annual coral-cover change

## Transition estimand

The annual dataset uses consecutive surveys 240--550 days apart. The response
is both next coral cover and absolute percentage-point change:

\[
\Delta C = 100(C_{t+1} - C_t).
\]

Unlike the bleaching-mortality table, gains are retained and no observations
are removed because another disturbance is recorded. Current cover, previous
change, interval duration, Acropora composition, DHW, WQC, cyclone waves and
COTS are all available as candidate predictors. Acropora fallbacks use only
the baseline year or the preceding two years.

```{r transition-manifest}
transition_manifest |>
  mutate(
    `Positive changes` = positive_changes,
    `Negative changes` = negative_changes,
    `Median interval (days)` = median_interval_days
  ) |>
  select(
    Programme = programme_key, transitions, reefs,
    `Positive changes`, `Negative changes`, `Median interval (days)`
  ) |>
  kable(caption = 'Annual coral-cover transition dataset')
```

```{r change-distribution, fig.width=10, fig.height=5.5}
transitions |>
  mutate(
    display_change = pmin(pmax(cover_change_pp, -35), 25),
    Programme = toupper(programme_key)
  ) |>
  ggplot(aes(display_change, fill = Programme)) +
  geom_histogram(binwidth = 2, position = 'identity', alpha = 0.48) +
  geom_vline(xintercept = 0, colour = 'grey35') +
  facet_wrap(~ Programme, scales = 'free_y') +
  labs(
    x = 'Annual coral-cover change (percentage points; tails clipped for display)',
    y = 'Transitions',
    title = 'Annual change retains both recovery and loss'
  )
```

Of the 3,089 transitions, 54.2% increase and 45.7% decline. The median change
is +0.44 percentage points, with a 5th--95th percentile range of approximately
-12.9 to +9.8 points.

## Why absolute change is complementary

```{r low-cover-table}
cover_bins |>
  mutate(
    across(starts_with('relative_change_'), ~ round(.x, 2)),
    median_absolute_change_pp = round(median_absolute_change_pp, 2)
  ) |>
  kable(caption = 'Change distribution by starting coral cover')
```

Below 5% initial cover, the 5th--95th percentile relative-change range is
approximately -48% to +143%. Absolute change remains on the interpretable
percentage-point scale. Absolute change should therefore complement, not
replace, event-attributable relative mortality: the two answer different
ecological questions.

## Initial predictive benchmarks

```{r change-metric-plot, fig.width=11, fig.height=6}
change_metrics |>
  filter(programme_key == 'all') |>
  mutate(Scheme = scheme_labels[scheme]) |>
  ggplot(aes(learner, post_cover_rmse_pp, fill = learner)) +
  geom_col(width = 0.72) +
  facet_wrap(~ Scheme, scales = 'free_x') +
  scale_fill_brewer(palette = 'Set2', guide = 'none') +
  labs(
    x = NULL, y = 'Next-cover RMSE (percentage points)',
    title = 'No single annual-change model wins every transfer problem'
  ) +
  theme(axis.text.x = element_text(angle = 22, hjust = 1))
```

```{r change-metric-table}
change_metrics |>
  filter(programme_key == 'all') |>
  mutate(
    Scheme = scheme_labels[scheme],
    across(
      c(post_cover_rmse_pp, change_mae_pp, change_bias_pp,
        severe_loss_mae_pp),
      ~ round(.x, 2)
    ),
    direction_accuracy = percent(direction_accuracy, accuracy = 0.1)
  ) |>
  select(
    Scheme, Model = learner,
    `Cover RMSE (pp)` = post_cover_rmse_pp,
    `Change MAE (pp)` = change_mae_pp,
    `Change bias (pp)` = change_bias_pp,
    `Direction accuracy` = direction_accuracy,
    `Severe-loss MAE (pp)` = severe_loss_mae_pp
  ) |>
  kable(caption = 'Annual transition validation')
```

For forward prediction of 2024--25, the growth/environment model improves
next-cover RMSE from 11.13 points under persistence to 10.62. For recent
leave-one-year-out prediction it improves 8.66 to 8.54. For reefs withheld
from every year, the change BRT is best at 6.98 points versus 7.61 for
persistence.

```{r event-mean-plot, fig.width=10, fig.height=5.5}
change_predictions |>
  filter(scheme == 'forward_2024_2025') |>
  group_by(learner, event_year) |>
  summarise(
    Observed = mean(cover_change_pp),
    Predicted = mean(predicted_change_pp),
    .groups = 'drop'
  ) |>
  tidyr::pivot_longer(
    c(Observed, Predicted), names_to = 'Series', values_to = 'Change'
  ) |>
  ggplot(aes(factor(event_year), Change, fill = Series)) +
  geom_col(position = 'dodge') +
  facet_wrap(~ learner) +
  geom_hline(yintercept = 0, colour = 'grey35') +
  labs(
    x = 'Survey event year', y = 'Mean coral-cover change (percentage points)',
    title = 'The change BRT captures the mean 2024 loss but not reef-level extremes'
  )
```

Observed mean change in 2024 is -5.89 points. The forward change BRT predicts
-4.71, compared with -1.60 for the growth model and zero for persistence.
Nevertheless, its individual 2024 RMSE is 13.4 points; the growth model is
better at 11.6. The model recognises an event-wide loss but allocates it poorly
among reefs.

## Influential predictors and residual priorities

```{r importance-plot, fig.width=10, fig.height=6}
importance |>
  group_by(learner) |>
  slice_max(influence, n = 8) |>
  ungroup() |>
  mutate(variable = reorder(variable, influence)) |>
  ggplot(aes(influence, variable, fill = learner)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ learner, scales = 'free_y') +
  labs(
    x = 'Relative influence', y = NULL,
    title = 'Current cover dominates next cover; change uses a broader process set'
  )
```

Current cover contributes 82% of post-cover BRT influence. In the change BRT,
current cover (21%), previous change (18%), DHW (17%), cyclone waves (9%) and
interval length (9%) dominate. Acropora contributes 4.4%, WQC 3.8% and COTS
3.8%. These are predictive influences, not causal partitions.

```{r outlier-table}
change_outliers |>
  filter(learner == 'Change BRT') |>
  transmute(
    Reef = ReefName, Programme = toupper(programme_key),
    Year = event_year, DHW = round(sst_maxdhw, 2),
    `Starting cover` = percent(pre_cover, accuracy = 0.1),
    `Observed change (pp)` = round(cover_change_pp, 1),
    `Predicted change (pp)` = round(predicted_change_pp, 1),
    `Error (pp)` = round(change_error_pp, 1),
    Disturbance = disturbance_text
  ) |>
  head(15) |>
  kable(caption = 'Largest forward 2024-25 change-BRT misses')
```

Snapper, North Direction, One Tree Island, Linnet, Broomfield and corrected
Mackay are among the large 2024 misses. Some reefs also show observed gains
where the model predicts event loss. This is the correct residual set for
examining fine-scale exposure, taxonomic recovery state, survey timing and
unrecorded disturbance.

# Pipeline decision

1. Keep the shared latent model as an MMP calibration/ensemble candidate, not
   the sole operational mortality model.
2. Retain both relative bleaching mortality and annual absolute cover change.
3. Use the annual transition table as the input to a formal dynamic model with
   separate growth and disturbance-loss components.
4. Constrain growth by current cover and available space; allow DHW,
   freshwater/WQC, cyclone waves, COTS and community composition to affect
   losses.
5. Preserve programme observation layers and integrate unknown reef/year
   effects for GBR-wide forecasts.
6. Require improvement over persistence in forward years and over the change
   BRT for unseen reefs before operational promotion.
