---
title: 'Local-first NOAA DHW calibration'
subtitle: 'Peak-qualified AIMS logger evidence, ENSO diagnostics and INLA validation'
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)
library(tidyr)

result_dir <- '../../output/dhw_correction_local_first'
audit <- read_csv(file.path(result_dir, 'series_peak_coverage_audit.csv'),
  show_col_types = FALSE)
site_events <- read_csv(file.path(result_dir, 'site_event_discrepancies.csv'),
  show_col_types = FALSE)
radius_metrics <- read_csv(file.path(result_dir, 'local_radius_metrics.csv'),
  show_col_types = FALSE)
enso_summary <- read_csv(file.path(result_dir, 'enso_event_discrepancy_summary.csv'),
  show_col_types = FALSE)
config <- read_csv(file.path(result_dir, 'configuration.csv'), show_col_types = FALSE)
layer <- read_csv('../data/processed/noaa_dhw_correction_layer_local_first_validation.csv',
  show_col_types = FALSE)
old_layer <- read_csv('../data/processed/noaa_dhw_correction_layer_validation.csv',
  show_col_types = FALSE)
metrics <- read_csv('../../output/inla_spatiotemporal/model_comparison.csv',
  show_col_types = FALSE)
predictions <- read_csv('../../output/inla_spatiotemporal/cv_predictions.csv',
  show_col_types = FALSE)
uncertainty_dir <- '../../output/local_calibrated_operational'
uncertainty_predictions <- read_csv(
  file.path(uncertainty_dir, 'uncertainty_predictions.csv'),
  show_col_types = FALSE)
uncertainty_calibration <- read_csv(
  file.path(uncertainty_dir, 'uncertainty_calibration.csv'),
  show_col_types = FALSE)
updated_misses <- read_csv(
  file.path(uncertainty_dir, 'largest_misses_local_calibrated.csv'),
  show_col_types = FALSE)
residual_priorities <- read_csv(
  file.path(uncertainty_dir, 'residual_feature_priorities.csv'),
  show_col_types = FALSE)

base_name <- 'persistent_rw1_cyclone_cots_partial_pool'
fixed_name <- 'persistent_rw1_cyclone_cots_lizard_adjusted_partial_pool'
local_name <- 'persistent_rw1_cyclone_cots_local_first_dhw_partial_pool'
operational_name <- 'persistent_rw1_cyclone_cots_operational_local_update_partial_pool'
decomposed_name <- 'persistent_rw1_local_dhw_decomposed_hazards_partial_pool'
decomposed_freshwater_name <-
  'persistent_rw1_local_dhw_decomposed_hazards_freshwater_partial_pool'
decomposed_operational_name <-
  'persistent_rw1_operational_update_decomposed_hazards_freshwater_partial_pool'

# The central operational candidate now uses interval-aligned disturbance
# hazards and continuous freshwater interactions.
previous_local_name <- local_name
previous_operational_name <- operational_name
local_name <- decomposed_freshwater_name
operational_name <- decomposed_operational_name
candidate_labels <- c(
  'NOAA unchanged',
  'Fixed +4.82 DHW stress-test',
  'Previous local-calibrated',
  'Interval hazards',
  'Interval hazards + freshwater',
  'Current-event local update (interval hazards + freshwater)')
names(candidate_labels) <- c(
  base_name, fixed_name, previous_local_name, decomposed_name,
  local_name, operational_name)

metric_table <- metrics |>
  filter(candidate %in% names(candidate_labels)) |>
  mutate(
    DHW_input = recode(candidate, !!!candidate_labels),
    validation = recode(scheme,
      leave_one_event_out = 'Leave-one-event-out',
      reef_blocked_5fold = 'Reef-blocked 5-fold'))

cluster_ids <- c('14-114', '14-116a', '14-118', '14-123', '14-126', '14-143')
cluster_layer <- layer |>
  filter(event_year == 2024, ReefID %in% cluster_ids) |>
  select(ReefID, ReefName, NOAA_correction_DHW = local_first_correction,
    correction_source, nearest_local_logger_km,
    effective_local_loggers, local_loggers_used) |>
  left_join(
    old_layer |>
      filter(event_year == 2024, ReefID %in% cluster_ids) |>
      select(ReefID, old_correction_DHW = within_event_correction),
    by = 'ReefID')

cluster_predictions <- predictions |>
  filter(
    event_year == 2024, ReefID %in% cluster_ids,
    scheme == 'leave_one_event_out',
    candidate %in% c(base_name, fixed_name, local_name, operational_name)) |>
  mutate(DHW_input = recode(candidate, !!!candidate_labels)) |>
  group_by(DHW_input, ReefID, ReefName) |>
  summarise(
    observed = mean(observed_mortality),
    predicted = mean(predicted_mortality),
    NOAA_DHW = mean(ann_maxdhw_original),
    uplift_DHW = mean(applied_dhw_uplift),
    model_DHW = mean(ann_maxdhw),
    .groups = 'drop')

coverage_case <- audit |>
  filter(
    event_year == 2024, site == 'Lizard Island',
    series %in% c(
      'LIZFL1',
      'Lizard Island Relay Pole 2 Water Temperature @5m')) |>
  mutate(
    true_peak_discrepancy = unconstrained_peak_dhw - noaa_official_max_dhw,
    retained = if_else(peak_qualified, 'Yes', 'No'))

event_coverage <- layer |>
  group_by(event_year) |>
  summarise(
    reefs = n(),
    direct = sum(correction_source == 'current_direct'),
    local_interpolation = sum(correction_source == 'current_local'),
    unchanged_NOAA = sum(correction_source == 'none'),
    corrected = direct + local_interpolation,
    corrected_fraction = corrected / reefs,
    .groups = 'drop')

event_rmse <- predictions |>
  filter(
    scheme == 'leave_one_event_out',
    candidate %in% c(base_name, local_name)) |>
  group_by(candidate, event_year) |>
  summarise(
    RMSE = sqrt(mean((observed_mortality - predicted_mortality)^2)),
    .groups = 'drop') |>
  mutate(model = recode(candidate, !!!candidate_labels)) |>
  select(-candidate) |>
  pivot_wider(names_from = model, values_from = RMSE) |>
  mutate(change = .data[['Interval hazards + freshwater']] -
    .data[['NOAA unchanged']])

hazard_candidates <- c(
  previous_local_name, decomposed_name, local_name, operational_name
)
hazard_metrics <- metrics |>
  filter(
    candidate %in% hazard_candidates,
    scheme == 'leave_one_event_out') |>
  transmute(
    Candidate = recode(candidate, !!!candidate_labels),
    rmse, mae, predictive_r2, severe_rmse, false_extreme_rate) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

named_miss_checks <- updated_misses |>
  filter(
    ReefID %in% c('14-137', '14-138', '15-099c', '21-025', '21-556', '16-015'),
    event_year %in% c(2016, 2017, 2020, 2024)) |>
  select(
    programme_key, ReefName, event_year, observed_mortality,
    predicted_mortality, residual, cyc_interval_maxHrs4mw,
    cyc_interval_peak_year, tc_interval_peak_name,
    cot_interval_idw_max, cots_outbreak_probability,
    primary_explanation) |>
  arrange(desc(residual))
```

## Decision

The earlier sub-1 DHW corrections were an artefact of the correction procedure,
not a disappearance of the 2024 logger signal. The original shallow reef-flat
comparisons remain large: Lizard Island is +4.26 DHW, Martin +3.84 DHW and
North Direction +5.38 DHW relative to NOAA.

The revised layer implements a strict hierarchy:

1. use a coverage-qualified reef logger at the site-event;
2. use an automated-station temperature only when no reef logger exists;
3. match a survey reef directly only when its normalised name agrees and it is
   within 5 km of the logger site;
4. otherwise interpolate from at most four current-event sites inside 25 km;
5. otherwise retain NOAA unchanged;
6. permit an ENSO-matched historical fallback only after at least two previous
   same-phase events at a site with at least 75% sign agreement.

No current event satisfies the historical fallback rule. Consequently, this
version applies **no GBR-wide event offset and no historical correction
surface**. Large-scale products remain the base exposure where local
measurements are absent; they do not override local observations.

## Why the previous corrections became small

Three effects attenuated the earlier 2024 Lizard correction.

- The automated 5 m relay-pole series was summarised at an earlier date with
  enough observations, giving an apparent -2.37 DHW discrepancy. Its actual
  event peak was later and implied +2.18 DHW, but only 66 of 84 days were
  observed. It is now excluded rather than moved to an earlier peak.
- The automated value was pooled with the +4.26 DHW reef-flat logger even
  though a direct reef logger existed, reducing the site estimate to about
  +0.94 DHW. The reef logger now takes precedence.
- A broad 200 km current-event surface was then combined with a negative GBR
  event offset and a negative historical surface. Those terms are no longer
  defensible and have been removed.

```{r}
coverage_case |>
  transmute(
    Source = recode(source,
      temperature_logger = 'Reef logger',
      automated_weather = 'Automated station'),
    Series = series,
    `Depth (m)` = depth_m,
    `Earlier eligible peak DHW` = logger_max_dhw,
    `Actual event peak DHW` = unconstrained_peak_dhw,
    `Days at actual peak` = true_peak_days,
    `Actual logger minus NOAA DHW` = true_peak_discrepancy,
    `Retained in correction` = retained) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  kable(caption = 'The 2024 Lizard coverage check that changed the result')
```

## Spatial scale and validation rule

The 25 km radius is selected because it has the best held-out MAE and
severe-discrepancy RMSE among the local candidates. A 50 km radius has a
slightly lower overall RMSE, but spreads information much farther and performs
worse for absolute and extreme errors. Ten kilometres leaves almost two thirds
of held-out sites unchanged.

```{r}
radius_metrics |>
  transmute(
    `Radius (km)` = local_radius_km,
    RMSE = rmse, MAE = mae, bias,
    `Extreme RMSE` = extreme_rmse,
    `Local support` = local_support_fraction,
    `Unchanged NOAA` = no_correction_fraction) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable(caption = 'Leave-one-site-out validation of current-event interpolation')
```

```{r, fig.width=8.5, fig.height=4.8}
radius_metrics |>
  select(local_radius_km, RMSE = rmse, `Extreme RMSE` = extreme_rmse) |>
  pivot_longer(-local_radius_km, names_to = 'metric', values_to = 'error') |>
  ggplot(aes(local_radius_km, error, colour = metric)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.8) +
  geom_vline(xintercept = config$selected_local_radius_km,
    linetype = 2, colour = 'grey35') +
  scale_x_continuous(breaks = radius_metrics$local_radius_km) +
  labs(
    x = 'Hard interpolation radius (km)', y = 'Held-out error (DHW)',
    colour = NULL,
    title = 'Twenty-five kilometres retains local signal and protects extremes',
    subtitle = 'Dashed line marks the selected radius') +
  theme_bw(base_size = 11) +
  theme(legend.position = 'bottom')
```

## ENSO is context, not a global offset

The event medians do not support a simple GBR-wide ENSO correction. The strong
2016 El Nino median is slightly positive, but the weak 2024 El Nino median is
negative across the GBR logger network. Neutral 2017 and 2020 are positive,
whereas 2022 La Nina is near zero. The 2016/2024 positive discrepancies near
Lizard can therefore be treated as a hypothesis about a local ENSO-linked
mechanism, but not yet as a transferable GBR adjustment.

```{r, fig.width=8.5, fig.height=5.2}
ggplot(
  enso_summary,
  aes(event_year, median_discrepancy, colour = enso_phase)
) +
geom_hline(yintercept = 0, colour = 'grey55') +
geom_linerange(aes(ymin = q10, ymax = q90), linewidth = 1) +
geom_point(size = 3) +
scale_x_continuous(breaks = enso_summary$event_year) +
labs(
  x = 'Bleaching event', y = 'Logger minus NOAA DHW', colour = 'ENSO phase',
  title = 'GBR-wide logger discrepancy is not explained by a single ENSO offset',
  subtitle = 'Points are event medians; bars are the 10th to 90th percentiles') +
theme_bw(base_size = 11) +
theme(legend.position = 'bottom')
```

```{r}
enso_summary |>
  transmute(
    Event = event_year, `ENSO category` = enso_category,
    Sites = sites, `Median discrepancy` = median_discrepancy,
    `Mean discrepancy` = mean_discrepancy,
    `Positive fraction` = positive_fraction) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable(caption = 'Event-level ENSO context and logger discrepancy')
```

## The revised 2024 Lizard-region layer

The six target reefs receive +0.33 to +5.38 DHW, compared with +0.55 to +0.96
DHW under the previous broad layer. MacGillivray, Lizard NW, Martin and North
Direction use their matched direct local measurement. Eyrie and Linnet use the
four nearest current-event sources inside 25 km. These are local calibrations,
not replacements for measured reef temperatures and not corrections exported
to unsupported parts of the GBR.

```{r}
cluster_layer |>
  transmute(
    ReefID, ReefName,
    `Previous broad correction` = old_correction_DHW,
    `Local-first correction` = NOAA_correction_DHW,
    Source = correction_source,
    `Nearest logger (km)` = nearest_local_logger_km,
    `Effective logger count` = effective_local_loggers,
    `Logger sites used` = local_loggers_used) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  arrange(ReefID) |>
  kable(caption = '2024 local-first corrections at the six Lizard-region reefs')
```

```{r, fig.width=8.5, fig.height=7}
layer |>
  filter(event_year == 2024) |>
  ggplot(aes(lon, lat, colour = local_first_correction)) +
  geom_point(size = 1.7, alpha = 0.85) +
  geom_point(
    data = filter(layer, event_year == 2024, ReefID %in% cluster_ids),
    shape = 21, fill = NA, colour = 'black', size = 3, stroke = 0.8) +
  scale_colour_gradient2(
    low = '#2c7bb6', mid = 'grey92', high = '#d7191c', midpoint = 0,
    name = 'DHW\ncorrection') +
  coord_equal() +
  labs(
    x = 'Longitude', y = 'Latitude',
    title = 'The local-first layer leaves unsupported 2024 reefs unchanged',
    subtitle = 'Outlined points are the six Lizard-region target reefs') +
  theme_bw(base_size = 11) +
  theme(legend.position = 'bottom')
```

## Mortality-model consequences

There are two useful deployment comparisons. The fully local-calibrated model
uses available event-specific logger evidence for both historical fitting rows
and the held-out/current event. It is the preferred local-evidence model because
the same measurement hierarchy is applied consistently. In leave-one-event-out
validation it reduces overall RMSE from
`r number(metric_table$rmse[metric_table$candidate == base_name & metric_table$scheme == 'leave_one_event_out'], accuracy = 0.001)`
to
`r number(metric_table$rmse[metric_table$candidate == local_name & metric_table$scheme == 'leave_one_event_out'], accuracy = 0.001)`,
changes predictive R2 from
`r number(metric_table$predictive_r2[metric_table$candidate == base_name & metric_table$scheme == 'leave_one_event_out'], accuracy = 0.001)`
to
`r number(metric_table$predictive_r2[metric_table$candidate == local_name & metric_table$scheme == 'leave_one_event_out'], accuracy = 0.001)`,
and severe-case RMSE from
`r number(metric_table$severe_rmse[metric_table$candidate == base_name & metric_table$scheme == 'leave_one_event_out'], accuracy = 0.001)`
to
`r number(metric_table$severe_rmse[metric_table$candidate == local_name & metric_table$scheme == 'leave_one_event_out'], accuracy = 0.001)`.
The false-extreme rate decreases from
`r percent(metric_table$false_extreme_rate[metric_table$candidate == base_name & metric_table$scheme == 'leave_one_event_out'], accuracy = 0.1)`
to
`r percent(metric_table$false_extreme_rate[metric_table$candidate == local_name & metric_table$scheme == 'leave_one_event_out'], accuracy = 0.1)`.

The current-event-only update leaves legacy training rows on NOAA and updates
only the held-out event. It gives a lower severe-case RMSE of
`r number(metric_table$severe_rmse[metric_table$candidate == operational_name & metric_table$scheme == 'leave_one_event_out'], accuracy = 0.001)`,
but worse overall RMSE and a higher false-extreme rate. It is retained as a
severe-tail sensitivity or ensemble input, not the primary calibration.
Reef-blocked results for this update-only candidate are not primary because the
split can mix corrected assessment rows with uncorrected training rows from the
same event.

```{r}
metric_table |>
  select(
    DHW_input, validation, RMSE = rmse, MAE = mae,
    predictive_R2 = predictive_r2, bias,
    occurrence_Brier = occurrence_brier,
    severe_RMSE = severe_rmse,
    false_extreme_rate) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  arrange(validation, RMSE) |>
  kable(caption = 'INLA validation of NOAA and alternative local corrections')
```

```{r, fig.width=8.5, fig.height=5.8}
ggplot(cluster_predictions, aes(observed, predicted, colour = DHW_input)) +
  geom_abline(slope = 1, intercept = 0, colour = 'grey50') +
  geom_point(size = 2.8) +
  geom_text(
    aes(label = ReefID), nudge_y = 0.018, size = 3,
    check_overlap = TRUE, show.legend = FALSE) +
  coord_equal(xlim = c(0, 0.65), ylim = c(0, 0.65)) +
  labs(
    x = 'Observed relative mortality', y = 'Predicted relative mortality',
    colour = 'DHW input',
    title = 'Local event calibration raises the 2024 Lizard predictions',
    subtitle = 'Predictions are from the fold withholding the complete 2024 event') +
  theme_bw(base_size = 11) +
  theme(legend.position = 'bottom')
```

The calibrated alternatives materially raise predictions at Eyrie, Martin,
Linnet and North Direction, while residual misses remain at Lizard NW and some
MacGillivray observations. This supports the ecological interpretation
already developed elsewhere in the pipeline: local thermal underestimation is
one part of the 2024 signal, but freshwater/coloured water, Acropora or weedy
community composition, cyclone exposure and COTS still need to act as compound
modifiers.

## Production rule

Use `ann_maxdhw_original` as the immutable NOAA field and store the local
calibration separately with its source, distance, logger count and uncertainty.
For historical fitting and a live or recently completed event, rebuild the
same calibration after enough logger observations span the actual DHW peak.
Do not extrapolate beyond 25 km.
If no supported local evidence exists, use NOAA unchanged. Revisit the
ENSO-matched fallback only when repeated same-phase events demonstrate stable
site-specific sign and magnitude; the current record is insufficient.

## Formal multi-event status

The fully local-calibrated candidate is not a 2024-only adjustment. It supplies
direct or locally interpolated corrections in every bleaching event used by the
model. Coverage ranges from 70.5% of reefs in 2022 to 97.4% in 2017. The
remaining reefs retain NOAA as their point exposure and receive wider
uncertainty rather than an invented correction.

```{r}
event_coverage |>
  transmute(
    Event = event_year, Reefs = reefs, Direct = direct,
    `Local interpolation` = local_interpolation,
    `NOAA unchanged` = unchanged_NOAA,
    `Corrected reefs` = corrected,
    `Corrected fraction` = percent(corrected_fraction, accuracy = 0.1)) |>
  kable(caption = 'Local-calibration coverage in every fitted bleaching event')
```

The pooled improvement is real but not uniform through time. Local calibration
improves event-transfer RMSE in 2020, 2022 and especially 2024, while it worsens
2016 and 2017. The correction itself should not be weakened to accommodate
non-thermal losses. Instead, the formal mortality model needs to stop asking
the thermal response to explain storm, COTS and other disturbance mortality.

```{r}
event_rmse |>
  rename(
    Event = event_year, `NOAA RMSE` = `NOAA unchanged`,
    `Local-calibrated RMSE` = `Interval hazards + freshwater`,
    `Change (local minus NOAA)` = change) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable(caption = 'Leave-one-event-out performance by bleaching event')
```

## Interval-aligned disturbance hazards and continuous freshwater response

The central model now keeps the same shared thermal response but gives
non-thermal mechanisms their own additive pressure block. Cyclone exposure is
the maximum RRN damaging-wave hours and track/intensity signal overlapping the
actual baseline-to-post-survey interval; COTS is represented by the
interval-maximum RRN surface and the GBR-wide outbreak hindcast. This is a
predictive decomposition, not a claimed causal allocation for an individual
loss: disturbance notes remain diagnostic labels and are never supplied to the
GBR-wide prediction model.

That timing change matters. Carter and Yonge were formerly assigned zero
cyclone hours in 2016 because the join used only the 2015--16 summer. Their
cover intervals started in late 2014 and include Cyclone Nathan; the corrected
RRN exposures are 68.72 and 64.44 damaging-wave hours respectively. Agincourt
does not share that exposure because its baseline followed Nathan.

Freshwater is deliberately continuous. The candidate includes current WQC
frequency, the reef-relative preceding-decade percentile, the unthresholded
10-year WQC sum, and coastal rainfall, each as a DHW modifier. The previous
single freshwater composite is retained only in the comparison candidate. This
avoids treating WQC = 0.50 as a universal ecological threshold.

```{r}
hazard_metrics |>
  rename(
    `Predictive R2` = predictive_r2,
    `Severe RMSE` = severe_rmse,
    `False-extreme rate` = false_extreme_rate) |>
  kable(caption = 'Leave-one-event-out comparison of disturbance and freshwater candidates')
```

The interval-hazard plus freshwater model is the new central candidate. It
reduces leave-one-event-out RMSE from 0.172 to 0.158 and raises predictive R2
from 0.060 to 0.209 relative to the previous fully local-calibrated model.
Severe-case RMSE improves from 0.356 to 0.346, while the false-extreme rate
falls from 14.3% to 6.0%. The 2024 severe-tail RMSE is essentially unchanged
(0.316 versus 0.310), so this is not an apparent average-error gain purchased
by suppressing the severe 2024 event.

Gannett Cay remains a deliberately visible failure: its 2020 COTS hindcast is
high, but a COTS effect learned from other events is not strong enough to
predict this held-out case. That preserves it as a high-priority COTS
validation case rather than using its labelled explanation to tune the model.

```{r}
named_miss_checks |>
  head(18) |>
  transmute(
    Programme = toupper(programme_key), Reef = ReefName, Year = event_year,
    Observed = percent(observed_mortality, accuracy = 0.1),
    Predicted = percent(predicted_mortality, accuracy = 0.1),
    Residual = percent(residual, accuracy = 0.1),
    `Interval wave hours` = round(cyc_interval_maxHrs4mw, 2),
    `Peak cyclone` = tc_interval_peak_name,
    `COTS interval surface` = round(cot_interval_idw_max, 3),
    `COTS hindcast` = round(cots_outbreak_probability, 3),
    Explanation = coalesce(primary_explanation, 'Unresolved')) |>
  kable(caption = 'Named disturbance and COTS diagnostic cases')
```

## Operational uncertainty propagation

The operational output now carries two uncertainty sources. Logger-correction
error is estimated without mortality outcomes: direct matches use replicate
logger dispersion with a pooled 0.61 DHW floor, interpolation combines source
spread with a 1.59 DHW held-out floor, and unsupported reefs carry a zero-mean
3.47 DHW uncertainty derived from omitted-site errors. A first-order response
slope from the current-event-only INLA contrast translates that DHW uncertainty
onto mortality. It is then combined with the largest programme-specific 90%
conformal residual radius among the other bleaching events.

This gives 94.6% empirical coverage for a nominal 90% interval across the 537
held-out observations. The mean total interval width is 0.495 mortality units,
whereas the mean DHW-only contribution is 0.115. Remaining ecological and
observation-process uncertainty is therefore much larger than uncertainty in
the local DHW correction. Manta 2024 remains less well calibrated than the
pooled result, so the interval procedure is conservative overall but not yet
exchangeable among programmes and events.

```{r}
uncertainty_predictions |>
  group_by(correction_source) |>
  summarise(
    Observations = n(),
    `Median correction SD (DHW)` = median(correction_sd),
    `Empirical 90% coverage` = mean(covered90),
    `Mean total width` = mean(interval_width90),
    `Mean DHW-only width` = mean(dhw_interval_width90),
    .groups = 'drop') |>
  mutate(
    `Empirical 90% coverage` = percent(
      `Empirical 90% coverage`, accuracy = 0.1
    ),
    across(where(is.numeric), ~ round(.x, 3))) |>
  kable(caption = 'Uncertainty calibration by local-correction source')
```

This is now suitable for operational uncertainty bands. The formal Bayesian
endpoint remains repeated correction-layer imputation with INLA predictions
pooled across imputations, because the DHW predictor also enters hinges,
novelty and ecological interactions. A single classical measurement-error
coefficient would not propagate uncertainty through all those derived terms.

## Updated largest misses

Predictions are averaged within reef, event and programme before ranking, so
repeated observations such as Snapper do not dominate the table. The prediction
gain column shows directly whether local temperature calibration resolved the
miss. `Outside 90%` identifies observations that remain surprising even after
both correction and held-out process uncertainty are included.

```{r}
updated_misses |>
  arrange(desc(outside_upper90), desc(residual)) |>
  head(20) |>
  transmute(
    Programme = toupper(programme_key), Reef = ReefName, Year = event_year,
    Observations = observations,
    NOAA_DHW = round(noaa_dhw, 2),
    Corrected_DHW = round(corrected_dhw, 2),
    `Correction source` = correction_source,
    Observed = percent(observed_mortality, accuracy = 0.1),
    Predicted = percent(predicted_mortality, accuracy = 0.1),
    `Gain from local DHW` = percent(
      prediction_gain_from_local_dhw, accuracy = 0.1
    ),
    `Observed minus predicted` = percent(residual, accuracy = 0.1),
    `Upper 90% limit` = percent(prediction_q95, accuracy = 0.1),
    `Outside 90%` = outside_upper90,
    `Registered explanation` = coalesce(primary_explanation, 'Unresolved')) |>
  kable(caption = 'Largest remaining misses after full local calibration')
```

The corrected model now brings Eyrie and Linnet inside their uncertainty bands,
and substantially raises their point predictions. Lizard NW improves but
remains outside the 90% interval. Mackay remains a major miss despite a direct
+2.99 DHW correction, strengthening the freshwater/Jasper interpretation.
Penrith remains the largest miss because its direct logger correction lowers
DHW while the observed loss was associated with Cyclone Debbie. Carter and
Yonge show the same diagnostic tension in 2016—high mortality despite local
evidence that NOAA overestimated thermal exposure—and need targeted ecological
review before adding another thermal term.

```{r, fig.width=8.5, fig.height=5.5}
residual_priorities |>
  filter(is.finite(spearman_with_residual)) |>
  head(10) |>
  mutate(feature = reorder(feature, spearman_with_residual)) |>
  ggplot(aes(spearman_with_residual, feature)) +
  geom_vline(xintercept = 0, colour = 'grey55') +
  geom_col(aes(fill = spearman_with_residual > 0), width = 0.7) +
  scale_fill_manual(values = c(`TRUE` = '#b2182b', `FALSE` = '#2166ac'),
    guide = 'none') +
  labs(
    x = 'Spearman correlation with observed minus predicted mortality',
    y = NULL,
    title = 'Residual priorities after local DHW calibration',
    subtitle = 'Diagnostic associations among reef-events with at least 30% mortality') +
  theme_bw(base_size = 11)
```

## Priorities for the next model stage

1. **Separate thermal and other-disturbance mortality contributions.** Retain
   the observations, but model additive or competing cyclone, COTS and thermal
   hazards so storm-driven rows cannot flatten the DHW response. Penrith and
   Gannett Cay provide strong labelled test cases; Carter, Yonge and Agincourt
   require review.
2. **Strengthen the DHW-by-freshwater/coloured-water response.** Mackay remains
   the largest 2024 miss after direct thermal correction, and upper-quartile
   coastal rainfall is associated with about 12 percentage points more
   underprediction among high-mortality reef-events. Keep WQC cumulative
   exposure, but avoid imposing an unsupported universal threshold.
3. **Replace the coarse cyclone representation.** Refit unchanged models when
   the Jasper/Kirrily-resolving track and intensity layer arrives. Proximity,
   intensity, wave exposure and rainfall/runoff should remain separate inputs.
4. **Retain the current-event-only candidate as a severe-tail ensemble member.**
   It has better severe-case RMSE than the fully calibrated central model but
   more false extremes, making it useful for tail risk rather than the mean.
5. **Audit 2016--2017 northern misses before expanding thermal complexity.**
   Determine whether Carter, Yonge, Agincourt and St Crispin reflect storms,
   COTS, cover-estimation error or unresolved community composition. Their
   logger corrections do not support extra heat as the explanation.
6. **Refine composition only where it adds information beyond total
   Acropora.** The existing Acropora predictor is influential and is not lost;
   the next candidate should distinguish tabular Acropora and weedy recovery
   assemblages rather than adding another correlated total-cover metric.
