---
title: 'Can rainfall reconstruct missing freshwater exposure?'
subtitle: 'ERA5 rainfall, wind, IMOS optics, and event-held-out eReefs validation'
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)

input_dir <- '../../output/salinity_rainfall'
cells <- read_csv(file.path(input_dir, 'cell_year_inputs.csv'), show_col_types = FALSE)
occurrence <- read_csv(
  file.path(input_dir, 'occurrence_metrics.csv'), show_col_types = FALSE
)
magnitude <- read_csv(
  file.path(input_dir, 'magnitude_metrics.csv'), show_col_types = FALSE
)
predictions <- read_csv(
  file.path(input_dir, 'occurrence_predictions.csv'), show_col_types = FALSE
)
risk_2024 <- read_csv(
  '../data/processed/ereefs_salinity_2024_rainfall_sensitivity.csv',
  show_col_types = FALSE
)
```

# Decision rule

Thirty PSU is the primary low-salinity screening threshold. This deliberately
increases the number of positive cell-events available for training. Exposure
below 28 and 26 PSU remains nested evidence of progressively more severe
freshwater conditions; it has not been discarded.

The rainfall model must improve event-held-out average precision and balanced
Brier score over the spatial prior, and it must retain useful recall for the
below-28-PSU tail. A low all-cell salinity error is not sufficient because
ordinary oceanic cells dominate the GBR.

## ERA5 versus ACCESS source choice

[ERA5](https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels)
is a consistent hourly reanalysis spanning every event year. The official CDS
retrieval API was tested first, but this workstation has no CDS credential and
the service requires a logged-in account with its licence accepted. Extraction
therefore uses the time-series-optimised layout of the anonymous
[AWS Open Data Icechunk ERA5 archive](https://registry.opendata.aws/earthmover-era5/),
which is losslessly derived from ECMWF/NCAR source files. A documented ERA5-only
mirror was also tested, but it altered rainfall totals in a five-cell check and
was rejected rather than mixed with the AWS values.

The Bureau's publicly documented
[ACCESS products](https://www.bom.gov.au/nwp/doc/access/docs/)
are operational forecasts rather than a single reanalysis covering all five
events. Historical ACCESS archives are available through registered Bureau/NCI
services, so ACCESS is not mixed with ERA5 in this first consistent series. It
can be added later as an independent 2024 sensitivity if archive access is
provided.

# Coverage and class size

```{r coverage}
cells |>
  group_by(year) |>
  summarise(
    `eReefs cells` = n(),
    `Complete Q1 salinity` = first(salinity_complete_q1),
    `Forcing warning` = first(salinity_forcing_warning),
    `Cells below 30 PSU` = sum(flood30),
    `Cells below 28 PSU` = sum(flood28),
    `Minimum salinity` = round(min(salinity_min), 2),
    .groups = 'drop'
  ) |>
  kable(caption = 'Low-salinity sample size under the expanded threshold')
```

# Rainfall and freshwater exposure

```{r rainfall-relationship, fig.width=10, fig.height=5.5}
cells |>
  filter(year %in% c(2016, 2017, 2020)) |>
  ggplot(aes(coastal_rain_max_30day, salinity_min)) +
  geom_point(colour = 'grey55', alpha = 0.2, size = 0.8) +
  geom_point(
    data = ~ filter(.x, flood30),
    aes(colour = freshwater_exposure_30),
    size = 2
  ) +
  geom_hline(yintercept = 28, linetype = 3) +
  geom_hline(yintercept = 30, linetype = 2) +
  facet_wrap(~ year) +
  scale_x_log10() +
  scale_colour_viridis_c(option = 'C', trans = 'sqrt') +
  labs(
    x = 'Nearest-coastal ERA5 maximum 30-day rain (mm)',
    y = 'eReefs Q1 minimum surface salinity (PSU)',
    colour = 'Exposure below\n30 PSU-days',
    title = 'Rainfall is tested against event-specific low-salinity exposure'
  )
```

The coastal series is the nearest ERA5 land-cell centre. It is more defensible
than rainfall at an offshore reef pixel, but it is still not catchment rainfall,
runoff, river discharge, or plume transport. Wind and Kd490 help represent
transport and optical plume conditions without changing that limitation.

# Leave-one-event-out occurrence performance

```{r occurrence-table}
occurrence |>
  filter(held_out_year == 'Event mean') |>
  select(
    Model = model,
    `Average precision <30` = average_precision30,
    `Balanced Brier <30` = balanced_brier30,
    `Top 1% recall <30` = top_1pct_recall30,
    `Top 5% recall <30` = top_5pct_recall30,
    `Top 1% recall <28` = severe28_top_1pct_recall,
    `Top 5% recall <28` = severe28_top_5pct_recall
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  arrange(desc(`Average precision <30`)) |>
  kable(caption = 'Event-balanced freshwater-occurrence validation')
```

```{r occurrence-event, fig.width=10, fig.height=5.5}
occurrence |>
  filter(held_out_year != 'Event mean') |>
  mutate(held_out_year = factor(held_out_year)) |>
  ggplot(aes(model, average_precision30, colour = held_out_year, group = held_out_year)) +
  geom_line() +
  geom_point(size = 2.5) +
  scale_colour_viridis_d(option = 'C', name = 'Held-out event') +
  coord_flip() +
  labs(
    x = NULL,
    y = 'Average precision for exposure below 30 PSU',
    title = 'Transfer must be consistent across events'
  )
```

# Minimum-salinity reconstruction

```{r magnitude-table}
magnitude |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  select(
    Model = model,
    `Held-out event` = held_out_year,
    `All-cell MAE` = all_cell_mae,
    `Below-30 cell MAE` = flood30_cell_mae,
    `Below-30 bias` = flood30_cell_bias,
    `Below-30 detected` = flood30_detection,
    `Below-28 detected` = flood28_detection
  ) |>
  kable(caption = 'Continuous reconstruction of minimum salinity')
```

# 2024 sensitivity layer

```{r map-2024, fig.width=8.5, fig.height=8}
ggplot(risk_2024, aes(lon, lat, colour = risk30_probability)) +
  geom_point(size = 0.55) +
  geom_point(
    data = filter(risk_2024, partial_ereefs_flood30),
    shape = 1, colour = 'black', size = 2, stroke = 0.7
  ) +
  coord_equal() +
  scale_colour_viridis_c(option = 'C', trans = 'sqrt') +
  labs(
    x = 'Longitude', y = 'Latitude',
    colour = 'Predicted P\n(<30 PSU)',
    title = 'Rainfall-assisted 2024 freshwater-risk sensitivity',
    subtitle = 'Black rings: direct eReefs exposure observed during 1--17 January'
  )
```

The map remains a sensitivity layer until the event-held-out results above
demonstrate improvement over the spatial prior. It is not automatically added
to the mortality model, and the partial January 2024 eReefs record is not
treated as a full-quarter validation target.
