---
title: 'Where the selected mortality model still fails'
subtitle: 'A critical audit of event, region, reef, cause and observation-scale residual variation'
date: 2026-09-15
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)
library(tidyr)
library(yaml)

theme_set(theme_minimal(base_size = 12))

root <- if (file.exists('config/model_registry.yml')) '.' else '../..'
audit_dir <- file.path(root, 'output', 'selected_model_residual_gap_audit')
overview <- read_csv(file.path(audit_dir, 'overview.csv'), show_col_types = FALSE)
events <- read_csv(file.path(audit_dir, 'event_summary.csv'), show_col_types = FALSE)
programmes <- read_csv(file.path(audit_dir, 'programme_summary.csv'), show_col_types = FALSE)
regions <- read_csv(file.path(audit_dir, 'region_summary.csv'), show_col_types = FALSE)
event_regions <- read_csv(
  file.path(audit_dir, 'event_region_summary.csv'), show_col_types = FALSE
)
causes <- read_csv(file.path(audit_dir, 'cause_summary.csv'), show_col_types = FALSE)
reefs <- read_csv(file.path(audit_dir, 'reef_summary.csv'), show_col_types = FALSE)
reef_events <- read_csv(
  file.path(audit_dir, 'reef_event_summary.csv'), show_col_types = FALSE
)
priority <- read_csv(
  file.path(audit_dir, 'priority_reef_events.csv'), show_col_types = FALSE
)
associations <- read_csv(
  file.path(audit_dir, 'residual_associations.csv'), show_col_types = FALSE
)
decomposition <- read_csv(
  file.path(audit_dir, 'error_decomposition.csv'), show_col_types = FALSE
)
discordance <- read_csv(
  file.path(audit_dir, 'cross_programme_discordance.csv'),
  show_col_types = FALSE
)
persistence <- read_csv(
  file.path(audit_dir, 'reef_persistence_summary.csv'),
  show_col_types = FALSE
)
coverage <- read_csv(
  file.path(audit_dir, 'mechanism_coverage.csv'), show_col_types = FALSE
)
model_registry <- read_yaml(file.path(root, 'config', 'model_registry.yml'))
local_dhw_layer <- read_csv(
  file.path(
    root, 'data', 'processed',
    'noaa_dhw_correction_layer_local_first_validation.csv'
  ),
  show_col_types = FALSE
)
local_dhw_site_events <- read_csv(
  file.path(
    root, 'output', 'dhw_correction_local_first',
    'site_event_discrepancies.csv'
  ),
  show_col_types = FALSE
)
inla_predictions <- read_csv(
  file.path(root, 'output', 'inla_spatiotemporal', 'cv_predictions.csv'),
  show_col_types = FALSE
)
inla_metrics <- read_csv(
  file.path(root, 'output', 'inla_spatiotemporal', 'model_comparison.csv'),
  show_col_types = FALSE
)
selected_predictions <- read_csv(
  file.path(root, 'output', 'prospective_cots_nowcast', 'cv_predictions.csv'),
  show_col_types = FALSE
) |>
  filter(
    candidate == model_registry$best_model$id,
    scheme == 'leave_one_event_out'
  ) |>
  left_join(
    local_dhw_layer |>
      select(
        ReefID, event_year, local_first_correction, correction_source,
        correction_sd
      ),
    by = c('ReefID', 'event_year'), relationship = 'many-to-one'
  ) |>
  mutate(
    local_first_correction = coalesce(local_first_correction, 0),
    correction_source = coalesce(correction_source, 'NOAA unchanged'),
    correction_band = case_when(
      local_first_correction < -0.25 ~ 'Negative correction',
      local_first_correction > 0.25 ~ 'Positive correction',
      TRUE ~ 'NOAA/near zero'
    )
  )
raw_dhw_name <- 'persistent_rw1_cyclone_cots_partial_pool'
local_dhw_name <- 'persistent_rw1_cyclone_cots_local_first_dhw_partial_pool'
dhw_pooled_metrics <- inla_metrics |>
  filter(candidate %in% c(raw_dhw_name, local_dhw_name)) |>
  transmute(
    candidate = if_else(
      candidate == raw_dhw_name, 'NOAA unchanged', 'Local-first DHW'
    ),
    validation = recode(
      scheme,
      leave_one_event_out = 'Leave-one-event-out',
      reef_blocked_5fold = 'Reef-blocked 5-fold'
    ),
    rmse, predictive_r2, severe_rmse, false_extreme_rate
  )
dhw_event_metrics <- inla_predictions |>
  filter(
    scheme == 'leave_one_event_out',
    candidate %in% c(raw_dhw_name, local_dhw_name)
  ) |>
  group_by(candidate, event_year) |>
  summarise(
    rmse = sqrt(mean((observed_mortality - predicted_mortality)^2)),
    bias = mean(predicted_mortality - observed_mortality),
    .groups = 'drop'
  ) |>
  mutate(candidate = if_else(
    candidate == raw_dhw_name, 'NOAA unchanged', 'Local-first DHW'
  )) |>
  select(event_year, candidate, rmse, bias) |>
  pivot_wider(names_from = candidate, values_from = c(rmse, bias)) |>
  mutate(
    rmse_change = `rmse_Local-first DHW` - `rmse_NOAA unchanged`
  )
dhw_2024_residual <- selected_predictions |>
  filter(event_year == 2024) |>
  group_by(correction_band) |>
  summarise(
    observations = n(), reefs = n_distinct(ReefID),
    mean_correction = mean(local_first_correction),
    mean_residual = mean(residual),
    rmse = sqrt(mean(residual^2)),
    sse = sum(residual^2), .groups = 'drop'
  ) |>
  mutate(sse_share = sse / sum(sse))
dhw_2024_correlation <- selected_predictions |>
  filter(event_year == 2024) |>
  summarise(value = cor(
    local_first_correction, residual,
    method = 'spearman', use = 'complete.obs'
  )) |>
  pull(value)
local_2024_coverage <- local_dhw_layer |>
  filter(event_year == 2024) |>
  summarise(
    reefs = n(),
    direct = sum(correction_source == 'current_direct'),
    interpolated = sum(correction_source == 'current_local'),
    unchanged = sum(correction_source == 'none')
  )
local_2024_position <- local_dhw_site_events |>
  filter(event_year == 2024) |>
  group_by(position_rule) |>
  summarise(
    site_events = n(), mean_correction = mean(discrepancy),
    median_correction = median(discrepancy), .groups = 'drop'
  )

fmt <- function(x, digits = 3) format(round(x, digits), nsmall = digits)
event_2024 <- events |> filter(event_year == 2024)
north_2024 <- event_regions |>
  filter(event_year == 2024, region_block == 'Northern GBR')
south_2024 <- event_regions |>
  filter(event_year == 2024, region_block == 'Southern GBR')
central_2024 <- event_regions |>
  filter(event_year == 2024, region_block == 'Central GBR')
cyclone_rows <- causes |> filter(cause_group == 'Cyclone label')
multiple_rows <- causes |>
  filter(cause_group == 'Multiple non-thermal labels')
```

## Decision

Do **not** respond to the recent lack of model improvement by widening the
predictor search. The selected cause-aware model already contains most of the
available broad thermal, freshwater, optical, cyclone, COTS, composition and
history fields. The next useful work is narrower:

1. measure event-timed freshwater exposure at the affected reef rather than
   adding another regional rainfall or colour proxy;
2. rebuild cyclone exposure as local, uncertainty-aware physical forcing and
   recalibrate cyclone *magnitude* separately from activation; and
3. reconcile monitoring footprints and observation error before interpreting
   all residual variation as ecological process.

These are conditional development priorities, not grounds to change the
selected model. More heat summaries, an unconstrained residual learner, a
larger interaction search, a reef identity effect, or a new event intercept are
not supported by the current evidence.

## Scope and interpretation

This audit uses the `r comma(overview$observations)` observation-level
leave-one-event-out predictions for
`operational_rrn_raw_plus_manta_state`: `r comma(overview$reefs)` reefs across
`r overview$events` events. A positive residual is observed minus predicted
mortality, so it represents underprediction. All summaries are cross-fitted;
they are not fitted residuals.

Squared-error share locates the current predictive burden but is influenced by
sample size. RMSE and bias show local severity and direction. Reef-event
summaries average repeated observations so that a heavily sampled reef does
not automatically become the highest priority. Cause labels are supporting
field evidence, not perfect causal truth.

```{r overview-table}
overview |>
  transmute(
    Observations = observations, Reefs = reefs, Events = events,
    RMSE = rmse, MAE = mae, `Predictive R2` = predictive_r2,
    `Severe RMSE` = severe_rmse,
    `Underprediction outliers` = underprediction_outliers,
    `Overprediction outliers` = overprediction_outliers
  ) |>
  kable(digits = 3, caption = 'Selected-model event-held-out performance')
```

## Why recent additions have not moved the model

```{r recent-ledger}
registry <- read_yaml(file.path(root, 'config', 'experiments.yml'))
recent_ids <- c(
  'imos_thermal_metrics', 'early_bleaching_update',
  'prospective_cots_brt', 'standalone_brt_envelope',
  'rainfall_salinity', 'salinity_kd490', 'sst_chla_screen',
  'zero_mortality_curves'
)
recent <- bind_rows(lapply(registry$experiments, function(x) {
  if (!x$id %in% recent_ids) return(NULL)
  tibble(
    Investigation = x$title, Status = x$status, Finding = x$finding
  )
}))
kable(recent, caption = 'Relevant completed investigations from the experiment ledger')
```

The common failure mode is not lack of flexibility. Richer BRTs and residual
learners improve some reef-blocked interpolation but lose skill when an entire
event is withheld. Thermal duration, peaks, MHW categories and empirical
frequency describe stress more completely, but did not provide stable new-event
information. Broad rainfall, colour and optical fields are useful risk proxies,
yet they do not observe the short, local salinity tail. The current ceiling is
therefore dominated by transfer, exposure measurement and outcome
heterogeneity rather than a shortage of candidate terms.

## Events: 2024 dominates, but the direction changes through time

```{r event-table}
events |>
  arrange(event_year) |>
  transmute(
    Event = event_year, N = observations, Reefs = reefs,
    `Observed mean` = observed_mean, `Predicted mean` = predicted_mean,
    Bias = bias_observed_minus_predicted, RMSE = rmse,
    `Severe N` = severe_observations, `Severe RMSE` = severe_rmse,
    `Under outliers` = underprediction_outliers,
    `Over outliers` = overprediction_outliers,
    `Error share` = percent(squared_error_share, accuracy = 0.1)
  ) |>
  kable(digits = 3, caption = 'Error by held-out bleaching event')
```

```{r event-plot, fig.width=9, fig.height=5}
events |>
  mutate(
    direction = if_else(
      bias_observed_minus_predicted >= 0,
      'Underprediction', 'Overprediction'
    ),
    event_year = factor(event_year)
  ) |>
  ggplot(aes(event_year, squared_error_share, fill = direction)) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = percent(squared_error_share, accuracy = 0.1)),
    vjust = -0.35
  ) +
  scale_y_continuous(labels = percent, limits = c(0, 0.65)) +
  scale_fill_manual(values = c(Underprediction = '#B2182B',
                               Overprediction = '#2166AC')) +
  labs(
    x = 'Held-out event', y = 'Share of total squared error', fill = NULL,
    title = 'The 2024 event carries most of the remaining predictive error',
    subtitle = 'Bar colour reports mean residual direction, not every reef'
  ) +
  theme(legend.position = 'top')
```

The 2024 observations contribute
`r percent(event_2024$squared_error_share, accuracy = 0.1)` of total squared
error and have a mean underprediction of
`r fmt(event_2024$bias_observed_minus_predicted)`. The 2016 event is also
underpredicted and has the worst severe-case RMSE. Conversely, 2020 and 2022
are low-mortality events with systematic overprediction. A single global
uplift would therefore improve 2016/2024 while making 2020/2022 worse.

This alternating event pattern is why event identity, retrospective event
intercepts and broad post hoc calibration are unsafe. Any event-scale addition
must exist before prediction and transfer across an entirely withheld event.

## Regions: broad northern and southern 2024 misses, mixed Central error

```{r event-region-plot, fig.width=9, fig.height=5.5}
event_regions |>
  mutate(
    Event = factor(event_year),
    Region = factor(
      region_block,
      levels = c('Northern GBR', 'Central GBR', 'Southern GBR')
    )
  ) |>
  ggplot(aes(Event, Region, fill = bias_observed_minus_predicted)) +
  geom_tile(colour = 'white', linewidth = 0.8) +
  geom_text(aes(label = sprintf('%.3f\\nRMSE %.3f',
                                bias_observed_minus_predicted, rmse)),
            size = 3.2) +
  scale_fill_gradient2(
    low = '#2166AC', mid = 'white', high = '#B2182B', midpoint = 0,
    name = 'Observed -\\npredicted'
  ) +
  labs(
    x = 'Held-out event', y = NULL,
    title = 'Residual direction is event-by-region specific',
    subtitle = 'Cell labels give mean bias and RMSE'
  ) +
  theme(panel.grid = element_blank())
```

Across all years, Central GBR contributes the largest raw error share because
it contains `r comma(regions$observations[regions$region_block == 'Central GBR'])`
observations and many high-variance reefs. That is not a uniform Central bias.
In 2024 its mean bias is only `r fmt(central_2024$bias_observed_minus_predicted)`,
because strong under- and overpredictions cancel. Northern and Southern 2024
are more coherent underpredictions: `r fmt(north_2024$bias_observed_minus_predicted)`
and `r fmt(south_2024$bias_observed_minus_predicted)`, respectively.

This pattern argues against adding a coarse region effect. The model needs to
distinguish local exposure and response within a region, especially around
Jasper-affected Central/Northern reefs and the southern 2024 mortality tail.

## Causes: cyclone and multi-disturbance rows remain disproportionately hard

```{r cause-table}
causes |>
  transmute(
    `Field-attributed group` = cause_group,
    N = observations, Reefs = reefs,
    Bias = bias_observed_minus_predicted, RMSE = rmse,
    `Severe N` = severe_observations,
    `Cyclone uplift` = cyclone_marginal_uplift_mean,
    `COTS uplift` = cots_marginal_uplift_mean,
    `Error share` = percent(squared_error_share, accuracy = 0.1)
  ) |>
  kable(digits = 3, caption = 'Residual burden by recorded disturbance context')
```

Cyclone-labelled and multiple-non-thermal rows comprise only
`r cyclone_rows$observations + multiple_rows$observations` observations but
account for
`r percent(cyclone_rows$squared_error_share + multiple_rows$squared_error_share, accuracy = 0.1)`
of total squared error. Their RMSE values are
`r fmt(cyclone_rows$rmse)` and `r fmt(multiple_rows$rmse)`. The cyclone
component is not simply absent: cyclone-labelled rows receive a mean marginal
uplift of `r fmt(cyclone_rows$cyclone_marginal_uplift_mean)`, yet mortality magnitude
is still underestimated. Multiple-cause rows receive much less cyclone uplift
and have the largest positive bias.

The bleaching-only group also remains difficult. That label should not be read
as proof of a purely thermal mechanism: unrecorded freshwater, disease,
mechanical injury, composition error and survey-scale differences can remain
inside it.

## Reef-events: a small set is ecologically informative

```{r priority-under-table}
priority |>
  filter(direction == 'underpredicted') |>
  slice_head(n = 15) |>
  transmute(
    Event = event_year, Reef = ReefName, Region = region_block,
    Programmes = programmes,
    Observed = observed_mortality, Predicted = predicted_mortality,
    Residual = residual, DHW = ann_maxdhw,
    `Candidate mechanism evidence` = candidate_mechanism_flags
  ) |>
  kable(digits = 3, caption = 'Largest equal-weighted reef-event underpredictions')
```

The highest-priority misses are not one mechanism:

- **Penrith 2017** has an explicit Debbie label and a strongly activated
  cyclone hazard, but the uplift is too small for observed mortality. This is
  a cyclone magnitude failure, not a missing gate.
- **Mackay 2024** combines explicit flood and cyclone labels, low DHW and high
  rain, but its cyclone activation is near zero. The current broad proxies miss
  its local compound exposure.
- **Gannett Cay 2020** remains underpredicted even with maximum COTS activation,
  so timing and magnitude remain uncertain rather than absent.
- **Carter and Yonge 2016** combine high DHW with activated Nathan exposure;
  neither a thermal-only nor cyclone-only explanation is adequate.
- **Swinger, Eyrie and Linnet 2024** carry rainfall or water-colour evidence
  without direct salinity observations. They are freshwater/plume priorities,
  not confirmation that the proxy is causal.
- **Broomfield and Orpheus 2024** have cyclone labels but very low current
  activation, suggesting missed local exposure, attribution mismatch, or both.

The largest overpredictions are equally informative. The Jasper activation is
high at Thetford, St Crispin and Hastings in 2024, yet predicted mortality is
far above observed mortality. Some high-DHW or high-COTS reefs in 2017--2020
also have little observed mortality. Increasing cyclone, freshwater or thermal
coefficients globally would amplify these failures.

## Reef persistence is limited

```{r persistent-table}
persistence |>
  filter(events >= 2) |>
  slice_head(n = 12) |>
  transmute(
    Reef = ReefName, Region = region_block, Events = events,
    `Event years` = event_years,
    `Mean residual` = mean_event_residual,
    `Event RMSE` = event_residual_rmse,
    `Maximum error` = maximum_event_absolute_error,
    `Under events` = underpredicted_events,
    `Over events` = overpredicted_events,
    `Sign consistency` = percent(same_direction_fraction, accuracy = 1)
  ) |>
  kable(digits = 3, caption = 'Repeated reefs with the largest event-level residual RMSE')
```

Penrith is consistently underpredicted in its two represented events. Most
other repeated high-error reefs change residual direction between events.
Mackay, for example, is not a stable positive reef effect across four events;
its 2024 compound event is the dominant miss. A reef-identity predictor would
therefore memorise a small and inconsistent history and would not transfer to
unmonitored reefs.

## Observation disagreement sets a real ceiling

The exact squared-error decomposition assigns
`r percent(decomposition$reef_event_mean_error_share, accuracy = 0.1)` to
reef-event mean error and
`r percent(decomposition$within_reef_event_disagreement_share, accuracy = 0.1)`
to differences among observations of the same reef-event. The latter includes
programme footprint, depth, survey timing, aggregation and ordinary
measurement error. No reef-level environmental predictor can make conflicting
observations simultaneously correct.

```{r discordance-table}
discordance |>
  slice_head(n = 12) |>
  transmute(
    Event = event_year, Reef = ReefName, Region = region_block,
    Programmes = programmes,
    `Observed range` = observed_range,
    `Residual range` = residual_range,
    `Residual signs differ` = residual_directions_disagree
  ) |>
  kable(digits = 3, caption = 'Largest cross-programme reef-event disagreements')
```

Lizard Island NW, Roxburgh, Penrith, MacGillivray and several other reefs have
large cross-programme ranges. The selected model already includes
programme-specific observation layers, but the remaining disagreement argues
for explicit footprint/timing uncertainty or a latent reef-event state before
adding fine-grained ecological interactions.

## Residual associations do not justify more copies of existing proxies

```{r association-plot, fig.width=9, fig.height=7}
associations |>
  filter(model_status == 'already represented') |>
  mutate(label = reorder(label, pooled_within_event_rank_r)) |>
  ggplot(aes(pooled_within_event_rank_r, label, colour = mechanism)) +
  geom_vline(xintercept = 0, colour = 'grey55') +
  geom_point(size = 2.5) +
  labs(
    x = 'Pooled within-programme/event rank association with residual',
    y = NULL, colour = NULL,
    title = 'Most remaining associations are weak or inconsistent',
    subtitle = 'Every field shown is already represented in the selected model'
  ) +
  theme(legend.position = 'bottom')
```

These are diagnostics, not independent tests of new predictors. Rainfall has a
pooled within-event rank association near zero, and the water-colour signs vary
by event and programme. The 2024 water-colour association is positive, but it
is negative in several earlier groups. Freshwater is therefore a plausible
local missing exposure, not a demonstrated GBR-wide residual slope.

Cyclone activation is negatively associated with residuals within every 2024
programme: high-activation Jasper reefs tend to be overpredicted, while several
labelled or field-implicated reefs receive low activation and are
underpredicted. This is evidence against merely lowering the activation
threshold or increasing the cyclone coefficient. The spatial exposure surface
and response magnitude need rebuilding together.

## Local-first logger-adjusted DHW is already in the selected model

**Yes: the registry-selected model already uses local-first DHW.** The model
registry sets `local_dhw_adjustment` to
`r model_registry$best_model$local_dhw_adjustment`, and its thermal core is
`r model_registry$best_model$thermal_core_id`. In the fitted data,
`local_first_correction` is added to immutable NOAA `ann_maxdhw`; the 4- and
8-DHW hinges and current-event novelty are then recomputed. The remaining 2024
underprediction therefore occurs *after* the logger adjustment, not because it
was omitted.

The correction is more conservative than a GBR-wide logger offset. In the 2024
reef-event layer, `r local_2024_coverage$direct` of
`r local_2024_coverage$reefs` reefs have a direct site match,
`r local_2024_coverage$interpolated` use at most four same-event sources within
25 km, and `r local_2024_coverage$unchanged` retain NOAA unchanged. No historical
or event-wide uplift is applied. The layer also carries correction uncertainty.

### The current hierarchy already favours flats and shallow loggers

At each logger site-event, reef-temperature loggers take precedence over an
automated station. If a shallow 0--5 m or reef-flat record exists, only that
subset is summarised; slope or other positions enter the site median only when
no shallow/flat series is available. Thus a strict flat-first idea is already
largely present. What is *not* implemented is an explicit reliability weight
that shrinks the fallback slope/all-position estimate toward zero.

```{r local-dhw-position}
local_2024_position |>
  transmute(
    `Position rule` = recode(
      position_rule,
      shallow_or_flat_preferred = 'Shallow/flat preferred',
      all_available = 'All available (fallback)'
    ),
    `Site-events` = site_events,
    `Mean logger - NOAA DHW` = mean_correction,
    `Median logger - NOAA DHW` = median_correction
  ) |>
  kable(digits = 3, caption = '2024 site-event inputs to the local-first layer')
```

The fallback group is more negative in 2024, so habitat-aware shrinkage is a
reasonable *construct sensitivity*. A blanket flat-only rule is less sound: it
would discard spatial coverage, and shallow flats can be hotter and less
representative of slope-based survey footprints. Matching correction evidence
to monitoring depth/habitat is preferable to assuming one habitat is correct
for every programme.

### The adjustment helped transfer, but not uniformly

Within the same earlier cyclone/COTS INLA architecture, local-first DHW reduced
pooled leave-one-event-out error, while slightly worsening reef-blocked error.
This is why it was retained as an exposure correction but not treated as a
universally superior replacement for NOAA.

```{r local-dhw-pooled}
dhw_pooled_metrics |>
  transmute(
    `DHW input` = candidate, Validation = validation, RMSE = rmse,
    `Predictive R2` = predictive_r2, `Severe RMSE` = severe_rmse,
    `False-extreme rate` = false_extreme_rate
  ) |>
  arrange(Validation, RMSE) |>
  kable(digits = 3, caption = 'Existing validation of the local-first correction')
```

The gain is concentrated in 2020, 2022 and especially 2024; 2016 and 2017
worsen. This pattern argues for retaining the correction and its uncertainty,
but against tuning a logger-habitat weight to the pooled mortality outcome.

```{r local-dhw-event}
dhw_event_metrics |>
  transmute(
    Event = event_year,
    `NOAA RMSE` = `rmse_NOAA unchanged`,
    `Local-first RMSE` = `rmse_Local-first DHW`,
    `Local - NOAA RMSE` = rmse_change,
    `NOAA bias (predicted - observed)` = `bias_NOAA unchanged`,
    `Local bias (predicted - observed)` = `bias_Local-first DHW`
  ) |>
  kable(digits = 3, caption = 'Event-held-out DHW correction comparison')
```

An earlier fixed +4.82-DHW flat-logger stress test at six Lizard-region reefs
also raised their held-out 2024 predictions, but it was a local stress test,
not a general flat-only correction layer. Its modest model-wide improvement
cannot justify exporting that positive uplift to other reefs.

### Would downweighting slopes close the current 2024 gap?

Probably only marginally, if at all. In the *selected* model's 2024 held-out
predictions, the rank correlation between the applied correction and remaining
residual is `r fmt(dhw_2024_correlation)`. Negative-correction reefs are
underpredicted on average, so shrinking some fallbacks toward NOAA could help
individual cases. But positive-correction reefs have at least as much residual
burden, and Mackay remains the largest 2024 miss despite a direct +2.99-DHW
correction. The residual pattern is not a monotone signature of logger
corrections being too negative.

```{r local-dhw-residual}
dhw_2024_residual |>
  transmute(
    `Correction band` = correction_band, Observations = observations,
    Reefs = reefs, `Mean correction (DHW)` = mean_correction,
    `Mean residual` = mean_residual, RMSE = rmse,
    `Share of 2024 SSE` = sse_share
  ) |>
  arrange(`Mean correction (DHW)`) |>
  kable(digits = 3, caption = 'Residual error after local-first DHW is already applied')
```

The defensible conclusion is to **keep local-first DHW in the current model**,
not replace it with raw NOAA and not promote a flat-only rule. If this layer is
revisited, the narrow test should partially pool fallback slope/all-position
corrections toward zero using held-out *logger-versus-NOAA* accuracy and
survey-habitat/depth compatibility, then rerun the fixed mortality folds. The
shrinkage must be learned without mortality outcomes. Given the near-zero
residual association and the already positive correction at Mackay, this is a
secondary calibration refinement, not a likely substitute for event-timed
freshwater and local cyclone forcing.

## Only three next steps survive this audit

### 1. Direct, event-timed freshwater exposure at priority reefs

Build a two-part freshwater layer representing probability of an ecologically
important low-salinity pulse and conditional deficit/duration. Use direct
salinity when available and treat rainfall, river discharge, colour and
hydrodynamic connectivity as uncertain predictors of that latent exposure—not
as interchangeable salinity measurements. Prioritise Mackay, Swinger, Eyrie,
Linnet and appropriate low-impact controls in the same event and region.

Promotion should require leave-one-event-out validation, a river/catchment or
spatial holdout, improved severe-case error, and credible low-salinity-tail
calibration. The current selected-model rows have complete broad rainfall and
water-colour proxies but **no direct salinity predictor**. Adding further sums,
lags or interactions of the same proxy products is unlikely to close the gap.

### 2. Local cyclone forcing and a separate damage-magnitude calibration

Replace the single deterministic activation view with an uncertainty-aware
exposure distribution combining storm track, event timing, wave energy and
direction, bathymetry, reef aspect/sheltering and depth. Keep forcing,
activation and mortality magnitude distinct. Penrith/Carter/Yonge indicate
underestimated magnitude despite activation; Mackay/Broomfield/Orpheus indicate
possible missed activation; Thetford/St Crispin/Hastings are essential
high-exposure, lower-loss negative controls.

Promotion should be based on held-out cyclone events or storms, labelled and
unlabelled negative controls, severe RMSE, false-extreme rate and calibration of
the cyclone component. A more permissive gate alone is contraindicated by the
2024 overpredictions.

### 3. Observation-footprint and latent reef-event reconciliation

Quantify how LTMP, manta and MMP differ in spatial footprint, depth, survey
date, baseline uncertainty and taxonomic aggregation. Carry those uncertainties
into a latent reef-event mortality layer or observation-error sensitivity.
This is not a device for shrinking inconvenient outcomes: the latent state must
be estimated without using event identity as an operational predictor, and the
raw programme observations must remain auditable.

This work could explain the 27% within-reef-event error component and prevent
environmental predictors from being selected to fit observation mismatch.

## Steps that should not proceed now

| Proposed move | Why it is not justified |
|---|---|
| Add more generic thermal metrics | MHW, intensity-duration, empirical frequency and cooling screens did not transfer consistently; local-first DHW is already present. |
| Replace local-first DHW with flat-only loggers | The current hierarchy already prefers shallow/flat records; a blanket restriction loses coverage and may mismatch slope surveys. |
| Add more rainfall/colour lags and interactions | These axes are already richly represented; residual signs vary by event and programme. |
| Lower the cyclone gate or increase its coefficient | High-activation Jasper reefs are already overpredicted while some labelled reefs remain unactivated. |
| Reintroduce a residual BRT | It improves known-event interpolation but worsens event-held-out and severe error. |
| Add reef identity or event identity | Reef residuals are often sign-inconsistent, and event identity is unavailable prospectively. |
| Downsample zeros or delete low-DHW losses | Previous tests worsened transfer and would discard plausible compound events. |
| Treat optical plume products as salinity | Their low-salinity-tail calibration is inadequate and the apparent skill relies on spatial priors. |
| Promote satellite night cooling | The proposed day/night contrast failed hourly-logger construct validation. |

## Promotion sequence

The next comparison should not be one large model containing all three changes.
Run locked ablations in this order:

1. selected model versus a new freshwater *measurement* layer;
2. selected model versus rebuilt cyclone exposure, then exposure plus a new
   cyclone magnitude layer;
3. repeat both comparisons under the observation-footprint sensitivity; and
4. test the combined freshwater-plus-cyclone candidate only if each component
   improves its relevant held-out cases without worsening overall event
   transfer or false extremes.

The operational model remains unchanged unless those gates pass.
