---
title: "Repeated heat exposure and susceptible coral composition"
subtitle: "Implications of McWilliam et al. (2026) for the GBR mortality models"
date: today
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)

theme_set(theme_minimal(base_size = 12))

exposure <- read_csv(
  '../data/processed/repeated_dhw_exposure_audit.csv',
  show_col_types = FALSE
)
composition <- read_csv(
  '../data/processed/acropora_resolution_audit.csv',
  show_col_types = FALSE
)
```

## Decision summary

McWilliam et al. (2026) strengthens the case that present DHW is not enough.
Two pre-event histories need to remain distinct:

1. **Years since the last >6-DHW event** represents the interval available for
   coral recovery and reassembly of susceptible taxa.
2. **Number of prior >6-DHW events since 2016** represents repeated exposure
   and selective loss of susceptible communities.

The pipeline previously had a leakage-safe years-since variable and a ten-year
thermal-load summary, but it did not use an explicit fixed-origin event count.
That count is now reconstructed for every operational reef feature. A rolling
eight-year count is also retained as a future-facing sensitivity metric, but it
should not enter the same small BRMS formula as the fixed count unless blocked
validation shows that both add independent information.

The paper also shows that total *Acropora* is an incomplete susceptibility
measure. Tabular and staghorn *Acropora* carry much of the signal. The local
AIMS workspace only resolves the relevant field to total *Acropora*; it does
not contain a tabular/staghorn split. Total *Acropora* remains the operational
fallback, but the missing morphology layer is now explicit rather than hidden.

## What the paper shows

The 2024 response was shaped by ecological memory. Reefs with repeated severe
heat exposure since 2016 often had depleted heat-sensitive taxa and lower 2024
bleaching or mortality. Reefs with a seven-to-eight-year recovery interval had
rebuilt fast-growing branching and tabular *Acropora* and could suffer severe
loss when exposed again. At a given DHW, high tabular/staghorn *Acropora* and
low massive or branching *Porites* were associated with positive bleaching
residuals.

This is best treated as a mechanism and prior expectation, not as a coefficient
to copy directly. The paper's intensive 2024 community analysis covers 23 reefs
and immediate bleaching/mortality, whereas this pipeline predicts broader
programme observations and annual cover change across multiple events.

## Leakage-safe exposure reconstruction

All history fields stop at 31 December of the year before the response. The
current event's DHW remains a separate predictor. This matters most for 2024:
including the 2024 event in the event-frequency count would leak the response
year's thermal exposure into a variable described as ecological history.

The new fields are:

| Field | Definition | Default role |
|---|---|---|
| `dhw_events_since2016_n6` | Count of annual maxima >6 DHW from 2016 through target year minus one | Primary repeated-exposure predictor |
| `dhw_years_since_last_n6` | Uncapped years since the last pre-event annual maximum >6 DHW | Audit and interpretation |
| `dhw_years_since_last_n6_capped8` | Years since last event, capped at eight; no previous event also maps to eight and is identified separately | Primary recovery-interval predictor |
| `dhw_no_prior_n6` | Censoring flag where no previous >6-DHW year exists | Keeps “no event observed” distinct from a long known interval |
| `dhw_events_prior8_n6` | Count in the preceding eight complete years | Operational sensitivity after the fixed 2016 origin ages |
| `dhw_history_years_since2016` | Number of observed historical years contributing to the fixed count | Provenance/coverage, not an ecological effect |

```{r}
exposure_2024 <- exposure |> filter(year == 2024)

exposure_2024 |>
  count(dhw_events_since2016_n6) |>
  mutate(proportion = n / sum(n)) |>
  ggplot(aes(factor(dhw_events_since2016_n6), proportion)) +
  geom_col(fill = '#2166ac') +
  scale_y_continuous(labels = percent) +
  labs(
    title = 'Prior >6-DHW event frequency varies strongly across the GBR',
    subtitle = 'History is 2016--2023; the 2024 response year is excluded',
    x = 'Prior events since 2016', y = 'GBR reef features'
  )
```

In 2024, `r percent(mean(exposure_2024$dhw_events_since2016_n6 >= 3), accuracy = 0.1)`
of operational reef features had at least three prior >6-DHW years, while
`r percent(mean(exposure_2024$dhw_events_since2016_n6 == 0), accuracy = 0.1)`
had none.

```{r}
target_history <- exposure_2024 |>
  filter(grepl('Mackay|Linnet|Lizard|Eyrie|Swinger', LOC_NAME_S,
               ignore.case = TRUE)) |>
  transmute(
    Reef = LOC_NAME_S, DHW_2024 = round(ann_maxdhw, 2),
    `Prior events since 2016` = dhw_events_since2016_n6,
    `Years since last >6 DHW` = dhw_years_since_last_n6,
    `No prior >6 DHW` = as.logical(dhw_no_prior_n6)
  ) |>
  arrange(Reef)
kable(target_history, caption = 'Exposure history at reefs under review')
```

## Composition resolution

```{r}
composition |>
  select(composition_field, available, rows, reefs, first_year, last_year,
         source_categories, source_variables) |>
  kable(caption = 'Resolution available in the local AIMS community workspace')
```

The audit finds total *Acropora* observations but no category or source
variable identifying tabular, staghorn or branching *Acropora*. Those values
must come from a more resolved image/transect product or a separately validated
community model. They must not be inferred by setting a fixed fraction of total
*Acropora* or by filling missing values with zero.

For operational GBR-wide predictions, the staged composition contract is:

- retain past-only total *Acropora* and total coral cover everywhere;
- add observed tabular/staghorn cover when a taxonomically resolved source is
  recovered;
- if a GBR-wide morphology model is later built, carry its prediction and
  uncertainty separately from observed composition; and
- report performance for observed-morphology and imputed-morphology subsets.

## Model comparison to run next

The paper implies both direct history and composition-mediated pathways. The
next blocked comparison should therefore be deliberately small:

1. current production history (`yrsince6`, chronic load and novelty);
2. add prior event count plus its DHW interaction;
3. add total *Acropora* plus DHW--*Acropora* interaction;
4. add tabular/staghorn *Acropora* only when a defensible source exists.

BRMS should prespecify the DHW interactions and use regularising priors; BRT can
learn the corresponding nonlinear interactions. Evaluation must retain
leave-one-event-out, unseen-reef and reef-blocked-2024 views, with severe-case
bias and calibration alongside RMSE, MAE, predictive R2 and information
criteria for comparable likelihood models.

The key diagnostic is whether repeated exposure reduces predicted response at
a fixed DHW because susceptible communities have been filtered, while a long
recovery interval and high susceptible composition increase it. If the history
effect disappears after resolved composition is added, that is scientifically
informative mediation rather than a failed predictor.
