---
title: 'RRN water-colour and cyclone pressure assessment'
subtitle: 'Freshwater proxy validation and blocked mortality-model ablations'
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)
library(tidyr)

input_dir <- '../../output/rrn_pressure_assessment'
coverage <- read_csv(file.path(input_dir, 'proxy_coverage.csv'), show_col_types = FALSE)
rank_correlations <- read_csv(
  file.path(input_dir, 'proxy_rank_correlations.csv'), show_col_types = FALSE
)
proxy_metrics <- read_csv(
  file.path(input_dir, 'proxy_loyo_metrics.csv'), show_col_types = FALSE
)
proxy_predictions <- read_csv(
  file.path(input_dir, 'proxy_loyo_predictions.csv'), show_col_types = FALSE
)
mortality_metrics <- read_csv(
  file.path(input_dir, 'mortality_brt_metrics.csv'), show_col_types = FALSE
)
programme_metrics <- read_csv(
  file.path(input_dir, 'mortality_brt_programme_metrics.csv'), show_col_types = FALSE
)
paired <- read_csv(
  file.path(input_dir, 'mortality_brt_paired_rows.csv'), show_col_types = FALSE
)
severe_2024 <- read_csv(
  file.path(input_dir, 'severe_2024_reef_diagnostics.csv'), show_col_types = FALSE
)
brms_effects <- read_csv(
  file.path(input_dir, 'rrn_brms_effects.csv'), show_col_types = FALSE
)
brms_diagnostics <- read_csv(
  file.path(input_dir, 'rrn_brms_diagnostics.csv'), show_col_types = FALSE
)
brms_comparison <- read_csv(
  file.path(input_dir, 'rrn_brms_model_comparison.csv'), show_col_types = FALSE
)

model_labels <- c(
  core = 'Core ecological BRT',
  weather = 'Core + rainfall/wind',
  rrn_wq = 'Core + RRN water colour',
  rrn_cyclone = 'Core + cyclone waves',
  rrn_both = 'Core + both RRN pressures',
  weather_rrn = 'Weather + raw WQC + cyclone waves',
  weather_rrn_relative = 'Weather + relative WQC + cyclone waves',
  weather_rrn_full = 'Weather + relative WQC + cyclone waves + COTS'
)
programme_labels <- c(
  ltmp = 'LTMP benthic', manta = 'Manta tow', mmp = 'MMP inshore'
)
proxy_labels <- c(
  spatial = 'Spatial baseline',
  rainfall = 'ERA5 rainfall',
  imos_kd490 = 'IMOS Kd490',
  rrn_wq = 'RRN water colour',
  rrn_wq_rainfall = 'RRN water colour + rainfall',
  all_proxies = 'Water colour + rainfall + Kd490'
)
```

# Decision

The RRN water-colour layer is a **useful additional freshwater/plume proxy**,
but not a replacement for IMOS Kd490 or direct salinity. Absolute WQC has a
positive all-event Bayesian association with mortality and gives more
consistent unseen-event prediction than a within-reef preceding-decade
percentile. The DHW-by-WQC interaction remains uncertain.

The cyclone-wave layer is ecologically useful for separating potential
mechanical cyclone exposure from thermal mortality. Its predictive gain is
small and programme-specific, and the Bayesian mean-response effect is not
positive after adjustment for the core ecological variables.

The weather + raw WQC + cyclone candidate is retained for the predictive
ensemble: it improves overall leave-one-event-out RMSE relative to weather
alone and has the lowest overall reef-blocked 2024 RMSE. It is not used as the
sole production route because gains vary by programme and WQC arrives one wet
season late. Rainfall/Kd490 remains the operational fallback.

Modelled COTS pressure is retained in the research data but is not promoted to
the predictive route. Adding it worsens blocked BRT prediction and produces no
meaningful LOO improvement or clearly supported BRMS mean effect.

# Reef identity and disturbance-lineage correction

The earlier cleaned-name crosswalk collapsed `MACKAY REEF` and
`MACKAY REEFS` to the same key. Exact-name-first matching now assigns the raw
2024 AIMS survey and its Jasper/flood-water tooltip to **Mackay Reef
(`16-015`)**, not Mackay Reefs (`15-024`). Both manta and LTMP benthic rows now
use the correct `16-015` coordinates and an observed pre-event Acropora
proportion of 0.111. The raw disturbance tooltip is retained in the canonical
outcome tables and provides explicit bleaching, cyclone and flood flags.

# Data alignment and coverage

The supplied workbook uses austral-summer codes. Summer `202324` ends in
2024 and is therefore aligned to the 2024 bleaching-mortality event. This
season is present for WQC, cyclone waves and the observed/modelled COTS
layers. The release delay affects when the
water-colour layer becomes available, not its retrospective coverage of 2024.

The reef-relative WQC feature is the mid-rank of the current frequency against
only the preceding ten event years; the current event is excluded from its
reference distribution. Modelled IDW COTS is used for prediction because raw
manta-tow counts are absent at unsurveyed reefs. Sixteen 2024 MMP observations
at four reefs lack the modelled COTS surface and are imputed from each analysis
fold rather than from the assessment data.

The extracted layer contains 3,862 GBRMPA reef IDs. It matches all 537
mortality observations directly by `LABEL_ID` and event year; no coordinate
matching or imputation is needed. In the larger eReefs grid, the RRN reef set
covers about 53.5% of cells in each complete salinity year.

```{r coverage-table}
coverage |>
  mutate(overlap_percent = round(overlap_percent, 1)) |>
  rename(
    Year = year,
    `eReefs cells` = ereefs_reef_cells,
    `RRN water-colour cells` = rrn_wq_cells,
    `RRN cyclone cells` = rrn_cyclone_cells,
    `Overlap (%)` = overlap_percent
  ) |>
  kable(caption = 'Overlap used for salinity-proxy validation')
```

# Does water colour reproduce low-salinity patterns?

Proxy validation uses only the complete 2016, 2017, 2020 and 2022 eReefs
January--March salinity fields. The partial 2024 salinity field is excluded.
There are only 47 below-30 PSU reef-cells among 15,120 matched reef-years, so
ordinary classification accuracy would be uninformative. Models are compared
by leave-one-event-out average precision, ROC AUC, balanced Brier score, and
continuous minimum-salinity error.

```{r proxy-metrics-table}
proxy_metrics |>
  mutate(
    Proxy = proxy_labels[model],
    across(
      c(average_precision, roc_auc, balanced_brier, salinity_rmse,
        salinity_mae, salinity_r2),
      ~ round(.x, 3)
    )
  ) |>
  select(
    Proxy, n, `Below-30 cells` = flood30_cells,
    `Average precision` = average_precision,
    `ROC AUC` = roc_auc,
    `Balanced Brier` = balanced_brier,
    `Minimum-salinity RMSE` = salinity_rmse,
    `Minimum-salinity MAE` = salinity_mae,
    `Minimum-salinity R2` = salinity_r2
  ) |>
  kable(caption = 'Leave-one-event-out salinity-proxy performance')
```

Water colour alone identifies rare below-30 PSU cells better than rainfall
alone (average precision 0.133 versus 0.101) and is similar to the spatial
baseline. Combining water colour with rainfall improves average precision to
0.146. IMOS Kd490 remains the best single classifier at 0.159. For continuous
minimum salinity, the combined water-colour, rainfall and Kd490 model performs
best (RMSE 0.610 PSU, predictive R2 0.316).

```{r proxy-plot, fig.width=10, fig.height=5.8}
proxy_metrics |>
  mutate(Proxy = factor(proxy_labels[model], levels = rev(proxy_labels))) |>
  ggplot(aes(average_precision, Proxy)) +
  geom_col(fill = '#287D8E') +
  geom_text(aes(label = sprintf('%.3f', average_precision)), hjust = -0.15) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(
    x = 'Leave-one-event-out average precision for <30 PSU',
    y = NULL,
    title = 'Water colour improves on rainfall, but IMOS Kd490 remains stronger'
  )
```

```{r salinity-scatter, fig.width=9.5, fig.height=6}
proxy_predictions |>
  filter(model == 'spatial') |>
  ggplot(aes(wqc_freqcc12, salinity_min)) +
  geom_bin_2d(bins = 35) +
  geom_hline(yintercept = 30, linetype = 2, colour = '#B2182B') +
  facet_wrap(~ year) +
  scale_fill_viridis_c(trans = 'log10') +
  labs(
    x = 'RRN wet-season coloured-water frequency',
    y = 'eReefs minimum Q1 salinity (PSU)',
    fill = 'Reef-years',
    title = 'The water-colour relationship is event dependent'
  )
```

Spearman correlation with minimum salinity is -0.397 in 2016 and -0.658 in
2017, but weak in 2022. The product is therefore informative without being a
direct salinity substitute.

# Does either layer improve mortality prediction?

Every BRT candidate retains DHW response shape, preceding Acropora proportion,
initial coral cover, prior heat exposure, IMOS Kd490, cloud, depth and
programme effects. The only difference is the added pressure block. Each
candidate is tuned within identical outer folds.

```{r mortality-overall-plot, fig.width=10, fig.height=6}
mortality_metrics |>
  mutate(
    Model = factor(model_labels[model], levels = rev(model_labels)),
    Target = recode(
      scheme,
      leave_one_event_out = 'Unseen event',
      reef_blocked_2024 = '2024 reef-blocked'
    )
  ) |>
  ggplot(aes(rmse, Model, fill = Target)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.72) +
  geom_text(
    aes(label = sprintf('%.3f', rmse)),
    position = position_dodge(width = 0.8), hjust = -0.1, size = 3
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.14))) +
  scale_fill_manual(values = c(
    'Unseen event' = '#4D4D4D', '2024 reef-blocked' = '#D95F02'
  )) +
  labs(
    x = 'RMSE', y = NULL, fill = NULL,
    title = 'Predictive gains are small and depend on the validation target'
  ) +
  theme(legend.position = 'top')
```

For a completely unseen event, core + raw water colour has the lowest overall
RMSE (0.179). Adding raw WQC and cyclone waves to the weather model improves
RMSE from 0.185 to 0.180. Replacing raw WQC with the preceding-decade
percentile is weaker (0.183), and adding modelled COTS worsens RMSE to 0.191.

For strict reef-blocked 2024 prediction, weather + raw WQC + cyclone waves has
the lowest overall RMSE (0.174) and predictive R2 of 0.285. The corresponding
COTS candidate has RMSE 0.175 and higher severe-case MAE, so COTS does not
clear the predictive gate.

```{r programme-table}
programme_metrics |>
  filter(scheme == 'reef_blocked_2024') |>
  mutate(
    Programme = programme_labels[programme_key],
    Model = model_labels[model],
    across(
      c(rmse, mae, severe_mae, bias_observed_minus_predicted),
      ~ round(.x, 3)
    )
  ) |>
  select(
    Programme, Model, RMSE = rmse, MAE = mae,
    `Severe MAE` = severe_mae,
    `Observed - predicted` = bias_observed_minus_predicted
  ) |>
  arrange(Programme, RMSE) |>
  kable(caption = 'Strict reef-blocked 2024 performance by programme')
```

The programme split matters:

- LTMP overall RMSE remains lowest for the core model (0.110), while relative
  WQC gives the lowest severe-case MAE (0.147).
- Manta is best with weather + relative WQC (RMSE 0.158), virtually tied with
  weather alone; both are substantially better than the corrected core model.
- MMP overall RMSE remains lowest for the core model (0.246), while weather +
  raw WQC + cyclone gives the lowest severe-case MAE (0.238).

# What happens at severe 2024 manta reefs?

```{r gain-plot, fig.width=10, fig.height=6}
paired |>
  filter(
    scheme == 'reef_blocked_2024',
    programme_key == 'manta',
    observed_mortality >= 0.2
  ) |>
  select(
    ReefID,
    `Raw WQC + waves vs weather` = weather_rrn_gain_vs_weather,
    `Relative WQC vs raw WQC` = relative_wq_gain_vs_weather_rrn,
    `Add COTS to relative WQC` = cots_gain_vs_relative
  ) |>
  pivot_longer(
    -ReefID, names_to = 'Candidate change',
    values_to = 'Error reduction'
  ) |>
  ggplot(aes(`Candidate change`, `Error reduction`)) +
  geom_hline(yintercept = 0, colour = 'grey45') +
  geom_boxplot(fill = '#A6CEE3', outlier.shape = NA) +
  geom_jitter(width = 0.12, alpha = 0.45, colour = '#2C7FB8') +
  coord_flip() +
  labs(
    x = NULL,
    y = 'Reduction in absolute error for the stated comparison',
    title = 'COTS does not systematically improve severe manta predictions'
  )
```

```{r severe-table}
severe_2024 |>
  filter(programme_key == 'manta') |>
  transmute(
    Reef = ReefName,
    Region = region_block,
    DHW = round(ann_maxdhw, 2),
    `Pre-event Acropora` = percent(prop_acropora_pre, accuracy = 1),
    `Water colour` = round(wqc_freqcc12, 3),
    `Within-reef WQC percentile` = percent(
      wqc_prior10_percentile, accuracy = 1
    ),
    `Cyclone hours` = round(cyc_maxHrs4mw, 1),
    `Modelled COTS/tow` = round(cot_idwmeanpertow, 4),
    Observed = round(observed_mortality, 3),
    `Core underprediction` = round(core_underprediction, 3),
    `Water-colour underprediction` = round(wq_underprediction, 3),
    `Cyclone underprediction` = round(cyclone_underprediction, 3),
    `Weather underprediction` = round(weather_underprediction, 3),
    `Full candidate underprediction` = round(full_underprediction, 3),
    Disturbance = disturbance_text
  ) |>
  head(15) |>
  kable(caption = 'Largest severe 2024 manta underpredictions')
```

Hastings Reef has substantial modelled 4 m-wave exposure and benefits from the
cyclone candidate. High-WQC cases also improve slightly on average. However,
the variables do not explain the largest residuals consistently. Corrected
Mackay (`16-015`) remains underpredicted by about 0.44 even with rainfall,
relative WQC, cyclone waves and COTS. Its retained tooltip explicitly identifies
bleaching, Cyclone Jasper and flood waters, indicating that seasonal/grid-scale
proxies still miss the acute local freshwater exposure. Linnet, Lizard Island
NW, Eyrie and Swinger also remain ecological review priorities.

# Bayesian mechanism check

```{r brms-effect-table}
brms_effects |>
  filter(model == 'weather_rrn_cots') |>
  mutate(
    Effect = recode(
      effect,
      rainfall_mean = 'Rainfall: positive-mortality mean',
      wq_mean = 'Water colour: positive-mortality mean',
      cyclone_mean = 'Cyclone waves: positive-mortality mean',
      dhw_by_rainfall_mean = 'DHW x rainfall: positive-mortality mean',
      dhw_by_wq_mean = 'DHW x water colour: positive-mortality mean',
      wq_zero_one = 'Water colour: zero/one probability',
      cyclone_zero_one = 'Cyclone waves: zero/one probability',
      cots_mean = 'COTS: positive-mortality mean',
      cots_zero_one = 'COTS: zero/one probability'
    ),
    across(
      c(posterior_mean, q025, q975, probability_positive),
      ~ round(.x, 3)
    )
  ) |>
  select(
    Effect, Estimate = posterior_mean,
    `2.5%` = q025, `97.5%` = q975,
    `P(positive)` = probability_positive
  ) |>
  kable(caption = 'Regularised all-event RRN pressure BRMS')
```

Rainfall has strong positive support (standardised logit-scale estimate 0.457,
95% interval 0.228--0.682). Absolute WQC is also probably positive (0.287,
interval -0.006--0.587, P[positive] = 0.971), while its DHW interaction remains
uncertain. The cyclone-wave mean effect is weakly negative and uncertain.

The COTS mean effect is small and uncertain (0.050, interval -0.096--0.190,
P[positive] = 0.754). Its addition changes Bayesian R2 only from 0.334 to
0.338 and improves expected log predictive density by 0.1 with a standard
error of 1.6: effectively no LOO difference. Because both fits contain many
high Pareto-k observations, the blocked BRT comparison is the stronger
predictive evidence and also rejects COTS as a production improvement.

```{r brms-fit-comparison}
brms_comparison |>
  transmute(
    Model = recode(
      model,
      weather_rrn = 'Rainfall + WQC + cyclone waves',
      weather_rrn_cots = 'Rainfall + WQC + cyclone waves + COTS'
    ),
    `ELPD difference` = round(elpd_diff, 2),
    `SE difference` = round(se_diff, 2),
    LOOIC = round(looic.x, 1),
    `Bayesian R2` = round(bayes_r2_mean, 3)
  ) |>
  kable(caption = 'Matched BRMS model-fit comparison')
```

```{r brms-diagnostics}
brms_diagnostics |>
  mutate(
    `Pass gate` = max_rhat <= 1.01 & min_bulk_ess >= 100 &
      min_tail_ess >= 100 & divergences == 0 & max_treedepth_hits == 0,
    across(c(max_rhat, min_bulk_ess, min_tail_ess), ~ round(.x, 3))
  ) |>
  rename(
    `Maximum R-hat` = max_rhat,
    `Minimum bulk ESS` = min_bulk_ess,
    `Minimum tail ESS` = min_tail_ess,
    Divergences = divergences,
    `Tree-depth hits` = max_treedepth_hits
  ) |>
  kable(caption = 'RRN pressure BRMS diagnostics')
```

# Recommended pipeline use

1. Retain rainfall, raw `wqc_freqcc12` and `cyc_maxHrs4mw` as distinct
   predictors in the RRN ensemble candidate. Do not call WQC measured salinity
   or treat cyclone wave-hours as a freshwater metric.
2. Retain the preceding-decade WQC percentile as an anomaly diagnostic and
   future candidate, but prefer raw WQC in the current predictive route.
3. Retain observed and IDW-modelled COTS fields in the research data. Do not
   require COTS in the production route until blocked validation improves.
4. Retain IMOS Kd490 and rainfall as operational inputs when annual WQC is not
   yet available.
5. Preserve the largest unexplained severe residuals as ecological review
   targets rather than flattening the DHW response or treating them as noise.

# Status of the shared latent ecological response

The current joint BRMS is a **shared-effects pooled response model**, not yet a
full latent ecological-state model. It shares ecological slopes and a
reef-event random effect while allowing programme-specific mean intercepts,
beta precision and zero/one probabilities. The BRT likewise includes programme
as a predictor and balances reef-event weights.

The remaining architecture is to estimate one latent reef-event mortality
state and link manta tow, LTMP photo transects and MMP observations to it with
separate programme-specific observation equations, bias, precision and
detection behaviour. That is the next formal modelling stage now that the
pressure feature set has been screened.
