---
title: 'Expanded surface-salinity and freshwater mortality assessment'
subtitle: 'Source audit plus 2011–2024 BRT and INLA screens of minimum SSS and hours below 30 and 26 PSU'
date: 2026-09-28
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(knitr)
  library(readr)
  library(scales)
  library(tidyr)
})

theme_set(theme_minimal(base_size = 12))
root <- if (file.exists('config/model_registry.yml')) '.' else '../..'
out <- file.path(root, 'output', 'surface_salinity_mortality')
processed <- file.path(root, 'data', 'processed')

source_audit <- read_csv(
  file.path(out, 'salinity_source_audit.csv'), show_col_types = FALSE
)
source_contract <- read_csv(
  file.path(out, 'salinity_source_contract.csv'), show_col_types = FALSE
)
grid_reconstruction <- read_csv(
  file.path(out, 'salinity_grid_reconstruction_audit.csv'),
  show_col_types = FALSE
)
legacy_audit <- read_csv(
  file.path(out, 'salinity_legacy_feature_mixing_audit.csv'),
  show_col_types = FALSE
)
legacy_example <- read_csv(
  file.path(out, 'salinity_legacy_18_014_example.csv'),
  show_col_types = FALSE
)
reef_coverage <- read_csv(
  file.path(processed, 'ereefs_surface_salinity_reef_grid_coverage.csv'),
  show_col_types = FALSE
)
bleaching_data <- read_csv(
  file.path(out, 'expanded_bleaching_salinity_data.csv'),
  show_col_types = FALSE
)
annual_data <- read_csv(
  file.path(out, 'expanded_annual_transition_salinity_data.csv'),
  show_col_types = FALSE
)
support <- read_csv(
  file.path(out, 'expanded_salinity_support.csv'), show_col_types = FALSE
)
correlations <- read_csv(
  file.path(out, 'expanded_salinity_correlations.csv'),
  show_col_types = FALSE
)
brt_metrics <- read_csv(
  file.path(out, 'expanded_salinity_brt_metrics.csv'), show_col_types = FALSE
)
brt_event <- read_csv(
  file.path(out, 'expanded_salinity_brt_event_metrics.csv'),
  show_col_types = FALSE
)
brt_deltas <- read_csv(
  file.path(out, 'expanded_salinity_brt_deltas.csv'), show_col_types = FALSE
)
inla_comparison <- read_csv(
  file.path(out, 'expanded_salinity_inla_fit_comparison.csv'),
  show_col_types = FALSE
)
thresholds <- read_csv(
  file.path(out, 'expanded_salinity_inla_threshold_summary.csv'),
  show_col_types = FALSE
)
extreme_rows <- read_csv(
  file.path(out, 'expanded_salinity_extreme_rows.csv'),
  show_col_types = FALSE
)

bleach_h30 <- brt_deltas |>
  filter(
    analysis_context == 'bleaching_mortality',
    validation_scheme == 'leave_one_event_out',
    candidate == 'exposures_h30'
  )
bleach_h26 <- brt_deltas |>
  filter(
    analysis_context == 'bleaching_mortality',
    validation_scheme == 'leave_one_event_out',
    candidate == 'exposures_h26'
  )
bleach_sss <- brt_deltas |>
  filter(
    analysis_context == 'bleaching_mortality',
    validation_scheme == 'leave_one_event_out',
    candidate == 'exposures_sss'
  )
annual_h26 <- brt_deltas |>
  filter(
    analysis_context == 'annual_transition_sensitivity',
    validation_scheme == 'leave_one_event_out',
    candidate == 'exposures_h26'
  )

candidate_labels <- c(
  exposures = 'Heat + COTS + cyclone baseline',
  exposures_sss = 'Baseline + minimum SSS',
  exposures_h30 = 'Baseline + hours <30 PSU',
  exposures_h26 = 'Baseline + hours <26 PSU',
  exposures_sss_h26 = 'Baseline + SSSmin + hours <26',
  exposures_all = 'Baseline + all salinity metrics',
  labels = 'Baseline + retrospective cause labels',
  labels_sss = 'Labels model + minimum SSS',
  labels_h30 = 'Labels model + hours <30 PSU',
  labels_h26 = 'Labels model + hours <26 PSU',
  labels_all = 'Labels model + all salinity metrics'
)
context_labels <- c(
  bleaching_mortality = 'Bleaching-mortality outcome (5 events)',
  annual_transition_sensitivity = 'Annual cover-loss sensitivity (14 years)'
)
metric_labels <- c(
  sss_min = 'Minimum SSS (PSU)',
  log_hours_below30 = 'Hours below 30 PSU',
  log_hours_below26 = 'Hours below 26 PSU'
)
```

## Decision

**Advance low-salinity duration as a formal candidate, but do not add it to the
selected model yet.** The expanded reef-only delivery changes the earlier
conclusion: minimum SSS no longer improves transfer, whereas duration below 30
and 26 PSU produces small, repeatable leave-one-event-out gains.

In the unchanged five-event bleaching-mortality dataset, adding hours below 30
PSU lowers direct-BRT RMSE by
`r number(-bleach_h30$delta_rmse, accuracy = 0.0001)` (reef-bootstrap 95%
interval `r number(-bleach_h30$delta_rmse_q975, accuracy = 0.0001)` to
`r number(-bleach_h30$delta_rmse_q025, accuracy = 0.0001)`). Hours below 26 PSU
lowers RMSE by `r number(-bleach_h26$delta_rmse, accuracy = 0.0001)` (interval
`r number(-bleach_h26$delta_rmse_q975, accuracy = 0.0001)` to
`r number(-bleach_h26$delta_rmse_q025, accuracy = 0.0001)`). Minimum SSS changes
RMSE by `r number(bleach_sss$delta_rmse, accuracy = 0.0001)` and its interval
crosses zero.

This is promising screening evidence, not promotion evidence. Only
`r sum(bleaching_data$hours_below30 > 0)` of `r nrow(bleaching_data)`
reef-event-programme-depth outcomes have non-zero area-mean exposure below 30
PSU, and only `r sum(bleaching_data$hours_below26 > 0)` have exposure below 26
PSU. The apparent bleaching-outcome threshold for <26 PSU is supported by one
reef. The source delivery also omits the start and end dates used to accumulate
hours, which must be recovered before duration can become an operational term.

The selected model in `config/model_registry.yml` is unchanged.

## What changed in the source data

The expanded delivery contains 14 paired reef and GBR4 layers from 2010–11
through 2023–24. Every reef layer has 3,862 unique reef polygons, of which
3,657 have valid salinity values. Reef IDs, grid IDs and the grid missing-value
mask are stable across all seasons. Hours below 26 PSU is always nested within
hours below 30 PSU.

```{r source-summary}
source_audit |>
  transmute(
    Season = season,
    `Valid reefs` = reef_rows - missing_metric_rows,
    `Reefs with hours <30` = positive_hours30_reefs,
    `Reefs with hours <26` = positive_hours26_reefs,
    `Minimum SSS` = number(minimum_sss_psu, accuracy = 0.1),
    `Maximum hours <30` = number(maximum_hours30, accuracy = 1),
    `Maximum hours <26` = number(maximum_hours26, accuracy = 1)
  ) |>
  kable(caption = 'Expanded reef-layer audit by summer')
```

The supplied reef aggregates exactly reproduce area-weighted values from the
4 km grid for the seasonal extremes and known influential reefs: the maximum
absolute reconstruction differences are
`r scientific(max(abs(grid_reconstruction$sss_difference)), digits = 2)` PSU,
`r scientific(max(abs(grid_reconstruction$hours30_difference)), digits = 2)`
hours and
`r scientific(max(abs(grid_reconstruction$hours26_difference)), digits = 2)`
hours.

### Why the earlier 2022–24 result is superseded

The older files at the root of `data/SalinityData` contain 5,362 features, not
3,862 reefs. They include 1,500 islands, rocks, cays, sand or mainland features
and 670 duplicated `LABEL_ID` values. The earlier extractor combined every
feature sharing an ID by area, so some reef exposure values were dominated by
nearby land polygons.

For `18-014`, the old delivery combined the 3.9 km² Hinchinbrook Reef polygon
with a 388 km² Hinchinbrook Island polygon. The new reef-only value is 26.54
PSU and 138 hours below 30; the island record was 18.41 PSU and 2,385 hours.
This explains the largest discrepancies and removes the earlier high-SSS BRT
artefact.

```{r legacy-example}
legacy_example |>
  transmute(
    Delivery = delivery, Feature = feature_type, Name = feature_name,
    `Area (km²)` = number(area_m2 / 1e6, accuracy = 0.1),
    `Minimum SSS` = number(sss_min_psu, accuracy = 0.01),
    `Hours <30` = number(hours_below30, accuracy = 1)
  ) |>
  kable(caption = 'Feature-mixing example for LABEL_ID 18-014')
```

### Is the 4 km cell product needed?

Not for the main model join. The reef-only aggregates pass exact grid
reconstruction and all `r n_distinct(bleaching_data$ReefID)` salinity-matched
bleaching reefs used here have at least 50% valid-grid coverage. The cell data
remain necessary for source QA, calculating valid-area coverage, and extracting
future monitoring locations that lack a reef polygon match.

Five monitored reefs fail the coverage rule and were excluded from the annual
sensitivity: Shute, Dent No. 1 and Hook Island No. 4 have 25.9%, 35.3% and 42.0%
valid area; Magnetic Island No. 2 and Lizard Island North West have none. Four
of these occur in the bleaching-outcome source. Across the full reef layer, 74
reefs have less than 50% valid coverage and 202 have none. Raw maxima from such
polygons should not be interpreted without this coverage screen.

The only unresolved source-contract item is temporal provenance: no machine-
readable start/end dates accompany the summary layers. Consequently, absolute
hours cannot yet be converted to a fraction of the exposure window, and the
2010–11 maximum of more than 5,000 hours cannot be fully audited against its
intended season length.

## Analysis design

Two outcomes are kept separate:

1. **Primary screening outcome:** the existing bleaching-restricted relative
   mortality rows for 2016, 2017, 2020, 2022 and 2024. Rows are averaged to one
   equal-weight ReefID-event-programme-depth outcome. The baseline contains
   survey depth/programme, local-first DHW, COTS hazard and interval cyclone
   exposure.
2. **Annual-transition sensitivity:** consecutive annual coral-cover losses
   for 2011–24, including the very wet 2010–11 summer. It uses raw annual DHW,
   COTS and cyclone context and retains only pre-cover of at least 2%. This is a
   different response and cannot select the production bleaching model.

Direct BRTs use leave-one-event/year-out validation as the primary comparison
and fixed reef-blocked folds as secondary evidence. Preprocessing is learned
inside each fold. Salinity is never imputed. Retrospective disturbance labels
are evaluated separately and are not operational predictors. Full-data
Bernoulli-beta INLA fits provide smooth marginal relationships and supplementary
WAIC/DIC only.

## Multi-event BRT performance

```{r brt-performance}
brt_metrics |>
  filter(
    validation_scheme == 'leave_one_event_out',
    candidate %in% c(
      'exposures', 'exposures_sss', 'exposures_h30', 'exposures_h26',
      'exposures_sss_h26', 'exposures_all'
    )
  ) |>
  mutate(
    Outcome = recode(analysis_context, !!!context_labels),
    Candidate = recode(candidate, !!!candidate_labels),
    RMSE = number(rmse, accuracy = 0.001),
    `Predictive R²` = number(predictive_r2, accuracy = 0.001),
    `Severe RMSE` = number(severe_rmse, accuracy = 0.001)
  ) |>
  select(Outcome, Candidate, RMSE, `Predictive R²`, `Severe RMSE`) |>
  arrange(Outcome, RMSE) |>
  kable(caption = 'Leave-one-event/year-out direct-BRT performance')
```

The duration gains are small but directionally consistent across four of five
bleaching events. Hours below 30 improves 2017, 2020, 2022 and 2024, but worsens
2016 slightly. Hours below 26 is nearly neutral in 2016 and improves the other
four. In contrast, minimum SSS helps 2020 and 2024 but substantially worsens
2017, which explains its failure to transfer.

```{r event-deltas}
baseline_event <- brt_event |>
  filter(
    validation_scheme == 'leave_one_event_out', candidate == 'exposures'
  ) |>
  select(analysis_context, event_year, baseline_rmse = rmse)

brt_event |>
  filter(
    validation_scheme == 'leave_one_event_out',
    analysis_context == 'bleaching_mortality',
    candidate %in% c('exposures_sss', 'exposures_h30', 'exposures_h26')
  ) |>
  left_join(
    baseline_event,
    by = c('analysis_context', 'event_year'), relationship = 'many-to-one'
  ) |>
  transmute(
    Event = event_year,
    Metric = recode(candidate, !!!candidate_labels),
    `RMSE change` = number(rmse - baseline_rmse, accuracy = 0.0001)
  ) |>
  pivot_wider(names_from = Metric, values_from = `RMSE change`) |>
  kable(caption = 'Event-held-out RMSE change; negative values improve')
```

Reef-blocked interpolation tells a different story: minimum SSS improves the
bleaching outcome by 0.0053 RMSE, while durations are nearly neutral. This is
exactly why the earlier 2024-only SSS result was not sufficient. SSS carries
within-event spatial information, but duration is the metric with better event
transfer.

When retrospective bleaching, flood, cyclone and COTS labels are added, the
bleaching-outcome gain shrinks to 0.0005 RMSE for hours below 30 and reverses
slightly for hours below 26; both intervals cross zero. This indicates that
part of the duration signal overlaps recorded disturbance causes. The labels
are incomplete, sometimes depth-specific text collapsed to a reef event, and
unavailable prospectively, so this does not invalidate the exposure signal—but
it sharply limits a causal freshwater claim.

Adding SSS and duration together does not improve on duration alone. Hours
below 30 and below 26 are nested and have Spearman correlation
`r number(correlations |> filter(analysis_context == 'bleaching_mortality') |> pull(spearman_hours30_hours26), accuracy = 0.01)`
in the bleaching data. They should remain alternative candidates, not be fit
as interchangeable terms in the same operational model.

## What does 2010–11 add?

The annual-transition sensitivity contains `r nrow(annual_data)` outcomes from
`r n_distinct(annual_data$ReefID)` reefs across all 14 years. Adding hours below
26 lowers pooled leave-one-year-out RMSE by
`r number(-annual_h26$delta_rmse, accuracy = 0.0001)`, but its bootstrap
interval crosses zero (`r number(-annual_h26$delta_rmse_q975, accuracy = 0.0001)`
to `r number(-annual_h26$delta_rmse_q025, accuracy = 0.0001)`). The pooled
evidence is therefore suggestive, not conclusive.

It is nevertheless informative for 2011: the <26 candidate lowers held-out
2011 RMSE from 0.338 to 0.298. The event includes 43 programme-depth outcomes,
with mean relative loss of 33.0% at MMP 2 m and 24.3% at MMP 5 m.

```{r support-2011}
support |>
  filter(
    analysis_context == 'annual_transition_sensitivity', event_year == 2011
  ) |>
  transmute(
    Programme = toupper(programme_key), Depth = depth,
    Rows = rows, Reefs = reefs,
    `Mean relative loss` = percent(mean_mortality, accuracy = 0.1),
    `Flood-labelled rows` = flood_rows,
    `Minimum SSS` = number(minimum_sss, accuracy = 0.1),
    `Maximum hours <26` = number(maximum_hours_below26, accuracy = 1)
  ) |>
  kable(caption = '2011 annual-transition support')
```

The clearest depth pattern is biological plausible but not cleanly causal:
Pelican Island Reef had 1,385 hours below 26 PSU and 98.5% versus 28.8% loss at
2 m and 5 m; Humpy Reef had 626 hours and 86.1% versus 14.8% loss. Stingaree and
Tobias were also affected by Cyclone Yasi, and other reefs show weak or reversed
depth differences. The 2011 result supports retaining depth-specific duration,
but it does not identify a stable duration-by-depth coefficient by itself.

## Smoothed INLA relationships

All three metrics improve full-data INLA WAIC relative to the exposure-only
baseline. In the bleaching outcome, delta WAIC is -14.0 for minimum SSS, -5.9
for hours below 30 and -10.0 for hours below 26. In the annual sensitivity the
corresponding values are -44.9, -51.5 and -55.6. These fit statistics confirm a
smooth association but cannot override held-out transfer.

```{r inla-comparison}
inla_comparison |>
  mutate(
    Outcome = recode(analysis_context, !!!context_labels),
    Metric = recode(candidate,
      baseline = 'Exposure-only baseline', !!!metric_labels
    ),
    `Combined WAIC` = number(combined_waic, accuracy = 0.1),
    `Delta WAIC` = number(delta_waic_from_baseline, accuracy = 0.1)
  ) |>
  select(Outcome, Metric, `Combined WAIC`, `Delta WAIC`) |>
  arrange(Outcome, `Combined WAIC`) |>
  kable(caption = 'Full-data explanatory INLA fit')
```

```{r combined-curves, fig.alt='BRT and INLA partial dependence curves for minimum SSS and duration below 30 and 26 PSU, separated by outcome and survey depth.'}
knitr::include_graphics(file.path(
  root, 'output', 'surface_salinity_mortality',
  'expanded_salinity_brt_inla_partial_dependence.png'
))
```

Both frameworks now show the expected direction: mortality rises with longer
low-salinity duration and toward lower minimum SSS. The previous high-SSS
upturn disappears after removing mixed land features and adding multiple
events. Curves for MMP 2 m generally sit above MMP 5 m, but programme/depth
intercepts and surface exposure do not prove depth attenuation.

## Where does mortality increase?

Posterior contrasts compare each duration value with zero exposure and each SSS
value with the highest observed SSS. The table reports the first grid value
whose lower 95% posterior contrast exceeds zero. These are model-dependent
screening crossings, not physiological thresholds.

```{r threshold-table}
thresholds |>
  filter(stratum %in% c('MMP 2 m', 'MMP 5 m')) |>
  mutate(
    Outcome = recode(analysis_context, !!!context_labels),
    Metric = recode(metric, !!!metric_labels),
    `Programme/depth` = stratum,
    Crossing = if_else(
      is.na(threshold_original_value), 'None detected',
      number(threshold_original_value, accuracy = 0.1)
    ),
    `Estimated uplift` = if_else(
      is.na(difference_at_threshold), '—',
      percent(difference_at_threshold, accuracy = 0.1)
    ),
    `Supporting reefs` = empirical_reefs_beyond_threshold
  ) |>
  select(
    Outcome, Metric, `Programme/depth`, Crossing,
    `Estimated uplift`, `Supporting reefs`
  ) |>
  kable(caption = 'Posterior screening crossings and empirical support')
```

The annual sensitivity yields much better support: about 180 hours below 30
PSU (14 reefs), 38 hours below 26 PSU (9 reefs), or minimum SSS below 28.1 PSU
(18 reefs). In the bleaching-restricted data, the analogous crossings are
about 212 hours below 30 or SSS below 24.9 PSU, each supported by only three
reefs. The <26 crossing is the maximum observed 189 hours at Snapper Reef and
is supported by that reef alone. It must not be encoded as a threshold.

Stingaree remains the essential counterexample: in 2024 it had 849 hours below
30 and 23 hours below 26, but only 4.2% and 5.7% MMP loss at 2 m and 5 m. Humpy
Reef in 2017 had 203 hours below 30 and 38 below 26 with 0% and 7.7% loss. These
observations argue for a graded duration response with uncertainty, not a hard
30 or 26 PSU cutoff.

## Critical next steps

1. **Test one duration metric at a time in the selected cause-aware INLA
   architecture.** Hours below 30 should be the primary formal candidate
   because three bleaching reefs support its elevated range and its event-held-
   out gain is slightly larger. Hours below 26 should be the mechanism-focused
   replacement sensitivity, not an additional collinear term.
2. **Do not promote from direct-BRT gains.** The improvement is only 0.002–0.003
   RMSE and comes from 10–18 exposed outcomes. Promotion requires the same
   fixed leave-one-event-out comparison in the registered operational model,
   including severe-event RMSE and false-extreme rate.
3. **Recover the exact accumulation windows and code.** Start/end dates,
   timestep handling, missing-hour treatment and whether durations are annual
   or wet-season totals are prerequisites for interpreting absolute hours.
4. **Retain the 2011 annual analysis as external mechanism evidence.** It
   demonstrates that the layer can recover a major freshwater year and a
   plausible shallow-depth pattern, but its annual cover-loss response is not
   interchangeable with bleaching mortality.
5. **Do not fit a hard threshold or a depth interaction yet.** The bleaching
   threshold tail contains one to three reefs and several cyclone/bleaching
   co-labels. More independent flood reefs are required before estimating a
   stable surface-exposure-by-depth interaction.
6. **Keep the 4 km cells as an audit and fallback product.** Reef aggregates
   are adequate after the valid-area filter. Cell extraction is warranted for
   new monitoring sites, reef IDs without a polygon match, or sensitivity to
   within-reef exposure heterogeneity.

## Promotion boundary

This investigation does not change `config/model_registry.yml`. A duration
metric can be promoted only if it improves the selected cause-aware INLA model
under fixed leave-one-event-out validation, retains or improves severe-event
RMSE and false-extreme rate, has documented temporal provenance, and does not
create excessive mortality at counterexamples such as Stingaree or Humpy Reef.
