---
title: "Freshwater, cyclone and upwelling predictors for GBR mortality"
subtitle: "AIMS extraction, 2024 context and a practical GBR-wide design"
date: today
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
    code-summary: "Show analysis code"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)

theme_set(theme_minimal(base_size = 12))

aims <- read_csv(
  '../data/processed/aims_freshwater_training_points.csv',
  show_col_types = FALSE
)
discrete <- read_csv(
  '../data/processed/aims_wq_discrete_samples_2015_2025.csv',
  show_col_types = FALSE
)
projects <- read_csv(
  '../data/processed/aims_wq_project_coverage.csv',
  show_col_types = FALSE
)
loggers <- read_csv(
  '../data/processed/aims_wq_logger_hourly_coverage.csv',
  show_col_types = FALSE
)
logger_events <- read_csv(
  '../data/processed/aims_wq_logger_hourly_event_summary.csv',
  show_col_types = FALSE
)
occurrence_metrics <- read_csv(
  '../output/aims_freshwater_proxy/occurrence_metrics.csv',
  show_col_types = FALSE
)
magnitude_metrics <- read_csv(
  '../output/aims_freshwater_proxy/magnitude_metrics.csv',
  show_col_types = FALSE
)
cyclones <- read_csv(
  '../data/processed/bom_cyclone_reef_year.csv',
  show_col_types = FALSE
)
tracks <- read_csv(
  '../data/processed/bom_cyclone_tracks_gbr.csv',
  show_col_types = FALSE
)
reef_grid <- read_csv(
  '../data/processed/freshwater_predictor_grid_2016_2025.csv',
  show_col_types = FALSE
)
```

## Decision summary

The AIMS data are useful, but they should be a **calibration and validation
layer**, not the only freshwater predictor. The portal provides direct salinity,
suspended solids, CDOM, Secchi depth and chlorophyll observations at monitored
stations. These observations are too sparse to interpolate blindly across the
GBR, and many low-salinity stations are near river mouths rather than on reefs.

The practical GBR-wide freshwater model is therefore a two-part model:

1. predict the probability of ecologically important freshwater exposure
   (minimum salinity below 30 PSU); and
2. conditional on exposure, predict its magnitude or freshwater deficit.

The predictors should combine event-timed rainfall, IMOS optical plume
information, cyclone exposure and spatial connectivity to river sources. The
historical eReefs salinity field remains useful as a broad physical prior, but
the AIMS measurements should calibrate its bias. The final reef prediction must
carry uncertainty that increases with distance from AIMS observations.

The first reef-blocked screen is a useful negative result. Rainfall, IMOS
Kd490, RRN water colour and cyclone proximity only modestly improve detection
of salinity below 30 PSU, and none of the candidates estimates freshwater
severity accurately enough for production use.

The cyclone track layer is immediately useful and resolves an exposure signal
that the existing RRN 4 m wave-hours variable did not consistently recover. It
should remain separate from freshwater and wave damage: track proximity is a
forcing indicator, not proof of either mechanism.

For upwelling, a single static shelf category is not adequate. We need an
event-specific cooling/refugia layer derived from high-resolution daily SST,
with a second subsurface intrusion component. This is an ameliorating modifier
of DHW, so its main model role is an interaction with heat stress.

## What the 2024 manuscript changes

The attached 2024 impact manuscript reports surveys of 124 reefs and record
single-year coral-cover declines in the Northern and Southern GBR. Acropora
accounted for much of both the preceding gains and subsequent losses. The most
important modelling result is the contrast around Lizard Island: outer-shelf
reefs with approximately 7.3--7.6 DHW had negligible mortality, while inshore
reefs with lower heat exposure experienced severe mortality. The outer-shelf
reefs experienced a local 1--1.5 degrees C cooling during the heatwave,
consistent with hydrodynamic relief. Jasper, flooding and low salinity were
also implicated in the northern losses.

This supports four changes already being developed in the mortality pipeline:

- retain pre-event coral composition, especially Acropora;
- represent freshwater as an event-specific DHW modifier;
- add cyclone forcing without conflating wave damage and flood exposure; and
- add upwelling/cooling relief as a DHW interaction.

The manuscript also supports using aerial bleaching severity as a within-event
nowcast gate. It should improve post-event forecasts, but cannot be used for a
fully pre-season prediction.

## AIMS data recovered

The extraction uses the public [AIMS water-quality portal](https://data.aims.gov.au/water-quality/)
and is reproducible with `scripts/extract_aims_water_quality.R`. Inspection of
the portal application recovered its all-programme/MMP-only,
depth-average/depth-weighted and hourly/daily query variants. The primary
calibration table uses all programmes, depth averages and observations from
0--10 m; the alternative summaries remain available for sensitivity checks.

```{r}
aims_event <- aims |>
  group_by(event_year) |>
  summarise(
    station_events = n(),
    locations = n_distinct(LOCATION_NAME),
    with_salinity = sum(is.finite(salinity_min_psu)),
    below_30 = sum(salinity_min_psu < 30, na.rm = TRUE),
    below_25 = sum(salinity_min_psu < 25, na.rm = TRUE),
    within_20km_reef = sum(reef_proximal_20km),
    .groups = 'drop'
  )
kable(aims_event, digits = 0)
```

The all-programme depth-averaged download contains `r comma(nrow(discrete))`
sample-depth rows. After retaining ecologically relevant shallow observations
(0--10 m), there are `r comma(nrow(aims))` station-event summaries;
`r sum(aims$reef_proximal_5km)` are within 5 km and
`r sum(aims$reef_proximal_20km)` are within 20 km of a GBR reef registry point.
The processed archive contains `r sum(aims$sample_dates, na.rm = TRUE)` station
sampling dates after shallow samples are aggregated.

```{r}
projects |>
  group_by(PROJECT) |>
  summarise(
    rows = sum(rows), locations = sum(locations),
    salinity_rows = sum(salinity_rows),
    first_event = min(event_year), last_event = max(event_year),
    .groups = 'drop'
  ) |>
  arrange(desc(salinity_rows)) |>
  kable(caption = 'All-programme depth-average coverage by project')
```

The expanded query adds MBM and LTMP-DIC reef observations that are absent
from an MMP-only export. A server quirk also matters for reproducibility:
`mmpOnly` is not encoded in the generated archive filename, so the extraction
uses a distinct bounding-box string for the all-programme request to prevent a
cached MMP-only archive masking the broader result.

```{r}
#| fig-width: 9
#| fig-height: 5.5
aims_2024 <- aims |> filter(event_year == 2024, is.finite(salinity_min_psu))

ggplot() +
  geom_point(
    data = reef_grid |> distinct(ReefID, .keep_all = TRUE),
    aes(reef_longitude, reef_latitude),
    colour = 'grey80', size = 0.25
  ) +
  geom_point(
    data = aims_2024,
    aes(LONGITUDE, LATITUDE, colour = salinity_min_psu),
    size = 2.2, alpha = 0.9
  ) +
  scale_colour_gradientn(
    colours = c('#7f0000', '#d7301f', '#fdae61', '#2c7fb8'),
    limits = c(0, 36), oob = squish,
    name = 'Minimum\nsalinity (PSU)'
  ) +
  coord_equal(xlim = c(142.5, 154), ylim = c(-25, -10)) +
  labs(
    title = 'AIMS shallow water-quality observations in the 2024 wet season',
    subtitle = 'Samples are from 0--10 m; grey points are GBR registry features',
    x = 'Longitude', y = 'Latitude'
  )
```

### Important limitation: the continuous logger gap

```{r}
#| fig-width: 9
#| fig-height: 4.5
loggers |>
  mutate(
    first_date = as.Date(first_time),
    last_date = as.Date(last_time),
    site = reorder(site, site_latitude)
  ) |>
  ggplot(aes(y = site)) +
  geom_segment(
    aes(x = first_date, xend = last_date, yend = site),
    linewidth = 2.5, colour = '#2166ac'
  ) +
  geom_vline(xintercept = as.Date('2024-01-01'), linetype = 2) +
  labs(
    title = 'AIMS continuous logger coverage available from the portal',
    subtitle = 'The public daily logger files end by October 2023; they do not observe Jasper in 2024',
    x = NULL, y = 'Logger site'
  )
```

This is not a join error: the latest timestamp in the downloaded hourly logger
files is `r max(as.POSIXct(loggers$last_time, tz = 'UTC'))`. Hourly data matter:
`r sum(logger_events$salinity_hours_below30 > 0, na.rm = TRUE)` historical
site-events contain at least one hour below 30 PSU, and short pulses can be
hidden by daily averages. The discrete record does include 2024, but it is
intermittent. It also includes LTMP-DIC, MBM and legacy projects when the API's
`mmpOnly=false` option is used; this is why it recovers reef samples that an
MMP-only export misses.

```{r}
target_wq <- aims |>
  filter(
    event_year == 2024,
    grepl('Mackay|Linnet|Eyrie|Swinger|Lizard', LOCATION_NAME,
          ignore.case = TRUE),
    is.finite(salinity_min_psu)
  ) |>
  transmute(
    Project = PROJECT, Location = LOCATION_NAME,
    `Matched reef` = ReefName,
    `Depth range (m)` = paste0(sample_depth_min_m, '--', sample_depth_max_m),
    `Minimum salinity (PSU)` = round(salinity_min_psu, 2),
    `Sampling dates` = sample_dates,
    `Match method` = reef_match_method,
    `Distance to reef (km)` = round(nearest_reef_distance_km, 1)
  ) |>
  arrange(Location)
kable(target_wq, caption = 'Recovered 2024 observations at target northern reefs')
```

These measurements demonstrate that the apparent absence at the target reefs
was partly an API-filter and depth-filter problem. They do not rule out an
earlier Jasper freshwater pulse: most are single visits, and a later value near
34 PSU cannot establish the event minimum. Mackay's measurement was at 5 m in
late February, roughly two months after Jasper's December flood pulse.

```{r}
#| fig-width: 9
#| fig-height: 5
logger_events |>
  filter(is.finite(salinity_min_psu)) |>
  ggplot(aes(event_year, salinity_min_psu, colour = site)) +
  geom_hline(yintercept = 30, linetype = 2, colour = 'grey35') +
  geom_point(size = 2, alpha = 0.8) +
  scale_x_continuous(breaks = sort(unique(logger_events$event_year))) +
  labs(
    title = 'Hourly loggers expose freshwater pulses hidden by seasonal means',
    x = 'Event year', y = 'Minimum hourly salinity (PSU)', colour = 'Site'
  )
```

## Freshwater proxy pilot

The first compact hurdle-model screen is complete. It uses five whole-reef
folds and keeps 2024 in the validation pool. The occurrence target has
`r comma(max(occurrence_metrics$rows))` complete shallow station-events and
`r comma(max(occurrence_metrics$flood30_rows))` events below 30 PSU.

```{r}
occurrence_metrics |>
  filter(evaluation == 'All reef-blocked rows') |>
  ggplot(aes(reorder(model, average_precision), average_precision)) +
  geom_col(fill = '#2166ac') +
  geom_hline(
    yintercept = unique(occurrence_metrics$prevalence[
      occurrence_metrics$evaluation == 'All reef-blocked rows'
    ]), linetype = 2
  ) +
  coord_flip() +
  labs(
    title = 'Reef-blocked detection of salinity below 30 PSU',
    subtitle = 'Dashed line is event prevalence; higher average precision is better',
    x = NULL, y = 'Average precision'
  )
```

The hybrid has the best overall average precision
(`r round(max(occurrence_metrics$average_precision[occurrence_metrics$evaluation == 'All reef-blocked rows']), 3)`),
but the gain over eReefs and operational proxies is modest. More importantly,
none of the candidates estimates low-salinity magnitude adequately: conditional
RMSE is `r round(min(magnitude_metrics$conditional_rmse_psu[magnitude_metrics$evaluation == 'All reef-blocked rows']), 1)`
PSU and predictions are biased about
`r round(min(magnitude_metrics$conditional_bias_psu[magnitude_metrics$evaluation == 'All reef-blocked rows']), 1)`
PSU too high. The pilot can rank some exposure risk, but it is not yet suitable
for supplying mortality-model salinity magnitudes. Event timing and
river-to-reef connectivity are the next necessary additions.

## Cyclone best tracks

The extraction uses the Bureau of Meteorology's Australian tropical cyclone
best-track archive. The Bureau describes these as post-event reanalysed tracks;
the source includes time, position, pressure, maximum wind and quadrant wind
radii. `scripts/extract_bom_cyclone_exposure.R` creates a complete reef-by-event
year table rather than matching only reefs already present in the mortality data.

```{r}
target_names <- tibble::tribble(
  ~ReefID, ~display_name,
  '15-024', 'Mackay Reefs (15-024)',
  '16-015', 'Mackay Reef (16-015)',
  '14-126', 'Linnet Reef (14-126)',
  '14-116a', 'Lizard Island reef (14-116a)',
  '14-116b', 'Lizard Island reef (14-116b)',
  '14-116d', 'Lizard Island reef (14-116d)',
  '14-118', 'Eyrie Reef (14-118)',
  '15-030', 'Swinger Reef (15-030)'
)
target_cyclone <- cyclones |>
  filter(event_year == 2024) |>
  inner_join(target_names, by = 'ReefID') |>
  transmute(
    Reef = display_name,
    `Nearest cyclone` = tc_nearest_name,
    `Minimum track distance (km)` = round(tc_min_distance_km, 1),
    `Wind at nearest fix (m/s)` = tc_nearest_max_wind_ms,
    `Inside reported gale radius` = tc_inside_reported_gale_radius
  ) |>
  arrange(`Minimum track distance (km)`)
kable(target_cyclone)
```

```{r}
#| fig-width: 9
#| fig-height: 6
tracks_2024 <- tracks |>
  filter(event_year == 2024, NAME %in% c('Jasper', 'Kirrily'))
reefs_2024 <- cyclones |> filter(event_year == 2024)

ggplot() +
  geom_point(
    data = reefs_2024,
    aes(reef_longitude, reef_latitude,
        colour = pmin(tc_min_distance_km, 300)),
    size = 0.45
  ) +
  geom_path(
    data = tracks_2024,
    aes(LON, LAT, group = DISTURBANCE_ID),
    linewidth = 1, colour = 'black'
  ) +
  geom_point(
    data = target_names |>
      inner_join(reefs_2024, by = 'ReefID'),
    aes(reef_longitude, reef_latitude),
    shape = 21, fill = '#ffff33', colour = 'black', size = 2.5
  ) +
  scale_colour_gradient(
    low = '#b2182b', high = '#2166ac',
    name = 'Nearest track\ndistance (km)'
  ) +
  coord_equal(xlim = c(142.5, 154), ylim = c(-25, -10)) +
  labs(
    title = '2024 cyclone tracks and GBR reef proximity',
    subtitle = 'Highlighted points are the severe northern reefs under review',
    x = 'Longitude', y = 'Latitude'
  )
```

Mackay (16-015), Mackay (15-024) and Swinger have strong Jasper proximity,
and the first two are inside or close to the reported gale footprint depending
on the quadrant at the nearest fix. Linnet and the Lizard/Eyrie group are still
within roughly 120--132 km of Jasper's track. This supports a regional Jasper
forcing variable, but it does not by itself distinguish rainfall/flood plume
from wind or wave damage.

## How to generalise freshwater exposure to all GBR reefs

The new file `data/processed/freshwater_predictor_grid_2016_2025.csv` contains
the candidate predictors at all `r n_distinct(reef_grid$ReefID)` GBR registry
features for each AIMS event year. Complete ERA5/eReefs fields currently exist
for 2016, 2017, 2020, 2022 and 2024; the RRN and cyclone layers cover a longer
period.

The modelling sequence should stay simple and auditable:

1. **Measured target.** Use AIMS minimum salinity and the below-30-PSU flag.
   Retain sample count and distance to reef. Give river-mouth measurements less
   leverage on reef-level magnitude, while retaining them for plume detection.
2. **Broad physical prior.** Use eReefs minimum salinity where the forcing is
   complete, but calibrate it to AIMS rather than assuming it is truth.
3. **Operational proxy model.** Predict exposure from event rainfall, IMOS
   Kd490/coloured-water anomalies, BoM cyclone forcing and a river-plume
   connectivity term. RRN coloured-water frequency is a valuable retrospective
   predictor but must be marked unavailable for real-time use when its release
   is delayed.
4. **Validation.** Keep 2024 in model development. Use held-out spatial groups
   or catchments so the model is tested at unseen locations while retaining the
   only well-observed severe compound event. Report sensitivity with and without
   2024, not a single 2024 holdout score.
5. **Mortality model.** Enter the predicted freshwater distribution, not just a
   plug-in mean, and interact it with DHW. This lets freshwater raise mortality
   under heat stress without inventing widespread thermal mortality at low DHW.

The one important missing structural predictor is hydrological connectivity:
distance from each reef to river mouths is not enough because plume transport is
directional. A compact first version can use nearest major river, coastal
rainfall/discharge, distance and prevailing current direction. If that performs
poorly, replace it with an eReefs-derived source-to-reef connectivity matrix.

## Upwelling and thermal-refuge layer

High-resolution model and satellite work has already identified GBR climate
refugia caused by cooler-water upwelling through interactions among tides,
currents and reef structure. A separate Central GBR study found 64 intrusive
upwelling events in six years: near-bottom temperature commonly fell 1--3
degrees C over about a week, and the cooling was not necessarily visible at the
surface. This is why neither shelf position nor SST alone is sufficient.

The first operational version should contain two interpretable components:

| Component | Calculation | Role |
|---|---|---|
| Surface cooling relief | Daily IMOS 0.02-degree L3S SST at each reef minus the same-day median in its surrounding shelf/latitude neighbourhood; retain minimum anomaly, days below -0.5/-1 degree C and longest cool spell during peak DHW | Detect the 2024 outer-Lizard cooling signal at prediction scale |
| Subsurface intrusion forcing | Daily vertical velocity, temperature at several depths, vertical temperature gradient and currents from Copernicus Marine; calibrate against Palm Passage/Myrmidon and any suitable AIMS/IMOS temperature loggers | Recover bottom intrusions missed by surface SST |

The surface record is available operationally and at useful reef-scale
resolution. Copernicus is coarser (about 0.083 degrees), so it should represent
large-scale forcing rather than claim reef-scale truth. A static historical
refugia probability can be estimated from repeat cool anomalies and used when
daily data are absent.

In the mortality model, compare:

```text
thermal_effect = f(DHW, composition, prior exposure)
                 + DHW x freshwater_exposure
                 + DHW x cooling_relief
```

The signs should be constrained ecologically: freshwater can exacerbate heat
loss and cooling relief can reduce it. A flexible residual term can still learn
exceptions, but these interactions prevent the low-DHW background from
flattening the severe-loss response.

## Immediate implementation order

1. Add river-source/connectivity fields to the new GBR predictor grid.
2. Add river-source connectivity and event-timed rainfall/optical anomalies to
   the completed below-30-PSU pilot, then repeat whole-reef validation and the
   no-2024 sensitivity analysis.
3. Extract the 2024 IMOS SST cooling-residual pilot around Lizard, then extend
   it to all event years if it reproduces the manuscript contrast.
4. Add Copernicus vertical-velocity/thermal-gradient forcing and validate it at
   the Central GBR moorings.
5. Pass freshwater and cooling posterior/ensemble predictions into the
   beta-binomial, BRMS and BRT mortality ensemble as DHW interactions.

The habitat map is intentionally excluded from this sequence. It does not solve
the dynamic freshwater or upwelling problem and would add complexity without a
clear predictive target.

## Sources

- [AIMS water-quality portal](https://data.aims.gov.au/water-quality/)
- [Bureau of Meteorology tropical cyclone databases](https://www.bom.gov.au/cyclone/tropical-cyclone-knowledge-centre/databases/)
- [IMOS satellite SST products](https://imos.org.au/srs-sst-data)
- [Copernicus Global Ocean Physics Analysis and Forecast](https://data.marine.copernicus.eu/product/GLOBAL_ANALYSISFORECAST_PHY_001_024/description)
- [Sun et al. (2024), climate refugia in the GBR](https://pmc.ncbi.nlm.nih.gov/articles/PMC11606497/)
- [Benthuysen et al. (2016), intrusive upwelling in the Central GBR](https://doi.org/10.1002/2016JC012294)
