---
title: "GBR coral mortality: formal BRMS-BRT prediction models"
subtitle: "Blocked validation, cross-fitted stacking, and all-event production refits"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(readr)
library(scales)
library(knitr)

formal_dir <- "../output/formal_models"
files <- c(
  comparison_metrics = file.path(formal_dir, "model_comparison_metrics.csv"),
  comparison_predictions = file.path(formal_dir, "model_comparison_predictions.csv"),
  production_weights = file.path(formal_dir, "production_ensemble_weights.csv"),
  fold_weights = file.path(formal_dir, "ensemble_fold_weights.csv"),
  brms_diagnostics = file.path(formal_dir, "brms_diagnostics.csv"),
  brms_production_diagnostics = file.path(formal_dir, "brms_production_diagnostics.csv"),
  brt_tuning = file.path(formal_dir, "brt_tuning.csv"),
  brt_influence = file.path(formal_dir, "brt_production_influence.csv")
)
files <- c(
  files,
  brms_fit_statistics = file.path(formal_dir, 'brms_fit_statistics.csv'),
  explanatory_metrics = '../output/model_validation/candidate_explanatory_metrics.csv',
  acropora_coverage = '../data/processed/pre_event_acropora_coverage.csv'
)
files <- c(
  files,
  event_context = file.path(formal_dir, 'event_context_predictor_audit.csv'),
  event_calibration = file.path(formal_dir, 'event_component_calibration.csv')
)
if (any(!file.exists(files))) {
  stop("Formal model results are incomplete. Run the BRMS, BRT, and ensemble scripts first.")
}

metrics <- read_csv(files[["comparison_metrics"]], show_col_types = FALSE)
predictions <- read_csv(files[["comparison_predictions"]], show_col_types = FALSE)
production_weights <- read_csv(files[["production_weights"]], show_col_types = FALSE)
fold_weights <- read_csv(files[["fold_weights"]], show_col_types = FALSE)
brms_diagnostics <- read_csv(files[["brms_diagnostics"]], show_col_types = FALSE)
brms_production_diagnostics <- read_csv(
  files[["brms_production_diagnostics"]], show_col_types = FALSE
)
brt_tuning <- read_csv(files[["brt_tuning"]], show_col_types = FALSE)
brt_influence <- read_csv(files[["brt_influence"]], show_col_types = FALSE)

brms_fit_statistics <- read_csv(
  files[['brms_fit_statistics']], show_col_types = FALSE
)
explanatory_metrics <- read_csv(
  files[['explanatory_metrics']], show_col_types = FALSE
)
acropora_coverage <- read_csv(
  files[['acropora_coverage']], show_col_types = FALSE
)
event_context <- read_csv(files[['event_context']], show_col_types = FALSE)
event_calibration <- read_csv(
  files[['event_calibration']], show_col_types = FALSE
)

programme_labels <- c(
  manta = "LTMP manta tow",
  ltmp = "LTMP benthic",
  mmp = "MMP inshore"
)
scheme_labels <- c(
  leave_one_event_out = "Leave one event out",
  reef_blocked_5fold = "Reef blocked",
  region_blocked = "Region blocked"
)
learner_labels <- c(
  benchmark = "Benchmark",
  brms = "BRMS",
  brt = "BRT",
  formal_ensemble = "BRMS-BRT ensemble",
  stacked_ensemble = "Benchmark-BRMS-BRT stack"
)

theme_set(theme_bw(base_size = 11))
```

# What was fitted

The response tables retain the established bleaching-attribution restriction:
disturbance codes `b`, `m`, `n`, or an unrecorded disturbance are eligible,
while records explicitly assigned to COTS, storms/cyclones, disease, or other
excluded disturbances are not used. The meaning of `m` is unchanged, and blank
disturbance records remain eligible when mortality can plausibly be attributed
to the bleaching event. Genuine zero-loss observations are retained as controls.

The benchmark models remain interpretable references. They do not limit the
production learners. The formal BRMS model is boundary-aware and hierarchical,
with a nonlinear DHW response, all attainable environmental predictors,
preceding-year Acropora proportion and absolute Acropora cover, a DHW--Acropora
interaction, and prespecified DHW interactions with cloud, Secchi depth,
current speed, prior severe heat exposure, ten-year cumulative heat load,
signed thermal novelty, and an acute IMOS optical-extreme metric. Reef, event,
and region effects capture repeated observations in the positive beta mean.
Event and region effects are now also fitted in the zero-boundary component,
where the earlier models most strongly underpredicted 2024 occurrence.

The BRT uses the same predictors in a two-part occurrence/magnitude model.
Its positive component compares Gaussian mortality against a logit-Gaussian
bounded response. Response scale, tree depth, learning rate, minimum node size,
and tree number are selected only within each outer analysis fold.

All observations receive one out-of-fold prediction per validation scheme.
Ensemble weights for a held-out fold are learned only from the other folds. Two
ensembles are retained: a BRMS-BRT framework blend and a prediction-focused
stack that may also use the low-variance benchmark. Final production weights use
the event- and reef-blocked predictions; region blocking remains a geographic
stress test. The final models are refitted on every modern event, including 2024.

Two prediction targets are kept explicit. Leave-one-event-out predictions are
future-event forecasts and exclude event and region random effects. Reef-blocked
predictions represent mapping an observed event to unsurveyed reefs and include
the known event and region effects. Region-blocked predictions include the known
event effect but exclude the held-out region effect. GBR-wide reconstruction
should therefore be judged primarily from reef-blocked validation, while
leave-one-event-out remains the stricter test of forecasting a new event.

## Pre-event Acropora reconstruction

Acropora composition is aligned to the survey used as the pre-bleaching coral
cover baseline. Exact reef, depth, and baseline-year observations are preferred.
Where these are unavailable, only past observations are used: first the same
reef, then a past-only spatial-temporal interpolation. No future community
composition is allowed to enter a validation row.

```{r acropora-coverage}
acropora_coverage |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Source = recode(
      acropora_source,
      observed_depth_year = 'Observed reef-depth-year',
      past_same_reef = 'Earlier observation at same reef',
      past_only_spatiotemporal_idw = 'Past-only spatial-temporal estimate'
    ),
    Coverage = percent(proportion, accuracy = 0.1)
  ) |>
  select(Programme, Source, Rows = rows, Coverage) |>
  arrange(Programme, desc(Rows)) |>
  kable()
```

All 537 validation observations now have a preceding-year Acropora estimate.
The source year is checked to be no later than the baseline year. LTMP benthic
is fully observed, MMP is 97.8% observed, and manta tow is 69.8% observed; the
remaining manta estimates therefore deserve a source-specific sensitivity
analysis before ecological effect sizes are treated as definitive.

# Blocked predictive performance

```{r performance-plot, fig.width=11, fig.height=7}
plot_metrics <- metrics |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Assessment = recode(scheme, !!!scheme_labels),
    Learner = recode(learner, !!!learner_labels)
  )

ggplot(plot_metrics, aes(x = Learner, y = rmse, colour = Learner)) +
  geom_point(size = 2.8) +
  geom_line(aes(group = Assessment), colour = "grey70", linewidth = 0.4) +
  facet_grid(Programme ~ Assessment, scales = "free_y") +
  scale_colour_brewer(palette = "Dark2", guide = "none") +
  labs(x = NULL, y = "Out-of-fold RMSE", title = "No learner is judged from in-sample fit") +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

```{r predictive-r2-plot, fig.width=11, fig.height=7}
ggplot(plot_metrics, aes(x = Learner, y = predictive_r2, colour = Learner)) +
  geom_hline(yintercept = 0, colour = 'grey55', linetype = 2) +
  geom_point(size = 2.8) +
  geom_line(aes(group = Assessment), colour = 'grey70', linewidth = 0.4) +
  facet_grid(Programme ~ Assessment, scales = 'free_y') +
  scale_colour_brewer(palette = 'Dark2', guide = 'none') +
  labs(
    x = NULL, y = expression(R[predictive]^2),
    title = 'Variance explained in blocked assessment data'
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

```{r performance-table}
metrics |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Assessment = recode(scheme, !!!scheme_labels),
    Learner = recode(learner, !!!learner_labels)
  ) |>
  select(
    Programme, Assessment, Learner, n,
    MAE = mae, RMSE = rmse, `Predictive R2` = predictive_r2,
    `Balanced MAE` = balanced_mae, `Severe-event MAE` = severe_mae,
    `Occurrence Brier` = occurrence_brier, Bias = bias,
    `Observed mean` = observed_mean, `Predicted mean` = predicted_mean
  ) |>
  arrange(Programme, Assessment, RMSE) |>
  kable(digits = 3)
```

## Current predictive choice

The table below pools the two prespecified selection tasks: transfer to another
event and transfer to unsurveyed reefs. It is an honest routing decision, not a
claim that the winning learner is the most ecologically informative model.

```{r current-predictive-choice}
selection_scores <- predictions |>
  filter(scheme %in% c("leave_one_event_out", "reef_blocked_5fold")) |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Learner = recode(learner, !!!learner_labels)
  ) |>
  group_by(Programme, Learner) |>
  summarise(
    n = n(),
    MAE = mean(abs(predicted_mortality - observed_mortality)),
    RMSE = sqrt(mean((predicted_mortality - observed_mortality)^2)),
    `Predictive R2` = 1 -
      sum((predicted_mortality - observed_mortality)^2) /
      sum((observed_mortality - mean(observed_mortality))^2),
    `Balanced MAE` = mean(c(
      mean(abs(predicted_mortality - observed_mortality)[observed_mortality == 0]),
      mean(abs(predicted_mortality - observed_mortality)[observed_mortality > 0])
    )),
    `Severe-event MAE` = mean(
      abs(predicted_mortality - observed_mortality)[observed_mortality >= 0.2]
    ),
    `Occurrence Brier` = mean(
      (predicted_occurrence - as.numeric(observed_mortality > 0))^2
    ),
    Bias = mean(predicted_mortality - observed_mortality),
    .groups = "drop"
  )

production_route <- tibble::tribble(
  ~Programme, ~Learner,
  'LTMP benthic', 'Benchmark-BRMS-BRT stack',
  'LTMP manta tow', 'BRMS-BRT ensemble',
  'MMP inshore', 'BRMS-BRT ensemble'
)

selection_scores |>
  inner_join(production_route, by = c('Programme', 'Learner')) |>
  arrange(Programme) |>
  kable(digits = 3)
```

RMSE remains the primary measure for the continuous expected-mortality target:
it penalises large errors in the high-mortality observations that matter most
for GBR-wide loss estimates. It is not sufficient alone for this zero-heavy
response. Balanced MAE gives zero and positive observations equal influence,
severe-event MAE targets observations with at least 20% mortality, and the
occurrence Brier score assesses the boundary component.

LTMP and manta select the displayed routes consistently across the continuous
error measures. MMP is less decisive: BRMS has the lowest pooled RMSE (0.231),
but the BRMS--BRT ensemble is within 1%, improves balanced MAE and occurrence
Brier score, and retains both modelling frameworks. It is therefore the
provisional MMP production route. This explicit trade-off is preferable to
declaring a DHW-only benchmark the winner from one metric in a small,
three-event dataset.

# Event context added in this refit

Ten-year cumulative load is the sum of annual DHW above 4 within the ten full
calendar years before an event. Signed novelty is current annual maximum DHW
minus the greatest annual maximum in that preceding window. Negative values
therefore identify events weaker than recent exposure, while positive values
identify unprecedented exposure. The acute optical predictor is
`1.7 / Q90(Kd490)` during January--March: lower values indicate unusually low
water clarity. It is a consistent IMOS plume/turbidity proxy, not a direct
salinity measurement.

```{r event-context-plot, fig.width=10, fig.height=5.5}
event_context |>
  filter(event_year %in% c(2022, 2024)) |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Event = factor(event_year)
  ) |>
  ggplot(aes(
    x = novelty10, y = secchi_p10,
    colour = Event, size = mortality, shape = region_block
  )) +
  geom_vline(xintercept = 0, colour = 'grey65', linetype = 2) +
  geom_point(alpha = 0.85) +
  facet_wrap(~ Programme, scales = 'free') +
  scale_size_continuous(labels = percent_format(), range = c(2, 7)) +
  labs(
    x = 'Signed thermal novelty (DHW)',
    y = 'Acute IMOS Secchi proxy (m; lower = poorer clarity)',
    size = 'Observed mortality', shape = 'Region',
    title = 'Thermal novelty and optical conditions separate parts of 2022 and 2024'
  )
```

Southern 2024 observations have positive mean novelty, including the
Capricorn--Bunker signal, whereas 2022 is negative across the observed regions.
Northern and central 2024 observations also have lower IMOS Secchi proxies than
their 2022 counterparts. These summaries support the proposed mechanisms, but
the optical metric cannot distinguish freshwater from sediment, coloured
dissolved matter, or other sources of attenuation. Direct salinity remains a
sensitivity analysis when a complete, consistent 2024 product is available.

# Forecasting and reconstructing the 2024 event

```{r predictions-2024, fig.width=10, fig.height=4.8}
predictions_2024 <- predictions |>
  filter(scheme == "leave_one_event_out", fold == "2024") |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Learner = recode(learner, !!!learner_labels)
  )

ggplot(
  predictions_2024,
  aes(x = observed_mortality, y = predicted_mortality, colour = Learner)
) +
  geom_abline(slope = 1, intercept = 0, colour = "grey55", linetype = 2) +
  geom_point(alpha = 0.65, size = 1.7) +
  facet_wrap(~ Programme) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  scale_colour_brewer(palette = "Dark2") +
  labs(
    x = "Observed mortality proportion",
    y = "Prediction trained without 2024",
    title = "The 2024 fold remains the central temporal-transfer test"
  )
```

```{r means-2024}
predictions_2024 |>
  group_by(Programme, Learner) |>
  summarise(
    n = n(),
    `Observed mean` = mean(observed_mortality),
    `Predicted mean` = mean(predicted_mortality),
    Bias = mean(predicted_mortality - observed_mortality),
    RMSE = sqrt(mean((predicted_mortality - observed_mortality)^2)),
    `Predictive R2` = 1 -
      sum((predicted_mortality - observed_mortality)^2) /
      sum((observed_mortality - mean(observed_mortality))^2),
    `Balanced MAE` = mean(c(
      mean(abs(predicted_mortality - observed_mortality)[observed_mortality == 0]),
      mean(abs(predicted_mortality - observed_mortality)[observed_mortality > 0])
    )),
    .groups = "drop"
  ) |>
  arrange(Programme, RMSE) |>
  kable(digits = 3)
```

## Component calibration in 2022 and 2024

```{r event-component-plot, fig.width=10, fig.height=6}
event_calibration |>
  filter(learner == 'brms') |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Assessment = recode(scheme, !!!scheme_labels),
    Event = factor(event_year)
  ) |>
  select(Programme, Assessment, Event, obs, pred) |>
  tidyr::pivot_longer(
    c(obs, pred), names_to = 'Series', values_to = 'Mean_mortality'
  ) |>
  mutate(Series = recode(Series, obs = 'Observed', pred = 'BRMS')) |>
  ggplot(aes(x = Event, y = Mean_mortality, colour = Series, group = Series)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.5) +
  facet_grid(Programme ~ Assessment, scales = 'free_y') +
  scale_y_continuous(labels = percent_format()) +
  scale_colour_manual(values = c(Observed = 'black', BRMS = '#2C7FB8')) +
  labs(
    x = 'Bleaching event', y = 'Mean mortality', colour = NULL,
    title = 'Known-event effects improve reconstruction; unseen-event forecasting remains harder'
  )
```

```{r event-component-table}
event_calibration |>
  filter(learner == 'brms') |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Assessment = recode(scheme, !!!scheme_labels)
  ) |>
  select(
    Programme, Assessment, Event = event_year, n,
    `Observed mean` = obs, `Predicted mean` = pred,
    `Observed occurrence` = obs_occ, `Predicted occurrence` = pred_occ,
    `Observed positive magnitude` = obs_pos,
    `Predicted positive magnitude` = pred_pos
  ) |>
  arrange(Programme, Assessment, Event) |>
  kable(digits = 3)
```

For known-event reef-blocked reconstruction, BRMS reproduces 2024 mean
mortality closely: 0.112 versus 0.113 observed for LTMP benthic, 0.171 versus
0.191 for manta tow, and 0.246 versus 0.238 for MMP. Occurrence is also close.
Leave-one-event-out forecasting remains more conservative, particularly for
MMP. The 2022 positive magnitude is still overpredicted by BRMS, although the
known-event effect reduces the error. This residual is an explicit target for
future event-mechanism work, not a reason to discard the bounded beta response.

# Ensemble weights

```{r ensemble-weights, fig.width=7.5, fig.height=4.5}
weights_plot <- production_weights |>
  filter(learner == "stacked_ensemble") |>
  mutate(Programme = recode(programme_key, !!!programme_labels)) |>
  select(
    Programme,
    Benchmark = weight_benchmark,
    BRMS = weight_brms,
    BRT = weight_brt
  ) |>
  tidyr::pivot_longer(
    c(Benchmark, BRMS, BRT),
    names_to = "Learner",
    values_to = "Weight"
  )

ggplot(weights_plot, aes(x = Programme, y = Weight, fill = Learner)) +
  geom_col() +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Production ensemble weight")
```

The weights are performance estimates, not ecological importance measures. A
weight near zero means that one learner did not add squared-error skill to the
other under the current blocked data; it does not invalidate that framework's
uncertainty estimates or mechanistic interpretation.

```{r ensemble-weight-table}
production_weights |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Ensemble = recode(learner, !!!learner_labels)
  ) |>
  select(
    Programme, Ensemble,
    `Cross-fitted predictions` = n_cross_fitted_predictions,
    Benchmark = weight_benchmark, BRMS = weight_brms, BRT = weight_brt
  ) |>
  arrange(Programme, Ensemble) |>
  kable(digits = 3)
```

# Predictive influence

```{r brt-influence, fig.width=10, fig.height=9}
top_brt_influence <- brt_influence |>
  group_by(programme_key, component) |>
  slice_max(relative_influence, n = 6, with_ties = FALSE) |>
  ungroup() |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Component = recode(
      component,
      occurrence = "Any mortality",
      magnitude = "Magnitude when positive"
    )
  )

ggplot(
  top_brt_influence,
  aes(x = relative_influence, y = reorder(predictor, relative_influence))
) +
  geom_col(fill = "#2C7FB8") +
  facet_grid(Programme ~ Component, scales = "free_y", space = "free_y") +
  labs(
    x = "Relative BRT influence (%)",
    y = NULL,
    title = "Leading predictors in the full-data BRT refits"
  )
```

BRT influence measures predictive use, not causal effect or effect direction.
Correlated thermal-history variables can share or exchange influence. The BRMS
model therefore remains the appropriate framework for response-shape and
uncertainty summaries, while blocked predictions decide production routing.

# Ecological explanatory fit

Predictive validation and ecological explanation answer different questions.
For the nested two-part GLM benchmarks, AICc and McFadden pseudo-R2 are reported
for the binomial occurrence component. The positive-magnitude component is
quasibinomial, so it has no likelihood-based AIC; its deviance explained is
shown instead. The apparent response R2 is an in-sample descriptive statistic,
not an estimate of transfer skill.

```{r explanatory-benchmark-fit}
explanatory_metrics |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Model = recode(
      model,
      thermal_linear = 'Linear DHW',
      thermal_nonlinear = 'Nonlinear DHW',
      thermal_optical = 'Nonlinear DHW + optics',
      thermal_context = 'DHW + physical context',
      thermal_composition = 'DHW + context + Acropora'
    )
  ) |>
  select(
    Programme, Model,
    `Occurrence AICc` = occurrence_aicc,
    `Occurrence McFadden R2` = occurrence_mcfadden_r2,
    `Positive deviance explained` = positive_deviance_explained,
    `Apparent response R2` = apparent_response_r2
  ) |>
  arrange(Programme, `Occurrence AICc`) |>
  kable(digits = 3)
```

Adding preceding-year Acropora improves the apparent explanatory fit and the
occurrence AICc in all three programmes. It does not uniformly improve blocked
prediction in an unregularised GLM, especially for MMP. That combination is
evidence of ecological signal plus overfitting risk, not evidence that
Acropora should be removed. The predictor is therefore retained in the
regularised BRT and BRMS learners.

```{r acropora-brt-influence}
brt_influence |>
  filter(grepl('acropora', predictor)) |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Component = recode(
      component,
      occurrence = 'Any mortality',
      magnitude = 'Magnitude when positive'
    ),
    Predictor = recode(
      predictor,
      prop_acropora_pre = 'Pre-event Acropora proportion',
      acropora_cover_pre = 'Pre-event absolute Acropora cover'
    )
  ) |>
  select(Programme, Component, Predictor, `Relative influence (%)` = relative_influence) |>
  arrange(Programme, Component, desc(`Relative influence (%)`)) |>
  kable(digits = 2)
```

Both Acropora measures are used by both BRT components. Absolute pre-event
Acropora cover is particularly influential for occurrence in manta tow and
MMP. Relative influence is predictive usage rather than a causal effect, and
the two correlated composition measures may divide influence between them.

```{r brms-fit-statistics}
brms_fit_statistics |>
  mutate(Programme = recode(programme_key, !!!programme_labels)) |>
  select(
    Programme,
    `Marginal Bayesian R2` = marginal_bayesian_r2,
    `Marginal R2 90% lower` = marginal_bayesian_r2_q05,
    `Marginal R2 90% upper` = marginal_bayesian_r2_q95,
    `Conditional Bayesian R2` = conditional_bayesian_r2,
    LOOIC = looic,
    `Pareto k > 0.7` = pareto_k_over_0_7,
    `Maximum Pareto k` = max_pareto_k
  ) |>
  arrange(Programme) |>
  kable(digits = 3)
```

Marginal Bayesian R2 describes the variance captured by the population-level
environmental response; conditional Bayesian R2 also includes reef, event, and
region effects. LOOIC is reported for completeness but is not used to compare
programmes or route predictions. Each production fit has influential
observations with Pareto k above 0.7, so these raw LOO values require moment
matching or exact refits before they can support fine model comparisons.

AIC, LOOIC, and R2 are not interchangeable across the GLM, BRT, and BRMS
frameworks. The models use different likelihoods and boundary constructions.
Blocked scores on identical assessment rows remain the valid cross-framework
comparison; model-family-specific fit statistics support interpretation within
each framework.

# Diagnostics

```{r brms-diagnostics}
brms_diagnostics |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Assessment = recode(scheme, !!!scheme_labels),
    diagnostic_pass = divergences == 0 & max_treedepth_hits == 0 &
      max_rhat <= 1.01 & min_bulk_ess >= 100 & min_tail_ess >= 100
  ) |>
  group_by(Programme, Assessment) |>
  summarise(
    Folds = n(),
    `Maximum R-hat` = max(max_rhat),
    `Minimum bulk ESS` = min(min_bulk_ess),
    `Minimum tail ESS` = min(min_tail_ess),
    Divergences = sum(divergences),
    `Tree-depth hits` = sum(max_treedepth_hits),
    `Passing folds` = sum(diagnostic_pass),
    .groups = "drop"
  ) |>
  kable(digits = 2)
```

```{r brms-production-diagnostics}
brms_production_diagnostics |>
  mutate(Programme = recode(programme_key, !!!programme_labels)) |>
  select(
    Programme, n, event_years,
    `Maximum R-hat` = max_rhat,
    `Minimum bulk ESS` = min_bulk_ess,
    `Minimum tail ESS` = min_tail_ess,
    Divergences = divergences,
    `Tree-depth hits` = max_treedepth_hits
  ) |>
  kable(digits = 2)
```

All 36 outer BRMS fits pass the prespecified diagnostic gate: maximum R-hat
1.01, minimum bulk ESS 167, minimum tail ESS 301, zero divergences, and zero
maximum-tree-depth hits. All three longer production fits also pass.

```{r brt-tuning}
brt_tuning |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Assessment = recode(scheme, !!!scheme_labels)
  ) |>
  count(
    Programme, Assessment, interaction_depth, shrinkage, minobs,
    magnitude_mode, n_trees,
    name = "Outer folds selecting setting"
  ) |>
  arrange(Programme, Assessment, desc(`Outer folds selecting setting`)) |>
  kable(digits = 3)
```

# Interpretation and next use

The formal comparison answers a predictive question: how well can each learner
transfer to another bleaching event, another set of reefs, or another GBR
region? Coefficients and variable influence are secondary and must be read in
the context of correlated predictors. GBR-wide application should generate
predictions from every saved full-data component and then apply the validated
programme-specific route shown above. Environmental-novelty flags are still
required for extrapolation checks even though signed novelty is now a fitted
predictor, and BRT resampling uncertainty should be propagated alongside the
BRMS posterior before publication.

IMOS MODIS-Aqua Kd490 is used consistently as the water-clarity source. It is
preferred to the modelled eReefs chlorophyll/Kd490 field for this application
and provides the required 2024 coverage. Recovered nearshore Secchi estimates
retain their extraction radius and observation counts for sensitivity checks.
The new acute optical metric uses the Q1 Kd490 90th percentile from the same
source and extraction geometry; it should not be labelled freshwater exposure
without independent salinity validation.
