---
title: "Fast Bayesian mortality-model screen with R-INLA"
subtitle: "Blocked prediction, severe-event behaviour, and repeated-DHW history"
date: today
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)
library(tidyr)

theme_set(theme_minimal(base_size = 12))

result_dir <- '../output/inla_history_screen'
predictions <- read_csv(file.path(result_dir, 'predictions.csv'), show_col_types = FALSE)
metrics <- read_csv(file.path(result_dir, 'metrics.csv'), show_col_types = FALSE)
metrics_2024 <- read_csv(file.path(result_dir, 'metrics_2024.csv'), show_col_types = FALSE)
criteria <- read_csv(file.path(result_dir, 'production_criteria.csv'), show_col_types = FALSE)
effects <- read_csv(file.path(result_dir, 'fixed_effects.csv'), show_col_types = FALSE)
runtime <- read_csv(file.path(result_dir, 'runtime.csv'), show_col_types = FALSE)

candidate_labels <- c(
  legacy_history = 'Legacy history',
  paper_history = 'Event count + recovery interval'
)
programme_labels <- c(manta = 'Manta tow', ltmp = 'LTMP', mmp = 'MMP')

predictions <- predictions |>
  mutate(
    Candidate = recode(candidate, !!!candidate_labels),
    Programme = recode(programme_key, !!!programme_labels),
    residual = observed_mortality - predicted_mortality
  )
```

## Decision summary

R-INLA is fast enough to be useful as a **screening and model-development
engine**. The 56 blocked fits completed in
`r round(sum(runtime$elapsed_seconds) / 60, 1)` minutes, with a median of
`r round(median(runtime$elapsed_seconds), 1)` seconds per two-component fit.
This makes repeated ecological hypothesis tests practical before committing to
the longer BRMS fits.

The new paper-motivated history is not a universal replacement for the legacy
history variables. It improves manta-tow prediction under both blocked schemes,
including 2024, and modestly improves MMP when a whole event is omitted. It
worsens LTMP prediction and slightly worsens reef-blocked MMP prediction.
Pooled reef-blocked RMSE is effectively unchanged. The appropriate next move is
therefore to keep both history definitions as competing candidates and use INLA
to screen the expanded stressor model; only the best stable specification should
then receive the formal BRMS fit.

Most importantly, neither history formulation resolves severe-event
underprediction. Across reef-blocked predictions, mortality observations at or
above 50% are underpredicted by about 26 percentage points on average. This is
larger than the difference between the history formulations and points back to
missing event modifiers, susceptible community composition, and/or the response
curve rather than inference speed.

## What was fitted

Each programme was fitted separately using a hurdle-style approximation:

1. Bernoulli model for any mortality versus zero mortality.
2. Beta model for mortality magnitude conditional on mortality being positive.
3. Expected mortality = probability of occurrence x conditional magnitude.

Both components include independent reef, event-year, and region effects. Fixed
effects use weakly regularising Gaussian priors; random-effect standard deviations
use penalised-complexity priors. All imputation and standardisation are learned
inside each analysis fold.

This is close to the ecological decomposition in the formal BRMS model, but it
is not an exact replacement for BRMS zero-inflated and zero-one-inflated beta
likelihoods. In particular, the two exact MMP values of one are adjusted into the
open beta interval in the conditional component. The INLA fit is consequently a
fast ranking tool, not the final uncertainty product.

The compact shared predictor set contains current DHW, thresholded DHW terms,
pre-event total *Acropora*, pre-event coral cover, ten-year heat load and novelty,
low-water-clarity exposure, and cloud cover. The comparison changes only thermal
history:

| Candidate | Thermal-history terms |
|---|---|
| Legacy history | Historical mean DHW above 6 and years since DHW exceeded 6 |
| Event count + recovery interval | Number of >6-DHW events since 2016, capped years since the last event, and a no-prior-event flag |

Both candidates include interactions of current DHW with composition and thermal
history. Freshwater, cyclone, COTS, wind, and upwelling candidates are not yet in
the validation rows used here and are therefore not silently imputed into this
test.

## Blocked predictive performance

```{r}
metric_table <- metrics |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Candidate = recode(candidate, !!!candidate_labels),
    Scheme = recode(
      scheme,
      reef_blocked_5fold = 'Reef-blocked 5-fold',
      leave_one_event_out = 'Leave one event out'
    )
  ) |>
  select(Programme, Candidate, Scheme, rows, rmse, mae, predictive_r2,
         bias, occurrence_brier, severe_bias) |>
  arrange(Scheme, Programme, Candidate)

metric_table |>
  mutate(across(c(rmse, mae, predictive_r2, bias, occurrence_brier, severe_bias),
                ~ round(.x, 3))) |>
  kable(
    col.names = c('Programme', 'History', 'Validation', 'n', 'RMSE', 'MAE',
                  'Predictive R2', 'Bias', 'Occurrence Brier', 'Severe bias'),
    caption = 'Out-of-sample metrics. Bias is prediction minus observation.'
  )
```

```{r}
pooled <- predictions |>
  group_by(Candidate, scheme) |>
  summarise(
    rows = n(),
    rmse = sqrt(mean(residual^2)),
    mae = mean(abs(residual)),
    predictive_r2 = 1 - sum(residual^2) /
      sum((observed_mortality - mean(observed_mortality))^2),
    bias = mean(-residual),
    severe_bias = mean(-residual[observed_mortality >= 0.5]),
    .groups = 'drop'
  ) |>
  mutate(Validation = recode(
    scheme,
    reef_blocked_5fold = 'Reef-blocked 5-fold',
    leave_one_event_out = 'Leave one event out'
  ))

pooled |>
  select(Candidate, Validation, rows, rmse, mae, predictive_r2, bias, severe_bias) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable(
    col.names = c('History', 'Validation', 'n', 'RMSE', 'MAE',
                  'Predictive R2', 'Bias', 'Severe bias'),
    caption = 'Pooled programme observations; programmes are not reweighted.'
  )
```

The pooled result should not choose the final model because it gives the largest
programme the most weight. Its value here is to show scale: changing the history
definition moves RMSE far less than the residual error at high mortality.

## The 2024 event

```{r}
metrics_2024 |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Candidate = recode(candidate, !!!candidate_labels)
  ) |>
  select(Programme, Candidate, rows, rmse, mae, predictive_r2, bias,
         severe_rows, severe_bias) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable(
    col.names = c('Programme', 'History', 'n', 'RMSE', 'MAE', 'Predictive R2',
                  'Bias', 'Severe n', 'Severe bias'),
    caption = 'Reef-blocked performance within 2024.'
  )
```

```{r}
predictions |>
  filter(scheme == 'reef_blocked_5fold', event_year == 2024) |>
  ggplot(aes(observed_mortality, predicted_mortality, colour = Candidate)) +
  geom_abline(slope = 1, intercept = 0, colour = 'grey45', linewidth = 0.5) +
  geom_point(alpha = 0.72, size = 1.8) +
  facet_wrap(~ Programme) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent) +
  scale_colour_manual(values = c('#7b3294', '#008837')) +
  labs(
    title = 'Both INLA candidates still shrink the most severe 2024 losses',
    subtitle = 'Predictions are out of sample for reef identity',
    x = 'Observed mortality', y = 'Predicted mortality', colour = NULL
  )
```

```{r}
predictions |>
  filter(
    scheme == 'reef_blocked_5fold', event_year == 2024,
    Candidate == 'Event count + recovery interval'
  ) |>
  arrange(desc(abs(residual))) |>
  slice_head(n = 15) |>
  transmute(
    Programme, ReefID, ReefName,
    Observed = percent(observed_mortality, accuracy = 0.1),
    Predicted = percent(predicted_mortality, accuracy = 0.1),
    `Observed - predicted` = percent(residual, accuracy = 0.1)
  ) |>
  kable(caption = 'Largest absolute 2024 reef-blocked residuals for the new history candidate.')
```

The familiar manta-tow misses remain visible at Linnet, Mackay, and Eyrie. The
largest MMP misses are repeated observations at Snapper Island. Conversely,
Fairfax, Hyde, Barren, Tobias, and Fitzroy include large overpredictions. These
are useful ecological audit targets: a single steeper DHW curve cannot fix both
directions without modifiers that separate exacerbated from ameliorated exposure.

## Full-data fit criteria

```{r}
criteria |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Candidate = recode(candidate, !!!candidate_labels)
  ) |>
  group_by(Programme) |>
  mutate(
    delta_waic = total_waic - min(total_waic),
    delta_dic = total_dic - min(total_dic)
  ) |>
  ungroup() |>
  select(Programme, Candidate, total_waic, delta_waic, total_dic, delta_dic) |>
  mutate(across(where(is.numeric), ~ round(.x, 1))) |>
  arrange(Programme, delta_waic) |>
  kable(
    col.names = c('Programme', 'History', 'WAIC', 'Delta WAIC', 'DIC', 'Delta DIC'),
    caption = 'Summed occurrence and positive-magnitude criteria from full-data fits.'
  )
```

WAIC favours the new history for manta tow and MMP, but the LTMP result favours
legacy history. These criteria describe full-data fit and cannot override the
blocked prediction results. They are most useful when they agree with the
out-of-sample comparison.

## Posterior history effects

```{r}
history_effects <- effects |>
  filter(
    candidate == 'paper_history',
    grepl('dhw_events|years_since|no_prior|dhw_x_history|dhw_x_recovery', term)
  ) |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Component = recode(component, occurrence = 'Any mortality'),
    Term = recode(
      term,
      dhw_events_since2016_n6_z = 'Prior event count',
      dhw_years_since_last_n6_capped8_z = 'Recovery interval',
      dhw_no_prior_n6 = 'No prior event',
      dhw_x_history_count = 'DHW x event count',
      dhw_x_recovery = 'DHW x recovery interval'
    )
  )

history_effects |>
  ggplot(aes(mean, Term, xmin = q025, xmax = q975, colour = Component)) +
  geom_vline(xintercept = 0, colour = 'grey60', linewidth = 0.5) +
  geom_errorbarh(height = 0, position = position_dodge(width = 0.45)) +
  geom_point(position = position_dodge(width = 0.45), size = 1.7) +
  facet_wrap(~ Programme) +
  scale_colour_manual(values = c('#2166ac', '#b2182b')) +
  labs(
    title = 'Most individual history effects remain uncertain',
    subtitle = 'Posterior means and 95% credible intervals on link scales',
    x = 'Coefficient', y = NULL, colour = 'Model component'
  )
```

The most consistent ecological signal is in manta tow: a longer recovery
interval is positively associated with the magnitude of mortality, compatible
with rebuilding susceptible communities between major events. Most other
history intervals overlap zero, especially in MMP. These terms should therefore
be judged as a grouped predictive hypothesis, not selected one coefficient at a
time.

## Recommended next implementation step

Use INLA for one controlled expansion of this screen:

1. Join the operationally available stressor fields to these exact validation
   rows: coloured-water anomaly or rainfall/runoff proxy, cyclone 4-m wave hours,
   COTS pressure, wind/doldrum exposure, and an upwelling or cooling proxy.
2. Compare a shared core, core plus exacerbating stressors, core plus
   ameliorating stressors, and the combined model under the same blocked folds.
3. Retain both history definitions during that comparison; do not choose a
   programme-wide winner from pooled RMSE alone.
4. Examine severe-loss sensitivity, calibration, and named reef-year residuals
   alongside RMSE, MAE, predictive R2, occurrence Brier score, WAIC, and DIC.
5. Promote the smallest set of candidates that materially improves blocked
   prediction—especially severe 2024 mortality—to formal BRMS fitting. BRMS then
   supplies the exact boundary-aware likelihood, posterior predictive checks,
   and final uncertainty intervals.

This preserves the speed advantage of INLA without treating its approximate
two-part likelihood as the final operational model.
