# Joint modifiers and extreme bleaching mortality

Coherent Acropora terms, a severe-tail learner, and INLA/BRT blending.

```{r}
#| echo: false
#| warning: false
#| message: false
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)
library(tidyr)

out <- '../output/joint_extreme_relative'
composition <- read_csv(file.path(out, 'composition_metrics.csv'), show_col_types = FALSE)
tail_all <- read_csv(file.path(out, 'ensemble_metrics.csv'), show_col_types = FALSE)
tail_2024 <- read_csv(file.path(out, 'ensemble_metrics_2024.csv'), show_col_types = FALSE)
framework <- read_csv(file.path(out, 'framework_metrics.csv'), show_col_types = FALSE)
framework_2024 <- read_csv(file.path(out, 'framework_metrics_2024.csv'), show_col_types = FALSE)
curves <- read_csv(file.path(out, 'joint_scenario_curves.csv'), show_col_types = FALSE)
influence <- read_csv(file.path(out, 'production_influence.csv'), show_col_types = FALSE)
errors <- read_csv(file.path(out, 'tail_error_changes.csv'), show_col_types = FALSE)
programme_labels <- c(ltmp = 'LTMP', manta = 'Manta tow', mmp = 'MMP')
theme_set(theme_minimal(base_size = 12))
```

## Decision

- Keep all zeros in the Bernoulli occurrence component.
- Use Acropora proportion, total pre-event cover and their DHW interaction.
- Retain the bounded-response INLA/BRMS model and nonlinear BRT as generalists.
  Add the tail BRT as an event-risk expert, not a universal replacement.
- Use conservative or nested-CV ensemble weights for routine maps and show the
  soft gate separately as a high-risk estimate.

Across 537 reef-blocked predictions, the standard BRT predicts 28.1% for cases
that exceed 50%. Conservative and soft tail learners raise this to 32.5% and
36.9%; pooled RMSE changes from 15.7% to 15.3% and 15.4%. False extremes among
observations below 30% rise from 4.1% to 5.9% and 8.3%.

## Acropora parameterisation

All alternatives include total pre-event cover and use identical reef-blocked
folds. Absolute Acropora cover is a useful sensitivity, but including it with
proportion is not consistently better and recreates correlated terms.

```{r}
#| echo: false
composition |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Parameterisation = recode(
      composition,
      proportion_only = 'Proportion + total cover',
      absolute_cover_only = 'Absolute Acropora cover',
      legacy_dual = 'Proportion + absolute cover'
    )
  ) |>
  ggplot(aes(Parameterisation, rmse, fill = Parameterisation)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Programme, scales = 'free_y') +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = 'Reef-blocked RMSE',
       title = 'The dual Acropora representation does not improve prediction') +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

Proportion plus total cover is the clearest common basis: proportion describes
susceptibility and total cover describes how much coral was available to lose.
Absolute cover is slightly better for LTMP and MMP but worse for Manta, so it
remains a sensitivity rather than an additional default term.

## Tail learner trade-off

The standard model predicts $E(M)=P(M>0)E(M\mid M>0)$. The specialist also
estimates the probability and magnitude of mortality at or above 30%. This
threshold is a modelling device, not an ecological discontinuity.

```{r}
#| echo: false
tail_all |>
  mutate(Programme = recode(programme_key, !!!programme_labels)) |>
  ggplot(aes(false_extreme_rate, severe_rmse, colour = learner)) +
  geom_point(aes(size = rmse), alpha = 0.9) +
  facet_wrap(~Programme, scales = 'free') +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent) +
  labs(x = 'False extreme rate', y = 'Severe-case RMSE',
       colour = 'Learner', size = 'Overall RMSE',
       title = 'Tail improvement has a measurable false-alarm cost') +
  theme(legend.position = 'bottom')
```

```{r}
#| echo: false
tail_2024 |>
  filter(learner %in% c(
    'Standard two-part BRT', 'Conservative tail ensemble',
    'Soft-gated tail ensemble'
  )) |>
  mutate(Programme = recode(programme_key, !!!programme_labels)) |>
  transmute(
    Programme, Model = learner, N = n, Severe_N = severe_n,
    RMSE = percent(rmse, accuracy = 0.1),
    Bias = percent(bias, accuracy = 0.1),
    Severe_observed = percent(severe_observed, accuracy = 0.1),
    Severe_predicted = percent(severe_predicted, accuracy = 0.1),
    False_extreme = percent(false_extreme_rate, accuracy = 0.1)
  ) |>
  kable(col.names = c(
    'Programme', 'Model', 'N', 'Severe N', 'RMSE', 'Bias',
    'Severe observed', 'Severe predicted', 'False extreme'
  ), caption = 'Reef-blocked performance for the 2024 event')
```

For 2024, the soft gate raises severe predictions from 39.8% to 52.9% in LTMP,
31.4% to 35.2% in Manta tow, and 23.9% to 35.0% in MMP. It reduces RMSE for
Manta and MMP while LTMP RMSE is essentially unchanged. MMP severe loss remains
substantially underestimated.

## Beta--Bernoulli and BRT are complementary

The existing INLA screen uses Bernoulli occurrence and conditional beta
magnitude. The comparison uses exactly matched held-out observations and the
best screened history definition per programme. Equal blends are fixed at
50:50; their weights were not tuned on these predictions.

```{r}
#| echo: false
models <- c(
  'INLA beta + Bernoulli', 'Standard two-part BRT',
  'Soft-gated tail ensemble',
  'Equal INLA + Soft-gated tail ensemble'
)
framework |>
  filter(learner %in% models) |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Model = factor(learner, levels = models)
  ) |>
  ggplot(aes(Model, rmse, fill = Model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Programme, scales = 'free_y') +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = 'Reef-blocked RMSE',
       title = 'No framework dominates every programme') +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r}
#| echo: false
framework_2024 |>
  filter(learner %in% models) |>
  mutate(Programme = recode(programme_key, !!!programme_labels)) |>
  transmute(
    Programme, Model = learner,
    RMSE = percent(rmse, accuracy = 0.1),
    Predictive_R2 = round(predictive_r2, 3),
    Severe_predicted = percent(severe_predicted, accuracy = 0.1),
    Severe_observed = percent(severe_observed, accuracy = 0.1),
    False_extreme = percent(false_extreme_rate, accuracy = 0.1)
  ) |>
  kable(col.names = c(
    'Programme', 'Model', 'RMSE', 'Predictive R2', 'Severe predicted',
    'Severe observed', 'False extreme'
  ), caption = 'Matched INLA and BRT predictions for 2024')
```

INLA is especially useful for LTMP 2024: its severe mean is 54.8% versus 62.0%
observed. The tail BRT adds more for Manta. For MMP, INLA improves general fit
while the tail BRT responds more strongly to severe cases but raises false
alarms. A single equal blend should therefore not be forced across programmes.

## Joint compound-risk curves

Unlike the earlier one-at-a-time curves, these profiles move Acropora, thermal
novelty, coastal rainfall, coloured-water exposure and cloud together. High
risk uses high quantiles of the first four and low cloud. These are conditional
predictions rather than causal effects.

```{r}
#| echo: false
curve_long <- curves |>
  select(programme_key, scenario, ann_maxdhw, outside_observed_dhw,
         predicted_mortality, conservative_prediction) |>
  pivot_longer(
    c(predicted_mortality, conservative_prediction),
    names_to = 'model', values_to = 'prediction'
  ) |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Model = recode(
      model, predicted_mortality = 'Standard BRT',
      conservative_prediction = 'Conservative tail'
    ),
    Support = if_else(outside_observed_dhw, 'Outside support', 'Observed')
  )

ggplot(curve_long,
       aes(ann_maxdhw, prediction, colour = scenario, linetype = Model,
           alpha = Support)) +
  geom_line(linewidth = 0.9) +
  facet_grid(Programme ~ scenario) +
  scale_y_continuous(labels = percent, limits = c(0, 0.65)) +
  scale_alpha_manual(values = c(Observed = 1, 'Outside support' = 0.25)) +
  labs(x = 'Annual maximum DHW', y = 'Predicted relative mortality',
       colour = 'Profile', linetype = 'Model', alpha = NULL,
       title = 'Joint modifiers recover a high-risk response near 12 DHW') +
  theme(legend.position = 'bottom')
```

At 11.5 DHW, inside every programme's observed range, high compound risk
predicts 35.9% for LTMP, 46.6% for Manta and 51.2% for MMP with the standard
BRT. The conservative tail version predicts 45.4%, 48.9% and 54.1%; low-risk
profiles remain below 7%. This is much closer to the ecological expectation
that about 12 DHW can produce at least 50% mortality in susceptible communities.

The profiles have small local reversals because interaction features are not
globally monotone, and trees plateau outside support. The formal BRMS model
should impose a smooth monotone thermal response within modifier profiles.

## Component influence

```{r}
#| echo: false
component_labels <- c(
  base_occurrence = 'Any loss', base_magnitude = 'Positive magnitude',
  tail_gate = 'At least 30% loss', tail_magnitude = 'Severe magnitude'
)
influence |>
  group_by(component) |>
  slice_max(relative_influence, n = 7, with_ties = FALSE) |>
  ungroup() |>
  mutate(
    Component = recode(component, !!!component_labels),
    predictor = reorder(predictor, relative_influence)
  ) |>
  ggplot(aes(relative_influence, predictor, fill = Component)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Component, scales = 'free_y') +
  labs(x = 'Relative BRT influence', y = NULL,
       title = 'Modifiers enter different response components')
```

Thermal novelty is the strongest occurrence and tail-gate predictor.
Freshwater joint exposure is third in the tail gate, Acropora-by-DHW is
important in both magnitude models, and cloud and rainfall are prominent in
positive magnitude. These ranks support retention but are not causal effects.

## Reefs helped and harmed

```{r}
#| echo: false
display_errors <- bind_rows(
  errors |> slice_head(n = 8) |> mutate(set = 'Largest improvement'),
  errors |> slice_tail(n = 6) |> arrange(error_reduction) |>
    mutate(set = 'Largest deterioration')
) |>
  mutate(Programme = recode(programme_key, !!!programme_labels)) |>
  transmute(
    set, Programme, reef = ReefName, year = event_year,
    dhw = round(ann_maxdhw, 2),
    observed = percent(observed_mortality, accuracy = 0.1),
    standard = percent(.data[['Standard two-part BRT']], accuracy = 0.1),
    conservative = percent(
      .data[['Conservative tail ensemble']], accuracy = 0.1
    ),
    error_change = percent(error_reduction, accuracy = 0.1)
  )
kable(display_errors, col.names = c(
  'Set', 'Programme', 'Reef', 'Year', 'DHW', 'Observed', 'Standard',
  'Conservative', 'Absolute-error improvement'
))
```

Largest gains include severe 2024 losses at Lady Musgrave, Fitzroy Island,
Humpy and North Keppel. The cost is overprediction of some low-loss
observations, including repeated MMP observations at Fitzroy. This within-reef
disagreement reinforces the need for a shared ecological state with
programme-specific observation layers.

## Formal next model

1. Use Acropora proportion, total pre-event cover and their DHW interaction;
   keep absolute Acropora cover as a sensitivity.
2. Retain Bernoulli occurrence and conditional beta magnitude. Add a continuous
   tail mixture or robust event-level scale rather than deleting observations.
3. Include DHW interactions for Acropora, novelty, freshwater/coloured water
   and cloud, partially pooled across programme observation layers.
4. Estimate BRT/BRMS stacking weights inside nested reef-blocked and
   leave-event-out validation. Keep 2024 in at least one operational candidate
   while retaining a forward-event stress test.
5. Publish general, conservative and high-risk predictions separately.

Implementation: scripts/fit_joint_extreme_relative_screen.R,
scripts/combine_relative_framework_predictions.R,
output/joint_extreme_relative/, and this report.
