---
title: 'Does freshwater-event context explain severe mortality misses?'
subtitle: 'Joint observation model and 2024 rainfall/calm-weather sensitivity'
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)

input_dir <- '../output/joint_compound_models'
metrics <- read_csv(file.path(input_dir, 'comparison_metrics.csv'), show_col_types = FALSE)
events <- read_csv(file.path(input_dir, 'event_metrics.csv'), show_col_types = FALSE)
ablation <- read_csv(file.path(input_dir, 'weather_ablation.csv'), show_col_types = FALSE)
effects <- read_csv(file.path(input_dir, 'brms_weather_effects.csv'), show_col_types = FALSE)
reefs <- read_csv(file.path(input_dir, 'manta_2024_severe_reefs.csv'), show_col_types = FALSE)
regions <- read_csv(file.path(input_dir, 'manta_2024_regions.csv'), show_col_types = FALSE)
diagnostics <- read_csv(file.path(input_dir, 'joint_brms_diagnostics.csv'), show_col_types = FALSE)
decision <- read_csv(file.path(input_dir, 'model_decision.csv'), show_col_types = FALSE)

programme_labels <- c(ltmp = 'LTMP benthic', manta = 'Manta tow', mmp = 'MMP inshore')
model_labels <- c(
  routed_benchmark = 'Current routed benchmark',
  formal_spatial_benchmark = 'Formal spatial benchmark',
  joint_core_brt = 'Joint BRT without weather',
  joint_compound_brt = 'Joint BRT with weather',
  joint_compound_brms = 'Joint BRMS with weather'
)
```

# Decision

The freshwater hypothesis receives **qualified rather than production-grade
support**. ERA5 rainfall and calm conditions improve the joint BRT's 2024
spatial predictions for manta tow, particularly at wet northern and central
reefs. They do not improve transfer to a completely unseen event, do not help
MMP, and the Bayesian model does not support a positive DHW--rainfall synergy.

```{r decision-table}
decision |>
  kable(caption = 'Model promotion decision by prediction target')
```

The existing programme-specific production route is therefore unchanged. The
joint weather BRT is retained as a 2024 reconstruction sensitivity, not as the
GBR-wide future-event predictor.

# What was tested

The joint dataset contains 537 observations and 306 unique reef-events. Of
these, 152 reef-events are observed by two programmes. The Bayesian model uses
shared ecological effects, a shared reef-event deviation, and
programme-specific mean offsets, boundary probabilities, complete-loss
probabilities, and precision.

The compact weather addition is deliberately limited to:

- ERA5 nearest-coastal maximum 30-day rainfall;
- ERA5 reef mean wind and fraction of calm hours below 3 m/s;
- distance to the coastal ERA5 cell;
- DHW by rainfall and DHW by calm-fraction interactions.

The partial January 2024 eReefs salinity record is not used as a predictor.
Disturbance codes and descriptions are retained for interpretation but do not
affect fitting or inclusion.

# Forecasting an unseen bleaching event

```{r event-performance}
metrics |>
  filter(scheme == 'leave_one_event_out') |>
  mutate(
    Programme = programme_labels[programme_key],
    Model = model_labels[learner],
    across(c(rmse, mae, predictive_r2, severe_mae, severe_bias), ~ round(.x, 3))
  ) |>
  select(
    Programme, Model, n, `Severe n` = severe_n,
    RMSE = rmse, MAE = mae, `Predictive R2` = predictive_r2,
    `Severe MAE` = severe_mae,
    `Severe observed - predicted` = severe_bias
  ) |>
  arrange(Programme, RMSE) |>
  kable(caption = 'Leave-one-event-out performance')
```

The current routed benchmark remains best. In manta tow, the weather BRT has
RMSE 0.152 versus 0.141 for the benchmark, while severe MAE increases from
0.221 to 0.251. The joint BRMS is poorer again. Freshwater context therefore
cannot yet be used to forecast the severity of a novel event from historical
events alone.

```{r event-calibration, fig.width=10, fig.height=6}
events |>
  filter(learner %in% c('routed_benchmark', 'joint_compound_brt', 'joint_compound_brms')) |>
  mutate(
    Programme = programme_labels[programme_key],
    Model = model_labels[learner]
  ) |>
  ggplot(aes(event_year, observed_mean - predicted_mean, colour = Model)) +
  geom_hline(yintercept = 0, colour = 'grey45') +
  geom_line() +
  geom_point(size = 2.3) +
  facet_wrap(~ Programme) +
  scale_colour_manual(values = c(
    'Current routed benchmark' = '#4D4D4D',
    'Joint BRT with weather' = '#1B9E77',
    'Joint BRMS with weather' = '#D95F02'
  )) +
  labs(
    x = 'Held-out event', y = 'Observed - predicted event mean',
    colour = NULL,
    title = 'New-event calibration remains the limiting test',
    subtitle = 'Positive values indicate event-wide underprediction'
  ) +
  theme(legend.position = 'top')
```

# Mapping 2024 after the event is represented

```{r spatial-performance}
metrics |>
  filter(scheme == 'reef_blocked_2024') |>
  mutate(
    Programme = programme_labels[programme_key],
    Model = model_labels[learner],
    across(c(rmse, mae, predictive_r2, severe_mae, severe_bias), ~ round(.x, 3))
  ) |>
  select(
    Programme, Model, n, `Severe n` = severe_n,
    RMSE = rmse, MAE = mae, `Predictive R2` = predictive_r2,
    `Severe MAE` = severe_mae,
    `Severe observed - predicted` = severe_bias,
    `Non-severe MAE` = nonsevere_mae
  ) |>
  arrange(Programme, RMSE) |>
  kable(caption = 'Strict reef-blocked reconstruction of 2024')
```

For manta tow, adding weather to the joint BRT reduces spatial RMSE from 0.161
to 0.151 and severe MAE from 0.179 to 0.165. It also reduces LTMP severe MAE,
but slightly worsens its overall error. MMP does not benefit.

```{r ablation-table}
ablation |>
  mutate(
    Programme = programme_labels[programme_key],
    Target = recode(
      scheme,
      leave_one_event_out = 'Unseen event',
      reef_blocked_2024 = '2024 spatial reconstruction'
    ),
    across(starts_with('delta_'), ~ round(.x, 3))
  ) |>
  select(
    Target, Programme,
    `Delta RMSE` = delta_rmse_compound_minus_core,
    `Delta severe MAE` = delta_severe_mae_compound_minus_core,
    `Delta non-severe MAE` = delta_nonsevere_mae_compound_minus_core
  ) |>
  kable(caption = 'Weather BRT minus core BRT; negative values are improvements')
```

# Is the effect a DHW interaction?

```{r weather-effects}
effects |>
  mutate(
    across(c(estimate, q025, q975, probability_ecological_direction), ~ round(.x, 3))
  ) |>
  select(
    Effect = effect, Estimate = estimate,
    `2.5%` = q025, `97.5%` = q975,
    `P(ecological direction)` = probability_ecological_direction
  ) |>
  kable(caption = 'All-event joint BRMS weather posterior')
```

Rainfall has a clear positive **main effect** on positive mortality. It also
tends to reduce zero inflation, making some mortality more likely. In contrast,
the positive DHW--rainfall synergy has posterior probability only 0.118 in the
magnitude component and 0.102 in the boundary component. Calm conditions show
the same broad pattern but with much greater uncertainty.

```{r effect-plot, fig.width=9, fig.height=5.5}
effects |>
  mutate(effect = factor(effect, levels = rev(effect))) |>
  ggplot(aes(estimate, effect)) +
  geom_vline(xintercept = 0, colour = 'grey45') +
  geom_errorbarh(aes(xmin = q025, xmax = q975), height = 0.15) +
  geom_point(aes(colour = ecological_direction), size = 2.5) +
  scale_colour_manual(values = c(positive = '#B2182B', negative = '#2166AC')) +
  labs(
    x = 'Standardised logit-scale coefficient', y = NULL,
    colour = 'Expected direction',
    title = 'The supported rainfall signal is additive, not synergistic'
  )
```

# Which severe manta reefs respond to weather?

```{r manta-regions}
regions |>
  mutate(
    Model = model_labels[learner],
    across(c(observed_mean, predicted_mean, severe_mae, severe_bias), ~ round(.x, 3))
  ) |>
  select(
    Model, Region = region_block, n,
    Observed = observed_mean, Predicted = predicted_mean,
    `Severe MAE` = severe_mae,
    `Observed - predicted` = severe_bias
  ) |>
  kable(caption = '2024 severe manta mortality by region and model')
```

```{r manta-gain-plot, fig.width=10, fig.height=6}
reefs |>
  ggplot(aes(coastal_rain30_mm, weather_gain_vs_core)) +
  geom_hline(yintercept = 0, colour = 'grey45') +
  geom_point(
    aes(colour = region_block, size = era5_wind_calm_fraction), alpha = 0.8
  ) +
  geom_text(
    data = ~ head(.x, 8), aes(label = ReefID),
    nudge_y = 0.012, size = 3, show.legend = FALSE
  ) +
  scale_x_log10() +
  scale_size_continuous(labels = percent, name = 'Calm fraction') +
  labs(
    x = 'ERA5 coastal maximum 30-day rainfall (mm)',
    y = 'Reduction in absolute error after adding weather',
    colour = 'Region',
    title = 'Weather helps most at several wet and calm manta reefs',
    subtitle = 'Positive values favour the compound-weather BRT'
  )
```

```{r manta-reef-table}
reefs |>
  transmute(
    Reef = ReefName, Region = region_block,
    DHW = round(ann_maxdhw, 2),
    Observed = round(observed_mortality, 3),
    `Formal spatial` = round(predicted_mortality_formal_spatial_benchmark, 3),
    `Core joint BRT` = round(predicted_mortality_joint_core_brt, 3),
    `Weather joint BRT` = round(predicted_mortality_joint_compound_brt, 3),
    `Joint BRMS` = round(predicted_mortality_joint_compound_brms, 3),
    `Rain 30d (mm)` = round(coastal_rain30_mm),
    `Calm fraction` = percent(era5_wind_calm_fraction, accuracy = 1),
    `Weather error reduction` = round(weather_gain_vs_core, 3),
    Disturbance = DISTURBANCE_TYPE,
    Notes = description
  ) |>
  head(25) |>
  kable(caption = 'Severe manta reefs ranked by weather-related error reduction')
```

Weather improves several northern and central reefs, including MacGillivray,
Swinger, South Direction, Hastings, Lizard Island NW, St Crispin, Mackay and
Eyrie. Hastings is explicitly associated with Jasper. The gain is not confined
to recorded multiple disturbances, and substantial underprediction remains.

# Sampling diagnostics and interpretation boundary

```{r diagnostics}
diagnostics |>
  mutate(
    `Pass gate` = max_rhat <= 1.01 & min_bulk_ess >= 100 &
      min_tail_ess >= 100 & divergences == 0 & max_treedepth_hits == 0,
    across(c(max_rhat, min_bulk_ess, min_tail_ess), ~ round(.x, 3))
  ) |>
  select(
    Target = scheme, Fold = fold,
    `Maximum R-hat` = max_rhat,
    `Minimum bulk ESS` = min_bulk_ess,
    `Minimum tail ESS` = min_tail_ess,
    Divergences = divergences,
    `Tree-depth hits` = max_treedepth_hits,
    `Pass gate`
  ) |>
  kable(caption = 'Joint BRMS screening-fold diagnostics')
```

Several two-chain screening folds exceed the strict R-hat gate, although none
has divergences or tree-depth failures and the four-chain production fit passes.
Because the candidate fails predictive promotion regardless, targeted refits
would not change the decision and are not warranted at this stage.

The evidence currently supports rainfall/calm weather as an additive 2024
reconstruction sensitivity. It does not yet support calling the inferred
effect measured salinity, imposing a positive DHW interaction, or changing the
GBR-wide production route.
