---
title: "GBR Coral Mortality: Blocked Validation Diagnostics"
subtitle: "Baseline models and the path to a BRMS–BRT prediction ensemble"
author: "Samuel Matthews"
date-modified: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    theme: cosmo
    code-fold: true
execute:
  warning: false
  message: false
  echo: false
---

```{r setup}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(scales)
library(knitr)

repo_root <- if (file.exists("output/model_validation/blocked_predictions.csv")) "." else ".."
result_path <- function(file) file.path(repo_root, "output", "model_validation", file)

required <- c(
    "blocked_predictions.csv", "candidate_metrics.csv",
    "candidate_fold_metrics.csv", "candidate_selection.csv",
    "selected_models.csv", "fit_warnings.csv"
)
if (!all(file.exists(vapply(required, result_path, character(1))))) {
    stop("Model results are missing. Run Rscript scripts/fit_candidate_models.R first.")
}

predictions <- read_csv(result_path("blocked_predictions.csv"), show_col_types = FALSE)
metrics <- read_csv(result_path("candidate_metrics.csv"), show_col_types = FALSE)
fold_metrics <- read_csv(result_path("candidate_fold_metrics.csv"), show_col_types = FALSE)
selection <- read_csv(result_path("candidate_selection.csv"), show_col_types = FALSE)
selected <- read_csv(result_path("selected_models.csv"), show_col_types = FALSE)
fit_warnings <- read_csv(result_path("fit_warnings.csv"), show_col_types = FALSE)

programme_labels <- c(
    manta = "LTMP manta tow",
    ltmp = "LTMP benthic",
    mmp = "MMP inshore"
)
model_labels <- c(
    thermal_linear = "Linear DHW",
    thermal_nonlinear = "Nonlinear DHW",
    thermal_optical = "DHW + optical context",
    thermal_context = "Full physical context"
)
model_labels <- c(
    model_labels,
    thermal_composition = 'DHW + context + Acropora'
)
scheme_labels <- c(
    leave_one_event_out = "Leave one event out",
    reef_blocked_5fold = "Reef blocked",
    region_blocked = "Region blocked"
)

label_results <- function(data) {
    data |>
        mutate(
            Programme = recode(programme_key, !!!programme_labels),
            Model = recode(model, !!!model_labels),
            Scheme = recode(scheme, !!!scheme_labels)
        )
}

metrics_plot <- label_results(metrics)
fold_metrics_plot <- label_results(fold_metrics)
selected_predictions <- predictions |>
    inner_join(select(selected, programme_key, model), by = c("programme_key", "model")) |>
    label_results()

theme_set(theme_classic(base_size = 12))
```

::: {.callout-warning}
## These are benchmark models, not the final prediction machine

The explanatory parsimony rule selects linear DHW for manta and MMP, but the
prediction-optimal manta benchmark is the full physical-context model. These
are reference GLMs, not ecological claims that mortality is truly linear in
DHW and not the GBR-wide prediction machine. The formal BRMS and BRT models use
the full predictor contract, including preceding-year Acropora, and ensemble
their cross-fitted predictions.
:::

# Data and validation design

All modern observations are used in modelling. Each event—including 2024—is
held out in turn for temporal validation. Reef-blocked folds test transfer to
unseen reefs, while broad region blocks provide a harder geographic stress
test. After validation, production models are refitted using every event.

```{r data-summary}
data_summary <- predictions |>
    distinct(programme_key, source_observation_id, ReefID, event_year) |>
    group_by(programme_key) |>
    summarise(
        Observations = n(),
        Reefs = n_distinct(ReefID),
        Events = paste(sort(unique(event_year)), collapse = ", "),
        .groups = "drop"
    ) |>
    mutate(Programme = recode(programme_key, !!!programme_labels)) |>
    select(Programme, everything(), -programme_key)

kable(data_summary, align = c("l", "r", "r", "l"))
```

# Candidate model comparison

Five nested hypotheses were evaluated. The comparison adds a preceding-year
Acropora composition model to the DHW, nonlinear, optical, and physical-context
benchmarks. It tests how much predictive accuracy is gained before moving to
the regularised BRMS and BRT frameworks.

```{r candidate-rmse, fig.width=11, fig.height=6}
ggplot(
    metrics_plot,
    aes(x = Model, y = rmse, colour = Scheme, group = Scheme)
) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 2.5) +
    facet_wrap(~Programme, scales = "free_y") +
    scale_colour_brewer(palette = "Dark2") +
    labs(
        x = NULL,
        y = "Blocked RMSE",
        colour = "Assessment",
        title = "Predictive performance depends on programme and validation target"
    ) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r selected-table}
selection_table <- selection |>
    mutate(
        Programme = recode(programme_key, !!!programme_labels),
        Model = recode(model, !!!model_labels)
    ) |>
    arrange(Programme, selection_rmse) |>
    select(
        Programme, Model, complexity, selection_rmse,
        selection_mae, within_two_percent, selected_for_prediction
    ) |>
    mutate(
        across(c(selection_rmse, selection_mae), ~ round(.x, 3))
    )

kable(
    selection_table,
    col.names = c(
        "Programme", "Candidate", "Complexity", "Selection RMSE",
        "Selection MAE", "Within 2%", "Selected benchmark"
    )
)
```

# Event-to-event transfer

Leave-one-event-out performance is the most relevant diagnostic for predicting
a future bleaching event. The figure below gives each event its own RMSE rather
than averaging away difficult years.

```{r event-rmse, fig.width=11, fig.height=7}
event_metrics <- fold_metrics_plot |>
    filter(scheme == "leave_one_event_out") |>
    mutate(Event = factor(fold, levels = sort(unique(as.integer(fold)))))

ggplot(event_metrics, aes(Event, rmse, colour = Model, group = Model)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2.4) +
    facet_wrap(~Programme, scales = "free_y") +
    scale_colour_brewer(palette = "Set1") +
    labs(
        x = "Held-out bleaching event",
        y = "RMSE",
        colour = "Candidate",
        title = "No candidate transfers uniformly across bleaching events"
    )
```

## The 2024 stress test

The widespread 2024 event is retained as a transparent out-of-event test and
then included in production fitting. The benchmark models trained on earlier
events underpredict mean 2024 mortality, especially for manta and MMP.

```{r event-2024-bars, fig.width=9, fig.height=5.5}
event_2024 <- selected_predictions |>
    filter(scheme == "leave_one_event_out", event_year == 2024) |>
    group_by(Programme, Model) |>
    summarise(
        n = n(),
        Observed = mean(observed_mortality),
        Predicted = mean(predicted_mortality),
        RMSE = sqrt(mean((observed_mortality - predicted_mortality)^2)),
        .groups = "drop"
    )

event_2024 |>
    pivot_longer(c(Observed, Predicted), names_to = "Series", values_to = "Mortality") |>
    ggplot(aes(Programme, Mortality, fill = Series)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.65) +
    geom_text(
        aes(label = percent(Mortality, accuracy = 0.1)),
        position = position_dodge(width = 0.75), vjust = -0.35, size = 3.5
    ) +
    scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.12))) +
    scale_fill_manual(values = c(Observed = "#C44E52", Predicted = "#4C72B0")) +
    labs(
        x = NULL, y = "Mean mortality", fill = NULL,
        title = "Earlier events do not fully anticipate 2024 mortality"
    )
```

```{r event-2024-table}
kable(
    event_2024 |>
        mutate(across(c(Observed, Predicted, RMSE), ~ round(.x, 3))) |>
        select(Programme, Model, n, Observed, Predicted, RMSE),
    col.names = c("Programme", "Benchmark", "N", "Observed mean", "Predicted mean", "RMSE")
)
```

# Out-of-fold prediction diagnostics

```{r observed-predicted, fig.width=10, fig.height=7}
selected_event_predictions <- selected_predictions |>
    filter(scheme == "leave_one_event_out")

ggplot(
    selected_event_predictions,
    aes(observed_mortality, predicted_mortality, colour = factor(event_year))
) +
    geom_abline(slope = 1, intercept = 0, colour = "grey35", linetype = "dashed") +
    geom_point(alpha = 0.62, size = 1.8) +
    facet_wrap(~Programme) +
    coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
    scale_x_continuous(labels = percent) +
    scale_y_continuous(labels = percent) +
    scale_colour_viridis_d(end = 0.9) +
    labs(
        x = "Observed mortality",
        y = "Out-of-event predicted mortality",
        colour = "Event",
        title = "Blocked predictions are compressed toward the mean"
    )
```

```{r dhw-predictions, fig.width=11, fig.height=7}
ggplot(
    selected_event_predictions,
    aes(ann_maxdhw, observed_mortality, colour = factor(event_year))
) +
    geom_point(alpha = 0.42, size = 1.5) +
    geom_point(
        aes(y = predicted_mortality),
        shape = 1, alpha = 0.75, size = 1.7
    ) +
    facet_wrap(~Programme) +
    scale_y_continuous(labels = percent, limits = c(0, 1)) +
    scale_colour_viridis_d(end = 0.9) +
    labs(
        x = "Annual maximum DHW (°C-weeks)",
        y = "Mortality",
        colour = "Held-out event",
        title = "Filled points are observed; open points are out-of-event predictions"
    )
```

# What the benchmark does—and does not—tell us

- DHW contains transferable signal, but it is not sufficient to reproduce the
  mortality distribution of an unprecedented event.
- Detailed LTMP benthic observations gain predictive accuracy from physical
  context, supporting continued environmental modelling.
- MMP cannot support a large standalone environmental model reliably because
  it has only three event years and nearly half of Secchi values are missing.
- Selection of a linear benchmark for manta or MMP is a statement about the
  current small candidate comparison and validation variance. It is not a
  mechanistic dose-response conclusion.

```{r warning-table}
if (nrow(fit_warnings) > 0) {
    warning_display <- fit_warnings |>
        mutate(
            Programme = recode(programme_key, !!!programme_labels),
            Candidate = recode(model, !!!model_labels),
            Assessment = recode(scheme, !!!scheme_labels)
        ) |>
        select(Programme, Assessment, fold, Candidate, warning)
    kable(warning_display)
}
```

# Production prediction architecture

The next comparison should retain the benchmark models but make prediction—not
parsimony—the objective:

1. **BRMS base learner:** a hierarchical boundary-aware model with nonlinear
   DHW, environmental modifiers, event effects, reef/region structure, and full
   posterior predictive uncertainty.
2. **BRT base learner:** a regularised nonlinear learner using the complete
   attainable predictor set, with depth and tree number tuned inside each
   analysis fold.
3. **Cross-fitted ensemble:** obtain BRMS and BRT predictions for every blocked
   assessment row, estimate non-negative ensemble weights without using that
   row's outcome, and compare the ensemble against each base learner and the
   benchmark.
4. **Full-data refit:** once the architecture and ensemble rule are frozen,
   refit BRMS and BRT on all events including 2024, then produce GBR-wide means,
   intervals, and environmental-extrapolation flags.

The extensive cached BRMS and BRT work in `output/models` has not been removed.
It remains valuable for model specification and diagnostics, but it must be
refitted against the canonical outcomes, reconstructed cloud values, and fixed
blocked folds before contributing to the production ensemble.
