---
title: 'GBR mortality prediction uncertainty and extrapolation contract'
subtitle: 'Cross-fitted intervals, severe-event guardrails, and 2024 environmental support'
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)

output_dir <- '../output/prediction_uncertainty'
files <- c(
  intervals = file.path(
    output_dir, 'selected_route_cross_calibrated_intervals.csv'
  ),
  coverage = file.path(
    output_dir, 'selected_route_interval_coverage.csv'
  ),
  brms = file.path(output_dir, 'brms_expected_interval_coverage.csv'),
  support = file.path(output_dir, 'environmental_support_2024.csv'),
  predictors = file.path(
    output_dir, 'environmental_support_predictor_summary.csv'
  )
)
if (any(!file.exists(files))) {
  stop('Run scripts/quantify_prediction_uncertainty.R first.')
}

intervals <- read_csv(files[['intervals']], show_col_types = FALSE)
coverage <- read_csv(files[['coverage']], show_col_types = FALSE)
brms_coverage <- read_csv(files[['brms']], show_col_types = FALSE)
support <- read_csv(files[['support']], show_col_types = FALSE)
predictor_support <- read_csv(files[['predictors']], show_col_types = FALSE)

programme_labels <- c(
  ltmp = 'LTMP benthic',
  manta = 'LTMP manta tow',
  mmp = 'MMP inshore'
)
scheme_labels <- c(
  leave_one_event_out = 'Leave one event out',
  reef_blocked_5fold = 'Reef blocked'
)
```

# Decision

The pipeline now distinguishes four things that should not be collapsed into a
single confidence band:

1. posterior uncertainty in the BRMS expected response;
2. disagreement between BRMS and BRT;
3. held-out residual/process error in the selected programme route; and
4. environmental and community-input support at the prediction location.

The saved BRMS Q05--Q95 limits came from `posterior_epred`; they are intervals
for expected mortality, not observation-level prediction intervals. Cross-
fitted residual intervals are therefore the current predictive uncertainty
layer. Even these remain marginal: their severe-event coverage is reported
separately and cannot be inferred from overall coverage.

A full GBR mortality surface is **not released yet**. Every 2024 registry row
lacks a complete, audited pre-event community layer for the model predictors
`prop_acropora_pre` and `acropora_cover_pre`; MMP additionally requires depth.
Environmental support maps below are diagnostics, not mortality maps.

# Expected-response intervals are too narrow

```{r brms-table}
brms_coverage |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Assessment = recode(scheme, !!!scheme_labels),
    `All-row coverage` = percent(expected_interval_coverage, accuracy = 0.1),
    `Severe-row coverage` = percent(
      severe_expected_interval_coverage, accuracy = 0.1
    ),
    `Mean width` = round(mean_expected_interval_width, 3)
  ) |>
  select(
    Programme, Assessment, n,
    `All-row coverage`, severe_n, `Severe-row coverage`, `Mean width`
  ) |>
  kable(
    caption = paste(
      'Empirical coverage of nominal BRMS 90% intervals for expected mortality.',
      'These are not posterior predictive intervals.'
    )
  )
```

Low coverage is expected when parameter uncertainty is presented as if it also
contained observation noise, unmeasured event mechanisms, and model-form error.
Renaming these limits is not cosmetic; it prevents overconfident maps.

# Cross-fitted predictive intervals

For each outer assessment fold, the interval radius is the finite-sample 90th
percentile of absolute residuals from *other* outer folds of the same programme
and validation scheme. No assessment outcome calibrates its own interval.

The severe upper guardrail is calibrated the same way using only mortality of
at least 20% in the other folds. It is intentionally conservative and should be
shown as a high-loss risk bound, not as another central interval.

```{r calibrated-table}
coverage |>
  filter(
    scheme == 'leave_one_event_out',
    subgroup %in% c('all', 'positive', 'severe')
  ) |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Subgroup = recode(
      subgroup,
      all = 'All observations',
      positive = 'Positive mortality',
      severe = 'Mortality >= 20%'
    ),
    `Marginal coverage` = percent(empirical_coverage, accuracy = 0.1),
    `Severe upper guard coverage` = percent(
      severe_upper_guard_coverage, accuracy = 0.1
    ),
    `Mean interval width` = round(mean_interval_width, 3),
    `Mean BRMS-BRT disagreement` = round(mean_model_disagreement, 3)
  ) |>
  select(
    Programme, Subgroup, n, `Marginal coverage`,
    `Severe upper guard coverage`, `Mean interval width`,
    `Mean BRMS-BRT disagreement`
  ) |>
  kable(caption = 'Event-held-out uncertainty performance')
```

```{r calibrated-plot, fig.width=10, fig.height=5.5}
intervals |>
  filter(scheme == 'leave_one_event_out') |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    severe = observed_mortality >= 0.2
  ) |>
  ggplot(aes(x = predicted_mortality, y = observed_mortality)) +
  geom_segment(
    aes(
      x = interval_lower_90, xend = interval_upper_90,
      y = observed_mortality, yend = observed_mortality
    ),
    alpha = 0.18, colour = '#4393C3'
  ) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = 'grey55') +
  geom_point(aes(colour = severe), alpha = 0.7, size = 1.3) +
  facet_wrap(~ Programme) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  scale_colour_manual(
    values = c('FALSE' = 'grey45', 'TRUE' = '#B2182B'),
    labels = c('No', 'Yes')
  ) +
  labs(
    x = 'Selected-route prediction',
    y = 'Observed mortality',
    colour = 'Observed >= 20%',
    title = 'Marginal intervals still under-cover severe event outcomes'
  )
```

The event-held-out results are the relevant test for a future or unusually
intense bleaching event. Reef-blocked coverage is generally better because it
interpolates within events already represented in training. MMP temporal
guardrail failure is retained as a formal limitation; three event years are not
enough to calibrate the 2024 severity shift from previous MMP events.

# Framework disagreement

```{r disagreement-plot, fig.width=10, fig.height=4.8}
intervals |>
  filter(scheme == 'leave_one_event_out') |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    absolute_error = abs(observed_mortality - predicted_mortality)
  ) |>
  ggplot(aes(x = model_disagreement, y = absolute_error)) +
  geom_point(alpha = 0.35, size = 1.1, colour = '#762A83') +
  geom_smooth(method = 'loess', se = FALSE, colour = 'black') +
  facet_wrap(~ Programme, scales = 'free_x') +
  labs(
    x = 'Absolute BRMS-BRT prediction disagreement',
    y = 'Absolute held-out error',
    title = 'Framework disagreement is a local epistemic warning, not an interval'
  )
```

Disagreement should accompany future prediction rows, but it is not a complete
uncertainty measure: learners can agree and still share the same missing event
mechanism.

# 2024 environmental support

Environmental support uses the 15 predictors available consistently in the
full registry. For each programme, values are standardised with its formal
training rows. A nearest-training-row distance is compared with leave-one-row-
out 95th and 99th percentiles, while individual predictors are also checked
against training ranges.

```{r support-table}
support |>
  count(programme_key, environmental_support_class, name = 'Features') |>
  group_by(programme_key) |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    Percent = percent(Features / sum(Features), accuracy = 0.1)
  ) |>
  ungroup() |>
  select(Programme, `Environmental support` = environmental_support_class,
         Features, Percent) |>
  arrange(Programme, `Environmental support`) |>
  kable()
```

```{r support-map, fig.width=10, fig.height=7}
support |>
  mutate(
    Programme = recode(programme_key, !!!programme_labels),
    environmental_support_class = factor(
      environmental_support_class,
      levels = c(
        'within environmental support', 'caution',
        'extreme extrapolation', 'missing environmental input'
      )
    )
  ) |>
  ggplot(aes(x = lon, y = lat, colour = environmental_support_class)) +
  geom_point(size = 0.65, alpha = 0.85) +
  coord_equal() +
  facet_wrap(~ Programme, nrow = 1) +
  scale_colour_manual(values = c(
    'within environmental support' = '#1B7837',
    'caution' = '#FDB863',
    'extreme extrapolation' = '#B2182B',
    'missing environmental input' = 'grey30'
  )) +
  labs(
    x = 'Longitude', y = 'Latitude', colour = 'Support class',
    title = 'A model can be well calibrated at survey reefs and unsupported elsewhere'
  ) +
  theme(legend.position = 'bottom')
```

MMP is an inshore programme with a narrow environmental training domain; only
18 of 7,063 registry features lie within its environmental support. MMP should
therefore not be extrapolated as a general GBR-wide surface. It is a distinct
inshore prediction target.

```{r predictor-table}
predictor_support |>
  mutate(
    outside_features = below_training_range + above_training_range,
    Programme = recode(programme_key, !!!programme_labels)
  ) |>
  group_by(programme_key) |>
  slice_max(outside_features, n = 5, with_ties = FALSE) |>
  ungroup() |>
  select(
    Programme, Predictor = predictor,
    `Outside training range` = outside_features,
    `Missing features` = missing_features,
    `Training minimum` = training_min,
    `Training maximum` = training_max
  ) |>
  mutate(across(c(`Training minimum`, `Training maximum`), ~ round(.x, 3))) |>
  kable(caption = 'Largest univariate support gaps in the 2024 registry')
```

# Production contract

Before a reef-level prediction is publishable, it must carry:

- the BRMS and BRT component predictions and their absolute disagreement;
- the selected programme route and fixed cross-fitted production weights;
- a cross-calibrated marginal interval and severe-event upper guardrail;
- environmental support class and the predictors outside training range;
- provenance and uncertainty for interpolated pre-event Acropora and total
  coral cover, plus depth for MMP; and
- a suppression flag when required inputs are missing or the target is outside
  the intended programme domain.

The immediate implementation gap is the full-registry community layer. The old
Acropora interpolation supplies a proportion but does not provide a fully
audited, past-only absolute pre-event Acropora cover for all registry features.
Using a median community value would produce a clean-looking map while removing
one of the most influential ecological predictors. The pipeline therefore stops
at environmental support rather than silently making that substitution.
