---
title: 'Where do the mortality models fail?'
subtitle: 'Formal event-held-out reef-year residual assessment'
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)
library(tidyr)

input_dir <- '../output/prediction_outliers'
primary <- read_csv(
  file.path(input_dir, 'formal_ensemble_residuals.csv'),
  show_col_types = FALSE
)
priority <- read_csv(
  file.path(input_dir, 'priority_reef_years.csv'),
  show_col_types = FALSE
)
event_summary <- read_csv(
  file.path(input_dir, 'event_outlier_summary.csv'),
  show_col_types = FALSE
)
event_weather <- read_csv(
  '../data/processed/era5_weather_reef_year.csv',
  show_col_types = FALSE
) |>
  group_by(year) |>
  summarise(
    `Median coastal Q1 rain` = median(era5_coastal_rain_q1_total),
    `Median max 30d rain` = median(era5_coastal_rain_dec_mar_max_30day),
    `Median reef wind` = median(era5_reef_wind_q1_mean),
    `Median calm fraction` = median(era5_reef_wind_q1_fraction_below_3),
    .groups = 'drop'
  )
family_comparison <- read_csv(
  file.path(input_dir, 'beta_binomial_family_comparison.csv'),
  show_col_types = FALSE
)
binomial_diagnostics <- read_csv(
  '../output/formal_models/binomial_diagnostics.csv',
  show_col_types = FALSE
)
family_rows <- read_csv(
  file.path(input_dir, 'all_cross_fitted_residuals.csv'),
  show_col_types = FALSE
) |>
  filter(learner %in% c('brms', 'brms_binomial_olre')) |>
  mutate(
    family = recode(
      learner,
      brms = 'Primary beta-family BRMS',
      brms_binomial_olre = 'Binomial-OLRE sensitivity'
    )
  )
family_overall <- family_rows |>
  group_by(programme_key, family) |>
  summarise(
    n = n(),
    MAE = mean(absolute_error),
    RMSE = sqrt(mean(residual^2)),
    `Predictive R2` = 1 - sum(residual^2) /
      sum((observed_mortality - mean(observed_mortality))^2),
    `Mean observed - predicted` = mean(residual),
    `Severe-loss MAE` = mean(
      absolute_error[observed_mortality >= 0.2], na.rm = TRUE
    ),
    `Low-DHW observed - predicted` = mean(
      residual[ann_maxdhw < 4], na.rm = TRUE
    ),
    .groups = 'drop'
  )

programme_labels <- c(
  ltmp = 'LTMP benthic', manta = 'Manta tow', mmp = 'MMP inshore'
)
primary <- primary |>
  mutate(
    programme = factor(
      programme_labels[programme_key], levels = unname(programme_labels)
    ),
    direction = if_else(
      residual >= 0, 'Underestimated mortality', 'Overestimated mortality'
    )
  )
priority <- priority |>
  mutate(programme = programme_labels[programme_key])
```

# Interpretation

Residuals are calculated from predictions made when the reef-year's entire
bleaching event was excluded from model fitting. A positive residual means the
model underestimated mortality; a negative residual means it overestimated
mortality. This is intentionally stricter than examining fitted residuals.

For each programme, learner, and held-out event, the 5th and 95th percentile
thresholds come only from cross-fitted residuals for the other events. A row is
therefore unusual relative to errors the same model made in other bleaching
events. Flags identify cases for ecological diagnosis and do **not** remove
them from model fitting.

# Event calibration

```{r event-calibration-table}
event_summary |>
  mutate(
    programme = programme_labels[programme_key],
    across(
      c(
        observed_mean, predicted_mean, bias_observed_minus_predicted, mae,
        soi_dec_mar_mean
      ),
      ~ round(.x, 3)
    ),
    roni_bleaching_summer_mean = round(roni_bleaching_summer_mean, 2),
    soi_dec_mar_mean = round(soi_dec_mar_mean, 1)
  ) |>
  select(
    Programme = programme, Event = event_year, n,
    `Unique reefs` = unique_reefs,
    ENSO = enso_category,
    RONI = roni_bleaching_summer_mean,
    SOI = soi_dec_mar_mean,
    `Observed mean` = observed_mean,
    `Predicted mean` = predicted_mean,
    `Observed - predicted` = bias_observed_minus_predicted,
    MAE = mae,
    `Underestimated outliers` = underestimated_outliers,
    `Overestimated outliers` = overestimated_outliers,
    `Severe 2024 misses` = severe_2024_underpredictions,
    `Severe 2024 reefs` = severe_2024_unique_reefs,
    `Low-DHW overpredictions` = low_dhw_overpredictions,
    `Low-DHW reefs` = low_dhw_unique_reefs
  ) |>
  kable(caption = 'Event-held-out calibration of the formal BRMS--BRT ensemble')
```

```{r enso-event-error, fig.width=9.5, fig.height=5.5}
event_summary |>
  mutate(programme = programme_labels[programme_key]) |>
  ggplot(
    aes(
      roni_bleaching_summer_mean,
      bias_observed_minus_predicted,
      colour = factor(event_year)
    )
  ) +
  geom_hline(yintercept = 0, colour = 'grey40') +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = 3, colour = 'grey60') +
  geom_point(size = 3) +
  geom_text(aes(label = event_year), nudge_y = 0.012, show.legend = FALSE) +
  facet_wrap(~ programme) +
  scale_colour_viridis_d(option = 'C', guide = 'none') +
  labs(
    x = 'Mean DJF--FMA Relative Oceanic Nino Index',
    y = 'Event mean observed - predicted mortality',
    title = 'ENSO state and held-out event calibration',
    subtitle = paste(
      'Dashed lines mark +/-0.5 phase thresholds;',
      'no trend is fitted to only five events'
    )
  )
```

The event sequence does not reduce to one ENSO gradient: 2020 is neutral and
2022 is Moderate La Nina, although both have lower mortality than predicted,
while 2024 is Weak El Nino with much higher-than-predicted mortality in manta
and MMP. RONI is therefore a candidate context modifier, not a replacement for
local rainfall, wind, community composition, or thermal history.

```{r enso-weather-context}
event_summary |>
  distinct(
    event_year, enso_category, roni_bleaching_summer_mean, soi_dec_mar_mean
  ) |>
  left_join(event_weather, by = c('event_year' = 'year')) |>
  mutate(
    across(where(is.numeric), ~ round(.x, 2)),
    `Median calm fraction` = percent(`Median calm fraction`, accuracy = 1)
  ) |>
  select(
    Event = event_year, ENSO = enso_category,
    RONI = roni_bleaching_summer_mean, SOI = soi_dec_mar_mean,
    everything()
  ) |>
  kable(caption = 'ENSO and GBR-wide ERA5 weather context by event')
```

GBR-wide rainfall is not a monotonic proxy for ENSO in these five events. For
example, 2022 is the only Moderate La Nina bleaching summer, but its median Q1
coastal rainfall is lower than in neutral 2017 and 2020 and Weak El Nino 2024.
This supports keeping local rainfall and basin-scale ENSO as distinct candidate
variables rather than using one to stand in for the other.

```{r observed-predicted, fig.width=10, fig.height=5.5}
ggplot(primary, aes(predicted_mortality, observed_mortality)) +
  geom_abline(slope = 1, intercept = 0, colour = 'grey40', linetype = 2) +
  geom_point(aes(colour = factor(event_year)), alpha = 0.7, size = 1.8) +
  facet_wrap(~ programme) +
  coord_equal(xlim = c(0, 0.8), ylim = c(0, 0.8)) +
  scale_colour_viridis_d(option = 'C', name = 'Event') +
  labs(
    x = 'Event-held-out predicted mortality',
    y = 'Observed mortality',
    title = 'Points above the diagonal are underestimated'
  )
```

```{r spatial-2024-errors, fig.width=10, fig.height=6.5}
primary |>
  filter(event_year == 2024) |>
  ggplot(aes(lon, lat)) +
  geom_point(
    aes(colour = residual, size = observed_mortality),
    alpha = 0.78
  ) +
  facet_wrap(~ programme) +
  coord_quickmap() +
  scale_colour_gradient2(
    low = '#2166AC', mid = 'grey90', high = '#B2182B', midpoint = 0,
    name = 'Observed -\npredicted'
  ) +
  scale_size_continuous(
    range = c(1.2, 5.5), labels = label_percent(), name = 'Observed\nmortality'
  ) +
  labs(
    x = 'Longitude', y = 'Latitude',
    title = 'Spatial pattern of held-out 2024 errors',
    subtitle = 'Red sites have more mortality than the model predicted'
  )
```

# Does error depend on DHW?

```{r residual-dhw, fig.width=10, fig.height=6}
ggplot(primary, aes(ann_maxdhw, residual)) +
  geom_hline(yintercept = 0, colour = 'grey35') +
  geom_hline(yintercept = c(-0.1, 0.1), linetype = 3, colour = 'grey55') +
  geom_vline(xintercept = 4, linetype = 2, colour = '#2166AC') +
  geom_point(aes(colour = factor(event_year)), alpha = 0.72, size = 1.8) +
  facet_wrap(~ programme, scales = 'free_x') +
  scale_colour_viridis_d(option = 'C', name = 'Event') +
  labs(
    x = 'Annual maximum DHW',
    y = 'Observed - predicted mortality',
    title = 'Signed prediction errors across thermal stress',
    subtitle = 'Positive: mortality underestimated; dashed vertical line: 4 DHW'
  )
```

# Priority ecological diagnoses

Two prespecified flags target the concerns motivating this assessment:

- **Severe 2024 miss:** observed mortality at least 20%, underestimated by at
  least 10 percentage points and beyond the other-event upper residual tail.
- **Low-DHW overprediction:** DHW below 4 and mortality overestimated by at
  least 10 percentage points and beyond the other-event lower residual tail.

```{r priority-counts}
priority |>
  summarise(
    `Severe 2024 reef-years` = sum(severe_2024_underprediction),
    `Low-DHW reef-years` = sum(low_dhw_overprediction),
    `All core models underestimate` = sum(
      core_model_consensus == 'all_three_underestimate'
    ),
    `All core models overestimate` = sum(
      core_model_consensus == 'all_three_overestimate'
    )
  ) |>
  pivot_longer(everything(), names_to = 'Flag', values_to = 'Reef-years') |>
  kable()
```

```{r severe-table}
priority |>
  filter(severe_2024_underprediction) |>
  arrange(desc(residual)) |>
  transmute(
    Programme = programme,
    Reef = ReefName,
    Region = region_block,
    DHW = round(ann_maxdhw, 1),
    Observed = round(observed_mortality, 3),
    Predicted = round(predicted_mortality, 3),
    `Observed - predicted` = round(residual, 3),
    `% Acropora before` = percent(prop_acropora_pre, accuracy = 0.1),
    `Freshwater-risk percentile` = percent(
      freshwater_risk30_percentile, accuracy = 0.1
    ),
    `Partial eReefs <30 PSU` = partial_ereefs_flood30,
    `Coastal rain max 30d` = round(era5_coastal_rain_max_30day),
    `ERA5 mean wind` = round(era5_wind_mean, 2),
    `Calm-day fraction` = percent(era5_wind_calm_fraction, accuracy = 1),
    Disturbance = DISTURBANCE_TYPE,
    `Core-model agreement` = core_model_consensus
  ) |>
  head(30) |>
  kable(caption = 'Largest severe-mortality underpredictions in 2024')
```

```{r severe-region-context}
priority |>
  filter(severe_2024_underprediction) |>
  group_by(programme, region_block) |>
  summarise(
    `Reef-years` = n(),
    Observed = mean(observed_mortality),
    Predicted = mean(predicted_mortality),
    DHW = mean(ann_maxdhw),
    `Pre-event Acropora` = mean(prop_acropora_pre, na.rm = TRUE),
    `Coastal rain max 30d` = mean(
      era5_coastal_rain_max_30day, na.rm = TRUE
    ),
    `Mean wind` = mean(era5_wind_mean, na.rm = TRUE),
    .groups = 'drop'
  ) |>
  mutate(
    across(c(Observed, Predicted, DHW, `Pre-event Acropora`), ~ round(.x, 3)),
    `Coastal rain max 30d` = round(`Coastal rain max 30d`),
    `Mean wind` = round(`Mean wind`, 2)
  ) |>
  kable(caption = 'Environmental context of severe 2024 misses by region')
```

These summaries separate two candidate failure modes without treating either
as proven causation: northern and central misses coincide with high coastal
rainfall, whereas southern misses combine higher DHW with high pre-event
Acropora composition. Both mechanisms should therefore be tested in the next
blocked refit rather than represented by a single steeper DHW curve.

```{r low-dhw-table}
priority |>
  filter(low_dhw_overprediction) |>
  arrange(residual) |>
  transmute(
    Programme = programme,
    Event = event_year,
    Reef = ReefName,
    Region = region_block,
    DHW = round(ann_maxdhw, 1),
    Observed = round(observed_mortality, 3),
    Predicted = round(predicted_mortality, 3),
    `Observed - predicted` = round(residual, 3),
    `% Acropora before` = percent(prop_acropora_pre, accuracy = 0.1),
    `Coastal rain max 30d` = round(era5_coastal_rain_max_30day),
    `ERA5 mean wind` = round(era5_wind_mean, 2),
    `Calm-day fraction` = percent(era5_wind_calm_fraction, accuracy = 1),
    Disturbance = DISTURBANCE_TYPE,
    `Core-model agreement` = core_model_consensus
  ) |>
  head(30) |>
  kable(caption = 'Largest overpredictions below 4 DHW')
```

The strict large-error rule identifies few individual low-DHW reefs, but the
event-calibration table shows a broader smaller bias: the ensemble overpredicts
mean mortality in the low-mortality 2020 and 2022 events for LTMP and manta,
and in the very small 2022 MMP sample. This supports testing a sharper
occurrence boundary or lower background component, while keeping that task
separate from the missing severe-2024 mechanism.

The December--March SOI separates the La Nina-like 2022 event from neutral 2020,
so it cannot by itself explain why both events had lower mortality than
predicted. It remains a useful atmospheric cross-check because SOI reflects the
circulation relevant to Australian rainfall. In 2024, mean summer RONI is
El Nino while mean SOI is neutral/mixed, so oceanic and atmospheric indicators
should not be treated as interchangeable.

# Complete hand-off table

The machine-readable file
`output/prediction_outliers/priority_reef_years.csv` contains the full context,
including descriptions, pre-event cover, Acropora composition, individual BRMS
and BRT residuals, and model-consensus flags. It is the appropriate table for
adding hypothesised ecological mechanisms such as freshwater exposure,
community susceptibility, unrecorded disturbance, or unusual prior exposure.
The companion `priority_observation_rows.csv` retains every contributing survey
row; `priority_reef_years.csv` selects the largest absolute error within each
programme--reef--event and records its number of flagged observations.

# Beta versus binomial response family

The binomial comparison uses 100 pseudo-trials solely to put the estimated
proportions on a common binomial scale. An observation-level random effect
absorbs extra-binomial variation. It is therefore a response-family
sensitivity, not a claim that every survey observed exactly 100 independent
colonies.

```{r binomial-diagnostics}
failed_binomial_diagnostics <- binomial_diagnostics |>
  filter(
    max_rhat > 1.01 | min_bulk_ess < 100 | min_tail_ess < 100 |
      divergences > 0 | max_treedepth_hits > 0
  )

failed_binomial_diagnostics |>
  mutate(across(c(max_rhat, min_bulk_ess, min_tail_ess), ~ round(.x, 3))) |>
  select(
    Programme = programme_key, Event = fold,
    `Maximum R-hat` = max_rhat,
    `Minimum bulk ESS` = min_bulk_ess,
    `Minimum tail ESS` = min_tail_ess,
    Divergences = divergences,
    `Tree-depth hits` = max_treedepth_hits
  ) |>
  kable(caption = 'Binomial folds that fail the prespecified diagnostic gate')

if (nrow(failed_binomial_diagnostics) == 0) {
  cat('All binomial sensitivity folds pass the diagnostic gate.')
} else {
  cat(paste(
    'The binomial sensitivity is not eligible for model promotion because',
    'at least one held-out fold remains diagnostically unresolved.'
  ))
}
```

```{r family-comparison}
family_overall |>
  mutate(
    Programme = programme_labels[programme_key],
    across(where(is.numeric), ~ round(.x, 3))
  ) |>
  select(Programme, Family = family, n, everything(), -programme_key) |>
  kable(caption = 'Overall event-held-out response-family comparison')

family_comparison |>
  mutate(
    Programme = programme_labels[programme_key],
    Family = recode(
      learner,
      brms = 'Primary beta-family BRMS',
      brms_binomial_olre = 'Binomial-OLRE sensitivity'
    ),
    across(
      c(
        observed_mean, predicted_mean, bias_observed_minus_predicted,
        mae, rmse, severe_mae, low_dhw_bias
      ),
      ~ round(.x, 3)
    )
  ) |>
  select(
    Programme, Family, Event = event_year,
    `Observed mean` = observed_mean,
    `Predicted mean` = predicted_mean,
    `Observed - predicted` = bias_observed_minus_predicted,
    MAE = mae, RMSE = rmse, `Severe-loss MAE` = severe_mae,
    `Low-DHW observed - predicted` = low_dhw_bias
  ) |>
  kable(caption = 'Like-for-like event-held-out family sensitivity')
```

```{r family-2024-summary}
family_comparison |>
  filter(event_year == 2024) |>
  mutate(
    Programme = programme_labels[programme_key],
    Family = recode(
      learner,
      brms = 'Primary beta-family BRMS',
      brms_binomial_olre = 'Binomial-OLRE sensitivity'
    ),
    across(
      c(observed_mean, predicted_mean, bias_observed_minus_predicted, rmse),
      ~ round(.x, 3)
    )
  ) |>
  select(
    Programme, Family, `Observed 2024` = observed_mean,
    `Predicted 2024` = predicted_mean,
    `Observed - predicted` = bias_observed_minus_predicted,
    RMSE = rmse
  ) |>
  kable(caption = 'Direct comparison for the held-out 2024 event')
```

```{r family-residual-dhw, fig.width=10, fig.height=6.5}
ggplot(
  family_rows,
  aes(ann_maxdhw, residual, colour = family)
) +
  geom_hline(yintercept = 0, colour = 'grey35') +
  geom_vline(xintercept = 4, linetype = 2, colour = 'grey55') +
  geom_point(alpha = 0.18, size = 1) +
  geom_smooth(method = 'loess', se = TRUE, linewidth = 0.9) +
  facet_grid(programme_key ~ event_year, scales = 'free_x') +
  scale_colour_manual(values = c('#2166AC', '#B2182B')) +
  labs(
    x = 'Annual maximum DHW',
    y = 'Observed - predicted mortality',
    colour = 'Response family',
    title = 'Response-family errors across DHW and held-out events',
    subtitle = paste(
      'Positive values are underprediction; smooths are diagnostics,',
      'not fitted ecological response curves'
    )
  )
```

Because the response is loss relative to estimated pre-event coral cover, not
a count of independent dead colonies, the binomial model cannot use a literal
survey denominator. Its 100 pseudo-trials and observation-level random effect
make this a useful shape sensitivity only. Model promotion still depends on
held-out predictive performance and clean sampling diagnostics. The current
manta-2022 fold has no divergences or tree-depth hits, but its reef-level
standard-deviation parameter remains above the R-hat gate after stricter
sampling; binomial results are therefore interpreted but not promoted.
