---
title: 'RRN audit of five severe 2024 mortality misses'
subtitle: 'Mackay, Linnet, Lizard Island NW, Eyrie and Swinger'
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)
library(tidyr)

input_dir <- '../output/rrn_target_reefs'
context <- read_csv(
  file.path(input_dir, 'target_reef_2024_context.csv'), show_col_types = FALSE
)
pressure <- read_csv(
  file.path(input_dir, 'target_reef_2024_pressure_percentiles.csv'),
  show_col_types = FALSE
)
pressure_history <- read_csv(
  file.path(input_dir, 'target_reef_all_individual_pressures.csv'),
  show_col_types = FALSE
)
metrics <- read_csv(
  file.path(input_dir, 'target_reef_metric_percentiles.csv'),
  show_col_types = FALSE
)
associations <- read_csv(
  file.path(input_dir, 'manta_2024_rrn_residual_associations.csv'),
  show_col_types = FALSE
)
acropora_scenarios <- read_csv(
  file.path(input_dir, 'manta_acropora_counterfactual_predictions.csv'),
  show_col_types = FALSE
)
acropora_summary <- read_csv(
  file.path(input_dir, 'manta_acropora_counterfactual_summary.csv'),
  show_col_types = FALSE
)
coverage <- read_csv(
  file.path(input_dir, 'extraction_coverage.csv'), show_col_types = FALSE
)

pressure_labels <- c(
  cot_meanpertow = 'Observed COTS',
  cot_idwmeanpertow = 'IDW COTS',
  cyc_maxHrs4mw = 'Cyclone waves',
  sst_maxdhw = 'RRN DHW',
  uq_coralsink1 = 'Low coral-sink connectivity',
  wqc_freqcc12 = 'Coloured water'
)
reef_order <- c('Linnet', 'Mackay', 'Eyrie', 'Lizard NW', 'Swinger')
```

# Main finding

The five reefs do not share one RRN explanation.

- **Linnet and Eyrie have strong low-level COTS signals.** Their observed and
  spatially modelled COTS values are unusually high relative to reefs in the
  same one-degree latitude band. These values should be treated as evidence of
  elevated COTS pressure, not automatically labelled a formal outbreak.
- **Linnet, Eyrie and Swinger experienced unusual coloured-water exposure
  relative to their own histories.** Each 2023--24 value is at the 95th
  percentile of its previous decade. Water colour also aligns positively with
  underprediction across the full set of 2024 manta reefs.
- **Lizard Island NW has very low coral-sink connectivity.** This suggests low
  connectivity-mediated rescue or recovery potential, although it is not an
  acute mortality agent and cannot by itself explain rapid post-bleaching loss.
- **Mackay is unexplained by the gridded RRN predictors, but not by its field
  record.** It has no modelled cyclone-wave exposure, no observed COTS, modest
  water colour and exceptionally low DHW relative to nearby reefs. However,
  its retained survey tooltip explicitly records coral bleaching, Cyclone
  Jasper and flood waters. The remaining miss therefore points to inadequate
  spatial or temporal resolution in the gridded freshwater/cyclone proxies.

The misses are **not simply caused by a missing Acropora column**. The model has
a preceding-year Acropora proportion for all five reefs. Linnet, Lizard NW and
Mackay use observed reef/depth-year composition; only Eyrie and Swinger use
past-only spatial-temporal interpolation. The earlier assignment of Mackay to
an interpolated 42.2% Acropora value was a crosswalk error: its corrected
observed value is 11.1%. Setting Acropora to 90% in each reef's original
held-out BRT still leaves 23--44 percentage points of underprediction.

# Is missing community composition the main cause?

```{r acropora-provenance}
acropora_summary |>
  mutate(
    Reef = sub(' Reef.*$', '', ReefName),
    Source = recode(
      acropora_source,
      observed_depth_year = 'Observed depth/year',
      observed_reef_year = 'Observed reef/year',
      past_only_spatiotemporal_idw = 'Past-only spatial-temporal IDW'
    ),
    `Assigned Acropora` = percent(
      assigned_prop_acropora_pre, accuracy = 0.1
    ),
    Observed = percent(observed_mortality, accuracy = 0.1),
    `Prediction at assigned` = percent(
      prediction_at_assigned, accuracy = 0.1
    ),
    `Prediction at 90% Acropora` = percent(
      prediction_at_90pct_acropora, accuracy = 0.1
    ),
    `Remaining miss at 90%` = percent(
      remaining_underprediction_at_90pct, accuracy = 0.1
    )
  ) |>
  select(
    Reef, Source, `Assigned Acropora`, Observed,
    `Prediction at assigned`, `Prediction at 90% Acropora`,
    `Remaining miss at 90%`
  ) |>
  kable(caption = 'Fold-specific Acropora counterfactuals')
```

```{r acropora-counterfactual, fig.width=10, fig.height=6}
acropora_scenarios |>
  mutate(Reef = sub(' Reef.*$', '', ReefName)) |>
  ggplot(aes(scenario_prop_acropora_pre, predicted_mortality)) +
  geom_hline(
    aes(yintercept = observed_mortality),
    linetype = 2, colour = '#B2182B'
  ) +
  geom_line(colour = '#2166AC', linewidth = 0.9) +
  geom_point(colour = '#2166AC', size = 1.8) +
  geom_point(
    data = ~ filter(
      .x,
      abs(scenario_prop_acropora_pre - assigned_prop_acropora_pre) < 1e-12
    ),
    colour = 'black', size = 2.8
  ) +
  facet_wrap(~ Reef, scales = 'free_y') +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent) +
  labs(
    x = 'Counterfactual preceding Acropora proportion',
    y = 'Held-out core-BRT mortality prediction',
    title = 'Even extreme Acropora values do not reproduce the observed losses',
    subtitle = paste(
      'Black point is the assigned value;',
      'red dashed line is observed mortality'
    )
  )
```

This result separates two issues:

1. **Acropora uncertainty:** Eyrie and Swinger do not have direct reef-year
   Acropora observations. Better composition data could increase or decrease
   their predicted susceptibility. Lizard NW and Mackay are the most
   sensitive to raising Acropora, but still remain substantially
   underestimated.
2. **Unrepresented weedy assemblages:** preceding `prop_acropora_pre` is only
   the fraction of hard coral assigned to *Acropora*. A reef dominated by other
   fast-growing, disturbance-sensitive or recently recruited taxa is not
   represented. If the post-recovery communities at these reefs were
   compositionally young or dominated by other weedy taxa, the current model
   would classify them as less susceptible than they actually were.

The next community improvement should therefore be a broader
**community-susceptibility/recovery-state index**, not merely replacement of
missing Acropora values. It should distinguish Acropora, other fast-growing
susceptible taxa or growth forms, massive/stress-tolerant corals, and the age or
time-since-recovery structure where attainable. Composition provenance and
uncertainty should accompany each prediction.

# What was extracted

All six individual-pressure variables and all four derived metric families
were extracted for all five reefs. This gives 860 individual pressure records
and 120 metric records. The original workbook is not modified.

The valid pressure window for the 2024 bleaching event is summer `202324`.
COTS uses fiscal July--June, cyclone and DHW use November--April, and coloured
water uses December--April. Coral-sink connectivity is static.

The newest disturbance (`cmex`) and combined resilience
(`cmfi`/`quad`) metrics include summer `202425`. They are retained in the
extract but flagged as post-event leakage and are not used to explain 2024
mortality. Recovery-potential `cmfr` metrics ending in `202324` can be
examined without that specific leakage.

```{r coverage-table}
coverage |>
  kable(caption = 'Extraction and reference-table coverage')
```

The AIMS monitoring reference assigns sectors to only 505 of the 3,862 RRN
reefs. Primary peer percentiles therefore use all RRN reefs in the same
one-degree latitude band. Sector-region percentiles are secondary.

# Event-level pressure profile

```{r pressure-heatmap, fig.width=11, fig.height=5.8}
pressure |>
  mutate(
    Reef = factor(reef_short, levels = rev(reef_order)),
    Pressure = factor(pressure_labels[Type], levels = pressure_labels),
    display_value = case_when(
      Type %in% c('cot_meanpertow', 'cot_idwmeanpertow', 'wqc_freqcc12',
                  'uq_coralsink1') ~ sprintf('%.3f', value),
      Type == 'cyc_maxHrs4mw' ~ sprintf('%.1f h', value),
      Type == 'sst_maxdhw' ~ sprintf('%.2f', value)
    )
  ) |>
  ggplot(aes(Pressure, Reef, fill = vulnerability_percentile)) +
  geom_tile(colour = 'white', linewidth = 0.8) +
  geom_text(aes(label = display_value), size = 3.2) +
  scale_fill_gradient2(
    low = '#2166AC', mid = '#F7F7F7', high = '#B2182B',
    midpoint = 50, limits = c(0, 100)
  ) +
  labs(
    x = NULL, y = NULL,
    fill = 'Locally adverse\npercentile',
    title = 'Different RRN pressures stand out at different reefs',
    subtitle = paste(
      'Percentiles are within the same one-degree latitude band;',
      'coral-sink percentile is reversed'
    )
  ) +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1),
    panel.grid = element_blank()
  )
```

```{r context-table}
context |>
  mutate(
    Reef = reef_short,
    `Observed mortality` = percent(observed_mortality, accuracy = 0.1),
    `Core underprediction` = percent(core_underprediction, accuracy = 0.1),
    `Pre-event Acropora` = percent(prop_acropora_pre, accuracy = 0.1)
  ) |>
  transmute(
    Reef, Region = region_block,
    `Observed mortality`, `Core underprediction`,
    DHW = round(ann_maxdhw, 2), `Pre-event Acropora`,
    `Observed COTS/tow` = round(cot_meanpertow, 4),
    `IDW COTS/tow` = round(cot_idwmeanpertow, 4),
    `Coloured-water frequency` = round(wqc_freqcc12, 3),
    `Cyclone 4 m-wave hours` = round(cyc_maxHrs4mw, 1),
    `Coral-sink index` = round(uq_coralsink1, 3),
    `Field disturbance record` = disturbance_text
  ) |>
  arrange(desc(parse_number(`Observed mortality`))) |>
  kable(caption = '2024 mortality and valid RRN event pressures')
```

# Reef-specific interpretation

## Linnet

Linnet lost 71.5% of expected coral cover and the core BRT underpredicted it by
42.3 percentage points. Its RRN COTS values are locally elevated: observed
COTS per tow is at the 86th latitude-band percentile and the IDW estimate is
at the 98.9th percentile. Coloured-water frequency is 0.537, the reef's 95th
within-decade percentile and 80th percentile GBR-wide. Pre-event Acropora is
also high at 63.4%.

This is the strongest RRN case for **compound thermal susceptibility,
freshwater/plume exposure and low-level COTS pressure**. Its one-year recovery
potential is also relatively low GBR-wide. The latest `quad = 4` and very low
`cmfi` are consistent with poor resilience, but include 2024--25 disturbance
information and must not be used as pre-event predictors.

## Eyrie

Eyrie's COTS signal is the strongest of the five reefs: observed COTS per tow
is at the 98th local percentile and the IDW estimate at the 99.7th. Its
coloured-water frequency of 0.388 is the reef's 95th within-decade percentile,
although only the 38th percentile among all latitude-band reefs. Pre-event
Acropora is high at 51.5%.

The RRN evidence therefore supports **elevated local COTS pressure plus an
unusually plume-affected year for this reef**, rather than an exceptionally
high regional plume exposure. Its recovery-potential metrics are high, so low
generic recovery potential is not a convincing explanation.

## Lizard Island NW

Lizard NW has no RRN cyclone-wave or COTS signal and low coloured-water
exposure. Its distinctive value is a coral-sink index of 0.105, only the 15.6th
latitude-band percentile; expressed as vulnerability, this is the 84.4th
percentile.

Low connectivity could reduce demographic rescue and longer-term recovery, but
it is not direct evidence of the acute mortality mechanism. The reef remains a
candidate for an unmeasured local freshwater pulse, finer-scale thermal
exposure, taxonomic susceptibility, disease or another local disturbance.

## Swinger

Swinger experienced only 3.87 DHW and has no observed COTS or cyclone-wave
exposure. Coloured-water frequency is 0.150: the 95th percentile of its own
decade, 68th percentile GBR-wide and 60th percentile locally.

The main RRN hypothesis is therefore a **reef-specific coloured-water anomaly**
acting additively with thermal stress. The signal is meaningful historically
but not extreme compared with other nearby reefs, and the recovery-potential
metrics are high. It remains a partial rather than complete explanation.

## Mackay

Mackay is the largest unexplained miss: 61.3% observed mortality and 50.8
percentage points of core-model underprediction. RRN DHW is 3.77, very close to
the model input of 3.52 and only the 2.6th percentile of its latitude band.
Cyclone exposure and observed COTS are zero; coloured-water frequency is only
the 29th local percentile, although it is at the 80th percentile relative to
the reef's preceding ten event years. Recovery-potential metrics are high.

There is **no convincing gridded RRN explanation for Mackay**, but the field
record explicitly lists `coral bleaching, Cyclone Jasper, flood waters`.
The likely issue is therefore that seasonal water-colour, rainfall and
grid-cell wave exposure do not resolve the acute local runoff/freshwater pulse.
This reef should remain a high-resolution exposure-review case rather than
being used to force a general DHW curve upward.

# Historical context

```{r pressure-history, fig.width=11, fig.height=8}
pressure_history |>
  filter(
    source_summer >= 201415,
    source_summer <= 202324,
    Type %in% c(
      'wqc_freqcc12', 'cot_meanpertow',
      'cot_idwmeanpertow', 'sst_maxdhw'
    )
  ) |>
  mutate(
    end_year = floor(source_summer / 100) + 1L,
    Pressure = recode(
      Type,
      wqc_freqcc12 = 'Coloured-water frequency',
      cot_meanpertow = 'Observed COTS per tow',
      cot_idwmeanpertow = 'IDW COTS per tow',
      sst_maxdhw = 'Maximum DHW'
    ),
    Reef = factor(reef_short, levels = reef_order)
  ) |>
  ggplot(aes(end_year, value, colour = Reef)) +
  geom_line(linewidth = 0.7, na.rm = TRUE) +
  geom_point(size = 1.6, na.rm = TRUE) +
  geom_vline(xintercept = 2024, linetype = 2, colour = 'grey40') +
  facet_wrap(~ Pressure, scales = 'free_y', ncol = 2) +
  scale_colour_brewer(palette = 'Dark2') +
  labs(
    x = 'Summer ending year', y = NULL, colour = NULL,
    title = 'The 2024 coloured-water anomalies are reef-specific',
    subtitle = 'Dashed line marks the mortality-event summer'
  ) +
  theme(legend.position = 'top')
```

# Do these variables explain residuals across manta reefs?

```{r residual-association-plot, fig.width=9.5, fig.height=5.5}
associations |>
  mutate(
    Pressure = factor(
      pressure_labels[Type],
      levels = rev(pressure_labels)
    )
  ) |>
  ggplot(aes(spearman_core_residual, Pressure)) +
  geom_vline(xintercept = 0, colour = 'grey45') +
  geom_col(aes(fill = spearman_core_residual > 0), width = 0.68) +
  scale_fill_manual(values = c('#2166AC', '#B2182B'), guide = 'none') +
  labs(
    x = 'Spearman correlation with core-model residual',
    y = NULL,
    title = 'Only coloured water has a clear positive residual pattern',
    subtitle = 'All 89 reef-blocked 2024 manta observations'
  )
```

Across all 89 2024 manta observations, coloured-water frequency correlates
0.317 with observed mortality and 0.170 with the core-model residual. Reefs in
its upper quartile average 5.7 percentage points more underprediction than the
remaining reefs. This is consistent with the earlier positive additive BRMS
water-colour effect, but remains observational.

COTS is not positively associated with residuals across the whole manta
dataset. Eyrie and Linnet are therefore **reef-specific COTS flags**, not
evidence that COTS generally explains the 2024 model miss. Cyclone-wave
exposure is negatively associated with residuals and does not explain these
five reefs.

# Recovery metrics and leakage boundary

```{r recovery-heatmap, fig.width=9.5, fig.height=5.5}
metrics |>
  filter(
    Type == 'cmfr',
    Year %in% c(202323, 202223, 201423, 200423)
  ) |>
  mutate(
    Reef = factor(reef_short, levels = rev(reef_order)),
    Window = recode(
      variable,
      cmfr_1y = '1 year', cmfr_2y = '2 years',
      cmfr_10y = '10 years', cmfr_20y = '20 years'
    )
  ) |>
  ggplot(aes(Window, Reef, fill = latitude_band_percentile)) +
  geom_tile(colour = 'white', linewidth = 0.8) +
  geom_text(aes(label = sprintf('%.3f', value)), size = 3.4) +
  scale_fill_gradient(
    low = '#FEE8C8', high = '#1B7837', limits = c(0, 100)
  ) +
  labs(
    x = 'Recovery-potential window', y = NULL,
    fill = 'Local percentile',
    title = 'Mackay and Swinger do not have low generic recovery potential'
  ) +
  theme(panel.grid = element_blank())
```

Linnet has the weakest recent recovery-potential value of the five
(`cmfr_1y = 0.468`; 16.6th percentile GBR-wide), while Mackay and Swinger are
consistently high. The `cmfr` metric combines coral-sink connectivity and
water colour; it is contextual rather than an independent mechanism.

# Recommended next checks

1. Add the raw 2023--24 COTS variables to a **targeted sensitivity**, keeping
   them distinct from disturbance attribution. Concentrate on Linnet and Eyrie;
   do not infer a GBR-wide COTS mechanism from these two reefs.
2. Retain water colour as an additive freshwater/plume predictor and inspect
   higher temporal-resolution plume/rainfall/salinity data for Linnet, Eyrie
   and Swinger.
3. Treat Lizard NW's low coral-sink index as a resilience/recovery flag, not an
   acute mortality exposure.
4. Mackay's lineage and composition crosswalk are now corrected. Prioritise
   fine-scale rainfall, runoff, salinity and flood-plume products around its
   survey interval; the field record already identifies Jasper and flood
   waters even though the gridded predictors miss them.
5. Never use the latest `cmex`, `cmfi` or `quad` metrics to predict 2024
   mortality because they include summer 2024--25.
