---
title: 'Can IMOS Kd490 reconstruct missing 2024 eReefs salinity?'
subtitle: 'Event-blocked validation of a deliberately limited optical sensitivity analysis'
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(knitr)
library(mgcv)
library(readr)
library(scales)
library(tidyr)
library(yardstick)

input_file <- '../data/processed/ereefs_salinity_imos_kd490_reef_year.csv'
if (!file.exists(input_file)) {
  stop('Run scripts/extract_ereefs_salinity.py before rendering this report.')
}

salinity <- read_csv(input_file, show_col_types = FALSE)

# Several reef features can share one 4-km eReefs cell. Salinity is identical
# within that cell, so modelling every feature would create pseudo-replication.
# IMOS Kd490 is at finer resolution and is represented by the within-cell median.
cell_year <- salinity |>
  group_by(ereefs_lat_index, ereefs_lon_index, year) |>
  summarise(
    lon = first(ereefs_longitude),
    lat = first(ereefs_latitude),
    n_reef_features = n(),
    k490_q90 = if (all(is.na(k490_q90))) {
      NA_real_
    } else {
      median(k490_q90, na.rm = TRUE)
    },
    salinity_min = first(salinity_min),
    salinity_q10 = first(salinity_q10),
    salinity_mean = first(salinity_mean),
    freshwater_exposure_26 = first(freshwater_exposure_26),
    freshwater_exposure_28 = first(freshwater_exposure_28),
    days_below_26 = first(days_below_26),
    days_below_28 = first(days_below_28),
    salinity_available_days = first(salinity_available_days),
    salinity_complete_q1 = first(salinity_complete_q1),
    salinity_forcing_warning = first(salinity_forcing_warning),
    salinity_period_end = first(salinity_period_end),
    .groups = 'drop'
  ) |>
  mutate(
    flood28 = freshwater_exposure_28 > 0,
    log_kd = log(k490_q90)
  )

complete_years <- c(2016, 2017, 2020)
historical <- cell_year |>
  filter(
    year %in% complete_years,
    salinity_complete_q1,
    !salinity_forcing_warning,
    is.finite(log_kd)
  )

coverage <- cell_year |>
  group_by(year) |>
  summarise(
    `eReefs cells` = n(),
    `Available days` = first(salinity_available_days),
    `Complete Q1` = first(salinity_complete_q1),
    `Forcing warning` = first(salinity_forcing_warning),
    `Cells below 28 PSU` = sum(flood28, na.rm = TRUE),
    `Cells below 26 PSU` = sum(freshwater_exposure_26 > 0, na.rm = TRUE),
    `Minimum salinity` = min(salinity_min, na.rm = TRUE),
    `Maximum exposure below 28` = max(freshwater_exposure_28, na.rm = TRUE),
    .groups = 'drop'
  )
```

# Decision

This analysis asks a narrow question: can a consistent, full-quarter IMOS
optical metric reconstruct the salinity information that is absent after 17
January 2024 in the public eReefs GBR4 v2 aggregation?

The answer is assessed by holding out one *complete event year* at a time. The
analysis is not allowed to call an imputation successful merely because it
fits the many normal-salinity cells. It must identify the rare cells exposed
below 28 PSU and predict their minimum salinity with useful accuracy.

The 2022 eReefs values are shown for transparency but excluded from calibration
because AIMS documents a river-forcing error spanning December 2021 with
carry-over into January 2022. The 2024 eReefs values cover only 1--17 January
and are descriptive, not a validation target for the full January--March IMOS
metric.

# Data coverage

```{r coverage-table}
coverage |>
  mutate(
    `Minimum salinity` = round(`Minimum salinity`, 2),
    `Maximum exposure below 28` = round(`Maximum exposure below 28`, 1)
  ) |>
  kable(
    caption = paste(
      'Surface salinity coverage and low-salinity tail.',
      'Exposure is accumulated daily deficit in PSU-days.'
    )
  )
```

[AIMS defines freshwater exposure](https://ereefs.aims.gov.au/gbr1/fresh-water-exposure/)
as the accumulated daily salinity deficit below a nominated threshold. The
4-km product is useful for a first sensitivity analysis, but the low-salinity
tail is very sparse. The
[eReefs extraction service](https://extraction.ereefs.aims.gov.au/extract-data)
also records that GBR4 v2 stops on 17 January 2024.

# What Kd490 and salinity measure

Kd490 is optical attenuation. It can respond to river plumes, but also to
suspended sediment, coloured dissolved material, phytoplankton, and other
sources of reduced water clarity. It therefore cannot be assumed to be a
one-to-one salinity measurement.

```{r relationship-plot, fig.width=10, fig.height=4.8}
relationship_summary <- historical |>
  group_by(year, flood28) |>
  summarise(
    cells = n(),
    median_kd490 = median(k490_q90),
    median_min_salinity = median(salinity_min),
    .groups = 'drop'
  )

ggplot(
  historical,
  aes(x = k490_q90, y = salinity_min)
) +
  geom_point(colour = 'grey55', alpha = 0.18, size = 0.75) +
  geom_point(
    data = filter(historical, flood28),
    aes(colour = freshwater_exposure_28),
    size = 2.2
  ) +
  geom_hline(yintercept = 28, linetype = 2, colour = '#B2182B') +
  facet_wrap(~ year) +
  scale_x_log10() +
  scale_colour_viridis_c(option = 'C', trans = 'sqrt') +
  labs(
    x = expression('IMOS Q1 ' * K[d] * '(490) Q90'),
    y = 'eReefs Q1 minimum surface salinity (PSU)',
    colour = 'Exposure below\n28 (PSU-days)',
    title = 'Optical attenuation is associated with flooding in some events, not all'
  )
```

```{r relationship-table}
relationship_summary |>
  mutate(
    flood28 = if_else(flood28, 'Any exposure below 28', 'No exposure below 28'),
    median_kd490 = round(median_kd490, 3),
    median_min_salinity = round(median_min_salinity, 2)
  ) |>
  rename(
    Year = year,
    Group = flood28,
    Cells = cells,
    `Median Kd490` = median_kd490,
    `Median minimum salinity` = median_min_salinity
  ) |>
  kable()
```

# Leave-one-event-out validation

Four occurrence models are compared. The spatial terms are a static
reconstruction prior: they learn where low-salinity exposure has occurred in
other events, but receive no information from the held-out event. This is
appropriate for reconstructing 2024 at known GBR locations, not for forecasting
novel locations.

Two continuous models then attempt the harder task of predicting minimum
salinity. Tail MAE and the fraction of truly flooded cells predicted below 28
PSU are the decisive diagnostics.

```{r blocked-models}
classification_predictions <- list()
salinity_predictions <- list()
prediction_index <- 1

for (held_out_year in complete_years) {
  training <- filter(historical, year != held_out_year)
  testing <- filter(historical, year == held_out_year)

  kd_centre <- mean(training$log_kd)
  kd_scale <- sd(training$log_kd)
  training <- mutate(training, log_kd_z = (log_kd - kd_centre) / kd_scale)
  testing <- mutate(testing, log_kd_z = (log_kd - kd_centre) / kd_scale)

  occurrence_fits <- list(
    'Intercept only' = glm(
      flood28 ~ 1, data = training, family = binomial()
    ),
    'Kd490 only' = glm(
      flood28 ~ log_kd_z, data = training, family = binomial()
    ),
    'Spatial prior only' = bam(
      flood28 ~ s(lon, lat, k = 40),
      data = training, family = binomial(), method = 'fREML',
      select = TRUE, discrete = TRUE
    ),
    'Kd490 + spatial prior' = bam(
      flood28 ~ s(log_kd_z, k = 5, bs = 'cr') + s(lon, lat, k = 40),
      data = training, family = binomial(), method = 'fREML',
      select = TRUE, discrete = TRUE
    )
  )

  for (model_name in names(occurrence_fits)) {
    classification_predictions[[prediction_index]] <- testing |>
      transmute(
        held_out_year,
        model = model_name,
        flood28,
        observed_exposure = freshwater_exposure_28,
        predicted = as.numeric(predict(
          occurrence_fits[[model_name]], newdata = testing, type = 'response'
        ))
      )
    prediction_index <- prediction_index + 1
  }

  magnitude_fits <- list(
    'Kd490 only' = gam(
      salinity_min ~ s(log_kd_z, k = 5, bs = 'cr'),
      data = training, method = 'REML', select = TRUE
    ),
    'Kd490 + spatial prior' = bam(
      salinity_min ~ s(log_kd_z, k = 5, bs = 'cr') +
        s(lon, lat, k = 40),
      data = training, method = 'fREML', select = TRUE, discrete = TRUE
    )
  )

  for (model_name in names(magnitude_fits)) {
    salinity_predictions[[prediction_index]] <- testing |>
      transmute(
        held_out_year,
        model = model_name,
        flood28,
        observed = salinity_min,
        predicted = as.numeric(predict(
          magnitude_fits[[model_name]], newdata = testing
        ))
      )
    prediction_index <- prediction_index + 1
  }
}

classification_predictions <- bind_rows(classification_predictions)
salinity_predictions <- bind_rows(salinity_predictions)

top_one_percent_recall <- function(truth, estimate) {
  top_n <- max(1, ceiling(length(truth) * 0.01))
  selected <- order(estimate, decreasing = TRUE)[seq_len(top_n)]
  sum(truth[selected]) / sum(truth)
}

# Integrate precision over the recall increments at each distinct threshold.
# Unlike trapezoidal interpolation, this returns prevalence for a constant
# classifier, which is the correct no-skill reference for a rare event.
average_precision <- function(truth, estimate) {
  ordered <- order(estimate, decreasing = TRUE)
  truth <- as.integer(truth[ordered])
  estimate <- estimate[ordered]
  threshold_ends <- c(which(diff(estimate) != 0), length(estimate))
  true_positives <- cumsum(truth)[threshold_ends]
  recall <- true_positives / sum(truth)
  precision <- true_positives / threshold_ends
  sum(diff(c(0, recall)) * precision)
}

classification_metrics <- classification_predictions |>
  group_by(model, held_out_year) |>
  summarise(
    cells = n(),
    flood_cells = sum(flood28),
    prevalence = mean(flood28),
    roc_auc = roc_auc_vec(
      factor(flood28, levels = c(FALSE, TRUE)),
      predicted,
      event_level = 'second'
    ),
    average_precision = average_precision(flood28, predicted),
    brier = mean((predicted - as.numeric(flood28))^2),
    balanced_brier = 0.5 * (
      mean((predicted[flood28] - 1)^2) +
        mean(predicted[!flood28]^2)
    ),
    top_one_percent_recall = top_one_percent_recall(flood28, predicted),
    .groups = 'drop'
  )

classification_event_mean <- classification_metrics |>
  group_by(model) |>
  summarise(
    held_out_year = 'Event mean',
    cells = sum(cells),
    flood_cells = sum(flood_cells),
    prevalence = mean(prevalence),
    roc_auc = mean(roc_auc),
    average_precision = mean(average_precision),
    brier = mean(brier),
    balanced_brier = mean(balanced_brier),
    top_one_percent_recall = mean(top_one_percent_recall),
    .groups = 'drop'
  )

salinity_metrics <- salinity_predictions |>
  group_by(model, held_out_year) |>
  summarise(
    cells = n(),
    flood_cells = sum(flood28),
    all_cell_mae = mean(abs(predicted - observed)),
    flood_cell_mae = mean(abs(predicted[flood28] - observed[flood28]), na.rm = TRUE),
    flood_cell_bias = mean(predicted[flood28] - observed[flood28], na.rm = TRUE),
    flood_detection = mean(predicted[flood28] < 28, na.rm = TRUE),
    .groups = 'drop'
  )
```

Average precision is the principal occurrence metric because fewer than 0.5%
of cells are exposed below 28 PSU. Its no-skill value is the event prevalence.
ROC AUC is retained but can appear optimistic under that imbalance. Balanced
Brier gives flooded and non-flooded cells equal weight, and top-1% recall asks
whether a practical risk screen actually captures the flooded cells.

```{r occurrence-results}
classification_event_mean |>
  select(
    Model = model,
    `ROC AUC` = roc_auc,
    `Average precision` = average_precision,
    `Balanced Brier` = balanced_brier,
    `Top 1% recall` = top_one_percent_recall
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  arrange(desc(`Average precision`)) |>
  kable(caption = 'Event-balanced leave-one-event-out occurrence performance')
```

```{r occurrence-plot, fig.width=10, fig.height=4.8}
classification_metrics |>
  select(
    model, held_out_year, prevalence,
    average_precision, top_one_percent_recall
  ) |>
  pivot_longer(
    c(average_precision, top_one_percent_recall),
    names_to = 'metric', values_to = 'value'
  ) |>
  mutate(
    metric = recode(
      metric,
      average_precision = 'Average precision (PR)',
      top_one_percent_recall = 'Recall within highest-risk 1%'
    ),
    held_out_year = factor(held_out_year)
  ) |>
  ggplot(aes(x = held_out_year, y = value, colour = model, group = model)) +
  geom_point(size = 2.4) +
  geom_line() +
  facet_wrap(~ metric, scales = 'free_y') +
  scale_y_continuous(labels = label_number(accuracy = 0.01)) +
  scale_colour_brewer(palette = 'Dark2') +
  labs(
    x = 'Held-out event year', y = NULL, colour = 'Model',
    title = 'Optical flood detection does not transfer uniformly among events'
  )
```

```{r salinity-results}
salinity_metrics |>
  mutate(
    across(c(all_cell_mae, flood_cell_mae, flood_cell_bias), ~ round(.x, 2)),
    flood_detection = percent(flood_detection, accuracy = 1)
  ) |>
  rename(
    Model = model,
    `Held-out year` = held_out_year,
    `All-cell MAE` = all_cell_mae,
    `Flood-cell MAE` = flood_cell_mae,
    `Flood-cell bias` = flood_cell_bias,
    `Flood cells predicted below 28` = flood_detection
  ) |>
  select(
    Model, `Held-out year`, `All-cell MAE`, `Flood-cell MAE`,
    `Flood-cell bias`, `Flood cells predicted below 28`
  ) |>
  kable(
    caption = paste(
      'Leave-one-event-out minimum-salinity prediction.',
      'Positive flood-cell bias means harmful freshening is underestimated.'
    )
  )
```

```{r salinity-tail-plot, fig.width=10, fig.height=4.8}
salinity_predictions |>
  filter(flood28) |>
  ggplot(aes(x = observed, y = predicted, colour = factor(held_out_year))) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = 'grey55') +
  geom_hline(yintercept = 28, linetype = 3, colour = '#B2182B') +
  geom_vline(xintercept = 28, linetype = 3, colour = '#B2182B') +
  geom_point(size = 2.3, alpha = 0.8) +
  facet_wrap(~ model) +
  coord_equal() +
  scale_colour_brewer(palette = 'Dark2') +
  labs(
    x = 'Observed minimum salinity (PSU)',
    y = 'Event-held-out prediction (PSU)',
    colour = 'Held-out year',
    title = 'Ordinary error is small because normal cells dominate; the flood tail decides utility'
  )
```

# Exploratory 2024 optical reconstruction

The following reconstruction is generated to show what the optical model
would imply, not to fill the primary mortality predictor table. It uses all
three complete, unflagged historical events and the full-Q1 2024 IMOS metric.
The observed eReefs overlay covers only 1--17 January.

```{r reconstruction-2024}
training <- historical
testing_2024 <- cell_year |>
  filter(year == 2024, is.finite(log_kd))

kd_centre <- mean(training$log_kd)
kd_scale <- sd(training$log_kd)
training <- mutate(training, log_kd_z = (log_kd - kd_centre) / kd_scale)
testing_2024 <- mutate(
  testing_2024,
  log_kd_z = (log_kd - kd_centre) / kd_scale
)

final_kd490 <- glm(
  flood28 ~ log_kd_z,
  data = training, family = binomial()
)
final_spatial <- bam(
  flood28 ~ s(lon, lat, k = 40),
  data = training, family = binomial(), method = 'fREML',
  select = TRUE, discrete = TRUE
)
final_occurrence <- bam(
  flood28 ~ s(log_kd_z, k = 5, bs = 'cr') + s(lon, lat, k = 40),
  data = training, family = binomial(), method = 'fREML',
  select = TRUE, discrete = TRUE
)
final_minimum <- bam(
  salinity_min ~ s(log_kd_z, k = 5, bs = 'cr') + s(lon, lat, k = 40),
  data = training, method = 'fREML', select = TRUE, discrete = TRUE
)

reconstruction_2024 <- testing_2024 |>
  mutate(
    kd490_only_flood28_probability = as.numeric(predict(
      final_kd490, newdata = testing_2024, type = 'response'
    )),
    kd490_only_risk_percentile = percent_rank(
      kd490_only_flood28_probability
    ),
    spatial_prior_flood28_probability = as.numeric(predict(
      final_spatial, newdata = testing_2024, type = 'response'
    )),
    spatial_prior_risk_percentile = percent_rank(
      spatial_prior_flood28_probability
    ),
    kd490_spatial_flood28_probability = as.numeric(predict(
      final_occurrence, newdata = testing_2024, type = 'response'
    )),
    kd490_spatial_risk_percentile = percent_rank(
      kd490_spatial_flood28_probability
    ),
    candidate_salinity_min_q1 = as.numeric(predict(
      final_minimum, newdata = testing_2024
    )),
    partial_ereefs_flood28 = freshwater_exposure_28 > 0
  )

reef_reconstruction_2024 <- salinity |>
  filter(year == 2024) |>
  left_join(
    reconstruction_2024 |>
      select(
        ereefs_lat_index, ereefs_lon_index,
        kd490_only_flood28_probability,
        kd490_only_risk_percentile,
        spatial_prior_flood28_probability,
        spatial_prior_risk_percentile,
        kd490_spatial_flood28_probability,
        kd490_spatial_risk_percentile,
        candidate_salinity_min_q1
      ),
    by = c('ereefs_lat_index', 'ereefs_lon_index')
  ) |>
  mutate(
    reconstruction_status = paste(
      'exploratory Kd490 and spatial sensitivity; salinity imputation',
      'not accepted for the primary mortality model'
    ),
    use_in_primary_mortality_model = FALSE
  )

write_csv(
  reef_reconstruction_2024,
  '../data/processed/ereefs_salinity_2024_optical_sensitivity.csv'
)
write_csv(
  classification_metrics,
  '../output/salinity_kd490_loyo_occurrence_metrics.csv'
)
write_csv(
  classification_predictions,
  '../output/salinity_kd490_loyo_occurrence_predictions.csv'
)
write_csv(
  salinity_metrics,
  '../output/salinity_kd490_loyo_minimum_metrics.csv'
)
write_csv(
  salinity_predictions,
  '../output/salinity_kd490_loyo_minimum_predictions.csv'
)
```

```{r reconstruction-map, fig.width=10, fig.height=6}
ggplot(
  reconstruction_2024,
  aes(x = lon, y = lat, colour = kd490_spatial_risk_percentile)
) +
  geom_point(size = 1.05) +
  geom_point(
    data = filter(reconstruction_2024, partial_ereefs_flood28),
    shape = 21, fill = NA, colour = 'black', size = 4, stroke = 1.1
  ) +
  coord_equal() +
  scale_colour_viridis_c(labels = percent, option = 'C') +
  labs(
    x = 'Longitude', y = 'Latitude',
    colour = 'Kd490 + spatial\nrisk percentile',
    title = 'Exploratory 2024 Kd490-plus-spatial risk surface',
    subtitle = 'Black rings: eReefs exposure below 28 PSU observed during 1--17 January only'
  )
```

All three low-salinity cells available in the partial 2024 eReefs record fall
at or above the `r percent(min(reconstruction_2024$kd490_spatial_risk_percentile[reconstruction_2024$partial_ereefs_flood28]), accuracy = 0.1)`
Kd490-plus-spatial risk percentile. Their Kd490-only percentiles span
`r paste(percent(range(reconstruction_2024$kd490_only_risk_percentile[reconstruction_2024$partial_ereefs_flood28]), accuracy = 0.1), collapse = ' to ')`,
showing how strongly the combined result relies on the spatial prior. This
supports retaining a ranked sensitivity layer, but not calling it satellite
salinity. It also does not rescue numeric salinity imputation: their observed
partial-period minima
range from `r round(min(reconstruction_2024$salinity_min[reconstruction_2024$partial_ereefs_flood28]), 1)`
to `r round(max(reconstruction_2024$salinity_min[reconstruction_2024$partial_ereefs_flood28]), 1)`
PSU, whereas the candidate full-quarter model predicts
`r paste(round(range(reconstruction_2024$candidate_salinity_min_q1[reconstruction_2024$partial_ereefs_flood28]), 1), collapse = ' to ')`
PSU. These periods are not temporally aligned, so the comparison is descriptive;
the held-out historical tail errors provide the formal reason for rejection.

# Recommendation

The optical analysis is retained as a *plume/flood-risk sensitivity*, because
it can contain ecologically relevant information that direct heat metrics do
not. It is not promoted to a quantitative salinity imputation for the primary
mortality model unless event-held-out results show consistent low-salinity-tail
performance. In particular:

1. Do not replace missing February--March 2024 salinity with the smooth
   `candidate_salinity_min_q1` values merely because all-cell MAE is small.
2. Keep the existing IMOS acute optical predictor in the formal BRMS and BRT
   models; it is observed consistently through the 2024 event.
3. Use the saved, explicitly separated Kd490-only, spatial-only, and combined
   risk percentiles only as sensitivity covariates or stratification layers,
   with `use_in_primary_mortality_model = FALSE`.
4. Revisit direct salinity only when a consistent 2024 hydrodynamic product or
   an independent in-situ/satellite salinity validation source is available.

This prevents a model-derived environmental covariate from being treated as a
measurement and avoids giving a visually plausible map more certainty than the
event-transfer tests support.
