---
title: 'Can the models recover severe mortality at high DHW?'
subtitle: 'Event-held-out comparison of the formal benchmark and shape-aware candidates'
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)
library(tidyr)

input_dir <- '../output/shape_aware_models'
metrics <- read_csv(file.path(input_dir, 'comparison_metrics.csv'), show_col_types = FALSE)
events <- read_csv(file.path(input_dir, 'event_metrics.csv'), show_col_types = FALSE)
rows <- read_csv(file.path(input_dir, 'comparison_rows.csv'), show_col_types = FALSE)
curves <- read_csv(file.path(input_dir, 'production_dose_curves.csv'), show_col_types = FALSE)
weights <- read_csv(file.path(input_dir, 'shape_production_weights.csv'), show_col_types = FALSE)
diagnostics <- read_csv(file.path(input_dir, 'shape_brms_diagnostics.csv'), show_col_types = FALSE)
routes <- read_csv(file.path(input_dir, 'recommended_model_routes.csv'), show_col_types = FALSE)

programme_labels <- c(ltmp = 'LTMP benthic', manta = 'Manta tow', mmp = 'MMP inshore')
model_labels <- c(
  benchmark_brms = 'Benchmark BRMS', benchmark_brt = 'Benchmark BRT',
  benchmark_ensemble = 'Benchmark ensemble', shape_brms = 'Shape-aware BRMS',
  shape_brt = 'Shape-aware BRT', shape_ensemble = 'Shape-aware ensemble',
  recommended_route = 'Recommended routed model'
)
```

# The candidate being tested

The shape-aware candidate is not a forced 50% mortality rule. It makes four
specific changes and leaves the existing benchmark untouched:

- proportional Acropora cover is the primary community-susceptibility term;
- total preceding coral cover replaces the strongly correlated absolute
  Acropora-cover term;
- the response can steepen above 8 DHW and the DHW--Acropora interaction enters
  both mortality occurrence and positive magnitude;
- event identity is excluded from the ecological mean so it cannot absorb a
  relationship that must transfer to a future event.

The BRT additionally constrains DHW and proportional Acropora effects to be
non-decreasing, gives each reef-year equal fitting weight, and tunes against
overall, event-balanced, and high-DHW errors. The Bayesian model uses
directional but non-binding ecological priors.

# Recommended production route

No single candidate wins for every observation programme. The recommended
predictor therefore uses the best event-held-out route for each programme and
does not ask a tree model to extrapolate beyond the largest DHW in its training
data.

```{r route-table}
routes |>
  mutate(
    Programme = programme_labels[programme_key],
    `Within observed DHW support` = model_labels[within_support_model],
    `Beyond observed DHW support` = model_labels[outside_support_model]
  ) |>
  select(
    Programme, `Within observed DHW support`,
    `Beyond observed DHW support`, Reason = reason
  ) |>
  kable(caption = 'Programme-specific model routing rule')
```

This promotes the shape-aware ensemble for LTMP within support, retains the
benchmark for manta tow, and promotes the shape-aware BRMS for MMP. BRMS alone
is used beyond observed DHW support because boosted trees cannot extrapolate a
new dose-response slope.

# Held-out predictive performance

```{r overall-table}
metrics |>
  filter(learner %in% c(
    'benchmark_ensemble', 'shape_brms', 'shape_ensemble',
    'recommended_route'
  )) |>
  mutate(
    Programme = programme_labels[programme_key],
    Model = model_labels[learner],
    across(where(is.numeric), ~ round(.x, 3))
  ) |>
  select(
    Programme, Model, n, RMSE = rmse, MAE = mae,
    `Predictive R2` = predictive_r2,
    `Observed - predicted` = bias_observed_minus_predicted,
    `Severe-loss MAE` = severe_mae,
    `High-DHW MAE` = high_dhw_mae,
    `High-DHW/Acropora MAE` = high_dhw_high_acropora_mae,
    `2020/22 MAE` = low_event_mae
  ) |>
  arrange(Programme, RMSE) |>
  kable(caption = 'Leave-one-event-out performance; lower error is better')
```

Overall RMSE is retained because it measures GBR-wide average error, but it is
not allowed to hide the severe tail. High-DHW/Acropora rows have at least 10 DHW
and at least 70% proportional Acropora cover. Positive bias below means the
model is still underpredicting mortality.

```{r bias-table}
metrics |>
  filter(learner %in% c('benchmark_ensemble', 'recommended_route')) |>
  mutate(
    Programme = programme_labels[programme_key],
    Model = model_labels[learner],
    across(
      c(high_dhw_bias, high_dhw_high_acropora_bias, low_event_bias),
      ~ round(.x, 3)
    )
  ) |>
  select(
    Programme, Model,
    `High-DHW observed - predicted` = high_dhw_bias,
    `High-DHW/Acropora observed - predicted` = high_dhw_high_acropora_bias,
    `2020/22 observed - predicted` = low_event_bias
  ) |>
  kable(caption = 'The high-tail versus low-event calibration trade-off')
```

```{r event-table}
events |>
  filter(learner %in% c('benchmark_ensemble', 'recommended_route')) |>
  mutate(
    Programme = programme_labels[programme_key],
    Model = model_labels[learner],
    across(c(observed_mean, predicted_mean, bias_observed_minus_predicted, rmse), ~ round(.x, 3))
  ) |>
  select(
    Programme, Model, Event = event_year,
    `Observed mean` = observed_mean, `Predicted mean` = predicted_mean,
    `Observed - predicted` = bias_observed_minus_predicted, RMSE = rmse
  ) |>
  kable(caption = 'Calibration in each held-out event')
```

# Does 12 DHW now produce ecologically plausible mortality?

The production curves below use all five events for fitting. "Acropora
dominated" means 90% proportional Acropora and 55% total pre-event coral cover;
other attainable predictors are held at programme-specific medians. These are
scenario predictions, not a replacement for held-out validation.

```{r dhw12-table}
curves |>
  filter(
    abs(ann_maxdhw - 12) < 1e-8,
    composition == 'Acropora dominated',
    learner %in% c('benchmark_ensemble', 'shape_brms', 'recommended_route')
  ) |>
  mutate(
    Programme = programme_labels[programme_key],
    Model = model_labels[learner],
    across(c(occurrence, positive, mortality), ~ round(.x, 3))
  ) |>
  select(
    Programme, Model,
    `Mortality occurrence` = occurrence,
    `Mortality when present` = positive,
    `Expected mortality` = mortality
  ) |>
  arrange(Programme, desc(`Expected mortality`)) |>
  kable(caption = 'Production response at 12 DHW in an Acropora-dominated community')
```

```{r dose-curves, fig.width=11, fig.height=8}
curves |>
  filter(learner %in% c(
    'benchmark_ensemble', 'shape_brms', 'recommended_route'
  )) |>
  mutate(
    Programme = programme_labels[programme_key],
    Model = model_labels[learner]
  ) |>
  ggplot(aes(ann_maxdhw, mortality, colour = Model)) +
  geom_hline(yintercept = 0.5, colour = 'grey55', linetype = 3) +
  geom_vline(xintercept = 12, colour = 'grey55', linetype = 3) +
  geom_line(linewidth = 1) +
  facet_grid(Programme ~ composition) +
  scale_colour_manual(values = c(
    'Benchmark ensemble' = '#4D4D4D',
    'Shape-aware BRMS' = '#D95F02',
    'Recommended routed model' = '#1B9E77'
  )) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  labs(
    x = 'Annual maximum DHW', y = 'Expected coral mortality',
    colour = NULL,
    title = 'Composition-specific production dose responses',
    subtitle = 'Other environmental predictors held at programme medians'
  ) +
  theme(legend.position = 'top')
```

# Where does held-out underprediction remain?

```{r residual-curves, fig.width=11, fig.height=7}
rows |>
  filter(learner %in% c('benchmark_ensemble', 'recommended_route')) |>
  mutate(
    Programme = programme_labels[programme_key],
    Model = model_labels[learner]
  ) |>
  ggplot(aes(ann_maxdhw, residual, colour = Model)) +
  geom_hline(yintercept = 0, colour = 'grey40') +
  geom_point(alpha = 0.22, size = 1) +
  geom_smooth(method = 'loess', se = TRUE, linewidth = 0.9) +
  facet_grid(Programme ~ event_year) +
  scale_colour_manual(values = c(
    'Benchmark ensemble' = '#4D4D4D',
    'Recommended routed model' = '#1B9E77'
  )) +
  labs(
    x = 'Annual maximum DHW',
    y = 'Observed - predicted mortality', colour = NULL,
    title = 'Event-held-out residuals across thermal stress',
    subtitle = 'Positive values indicate underprediction'
  ) +
  theme(legend.position = 'top')
```

# Ensemble and sampling diagnostics

```{r weights}
weights |>
  mutate(
    Programme = programme_labels[programme_key],
    across(c(weight_brms, weight_brt), ~ round(.x, 3))
  ) |>
  select(
    Programme, n,
    `Shape-aware BRMS weight` = weight_brms,
    `Shape-aware BRT weight` = weight_brt
  ) |>
  kable(caption = 'Candidate production weights estimated from event-held-out predictions')
```

```{r diagnostics}
diagnostics |>
  mutate(
    Programme = programme_labels[programme_key],
    across(c(max_rhat, min_bulk_ess, min_tail_ess), ~ round(.x, 3))
  ) |>
  select(
    Programme, Event = fold,
    `Maximum R-hat` = max_rhat,
    `Minimum bulk ESS` = min_bulk_ess,
    `Minimum tail ESS` = min_tail_ess,
    Divergences = divergences,
    `Tree-depth hits` = max_treedepth_hits
  ) |>
  kable(caption = 'Shape-aware BRMS sampling diagnostics')
```

Promotion requires more than producing a steeper curve. The candidate must
improve held-out high-DHW and severe-loss calibration, retain acceptable
overall error, avoid inflating the low-mortality 2020/22 events, and pass all
sampling diagnostics. Any failure remains visible rather than being averaged
away by the production scenario.
