---
title: 'Wind and doldrum-like conditions for GBR bleaching mortality'
subtitle: 'Coverage audit and source decision for eReefs, Sentinel-1 SAR, and Metop-B'
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)

ereefs_file <- '../data/processed/ereefs_wind_reef_year.csv'
metop_file <- '../data/processed/imos_metop_b_wind_reef_year.csv'
sar_daily_file <- '../output/wind_audit/sar_wind_2024_q1_daily_coverage.csv'
sar_reef_file <- '../output/wind_audit/sar_wind_2024_q1_reef_coverage.csv'
required <- c(ereefs_file, metop_file, sar_daily_file, sar_reef_file)
if (any(!file.exists(required))) {
  stop('Run the eReefs, Metop-B, and Sentinel-1 coverage scripts first.')
}

ereefs <- read_csv(ereefs_file, show_col_types = FALSE) |>
  mutate(lon_key = round(lon, 6), lat_key = round(lat, 6))
metop <- read_csv(metop_file, show_col_types = FALSE) |>
  mutate(lon_key = round(lon, 6), lat_key = round(lat, 6))
sar_daily <- read_csv(sar_daily_file, show_col_types = FALSE)
sar_reef <- read_csv(sar_reef_file, show_col_types = FALSE)

event_summary <- ereefs |>
  group_by(year) |>
  summarise(
    features = n(),
    available_days = first(ereefs_wind_available_days),
    complete_q1 = first(ereefs_wind_complete_q1),
    median_mean_wind = median(ereefs_wind_mean),
    median_q10_wind = median(ereefs_wind_q10),
    median_fraction_below_3 = median(ereefs_wind_fraction_below_3),
    median_longest_spell_below_3 = median(
      ereefs_wind_longest_spell_below_3
    ),
    .groups = 'drop'
  )

joined <- ereefs |>
  inner_join(
    metop |>
      select(
        LABEL_ID, LOC_NAME_S, lon_key, lat_key, year,
        metop_mean = wind_mean,
        metop_q10 = wind_q10,
        metop_calm_fraction = wind_fraction_below_3,
        metop_coverage = wind_day_coverage
      ),
    by = c('LABEL_ID', 'LOC_NAME_S', 'lon_key', 'lat_key', 'year')
  ) |>
  group_by(ereefs_lat_index, ereefs_lon_index, year) |>
  summarise(
    lon = first(ereefs_longitude),
    lat = first(ereefs_latitude),
    ereefs_mean = first(ereefs_wind_mean),
    ereefs_q10 = first(ereefs_wind_q10),
    ereefs_calm_fraction = first(ereefs_wind_fraction_below_3),
    metop_mean = median(metop_mean, na.rm = TRUE),
    metop_q10 = median(metop_q10, na.rm = TRUE),
    metop_calm_fraction = median(metop_calm_fraction, na.rm = TRUE),
    metop_coverage = max(metop_coverage, na.rm = TRUE),
    .groups = 'drop'
  ) |>
  filter(
    year != 2024,
    metop_coverage >= 0.70,
    is.finite(metop_mean)
  )

comparison <- joined |>
  group_by(year) |>
  summarise(
    cells = n(),
    mean_spearman = cor(ereefs_mean, metop_mean, method = 'spearman'),
    mean_rmse = sqrt(mean((ereefs_mean - metop_mean)^2)),
    ereefs_minus_metop_bias = mean(ereefs_mean - metop_mean),
    q10_spearman = cor(ereefs_q10, metop_q10, method = 'spearman'),
    calm_fraction_spearman = cor(
      ereefs_calm_fraction, metop_calm_fraction, method = 'spearman'
    ),
    .groups = 'drop'
  )
write_csv(
  comparison,
  '../output/wind_audit/wind_product_comparison_metrics.csv'
)

programme_coverage <- bind_rows(lapply(c('ltmp', 'manta', 'mmp'), function(key) {
  observations <- readRDS(paste0('../data/processed/validation_rows_', key, '.rds'))
  observations |>
    left_join(
      metop |>
        select(
          LOC_NAME_S, year, wind_n_days, wind_day_coverage
        ),
      by = c('ReefName' = 'LOC_NAME_S', 'event_year' = 'year')
    ) |>
    summarise(
      programme_key = key,
      model_rows = n(),
      any_metop_wind = sum(wind_n_days > 0, na.rm = TRUE),
      adequate_metop_wind = sum(wind_day_coverage >= 0.70, na.rm = TRUE)
    )
}))
```

# Decision

The linked Sentinel-1 SAR product should **not** be used to count calm days or
reconstruct a full Q1 wind history. The eReefs daily wind forcing is retained
for historical sensitivity analyses, but its 2024 value is explicitly partial.
Metop-B is a useful independent offshore comparison, not a GBR-wide replacement.

Wind should enter the formal BRMS/BRT candidate set only after a continuous,
single-source 2016--2024 reanalysis extraction (ERA5 or the corresponding ACCESS
product) passes comparison against eReefs and good-QC Metop observations.

# What is available

The [IMOS Sentinel-1 SAR wind record](https://portal-beta.aodn.org.au/details/b02b929f-2caf-45d4-ac60-d4632b7ca0ca?tab=summary)
contains calibrated 10-m neutral wind speed and direction at approximately 1-km
resolution. The official metadata dates it from 23 October 2017, which already
excludes the 2016 event. Its
[daily gridded THREDDS archive](https://thredds.aodn.org.au/thredds/catalog/IMOS/SRS/Surface-Waves/SAR_Wind/DELAYED/GRIDDED/SENTINEL-1/catalog.html)
combines only the satellite swaths present on each day; a nominal daily file is
not a wall-to-wall daily observation.

The DMS eReefs GBR4 store contains `mean_wspeed`, `wspeed_u`, and `wspeed_v`.
These are continuous model-forcing fields for complete Q1 periods in 2016,
2017, 2020, and 2022, but the public v2 record stops on 17 January 2024.

The adjacent IMOS Metop-B scatterometer archive spans every event from 2016 to
2024 and is calibrated against buoy winds. It has regular repeat sampling
offshore, but valid coastal coverage is insufficient for the inshore programme.

# Sentinel-1 coverage audit

The audit queried all 91 Q1 2024 daily files on a 0.1-degree GBR grid while
preserving the native swath gaps.

```{r sar-coverage}
sar_summary <- tibble(
  Measure = c(
    'Mean GBR domain covered on a day',
    'Median observed days per audit cell',
    'Median observed days at a registry feature',
    'Registry features with at least 10 observed days'
  ),
  Value = c(
    percent(mean(sar_daily$valid_grid_fraction), accuracy = 0.1),
    as.character(median(read_csv(
      '../output/wind_audit/sar_wind_2024_q1_grid_coverage.csv',
      show_col_types = FALSE
    )$observed_days)),
    as.character(median(sar_reef$audit_observed_days)),
    paste0(sum(sar_reef$audit_observed_days >= 10), ' / ', nrow(sar_reef))
  )
)
sar_summary |> kable()
```

Seven observed days out of 91 cannot identify sustained calm spells. Using a
wide spatial neighbourhood would increase counts by pooling different places,
not by observing the same reef more often.

# Metop-B coverage at modelling reefs

Good-QC Metop-B observations within 30 km were aggregated to daily medians. A
reef-year is called adequate only when at least 70% of Q1 days are represented.

```{r programme-coverage}
programme_coverage |>
  mutate(
    Programme = recode(
      programme_key,
      ltmp = 'LTMP benthic',
      manta = 'LTMP manta tow',
      mmp = 'MMP inshore'
    ),
    `Adequate coverage` = paste0(
      adequate_metop_wind, ' / ', model_rows,
      ' (', percent(adequate_metop_wind / model_rows, accuracy = 1), ')'
    )
  ) |>
  select(Programme, `Model rows` = model_rows, `Adequate coverage`) |>
  kable()
```

The zero coverage for MMP is a hard ecological limitation: replacing inshore
conditions with the nearest distant offshore pixel would erase the very
coastal wind and flood context that matters.

# Continuous eReefs historical metrics

```{r event-table}
event_summary |>
  mutate(
    Period = if_else(complete_q1, 'Complete Q1', '1--17 January only'),
    across(
      c(median_mean_wind, median_q10_wind),
      ~ round(.x, 2)
    ),
    median_fraction_below_3 = percent(
      median_fraction_below_3, accuracy = 0.1
    )
  ) |>
  select(
    Year = year, Period,
    `Median mean wind (m/s)` = median_mean_wind,
    `Median lower-decile wind (m/s)` = median_q10_wind,
    `Median fraction below 3 m/s` = median_fraction_below_3,
    `Median longest spell below 3 m/s` = median_longest_spell_below_3
  ) |>
  kable()
```

```{r event-plot, fig.width=10, fig.height=4.8}
ereefs |>
  mutate(
    Event = if_else(
      ereefs_wind_complete_q1,
      as.character(year),
      paste0(year, ' (partial)')
    )
  ) |>
  ggplot(aes(x = factor(Event), y = ereefs_wind_fraction_below_3)) +
  geom_boxplot(outlier.alpha = 0.08, fill = '#74ADD1') +
  scale_y_continuous(labels = percent) +
  labs(
    x = 'Event period',
    y = 'Fraction of available days below 3 m/s',
    title = 'eReefs represents continuous historical calm exposure',
    subtitle = 'The 2024 distribution covers only 1--17 January and is not comparable to full Q1'
  )
```

# eReefs versus Metop-B

Comparison is restricted to unique eReefs cells with at least 70% good-QC
Metop coverage. The 2024 products are not compared because their time windows
are different.

```{r comparison-table}
comparison |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  rename(
    Year = year,
    Cells = cells,
    `Mean Spearman` = mean_spearman,
    `Mean RMSE` = mean_rmse,
    `eReefs - Metop bias` = ereefs_minus_metop_bias,
    `Q10 Spearman` = q10_spearman,
    `Calm-fraction Spearman` = calm_fraction_spearman
  ) |>
  kable()
```

```{r comparison-plot, fig.width=10, fig.height=5}
ggplot(joined, aes(x = metop_mean, y = ereefs_mean)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = 'grey55') +
  geom_point(alpha = 0.25, size = 0.8, colour = '#2166AC') +
  facet_wrap(~ year) +
  coord_equal() +
  labs(
    x = 'Metop-B Q1 mean wind (m/s)',
    y = 'eReefs Q1 mean wind (m/s)',
    title = 'Mean wind agrees better than the low-wind tail'
  )
```

Mean wind agreement is adequate for a broad product check. It is not enough
for imputation of doldrum exposure: event-specific lower-decile and calm-
fraction correlations are weak and sometimes near zero. The products also
represent different spatial scales and coastal masks.

# Implementation recommendation

1. Preserve the extracted eReefs variables as historical sensitivity inputs:
   Q1 mean, Q10, minimum, fractions and days below 3/5 m/s, and longest calm
   spells. Do not label the 2024 record complete.
2. Preserve Sentinel-1 and Metop coverage fields and never interpret missing
   satellite pixels as calm conditions.
3. Extract hourly ERA5 or the matched ACCESS 10-m `u`/`v` winds for every event
   from one source, compute speed before daily aggregation, and use the same
   reef matching and Q1 definitions in all years.
4. Validate reanalysis mean and low-wind tails against complete eReefs years
   and good-QC Metop cells. Only then add a compact wind set to candidate models:
   mean speed plus either calm fraction or longest calm spell, avoiding several
   highly collinear threshold variants at once.
5. Re-run the fixed outer validation folds before promoting wind into the
   production BRMS/BRT ensemble. The ecological hypothesis is strong, but the
   predictor still has to improve temporal transfer, severe-event error, or
   interval calibration.
