## LTMP Benthic Photo-Transect (N=107, Offshore)

This section presents results for the LTMP Benthic photo-transect dataset — offshore reefs surveyed at 5 m depth with higher spatial precision. The dataset includes proportion Acropora as a community vulnerability metric.

```{r qbin-ltmp-model}
#| label: qbin-ltmp-model
#| code-fold: true
#| code-summary: "Data Preparation & Quasibinomial GLM Summary"

acrop_comp <- df.AIMS.full %>%
  filter(data_type == "photo-transect", domain_category == "reef",
         purpose == "COMPOSITION", variable == "HARD CORAL") %>%
  mutate(Reef_Name_Clean = gsub(" REEF(S)?| ISLAND| IS", "", toupper(domain_name))) %>%
  left_join(dat.Ref.Clean %>% select(AIMS_REEF_NAME_Clean, ReefName),
            by = c("Reef_Name_Clean" = "AIMS_REEF_NAME_Clean")) %>%
  filter(!is.na(ReefName)) %>%
  group_by(ReefName, report_year, depth) %>%
  summarise(total_hc = sum(mean, na.rm = TRUE),
    acrop_cover = sum(mean[grepl("Acropora", reefpage_category, ignore.case = TRUE)], na.rm = TRUE),
    .groups = "drop") %>%
  mutate(prop_acropora = ifelse(total_hc > 0, acrop_cover / total_hc, 0))

dat.ml.ltmp.clean <- dat.ml.ltmp %>%
  filter(!is.na(mcur_90), !is.na(Rel.Change)) %>%
  left_join(acrop_comp %>% select(ReefName, report_year, depth, prop_acropora),
            by = c("ReefName", "report_year", "depth")) %>%
  mutate(prop_acropora = replace_na(prop_acropora, median(prop_acropora, na.rm = TRUE)))
N_ltmp <- nrow(dat.ml.ltmp.clean)
dat.ml.ltmp.clean <- dat.ml.ltmp.clean %>%
  mutate(Mort.prop = pmin(pmax(-Rel.Change, 0), 1),
         Mort.prop.nudge = (Mort.prop * (N_ltmp - 1) + 0.5) / N_ltmp)
cat("LTMP Benthic model dataset:", N_ltmp, "observations\n")

fit_qbin_ltmp <- glm(
  Mort.prop.nudge ~ MaxDHW.mean * secc3m + MaxDHW.mean * cloudp_90 +
                    MaxDHW.mean * histmDHW6 + MaxDHW.mean * mcur_90 +
                    MaxDHW.mean * prop_acropora + winyear_sd,
  data = dat.ml.ltmp.clean, family = quasibinomial(link = "logit"))
print(summary(fit_qbin_ltmp))
```

### Effect Size Comparison

::: {.panel-tabset}
#### Bayesian Beta

```{r brms-halfeye-ltmp}
#| label: brms-halfeye-ltmp
#| fig-cap: "LTMP Benthic — Posterior distributions (Beta)."
#| fig-width: 10
#| fig-height: 6

library(brms); library(dplyr); library(tidybayes)
draws_ltmp <- as_draws_df(fit_brms_ltmp) %>%
  select(starts_with("b_")) %>%
  pivot_longer(everything(), names_to = "term", values_to = "estimate") %>%
  filter(term != "b_Intercept") %>%
  mutate(term = gsub("b_", "", term), term = gsub("MaxDHW_s", "DHW", term),
         term = gsub("histmDHW6_s", "Prior Exposure", term), term = gsub("secc3m_s", "Secchi Depth", term),
         term = gsub("cloudp_90_s", "Cloud Cover", term), term = gsub("winyear_sd_s", "Winter SST SD", term),
         term = gsub("mcur_90_s", "Current Speed", term), term = gsub("prop_acropora_s", "Prop. Acropora", term),
         term = gsub(":", " x ", term))
ggplot(draws_ltmp, aes(x = estimate, y = reorder(term, estimate))) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_qi, fill = "darkorange", alpha = 0.7, normalize = "xy") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(x = "Posterior Estimate (logit scale)", y = NULL, title = "LTMP Benthic — Posterior Effect Distributions (Beta)") +
  theme_bw(base_size = 12)
```

#### Bayesian Binomial-OLRE

```{r brms-halfeye-ltmp-olre}
#| label: brms-halfeye-ltmp-olre
#| fig-cap: "LTMP Benthic — Posterior distributions (Binomial-OLRE)."
#| fig-width: 10
#| fig-height: 6

library(brms); library(dplyr); library(tidybayes)
draws_ltmp_olre <- as_draws_df(fit_brmsb_ltmp) %>%
  select(starts_with("b_")) %>%
  pivot_longer(everything(), names_to = "term", values_to = "estimate") %>%
  filter(term != "b_Intercept") %>%
  mutate(term = gsub("b_", "", term), term = gsub("MaxDHW_s", "DHW", term),
         term = gsub("histmDHW6_s", "Prior Exposure", term), term = gsub("secc3m_s", "Secchi Depth", term),
         term = gsub("cloudp_90_s", "Cloud Cover", term), term = gsub("winyear_sd_s", "Winter SST SD", term),
         term = gsub("mcur_90_s", "Current Speed", term), term = gsub("prop_acropora_s", "Prop. Acropora", term),
         term = gsub(":", " x ", term))
ggplot(draws_ltmp_olre, aes(x = estimate, y = reorder(term, estimate))) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = median_qi, fill = "coral", alpha = 0.7, normalize = "xy") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(x = "Posterior Estimate (logit scale)", y = NULL, title = "LTMP Benthic — Posterior Effect Distributions (Binomial-OLRE)") +
  theme_bw(base_size = 12)
```

#### Quasibinomial GLM

```{r qbin-ltmp-effects}
#| label: qbin-ltmp-effects
#| fig-cap: "LTMP Benthic — Coefficient estimates (logit scale) with 90% CI."
#| fig-width: 8
#| fig-height: 5

coef_ltmp <- as.data.frame(coef(summary(fit_qbin_ltmp)))
names(coef_ltmp) <- c("Estimate", "SE", "t", "p")
coef_ltmp$Term <- rownames(coef_ltmp)
coef_ltmp <- coef_ltmp[coef_ltmp$Term != "(Intercept)", ]
z90 <- qnorm(0.95)
coef_ltmp$lo <- coef_ltmp$Estimate - z90 * coef_ltmp$SE
coef_ltmp$hi <- coef_ltmp$Estimate + z90 * coef_ltmp$SE
coef_ltmp$Term <- gsub("MaxDHW.mean", "DHW", coef_ltmp$Term)
coef_ltmp$Term <- gsub("histmDHW6", "Prior Exposure (>6 DHW)", coef_ltmp$Term)
coef_ltmp$Term <- gsub("secc3m", "Secchi Depth", coef_ltmp$Term)
coef_ltmp$Term <- gsub("cloudp_90", "Cloud Cover", coef_ltmp$Term)
coef_ltmp$Term <- gsub("winyear_sd", "Winter SST SD", coef_ltmp$Term)
coef_ltmp$Term <- gsub("mcur_90", "Current Speed", coef_ltmp$Term)
coef_ltmp$Term <- gsub("prop_acropora", "Prop. Acropora", coef_ltmp$Term)
coef_ltmp$Term <- gsub(":", " x ", coef_ltmp$Term)
coef_ltmp$sig <- ifelse(coef_ltmp$lo > 0 | coef_ltmp$hi < 0, "Significant", "Not significant")
coef_ltmp$Term <- reorder(coef_ltmp$Term, coef_ltmp$Estimate)
ggplot(coef_ltmp, aes(x = Estimate, y = Term, colour = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = lo, xmax = hi), size = 0.7, fatten = 3) +
  scale_colour_manual(values = c("Significant" = "black", "Not significant" = "grey60")) +
  labs(x = "Coefficient Estimate (logit scale)", y = NULL, colour = NULL,
       title = "LTMP Benthic — Effect Sizes (90% CI)") +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```
:::

### Dose-Response Curves

::: {.panel-tabset}
#### Bayesian Beta
