---
title: "Multidimensional Attribution of Coral Bleaching Mortality on the Great Barrier Reef"
author: 
  - name: Samuel Matthews
abstract: |
  [Abstract Placeholder]
bibliography: references.bib
format:
  html:
    toc: true
    theme: cosmo
  pdf:
    keep-tex: true
  docx: default  
---

# Introduction

Climate change is increasing the frequency and intensity of marine heatwaves, 
causing widespread coral bleaching and mortality across the Great Barrier Reef 
(GBR) [@eakin2019]. As thermal stress shifts from acute events to chronic 
disturbance, managers must transition from monitoring to targeted spatial 
interventions [@vanoppen2017]. Deploying resilience-based strategies---like 
localized solar radiation management, assisted gene flow, or culling benthic 
competitors---requires accurate attribution of coral mortality to specific thermal 
thresholds [@hughes2018; @vanoppen2017]. Treating the GBR uniformly masks the 
spatial variation in bleaching and survival observed among neighboring reefs 
[@wold2023]. Directing conservation resources at actionable scales requires 
spatially explicit models that identify the environmental and biological factors 
moderating survival during severe heatwaves.

Current diagnostic tools rely on static or coarse-scale environmental proxies 
that miss fine-scale physical and ecological mediators of thermal stress 
[@safaie2018]. Many vulnerability assessments evaluate Degree Heating Weeks (DHW) 
as an isolated, linear driver of mortality without adjusting for background coral 
recovery [@macneil2019]. This omission conflates baseline demographic turnover 
with bleaching-attributable loss, biasing mortality estimates. Most models also 
omit localized optical and hydrodynamic modifiers---such as water clarity (Secchi 
depth), cloud cover shading, and current flushing speeds---or shifts in community 
composition, such as the proportion of sensitive *Acropora* taxa 
[@cacciapaglia2016]. Deterministic forecasts trained only on historical heatwaves 
often fail under novel thermal regimes. This inflexibility led to widespread 
underprediction of mortality during the 2024 mass bleaching event, limiting the 
effectiveness of adaptive management.

We developed a growth-adjusted framework to quantify coral mortality across the 
GBR. We combined multidecadal benthic photo-transects and manta tow data with 
high-resolution environmental predictors, including winter temperature variability, 
historical thermal exposure, and acute optical modifiers. To isolate bleaching 
impacts, we adjusted baseline mortality estimates using reef-specific Gompertz 
coral growth parameters to account for expected biological recovery between 
surveys. We paired machine-learning algorithms (Random Forests and Boosted 
Regression Trees) to capture non-linear feature interactions with Bayesian 
hierarchical beta regression models to bound proportional mortality. The Bayesian 
models propagate parameter uncertainty and partition spatial variance across reef 
clusters. By explicitly modeling interactions between thermal stress and optical 
mediators---specifically how localized Secchi depth and cloud fraction buffer 
DHW---we link mortality attribution to measurable physical contexts.

We evaluated how physical context and community composition shape reef 
susceptibility to thermal stress under different exposure scenarios. We outline 
the growth-adjustment protocol, environmental feature engineering, and the 
comparative validation of our statistical and machine-learning ensembles. Using 
leave-one-event-out and spatially blocked cross-validation, we assessed model 
stability across historical bleaching events (1998--2022) and stress-tested the 
framework against the 2024 anomaly. Our analyses show that integrating water 
clarity and localized shading metrics improves mapping of regional disturbance 
trajectories and explains spatial variation in survival. Reefs in distinct optical 
and hydrodynamic contexts showed moderated mortality responses, indicating that 
historical thermal exposure and the physical environment interact to create 
localized spatial buffers. This framework provides a probabilistically bounded, 
spatially explicit tool for prioritizing resilient reef networks and designating 
conservation zones under changing climate regimes.

# Methods

## Benthic Data and Growth-Adjustment
We integrated multidecadal benthic photo-transect and broad-scale manta tow data 
across the Great Barrier Reef spanning 1998 to 2024. This expanded dataset 
($N=465$ reef-event pairs) captures the full spectrum of recent mass bleaching 
events. To isolate thermal impacts from background demographic turnover, we 
growth-adjusted baseline mortality estimates using reef-specific Gompertz coral 
growth parameters. This adjustment accounts for the biological recovery expected 
between survey intervals. We also constructed a strictly filtered subset of 
confirmed bleaching mortality ($N=305$) to evaluate model sensitivity when excluding 
compound disturbances such as cyclones and freshwater plumes.

## Environmental Covariates and Feature Engineering
We engineered a suite of high-resolution environmental and biological predictors. 
Thermal stress was parameterized using Degree Heating Weeks (DHW). To test for 
physical buffering, we integrated localized optical and hydrodynamic modifiers, 
specifically acute water clarity (Secchi depth), cloud cover fraction, and prevailing 
current speeds. We also included winter sea surface temperature (SST) variability 
and the preceding-year proportion of thermally sensitive *Acropora* taxa to account 
for community composition.

## Statistical Modeling and Validation
We applied a dual analytical architecture. We used Boosted Regression Trees (BRT) 
to map non-linear feature interactions and derive relative variable importance. We 
paired this with Bayesian hierarchical Beta and Binomial-OLRE regression models to 
estimate dose-response trajectories and enforce physical bounds on proportional 
mortality. The Bayesian framework propagates parameter uncertainty and partitions 
spatial variance across nested reef clusters. Model stability was assessed using 
leave-one-event-out and spatial sector-blocked cross-validation, and predictive 
skill was stress-tested on out-of-sample data from the extreme 2024 mass bleaching 
event.

```{r}
#| label: load-data
#| echo: false
#| message: false

library(readr)
library(dplyr)
library(knitr)

# Load comparison metrics for the summary table
metrics <- read_csv("output/formal_models/model_comparison_metrics.csv", show_col_types = FALSE)
```

# Results

## Model Validation and Variable Importance
The dual analytical framework reliably predicted proportional coral mortality 
across historical bleaching eras. Spatial sector-blocked cross-validation showed 
the Bayesian models achieved a strong predictive $R^2$ and minimized mean absolute 
error compared to isolated univariate frameworks. Boosted Regression Trees identified 
Degree Heating Weeks as the primary driver of mortality. However, community 
composition (proportion *Acropora*) and optical modifiers (Secchi depth and cloud 
cover) contributed substantial relative influence, confirming that thermal stress 
does not act in isolation.

## Dose-Response Trajectories and Environmental Buffering
Bayesian dose-response curves demonstrated that physical context strongly moderates 
reef susceptibility to thermal anomalies. High cloud cover and reduced water clarity 
(lower Secchi depth) attenuated mortality at severe DHW thresholds. This confirms 
that localized shading and turbidity buffer acute thermal exposure. Conversely, 
reefs dominated by *Acropora* in clear-water, unshaded environments experienced 
steeper mortality trajectories, matching the known physiological sensitivity of 
these taxa.

## 2024 Mass Bleaching Generalization
When stress-tested against the unprecedented 2024 mass bleaching event as a strict 
out-of-sample forecast, the framework maintained predictive accuracy. While traditional 
DHW-only linear models severely underpredicted 2024 mortality, integrating optical 
modifiers and community composition allowed the BRT and Bayesian models to map the 
extreme spatial heterogeneity observed during the anomaly.

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary of model performance metrics across validation folds."
#| echo: false

metrics %>%
  filter(scheme == "leave_one_event_out") %>%
  select(programme_key, learner, rmse, predictive_r2, bias) %>%
  head(10) %>%
  kable(digits = 3, col.names = c("Dataset", "Model", "RMSE", "Predictive R2", "Bias"))
```

# Discussion

# References

::: {#refs}
:::
