---
title: 'Experiment findings'
subtitle: 'A lightweight, reproducible decision index'
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: false
  warning: false
---

```{r setup}
registry_path <- 'config/experiments.yml'
if (!file.exists(registry_path)) registry_path <- '../config/experiments.yml'
registry <- yaml::read_yaml(registry_path)
experiments <- registry$experiments

status_labels <- c(
  selected = 'Selected',
  retained = 'Retained comparator or sensitivity',
  not_promoted = 'Tested, not promoted',
  superseded = 'Superseded',
  diagnostic = 'Diagnostic',
  data_pipeline = 'Data pipeline'
)

experiment_rows <- do.call(rbind, lapply(experiments, function(x) {
  data.frame(
    id = x$id, title = x$title, track = x$track, status = x$status,
    finding = x$finding, report = x$report, updated = as.character(x$updated),
    stringsAsFactors = FALSE
  )
}))
```

This page is generated only from `config/experiments.yml`. It answers “what has already been tested?” before detailed reports, model outputs or raw data are opened. Last registry review: `r registry$updated`.

## Current synthesis

```{r current-synthesis, results='asis'}
current <- experiments[vapply(experiments, function(x) identical(x$track, 'current'), logical(1))]
for (x in current) {
  cat('### ', x$title, '\n\n', sep = '')
  cat('**', status_labels[[x$status]], '.** ', x$finding, '\n\n', sep = '')
  cat('Detailed report: `', x$report, '`\n\n', sep = '')
}
```

## Experiment index

```{r experiment-index}
index <- experiment_rows[, c('id', 'title', 'track', 'status', 'updated')]
index$status <- unname(status_labels[index$status])
names(index) <- c('ID', 'Experiment', 'Track', 'Decision', 'Updated')
knitr::kable(index, row.names = FALSE)
```

## Reproducible findings blocks

Each block is intentionally brief. The report contains methods and diagnostics; the evidence paths identify the smallest set of generated artefacts needed to verify the finding.

```{r findings-blocks, results='asis'}
for (x in experiments) {
  evidence <- unlist(x$evidence, use.names = FALSE)
  available <- file.exists(evidence)
  evidence_text <- paste0(
    '`', evidence, '`',
    ifelse(available, ' (available locally)', ' (generated artefact not present)')
  )
  cat('<details>\n<summary><strong>', x$title, '</strong> — ',
      status_labels[[x$status]], '</summary>\n\n', sep = '')
  cat('- **ID:** `', x$id, '`\n', sep = '')
  cat('- **Question:** ', x$question, '\n', sep = '')
  cat('- **Finding:** ', x$finding, '\n', sep = '')
  cat('- **Report:** `', x$report, '`\n', sep = '')
  cat('- **Evidence:** ', paste(evidence_text, collapse = '; '), '\n', sep = '')
  cat('- **Reviewed:** ', as.character(x$updated), '\n\n', sep = '')
  cat('</details>\n\n')
}
```

## Adding an experiment

Add or update the detailed analysis, then add one compact record to `config/experiments.yml`. Run:

```powershell
Rscript src/evaluation/validate_experiment_registry.R
quarto render analysis/exploration.qmd
```

Evidence availability is informative rather than a validation failure because generated `output/` files are deliberately not committed.
